### 2019-01-22 paper/ICPC19/icpc19.tex
paper: sections, dataset description, first barchart

\documentclass[10pt,conference]{IEEEtran} 
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{booktabs}
\usepackage[table,xcdraw]{xcolor}
\usepackage{xcolor}
\definecolor{mypink3}{cmyk}{0, 0.7808, 0.4429, 0.1412}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\begin{document}

\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code\\
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{
    Anonymou(s) Author(s)
%     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
% \and
% \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address}
}

\maketitle

\begin{abstract}

Lorem Ipsum is simply dummy text of the printing and typesetting industry. 
Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, 
when an unknown printer took a galley of type and scrambled it to make a type 
specimen book. It has survived not only five centuries, but also the leap into 
electronic typesetting, remaining essentially unchanged. It was popularised in 
the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, 
and more recently with desktop publishing software like Aldus PageMaker including 
versions of Lorem Ipsum. It was popularised in the 1960s with the release of 
Letraset sheets containing Lorem Ipsum passages, and more recently with desktop 
publishing software like Aldus PageMaker including versions of Lorem Ipsum.

\end{abstract}

\begin{IEEEkeywords}
Security, Maintainability, Open-Source
\end{IEEEkeywords}

\section{Introduction}

Add a point for motivation with one of the examples of the poster.

\section{Related Work}

\section{Methodology}

\subsection{Dataset}

\begin{table*}[h]
\centering
\caption{Descriptive statistics of the dataset projects\textcolor{mypink3}{@update}} \label{tab:dataset} 
\begin{tabular}{@{}ccccccccccc@{}}
\toprule
     & forks   & stars   & watchers & contributors & commits  & branches & releases & size      & issues  & pull requests \\ \midrule
Mean & 1734.79 & 5262.97 & 395.11   & 152.48       & 22257.82 & 44.08    & 126.31   & 143243.64 & 3725.4  & 1976.92       \\
Min     & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0       & 0             \\
25 \%     & 374.75  & 1571.5  & 117.75   & 48           & 1347     & 4        & 22       & 8339.5    & 321.25  & 151.75        \\
Median     & 887     & 2836.5  & 258.5    & 102.5        & 5823     & 9        & 59       & 38331.5   & 1652.5  & 515.5         \\
75 \%     & 2196    & 6455.75 & 462.5    & 259          & 22786.25 & 22.25    & 142.25   & 164442.5  & 4152.75 & 1944.25       \\
Max     & 16366   & 31841   & 3446     & 413          & 717449   & 1227     & 1114     & 2042017   & 33970   & 19329         \\
Total     & 180418  & 547349  & 41091    & 15858        & 2314813  & 4584     & 13136    & 14897339  & 387442  & 205600        \\ \bottomrule
\end{tabular}
\end{table*}

\subsection{Security vs. Baseline Commits}

\subsection{Maintainability}

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.5\textwidth]{figures/maintainability.pdf}
 	\caption{Maintainability difference for security and baseline commits \textcolor{mypink3}{@update}}
\end{figure}

\section{Results}

\section{Discussion}

\section{Threats to Validity}

\section{Conclusion}

\section*{Acknowledgment}

{
  \bibliography{paper}
}

\end{document}

### 2019-01-23 paper/ICPC19/icpc19.tex
paper
paper running
abstract --> sofia, add numbers and review

@@
 \usepackage[table,xcdraw]{xcolor}
+\usepackage{framed}
 \usepackage{xcolor}
@@
 
-\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code\\
-\thanks{Identify applicable funding agency here. If none, delete this.}
+\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code%\\
+%\thanks{Identify applicable funding agency here. If none, delete this.}
 }
@@
 \begin{abstract}
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry. 
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, 
-when an unknown printer took a galley of type and scrambled it to make a type 
-specimen book. It has survived not only five centuries, but also the leap into 
-electronic typesetting, remaining essentially unchanged. It was popularised in 
-the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, 
-and more recently with desktop publishing software like Aldus PageMaker including 
-versions of Lorem Ipsum. It was popularised in the 1960s with the release of 
-Letraset sheets containing Lorem Ipsum passages, and more recently with desktop 
-publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
+  Security is a crucial non-functionality requirement for software applications.
+  However, building secure software is far from trivial as developers lack both
+  the knowledge and tools to effectively address this concern. In this paper, we
+  study the impact of changes to improve security on the maintainability of several
+  open source applications. Using a dataset containing XXX security-oriented
+  commits, we measure maintainability --- as computed by the Software Improvement
+  Group's web-based source code analysis service \emph{Better Code Hub} (BCH) ---
+  before and after the security refactoring. Results show that making software
+  more secure comes at a cost on  maintainability. This is particularly evident
+  in refactorings to deal with \emph{XXX} and \emph{YYY} attacks.
+  Furthermore, we have found evidence that security-related changes are more
+  likely to be modified in the future than regular code changes.
 \end{abstract}
@@
 
-Add a point for motivation with one of the examples of the poster.
+\begin{framed}
+\textit{\textbf{RQ1} What is the impact }
+\end{framed}
 
@@
  	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/maintainability.pdf}
+ 	\includegraphics[width=0.4\textwidth]{figures/maintainability.pdf}
  	\caption{Maintainability difference for security and baseline commits \textcolor{mypink3}{@update}}
 \end{figure}
+
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.55\textwidth]{figures/category.pdf}
+ 	\caption{Maintainability difference by type of security commit \textcolor{mypink3}{@update}}
+\end{figure}
+
+\cite{Wang:2008:DSJ:1330017.1330021}
 
@@
 {
-  \bibliography{paper}
+ \bibliographystyle{IEEEtran}
+  \bibliography{icpc19}
 }

### 2019-01-23 paper/ICPC19/icpc19.tex
merge

--- 2019-01-22 paper/ICPC19/icpc19.tex
@@
 
-Add a point for motivation with one of the examples of the poster.
+The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability,
+functional suitability, performance efficiency, compatibility, usability, reliability, security,
+and portability. This paper focuses solely on maintainability.
+
+\begin{framed}
+\textit{\textbf{RQ1} What is the impact of fixing vulnerabilities to improve security }
+\end{framed}
 

### 2019-01-23 paper/ICPC19/icpc19.tex

--- 2019-01-23 paper/ICPC19/icpc19.tex
@@
 
+The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability,
+functional suitability, performance efficiency, compatibility, usability, reliability, security,
+and portability. This paper focuses solely on maintainability.
+
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact }
+\textit{\textbf{RQ1} What is the impact of fixing vulnerabilities to improve security }
 \end{framed}

### 2019-01-23 paper/ICPC19/icpc19.tex
abstrasct and keywords

@@
   study the impact of changes to improve security on the maintainability of several
-  open source applications. Using a dataset containing XXX security-oriented
+  open source applications. Using a dataset containing 643 security-oriented
   commits, we measure maintainability --- as computed by the Software Improvement
@@
   more secure comes at a cost on maintainability. This is particularly evident
-  in refactorings to deal with \emph{XXX} and \emph{YYY} attacks.
+  in refactorings to deal with Broken Authentication and Cross-Site Request Forgery attacks.
   Furthermore, we have found evidence that security-related changes are more
@@
 \begin{IEEEkeywords}
-Security, Maintainability, Open-Source
+Security, Software Maintenance, Open-Source Software
 \end{IEEEkeywords}

### 2019-01-23 paper/ICPC19/icpc19.tex
update paper + cleaning

@@
   more secure comes at a cost on maintainability. This is particularly evident
-  in refactorings to deal with Broken Authentication and Cross-Site Request Forgery attacks.
+  in refactorings to deal with \textit{Broken Authentication} and \textit{Cross-Site Request Forgery} attacks.
   Furthermore, we have found evidence that security-related changes are more
@@
 
+\section{Motivation and Research Questions}
+
+\textcolor{mypink3}{Ver que exemplos existem na amostra da secbench que vamos utilizar para dar como exemplo de motivacao}
+
 \begin{framed}
@@
 \end{framed}
-
-\section{Related Work}
 
@@
 
+\cite{Wang:2008:DSJ:1330017.1330021}
+
+\section{Results}
+
 \begin{figure}[h]
@@
 
-\cite{Wang:2008:DSJ:1330017.1330021}
-
-\section{Results}
-
 \section{Discussion}
@@
 
-\section{Conclusion}
+\section{Related Work}
+
+\section{Conclusion and Future Work}
 
 \section*{Acknowledgment}
+
+Better Code Hub?
 

### 2019-01-23 paper/ICPC19/icpc19.tex
RQs draft

@@
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact of fixing vulnerabilities to improve security }
+\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability of open-source software?}
+\end{framed}
+
+\begin{framed}
+\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to affect open-source software maintainability?}
+\end{framed}
+
+\begin{framed}
+\textit{\textbf{RQ3} }
 \end{framed}

### 2019-01-23 paper/ICPC19/icpc19.tex

@@
   in refactorings to deal with \textit{Broken Authentication} and \textit{Cross-Site Request Forgery} attacks.
-  Furthermore, we have found evidence that security-related changes are more
-  likely to be modified in the future than regular code changes.
+  \textcolor{mypink3}{Furthermore, we have found evidence that security-related changes are more
+  likely to be modified in the future than regular code changes.}
 \end{abstract}
@@
 
+In this chapter, the methodology used to measure the impact of security refactorings on the maintainability of open-source software is presented in depth in the following sections and illustrated in Figure \ref{fig:met}.
+Aiming to answer our research questions, we use a dataset containing more than 600 security refactorings collected in a previous study from open-source software publicly available on GitHub.
+We use as the baseline a collection of randomly chosen regular commits from a list of projects retrieved from the main dataset.
+Then, we Better Code Hub to collect the maintainability reports for both security and regular commits.
+
+%\footnote{SIG's website is available at https://www.softwareimprovementgroup.com/ (Accessed on January 23, 2019)}
+
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
+ 	\caption{Maintainability difference for security and baseline commits \textcolor{mypink3}{@update}}
+	\label{fig:met}
+\end{figure}
+
 \subsection{Dataset}
+
+To evaluate the impact of security refactorings on the maintainability of open-source software, we use a dataset of security flaws provided by a study conducted in 2017 \cite{Reis:2017:IJSSE} where the authors mined more than 248 GitHub repositories for several different patterns (e.g., injection, cross-site scripting, cross-site request forgery, memory leaks).
+Reis and Abreu (2017) mined open-source software aiming the extraction of real test cases - created by real developers on their daily basis development - to test and assess the performance of static analysis tools since using hand-seeded test cases or mutations could lead to misleading assessments of the tools' capabilities.
+The study yielded to a dataset of more than 600 test cases for 16 different patterns.
+Each test case of the dataset is a triplet of folders: the commit before the refactoring, the commit responsible for the refactoring and the snippets of code that differ from one version to another (usually, called \textit{diff}) - where one can easily identify the code used to fix the security flaw.
+In this study, we focus on computing the maintainability of the commits before and after the security refactoring to calculate if the impact was positive, negative or none.
 
@@
 
-\cite{Wang:2008:DSJ:1330017.1330021}
-
 \section{Results}
@@
 \end{figure}
+
+\begin{framed}
+\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability of open-source software?}
+\end{framed}
+
+\begin{framed}
+\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to affect open-source software maintainability?}
+\end{framed}
 

### 2019-01-24 paper/ICPC19/icpc19.tex
paper update

--- 2019-01-23 paper/ICPC19/icpc19.tex
@@
 
+In this section, the methodology used to measure the impact of security refactorings on the maintainability of open-source software is presented in depth in the following sections and illustrated in Figure \ref{fig:met}.
+Aiming to answer the research questions presented in this paper, we use a dataset containing more than 700 security refactorings collected from open-source software available on GitHub and where 648 were reported in a previous study conducted in 2017 \cite{Reis:2017:IJSSE}.
+In parallel, a baseline of regular commits was randomly collected from the list of projects of the main dataset to evaluate what is the impact of regular commits on the maintainability of open-source software.
+The Software Improvement Group's\footnote{SIG's website is available at https://www.softwareimprovementgroup.com/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.
+
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
+ 	\caption{Study Methodology}
+	\label{fig:met}
+\end{figure}
+
 \subsection{Dataset}
+
+To evaluate the impact of security refactorings on the maintainability of open-source software, we use a dataset of security flaws which is the outcome of mining more than 248 GitHub\footnote{Github's website available at https://github.com/ (Accessed on January 23, 2019)} repositories.
+Reis and Abreu (2017) mined open-source software aiming the extraction of real test cases - created by real developers on their daily basis work - to test and assess the performance of static analysis tools since using hand-seeded test cases or mutations could lead to misleading assessments of the tools' capabilities.
+The study yielded to a dataset of more than 600 test cases for 16 different patterns.
+Each test case of the dataset is a triplet of folders: the commit before the refactoring, the commit responsible for the refactoring and the snippets of code that differ from one version to another (usually, called \textit{diff}) - where one can easily identify the code used to fix the security flaw.
+In this study, we focus on computing the maintainability of the commits before and after the security refactoring to evaluate if the impact was positive, negative or none.
+
+\begin{table}[h]
+	\centering
+\caption{Current Dataset Patterns Distribution} \label{tab:patterns}
+\begin{tabular}{@{}ll@{}}
+\toprule
+Pattern & \# Commits\\
+\midrule
+Injection & 101\\
+Broken Authentication and Session Management& 45\\
+Cross-Site Scripting& 142\\
+Broken Access Control& 2\\
+Security Misconfiguration& 9\\
+Sensitive Data Exposure& 24\\
+Insufficient Attack Protection& 18\\
+Cross-Site Request Forgery& 34\\
+Using Components With Known Vulnerabilities& 27\\
+Unprotected APIs& 7\\
+Memory Leak& 91\\
+Overflow& 19\\
+Denial-of-Service& 43\\
+Path Traversal& 19\\
+Resource Leak& 33\\
+SHA-1 Hash Function & 1\\
+Miscellaneous& 101\\\midrule
+Total& 716\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Reis and Abreu (2017) dataset commits were manually validated, flagged as security flaws and classified with one of 16 the following patterns: \textit{Injection}, \textit{Broken Authentication and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery}, \textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs}, \textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service}, \textit{Path Traversal} and \textit{Miscellaneous}.
+However, the dataset evolved and now integrates new patterns, such as \textit{Resource Leak} and \textit{SHA-1 Hash Function}.
+
+In total, the dataset contains 716 security refactorings, as shown in Table \ref{tab:patterns}. This is the dataset used
+by the BCH tool to compute the maintainability of security refactorings. However, some of the projects are not considered for
+the final results because the limitations of the tool, such as the project size and large time of computation.
+To calculate the significance of our results by pattern, we perform the wilcoxon statistical test - described in more depth in subsection \ref{sec:statsval} - which needs to comprise more than 20 commits to be effective.
+The patterns not comprising the number of necessary commits were reallocated to the \textit{Miscellaneous} pattern.
+Due to the tool limitations and the statistical test requirements, this study only distinguishes nine of the patterns presented in Table \ref{tab:patterns}.
+In the
+end, we used the following categorization:
+
+\begin{itemize}
+	\item \textbf{Using Components with Known Vulnerabilities.} Updating the libraries, frameworks and other modules.
+
+	Components, such as libraries, frameworks, and other software modules, run with the same
+	privileges as the application. If a vulnerable component is exploited, such an attack can facilitate
+	serious data loss or server takeover.
+
+	\item \textbf{Resource Leak}
+	\item \textbf{Broken Authentication \& Session Management}
+	\item \textbf{Cross-Site Request Forgery}
+	\item \textbf{Denial-of-Service}
+	\item \textbf{Miscellaneous}
+	\item \textbf{Cross-Site Scripting}
+	\item \textbf{Injection}
+	\item \textbf{Memory Leak}
+\end{itemize}
 
@@
 
-\subsection{Maintainability}
+\subsection{Maintainability Analysis}
 
-\cite{Wang:2008:DSJ:1330017.1330021}
+BCH is used to collect the maintanability reports of the commits of each project. The tool 
+evaluation is based on a model of 10 different guidelines \cite{Visser:2016:OREILLY}:
+
+\begin{itemize}
+	\item \textbf{Write short units of code.} Long units are hard to test, reuse, and understand.
+	\item \textbf{Write simple units of code.} Keeping the number of branch points low makes units easier to modify and test.
+	\item \textbf{Write code once.} When code is duplicated, bugs need to be fixed in multiple places, which is inefficient and
+prone to errors.
+	\item \textbf{Keep unit interfaces small.} Keeping the number of parameters low makes units easier to understand and reuse.
+	\item \textbf{Separate concerns in modules.} Changes in a loosely
+coupled codebase are much easier to oversee and execute
+than changes in a tightly coupled codebase.
+	\item \textbf{Couple architecture components loosely.} Independent
+components ease isolated maintenance.
+	\item \textbf{Keep architecture components balanced.} Balanced
+components ease locating code and foster isolation,
+improving maintenance activities.
+	\item \textbf{Keep your codebase small.} Small systems are easier to
+search through, analyze, and understand code.
+	\item \textbf{Automate tests.} Automated testing makes development
+predictable and less risky.
+	\item \textbf{Write clean code.} Code without code smells is less likely to bring maintainability issues.
+\end{itemize}
+
+For each guideline, BCH evaluates the compliance against a particular guideline by setting boundaries for the percentage of code allowed to fall in each of the four risk severity categories (low risk, medium risk, high risk, and very high risk).
+If the thresholds are not violated, the project is considered to be compliant with the guideline.
+According to BCH, the guideline thresholds are calibrated yearly based on a representative benchmark of closed and open-source software systems.
+Being compliant with a guideline means that the project under analysis is at least better than 65\% of the software systems in BCH’s benchmark.
+
+The BCH report of the $<INSERT\_EXEMPLO>$ for a non-compliant guideline can be seen in Fig.
+[X].
+This was extracted from the report of the app $<INSERT\_EXEMPLO>$, used in the motivating example of Section II [MAYBE?].
+The green bar represents the percentage of compliant lines of code.
+These lines of code are considered to be compliant with ISO 25010 standard for maintainability [21].
+The yellow, orange and red bars represent non-compliant lines of code with medium, high, and very high severity levels, respectively.
+Along the bars, there are also marks that refer to the compliance thresholds for each severity level.
+The report is equivalent to the information reported in Table II: a set of thresholds, number of lines of code (LOC), and percentage of the project for each severity level.
+Nonetheless, thresholds provided by BCH do not sum to 100\%: non-compliant levels are provided in a cumulative way (e.g., the threshold for the medium level includes high and very high levels); the compliant-level threshold is the complement of the medium-level threshold.
+
+Since we want to analyze maintainability regression, we
+use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the energy commit ($v_{E−1}$) and b) the version immediately after the energy commit ($v_E$).
+This is illustrated in Fig.
+[X].
+Although BCH provides a detailed report of the maintainability of the project, it does not compute a final score that we can use to compare maintainability amongst different projects.
+Thus, based in previous work CITE?, we designed an equation to capture the distance between the current state of the project and the standard thresholds.
+We have adjusted the equation to meet the following requirements:
+
+\begin{itemize}
+	\item \textbf{The maintainability difference between two versions of the same project is not affected by its size.} In this work, we want to evaluate the identical energy patterns occurring in different projects.
+Thus, the metric cannot use normalization based on its size – we convert percentage data to the respective number of lines of code.
+	\item \textbf{Distance to the thresholds in high severity levels is more penalized than in low severity levels.} We use weights based on the severity level to count lines of code that violate maintainability guidelines.
+\end{itemize}
+
+We compute the mean average of the maintainability score $M(v)$ for all the selected guidelines, as follows:
+
+\begin{equation}
+    M(v) = \sum_{g \in G}^{} M_{g}(v)
+\end{equation}
+
+where $G$ is the group of selected maintainability guidelines from BCH (e.g., Write short units of code, etc.) and $v$ is the
+version of the software under evaluation. The maintenance $M$ based on the guideline $g$ for a given version of a project is computed with the following equation:
+
+\begin{equation}
+    M_{g} = \frac{1}{|L|} \sum_{l \in L}^{} C(l) , L = \{medium, high, veryHigh\}
+\end{equation}
+
+where $C$ is the compliance with the maintainability guideline for the given severity level (medium, high, and very high) and
+$L$ is the group of severity levels of maintainability infractions. The compliance $C$ for a given severity level $l$ is derived by:
+
+\begin{equation}\label{eq:3}
+    C(l) = LOC_{compliant}(l) - w(l) * LOC_{\neg compliant}(l)
+\end{equation}
+
+where $LOC_{compliant}(l)$ are the lines of code that comply with the guideline at the given severity level $l$, $LOC_{\neg compliant}(l)$ are the lines of code that do not comply with the guideline at the given
+severity level $l$ and $w(l)$ is the weight factor to boost the impact of
+non-compliant lines in comparison to compliant lines. Finally, the term $w(l)$ is calculated as follows:
+
+\begin{equation}
+    w(l) = \frac{1 - T(l)}{T(l)}
+\end{equation}
+
+where $T(l)$ is the threshold in percentage of the lines of code that are accepted to be non-compliant with the guideline for the severity level $l$.
+This is a standard value defined by BCH.
+In other words, the factor $w$ is used in Eq.
+\ref{eq:3} to highlight the lines of code that are not complying with the guideline.
+For instance, the threshold for the severity level veryHigh is defined in Table II as T(veryHigh) = 6.9\%, which derives to a weight of w(veryHigh) = 13.5.
+This means that, in this example, one non-compliant guideline is decreasing maintainability score by 13.5 points while a compliant guideline is increasing by 1.0 point.
+In addition, a version that is perfectly aligned with the standard thresholds has a maintainability score of zero.
+Then, we compute the difference of maintainability between the energy commit ($v_E$) and its parent commit ($v_{E−1}$), as illustrated in Fig.
+[X].
+
+\subsection{Statistical Validation}\label{sec:statsval}
+
+To validate the maintainability differences in different groups of commits (e.g., baseline and energy commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
+In other words, we test the null hypothesis that the maintainability difference between pairs of versions vE−1, vE (i.e., before and after an energy-commit) follows a symmetric distribution around 0.
+To understand the effect-size, as advocated by the Common-language effect sizes [X], we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
+
+\textcolor{mypink3}{@todo: Discuss pratt improvement and accuracy}
 
@@
  	\centering
- 	\includegraphics[width=0.4\textwidth]{figures/maintainability.pdf}
+ 	\includegraphics[width=0.45\textwidth]{figures/maintainability.pdf}
  	\caption{Maintainability difference for security and baseline commits \textcolor{mypink3}{@update}}

### 2019-01-24 paper/ICPC19/icpc19.tex
fixing unicodes..

@@
 Since we want to analyze maintainability regression, we
-use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the energy commit ($v_{E−1}$) and b) the version immediately after the energy commit ($v_E$).
+use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the energy commit ($v_{E-1}$) and b) the version immediately after the energy commit ($v_E$).
 This is illustrated in Fig.
@@
 In addition, a version that is perfectly aligned with the standard thresholds has a maintainability score of zero.
-Then, we compute the difference of maintainability between the energy commit ($v_E$) and its parent commit ($v_{E−1}$), as illustrated in Fig.
+Then, we compute the difference of maintainability between the energy commit ($v_E$) and its parent commit ($v_{E-1}$), as illustrated in Fig.
 [X].
@@
 To validate the maintainability differences in different groups of commits (e.g., baseline and energy commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
-In other words, we test the null hypothesis that the maintainability difference between pairs of versions vE−1, vE (i.e., before and after an energy-commit) follows a symmetric distribution around 0.
+In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{E-1}$, $v_E$ (i.e., before and after an energy-commit) follows a symmetric distribution around 0.
 To understand the effect-size, as advocated by the Common-language effect sizes [X], we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.

### 2019-01-24 paper/ICPC19/icpc19.tex
baseline + latex fixed

--- 2019-01-24 paper/ICPC19/icpc19.tex
@@
 \begin{itemize}
-	\item \textbf{Using Components with Known Vulnerabilities.} Updating the libraries, frameworks and other modules.
-
-	Components, such as libraries, frameworks, and other software modules, run with the same
-	privileges as the application. If a vulnerable component is exploited, such an attack can facilitate
-	serious data loss or server takeover.
+	\item \textbf{Using Components with Known Vulnerabilities.}
 
@@
 \subsection{Security vs. Baseline Commits}
+
+The baseline dataset uses the security commits dataset as input.
+For each security commit, one random commit is selected from the list of all commits of the belonging project.
+We originate the regular commits from the security commits to ensure that differences in maintainability are not consequence of characteristics
+of different projects.
 
@@
 
-To validate the maintainability differences in different groups of commits (e.g., baseline and energy commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
+To validate the maintainability differences in different groups of commits (e.g., baseline and security commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
 In other words, we test the null hypothesis that the maintainability difference between pairs of versions vE−1, vE (i.e., before and after an energy-commit) follows a symmetric distribution around 0.

### 2019-01-24 paper/ICPC19/icpc19.tex
cleaning

@@
 Since we want to analyze maintainability regression, we
-use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the energy commit ($v_{E−1}$) and b) the version immediately after the energy commit ($v_E$).
+use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the security commit ($v_{E−1}$) and b) the version immediately after the security commit ($v_E$).
 This is illustrated in Fig.
@@
 \begin{itemize}
-	\item \textbf{The maintainability difference between two versions of the same project is not affected by its size.} In this work, we want to evaluate the identical energy patterns occurring in different projects.
+	\item \textbf{The maintainability difference between two versions of the same project is not affected by its size.} In this work, we want to evaluate the identical security patterns occurring in different projects.
 Thus, the metric cannot use normalization based on its size – we convert percentage data to the respective number of lines of code.
@@
 In addition, a version that is perfectly aligned with the standard thresholds has a maintainability score of zero.
-Then, we compute the difference of maintainability between the energy commit ($v_E$) and its parent commit ($v_{E−1}$), as illustrated in Fig.
+Then, we compute the difference of maintainability between the security commit ($v_E$) and its parent commit ($v_{E−1}$), as illustrated in Fig.
 [X].
@@
 To validate the maintainability differences in different groups of commits (e.g., baseline and security commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
-In other words, we test the null hypothesis that the maintainability difference between pairs of versions vE−1, vE (i.e., before and after an energy-commit) follows a symmetric distribution around 0.
+In other words, we test the null hypothesis that the maintainability difference between pairs of versions vE−1, vE (i.e., before and after an security-commit) follows a symmetric distribution around 0.
 To understand the effect-size, as advocated by the Common-language effect sizes [X], we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.

### 2019-01-24 paper/ICPC19/icpc19.tex
update
fixing

@@
 	\item \textbf{Using Components with Known Vulnerabilities.}
-
 	\item \textbf{Resource Leak}
@@
 Since we want to analyze maintainability regression, we
-use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the security commit ($v_{E−1}$) and b) the version immediately after the security commit ($v_E$).
+use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the security commit ($v_{E-1}$) and b) the version immediately after the security commit ($v_E$).
 This is illustrated in Fig.
@@
 In addition, a version that is perfectly aligned with the standard thresholds has a maintainability score of zero.
-Then, we compute the difference of maintainability between the security commit ($v_E$) and its parent commit ($v_{E−1}$), as illustrated in Fig.
+Then, we compute the difference of maintainability between the security commit ($v_E$) and its parent commit ($v_{E-1}$), as illustrated in Fig.
 [X].
@@
 To validate the maintainability differences in different groups of commits (e.g., baseline and security commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
-In other words, we test the null hypothesis that the maintainability difference between pairs of versions vE−1, vE (i.e., before and after an security-commit) follows a symmetric distribution around 0.
+In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{E-1}$, $v_E$ (i.e., before and after an security-commit) follows a symmetric distribution around 0.
 To understand the effect-size, as advocated by the Common-language effect sizes [X], we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
@@
 \section{Results}
-
+a
 \begin{figure}[h]

### 2019-01-24 scripts/get_maintainability.py
updating scripts

@@ -5,6 +5,20 @@ import collections
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
+from scipy.stats import wilcoxon
+
+def _print_hypothesis_test(differences):
+    """Paired Wilcoxon signed-rank test (N should be > 20)"""
+    #diff = pd.DataFrame({'diff': differences})
+    test, pvalue = _hypothesis_test(differences)
+    #print(diff.median())
+    stats = {'test': [test], 'med':[pd.DataFrame({'diff': differences})['diff'].median()],'pvalue': [pvalue]}
+    df = pd.DataFrame(stats)
+    print("Wilcoxon signed-rank test {}, p-value: {}".format(test,pvalue))
+    return df
+
+def _hypothesis_test(differences):
+    return wilcoxon(x=differences, zero_method="pratt")
 
 def readBCHCache(path):
     return bch.BCHCache(path)
@@ -35,15 +49,52 @@ def count(t_id, id, total, lang, pattern, year, l, p, y):
     pattern[p][id] += 1
     year[y][id] += 1
 
-def total_main_barchart(total, graphics):
-    d = {'count' : pd.Series([total['neg'], total['pos'], total['nul']]),
-        'type' : pd.Series(['neg', 'pos', 'nul'])}
+def total_main_barchart(total_sec, total_reg, graphics, result):
+    
+    sec = sum([value for key,value in total_sec.items()])
+    reg = sum([value for key,value in total_reg.items()]) 
+    
+    d = {'neg' : pd.Series([total_reg['neg']/reg, total_sec['neg']/sec]),
+        'pos' : pd.Series([total_reg['pos']/reg, total_sec['pos']/sec]),
+        'none' : pd.Series([total_reg['nul']/reg, total_sec['nul']/sec]),
+        'type' : pd.Series(['Regular Change','Security Change'])}
+
+    # d = {'neg' : pd.Series([total_reg['neg'], total_sec['neg']]),
+    #     'pos' : pd.Series([total_reg['pos'], total_sec['pos']]),
+    #     'none' : pd.Series([total_reg['nul'], total_sec['nul']]),
+    #     'type' : pd.Series(['Regular Change','Security Change'])}
+    
+    index = np.arange(len(d['type']))
+    
     df = pd.DataFrame(d)
-    plt.bar(df['type'], df['count'], align='center', width=0.5, color=['red', 'green', 'orange'])
-    plt.title("Secbench")
[…]

### 2019-01-24 scripts/histogram.py

@@ -1,0 +1,57 @@
+import maintainability.better_code_hub as bch
+import argparse
+import csv
+import collections
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+
+    
+    
+
+def main(projects):    
+    df = pd.read_csv(projects)
+    
+    lang_trans = {'objc':'Objective-C', 'php':'PHP', 'ruby':'Ruby', 'c':'C', 'c++':'C++', 'groovy':'Groovy', 'javascript':'JavaScript',
+                    'python':'Python', 'java':'Java', 'objc++':'Objective-C++', 'scala':'scala', 'swift':'Swift', 'smarty':'Smarty'}
+    
+    
+    dic = {}
+    for i in df['language']:
+        if i not in dic:
+            dic[i] = 1
+        else:
+            dic[i] += 1
+    
+    dic_lang = {'lang':[], 'no':[]}
+    for i in dic:
+        if i == 'others':
+            no = dic['others']
+        else:
+            dic_lang['lang'].append(lang_trans[i])
+            dic_lang['no'].append(dic[i]/len(df['language']))
+    
+    dic_lang['lang'].append('Others')
+    dic_lang['no'].append(dic['others']/len(df['language']))
+    
+    lang = pd.DataFrame(dic_lang)
+    
+    ax = lang['no'].plot(kind='bar', title ="Language Distribution", figsize=(10, 7), fontsize=11, color="#2541F0")
+    ax.set_xlabel("Language", fontsize=11)
+    ax.set_ylabel("", fontsize=11)
+    vals = ax.get_yticks()
+    ax.set_yticklabels(['{:,.0%}'.format(x) for x in vals])
+    ax.set_xticklabels(dic_lang['lang'], minor=False)
+    plt.subplots_adjust(left=0.1, right=0.9, top=0.9, bottom=0.3)
+    plt.gca().yaxis.grid(True, linestyle='--')
+    
+    plt.savefig('../paper/ICPC19/figures/language_dist.pdf')
+    
+
+if __name__ == "__main__":
+    
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--projects-csv',metavar='projects-csv',required=True,help='the projects filename')
+
+    args = parser.parse_args()
+    main(projects=args.projects_csv)

### 2019-01-24 scripts/histogram.py
check todo

@@ -36,7 +36,7 @@ def main(projects):
     
     lang = pd.DataFrame(dic_lang)
     
-    ax = lang['no'].plot(kind='bar', title ="Language Distribution", figsize=(10, 7), fontsize=11, color="#2541F0")
+    ax = lang['no'].plot(kind='bar', figsize=(10, 7), fontsize=11, color="#2541F0")
     ax.set_xlabel("Language", fontsize=11)
     ax.set_ylabel("", fontsize=11)
     vals = ax.get_yticks()

### 2019-01-24 scripts/get_maintainability.py
problem with pair of security and baseline fixed + new reports

@@ -9,9 +9,7 @@ from scipy.stats import wilcoxon
 
 def _print_hypothesis_test(differences):
     """Paired Wilcoxon signed-rank test (N should be > 20)"""
-    #diff = pd.DataFrame({'diff': differences})
     test, pvalue = _hypothesis_test(differences)
-    #print(diff.median())
     stats = {'test': [test], 'med':[pd.DataFrame({'diff': differences})['diff'].median()],'pvalue': [pvalue]}
     df = pd.DataFrame(stats)
     print("Wilcoxon signed-rank test {}, p-value: {}".format(test,pvalue))
@@ -136,21 +134,42 @@ def analysis(dataset, results, CACHE):
             for l in lines:
                 data = l.split(',')
                 owner, proj, sha, sha_p, lang, pattern, year = l.split(',')[0:7]
-                        
+                       
                 info_f = CACHE.get_stored_commit_analysis(owner, proj, sha)
                 info_p = CACHE.get_stored_commit_analysis(owner, proj, sha_p)
 
                 if info_f is None or info_p is None:
                     none+=1
+                    writer.writerow({'owner' : owner, 'project': proj,
+                                    'sha': sha, 'sha-p': sha_p,
+                                    'main(sha)': 'error',
+                                    'main(sha-p)': 'error',
+                                    'main(diff)': 'error'
+                                    })
                     continue
                 if info_f.get('error') or info_p.get('error'):
                     error+=1
+                    writer.writerow({'owner' : owner, 'project': proj,
+                                    'sha': sha, 'sha-p': sha_p,
+                                    'main(sha)': 'error',
+                                    'main(sha-p)': 'error',
+                                    'main(diff)': 'error'
+                                    })
                     continue
                 
-                #print(owner, proj, sha, sha_p)
-            
-                main = bch.compute_maintainability_score(info_f)
-                main_p = bch.compute_maintainability_score(info_p)
+                try:
+                    main = bch.compute_maintainability_score(info_f)
+                    main_p = bch.compute_maintainability_score(info_p)
+                except ZeroDivisionError as e:
+                    writer.writerow({'owner' : owner, 'project': proj,
[…]

### 2019-01-25 paper/ICPC19/icpc19.tex
language chart
adding more to S3.B
check todo
paper update
shorten SIG's website

@@
 In parallel, a baseline of regular commits was randomly collected from the list of projects of the main dataset to evaluate what is the impact of regular commits on the maintainability of open-source software.
-The Software Improvement Group's\footnote{SIG's website is available at https://www.softwareimprovementgroup.com/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.
+The Software Improvement Group's\footnote{SIG's website is available at https://www.sig.eu/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.
 
@@
 
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
+ 	\caption{Language Distribution}
+	\label{fig:lang}
+\end{figure}
+
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
+ 	\caption{Type Distribution}
+	\label{fig:type}
+\end{figure}
+
 Reis and Abreu (2017) dataset commits were manually validated, flagged as security flaws and classified with one of 16 the following patterns: \textit{Injection}, \textit{Broken Authentication and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery}, \textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs}, \textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service}, \textit{Path Traversal} and \textit{Miscellaneous}.
@@
 \subsection{Security vs. Baseline Commits}
+%
+Although we want to assess the maintainability of security-related commits,
+\textcolor{mypink3}{there is no evidence in previous work on the impact of \textit{regular} commits
+in the maintainability of projects}. Thus, we analyze the maintainability of
+regular commits and use them as a baseline.
 
-The baseline dataset uses the security commits dataset as input.
-For each security commit, one random commit is selected from the list of all commits of the belonging project.
-We originate the regular commits from the security commits to ensure that differences in maintainability are not consequence of characteristics
-of different projects.
-
+The baseline dataset uses the security commits dataset as input. For each
+security commit, one random commit is selected from the list of all commits of
+the belonging project. We originate the regular commits from the security
+commits to ensure that differences in maintainability are not consequence of
+characteristics of different projects.
+%
 \subsection{Maintainability Analysis}

### 2019-01-25 paper/ICPC19/icpc19.tex
final data reports and patterns

--- 2019-01-24 paper/ICPC19/icpc19.tex
@@
   study the impact of changes to improve security on the maintainability of several
-  open source applications. Using a dataset containing 643 security-oriented
+  open source applications. Using a dataset containing $607$ security-oriented
   commits, we measure maintainability --- as computed by the Software Improvement
@@
 In this section, the methodology used to measure the impact of security refactorings on the maintainability of open-source software is presented in depth in the following sections and illustrated in Figure \ref{fig:met}.
-Aiming to answer the research questions presented in this paper, we use a dataset containing more than 700 security refactorings collected from open-source software available on GitHub and where 648 were reported in a previous study conducted in 2017 \cite{Reis:2017:IJSSE}.
+Aiming to answer the research questions presented in this paper, we use a dataset containing $716$ security refactorings collected from open-source software available on GitHub.
+$648$ of those refactorings were reported in a previous study conducted in 2017 \cite{Reis:2017:IJSSE}.
 In parallel, a baseline of regular commits was randomly collected from the list of projects of the main dataset to evaluate what is the impact of regular commits on the maintainability of open-source software.
@@
 
-To evaluate the impact of security refactorings on the maintainability of open-source software, we use a dataset of security flaws which is the outcome of mining more than 248 GitHub\footnote{Github's website available at https://github.com/ (Accessed on January 23, 2019)} repositories.
-Reis and Abreu (2017) mined open-source software aiming the extraction of real test cases - created by real developers on their daily basis work - to test and assess the performance of static analysis tools since using hand-seeded test cases or mutations could lead to misleading assessments of the tools' capabilities.
-The study yielded to a dataset of more than 600 test cases for 16 different patterns.
+To evaluate the impact of security refactorings on the maintainability of open-source software, we use a dataset of security flaws which is the outcome of mining more than $248$ GitHub\footnote{Github's website available at https://github.com/ (Accessed on January 23, 2019)} projects.
+Reis and Abreu (2017) mined open-source software aiming the extraction of real test cases - created by real developers on their daily basis development - to test and assess the performance of static analysis tools since using hand-seeded test cases or mutations could lead to misleading assessments of the capabilities of the tools.
+The study yielded to a dataset of $648$ test cases for $16$ different patterns.
 Each test case of the dataset is a triplet of folders: the commit before the refactoring, the commit responsible for the refactoring and the snippets of code that differ from one version to another (usually, called \textit{diff}) - where one can easily identify the code used to fix the security flaw.
@@
 
-\begin{table}[h]
+% \begin{table}[h]
+% 	\centering
+% \caption{Current Dataset Patterns Distribution} \label{tab:patterns}
+% \begin{tabular}{@{}ll@{}}
+% \toprule
+% Pattern & \# Commits\\
+% \midrule
+% Injection & 89\\
+% Broken Authentication and Session Management& 45\\
+% Cross-Site Scripting& 142\\
+% Broken Access Control& 2\\
+% Security Misconfiguration& 9\\
+% Sensitive Data Exposure& 24\\
+% Insufficient Attack Protection& 18\\
+% Cross-Site Request Forgery& 34\\
+% Using Components With Known Vulnerabilities& 27\\
+% Unprotected APIs& 7\\
+% Memory Leak& 91\\
+% Overflow& 19\\
+% Denial-of-Service& 43\\
+% Path Traversal& 19\\
+% Resource Leak& 33\\
+% SHA-1 Hash Function & 1\\
+% Miscellaneous& 101\\\midrule
+% Total& 716\\
+% \bottomrule
+% \end{tabular}
+% \end{table}
+
+\begin{figure}[h]
  	\centering
-\caption{Current Dataset Patterns Distribution} \label{tab:patterns}
-\begin{tabular}{@{}ll@{}}
-\toprule
-Pattern & \# Commits\\
-\midrule
-Injection & 101\\
-Broken Authentication and Session Management& 45\\
-Cross-Site Scripting& 142\\
-Broken Access Control& 2\\
-Security Misconfiguration& 9\\
-Sensitive Data Exposure& 24\\
-Insufficient Attack Protection& 18\\
-Cross-Site Request Forgery& 34\\
-Using Components With Known Vulnerabilities& 27\\
-Unprotected APIs& 7\\
-Memory Leak& 91\\
-Overflow& 19\\
-Denial-of-Service& 43\\
-Path Traversal& 19\\
-Resource Leak& 33\\
-SHA-1 Hash Function & 1\\
-Miscellaneous& 101\\\midrule
-Total& 716\\
-\bottomrule
-\end{tabular}
-\end{table}
+ 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
+ 	\caption{Security Refactorings Language Distribution}
+	\label{fig:lang}
+\end{figure}
 
-Reis and Abreu (2017) dataset commits were manually validated, flagged as security flaws and classified with one of 16 the following patterns: \textit{Injection}, \textit{Broken Authentication and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery}, \textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs}, \textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service}, \textit{Path Traversal} and \textit{Miscellaneous}.
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
+ 	\caption{Type Distribution}
+	\label{fig:type}
+\end{figure}
+
+Reis and Abreu ($2017$) dataset commits were manually validated, flagged as security flaws and classified with one of the $16$ following patterns: \textit{Injection}, \textit{Broken Authentication and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery}, \textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs}, \textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service}, \textit{Path Traversal} and \textit{Miscellaneous}.
 However, the dataset evolved and now integrates new patterns, such as \textit{Resource Leak} and \textit{SHA-1 Hash Function}.
+A considerable amount of these patterns are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 2017~\cite{oswap:2017}.
 
-In total, the dataset contains 716 security refactorings, as shown in Table \ref{tab:patterns}. This is the dataset used
-by the BCH tool to compute the maintainability of security refactorings. However, some of the projects are not considered for
-the final results because the limitations of the tool, such as the project size and large time of computation.
-To calculate the significance of our results by pattern, we perform the wilcoxon statistical test - described in more depth in subsection \ref{sec:statsval} - which needs to comprise more than 20 commits to be effective.
-The patterns not comprising the number of necessary commits were reallocated to the \textit{Miscellaneous} pattern.
-Due to the tool limitations and the statistical test requirements, this study only distinguishes nine of the patterns presented in Table \ref{tab:patterns}.
-In the
-end, we used the following categorization:
+In total, the dataset contains 716 security refactorings.
+These refactorings were computed by the BCH tool to calculate maintainability.
+However, some of the refactorings are not considered in the final results due to limitations of BCH: lack of language support and project size.
+The Wilcoxon statistical test - described more in-depth in subsection \ref{sec:statsval} - is performed to evaluate the significance of our results.
+Each type of security flaw is tested, Figure \ref{fig:pat}.
+One of the main requirements of this test is that the input sample needs to comprise more than 20 commits.
+Thus, the patterns not complying with this requirement were reallocated to the \textit{Miscellaneous} pattern.
+In the end, we highlight the following patterns:
 
 \begin{itemize}
-	\item \textbf{Using Components with Known Vulnerabilities.}
-	\item \textbf{Resource Leak}
-	\item \textbf{Broken Authentication \& Session Management}
-	\item \textbf{Cross-Site Request Forgery}
-	\item \textbf{Denial-of-Service}
-	\item \textbf{Miscellaneous}
-	\item \textbf{Cross-Site Scripting}
-	\item \textbf{Injection}
-	\item \textbf{Memory Leak}
+	\item \textbf{Using Components with Known Vulnerabilities.} The majority of software produced today integrates several components, such as libraries and frameworks, which may affect the software if developers use vulnerable versions - usually known and disclosed somewhere on the Internet.
+\textit{Based on the OWASP Top 10 of 2017.}
+	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly implemented functionalities related to authentication and session management, allowing crackers to gain access to session tokens, passwords, keys and other sensitive data \textit{- Based on the OWASP Top 10 of 2013.}
+	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow crackers to send forged HTTP requests including authentication information from the victim to vulnerable web application \textit{- Based on the OWASP Top 10 of 2013.}
+	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to flood or crashing services.
+This attacks usually occur when the system receives much traffic causing it to slow down and eventually stop.
+For example, flaws allowing crackers to trigger memory allocations corresponding to large length values (CVE-2014-3506)\footnote{CVE-2014-3506 details available at https://nvd.nist.gov/vuln/detail/CVE-2014-3506 (Accessed on January 25, 2019)}.
+	\item \textbf{Miscellaneous} This pattern comprises several others categories of security refactorings (e.g., \textit{path traversal and buffer overflow}).
+It contains several other security refactorings that do not have their own pattern yet.
+The patterns not satisfying the size requirement of more than 20 refactorings are contemplated here.
+	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow crackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites \textit{- Based on the OWASP Top 10 of 2017.}
+	\item \textbf{Injection.} When developers do not keep untrusted data separate from commands and queries.
+If a cracker sends a string that exploits the syntax of the interpreter, then an injection attack is possible (e.g., SQL and LDAP injection)\textit{- Based on the OWASP Top 10 of 2017.}
+	\item \textbf{Memory Leak.} Memory management issue found more frequently in programming languages that do not manage memory automatically (e.g., C/C++ and Objective-C), i.e., where instead developers are responsible for handling it.
+One of the main causes of DoS attacks.
 \end{itemize}
@@
 \section{Results}
-a
+
 \begin{figure}[h]
@@
  	\includegraphics[width=0.45\textwidth]{figures/maintainability.pdf}
- 	\caption{Maintainability difference for security and baseline commits \textcolor{mypink3}{@update}}
+ 	\caption{Maintainability difference for security and baseline refactorings}
+	\label{fig:secvsreg}
 \end{figure}
@@
  	\includegraphics[width=0.55\textwidth]{figures/category.pdf}
- 	\caption{Maintainability difference by type of security commit \textcolor{mypink3}{@update}}
+ 	\caption{Maintainability difference by type after security refactorings}
+	\label{fig:pat}
 \end{figure}

### 2019-01-25 scripts/get_maintainability.py

@@ -41,11 +41,8 @@ def clean_plot():
     plt.gcf().clear()
 
 # TODO: IMPROVE
-def count(t_id, id, total, lang, pattern, year, l, p, y):
-    total[t_id] += 1
-    lang[l][id] += 1
+def count(id, pattern, p):
     pattern[p][id] += 1
-    year[y][id] += 1
 
 def total_main_barchart(total_sec, total_reg, graphics, result):
     
@@ -124,7 +121,6 @@ def analysis(dataset, results, CACHE):
         fields = ['owner', 'project', 'sha', 'sha-p', 'main(sha)', 'main(sha-p)', 'main(diff)']
         writer = csv.DictWriter(out, fieldnames=fields)    
         writer.writeheader()
-        total = {'neg': 0, 'pos': 0, 'nul': 0}
 
         langs, patterns, years, differences, diff_by_pat = {}, {}, {}, [], {}
         none = 0; error = 0
@@ -179,16 +175,14 @@ def analysis(dataset, results, CACHE):
                                 'main(diff)': main_d
                                 })          
             
-                langs = check_if_in(lang, langs)
                 patterns = check_if_in(pattern, patterns)
-                years = check_if_in(year, years)
-
+                
                 if main_d < 0:
-                    count('neg', 0, total, langs, patterns, years, lang, pattern, year)
+                    count(0, patterns, pattern)
                 elif main_d > 0:
-                    count('pos', 1, total, langs, patterns, years, lang, pattern, year)
+                    count(1, patterns, pattern)
                 else:
-                    count('nul', 2, total, langs, patterns, years, lang, pattern, year)
+                    count(2, patterns, pattern)
 
                 differences.append(main_d)
                 if pattern not in diff_by_pat:
@@ -196,26 +190,26 @@ def analysis(dataset, results, CACHE):
                 else:
                     diff_by_pat[pattern].append(main_d)
                     
-            return patterns, none, error, total, diff_by_pat 
+            return patterns, none, error, diff_by_pat 
 
 def sec_vs_rg_commits(CACHE, graphics, dataset):
     
[…]

### 2019-01-25 scripts/histogram.py

@@ -8,6 +8,7 @@ import matplotlib.pyplot as plt
 
 
 def language_dist(commits):
+    
     df = pd.read_csv(commits)
     
     lang_trans = {'objc':'Objective-C', 'php':'PHP', 'ruby':'Ruby', 'c':'C', 'c++':'C++', 'groovy':'Groovy', 'javascript':'JavaScript',
@@ -43,6 +44,7 @@ def language_dist(commits):
     plt.savefig('../paper/ICPC19/figures/language_dist.pdf')
 
 def type_dist(projects):
+    
     df = pd.read_csv(projects)
     
     dic = {}
@@ -57,7 +59,7 @@ def type_dist(projects):
     for i in dic:
         dic_type['type'].append(i)
         dic_type['no'].append(dic[i]) 
-        
+                
     app_type = pd.DataFrame(dic_type)
     
     ax = app_type['no'].plot(kind='bar', figsize=(10, 7), fontsize=11, color="#48B963")
@@ -78,16 +80,13 @@ def main(projects, commits):
     
     type_dist(projects)
     
-    
-    
-    
 
 if __name__ == "__main__":
     
     parser = argparse.ArgumentParser()
     parser.add_argument('--projects-csv',metavar='projects-csv',required=True,help='the projects filename')
     parser.add_argument('--commits-csv',metavar='commits-csv',required=True,help='the commits filename')
+    #parser.add_argument('--paper-fig-folder',metavar='commits-csv',required=True,help='the figures folder')
     
-
     args = parser.parse_args()
     main(projects=args.projects_csv, commits=args.commits_csv)

### 2019-01-25 paper/ICPC19/icpc19.tex
another commit

@@
 
+The dataset contemplates security flaws for more than 13 different languages being PHP the most prevalent one, Figure \ref{fig:lang}.
+This may be due to the abundance of web software (e.g., Library, Frameworks, Platform, Server), Figure \ref{fig:type}.
+
 \begin{table*}[h]
 	\centering
-\caption{Descriptive statistics of the dataset projects\textcolor{mypink3}{@update}} \label{tab:dataset}
-\begin{tabular}{@{}ccccccccccc@{}}
+	\caption{Descriptive statistics of the dataset projects}
+\begin{tabular}{@{}lllllllllll@{}}
 \toprule
       & forks   & stars   & watchers & contributors & commits  & branches & releases & size      & issues & pull requests  \\ \midrule
-Mean & 1734.79 & 5262.97 & 395.11   & 152.48       & 22257.82 & 44.08    & 126.31   & 143243.64 & 3725.4  & 1976.92       \\
-Min     & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0       & 0             \\
-25 \%     & 374.75  & 1571.5  & 117.75   & 48           & 1347     & 4        & 22       & 8339.5    & 321.25  & 151.75        \\
-Median     & 887     & 2836.5  & 258.5    & 102.5        & 5823     & 9        & 59       & 38331.5   & 1652.5  & 515.5         \\
-75 \%     & 2196    & 6455.75 & 462.5    & 259          & 22786.25 & 22.25    & 142.25   & 164442.5  & 4152.75 & 1944.25       \\
-Max     & 16366   & 31841   & 3446     & 413          & 717449   & 1227     & 1114     & 2042017   & 33970   & 19329         \\
-Total     & 180418  & 547349  & 41091    & 15858        & 2314813  & 4584     & 13136    & 14897339  & 387442  & 205600        \\ \bottomrule
+mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
+std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
+min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
+25\%  & 391.50  & 1581    & 117.25   & 49           & 1440.50  & 4        & 19       & 8466.75   & 313.75    & 143.25  \\
+median  & 838.50  & 2836.50 & 248      & 99           & 5504.50  & 9        & 59       & 37372.50  & 1792.50   & 650     \\
+75\%  & 2155    & 6828.50 & 459.50   & 261          & 18579.25 & 20       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
+max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
+Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
 \end{tabular}
+\label{tab:dataset}
 \end{table*}
+
+Table \ref{tab:dataset} presents the final dataset projects statistics.
+Refactorings from 94 projects were evaluated by BCH.
+According to the statistics, the projects of this dataset are popular and the developers are active.
 

### 2019-01-25 paper/ICPC19/icpc19.tex
another commit 2
small fix

@@
 \definecolor{mypink3}{cmyk}{0, 0.7808, 0.4429, 0.1412}
+
+\usepackage{listings}
+
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\definecolor{lightgreen}{rgb}{0.6,0.9,0.6}
+\definecolor{lightyellow}{rgb}{0.9,0.9,0.6}
+\definecolor{lightorange}{rgb}{0.9,0.8,0.6}
+\definecolor{lightred}{rgb}{0.9,0.7,0.7}
+\definecolor{mygray}{rgb}{0.5,0.5,0.5}
+\definecolor{lightgray}{rgb}{0.8,0.8,0.8}
+\definecolor{mymauve}{rgb}{0.58,0,0.82}
+
+\usepackage{pifont}
+\newcommand{\lstbg}[3][0pt]{{\fboxsep#1\colorbox{#2}{\strut #3}}}
+\lstset{ %
+  language=Java,
+  backgroundcolor=\color{white},   % choose the background color
+  basicstyle=\ttfamily\scriptsize,        % size of fonts used for the code
+  breaklines=false,                 % automatic line breaking only at whitespace
+  captionpos=b,                    % sets the caption-position to bottom
+  commentstyle=\color{mygray}\bfseries,    % comment style
+  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+  keywordstyle=\color{blue},       % keyword style
+  stringstyle=\color{mymauve},     % string literal style
+  frame=single,
+  escapeinside={/*!}{!*/},
+  morekeywords={@Override,@TargetApi},
+  moredelim=**[l][\color{mygreen}]{+},
+  moredelim=*[l][\color{red}]{-}
+}
 
@@
 
+%\lstinputlisting[language=Python]{source_filename.py}
+
 \textcolor{mypink3}{Ver que exemplos existem na amostra da secbench que vamos utilizar para dar como exemplo de motivacao}
@@
 In parallel, a baseline of regular commits was randomly collected from the list of projects of the main dataset to evaluate what is the impact of regular commits on the maintainability of open-source software.
-The Software Improvement Group's\footnote{SIG's website is available at https://www.softwareimprovementgroup.com/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.
+The Software Improvement Group's\footnote{SIG's website is available at https://www.sig.eu/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.
 
 \begin{figure}[h]
- 	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
  	\caption{Study Methodology}
@@
 Refactorings from 94 projects were evaluated by BCH.
-According to the statistics, the projects of this dataset are popular and the developers are active.
+According to the statistics, the projects of this dataset are \textcolor{mypink3}{popular and the developers are active}.
 

### 2019-01-26 paper/ICPC19/icpc19.tex
fixed indent

@@
 
+\noindent
 where $G$ is the group of selected maintainability guidelines from BCH (e.g., Write short units of code, etc.) and $v$ is the

### 2019-01-26 paper/ICPC19/icpc19.tex
fixed indent

@@
 
+\noindent
 where $C$ is the compliance with the maintainability guideline for the given severity level (medium, high, and very high) and
@@
 
+\noindent
 where $LOC_{compliant}(l)$ are the lines of code that comply with the guideline at the given severity level $l$, $LOC_{\neg compliant}(l)$ are the lines of code that do not comply with the guideline at the given
@@
 
+\noindent
 where $T(l)$ is the threshold in percentage of the lines of code that are accepted to be non-compliant with the guideline for the severity level $l$.

### 2019-01-26 paper/ICPC19/icpc19.tex
another update

--- 2019-01-25 paper/ICPC19/icpc19.tex
@@
 
-\usepackage{pifont}
 \newcommand{\lstbg}[3][0pt]{{\fboxsep#1\colorbox{#2}{\strut #3}}}
 \lstset{ %
-  language=Java,
+  language=C,
   backgroundcolor=\color{white},   % choose the background color
@@
 
-%\lstinputlisting[language=Python]{source_filename.py}
+Improving energy efficiency of apps revolves around chang- ing their codebases.
+Previous work has studied existing energy patterns for mobile applications [18].
+It cataloged typical coding practices developers adopt to address energy efficiency.
+An example of an energy pattern is the Power Save Mode: the app features a mode that can be activated upon low battery and uses fewer resources while providing the minimum
++ functionality that is indispensable to the user.
+
+\begin{lstlisting}
+
++ static unsigned long dtls1_max_handshake_message_len(
++ 	const SSL *s)
++	{
++	unsigned long max_len = DTLS1_HM_HEADER_LENGTH + 
++ 			SSL3_RT_MAX_ENCRYPTED_LENGTH;
++	if (max_len < (unsigned long)s->max_cert_list)
++		return s->max_cert_list;
++	return max_len;
++	}
+
+// [snip]
+
+ 	pitem *item = NULL;
+ 	int i = -1, is_complete;
+ 	unsigned char seq64be[8];
+-	unsigned long frag_len = msg_hdr->frag_len, max_len;
+-
+-	if ((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
+-		goto err;
+-
+-	/* Determine maximum allowed message size. Depends on (user set)
+-	 * maximum certificate length, but 16k is minimum.
+-	 */
+-	if (DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
+-		max_len = s->max_cert_list;
+-	else
+-		max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
++	unsigned long frag_len = msg_hdr->frag_len;
+
+-	if ((msg_hdr->frag_off+frag_len) > max_len)
++	if ((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
++	    msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
+ 		goto err;
+
+// [snip]
+
+ 		if (frag_len && frag_len < msg_hdr->msg_len)
+ 			return dtls1_reassemble_fragment(s, msg_hdr, ok);
+
++		if (frag_len > dtls1_max_handshake_message_len(s))
++			goto err;
++
+ 		frag = dtls1_hm_fragment_new(frag_len, 0);
+ 		if ( frag == NULL)
+ 			goto err;
+
+\end{lstlisting}
 

### 2019-01-26 paper/ICPC19/icpc19.tex
read only between ready start and finish

--- 2019-01-26 paper/ICPC19/icpc19.tex
@@
 \usepackage{framed}
+\usepackage{float}
 \usepackage{xcolor}
@@
 \lstset{ %
-  language=Java,
+  language=C,
   backgroundcolor=\color{white},   % choose the background color
   basicstyle=\ttfamily\scriptsize,        % size of fonts used for the code
-  breaklines=false,                 % automatic line breaking only at whitespace
+  breaklines=true,                 % automatic line breaking only at whitespace
   captionpos=b,                    % sets the caption-position to bottom
@@
   frame=single,
+  numbers=left,
+  stepnumber=1,
+  xleftmargin=2em,
   escapeinside={/*!}{!*/},
-  morekeywords={@Override,@TargetApi},
   moredelim=**[l][\color{mygreen}]{+},
@@
 
+\newcounter{lstannotation}
+\setcounter{lstannotation}{0}
+\renewcommand{\thelstannotation}{\ding{\number\numexpr181+\arabic{lstannotation}}}
+\newcommand{\annotation}[1]{\refstepcounter{lstannotation}\label{#1}\thelstannotation}
+
 \def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
@@
 
+\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{cyan} #1}\textbf{]]}}
+
 \begin{document}
 
-\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code%\\
-%\thanks{Identify applicable funding agency here. If none, delete this.}
-}
+\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code}
 
@@
 
-\section{Motivation and Research Questions}
+\Sof{=============== READY START!}
 
-%\lstinputlisting[language=Python]{source_filename.py}
+\section{Motivation and Research Questions}\label{sec:motivation}
 
-\textcolor{mypink3}{Ver que exemplos existem na amostra da secbench que vamos utilizar para dar como exemplo de motivacao}
+Under the rush of companies trying to outdo each other and due to the lack of expertise in
+the security field, security flaws are usually detected only when hackers exploit them which happens
+because developers create buggy software without knowing. Gladly, fixing these
+issues usually is as simple as changing the software codebase. However,
+these refactorings can sometimes have a negative impact on software maintenance, not
+only because developers are under the rush of "protecting" their software but also because 
+they tend to take the easiest path and not the most efficient one. Thus, we
+want to understand how security refactorings impact the maintainability of software.
+
+One example is the refactoring of the DTLS protocol implementation in OpenSSL after
+Adam Langley detect that remote attackers could cause a Denial-of-Service (DoS) attack
+when crafting DTLS handshake messages to trigger memory allocations corresponding to large length values. 
+This vulnerability is listed at the Common Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on January 25, 2019)} and
+it is one of the vulnerabilities under evaluation in this study. The impact of this refactoring
+on the OpenSSL maintainability was negative. The following snippet presents the codebase changes 
+performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636} by the OpenSLL developers to fix the vulnerability.
+
+\setcounter{lstannotation}{0}
+\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability},label={lst:vuln}]
++static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
++
++ unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
++ if(max_len < (unsigned long)s->max_cert_list) /*!\annotation{lst:func2}!*/
++    return s->max_cert_list;
++ return max_len;
++}
+
+static int dtls1_reassemble_fragment(SSL *s, struct hm_header_st* msg_hdr, int *ok){
+
+// [snip]
+
+- unsigned long frag_len = msg_hdr->frag_len, max_len; /*!\annotation{lst:func3}!*/
+-
+- if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len) 
+-    goto err;
+
+- if(DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
+-    max_len = s->max_cert_list;
+- else
+-    max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
+
++ unsigned long frag_len = msg_hdr->frag_len;
+
+- if((msg_hdr->frag_off+frag_len) > max_len)
+
++ if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
++ msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
+     goto err; /*!\annotation{lst:func5}!*/
+
+  memset(seq64be,0,sizeof(seq64be));
+  seq64be[6] = (unsigned char) (msg_hdr->seq>>8);
+  seq64be[7] = (unsigned char) msg_hdr->seq;
+  item = pqueue_find(s->d1->buffered_messages, seq64be);
+  if(item == NULL)
+  {
+    frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);
+
+// [snip]
+
+ static int
+ dtls1_process_out_of_seq_message(SSL *s, struct hm_header_st* msg_hdr, int *ok)
+ {
+
+// [snip]
+
+  if(frag_len && frag_len < msg_hdr->msg_len)
+     return dtls1_reassemble_fragment(s, msg_hdr, ok);
+
++ if(frag_len > dtls1_max_handshake_message_len(s))
++    goto err; /*!\annotation{lst:func4}!*/
++
+  frag = dtls1_hm_fragment_new(frag_len, 0);
+
+\end{lstlisting}
+
+In the \emph{dtls1\_reassemble\_fragment} function the value of \emph{msg\_hdr-$>$frag\_off+frag\_len} was checked against the
+maximum hadshake message size but \emph{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
+the allowed size, the pending handshake message could consume $16MB$+$2MB$. According with the message commit, 10 handshake messages
+are allowed, so the attacker could consume approximately $180MB$ per DTLS connection.
+
+The code changes performed to fix the vulnerability were the following:
+
+\ref{lst:func1} $dtls1\_max\_handshake\_message\_len$ function was inserted to limit the number of bytes allowed in a DTLS handshake message for \emph{s}. This function replaces the lines 18-21.
+
+\ref{lst:func2} Checks if the maximum certificate list size (\emph{s-$>$max\_cert\_list}) requires more length than the maximum length (\emph{max\_len}).
+If yes, it return the certificate list size, if not it return the maximum length.
+
+\ref{lst:func3} Replaced by attribution in line 23.
+
+\ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message length in never higher than the maxiumum allowed by the DTSL protocol in different functions.
+
+Although limiting the size of a handshake message seems an easy problem to fix, it is possible to see that there was a considerable amount of changes performed
+in the codebase which led to a decrease in the project maintainability.
+
+\Sof{=============== READY FINISH!}
 
@@
 \end{framed}
+
+\Sof{=============== READY START!}
 
@@
  	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
- 	\caption{Type Distribution}
-	\label{fig:type}
+ 	\caption{Projects Domain Distribution}
+	\label{fig:domain}
 \end{figure}
@@
 This attacks usually occur when the system receives much traffic causing it to slow down and eventually stop.
-For example, flaws allowing crackers to trigger memory allocations corresponding to large length values (CVE-2014-3506)\footnote{CVE-2014-3506 details available at https://nvd.nist.gov/vuln/detail/CVE-2014-3506 (Accessed on January 25, 2019)}.
+For example, flaws allowing crackers to trigger memory allocations corresponding to large length values (Listing \ref{lst:vuln}).
 	\item \textbf{Miscellaneous} This pattern comprises several others categories of security refactorings (e.g., \textit{path traversal and buffer overflow}).
@@
 
-The dataset contemplates security flaws for more than 13 different languages being PHP the most prevalent one, Figure \ref{fig:lang}.
-This may be due to the abundance of web software (e.g., Library, Frameworks, Platform, Server), Figure \ref{fig:type}.
+The dataset contemplates security flaws for more than 13 different languages being PHP (39\%) and C (24\%) the most prevalent ones, Figure \ref{fig:lang}.
+To classify the software of the dataset, we use a taxonomy for open-source software published in an older study~\cite{7816479}.
+Figure \ref{fig:domain} presents the projects domain distribution: \textit{Aplication Software} ($25$), software that provides end-users with functional systems; \textit{Web libraries and frameworks} ($20$); \textit{System Software} ($18$), software that provides services and infrustructures (e.g., operating systems, servers and databases); \textit{Software Tool} ($13$), software that supports development (e.g., programming languages, compilers, package managers, IDEs); and, \textit{Non-web libraries and frameworks} ($18$), for desktop and mobile software.
 
@@
 
-Table \ref{tab:dataset} presents the final dataset projects statistics.
-Refactorings from 94 projects were evaluated by BCH.
-According to the statistics, the projects of this dataset are \textcolor{mypink3}{popular and the developers are active}.
+Table \ref{tab:dataset} presents the descriptive statistics of the 94 open-source projects involved in this study.
 
@@
 %
-Although we want to assess the maintainability of security-related commits,
-\textcolor{mypink3}{there is no evidence in previous work on the impact of \textit{regular} commits
-in the maintainability of projects}. Thus, we analyze the maintainability of
-regular commits and use them as a baseline.
+Previous studies tried to measure the impact of refactorings on open-source software maintainability~\cite{HEGEDUS2018313} before.
+But to the best of our knowledge, none of them used the BCH model and tried to evaluate the impact of security refactorings compared with regular refactorings as we do.
+Thus, we analyze the maintainability of regular commits and use them as a baseline.
 
@@
 
-BCH is used to collect the maintanability reports of the commits of each project. The tool
-evaluation is based on a model of 10 different guidelines \cite{Visser:2016:OREILLY}:
+Better Code Hub is used to collect the maintanability reports of the refactorings of each project.
+Figure \ref{fig:guidelines} presents the 10 different guidelines proposed by BCH authors for delivering software that is easy to maintain \cite{Visser:2016:OREILLY}.
+The reports provided by the tool are based on the following guidelines:
 
-\begin{itemize}
-	\item \textbf{Write short units of code.} Long units are hard to test, reuse, and understand.
-	\item \textbf{Write simple units of code.} Keeping the number of branch points low makes units easier to modify and test.
-	\item \textbf{Write code once.} When code is duplicated, bugs need to be fixed in multiple places, which is inefficient and
-prone to errors.
-	\item \textbf{Keep unit interfaces small.} Keeping the number of parameters low makes units easier to understand and reuse.
-	\item \textbf{Separate concerns in modules.} Changes in a loosely
-coupled codebase are much easier to oversee and execute
-than changes in a tightly coupled codebase.
-	\item \textbf{Couple architecture components loosely.} Independent
-components ease isolated maintenance.
-	\item \textbf{Keep architecture components balanced.} Balanced
-components ease locating code and foster isolation,
-improving maintenance activities.
-	\item \textbf{Keep your codebase small.} Small systems are easier to
-search through, analyze, and understand code.
-	\item \textbf{Automate tests.} Automated testing makes development
-predictable and less risky.
-	\item \textbf{Write clean code.} Code without code smells is less likely to bring maintainability issues.
-\end{itemize}
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/guidelines.pdf}
+ 	\caption{10 guidelines to produce maintainable code}
+	\label{fig:guidelines}
+\end{figure}
+
+\Sof{=============== READY FINISH!}
+
+The tool calculates the compliance against each particular guideline.
 

### 2019-01-27 paper/ICPC19/icpc19.tex
section 2 finished

@@
 
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
 \Sof{=============== READY START!}
+
+This paper is structured as follows: section \ref{sec:motivation} introduces an example of a security refactoring of a known vulnerability found in OpenSSL\footnote{Repository details available at https://github.com/openssl/openssl (Accessed on January 27, 2019)} software; section \ref{sec:methodology} describes the methodology used to answer the research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.
 
@@
 when crafting DTLS handshake messages to trigger memory allocations corresponding to large length values. 
-This vulnerability is listed at the Common Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on January 25, 2019)} and
+This vulnerability is listed at the Common Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on January 27, 2019)} and
 it is one of the vulnerabilities under evaluation in this study. The impact of this refactoring
 on the OpenSSL maintainability was negative. The following snippet presents the codebase changes 
-performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636} by the OpenSLL developers to fix the vulnerability.
+performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636 (Accessed on January 27, 2019)} by the
+OpenSLL developers to fix the vulnerability.
 
@@
 
-In the \emph{dtls1\_reassemble\_fragment} function the value of \emph{msg\_hdr-$>$frag\_off+frag\_len} was checked against the
-maximum hadshake message size but \emph{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
+In the \texttt{dtls1\_reassemble\_fragment} function the value of \texttt{msg\_hdr-$>$frag\_off+frag\_len} was checked against the
+maximum hadshake message size but \texttt{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
 the allowed size, the pending handshake message could consume $16MB$+$2MB$. According with the message commit, 10 handshake messages
@@
 
-\ref{lst:func1} $dtls1\_max\_handshake\_message\_len$ function was inserted to limit the number of bytes allowed in a DTLS handshake message for \emph{s}. This function replaces the lines 18-21.
+\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to limit the number of bytes allowed in a DTLS handshake message for \texttt{s}.
+This function replaces the lines 18-21.
 
-\ref{lst:func2} Checks if the maximum certificate list size (\emph{s-$>$max\_cert\_list}) requires more length than the maximum length (\emph{max\_len}).
+\ref{lst:func2} Checks if the maximum certificate list size (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length (\texttt{max\_len}).
 If yes, it return the certificate list size, if not it return the maximum length.
@@
 
-Although limiting the size of a handshake message seems an easy problem to fix, it is possible to see that there was a considerable amount of changes performed
-in the codebase which led to a decrease in the project maintainability.
+Although limiting the size of a handshake message seems an elementary problem to repair, the example shows that there was a significant amount of changes performed in the codebase which produced a negative impact in the project maintainability.
+Even though the developer tries to simplify the codebase with the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it also disrupts 2 of the guidelines proposed by the Software Improvement Group (SIG) for building maintainable software~\cite{Visser:2016:OREILLY}.
+It adds a new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the \emph{Write Simple Units of Code} guideline.
+And, it increases the codebase size in 4 LOCs (13 lines added, 9 lines deleted) which infringes the \emph{Keep Your Codebase Small} guideline.
 
-\Sof{=============== READY FINISH!}
+In this study, our concern is if while improving software security, developers are also inserting a negative impact on the maintainability of their software which may increase technical debt when choosing the easiest solution instead of the using the best approach.
+To answer the following two research questions, we use a dataset of security refactorings to measure the impact of the changes in the maintainability of open-source software.
 
@@
 
+Often security flaws require refactoring code to make software more secure.
+However, there is no evidence yet of how security refactorings impact the maintainability of open-source software.
+Our suspicion is that developers tend to introduce technical debt in their software when refactoring the codebose to address a security flaw because they tend to choose the easiest path to solve it.
+To address it, we compute the maintainability value of $607$ commits using \emph{Better Code Hub}.
+The same approach is applied to a randomly generated dataset of regular commits - baseline - that we use to understand how maintainability evolves when security refactorings are performed versus when they are not.
+
 \begin{framed}
@@
 
-\begin{framed}
-\textit{\textbf{RQ3} }
-\end{framed}
+There are security flaws that are more difficult to refactor than others.
+For instance, introducing a mechanism to secure authentication is not so easy to implement as fixing a cross-site scripting vulnerability with a \texttt{escape} function.
+Knowing which patterns are more likely to increase maintainability issues is one step forward to bring awareness to security engineers of what patterns need more attention.
+The taxonomy of security patterns used to answer this question is the same as the one provided by the authors of the dataset.
+Maintainability is measured separately for each pattern.
 
-\Sof{=============== READY START!}
-
-\section{Methodology}
+\section{Methodology}\label{sec:methodology}
 
@@
 However, some of the refactorings are not considered in the final results due to limitations of BCH: lack of language support and project size.
+The final dataset contains $607$ security refactorings.
 The Wilcoxon statistical test - described more in-depth in subsection \ref{sec:statsval} - is performed to evaluate the significance of our results.
@@
 
-\textcolor{mypink3}{@todo: Discuss pratt improvement and accuracy}
+\textcolor{mypink3}{@todo: Discuss pratt improvement and accuracy} Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
-\section{Results}
+\section{Results}\label{sec:results}
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+
+\begin{framed}
+\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability of open-source software?}
+\end{framed}
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+
+\begin{framed}
+\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to affect open-source software maintainability?}
+\end{framed}
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
@@
 
+\section{Discussion}\label{sec:discussion}
+
 \begin{framed}
@@
 \end{framed}
+
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
@@
 
-\section{Discussion}
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
-\section{Threats to Validity}
+\section{Threats to Validity}\label{sec:threats}
 
-\section{Related Work}
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
 
-\section{Conclusion and Future Work}
+\section{Related Work}\label{sec:rw}
 
-\section*{Acknowledgment}
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
-Better Code Hub?
+\section{Conclusion and Future Work}\label{sec:conclusions}
+
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+
+\section*{Acknowledgment}\label{sec:ack}
+
+We thank to the Software Improvement Group (SIG) team for all the support and help in validating our methodology.
 

### 2019-01-28 paper/ICPC19/icpc19.tex
s3 finished

@@
 
-\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{cyan} #1}\textbf{]]}}
+\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{red} #1}\textbf{]]}}
 
@@
   study the impact of changes to improve security on the maintainability of several
-  open source applications. Using a dataset containing $607$ security-oriented
+  open source applications. Using a dataset containing 607 security-oriented
   commits, we measure maintainability --- as computed by the Software Improvement
@@
   in refactorings to deal with \textit{Broken Authentication} and \textit{Cross-Site Request Forgery} attacks.
-  \textcolor{mypink3}{Furthermore, we have found evidence that security-related changes are more
-  likely to be modified in the future than regular code changes.}
 \end{abstract}
@@
 
-The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability,
-functional suitability, performance efficiency, compatibility, usability, reliability, security,
-and portability. This paper focuses solely on maintainability.
-
 Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
@@
 It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
 Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
@@
 It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+
+The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability,
+functional suitability, performance efficiency, compatibility, usability, reliability, security,
+and portability. This paper focuses solely on maintainability.
 
@@
 
-Under the rush of companies trying to outdo each other and due to the lack of expertise in
-the security field, security flaws are usually detected only when hackers exploit them which happens
+\Sof{@todo: Improve this paragraph}Under the rush of companies trying to outdo each other and due to the lack of expertise in the security field, security flaws are usually detected only when hackers exploit them which happens
 because developers create buggy software without knowing. Gladly, fixing these
@@
 on the OpenSSL maintainability was negative. The following snippet presents the codebase changes 
-performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636 (Accessed on January 27, 2019)} by the
-OpenSLL developers to fix the vulnerability.
+performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636 (Accessed on January 27, 2019)} by the OpenSLL developers to fix the vulnerability.
 
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability},label={lst:vuln}]
+\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability \Sof{@todo: Fix the regex issue}},label={lst:vuln}]
 +static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
@@
 In the \texttt{dtls1\_reassemble\_fragment} function the value of \texttt{msg\_hdr-$>$frag\_off+frag\_len} was checked against the
-maximum hadshake message size but \texttt{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
+maximum handshake message size but \texttt{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
 the allowed size, the pending handshake message could consume $16MB$+$2MB$. According with the message commit, 10 handshake messages
@@
 It adds a new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the \emph{Write Simple Units of Code} guideline.
-And, it increases the codebase size in 4 LOCs (13 lines added, 9 lines deleted) which infringes the \emph{Keep Your Codebase Small} guideline.
+And, it increases the codebase size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the \emph{Keep Your Codebase Small} guideline.
 
@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
- 	\caption{Study Methodology}
+ 	\caption{Study Methodology \Sof{@todo: Improve it!}}
 	\label{fig:met}
@@
 
-% \begin{table}[h]
-% 	\centering
-% \caption{Current Dataset Patterns Distribution} \label{tab:patterns}
-% \begin{tabular}{@{}ll@{}}
-% \toprule
-% Pattern & \# Commits\\
-% \midrule
-% Injection & 89\\
-% Broken Authentication and Session Management& 45\\
-% Cross-Site Scripting& 142\\
-% Broken Access Control& 2\\
-% Security Misconfiguration& 9\\
-% Sensitive Data Exposure& 24\\
-% Insufficient Attack Protection& 18\\
-% Cross-Site Request Forgery& 34\\
-% Using Components With Known Vulnerabilities& 27\\
-% Unprotected APIs& 7\\
-% Memory Leak& 91\\
-% Overflow& 19\\
-% Denial-of-Service& 43\\
-% Path Traversal& 19\\
-% Resource Leak& 33\\
-% SHA-1 Hash Function & 1\\
-% Miscellaneous& 101\\\midrule
-% Total& 716\\
-% \bottomrule
-% \end{tabular}
-% \end{table}
-
 \begin{figure}[h]
- 	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
  	\caption{Security Refactorings Language Distribution}
@@
 \begin{figure}[h]
- 	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
  	\caption{Projects Domain Distribution}
@@
 
-In total, the dataset contains 716 security refactorings.
+In total, the dataset contains $716$ security refactorings.
 These refactorings were computed by the BCH tool to calculate maintainability.
@@
 Each type of security flaw is tested, Figure \ref{fig:pat}.
-One of the main requirements of this test is that the input sample needs to comprise more than 20 commits.
+One of the main requirements of this test is that the input sample needs to comprise more than $20$ commits.
 Thus, the patterns not complying with this requirement were reallocated to the \textit{Miscellaneous} pattern.
@@
 \textit{Based on the OWASP Top 10 of 2017.}
-	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly implemented functionalities related to authentication and session management, allowing crackers to gain access to session tokens, passwords, keys and other sensitive data \textit{- Based on the OWASP Top 10 of 2013.}
-	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow crackers to send forged HTTP requests including authentication information from the victim to vulnerable web application \textit{- Based on the OWASP Top 10 of 2013.}
+	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly implemented functionalities related to authentication and session management, allowing crackers to gain access to session tokens, passwords, keys and other sensitive data.
+\textit{Based on the OWASP Top 10 of 2013.}
+	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow crackers to send forged HTTP requests including authentication information from the victim to vulnerable web application.
+\textit{Based on the OWASP Top 10 of 2013.}
 	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to flood or crashing services.
@@
 The patterns not satisfying the size requirement of more than 20 refactorings are contemplated here.
-	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow crackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites \textit{- Based on the OWASP Top 10 of 2017.}
+	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow crackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites.
+\textit{Based on the OWASP Top 10 of 2017.}
 	\item \textbf{Injection.} When developers do not keep untrusted data separate from commands and queries.
-If a cracker sends a string that exploits the syntax of the interpreter, then an injection attack is possible (e.g., SQL and LDAP injection)\textit{- Based on the OWASP Top 10 of 2017.}
+If a cracker sends a string that exploits the syntax of the interpreter, then an injection attack is possible (e.g., SQL and LDAP injection).
+\textit{Based on the OWASP Top 10 of 2017.}
 	\item \textbf{Memory Leak.} Memory management issue found more frequently in programming languages that do not manage memory automatically (e.g., C/C++ and Objective-C), i.e., where instead developers are responsible for handling it.
@@
 Table \ref{tab:dataset} presents the descriptive statistics of the 94 open-source projects involved in this study.
+\Sof{@all: Should I expand this?
+I am not sure what it is important to say about this.}
 
 \subsection{Security vs. Baseline Commits}
-%
+
 Previous studies tried to measure the impact of refactorings on open-source software maintainability~\cite{HEGEDUS2018313} before.
@@
 characteristics of different projects.
-%
+
 \subsection{Maintainability Analysis}
@@
 
-\begin{figure}[h]
- 	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/guidelines.pdf}
- 	\caption{10 guidelines to produce maintainable code}
+\begin{figure}[H]
+ 	\centering 	\includegraphics[width=0.49\textwidth]{figures/guidelines.pdf}
+ 	\caption{Guidelines to produce maintainable code}
 	\label{fig:guidelines}
@@
 
-\Sof{=============== READY FINISH!}
+All guidelines are measured in the same evaluation by BCH.
+During each guideline evaluation, the tool determines the compliance towards one guideline by establishing limits for the percentage of code allowed to be in each of the 4 risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and \emph{very high risk}).
+If the project does not violate the thresholds, then it is compliant with the guideline.
+These thresholds are calibrated by BCH using their own data - open-source and closed software systems.
+If a project is compliant with a guideline, it means that it is at least $65\%$ better than the software used by BCH to calculate the thresholds\footnote{Check the answer to \emph{How can I adjust the threshold for passing/not passing a guideline?} at https://bettercodehub.com/docs/faq (Accessed on January 27, 2019)}.
+Figure \ref{fig:bchrep} shows an example of the  guidelines report provided by BCH for a project after finishing its evaluation.
+These are the results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring - motivation example discussed, previously, in section \ref{sec:motivation}.
+This version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write Clean Code} guideline.
 
-The tool calculates the compliance against each particular guideline.
+\begin{figure}[H]
+ 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
+ 	\caption{Maintainability report of OpenSSL CVE-2014-3506 vulnerability refactoring for the guideline \emph{Write Simple Units of Code} provided by \emph{Better Code Hub}.
+This version of OpenSSL does not comply with the guideline in the example since the bars do not reach the threshold points.
+This example only complies with 1/10 guidelines (\emph{Write Clean Code}).}
+	\label{fig:bchrep}
+\end{figure}
 
-For each guideline, BCH evaluates the compliance against a particular guideline by setting boundaries for the percentage of code allowed to fall in each of the four risk severity categories (low risk, medium risk, high risk, and very high risk).
-If the thresholds are not violated, the project is considered to be compliant with the guideline.
-According to BCH, the guideline thresholds are calibrated yearly based on a representative benchmark of closed and open-source software systems.
-Being compliant with a guideline means that the project under analysis is at least better than 65\% of the software systems in BCH’s benchmark.
+\emph{Units} are the smallest groups of code that can be maintained and executed independently \cite{Visser:2016:OREILLY}.
+McCabe \cite{1702388} is a software metric that calculates the number of linearly independent paths of the source code.
+BCH uses this metric to measure the number of branch points in the \emph{Write Simple Units of Code} guideline.
+The bar represents the top 30 of units that violates the guideline, sorted by severity, which is indicated by the colors of the checkboxes.
+The green bar represents the number of compliant branch points per unit (\emph{at most 5}), i.e., the units are compliant with ISO 25010 \cite{iso:2011}.
+Yellow, orange and red bars represent units that do not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high (\emph{above 25}) severity levels.
+In the bar, there are marks that pinpoint the compliance thresholds for each severity level.
 
-The BCH report of the $<INSERT\_EXEMPLO>$ for a non-compliant guideline can be seen in Fig.
-[X].
-This was extracted from the report of the app $<INSERT\_EXEMPLO>$, used in the motivating example of Section II [MAYBE?].
-The green bar represents the percentage of compliant lines of code.
-These lines of code are considered to be compliant with ISO 25010 standard for maintainability [21].
-The yellow, orange and red bars represent non-compliant lines of code with medium, high, and very high severity levels, respectively.
-Along the bars, there are also marks that refer to the compliance thresholds for each severity level.
-The report is equivalent to the information reported in Table II: a set of thresholds, number of lines of code (LOC), and percentage of the project for each severity level.
-Nonetheless, thresholds provided by BCH do not sum to 100\%: non-compliant levels are provided in a cumulative way (e.g., the threshold for the medium level includes high and very high levels); the compliant-level threshold is the complement of the medium-level threshold.
+\begin{figure}[H]
+ 	\centering 	\includegraphics[width=0.49\textwidth]{figures/commit.pdf}
+ 	\caption{Maintainability difference for security commits}
+	\label{fig:commit}
+\end{figure}
 
-Since we want to analyze maintainability regression, we
-use BCH to compute maintainability in two different versions of the Android app: a) the version of the project before the security commit ($v_{E-1}$) and b) the version immediately after the security commit ($v_E$).
-This is illustrated in Fig.
-[X].
-Although BCH provides a detailed report of the maintainability of the project, it does not compute a final score that we can use to compare maintainability amongst different projects.
-Thus, based in previous work CITE?, we designed an equation to capture the distance between the current state of the project and the standard thresholds.
-We have adjusted the equation to meet the following requirements:
+Aiming to analyze the impact of security refactorings, we use BCH to compute maintainability of two different versions of the project (Figure \ref{fig:commit}):
+\begin{itemize}
+	\item $v_{s-1}$, the version containing the security flaw, i.e., before the refactoring;
+	\item $v_{s}$, the version after fixing the security flaw, i.e., after the refactoring;
+\end{itemize}
 
+The BCH tool does not compute the final score that our study needs to compare maintainability amongst different project versions.
+Thus, our work proposes an equation to capture the distance between the current state of the project and the standard thresholds based on previous work \cite{Olivari:2018}.
 \begin{itemize}
-	\item \textbf{The maintainability difference between two versions of the same project is not affected by its size.} In this work, we want to evaluate the identical security patterns occurring in different projects.
-Thus, the metric cannot use normalization based on its size – we convert percentage data to the respective number of lines of code.
-	\item \textbf{Distance to the thresholds in high severity levels is more penalized than in low severity levels.} We use weights based on the severity level to count lines of code that violate maintainability guidelines.
+	\item \textbf{Size project changes do not affect the maintainability difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we aim to evaluate the alike security patterns occurring in different projects.
+Thus, normalization based on its size cannot be used - we convert relative data to the respective number of lines of code.
+	\item \textbf{Distance to lower severity levels are less penalized than to the thresholds in high severity levels.} Severity level weights based on the severity level to count lines of code that violate maintainability guidelines.
 \end{itemize}
@@
 \noindent
-where $G$ is the group of selected maintainability guidelines from BCH (e.g., Write short units of code, etc.) and $v$ is the
+where $G$ is the group of selected maintainability guidelines from BCH (Figure \ref{fig:guidelines}) and $v$ is the
 version of the software under evaluation. The maintenance $M$ based on the guideline $g$ for a given version of a project is computed with the following equation:
@@
 This is a standard value defined by BCH.
-In other words, the factor $w$ is used in Eq.
-\ref{eq:3} to highlight the lines of code that are not complying with the guideline.
-For instance, the threshold for the severity level veryHigh is defined in Table II as T(veryHigh) = 6.9\%, which derives to a weight of w(veryHigh) = 13.5.
-This means that, in this example, one non-compliant guideline is decreasing maintainability score by 13.5 points while a compliant guideline is increasing by 1.0 point.
-In addition, a version that is perfectly aligned with the standard thresholds has a maintainability score of zero.
-Then, we compute the difference of maintainability between the security commit ($v_E$) and its parent commit ($v_{E-1}$), as illustrated in Fig.
-[X].
+In other words, the factor $w$ is used in Equation \ref{eq:3} to highlight the lines of code that are not complying with the guideline.
+Then, we compute the difference of maintainability between the security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in Figure \ref{fig:commit}.
 
@@
 
-To validate the maintainability differences in different groups of commits (e.g., baseline and security commits) we use the Paired Wilcoxon signed-rank test with the significance level $\alpha = 0.05$.
-In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{E-1}$, $v_E$ (i.e., before and after an security-commit) follows a symmetric distribution around 0.
-To understand the effect-size, as advocated by the Common-language effect sizes [X], we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
-
-\textcolor{mypink3}{@todo: Discuss pratt improvement and accuracy} Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+To validate the maintainability differences in different groups of commits (e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank test \cite{10.2307/3001968} with the significance level $\alpha = 0.05$.
+In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a security-commit) follows a symmetric distribution around $0$.
+To understand the effect-size, as advocated by the Common-language effect sizes \cite{graw:1992}, we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
 
 \section{Results}\label{sec:results}
-
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
@@
 
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-
 \begin{framed}
@@
 \end{framed}
-
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 
@@
 
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-
 \section{Threats to Validity}\label{sec:threats}
-
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
 
@@
 
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-
 \section{Conclusion and Future Work}\label{sec:conclusions}
-
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
 

### 2019-01-28 paper/ICPC19/icpc19.tex
paper

@@
 In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a security-commit) follows a symmetric distribution around $0$.
+However, when the difference between groups is zero, the observations are discarded.
+Thus, we use a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which provides an alternative that incorporates the cases where the maintainability difference is zero \cite{10.2307/2282543}.
 To understand the effect-size, as advocated by the Common-language effect sizes \cite{graw:1992}, we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
+
+ when the difference between the groups is zero, the observations are discarded.
+This is of particular concern if the samples are taken from a discrete distribution.
+In these scenarios the modification to the Wilcoxon test by Pratt 1959, provides an alternative which incorporates the zero differences.[4][5] This modification is more robust for data on an ordinal scale.[5]
 

### 2019-01-28 paper/ICPC19/icpc19.tex
cleaning
starting results

--- 2019-01-28 paper/ICPC19/icpc19.tex
@@
 In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a security-commit) follows a symmetric distribution around $0$.
+However, when the difference between groups is zero, the observations are discarded.
+Thus, we use a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which provides an alternative that incorporates the cases where the maintainability difference is zero \cite{10.2307/2282543}.
 To understand the effect-size, as advocated by the Common-language effect sizes \cite{graw:1992}, we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
@@
 \section{Results}\label{sec:results}
+
+In total, this study evaluates $607$ security and baseline commits from 94 different open-source projects distributed among 5 different main types of software, Figure \ref{fig:domain}.
+The following section presents the results obtained for each research question.
 

### 2019-01-28 paper/ICPC19/icpc19.tex
fixing first paragraph

@@
 \section{Motivation and Research Questions}\label{sec:motivation}
-
-\Sof{@todo: Improve this paragraph}Under the rush of companies trying to outdo each other and due to the lack of expertise in the security field, security flaws are usually detected only when hackers exploit them which happens
-because developers create buggy software without knowing. Gladly, fixing these
-issues usually is as simple as changing the software codebase. However,
-these refactorings can sometimes have a negative impact on software maintenance, not
-only because developers are under the rush of "protecting" their software but also because 
-they tend to take the easiest path and not the most efficient one. Thus, we
-want to understand how security refactorings impact the maintainability of software.
+%
+Due to time-to-market pressure and the lack of security expertise, code-related
+security flaws are generally detected \textit{after the fact}, i.e., when
+hackers exploit them. It turns out that fixing these issues is as simple as
+modifying the software codebase. However, these refactorings may have a negative
+impact on software maintenance, mostly because developers follow the quickest
+solution to fix the problem and not necessarily the most elegant one. Thus, we
+want to understand how security refactorings impact the maintainability of
+software.
 

### 2019-01-28 paper/ICPC19/icpc19.tex
merge
several fixes S2

@@
 
-\title{Security vs. Maintainability: Fixing Vulnerabilities Obfuscates your Code}
+\title{Security vs. Maintainability: \\ Fixing Vulnerabilities Obfuscates your Code}
 
@@
 
-This paper is structured as follows: section \ref{sec:motivation} introduces an example of a security refactoring of a known vulnerability found in OpenSSL\footnote{Repository details available at https://github.com/openssl/openssl (Accessed on January 27, 2019)} software; section \ref{sec:methodology} describes the methodology used to answer the research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.
-
+This paper is structured as follows: section \ref{sec:motivation} introduces an
+example of a security refactoring of a known vulnerability found in
+OpenSSL\footnote{Repository available at
+https://github.com/openssl/openssl (Accessed on \today{})} software;
+section \ref{sec:methodology} describes the methodology used to answer the
+research questions; section \ref{sec:results} presents the results and section
+\ref{sec:discussion} discusses their implications; section \ref{sec:threats},
+threats to the validity of the study are enumerated; section \ref{sec:rw}
+describres the different work and existing literature in the field of study;
+and, finally section \ref{sec:conclusions} concludes the main findings and
+elaborates on future work.
+%
 \section{Motivation and Research Questions}\label{sec:motivation}
@@
 impact on software maintenance, mostly because developers follow the quickest
-solution to fix the problem and not necessarily the most elegant one. Thus, we
-want to understand how security refactorings impact the maintainability of
-software.
+solution to fix the problem and not necessarily the most elegant and performant
+one. Thus, we want to understand how security refactorings impact the
+maintainability of software.
 
-One example is the refactoring of the DTLS protocol implementation in OpenSSL after
-Adam Langley detect that remote attackers could cause a Denial-of-Service (DoS) attack
-when crafting DTLS handshake messages to trigger memory allocations corresponding to large length values.
-This vulnerability is listed at the Common Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on January 27, 2019)} and
-it is one of the vulnerabilities under evaluation in this study. The impact of this refactoring
-on the OpenSSL maintainability was negative. The following snippet presents the codebase changes
-performend on the \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636 (Accessed on January 27, 2019)} by the OpenSLL developers to fix the vulnerability.
+As an example, consider the refactoring of the Datagram Transport Layer Security
+(DTLS) protocol implementation in OpenSSL to address the fact that hackers could
+cause a Denial-of-Service (DoS) attack by crafting DTLS
+handshake messages to trigger memory allocations corresponding to large length
+values. This vulnerability is listed at the Common Vulnerabilities and Exposures
+dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at
+http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on
+\today{})}. It is amongst the vulnerabilities studied in our evaluation.
+The following snippet presents the changes performend on the
+\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
+https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
+(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.
 
@@
 
-In the \texttt{dtls1\_reassemble\_fragment} function the value of \texttt{msg\_hdr-$>$frag\_off+frag\_len} was checked against the
-maximum handshake message size but \texttt{msg\_len} is allocated to the fragment buffer in line 37. Thus, if the fragment was within
-the allowed size, the pending handshake message could consume $16MB$+$2MB$. According with the message commit, 10 handshake messages
-are allowed, so the attacker could consume approximately $180MB$ per DTLS connection.
+In the \texttt{dtls1\_reassemble\_fragment} function the value of
+\texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the maximum
+handshake message size, but \texttt{msg\_len} is allocated to the fragment
+buffer in line $37$. Thus, if the fragment was within the allowed size, the
+pending handshake message could consume $18MB$ ($16MB$+$2MB$). According to the
+message commit, $10$ handshake messages are allowed, so the attacker could
+consume approximately $180MB$ per DTLS connection.
 
@@
 
-\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to limit the number of bytes allowed in a DTLS handshake message for \texttt{s}.
-This function replaces the lines 18-21.
+\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
+inserted to limit the number of bytes allowed in a DTLS handshake message for
+\texttt{s}. This function replaces lines $18$---$21$.
 
-\ref{lst:func2} Checks if the maximum certificate list size (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length (\texttt{max\_len}).
-If yes, it return the certificate list size, if not it return the maximum length.
+\ref{lst:func2} Checks if the maximum certificate list size
+(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
+(\texttt{max\_len}). If yes, it return the certificate list size, if not it
+return the maximum length.
 
-\ref{lst:func3} Replaced by attribution in line 23.
+\ref{lst:func3} Replaced by attribution in line $23$.
 
-\ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message length in never higher than the maxiumum allowed by the DTSL protocol in different functions.
+\ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message
+length in never higher than the maxiumum allowed by the DTSL protocol in
+different functions.
 
-Although limiting the size of a handshake message seems an elementary problem to repair, the example shows that there was a significant amount of changes performed in the codebase which produced a negative impact in the project maintainability.
-Even though the developer tries to simplify the codebase with the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it also disrupts 2 of the guidelines proposed by the Software Improvement Group (SIG) for building maintainable software~\cite{Visser:2016:OREILLY}.
-It adds a new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the \emph{Write Simple Units of Code} guideline.
-And, it increases the codebase size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the \emph{Keep Your Codebase Small} guideline.
+Although limiting the size of a handshake message seems an elementary problem to
+repair, the example shows that there was a significant amount of changes
+performed in the codebase which produced a negative impact in the project
+maintainability. Even though the developer tries to simplify the codebase with
+the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it
+also disrupts $2$ of the guidelines proposed by the Software Improvement Group
+(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
+new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
+(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
+\emph{Write Simple Units of Code} guideline. And, it increases the codebase size
+in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the
+\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
+impact of this refactoring on the OpenSSL maintainability was negative.
 
-In this study, our concern is if while improving software security, developers are also inserting a negative impact on the maintainability of their software which may increase technical debt when choosing the easiest solution instead of the using the best approach.
-To answer the following two research questions, we use a dataset of security refactorings to measure the impact of the changes in the maintainability of open-source software.
+In this study, our concern is to study whether while improving software
+security, developers are also inserting a negative impact on the maintainability
+of their software which may increase technical debt when choosing the easiest
+solution instead of the using the best approach. To answer the following two
+research questions, we use a dataset of security refactorings to measure the
+impact of the changes in the maintainability of open-source software.
 
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability of open-source software?}
+\textit{\textbf{RQ1} What is the impact of security refactorings on the
+maintainability of open-source software?}
 \end{framed}
@@
 Often security flaws require refactoring code to make software more secure.
-However, there is no evidence yet of how security refactorings impact the maintainability of open-source software.
-Our suspicion is that developers tend to introduce technical debt in their software when refactoring the codebose to address a security flaw because they tend to choose the easiest path to solve it.
-To address it, we compute the maintainability value of $607$ commits using \emph{Better Code Hub}.
-The same approach is applied to a randomly generated dataset of regular commits - baseline - that we use to understand how maintainability evolves when security refactorings are performed versus when they are not.
+However, there is no evidence yet of how security refactorings impact the
+maintainability of open-source software. Our suspicion is that developers tend
+to introduce technical debt in their software when refactoring the codebose to
+address a security flaw because they tend to choose the easiest path to solve
+it. To address it, we compute the maintainability value of $607$ commits using
+\emph{Better Code Hub}. The same approach is applied to a randomly generated
+dataset of regular commits - baseline - that we use to understand how
+maintainability evolves when security refactorings are performed versus when
+they are not.
 
 \begin{framed}
-\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to affect open-source software maintainability?}
+\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
+affect open-source software maintainability?}
 \end{framed}
 
-There are security flaws that are more difficult to refactor than others.
-For instance, introducing a mechanism to secure authentication is not so easy to implement as fixing a cross-site scripting vulnerability with a \texttt{escape} function.
-Knowing which patterns are more likely to increase maintainability issues is one step forward to bring awareness to security engineers of what patterns need more attention.
-The taxonomy of security patterns used to answer this question is the same as the one provided by the authors of the dataset.
+There are security flaws that are more difficult to refactor than others. For
+instance, introducing a mechanism to secure authentication is not so easy to
+implement as fixing a cross-site scripting vulnerability with a \texttt{escape}
+function. Knowing which patterns are more likely to increase maintainability
+issues is one step forward to bring awareness to security engineers of what
+patterns need more attention. The taxonomy of security patterns used to answer
+this question is the same as the one provided by the authors of the dataset.
 Maintainability is measured separately for each pattern.
-
+%
 \section{Methodology}\label{sec:methodology}
-
-In this section, the methodology used to measure the impact of security refactorings on the maintainability of open-source software is presented in depth in the following sections and illustrated in Figure \ref{fig:met}.
-Aiming to answer the research questions presented in this paper, we use a dataset containing $716$ security refactorings collected from open-source software available on GitHub.
-$648$ of those refactorings were reported in a previous study conducted in 2017 \cite{Reis:2017:IJSSE}.
-In parallel, a baseline of regular commits was randomly collected from the list of projects of the main dataset to evaluate what is the impact of regular commits on the maintainability of open-source software.
-The Software Improvement Group's\footnote{SIG's website is available at https://www.sig.eu/ (Accessed on January 23, 2019)} web-based source code analysis service Better Code Hub (BCH) \footnote{BCH's website is available at https://bettercodehub.com/ (Accessed on January 23, 2019)} was used to collect the maintainability reports for both security and regular commits.
+%
+The methodology used to measure the impact of security refactorings on the
+maintainability of open-source software is discussed in the following sections
+and illustrated in Figure \ref{fig:met}. Aiming to answer the research questions
+outlined in the previous section, in this paper, we use a dataset containing
+$716$ security refactorings collected from open-source software available on
+GitHub. $648$ of those refactorings were reported in a previous study conducted
+in 2017 \cite{Reis:2017:IJSSE}. Moreover, a baseline of regular commits was
+randomly collected from the list of projects of the main dataset to evaluate
+what is the impact of regular commits on the maintainability of open-source
+software. The Software Improvement Group's\footnote{SIG's website is available
+at https://www.sig.eu/ (Accessed on \today{})} web-based source code
+analysis service Better Code Hub (BCH)\footnote{BCH's website is available at
+https://bettercodehub.com/ (Accessed on \today{})} was used to quantify
+maintainability for both security and regular commits.
 
@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
- 	\caption{Study Methodology \Sof{@todo: Improve it!}}
+ 	\caption{Study Methodology}
 	\label{fig:met}

### 2019-01-28 paper/ICPC19/icpc19.tex
pass S3

@@
 $716$ security refactorings collected from open-source software available on
-GitHub. $648$ of those refactorings were reported in a previous study conducted
-in 2017 \cite{Reis:2017:IJSSE}. Moreover, a baseline of regular commits was
+GitHub~\cite{Reis:2017:IJSSE}. Moreover, a baseline of regular commits was
 randomly collected from the list of projects of the main dataset to evaluate
@@
 software. The Software Improvement Group's\footnote{SIG's website is available
-at https://www.sig.eu/ (Accessed on \today{})} web-based source code
-analysis service Better Code Hub (BCH)\footnote{BCH's website is available at
+at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
+service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
 https://bettercodehub.com/ (Accessed on \today{})} was used to quantify
 maintainability for both security and regular commits.
-
+%
 \begin{figure}[h]
@@
 \end{figure}
-
+%
 \subsection{Dataset}
-
-To evaluate the impact of security refactorings on the maintainability of open-source software, we use a dataset of security flaws which is the outcome of mining more than $248$ GitHub\footnote{Github's website available at https://github.com/ (Accessed on January 23, 2019)} projects.
-Reis and Abreu (2017) mined open-source software aiming the extraction of real test cases - created by real developers on their daily basis development - to test and assess the performance of static analysis tools since using hand-seeded test cases or mutations could lead to misleading assessments of the capabilities of the tools.
-The study yielded to a dataset of $648$ test cases for $16$ different patterns.
-Each test case of the dataset is a triplet of folders: the commit before the refactoring, the commit responsible for the refactoring and the snippets of code that differ from one version to another (usually, called \textit{diff}) - where one can easily identify the code used to fix the security flaw.
-In this study, we focus on computing the maintainability of the commits before and after the security refactoring to evaluate if the impact was positive, negative or none.
+%
+To evaluate the impact of security refactorings on the maintainability of
+open-source software, we use a dataset of security flaws which is the outcome of
+mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
+software aiming at the extraction of real examples --- created by real
+developers --- to test and assess the performance of static analysis tools since
+using hand-seeded test cases or mutations could lead to misleading assessments
+of the capabilities of the tools \cite{just2014mutants}. The study yielded a
+dataset with $716$ test case for $16$ security patterns. Each test case of the
+dataset is a triplet: the commit before the refactoring, the commit responsible
+for the refactoring and the snippets of code that differ from one version to
+another (typically, called \textit{diff}) --- where one can easily identify the
+code used to fix the security flaw. In this study, we focus on computing the
+maintainability of the commits before and after the security refactoring to
+evaluate if the impact was positive, negative or none.
 
@@
 
-Reis and Abreu ($2017$) dataset commits were manually validated, flagged as security flaws and classified with one of the $16$ following patterns: \textit{Injection}, \textit{Broken Authentication and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery}, \textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs}, \textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service}, \textit{Path Traversal} and \textit{Miscellaneous}.
-However, the dataset evolved and now integrates new patterns, such as \textit{Resource Leak} and \textit{SHA-1 Hash Function}.
-A considerable amount of these patterns are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 2017~\cite{oswap:2017}.
+The test cases in the dataset are classified with one of the $16$ following
+patterns, as based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top
+10 of 2017~\cite{oswap:2017}: \textit{Injection}, \textit{Broken Authentication
+and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access
+Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure},
+\textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery},
+\textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs},
+\textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service},
+\textit{Path Traversal}, \textit{Miscellaneous}, \textit{Resource Leak} and
+\textit{SHA-1 Hash Function}.
 
-In total, the dataset contains $716$ security refactorings.
-These refactorings were computed by the BCH tool to calculate maintainability.
-However, some of the refactorings are not considered in the final results due to limitations of BCH: lack of language support and project size.
-The final dataset contains $607$ security refactorings.
-The Wilcoxon statistical test - described more in-depth in subsection \ref{sec:statsval} - is performed to evaluate the significance of our results.
-Each type of security flaw is tested, Figure \ref{fig:pat}.
-One of the main requirements of this test is that the input sample needs to comprise more than $20$ commits.
-Thus, the patterns not complying with this requirement were reallocated to the \textit{Miscellaneous} pattern.
-In the end, we highlight the following patterns:
+The $716$ refactorings in the dataset were run against the BCH toolset to
+calculate their maintainability reports. However, there are refactorings
+discarded from tour study due to limitations of BCH: in particular, lack of
+language support and project size. The final dataset used in this paper contains
+$607$ security refactorings, labelled as one of the following categories (we
+have decided to change the label to \textit{Miscellaneous} of the patterns in
+the dataset containing less than $20$ instances):
 
 \begin{itemize}
-	\item \textbf{Using Components with Known Vulnerabilities.} The majority of software produced today integrates several components, such as libraries and frameworks, which may affect the software if developers use vulnerable versions - usually known and disclosed somewhere on the Internet.
+	\item \textbf{Using Components with Known Vulnerabilities.} The majority of
+	software produced today integrates several components, such as libraries and
+	frameworks, which may affect the software if developers use vulnerable
+	versions - usually known and disclosed somewhere on the Internet.
 	\textit{Based on the OWASP Top 10 of 2017.}
-	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly implemented functionalities related to authentication and session management, allowing crackers to gain access to session tokens, passwords, keys and other sensitive data.
+
+	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly
+	implemented functionalities related to authentication and session management,
+	allowing crackers to gain access to session tokens, passwords, keys and other
+	sensitive data. \textit{Based on the OWASP Top 10 of 2013.}
+
+	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and
+	management usually allow crackers to send forged HTTP requests including
+	authentication information from the victim to vulnerable web application.
 	\textit{Based on the OWASP Top 10 of 2013.}
-	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow crackers to send forged HTTP requests including authentication information from the victim to vulnerable web application.
-\textit{Based on the OWASP Top 10 of 2013.}
-	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to flood or crashing services.
-This attacks usually occur when the system receives much traffic causing it to slow down and eventually stop.
-For example, flaws allowing crackers to trigger memory allocations corresponding to large length values (Listing \ref{lst:vuln}).
-	\item \textbf{Miscellaneous} This pattern comprises several others categories of security refactorings (e.g., \textit{path traversal and buffer overflow}).
-It contains several other security refactorings that do not have their own pattern yet.
-The patterns not satisfying the size requirement of more than 20 refactorings are contemplated here.
-	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow crackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites.
-\textit{Based on the OWASP Top 10 of 2017.}
-	\item \textbf{Injection.} When developers do not keep untrusted data separate from commands and queries.
-If a cracker sends a string that exploits the syntax of the interpreter, then an injection attack is possible (e.g., SQL and LDAP injection).
-\textit{Based on the OWASP Top 10 of 2017.}
-	\item \textbf{Memory Leak.} Memory management issue found more frequently in programming languages that do not manage memory automatically (e.g., C/C++ and Objective-C), i.e., where instead developers are responsible for handling it.
+
+	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to
+	flood or crashing services. This attacks usually occur when the system
+	receives much traffic causing it to slow down and eventually stop. For
+	example, flaws allowing crackers to trigger memory allocations corresponding
+	to large length values (Listing \ref{lst:vuln}).
+
+	\item \textbf{Miscellaneous} This pattern comprises several others categories
+	of security refactorings (e.g., \textit{path traversal and buffer overflow}).
+	It contains several other security refactorings that do not have their own
+	pattern yet. The patterns not satisfying the size requirement of more than 20
+	refactorings are contemplated here.
+
+	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
+	allow crackers to submit untrusted data to web browsers through malicious
+	scripts that can hijack the user sessions or redirect the user to malicious
+	sites. \textit{Based on the OWASP Top 10 of 2017.}
+
+	\item \textbf{Injection.} When developers do not keep untrusted data separate
+	from commands and queries. If a cracker sends a string that exploits the
+	syntax of the interpreter, then an injection attack is possible (e.g., SQL and
+	LDAP injection). \textit{Based on the OWASP Top 10 of 2017.}
+
+	\item \textbf{Memory Leak.} Memory management issue found more frequently in
+	programming languages that do not manage memory automatically (e.g., C/C++ and
+	Objective-C), i.e., where instead developers are responsible for handling it.
 	One of the main causes of DoS attacks.
@@
 
-The dataset contemplates security flaws for more than 13 different languages being PHP (39\%) and C (24\%) the most prevalent ones, Figure \ref{fig:lang}.
-To classify the software of the dataset, we use a taxonomy for open-source software published in an older study~\cite{7816479}.
-Figure \ref{fig:domain} presents the projects domain distribution: \textit{Aplication Software} ($25$), software that provides end-users with functional systems; \textit{Web libraries and frameworks} ($20$); \textit{System Software} ($18$), software that provides services and infrustructures (e.g., operating systems, servers and databases); \textit{Software Tool} ($13$), software that supports development (e.g., programming languages, compilers, package managers, IDEs); and, \textit{Non-web libraries and frameworks} ($18$), for desktop and mobile software.
+The dataset contemplates security flaws for more than 13 different languages
+being PHP (39\%) and C (24\%) the most prevalent ones; see Figure \ref{fig:lang}.
+To classify the applications in the dataset, we use a taxonomy for open-source
+software published in an older study~\cite{7816479}. Figure \ref{fig:domain}
+presents the projects domain distribution: \textit{Aplication Software} ($25$),
+software that provides end-users with functional systems; \textit{Web libraries
+and frameworks} ($20$); \textit{System Software} ($18$), software that provides
+services and infrustructures (e.g., operating systems, servers and databases);
+\textit{Software Tool} ($13$), software that supports development (e.g.,
+programming languages, compilers, package managers, IDEs); and, \textit{Non-web
+libraries and frameworks} ($18$), for desktop and mobile software.
+%
+Table \ref{tab:dataset} presents the descriptive statistics of the $94$
+open-source projects involved in this study, icluding number of forks, number of
+stars, number of watchers, number of contributors, number of commits, number of
+branches, number of releases,  size (cf. GitHub it is the size of the whole
+repository, including all of its history, in kilobytes), number of issues and pull
+requests.
 
@@
 \end{table*}
-
-Table \ref{tab:dataset} presents the descriptive statistics of the 94 open-source projects involved in this study.
-\Sof{@all: Should I expand this?
-I am not sure what it is important to say about this.}
-
+%
 \subsection{Security vs. Baseline Commits}
-
-Previous studies tried to measure the impact of refactorings on open-source software maintainability~\cite{HEGEDUS2018313} before.
-But to the best of our knowledge, none of them used the BCH model and tried to evaluate the impact of security refactorings compared with regular refactorings as we do.
-Thus, we analyze the maintainability of regular commits and use them as a baseline.
+%
+Previous studies attempted to measure the impact of code refactorings on
+open-source software maintainability~\cite{HEGEDUS2018313} before. But, to the
+best of our knowledge, no previous work has used the BCH model and focused on
+evaluating the impact of security refactorings and compare with regular
+refactorings as in this study. We analyze the maintainability of regular commits
+(i.e., commits not related with security fixes) and use them as a baseline.
 
@@
 security commit, one random commit is selected from the list of all commits of
-the belonging project. We originate the regular commits from the security
+that particular project. We originate the regular commits from the security
 commits to ensure that differences in maintainability are not consequence of
 characteristics of different projects.
-
+%
 \subsection{Maintainability Analysis}
 
-Better Code Hub is used to collect the maintanability reports of the refactorings of each project.
-Figure \ref{fig:guidelines} presents the 10 different guidelines proposed by BCH authors for delivering software that is easy to maintain \cite{Visser:2016:OREILLY}.
-The reports provided by the tool are based on the following guidelines:
+As saids before, the web-based source code analysis service \emph{Better Code
+Hub} (BCH) is used to collect the maintanability reports of the refactorings of
+each project. Figure \ref{fig:guidelines} presents the 10 guidelines
+proposed by BCH's authors for delivering software that is easy to maintain
+\cite{Visser:2016:OREILLY}.
 
@@
 
-All guidelines are measured in the same evaluation by BCH.
-During each guideline evaluation, the tool determines the compliance towards one guideline by establishing limits for the percentage of code allowed to be in each of the 4 risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and \emph{very high risk}).
-If the project does not violate the thresholds, then it is compliant with the guideline.
-These thresholds are calibrated by BCH using their own data - open-source and closed software systems.
-If a project is compliant with a guideline, it means that it is at least $65\%$ better than the software used by BCH to calculate the thresholds\footnote{Check the answer to \emph{How can I adjust the threshold for passing/not passing a guideline?} at https://bettercodehub.com/docs/faq (Accessed on January 27, 2019)}.
-Figure \ref{fig:bchrep} shows an example of the  guidelines report provided by BCH for a project after finishing its evaluation.
-These are the results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring - motivation example discussed, previously, in section \ref{sec:motivation}.
-This version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write Clean Code} guideline.
+All guidelines are computed in the same evaluation by BCH. During each guideline
+evaluation, the tool determines the compliance towards one guideline by
+establishing limits for the percentage of code allowed to be in each of the $4$
+risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and
+\emph{very high risk}). If the project does not violate the thresholds, then it
+is compliant with the guideline. These thresholds are calibrated by BCH using
+their own data/experience --- open-source and closed software systems. If a
+project is compliant with a guideline, it means that it is at least $65\%$
+better than the software used by BCH to calculate the thresholds\footnote{Check
+the answer to \emph{How can I adjust the threshold for passing/not passing a
+guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.
+Figure \ref{fig:bchrep} shows an example of the  guidelines report
+provided by BCH for a project after finishing its evaluation. These are the
+results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring -
+motivation example discussed, previously, in section \ref{sec:motivation}. This
+version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write
+Clean Code} guideline.
 
@@
  	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
- 	\caption{Maintainability report of OpenSSL CVE-2014-3506 vulnerability refactoring for the guideline \emph{Write Simple Units of Code} provided by \emph{Better Code Hub}.
-This version of OpenSSL does not comply with the guideline in the example since the bars do not reach the threshold points.
-This example only complies with 1/10 guidelines (\emph{Write Clean Code}).}
+ 	\caption{Maintainability report of OpenSSL CVE-2014-3506 vulnerability
+refactoring for the guideline \emph{Write Simple Units of Code} provided by
+\emph{Better Code Hub}. This version of OpenSSL does not comply with the
+guideline in the example since the bars do not reach the threshold points. This
+example only complies with 1/10 guidelines (\emph{Write Clean Code}).}
 	\label{fig:bchrep}
@@
 
-\emph{Units} are the smallest groups of code that can be maintained and executed independently \cite{Visser:2016:OREILLY}.
-McCabe \cite{1702388} is a software metric that calculates the number of linearly independent paths of the source code.
-BCH uses this metric to measure the number of branch points in the \emph{Write Simple Units of Code} guideline.
-The bar represents the top 30 of units that violates the guideline, sorted by severity, which is indicated by the colors of the checkboxes.
-The green bar represents the number of compliant branch points per unit (\emph{at most 5}), i.e., the units are compliant with ISO 25010 \cite{iso:2011}.
-Yellow, orange and red bars represent units that do not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high (\emph{above 25}) severity levels.
-In the bar, there are marks that pinpoint the compliance thresholds for each severity level.
+\emph{Units} are the smallest groups of code that can be maintained and executed
+independently \cite{Visser:2016:OREILLY}. McCabe \cite{1702388} is a software
+metric that calculates the number of linearly independent paths of the source
+code. BCH uses this metric to measure the number of branch points in the
+\emph{Write Simple Units of Code} guideline. The bar represents the top 30 of
+units that violates the guideline, sorted by severity, which is indicated by the
+colors of the checkboxes. The green bar represents the number of compliant
+branch points per unit (\emph{at most 5}), i.e., the units are compliant with
+ISO 25010 \cite{iso:2011}. Yellow, orange and red bars represent units that do
+not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
+(\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
+compliance thresholds for each severity level.
 
@@
 
-Aiming to analyze the impact of security refactorings, we use BCH to compute maintainability of two different versions of the project (Figure \ref{fig:commit}):
+Aiming to analyze the impact of security refactorings, we use BCH to compute
+maintainability of two different versions of the project (Figure
+\ref{fig:commit}):
 \begin{itemize}
-	\item $v_{s-1}$, the version containing the security flaw, i.e., before the refactoring;
-	\item $v_{s}$, the version after fixing the security flaw, i.e., after the refactoring;
+	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
+	refactoring;
+	\item $v_{s}$, the version after fixing the security flaw, i.e., after the
+	refactoring;
 \end{itemize}
 
-The BCH tool does not compute the final score that our study needs to compare maintainability amongst different project versions.
-Thus, our work proposes an equation to capture the distance between the current state of the project and the standard thresholds based on previous work \cite{Olivari:2018}.
+The BCH tool does not compute the final score that our study needs to compare
+maintainability amongst different project versions. Thus, our work proposes an
+equation to capture the distance between the current state of the project and
+the standard thresholds based on previous work \cite{Olivari:2018}.
 \begin{itemize}
-	\item \textbf{Size project changes do not affect the maintainability difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we aim to evaluate the alike security patterns occurring in different projects.
-Thus, normalization based on its size cannot be used - we convert relative data to the respective number of lines of code.
-	\item \textbf{Distance to lower severity levels are less penalized than to the thresholds in high severity levels.} Severity level weights based on the severity level to count lines of code that violate maintainability guidelines.
+	\item \textbf{Size project changes do not affect the maintainability
+	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we
+	aim to evaluate the alike security patterns occurring in different projects.
+	Thus, normalization based on its size cannot be used - we convert relative
+	data to the respective number of lines of code.
+	\item \textbf{Distance to lower severity levels are less penalized than to the
+	thresholds in high severity levels.} Severity level weights based on the
+	severity level to count lines of code that violate maintainability guidelines.
 \end{itemize}
 
-We compute the mean average of the maintainability score $M(v)$ for all the selected guidelines, as follows:
+We compute the mean average of the maintainability score $M(v)$ for all the
+selected guidelines, as follows:
 
@@
 \noindent
-where $G$ is the group of selected maintainability guidelines from BCH (Figure \ref{fig:guidelines}) and $v$ is the
-version of the software under evaluation. The maintenance $M$ based on the guideline $g$ for a given version of a project is computed with the following equation:
+where $G$ is the group of selected maintainability guidelines from BCH (Figure
+\ref{fig:guidelines}) and $v$ is the version of the software under evaluation.
+The maintenance $M$ based on the guideline $g$ for a given version of a project
+is computed with the following equation:
 
@@
 \noindent
-where $C$ is the compliance with the maintainability guideline for the given severity level (medium, high, and very high) and
-$L$ is the group of severity levels of maintainability infractions. The compliance $C$ for a given severity level $l$ is derived by:
+where $C$ is the compliance with the maintainability guideline for the given
+severity level (medium, high, and very high) and $L$ is the group of severity
+levels of maintainability infractions. The compliance $C$ for a given severity
+level $l$ is derived by:
 
@@
 \noindent
-where $LOC_{compliant}(l)$ are the lines of code that comply with the guideline at the given severity level $l$, $LOC_{\neg compliant}(l)$ are the lines of code that do not comply with the guideline at the given
-severity level $l$ and $w(l)$ is the weight factor to boost the impact of
-non-compliant lines in comparison to compliant lines. Finally, the term $w(l)$ is calculated as follows:
+where $LOC_{compliant}(l)$ are the lines of code that comply with the guideline
+at the given severity level $l$, $LOC_{\neg compliant}(l)$ are the lines of code
+that do not comply with the guideline at the given severity level $l$ and $w(l)$
+is the weight factor to boost the impact of non-compliant lines in comparison to
+compliant lines. Finally, the term $w(l)$ is calculated as follows:
 
@@
 \noindent
-where $T(l)$ is the threshold in percentage of the lines of code that are accepted to be non-compliant with the guideline for the severity level $l$.
-This is a standard value defined by BCH.
-In other words, the factor $w$ is used in Equation \ref{eq:3} to highlight the lines of code that are not complying with the guideline.
-Then, we compute the difference of maintainability between the security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in Figure \ref{fig:commit}.
+where $T(l)$ is the threshold in percentage of the lines of code that are
+accepted to be non-compliant with the guideline for the severity level $l$. This
+is a standard value defined by BCH. In other words, the factor $w$ is used in
+Equation \ref{eq:3} to highlight the lines of code that are not complying with
+the guideline. Then, we compute the difference of maintainability between the
+security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
+Figure \ref{fig:commit}.
 
 \subsection{Statistical Validation}\label{sec:statsval}
-
-To validate the maintainability differences in different groups of commits (e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank test \cite{10.2307/3001968} with the significance level $\alpha = 0.05$.
-In other words, we test the null hypothesis that the maintainability difference between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a security-commit) follows a symmetric distribution around $0$.
-However, when the difference between groups is zero, the observations are discarded.
-Thus, we use a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which provides an alternative that incorporates the cases where the maintainability difference is zero \cite{10.2307/2282543}.
-To understand the effect-size, as advocated by the Common-language effect sizes \cite{graw:1992}, we compute the mean difference, the median of the difference, and the percentage of cases that reduce maintainability.
-
+%
+To validate the maintainability differences in different groups of commits
+(e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
+test with the significance level $\alpha = 0.05$ \cite{10.2307/3001968} . In
+other words, we test the null hypothesis that the maintainability difference
+between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
+security-commit) follows a symmetric distribution around $0$. However, when the
+difference between groups is zero, the observations are discarded. Thus, we use
+a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which
+provides an alternative that incorporates the cases where the maintainability
+difference is zero \cite{10.2307/2282543}. To understand the effect-size, as
+advocated by the Common-language effect sizes \cite{graw:1992}, we compute the
+mean difference, the median of the difference, and the percentage of cases that
+reduce maintainability.
+%
 \section{Results}\label{sec:results}

### 2019-01-28 paper/ICPC19/icpc19.tex
acks
adding placehoder for t2v.

@@
 \section{Threats to Validity}\label{sec:threats}
+%
+In this section, we discuss potential threats to the validity of our work.
+%
+\subsection{Construct}
+%
+We use a metric derived from the reports provided by BCH as a construct of
+software maintainability. Different projects and contexts may require different
+maintainability standards. However, BCH uses a representative benchmark of
+closed and open source software systems to compute the thresholds used in each
+maintainability guideline\cite{Visser:2016:OREILLY,baggen2012standardized}.
+In addition, the benchmark is updated every year~\cite{Visser:2016:OREILLY}.
 
+The lifespan of a given energy commit is an approximation based on a single
+change in one of the lines of code submitted by that commit. Although changes in
+whitespace are not considered, we did not perform any evaluation on the type of
+change that is performed. Some changes may be trivial while others may have
+required more efforts from developers. Related work has proposed approaches to
+automatically classify changes (e.g., bug fixes\cite{da2017framework}, and
+refactorings~\cite{pascarella2018self}), but we have not explored them.
+Nevertheless, we understand that the differences in the lifespan between our
+dataset of energy changes and the baseline provide interesting leads for
+discussion and new research directions.
+
+\subsection{Internal}
+
+We resort to a set of security changes that were collected from previous
+work~\cite{Reis:2017:IJSSE}. These dataset was collected using the commit
+message provided by developers to classify a given commit as a security change.
+We understand that this approach discards security changes that did not have an
+explicit commit message describing them as such.
+
+In the case of baseline commits, they were collected by randomly select a commit
+in the same project that a given security change occurs. We use this approach to
+control differences incurred by the specificities of each project. Still,
+maintainability may be affected by different coding styles and experience level
+from developers of the same project. We do not evaluate differences at that
+level. Moreover, we do not evaluate the maintainbility difference for all
+regular commits in a project. We assume that the size of the dataset is enough
+to mitigate random variations in the maintainability differences of the baseline.
+
+\subsection{External}
+
+The dataset used in this work comprises code changes of free and open
+source applications. Our methodology requires access to data that is not
+publicly available for commercial apps. The extent to which this findings
+generalize to commercial applications with non-open source licenses remains
+a question to be answered in future work.
+
+Moreover, all the security commits in this work are described in English.
+%
 \section{Related Work}\label{sec:rw}
@@
 
-We thank to the Software Improvement Group (SIG) team for all the support and help in validating our methodology.
+\noindent
+We thank to the Software Improvement Group (SIG) team for all the support and
+help in validating our methodology.
 

### 2019-01-28 paper/ICPC19/icpc19.tex
minor fixes

--- 2019-01-28 paper/ICPC19/icpc19.tex
@@
   escapeinside={/*!}{!*/},
-  moredelim=**[l][\color{mygreen}]{+},
-  moredelim=*[l][\color{red}]{-}
+  moredelim=**[l][\color{mygreen}]{+\ },
+  moredelim=*[l][\color{red}]{-\ }
 }
@@
 \begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability \Sof{@todo: Fix the regex issue}},label={lst:vuln}]
-+static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
-+
++ static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
 +  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
@@
 +  return max_len;
-+}
++ }
 
@@
 - unsigned long frag_len = msg_hdr->frag_len, max_len; /*!\annotation{lst:func3}!*/
--
+
 - if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
@@
 
-- if(DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
+- if(DTLS1_HM_HEADER_LENGTH +
+-    	SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
 -    max_len = s->max_cert_list;
 - else
--    max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
+-    max_len = DTLS1_HM_HEADER_LENGTH +
+-   		SSL3_RT_MAX_ENCRYPTED_LENGTH;
 
@@
 reduce maintainability.
+
 %
@@
 
+The impact of security and baseline commits in the maintainability of open-source software is presented in Figure \ref{fig:secvsreg}.
+For each type of change, three types of impact of refactorings in the software maintainability are presented: \emph{negative}, i.e., maintainability decreased (\emph{red}); \emph{none}, i.e., the maintainability did not suffer any changes (\emph{yellow}); and, \emph{positive}, i.e., the maintainability increased (\emph{negative}).
+
+\begin{figure}[h]
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/maintainability.pdf}
+ 	\caption{Maintainability difference for security and baseline refactorings}
+	\label{fig:secvsreg}
+\end{figure}
+
 \begin{framed}
@@
 \end{framed}
-
-\begin{figure}[h]
- 	\centering
- 	\includegraphics[width=0.45\textwidth]{figures/maintainability.pdf}
- 	\caption{Maintainability difference for security and baseline refactorings}
-	\label{fig:secvsreg}
-\end{figure}
 

### 2019-01-28 paper/ICPC19/icpc19.tex
remove todo

@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability \Sof{@todo: Fix the regex issue}},label={lst:vuln}]
+\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the
+CVE-2014-3506 vulnerability},label={lst:vuln}]
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/

### 2019-01-28 paper/ICPC19/icpc19.tex
sfrac
more results

@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the
-CVE-2014-3506 vulnerability},label={lst:vuln}]
+\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability},label={lst:vuln}]
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
@@
 The impact of security and baseline commits in the maintainability of open-source software is presented in Figure \ref{fig:secvsreg}.
-For each type of change, three types of impact of refactorings in the software maintainability are presented: \emph{negative}, i.e., maintainability decreased (\emph{red}); \emph{none}, i.e., the maintainability did not suffer any changes (\emph{yellow}); and, \emph{positive}, i.e., the maintainability increased (\emph{negative}).
+For each type of change, three types of impact of refactorings in the software maintainability are presented: \emph{negative}, maintainability decreases (\emph{red}); \emph{none}, maintainability stays the same (\emph{yellow}); and, \emph{positive}, maintainability increases (\emph{green}).
+In front of each group, it is presented the mean ($\overline{x}$) and median ({$\widetilde{x}$) of the maintainability difference and the p-value resulting from the Wilcoxon signed-rank test.
 
@@
 \end{figure}
+
+Regarding security commits results, we observe that maintainability decreases in 40.20\% (244), increases in 29.33\% (178) and stays the same in 30.5\% (185) of the cases after security refactorings being applyed.
+The resulting p-value of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$.
+Since the p-value is below the significance level of $0.05$, it is fair to state that \emph{security refactorings have a negative impact in the maintainability of open-source software}.
+
+For regular commits, we also observe that maintainability decreases in 41.35\% (251) of the cases after regular refactorings being applyed.
+But in contrast to security commits, it increases in 58.16\% (353) and stays the same in 0.5\% (3).
+The resulting p-value of the Wilcoxon signed-rank test is $1.54$x$10^{-6}$.
+Since the p-value is below the significance level of $0.05$, it is fair to state that \emph{regular refactorings have a positive impact in the maintainability of open-source software}.
 

### 2019-01-29 paper/ICPC19/icpc19.tex
new update

@@
 \definecolor{mypink3}{cmyk}{0, 0.7808, 0.4429, 0.1412}
+
+\usepackage{xfrac}
 
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the CVE-2014-3506 vulnerability},label={lst:vuln}]
+\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the
+CVE-2014-3506 vulnerability},label={lst:vuln}]
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
@@
 guideline in the example since the bars do not reach the threshold points. This
-example only complies with 1/10 guidelines (\emph{Write Clean Code}).}
+example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
 	\label{fig:bchrep}
@@
 
-Regarding security commits results, we observe that maintainability decreases in 40.20\% (244), increases in 29.33\% (178) and stays the same in 30.5\% (185) of the cases after security refactorings being applyed.
+Regarding security commits results, we observe that maintainability decreases in 40.20\% (244), increases in 29.33\% (178) and stays the same in 30.5\% (185) of the cases after security refactorings being applied.
 The resulting p-value of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$.
@@
 
+\Sof{@todo: effect size}
+
 \begin{framed}
@@
 \end{framed}
+
+The impact of security refactorings on the maintainability of open-source software per group of security flaws is presented in Figure \ref{fig:pat}.
+The results of the Wilcoxon signed-rank test by pattern returned statistical evidence which supports the maintainability of open-source software decreases for 4 different patterns: Broken Autentication (p = 0.018), Denial-of-Service (p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
+The results also returned statistical significance that evidence that maintainability is more likely to not suffer differences for open-source software after removing Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site Scripting Vulnerabilities (p = 0.001).
+Injection and Cross-Site Request Forgery produced more cases where the refactorings had a negative impact in the maintainability.
+However, the results obtained for these patterns are not statistically significant.
+
+For the Broken Authentication pattern, 55.3\% of the refactorings (21) had a negative impact in the maintainability, 28.95\% (11) improved maintainability and 15.8\% (6) of the changes did not add any impact in the maintainability of open-source software.
+
+%    neg  pos  nul
+% ml   41   28   11
+% injec   32   31   26
+% xss   27   24   63
+% misc   84   62   49
+% dos   19   12    9
+% csrf   17    9    4
+% auth   21   11    6
+% ucwkv    3    1   17
+
+%        neg       pos       nul
+% ml  0.512500  0.350000  0.137500
+% injec  0.359551  0.348315  0.292135
+% xss  0.236842  0.210526  0.552632
+% misc  0.430769  0.317949  0.251282
+% dos  0.475000  0.300000  0.225000
+% csrf  0.566667  0.300000  0.133333
+% auth  0.552632  0.289474  0.157895
+% ucwkv  0.142857  0.047619  0.809524
 

### 2019-01-29 paper/ICPC19/icpc19.tex
abstract fixed, resuls finished, discussion introduced + rq brieflyanswered + threats to validity finished

@@
   more secure comes at a cost on maintainability. This is particularly evident
-  in refactorings to deal with \textit{Broken Authentication} and \textit{Cross-Site Request Forgery} attacks.
+  in refactorings to deal with \emph{Broken Authentication and Session Management}, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}.
 \end{abstract}
@@
 
-The impact of security refactorings on the maintainability of open-source software per group of security flaws is presented in Figure \ref{fig:pat}.
-The results of the Wilcoxon signed-rank test by pattern returned statistical evidence which supports the maintainability of open-source software decreases for 4 different patterns: Broken Autentication (p = 0.018), Denial-of-Service (p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
+The impact of each type of security refactoring on the maintainability of open-source software is presented in Figure \ref{fig:pat}.
+The results of the Wilcoxon signed-rank test by pattern returned statistical evidence which supports that the maintainability of open-source software decreases for 4 different patterns: Broken Autentication (p = 0.018), Denial-of-Service (p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
 The results also returned statistical significance that evidence that maintainability is more likely to not suffer differences for open-source software after removing Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site Scripting Vulnerabilities (p = 0.001).
@@
 
-For the Broken Authentication pattern, 55.3\% of the refactorings (21) had a negative impact in the maintainability, 28.95\% (11) improved maintainability and 15.8\% (6) of the changes did not add any impact in the maintainability of open-source software.
+Projects with Broken Authentication flaws are the most affected ones by security refactorings.
+55.3\% (21) of the refactorings had a negative impact in the maintainability while 35\% (11) had a positive impact and only 13.8\% (6) did not suffer any change.
+The results also show that fixing memory leaks affects the maintainability of 51.3\% (41) of the cases and leaves only 13.8\% (11) of the cases unaffected.
+35\% (28) are affected positively.
+Refactoring vulnerabilities that allow denial-of-service attacks (e.g., CVE-2014-3506) has a negative impact of 47.5\% (19) in the maintainability of open-source software.
+30.0\% (12) of the cases exhibit maintainability improvement while 22.5\% (9) show that overall maintainability does not suffer any change.
+Miscellaneous refactorings pattern integrates several types of security flaws whose impact is of 43.1\% (84 cases) negative in the maintainability of open-source sofware.
+31.8\% (62) of the cases do not exhibit any change in maintainability and 25.1\% (49) of cases have a positive impact on the maintainability of open-source software.
 
-%    neg  pos  nul
-% ml   41   28   11
-% injec   32   31   26
-% xss   27   24   63
-% misc   84   62   49
-% dos   19   12    9
-% csrf   17    9    4
-% auth   21   11    6
-% ucwkv    3    1   17
-
-%        neg       pos       nul
-% ml  0.512500  0.350000  0.137500
-% injec  0.359551  0.348315  0.292135
-% xss  0.236842  0.210526  0.552632
-% misc  0.430769  0.317949  0.251282
-% dos  0.475000  0.300000  0.225000
-% csrf  0.566667  0.300000  0.133333
-% auth  0.552632  0.289474  0.157895
-% ucwkv  0.142857  0.047619  0.809524
+It is also important to notice that overall the maintainability of open-source software is not affected by Cross-Site Scripting vulnerabilities and by Using Components with Know Vulnerabilities.
+Cross-Site Scripting does not suffer any decrease in the maintainability of 55.3\% (63) of cases and 21.1\% (24) have a positive effect on it.
+Only 23.7\% (27) of the cases hinder the software maintainability.
+From 21 cases evaluated for Using Componentes with Known Vulnerabilities, only 4.8\% (1) has a positive effect in the maintainability of software.
+14.3\% (3) harms maintainability software and 81\% has no effect in the maintainability.
 
@@
 
+In this section, we discuss the results and answer the proposed research questions.
+
 \begin{framed}
@@
 \end{framed}
+
+\textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
 
@@
 
+\textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}}
+
 \section{Threats to Validity}\label{sec:threats}
 %
-In this section, we discuss potential threats to the validity of our work.
+The following section presents the potential threats to validity of this study.
 %
@@
 %
-We use a metric derived from the reports provided by BCH as a construct of
-software maintainability. Different projects and contexts may require different
-maintainability standards. However, BCH uses a representative benchmark of
-closed and open source software systems to compute the thresholds used in each
-maintainability guideline\cite{Visser:2016:OREILLY,baggen2012standardized}.
-In addition, the benchmark is updated every year~\cite{Visser:2016:OREILLY}.
-
-The lifespan of a given energy commit is an approximation based on a single
-change in one of the lines of code submitted by that commit. Although changes in
-whitespace are not considered, we did not perform any evaluation on the type of
-change that is performed. Some changes may be trivial while others may have
-required more efforts from developers. Related work has proposed approaches to
-automatically classify changes (e.g., bug fixes\cite{da2017framework}, and
-refactorings~\cite{pascarella2018self}), but we have not explored them.
-Nevertheless, we understand that the differences in the lifespan between our
-dataset of energy changes and the baseline provide interesting leads for
-discussion and new research directions.
+The maintainability formula was inferred based on the reports of the Better Code Hub. 
+Since the dataset has a broad group of open-source projects with different backgrounds, this metric may not be suitable for all of them. However, BCH does have a representative benchmark of
+closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012} (Figure \ref{fig:guidelines}).
 
@@
 
-We resort to a set of security changes that were collected from previous
-work~\cite{Reis:2017:IJSSE}. These dataset was collected using the commit
-message provided by developers to classify a given commit as a security change.
-We understand that this approach discards security changes that did not have an
-explicit commit message describing them as such.
+The security refactorings dataset provided by the previous
+work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub commits produced by
+project developers to classify the changes perfomed by them to fix security flaws. This approach discards security refactorings that were not explicit in commits messages.
 
-In the case of baseline commits, they were collected by randomly select a commit
-in the same project that a given security change occurs. We use this approach to
-control differences incurred by the specificities of each project. Still,
-maintainability may be affected by different coding styles and experience level
-from developers of the same project. We do not evaluate differences at that
-level. Moreover, we do not evaluate the maintainbility difference for all
-regular commits in a project. We assume that the size of the dataset is enough
-to mitigate random variations in the maintainability differences of the baseline.
+Baseline commits are retrieved randomly by selecting one commit from the same project of a given security refactoring. This approach softens the differences that may result from the
+characteristics of each project.
+However, maintainability still may be affected by the developers experience, coding style and software contribution polices.
+But maintainability difference is not evaluated at this level.
 
@@
 
-The dataset used in this work comprises code changes of free and open
-source applications. Our methodology requires access to data that is not
-publicly available for commercial apps. The extent to which this findings
-generalize to commercial applications with non-open source licenses remains
-a question to be answered in future work.
+In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation.
+Thus, answering these questions to closed commercial applications may be a future work opportunity.
 
-Moreover, all the security commits in this work are described in English.
+All the security commits are described in English.
 %

### 2019-01-29 paper/ICPC19/icpc19.tex
balancing columns in last page
changes

@@
 
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+Unmaintainable code is hard to test, analyse and re-use.
 
-The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability,
-functional suitability, performance efficiency, compatibility, usability, reliability, security,
-and portability. This paper focuses solely on maintainability.
+It is hard to find bugs and vulnerabilities in high complex code, this some developers then to forget maintainability.
+Therefore, we want to bring awareness for the fact that developers need to write maintainable code to prevent the introduction of vulnerabilities in the code.
 
-\Sof{=============== READY START!}
+The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability, functional suitability, performance efficiency, compatibility, usability, reliability, security, and portability.
+This paper focuses solely on maintainability.
 
@@
 \toprule
-      & forks   & stars   & watchers & contributors & commits  & branches & releases & size      & issues & pull requests  \\ \midrule
+      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
 mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
@@
 
+40.2\% of the security refactoring have a negative impact on the maintainability of open-source software, Figure \ref{fig:secvsreg}.
+
 \begin{framed}

### 2019-01-29 scripts/get_maintainability.py

@@ -5,12 +5,17 @@ import collections
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
-from scipy.stats import wilcoxon
+from scipy.stats import wilcoxon, shapiro
+from statistics import mean, stdev
+from math import sqrt
 
 def _print_hypothesis_test(differences):
     """Paired Wilcoxon signed-rank test (N should be > 20)"""
+    _, pvalue = shapiro(differences)
+    print("Shapiro-Wilk test for normality: {}".format(pvalue))
     test, pvalue = _hypothesis_test(differences)
-    stats = {'test': [test], 'med':[pd.DataFrame({'diff': differences})['diff'].median()],'pvalue': [pvalue]}
+    stats = {'test': [test], 'med':[pd.DataFrame({'diff': differences})['diff'].median()],'pvalue': [pvalue], 
+                'mean':[pd.DataFrame({'diff': differences})['diff'].mean()]}
     df = pd.DataFrame(stats)
     print("Wilcoxon signed-rank test {}, p-value: {}".format(test,pvalue))
     return df
@@ -47,21 +52,17 @@ def count(id, pattern, p):
 def total_main_barchart(total_sec, total_reg, graphics, result):
     
     sec = sum([value for key,value in total_sec.items()])
-    reg = sum([value for key,value in total_reg.items()]) 
+    reg = sum([value for key,value in total_reg.items()])
     
     d = {'neg' : pd.Series([total_reg['neg']/reg, total_sec['neg']/sec]),
         'pos' : pd.Series([total_reg['pos']/reg, total_sec['pos']/sec]),
         'none' : pd.Series([total_reg['nul']/reg, total_sec['nul']/sec]),
         'type' : pd.Series(['Regular Change','Security Change'])}
-
-    # d = {'neg' : pd.Series([total_reg['neg'], total_sec['neg']]),
-    #     'pos' : pd.Series([total_reg['pos'], total_sec['pos']]),
-    #     'none' : pd.Series([total_reg['nul'], total_sec['nul']]),
-    #     'type' : pd.Series(['Regular Change','Security Change'])}
     
     index = np.arange(len(d['type']))
     
     df = pd.DataFrame(d)
[…]

### 2019-01-29 scripts/histogram.py

@@ -22,24 +22,27 @@ def language_dist(commits):
             dic[i] += 1
     
     dic_lang = {'lang':[], 'no':[]}
-    for i in dic:
-        if i == 'others':
-            no = dic['others']
-        else:
-            dic_lang['lang'].append(lang_trans[i])
-            dic_lang['no'].append(dic[i]/len(df['language']))
     
     dic_lang['lang'].append('Others')
     dic_lang['no'].append(dic['others']/len(df['language']))
     
-    lang = pd.DataFrame(dic_lang)
+    for i in dic:
+        if i != 'others':
+        #     no = dic['others']
+        # else:
+            dic_lang['lang'].append(lang_trans[i])
+            dic_lang['no'].append(dic[i]/len(df['language']))
     
-    ax = lang['no'].plot(kind='bar', figsize=(10, 7), fontsize=11, color="#2541F0")
-    vals = ax.get_yticks()
-    ax.set_yticklabels(['{:,.0%}'.format(x) for x in vals])
-    ax.set_xticklabels(dic_lang['lang'], minor=False)
-    plt.subplots_adjust(left=0.1, right=0.9, top=0.9, bottom=0.3)
-    plt.gca().yaxis.grid(True, linestyle='--')
+    
+    lang = pd.DataFrame(dic_lang)
+    print(lang)
+    
+    ax = lang['no'].plot(kind='barh', figsize=(10, 11), fontsize=12, color="#3F86DB")
+    vals = ax.get_xticks()
+    ax.set_xticklabels(['{:,.0%}'.format(x) for x in vals])
+    ax.set_yticklabels(dic_lang['lang'], minor=False)
+    plt.subplots_adjust(left=0.2, right=0.95, top=0.9, bottom=0.05)
+    plt.gca().xaxis.grid(True, linestyle='--')
     
     plt.savefig('../paper/ICPC19/figures/language_dist.pdf')
 
@@ -48,7 +51,7 @@ def type_dist(projects):
     df = pd.read_csv(projects)
     
     dic = {}
-    for i in df['type']:
+    for i in df['domain']:
         if i not in dic:
             dic[i] = 1
         else:
@@ -61,21 +64,26 @@ def type_dist(projects):
         dic_type['no'].append(dic[i]) 
                 
     app_type = pd.DataFrame(dic_type)
+    print(app_type)
     
[…]

### 2019-01-29 scripts/maintainability/better_code_hub.py

@@ -180,15 +180,20 @@ def _compute_maintainability_for_guideline(guideline, total_loc):
         ]
 
     bad_thresholds = guideline['qualityProfileComplianceThresholds'][1:]
+    print(bad_thresholds)
     return _compute_distance_to_thresholds(volumes, bad_thresholds)
 
 def _compute_distance_to_thresholds(volumes, bad_thresholds):
     results = []
     for index, threshold in enumerate(bad_thresholds):
+        print('threshold=', threshold)
         good_code_loc = sum(volumes[:index+1])
+        print('good_code_loc', good_code_loc)
         bad_code_loc = sum(volumes[index+1:])
+        print('bad_code_loc', bad_code_loc)
         granularity = 0.01 # thresholds have a minimum of 0.01
         bad_code_factor = ((1-threshold)/(threshold+granularity))
+        print('bad_code_factor', bad_code_factor)
         results.append(good_code_loc - bad_code_factor*bad_code_loc)
     return mean(results)
 

### 2019-01-29 paper/ICPC19/icpc19.tex
minor fixes

@@
 \usepackage{xcolor}
-\definecolor{mypink3}{cmyk}{0, 0.7808, 0.4429, 0.1412}
 
 \usepackage{xfrac}
+
+\usepackage{balance}
+
+\usepackage{array}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 \begin{lstlisting}[caption={Fix provided by OpenSSL developers to the
-CVE-2014-3506 vulnerability},label={lst:vuln}]
+\\CVE-2014-3506 vulnerability},label={lst:vuln}]
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
@@
 
-As saids before, the web-based source code analysis service \emph{Better Code
+As said before, the web-based source code analysis service \emph{Better Code
 Hub} (BCH) is used to collect the maintanability reports of the refactorings of
-each project. Figure \ref{fig:guidelines} presents the 10 guidelines
+each project. Table \ref{tab:guidelines} presents the 10 guidelines
 proposed by BCH's authors for delivering software that is easy to maintain
@@
 
-\begin{figure}[H]
- 	\centering 	\includegraphics[width=0.49\textwidth]{figures/guidelines.pdf}
+\begin{table}[H]
 	\caption{Guidelines to produce maintainable code}
-	\label{fig:guidelines}
-\end{figure}
+\begin{tabular}{L{2.5cm}L{5.5cm}}
+
+\toprule
+\textbf{10 Guidelines} & \textbf{Description}\\
+\midrule
+\textbf{Write Short Units of Code} & Limit code units to 15 LOCs because smaller units are easier to understand, reuse and test them\\\midrule
+\textbf{Write Simple Units of Code} & Limit branch points to 4 per unit because it makes units easier to test and modify \\\midrule
+\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at multiple places (inefficient and error-prone)\\\midrule
+\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most 4 because it makes units easier to understand and reuse\\\midrule
+\textbf{Separate Concerns in Modules} & Avoid large modules because changes in loosely coupled databases are easier to oversee and execute\\\midrule
+\textbf{Couple Architecture Components Loosely} & Minimize the amount of code within modules that is exposed to modules in other components\\\midrule
+\textbf{Keep Architecture Components Balanced} & Balancing the number of components ease locating code and allow for isolated maintenance\\\midrule
+\textbf{Keep your Codebase Small} & Reduce and avoid the system size because small products are easier to manage and maintain\\\midrule
+\textbf{Automate Tests} & Test your codebase because it makes development predictable and less risky\\\midrule
+\textbf{Write Clean Code} & Avoid producing software with code smells because it is more likely to be maintainable in the future\\
+\bottomrule
+\end{tabular}
+\label{tab:guidelines}
+\end{table}
 
@@
 compliance thresholds for each severity level.
+\Sof{Adress thresholds confusion}
 
@@
 \noindent
-where $G$ is the group of selected maintainability guidelines from BCH (Figure
-\ref{fig:guidelines}) and $v$ is the version of the software under evaluation.
+where $G$ is the group of selected maintainability guidelines from BCH (Table
+\ref{tab:guidelines}) and $v$ is the version of the software under evaluation.
 The maintenance $M$ based on the guideline $g$ for a given version of a project
@@
 30.0\% (12) of the cases exhibit maintainability improvement while 22.5\% (9) show that overall maintainability does not suffer any change.
-Miscellaneous refactorings pattern integrates several types of security flaws whose impact is of 43.1\% (84 cases) negative in the maintainability of open-source sofware.
+Miscellaneous refactorings pattern integrates several types of security flaws whose impact is of 43.1\% (84) negative in the maintainability of open-source sofware.
 31.8\% (62) of the cases do not exhibit any change in maintainability and 25.1\% (49) of cases have a positive impact on the maintainability of open-source software.
@@
 Since the dataset has a broad group of open-source projects with different backgrounds, this metric may not be suitable for all of them. However, BCH does have a representative benchmark of
-closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012} (Figure \ref{fig:guidelines}).
+closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012} (Table \ref{tab:guidelines}).
 
@@
 
+\balance
+
 {

### 2019-01-29 paper/ICPC19/icpc19.tex
paper

@@
   more secure comes at a cost on maintainability. This is particularly evident
-  in refactorings to deal with \emph{Broken Authentication and Session Management}, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}.
+  in refactorings to deal with \emph{Broken Authentication and Session Management}, 
+  \emph{Denial-of-Service} attacks and \emph{Memory Leaks}.
 \end{abstract}
@@
 This paper focuses solely on maintainability.
+
+In 2011, ISO isued an update for software product quality ISO/IEC 25010~\cite{iso:2011}.
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
  	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
@@
 (\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
-compliance thresholds for each severity level.
-\Sof{Adress thresholds confusion}
+compliance thresholds for each severity level. \Sof{@todo: Adress thresholds confusion}
 

### 2019-01-30 paper/ICPC19_Abstract/icpc19.tex
abstract

\documentclass[a4paper]{article} %
\usepackage{graphicx,amssymb} %

\textwidth=15cm \hoffset=-1.2cm %
\textheight=25cm \voffset=-2cm %

\date{\today}

\def\keywords#1{\begin{center}{\bf Keywords}\\{#1}\end{center}} %

\begin{document}

\title{Security vs. Maintainability: \\ Fixing Vulnerabilities Obfuscates your Code}

\author{Anonymou(s) Author(s)}

\maketitle

\thispagestyle{empty}

\begin{abstract}

Nowadays, security is an absolutely necessary requirement for the production of quality software. However, there is still a considerable amount of zero-day vulnerabilities published and fixed almost every week. The authors suspect that when developers address these vulnerabilities on their codebases, they actually affect the maintainability of software and even introduce more vulnerabilities. This paper evaluates the impact of refactorings to improve security on the maintainability of open-source software. Maintainability is measured based on the \emph{Better Code Hub}'s model of 10 guidelines for both before and after security commits for a dataset including 607 security commits. Results show that fixing software vulnerabilities harms the maintainability of open-source software. Specially for refactorings that deal with \emph{Broken Authentication and Session Management}, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks. Therefore, it is necessary to bring awareness for coding good practices and for the use of automated tools that have the potential to help developers with this issue.

\end{abstract}

\keywords{Security, Software Maintenance, Open-Source Software}

% \section{Introduction}

\end{document}

### 2019-01-30 paper/ICPC19/icpc19.tex
removing abstract folder, adding it to the paper

@@
 \begin{abstract}
-  Security is a crucial non-functionality requirement for software applications.
-  However, building secure software is far from trivial as developers lack both
-  the knowledge and tools to effectively address this concern. In this paper, we
-  study the impact of changes to improve security on the maintainability of several
-  open source applications. Using a dataset containing 607 security-oriented
-  commits, we measure maintainability --- as computed by the Software Improvement
-  Group's web-based source code analysis service \emph{Better Code Hub} (BCH) ---
-  before and after the security refactoring. Results show that making software
-  more secure comes at a cost on maintainability. This is particularly evident
-  in refactorings to deal with \emph{Broken Authentication and Session Management}, 
-  \emph{Denial-of-Service} attacks and \emph{Memory Leaks}.
+Nowadays, security is an absolutely necessary requirement for the production of quality software.
+However, there is still a considerable amount of zero-day vulnerabilities published and fixed almost every week.
+The authors suspect that when developers address these vulnerabilities on their codebases, they actually affect the maintainability of software and even introduce more vulnerabilities.
+This paper evaluates the impact of refactorings to improve security on the maintainability of open-source software.
+Maintainability is measured based on the \emph{Better Code Hub}'s model of 10 guidelines for both before and after security commits for a dataset including 607 security commits.
+Results show that fixing software vulnerabilities harms the maintainability of open-source software.
+Specially for refactorings that deal with \emph{Broken Authentication and Session Management}, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks.
+Therefore, it is necessary to bring awareness for coding good practices and for the use of automated tools that have the potential to help developers with this issue.
 \end{abstract}
@@
 
-This paper is structured as follows: section \ref{sec:motivation} introduces an
-example of a security refactoring of a known vulnerability found in
+% Our empirical study finds evidence that energy efficiency- oriented code changes have a negative impact on code main- tainability.
+In particular, careful thinking is required to im- plement the energy patterns Power Save Mode and Wakelock Addition.
+Furthermore, we show that energy patterns are more likely to require maintenance than regular code changes.
+
+This paper is structured as follows: section \ref{sec:motivation} introduces an example of a security refactoring of a known vulnerability found in
 OpenSSL\footnote{Repository available at
@@
 section \ref{sec:methodology} describes the methodology used to answer the
-research questions; section \ref{sec:results} presents the results and section
-\ref{sec:discussion} discusses their implications; section \ref{sec:threats},
-threats to the validity of the study are enumerated; section \ref{sec:rw}
-describres the different work and existing literature in the field of study;
-and, finally section \ref{sec:conclusions} concludes the main findings and
-elaborates on future work.
+research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.
 %
@@
 
-\begin{figure}[h]
+\begin{figure}[]
  	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}

### 2019-01-30 paper/ICPC19/icpc19.tex
abstract + title

@@
 
-\title{Security vs. Maintainability: \\ Fixing Vulnerabilities Obfuscates your Code}
+\title{Fixing Vulnerabilities Hinders Maintainability}
 
@@
 \begin{abstract}
-Nowadays, security is an absolutely necessary requirement for the production of quality software.
-However, there is still a considerable amount of zero-day vulnerabilities published and fixed almost every week.
-The authors suspect that when developers address these vulnerabilities on their codebases, they actually affect the maintainability of software and even introduce more vulnerabilities.
-This paper evaluates the impact of refactorings to improve security on the maintainability of open-source software.
-Maintainability is measured based on the \emph{Better Code Hub}'s model of 10 guidelines for both before and after security commits for a dataset including 607 security commits.
-Results show that fixing software vulnerabilities harms the maintainability of open-source software.
-Specially for refactorings that deal with \emph{Broken Authentication and Session Management}, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks.
-Therefore, it is necessary to bring awareness for coding good practices and for the use of automated tools that have the potential to help developers with this issue.
+Security is a requirement of utmost importance to produce software with high
+quality. However, there is still a considerable amount of zero-day
+vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
+developers address these vulnerabilities on their codebases, they actually
+affect the maintainability of software, potentially introducing other
+vulnerabilities. This paper evaluates the impact of refactorings to improve
+security on the maintainability of open-source software. Maintainability is
+measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
+dataset including $607$ security commits. Results show that fixing software
+vulnerabilities hinders the maintainability of open-source software. This is
+particularly the case for refactorings that address \emph{Broken Authentication
+and Session Management}, \emph{Memory Leaks} and \emph{Denial-of-Service}
+attacks. Therefore, it is necessary to bring awareness for coding practices and
+for the use of automated tools that have the potential to help developers with
+this issue.
 \end{abstract}

### 2019-01-30 paper/ICPC19/icpc19.tex
abstract fixed

@@
 particularly the case for refactorings that address \emph{Broken Authentication
-and Session Management}, \emph{Memory Leaks} and \emph{Denial-of-Service}
-attacks. Therefore, it is necessary to bring awareness for coding practices and
-for the use of automated tools that have the potential to help developers with
-this issue.
+and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
+attacks. Within the implications of our project, we highlight the need for tools 
+and documentation to help developers adopt security best practices.
 \end{abstract}
@@
 Unmaintainable code is hard to test, analyse and re-use.
+It is hard to find bugs and vulnerabilities in high complex code, this some developers then to forget maintainability.
+Therefore, we want to bring awareness for the fact that developers need to write maintainable code to prevent the introduction of new vulnerabilities in the code or other defects.
 
-It is hard to find bugs and vulnerabilities in high complex code, this some developers then to forget maintainability.
-Therefore, we want to bring awareness for the fact that developers need to write maintainable code to prevent the introduction of vulnerabilities in the code.
+Vulnerabilities in open-source software are similar to exploits found in closed products.
+These are portions of code that were developed with problems or features that attackers can take advantage to carry out harmful actions.
+In some cases, these issues are the advantage that hackers need to take services down or even get remote access to others systems.
+However, whereas closed software is written by developers who are following the same organization rules, open-source is less restrictive and more diverse since it depends on the community who writes, fixes, and generally maintains the codebase.
 
-The international standard ISO/IEC 25010:2011 breaks down software quality into eight characteristics: maintainability, functional suitability, performance efficiency, compatibility, usability, reliability, security, and portability.
-This paper focuses solely on maintainability.
+In 2011, the International Organization for Standardization (ISO) issued an update for software product quality ISO/IEC 25010~\cite{iso:2011} considering \emph{Security} as one of the main software product quality characteristics.
+In this work, we are concerned about the impact that fixing vulnerabilities may have in the maintainability of open-source software.
+Are developers affecting software maintainability when trying to improve software safety?
+We suspect that some of those fixes may have a negative impact in the software maintainability and possibly even be the cause of the introduction of new vulnerabilities.
 
-In 2011, ISO isued an update for software product quality ISO/IEC 25010~\cite{iso:2011}.
+ISO describes software maintainability as “the degree of effectiveness and efficiency with which a software product or system can be modified to improve it, correct it or adapt it to changes in environment, and in requirements” on software quality ISO/IEC 25010.
+ISO does not provide any specific guidelines or formula to calculate maintainability.
+Thus, we resort on Software Improvement Group’s web-based source code analysis service Better Code Hub (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given project.
+BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the base of BCH's evaluation.
+Two examples of guidelines used by BCH are the \emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small} based on the idea that smaller codebases are easier to maintain and less sensitive to defects~\cite{Visser:2016:OREILLY}.
 
-% Our empirical study finds evidence that energy efficiency- oriented code changes have a negative impact on code main- tainability.
-In particular, careful thinking is required to im- plement the energy patterns Power Save Mode and Wakelock Addition.
-Furthermore, we show that energy patterns are more likely to require maintenance than regular code changes.
+In this paper, we present the results of our analysis on the maintainability of 607 security changes collected from open-source software.
+The main contributions of the present work is an empirical study on the impact of security changes on software maintainability and what patterns need more attention.
+This empirical study exhibits proof that changes applied in the codebase to fix vulnerabilities affect code maintainability.
+Especially, for refactorings that address \emph{Broken Authentication and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks.
+With this study, we intend to highlight the need for tools and documentation to help developers
+adopt security best practices.
 
@@
 - unsigned long frag_len = msg_hdr->frag_len, max_len; /*!\annotation{lst:func3}!*/
-
 - if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
@@
 \ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message
-length in never higher than the maxiumum allowed by the DTSL protocol in
+length is never higher than the maxiumum allowed by the DTSL protocol in
 different functions.
@@
 
+\begin{table*}[h]
+	\centering
+	\caption{Descriptive statistics of the dataset projects}
+\begin{tabular}{@{}lllllllllll@{}}
+\toprule
+      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
+mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
+std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
+min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
+25\%  & 391.50  & 1581    & 117.25   & 49           & 1440.50  & 4        & 19       & 8466.75   & 313.75    & 143.25  \\
+median  & 838.50  & 2836.50 & 248      & 99           & 5504.50  & 9        & 59       & 37372.50  & 1792.50   & 650     \\
+75\%  & 2155    & 6828.50 & 459.50   & 261          & 18579.25 & 20       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
+max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
+Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
+\end{tabular}
+\label{tab:dataset}
+\end{table*}
+
 The dataset contemplates security flaws for more than 13 different languages
@@
 
-\begin{table*}[h]
-	\centering
-	\caption{Descriptive statistics of the dataset projects}
-\begin{tabular}{@{}lllllllllll@{}}
-\toprule
-      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
-mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
-std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
-min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
-25\%  & 391.50  & 1581    & 117.25   & 49           & 1440.50  & 4        & 19       & 8466.75   & 313.75    & 143.25  \\
-median  & 838.50  & 2836.50 & 248      & 99           & 5504.50  & 9        & 59       & 37372.50  & 1792.50   & 650     \\
-75\%  & 2155    & 6828.50 & 459.50   & 261          & 18579.25 & 20       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
-max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
-Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
-\end{tabular}
-\label{tab:dataset}
-\end{table*}
 %
@@
 
-\begin{table}[H]
+All guidelines are computed in the same evaluation by BCH. During each guideline
+evaluation, the tool determines the compliance towards one guideline by
+establishing limits for the percentage of code allowed to be in each of the $4$
+risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and
+\emph{very high risk}). If the project does not violate the thresholds, then it
+is compliant with the guideline. These thresholds are calibrated by BCH using
+their own data/experience --- open-source and closed software systems. If a
+project is compliant with a guideline, it means that it is at least $65\%$
+better than the software used by BCH to calculate the thresholds\footnote{Check
+the answer to \emph{How can I adjust the threshold for passing/not passing a
+guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.
+
+Figure \ref{fig:bchrep} shows an example of the  guidelines report
+provided by BCH for a project after finishing its evaluation. These are the
+results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring -
+motivation example discussed, previously, in section \ref{sec:motivation}. This
+version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write
+Clean Code} guideline.
+
+\begin{table}[h]
 	\caption{Guidelines to produce maintainable code}
@@
 
-All guidelines are computed in the same evaluation by BCH. During each guideline
-evaluation, the tool determines the compliance towards one guideline by
-establishing limits for the percentage of code allowed to be in each of the $4$
-risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and
-\emph{very high risk}). If the project does not violate the thresholds, then it
-is compliant with the guideline. These thresholds are calibrated by BCH using
-their own data/experience --- open-source and closed software systems. If a
-project is compliant with a guideline, it means that it is at least $65\%$
-better than the software used by BCH to calculate the thresholds\footnote{Check
-the answer to \emph{How can I adjust the threshold for passing/not passing a
-guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.
-Figure \ref{fig:bchrep} shows an example of the  guidelines report
-provided by BCH for a project after finishing its evaluation. These are the
-results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring -
-motivation example discussed, previously, in section \ref{sec:motivation}. This
-version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write
-Clean Code} guideline.
-
-\begin{figure}[]
+\begin{figure}[h]
  	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
  	\centering 	\includegraphics[width=0.49\textwidth]{figures/commit.pdf}

### 2019-01-30 paper/ICPC19/icpc19.tex
cleaning
intro - not finished

@@
 and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
-attacks. Within the implications of our project, we highlight the need for tools 
-and documentation to help developers adopt security best practices.
+attacks. The observations of this work lead us to conclude that security refactorings should be performed with extra care.
 \end{abstract}
@@
 
-Unmaintainable code is hard to test, analyse and re-use.
-It is hard to find bugs and vulnerabilities in high complex code, this some developers then to forget maintainability.
-Therefore, we want to bring awareness for the fact that developers need to write maintainable code to prevent the introduction of new vulnerabilities in the code or other defects.
+Security is a requirement of utmost importance to produce software with high
+quality. However, there is still a considerable amount of zero-day
+vulnerabilities being discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative\footnote{Zero Day Initiative website available  at https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})} website.
+These vulnerabilities are gateways for attackers to penetrate systems and exploit their resources.
 
-Vulnerabilities in open-source software are similar to exploits found in closed products.
-These are portions of code that were developed with problems or features that attackers can take advantage to carry out harmful actions.
-In some cases, these issues are the advantage that hackers need to take services down or even get remote access to others systems.
-However, whereas closed software is written by developers who are following the same organization rules, open-source is less restrictive and more diverse since it depends on the community who writes, fixes, and generally maintains the codebase.
+% Maintainability importance is being understimated.
+
+% The time required to implement an additional security functionality in an application is not always easy to estimate.
+Estimating whether such functionality requires any changes in the architecture of the system is even harder.
+For instance, if you think about refining the access control policies of your application, by moving from a function-level access control to a data-level access control (where access control is performed on the data field instead on the system functionality executed by the user), can you precisely estimate the amount of time required to perform such change or even if that change has an impact on the architecture of the system?
+
+% Having a resilient and scalable security architecture and design is crucial to prevent vulnerabilities.
+
+% High complex code is hard to test, thus being more difficult to find bugs and vulnerabilities. Therefore, writing maitainable code prevent the introduction of vulnerabilities in the code.
 
 In 2011, the International Organization for Standardization (ISO) issued an update for software product quality ISO/IEC 25010~\cite{iso:2011} considering \emph{Security} as one of the main software product quality characteristics.
+Unmaintainable code is hard to test, analyse and re-use, and thus more difficult to find vulnerabilities and defects.
+Therefore, writing maitainable code prevents the introduction of vulnerabilities in the code.
 In this work, we are concerned about the impact that fixing vulnerabilities may have in the maintainability of open-source software.
@@
 The maintainability formula was inferred based on the reports of the Better Code Hub.
-Since the dataset has a broad group of open-source projects with different backgrounds, this metric may not be suitable for all of them. However, BCH does have a representative benchmark of
-closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012} (Table \ref{tab:guidelines}).
+Since the dataset has a broad group of open-source projects with different backgrounds, this metric may not be suitable for all of them.
+\Sof{Improve it} However, BCH does have a representative benchmark of
+closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
 
@@
 
-All the security commits are described in English.
-%
+All the security commits are described in English. This approach discards security refactorings provided in other languages.
+
 \section{Related Work}\label{sec:rw}

### 2019-01-30 paper/ICPC19/icpc19.tex
main contributions

@@
 In this paper, we present the results of our analysis on the maintainability of 607 security changes collected from open-source software.
-The main contributions of the present work is an empirical study on the impact of security changes on software maintainability and what patterns need more attention.
+
+The main contributions of the present work is:
+\begin{itemize}
+	\item An empirical study on the impact of security changes on software maintainability and what patterns need more attention.
+	\item A software package with all scripts and data created to perform the empirical evaluation, for reproducibility. Available here: $<url>$.
+
+\end{itemize}
+
 This empirical study exhibits proof that changes applied in the codebase to fix vulnerabilities affect code maintainability.
@@
 https://github.com/openssl/openssl (Accessed on \today{})} software;
-section \ref{sec:methodology} describes the methodology used to answer the
-research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.
+section \ref{sec:methodology} describes the methodology used to answer the research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.
 %

### 2019-01-30 paper/ICPC19/icpc19.tex
bumping 1st paragraph

--- 2019-01-30 paper/ICPC19/icpc19.tex
@@
 and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
-attacks. The observations of this work lead us to conclude that security refactorings should be performed with extra care.
+attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
+need to be performed with extra care.
 \end{abstract}
@@
 
-Security is a requirement of utmost importance to produce software with high
-quality. However, there is still a considerable amount of zero-day
-vulnerabilities being discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative\footnote{Zero Day Initiative website available  at https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})} website.
-These vulnerabilities are gateways for attackers to penetrate systems and exploit their resources.
+Software quality is important because it is ultimately related to the overall
+cost for developing, extending and maintaing software applications. Software
+quality characteristics include, but are not limited to, functional correctness,
+reliability, usability, maintainability, and security. Security is an important
+non-functional requirement during the development of software systems.  However,
+there is still a considerable amount of zero-day vulnerabilities being
+discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
+website\footnote{Zero Day Initiative website available  at
+https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}.
+As an example, in the last week of January 2019, a severe vulnerability in Apple
+FaceTime was
+revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
+(Accessed on \today{})}. These vulnerabilities are gateways for attackers to
+penetrate systems and exploit their resources.
 

### 2019-01-30 paper/ICPC19/icpc19.tex
discussion
more things on intro

@@
 
-In 2011, the International Organization for Standardization (ISO) issued an update for software product quality ISO/IEC 25010~\cite{iso:2011} considering \emph{Security} as one of the main software product quality characteristics.
-Unmaintainable code is hard to test, analyse and re-use, and thus more difficult to find vulnerabilities and defects.
-Therefore, writing maitainable code prevents the introduction of vulnerabilities in the code.
-In this work, we are concerned about the impact that fixing vulnerabilities may have in the maintainability of open-source software.
-Are developers affecting software maintainability when trying to improve software safety?
-We suspect that some of those fixes may have a negative impact in the software maintainability and possibly even be the cause of the introduction of new vulnerabilities.
+In 2011, the International Organization for Standardization (ISO) issued an
+update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
+\emph{Security} as one of the main software product quality characteristics.
+Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
+to find vulnerabilities and defects. Therefore, writing maitainable code
+prevents the introduction of vulnerabilities in the code. In this work, we are
+concerned about the impact that fixing vulnerabilities may have in the
+maintainability of open-source software. Are developers affecting software
+maintainability when trying to improve software safety? We suspect that some of
+those fixes may have a negative impact in the software maintainability and
+possibly even be the cause of the introduction of new vulnerabilities.
 
-ISO describes software maintainability as “the degree of effectiveness and efficiency with which a software product or system can be modified to improve it, correct it or adapt it to changes in environment, and in requirements” on software quality ISO/IEC 25010.
-ISO does not provide any specific guidelines or formula to calculate maintainability.
-Thus, we resort on Software Improvement Group’s web-based source code analysis service Better Code Hub (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given project.
-BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the base of BCH's evaluation.
-Two examples of guidelines used by BCH are the \emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small} based on the idea that smaller codebases are easier to maintain and less sensitive to defects~\cite{Visser:2016:OREILLY}.
+ISO describes software maintainability as “the degree of effectiveness and
+efficiency with which a software product or system can be modified to improve
+it, correct it or adapt it to changes in environment, and in requirements” on
+software quality ISO/IEC 25010. ISO does not provide any specific guidelines or
+formula to calculate maintainability. Thus, we resort on Software Improvement
+Group’s web-based source code analysis service Better Code Hub
+(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
+project. BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
+base of BCH's evaluation. Two examples of guidelines used by BCH are the
+\emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
+and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
+easier to maintain and less sensitive to defects~\cite{Visser:2016:OREILLY}.
 
-In this paper, we present the results of our analysis on the maintainability of 607 security changes collected from open-source software.
+The particular subject of this paper is to explore whether there is a trade-off
+between refactoring source code to fix identified vulnerabilities while keeping
+(or even improving) the maintainability of the software system. In other words,
+we are interested in answering the following overarching question
+
+\begin{framed}
+\noindent
+Does improving security have a negative impact on code
+maintainability?
+\end{framed}
+
+\noindent
+In this paper, we present the results of our analysis on the maintainability of
+607 security changes collected from open-source software.
 
@@
 \begin{itemize}
-	\item An empirical study on the impact of security changes on software maintainability and what patterns need more attention.
-	\item A software package with all scripts and data created to perform the empirical evaluation, for reproducibility. Available here: $<url>$.
+	\item An empirical study on the impact of security changes on software
+	maintainability and what patterns need more attention.
+
+  \item An analysis of how likely energy changes need to be
+  maintained, in comparison to regular commits.
+
+	\item A replication package with all scripts and data created to perform the
+	empirical evaluation, for reproducibility. Available here: $<url>$.
 
@@
 
-This empirical study exhibits proof that changes applied in the codebase to fix vulnerabilities affect code maintainability.
-Especially, for refactorings that address \emph{Broken Authentication and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks.
-With this study, we intend to highlight the need for tools and documentation to help developers
-adopt security best practices.
+This empirical study exhibits proof that changes applied in the codebase to fix
+vulnerabilities affect code maintainability. Especially, for refactorings that
+address \emph{Broken Authentication and Session Management} issues, \emph{Memory
+Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to
+highlight the need for tools and documentation to help developers adopt security
+best practices.
 
-This paper is structured as follows: section \ref{sec:motivation} introduces an example of a security refactoring of a known vulnerability found in
-OpenSSL\footnote{Repository available at
-https://github.com/openssl/openssl (Accessed on \today{})} software;
-section \ref{sec:methodology} describes the methodology used to answer the research questions; section \ref{sec:results} presents the results and section \ref{sec:discussion} discusses their implications; section \ref{sec:threats}, threats to the validity of the study are enumerated; section \ref{sec:rw} describres the different work and existing literature in the field of study; and, finally section \ref{sec:conclusions} concludes the main findings and elaborates on future work.
+This paper is structured as follows: section \ref{sec:motivation} introduces an
+example of a security refactoring of a known vulnerability found in
+OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
+(Accessed on \today{})} software; section \ref{sec:methodology} describes the
+methodology used to answer the research questions; section \ref{sec:results}
+presents the results and section \ref{sec:discussion} discusses their
+implications; section \ref{sec:threats}, threats to the validity of the study
+are enumerated; section \ref{sec:rw} describres the different work and existing
+literature in the field of study; and, finally section \ref{sec:conclusions}
+concludes the main findings and elaborates on future work.
 %

### 2019-01-30 paper/ICPC19/icpc19.tex
fix
adding more to intro

@@
 
-ISO describes software maintainability as “the degree of effectiveness and
+ISO describes software maintainability as ``the degree of effectiveness and
 efficiency with which a software product or system can be modified to improve
-it, correct it or adapt it to changes in environment, and in requirements” on
-software quality ISO/IEC 25010. ISO does not provide any specific guidelines or
-formula to calculate maintainability. Thus, we resort on Software Improvement
-Group’s web-based source code analysis service Better Code Hub
+it, correct it or adapt it to changes in environment, and in requirements'' on
+software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
+formula to calculate maintainability, we resort on Software Improvement
+Group (SIG)’s web-based source code analysis service Better Code Hub
 (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
@@
 easier to maintain and less sensitive to defects~\cite{Visser:2016:OREILLY}.
+Compliance to the guidelines is derived from the SIG's
+industry benchmark which, according to official sources, consists of over 8
+billion lines of code in more than 180 different technologies. SIG analyzes
+around 15 million lines of code, per week.
 
@@
 
-  \item An analysis of how likely energy changes need to be
+  \item An analysis of how likely security-related changes need to be
   maintained, in comparison to regular commits.

### 2019-01-30 paper/ICPC19/icpc19.tex
more info

@@
 In this paper, we present the results of our analysis on the maintainability of
-607 security changes collected from open-source software.
+607 security changes collected from open-source software, from different
+programming languages.
 

### 2019-01-31 paper/ICPC19/icpc19.tex
small advances

--- 2019-01-30 paper/ICPC19/icpc19.tex
@@
 revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
-(Accessed on \today{})}. These vulnerabilities are gateways for attackers to
+(Accessed on \today{})}. Callers could see and hear others without them picking up the call. These vulnerabilities are gateways for attackers to
 penetrate systems and exploit their resources.
-
-% Maintainability importance is being understimated.
-
-% The time required to implement an additional security functionality in an application is not always easy to estimate.
-Estimating whether such functionality requires any changes in the architecture of the system is even harder.
-For instance, if you think about refining the access control policies of your application, by moving from a function-level access control to a data-level access control (where access control is performed on the data field instead on the system functionality executed by the user), can you precisely estimate the amount of time required to perform such change or even if that change has an impact on the architecture of the system?
-
-% Having a resilient and scalable security architecture and design is crucial to prevent vulnerabilities.
-
-% High complex code is hard to test, thus being more difficult to find bugs and vulnerabilities. Therefore, writing maitainable code prevent the introduction of vulnerabilities in the code.
 
@@
 
+% find work about the struggle to make systems secure
+
+We believe developers should be able to create secure code without affecting 
+the maintainability of their projects.
+
+\begin{itemize}
+	\item \textbf{Security Frameworks.} The most popular languagues (e.g., Java and c\#) 
+used in the development of enterprise applications provide stable
+frameworks to implement the security requirements and easily prevent some of the
+OWASP TOP 10 vulnerabilities.
+	\item \textbf{Documentation}
+	\item \textbf{Programming Languages}
+	\item \textbf{Proper authorization mechanisms}
+	\item \textbf{Remove dead and duplicated code}
+	\item \textbf{Configurations in files}
+	\item \textbf{Framework manages the session}
+
+\end{itemize}
+
+% #1 Security frameworks are your new best friend
+% The most popular software stacks (e.g.: Java and C#) used in the development of enterprise applications provide very good and stable frameworks to implement the security requirements and easily prevent some of the OWASP TOP 10 vulnerabilities.
+For instance, through a very simple annotation such as @EnableWebSecurity, Spring Security automatically and out of the box, implements mitigations against some of the most common web application vulnerabilities (e.g.: CSRF, XSS, etc.).
+The essential HTTP headers to prevent certain types of vulnerabilities are added to the requests and validated in the responses (e.g.: same-origin policy for the synchronization token to prevent CSRF).
+%
+% Although these frameworks may require some time to understand and use (the learning curve might be quite steep, sometimes), they have been improving quite a lot in the past few years, facilitating the implementation and maintenance of the security requirements.
+%
+% Nevertheless, do not blindly trust all the defaults of the frameworks and libraries your application depends on.
+More than often frameworks rely on weak default encryption algorithms such as DES (encryption standard deprecated since 1998).
+Thus, always check all the frameworks and libraries defaults, before releasing the application.
+%
+% Besides, some of the least maintained frameworks contain vulnerabilities that never get to be reported.
+This is because some bugs (this is an interesting discussion about a systemd null pointer dereferencing bug) are not even considered security vulnerabilities in the first place.
+Notice that categorisation of bugs as security vulnerabilities, is based on the ability of the development team to evaluate the impact of that bug in the security of the application.
+That’s why updating the libraries and frameworks of your application to their latest version helps improving its overall security.
+So, stay up-to-date.
+%
+% #2 Deny first, authorize after
+% Broken access control is one of the OWASP TOP 10 vulnerabilities.
+This is because most applications fail to implement proper authorization mechanisms.
+Authorization design and implementation is probably one of the most complex security features that exists.
+Making it right and scalable is challenging and requires the effort of all the team members.
+Nevertheless, adoption of certain mechanisms and design strategies may facilitate authorization maintenance in the long run:
+%
+% 1.
+Authorise only afterwards: access must always be denied in the first place, i.e., access must only be granted if the user is actually authorized to perform a certain action.
+So, access control policies must be checked before processing the request and not when producing the response.
+%
+% 2.
+Use and abuse of filters: HTTP filters can be used to filter out all the requests before processing them (e.g.: if the user doing the request is authorized to perform the designated action, then the request can be processed).
+Certain frameworks (e.g.: Spring Security) even provide a very elegant way of binding the access control policies to certain system functions by means of code annotations in the methods that expose the desired functionalities.
+In combination with the HTTP filters these annotations provide a very clean and easy to change implementation of the access control mechanisms.
+%
+% 3.
+Always document what you’re doing: either using role-based access control or a more fine-grained access control model, the functionalities enabled in each roles must be clearly defined and documented.
+Although, the code should speak for itself, the lack of documentation brings risks for code portability and transferability, specially when considering access control management.
+%
+% #3 Make the IT infrastructure your number one software project
+% A securely developed application is defenseless against an attacker with administrative privileges in the production environment.
+For instance, the WannaCry ransomware attack was based on an OS level exploit that was fairly easy to protect, just by upgrading the OS version.
+Nonetheless, it was very problematic because a lot of IT departments tend to disregard the importance and impact of OS security updates.
+%
+% In general, the longer you wait to upgrade, the harder it gets and eventually a system can get ‘un-upgradable’, therefore, unmaintainable.
+% When deploying the application, it is necessary to make sure that all 3rd party components that are part of the production environment can be upgraded or recreated from backup (in case of ransomware) at any point in time.
+This includes databases and web servers, deployment frameworks, the identity manager, log manager application and alerting system, the operating system, etc.
+%
+% The ‘upgrading is hard’ problem is similar to the ‘releasing is hard’ problem and should be solved the same way: by doing it often.
+% The way to do this is to treat the environment in which your application runs as part of the application and not something separate. As such it should be treated as a software project in itself.
+%
+% #4 Validations should be done in the backend
+% More and more web applications follow a service oriented architecture model, where the backend provides a RESTful API and the front-end is purely implemented in JavaScript (with HTML).
+Some modern JavaScript frameworks provide out of the box escaping and sanitization functionalities (e.g.: Angular 2 or above).
+Some developers are led to mistakenly believe that this free-feature, together with type checking the REST parameters (done in the corresponding controller, for instance) is enough to prevent introduction of malicious input in the application.
+But unfortunately, it is not.
+As much as the web and security frameworks try to enable security mitigations by default, a lot of effort is still needed from the development team.
+Therefore, it’s very important to validate all the user input in the backend and filter-out unwanted/invalid inputs.
+%management of the web application.
+%
+% Notice that the fact that in modern frameworks crypto primitives can be defined in a configuration file, makes it much easier and less time consuming, to, for instance, update those primitives when a new attack is discovered.
+Plus, if the default session manager has a known vulnerability, it is easier to patch the application by updating the library, then doing quick (and perhaps dirty) fixes in custom implementations, that can lead to even more exploits.
+%
+
 \begin{framed}
@@
 
+\subsection{Security Maintainability Models}
+
+~\cite{6616351,oswap:2009,common:2009}
+
 \section{Conclusion and Future Work}\label{sec:conclusions}

### 2019-01-31 paper/ICPC19/icpc19.tex
language vs maint chart

@@
 
-  \item An analysis of how likely security-related changes need to be
-  maintained, in comparison to regular commits.
+  % \item An analysis of how likely security-related changes need to be
+  % maintained, in comparison to regular commits.
 
@@
 
-\Sof{@todo: effect size}
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.55\textwidth]{figures/language.pdf}
+ 	\caption{Maintainability difference by language after security refactorings}
+	\label{fig:lang}
+\end{figure}
 
@@
  	\includegraphics[width=0.55\textwidth]{figures/category.pdf}
- 	\caption{Maintainability difference by type after security refactorings}
+ 	\caption{Maintainability difference by pattern after security refactorings}
 	\label{fig:pat}
@@
 
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
 \begin{framed}
@@
 \textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}}
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
 
@@
 
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
 \section{Conclusion and Future Work}\label{sec:conclusions}
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
 

### 2019-01-31 paper/ICPC19/icpc19.tex
more results

@@
 
+Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
+It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+
 In 2011, the International Organization for Standardization (ISO) issued an
@@
 
-In total, this study evaluates $607$ security and baseline commits from 94 different open-source projects distributed among 5 different main types of software, Figure \ref{fig:domain}.
-The following section presents the results obtained for each research question.
+In total, this study evaluates $607$ security and baseline commits from 94 
+different open-source projects distributed among 5 different main types of 
+software, Figure \ref{fig:domain}. The following section presents the results 
+obtained for each research question.
 
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability of open-source software?}
+\textit{\textbf{RQ1} What is the impact of security refactorings on the 
+maintainability of open-source software?}
 \end{framed}
 
-The impact of security and baseline commits in the maintainability of open-source software is presented in Figure \ref{fig:secvsreg}.
-For each type of change, three types of impact of refactorings in the software maintainability are presented: \emph{negative}, maintainability decreases (\emph{red}); \emph{none}, maintainability stays the same (\emph{yellow}); and, \emph{positive}, maintainability increases (\emph{green}).
-In front of each group, it is presented the mean ($\overline{x}$) and median ({$\widetilde{x}$) of the maintainability difference and the p-value resulting from the Wilcoxon signed-rank test.
+The impact of security and baseline commits in the maintainability of 
+open-source software is presented in Figure \ref{fig:secvsreg}. For each type 
+of change, three types of impact of refactorings in the software maintainability 
+are presented: \emph{negative}, maintainability decreases (\emph{red}); \emph{none}, 
+maintainability stays the same (\emph{yellow}); and, \emph{positive}, maintainability 
+increases (\emph{green}). In front of each group, it is presented the mean ($\overline{x}$) 
+and median ({$\widetilde{x}$) of the maintainability difference and the p-value 
+resulting from the Wilcoxon signed-rank test.
 
@@
 
-Regarding security commits results, we observe that maintainability decreases in 40.20\% (244), increases in 29.33\% (178) and stays the same in 30.5\% (185) of the cases after security refactorings being applied.
-The resulting p-value of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$.
-Since the p-value is below the significance level of $0.05$, it is fair to state that \emph{security refactorings have a negative impact in the maintainability of open-source software}.
-
-For regular commits, we also observe that maintainability decreases in 41.35\% (251) of the cases after regular refactorings being applyed.
-But in contrast to security commits, it increases in 58.16\% (353) and stays the same in 0.5\% (3).
-The resulting p-value of the Wilcoxon signed-rank test is $1.54$x$10^{-6}$.
-Since the p-value is below the significance level of $0.05$, it is fair to state that \emph{regular refactorings have a positive impact in the maintainability of open-source software}.
+Regarding security commits results, we observe that maintainability decreases 
+in 40.20\% (244), stays equal in 30.5\% (185) and increases in 29.33\% (178)  
+of the cases after security refactorings being applied. The resulting p-value 
+of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$. Since the p-value is below 
+the significance level of $0.05$, it is reasonable to affirm that \emph{security 
+refactorings have a negative impact in the maintainability of open-source software}.
+
+For regular commits, we also observe that maintainability decreases in 41.35\% 
+(251) of the cases after regular refactorings being applyed. But in contrast 
+to security commits, it increases in 58.16\% (353) and stays equal in 0.5\% 
+(3) of the cases. The resulting p-value of the Wilcoxon signed-rank test is 
+$1.54$x$10^{-6}$. Since the p-value is below the significance level of $0.05$, 
+it is reasonable to affirm that \emph{regular refactorings have a positive 
+impact in the maintainability of open-source software}.
 
@@
 \end{figure}
+
+In this study, we also evaluated the impact of each programming language on the 
+maintainability of open-source software, Figure \ref{fig:lang}. Although the
+dataset contains security refactorings for more than $13$ languages, we only
+present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP} because
+one of the requirements of the Wilcoxon signed-rank test is that the input sample
+has to have more than $20$ measurements. The other languages were integrated in the
+\emph{Others} set. The results of the Wilcoxon signed-rank test by pattern returned 
+statistical evidence which supports that the maintainability of open-source software 
+decreases for security refactorings in 3 different languages: \emph{C} 
+(p = $5.72$x$10^{-9}$), Ruby (p = 0.057) and PHP (p = $3.54$x$10^{-6}$). The results 
+also provide statistical significance for when security refactorings are performed
+in \emph{Java} (p = $3.61$x$10^{-2}$). \emph{Python} and the other languages 
+that integrate the \emph{Others} set contain more cases where the maintainability
+increases. However, the results obtained for these patterns are not statistically 
+significant.
 

### 2019-02-01 paper/ICPC19/icpc19.tex
discussion up

@@
 
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
+\Sof{Models} Lorem Ipsum is simply dummy text of the printing and typesetting industry.
 Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
 It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions.
 
@@
 \midrule
-\textbf{Write Short Units of Code} & Limit code units to 15 LOCs because smaller units are easier to understand, reuse and test them\\\midrule
-\textbf{Write Simple Units of Code} & Limit branch points to 4 per unit because it makes units easier to test and modify \\\midrule
-\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at multiple places (inefficient and error-prone)\\\midrule
-\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most 4 because it makes units easier to understand and reuse\\\midrule
-\textbf{Separate Concerns in Modules} & Avoid large modules because changes in loosely coupled databases are easier to oversee and execute\\\midrule
-\textbf{Couple Architecture Components Loosely} & Minimize the amount of code within modules that is exposed to modules in other components\\\midrule
-\textbf{Keep Architecture Components Balanced} & Balancing the number of components ease locating code and allow for isolated maintenance\\\midrule
-\textbf{Keep your Codebase Small} & Reduce and avoid the system size because small products are easier to manage and maintain\\\midrule
-\textbf{Automate Tests} & Test your codebase because it makes development predictable and less risky\\\midrule
-\textbf{Write Clean Code} & Avoid producing software with code smells because it is more likely to be maintainable in the future\\
+\textbf{Write Short Units of Code} & Limit code units to 15 LOCs because smaller
+ units are easier to understand, reuse and test them\\\midrule
+\textbf{Write Simple Units of Code} & Limit branch points to 4 per unit because 
+it makes units easier to test and modify \\\midrule
+\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at 
+multiple places (inefficient and error-prone)\\\midrule
+\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most 
+4 because it makes units easier to understand and reuse\\\midrule
+\textbf{Separate Concerns in Modules} & Avoid large modules because changes in 
+loosely coupled databases are easier to oversee and execute\\\midrule
+\textbf{Couple Architecture Components Loosely} & Minimize the amount of code 
+within modules that is exposed to modules in other components\\\midrule
+\textbf{Keep Architecture Components Balanced} & Balancing the number of 
+components ease locating code and allow for isolated maintenance\\\midrule
+\textbf{Keep your Codebase Small} & Reduce and avoid the system size because 
+small products are easier to manage and maintain\\\midrule
+\textbf{Automate Tests} & Test your codebase because it makes development 
+predictable and less risky\\\midrule
+\textbf{Write Clean Code} & Avoid producing software with code smells because 
+it is more likely to be maintainable in the future\\
 \bottomrule
@@
 guideline in the example since the bars do not reach the threshold points. This
-example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
+example only complies with $\frac{1}{10}$ guidelines (\emph{Write Clean Code}).}
 	\label{fig:bchrep}
@@
 reduce maintainability.
-
 %
@@
  	\caption{Maintainability difference by language after security refactorings}
-	\label{fig:lang}
+	\label{fig:lang_main}
 \end{figure}
@@
 In this study, we also evaluated the impact of each programming language on the 
-maintainability of open-source software, Figure \ref{fig:lang}. Although the
+maintainability of open-source software, Figure \ref{fig:lang_main}. Although the
 dataset contains security refactorings for more than $13$ languages, we only
@@
 \begin{framed}
-\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to affect open-source software maintainability?}
+\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to 
+affect open-source software maintainability?}
 \end{framed}
 
-The impact of each type of security refactoring on the maintainability of open-source software is presented in Figure \ref{fig:pat}.
-The results of the Wilcoxon signed-rank test by pattern returned statistical evidence which supports that the maintainability of open-source software decreases for 4 different patterns: Broken Autentication (p = 0.018), Denial-of-Service (p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
-The results also returned statistical significance that evidence that maintainability is more likely to not suffer differences for open-source software after removing Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site Scripting Vulnerabilities (p = 0.001).
-Injection and Cross-Site Request Forgery produced more cases where the refactorings had a negative impact in the maintainability.
+The impact of each type of security refactoring on the maintainability of 
+open-source software is presented in Figure \ref{fig:pat}. The results of the 
+Wilcoxon signed-rank test by pattern returned statistical evidence which 
+supports that the maintainability of open-source software decreases for 4 
+different patterns: Broken Autentication (p = 0.018), Denial-of-Service 
+(p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$). 
+The results also returned statistical significance that evidence that maintainability
+is more likely to not suffer differences for open-source software after removing 
+Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site 
+Scripting Vulnerabilities (p = 0.001). Injection and Cross-Site Request Forgery 
+produced more cases where the refactorings had a negative impact in the maintainability. 
 However, the results obtained for these patterns are not statistically significant.
 
-Projects with Broken Authentication flaws are the most affected ones by security refactorings.
-55.3\% (21) of the refactorings had a negative impact in the maintainability while 35\% (11) had a positive impact and only 13.8\% (6) did not suffer any change.
-The results also show that fixing memory leaks affects the maintainability of 51.3\% (41) of the cases and leaves only 13.8\% (11) of the cases unaffected.
-35\% (28) are affected positively.
-Refactoring vulnerabilities that allow denial-of-service attacks (e.g., CVE-2014-3506) has a negative impact of 47.5\% (19) in the maintainability of open-source software.
-30.0\% (12) of the cases exhibit maintainability improvement while 22.5\% (9) show that overall maintainability does not suffer any change.
-Miscellaneous refactorings pattern integrates several types of security flaws whose impact is of 43.1\% (84) negative in the maintainability of open-source sofware.
-31.8\% (62) of the cases do not exhibit any change in maintainability and 25.1\% (49) of cases have a positive impact on the maintainability of open-source software.
+Projects with Broken Authentication flaws are the most affected ones by security 
+refactorings. 55.3\% (21) of the refactorings had a negative impact in the maintainability 
+while 35\% (11) had a positive impact and only 13.8\% (6) did not suffer any 
+change. The results also show that fixing memory leaks affects the maintainability
+of 51.3\% (41) of the cases and leaves only 13.8\% (11) of the cases unaffected. 
+In 35\% (28) of the cases, maintainability increased. Refactoring vulnerabilities 
+that allow denial-of-service attacks (e.g., CVE-2014-3506) has a negative impact 
+of 47.5\% (19) in the maintainability of open-source software. 30.0\% (12) of 
+the cases exhibit maintainability improvement while 22.5\% (9) show that overall 
+maintainability does not suffer any change. Miscellaneous refactorings pattern 
+integrates several types of security flaws whose impact is of 43.1\% (84) negative 
+in the maintainability of open-source sofware. 31.8\% (62) of the cases do not 
+exhibit any change in maintainability and 25.1\% (49) of cases have a positive 
+impact on the maintainability of open-source software.
 
-It is also important to notice that overall the maintainability of open-source software is not affected by Cross-Site Scripting vulnerabilities and by Using Components with Know Vulnerabilities.
-Cross-Site Scripting does not suffer any decrease in the maintainability of 55.3\% (63) of cases and 21.1\% (24) have a positive effect on it.
-Only 23.7\% (27) of the cases hinder the software maintainability.
-From 21 cases evaluated for Using Componentes with Known Vulnerabilities, only 4.8\% (1) has a positive effect in the maintainability of software.
-14.3\% (3) harms maintainability software and 81\% has no effect in the maintainability.
+It is also important to notice that overall the maintainability of open-source 
+software is not affected by Cross-Site Scripting vulnerabilities and by Using 
+Components with Know Vulnerabilities. Cross-Site Scripting does not suffer any 
+decrease in the maintainability of 55.3\% (63) of cases and 21.1\% (24) have a 
+positive effect on it. Only 23.7\% (27) of the cases hinder the software 
+maintainability. From 21 cases evaluated for Using Componentes with Known 
+Vulnerabilities, only 4.8\% (1) has a positive effect in the maintainability 
+of software. 14.3\% (3) harms maintainability software and 81\% has no effect 
+in the maintainability.
 
@@
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability of open-source software?}
+\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability 
+of open-source software?}
 \end{framed}
@@
 
-40.2\% of the security refactoring have a negative impact on the maintainability of open-source software, Figure \ref{fig:secvsreg}.
-
-% find work about the struggle to make systems secure
-
-We believe developers should be able to create secure code without affecting 
-the maintainability of their projects.
-
+40.2\% of the security refactoring have a negative impact on the maintainability 
+of open-source software, Figure \ref{fig:secvsreg}. This raises a new tradeoff when
+developers need to fix vulnerabilities on their projects because they might be affecting 
+the software maintainability or even introducing new vulnerabilities. This study exhibits
+proof that developers may have to reduce maintainability for the sake of safety. We understand 
+that developers should be able to produce secure code without affecting the maintainability of 
+their projects. But there are still some concerns that need to be addressed: 
 \begin{itemize}
-	\item \textbf{Security Frameworks.} The most popular languagues (e.g., Java and c\#) 
-used in the development of enterprise applications provide stable
-frameworks to implement the security requirements and easily prevent some of the
-OWASP TOP 10 vulnerabilities.
-	\item \textbf{Documentation}
-	\item \textbf{Programming Languages}
-	\item \textbf{Proper authorization mechanisms}
-	\item \textbf{Remove dead and duplicated code}
-	\item \textbf{Configurations in files}
-	\item \textbf{Framework manages the session}
+	\item \textbf{Security Frameworks} need to be secure. Unfortunately, there are several 
+	vulnerabilities that stay undisclosed. Thus, developers should not trust them blindly and 
+	check defaults.
+	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
+	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
+	one of the most complex security features to implement.
+Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
+	\item \textbf{Deprecated code} should be avoided by developers.
+They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
+	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.
 
@@
 
-% #1 Security frameworks are your new best friend
-% The most popular software stacks (e.g.: Java and C#) used in the development of enterprise applications provide very good and stable frameworks to implement the security requirements and easily prevent some of the OWASP TOP 10 vulnerabilities.
-For instance, through a very simple annotation such as @EnableWebSecurity, Spring Security automatically and out of the box, implements mitigations against some of the most common web application vulnerabilities (e.g.: CSRF, XSS, etc.).
-The essential HTTP headers to prevent certain types of vulnerabilities are added to the requests and validated in the responses (e.g.: same-origin policy for the synchronization token to prevent CSRF).
-%
-% Although these frameworks may require some time to understand and use (the learning curve might be quite steep, sometimes), they have been improving quite a lot in the past few years, facilitating the implementation and maintenance of the security requirements.
-%
-% Nevertheless, do not blindly trust all the defaults of the frameworks and libraries your application depends on.
-More than often frameworks rely on weak default encryption algorithms such as DES (encryption standard deprecated since 1998).
-Thus, always check all the frameworks and libraries defaults, before releasing the application.
-%
-% Besides, some of the least maintained frameworks contain vulnerabilities that never get to be reported.
-This is because some bugs (this is an interesting discussion about a systemd null pointer dereferencing bug) are not even considered security vulnerabilities in the first place.
-Notice that categorisation of bugs as security vulnerabilities, is based on the ability of the development team to evaluate the impact of that bug in the security of the application.
-That’s why updating the libraries and frameworks of your application to their latest version helps improving its overall security.
-So, stay up-to-date.
-%
-% #2 Deny first, authorize after
-% Broken access control is one of the OWASP TOP 10 vulnerabilities.
-This is because most applications fail to implement proper authorization mechanisms.
-Authorization design and implementation is probably one of the most complex security features that exists.
-Making it right and scalable is challenging and requires the effort of all the team members.
-Nevertheless, adoption of certain mechanisms and design strategies may facilitate authorization maintenance in the long run:
-%
-% 1.
-Authorise only afterwards: access must always be denied in the first place, i.e., access must only be granted if the user is actually authorized to perform a certain action.
-So, access control policies must be checked before processing the request and not when producing the response.
-%
-% 2.
-Use and abuse of filters: HTTP filters can be used to filter out all the requests before processing them (e.g.: if the user doing the request is authorized to perform the designated action, then the request can be processed).
-Certain frameworks (e.g.: Spring Security) even provide a very elegant way of binding the access control policies to certain system functions by means of code annotations in the methods that expose the desired functionalities.
-In combination with the HTTP filters these annotations provide a very clean and easy to change implementation of the access control mechanisms.
-%
-% 3.
-Always document what you’re doing: either using role-based access control or a more fine-grained access control model, the functionalities enabled in each roles must be clearly defined and documented.
-Although, the code should speak for itself, the lack of documentation brings risks for code portability and transferability, specially when considering access control management.
-%
-% #3 Make the IT infrastructure your number one software project
-% A securely developed application is defenseless against an attacker with administrative privileges in the production environment.
-For instance, the WannaCry ransomware attack was based on an OS level exploit that was fairly easy to protect, just by upgrading the OS version.
-Nonetheless, it was very problematic because a lot of IT departments tend to disregard the importance and impact of OS security updates.
-%
-% In general, the longer you wait to upgrade, the harder it gets and eventually a system can get ‘un-upgradable’, therefore, unmaintainable.
-% When deploying the application, it is necessary to make sure that all 3rd party components that are part of the production environment can be upgraded or recreated from backup (in case of ransomware) at any point in time.
-This includes databases and web servers, deployment frameworks, the identity manager, log manager application and alerting system, the operating system, etc.
-%
-% The ‘upgrading is hard’ problem is similar to the ‘releasing is hard’ problem and should be solved the same way: by doing it often.
-% The way to do this is to treat the environment in which your application runs as part of the application and not something separate. As such it should be treated as a software project in itself.
-%
-% #4 Validations should be done in the backend
-% More and more web applications follow a service oriented architecture model, where the backend provides a RESTful API and the front-end is purely implemented in JavaScript (with HTML).
-Some modern JavaScript frameworks provide out of the box escaping and sanitization functionalities (e.g.: Angular 2 or above).
-Some developers are led to mistakenly believe that this free-feature, together with type checking the REST parameters (done in the corresponding controller, for instance) is enough to prevent introduction of malicious input in the application.
-But unfortunately, it is not.
-As much as the web and security frameworks try to enable security mitigations by default, a lot of effort is still needed from the development team.
-Therefore, it’s very important to validate all the user input in the backend and filter-out unwanted/invalid inputs.
-%management of the web application.
-%
-% Notice that the fact that in modern frameworks crypto primitives can be defined in a configuration file, makes it much easier and less time consuming, to, for instance, update those primitives when a new attack is discovered.
-Plus, if the default session manager has a known vulnerability, it is easier to patch the application by updating the library, then doing quick (and perhaps dirty) fixes in custom implementations, that can lead to even more exploits.
-%
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+We would expect that the negative impact of programming languages on maintainability 
+would be more intense due to the buzz of the last years surrounding the idea that 
+poor design of programming languages and lack of using the best practices lead to 
+more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. 
+However, Figure \ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on 
+software maintainability. We suspect that these values are the result of projects 
+contributions policies. 9 out of 10 projects --- the equivalent to 31.8\% of the commits --- 
+of the top-10 of dataset projects with more contributors have restrict contribution 
+policies regarding code standards. Some examples are the \texttt{laravel/framework}, 
+\texttt{php/php-src}, \texttt{openssl/openssl} and 
+\texttt{cakephp/cakephp}. For instance, \texttt{laravel/framework} requires the 
+\texttt{PSR-2 Standard}\footnote{PSR-2 Standard available at 
+https://www.php-fig.org/psr/psr-2/ (Accessed on \today{}}).
 
@@
 
-\textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}}
+\textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}} Although there are
+a few refactorings that harm software maintainability in the \emph{Cross-Site Scripting} and \emph{Using 
+Components with Know Vulnerabilities}, overall maintainability does not suffer any changes for these patterns.
+For the remaining patterns, \emph{Injection} and \emph{Cross-Site Request Forgery}, results are not statistically
+significant.
 
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
+The impact of a refactoring depends on its complexity, i.e., if the refactoring adds complexity to the codebase it is probably affecting the software maintainability.
+One evidence of that is the fact of \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings endure more cases without any impact in the maintainability.
+Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively.
+Whereas, Broken Authentication and vulnerabilities such as the one presented in Section \ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
+Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{}}, where any user could login as root with an empty password.
+Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{}} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
+In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability.
+Real efforts have been made in creating and incorporating coding standards and best practices in software production.
+However, security is still far from perfection and fixing vulnerabilities might hinder software maintainability.
+Thus, it is important to provide tool support to developers to help them applying these patterns without harming software maintainability.
 
@@
 
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
 \section{Conclusion and Future Work}\label{sec:conclusions}
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
-
-Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions of Lorem Ipsum.
 

### 2019-02-01 paper/ICPC19/icpc19.tex
conclusion up

@@
 
+This work presents an empirical study on the impact of 607 security refactorings on
+the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
+of each project version to calculate maintainability based on a model proposed in 
+previous work.
+
+Security refactorings significantly affect the software maintainability of open-source 
+software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
+this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
+overall maintainability increases (50.16\%). However, there is still a very considerable amout of
+cases that harm maintaniability (41.35\%). This study also exhibits evidence that some patterns
+need more attention than others. Broken Autentication issues, Denial-of-Service attacks 
+and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively.
+
+Following the implications of our work, we draw attention for the need of secure security frameworks,
+code best practices documentation to produce safe and maintainable code, proper authorization 
+mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
+maintenance and new or improved programming languages to allow developers to easily implement safety in
+their software without endangering maintainability.
+
+As future work, this study may be extended in many directions such as analyze which guidelines are
+more likely to affect security refactorings; evaluate the impact of programming languages of the
+maintainability of each refactoring pattern; investigate if there is any correlation between 
+software maintainability and projects popularity; expand our methodology with other software quality
+properties; and, validate these findings with closed or private software.
+
 \section*{Acknowledgment}\label{sec:ack}

### 2019-02-01 paper/ICPC19/icpc19.tex
clarifying what OpenSSL is

@@
 OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
-(Accessed on \today{})} software; section \ref{sec:methodology} describes the
+(Accessed on \today{})}; section \ref{sec:methodology} describes the
 methodology used to answer the research questions; section \ref{sec:results}
@@
 As an example, consider the refactoring of the Datagram Transport Layer Security
-(DTLS) protocol implementation in OpenSSL to address the fact that hackers could
-cause a Denial-of-Service (DoS) attack by crafting DTLS
-handshake messages to trigger memory allocations corresponding to large length
-values. This vulnerability is listed at the Common Vulnerabilities and Exposures
-dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at
-http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on
-\today{})}. It is amongst the vulnerabilities studied in our evaluation.
-The following snippet presents the changes performend on the
+(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
+contains open-source implementations of the SSL and TLS cryptographic
+protocols.} to address the fact that hackers could cause a Denial-of-Service
+(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
+corresponding to large length values. This vulnerability is listed at the Common
+Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
+details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
+(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
+evaluation. The following snippet presents the changes performend on the
 \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at

### 2019-02-01 paper/ICPC19/icpc19.tex
fixing

@@
 
-In this study, our concern is to study whether while improving software
-security, developers are also inserting a negative impact on the maintainability
-of their software which may increase technical debt when choosing the easiest
-solution instead of the using the best approach. To answer the following two
+In this paper, our concern is to study whether, while improving software
+security, developers are also negatively impacting the maintainability
+of their software application. To answer the following two
 research questions, we use a dataset of security refactorings to measure the

### 2019-02-02 paper/ICPC19/icpc19.tex
fixing infringe
adding )

@@
 (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
-\emph{Write Simple Units of Code} guideline. And, it increases the codebase size
-in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the
+\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
+size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
+to a potential violation of the
 \emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
@@
 Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{}}, where any user could login as root with an empty password.
-Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{}} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
+Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
 In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability.

### 2019-02-02 paper/ICPC19/icpc19.tex
remove lorem BS

@@
 
-\Sof{Models} Lorem Ipsum is simply dummy text of the printing and typesetting industry.
-Lorem Ipsum has been the industry's standard dummy text ever since the 1500s, when an unknown printer took a galley of type and scrambled it to make a type specimen book.
-It has survived not only five centuries, but also the leap into electronic typesetting, remaining essentially unchanged.
-It was popularised in the 1960s with the release of Letraset sheets containing Lorem Ipsum passages, and more recently with desktop publishing software like Aldus PageMaker including versions.
+\Sof{Models}
 

### 2019-02-02 paper/ICPC19/icpc19.tex
fixes

@@
 As an example, in the last week of January 2019, a severe vulnerability in Apple
-FaceTime was
-revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
-(Accessed on \today{})}. Callers could see and hear others without them picking up the call. These vulnerabilities are gateways for attackers to
-penetrate systems and exploit their resources.
+FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
+(Accessed on \today{})}. Callers could see and hear others without them picking 
+up the call. These vulnerabilities are gateways for attackers to penetrate 
+systems and exploit their resources.
 
@@
 OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
-(Accessed on \today{})}; section \ref{sec:methodology} describes the
+(Accessed on \today{})} software; section \ref{sec:methodology} describes the
 methodology used to answer the research questions; section \ref{sec:results}
@@
 As an example, consider the refactoring of the Datagram Transport Layer Security
-(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
-contains open-source implementations of the SSL and TLS cryptographic
-protocols.} to address the fact that hackers could cause a Denial-of-Service
-(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
-corresponding to large length values. This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
-details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
-(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-evaluation. The following snippet presents the changes performend on the
+(DTLS) protocol implementation in OpenSSL to address the fact that hackers could
+cause a Denial-of-Service (DoS) attack by crafting DTLS
+handshake messages to trigger memory allocations corresponding to large length
+values. This vulnerability is listed at the Common Vulnerabilities and Exposures
+dictionary as CVE-2014-3506\footnote{CVE-2014-3506 details available at
+http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506 (Accessed on
+\today{})}. It is amongst the vulnerabilities studied in our evaluation.
+The following snippet presents the changes performend on the
 \emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
@@
 (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
-\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
-size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
-to a potential violation of the
+\emph{Write Simple Units of Code} guideline. And, it increases the codebase size
+in $4$ lines of code ($13$ lines added, $9$ lines deleted) which infringes the
 \emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
@@
 
-In this paper, our concern is to study whether, while improving software
-security, developers are also negatively impacting the maintainability
-of their software application. To answer the following two
+In this study, our concern is to study whether while improving software
+security, developers are also inserting a negative impact on the maintainability
+of their software which may increase technical debt when choosing the easiest
+solution instead of the using the best approach. To answer the following two
 research questions, we use a dataset of security refactorings to measure the
@@
 \texttt{PSR-2 Standard}\footnote{PSR-2 Standard available at 
-https://www.php-fig.org/psr/psr-2/ (Accessed on \today{}}).
+https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.
 
@@
 Whereas, Broken Authentication and vulnerabilities such as the one presented in Section \ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
-Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{}}, where any user could login as root with an empty password.
+Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user could login as root with an empty password.
 Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
@@
 The maintainability formula was inferred based on the reports of the Better Code Hub.
-Since the dataset has a broad group of open-source projects with different backgrounds, this metric may not be suitable for all of them.
-\Sof{Improve it} However, BCH does have a representative benchmark of
-closed and open-source software projects that is under the calculation of the thresholds used in the computation of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
+The high amount of different projects and backgrounds may require different maintainablity
+standards. However, BCH does have a representative benchmark of closed and open-source 
+software projects that is under the calculation of the thresholds used in the computation 
+of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
 
@@
 
-The security refactorings dataset provided by the previous
-work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub commits produced by
-project developers to classify the changes perfomed by them to fix security flaws. This approach discards security refactorings that were not explicit in commits messages.
+The security refactorings dataset provided by the previous work~\cite{Reis:2017:IJSSE} 
+was collected based on the messages of GitHub commits produced by project developers 
+to classify the changes perfomed by them to fix security flaws. This approach 
+discards security refactorings that were not explicit in commits messages.
 
@@
 characteristics of each project.
-However, maintainability still may be affected by the developers experience, coding style and software contribution polices.
-But maintainability difference is not evaluated at this level.
+However, maintainability still may be affected by the developers experience, coding style and software contribution polices which it is not evaluated in this study.
+Furthermore, this evaluation considers that 607 regular commits are enough to alleviate random irregularities in the maintainability differences of the baseline.
 
@@
 In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation.
-Thus, answering these questions to closed commercial applications may be a future work opportunity.
+However, BCH thresholds are calculated from a set of both open and closed sotfware.
+Different programming languages may require different coding practices to address software safety.
+The dataset
+comprises more commits in C and PHP, i.e., the dataset may not be properly representative of the population.
 
 All the security commits are described in English. This approach discards security refactorings provided in other languages.
+
+\Sof{Talk about language tuning}
+% The total volume in a codebase is the volume in lines of code converted to man- months.
+A man-month is a standard measure of source code volume.
+It is the amount of source code that one developer with average productivity could write in one month.
+The advantage of “man-month” is that it allows for comparisons of source code volume between technologies.
+This is relevant because programming languages have different productivity measures, or “levels of verbosity.” Therefore, a system with multiple programming languages can be converted to an aggregate measure that tells you the approximate effort it would take to rebuild it: the “rebuild value.”
 
@@
 of each project version to calculate maintainability based on a model proposed in 
-previous work.
-
-Security refactorings significantly affect the software maintainability of open-source
+previous work. Security refactorings significantly affect the software maintainability of open-source 
 software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
@@
 need more attention than others. Broken Autentication issues, Denial-of-Service attacks 
-and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively.
-
-Following the implications of our work, we draw attention for the need of secure security frameworks,
+and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following the implications of our work, we draw attention for the need of secure security frameworks,
 code best practices documentation to produce safe and maintainable code, proper authorization 
@@
 We thank to the Software Improvement Group (SIG) team for all the support and
-help in validating our methodology.
+help in validating the methodology.
 

### 2019-02-02 paper/ICPC19/icpc19.tex

--- 2019-02-02 paper/ICPC19/icpc19.tex
@@
 https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}. 
+These vulnerabilities are gateways for attackers to penetrate 
+systems and exploit their resources. 
 As an example, in the last week of January 2019, a severe vulnerability in Apple
-FaceTime was
-revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
-(Accessed on \today{})}. Callers could see and hear others without them picking up the call. These vulnerabilities are gateways for attackers to
-penetrate systems and exploit their resources.
-
-\Sof{Models}
+FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
+(Accessed on \today{})} --- the outcome of a design/logical flaw that allowed callers to see and
+hear others without them picking up the call.
 
@@
 Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
-to find vulnerabilities and defects. Therefore, writing maitainable code
-prevents the introduction of vulnerabilities in the code. In this work, we are
-concerned about the impact that fixing vulnerabilities may have in the
-maintainability of open-source software. Are developers affecting software
+to find vulnerabilities and defects. Therefore, writing maintainable code
+prevents the introduction of vulnerabilities in the codebases. In this work, we are
+concerned about the impact that patching vulnerabilities may have in the
+open-source software maintainability. Are developers affecting software
 maintainability when trying to improve software safety? We suspect that some of
-those fixes may have a negative impact in the software maintainability and
+the patches may have a negative impact in the software maintainability and
 possibly even be the cause of the introduction of new vulnerabilities.
@@
 formula to calculate maintainability, we resort on Software Improvement
-Group (SIG)’s web-based source code analysis service Better Code Hub
+Group (SIG)'s web-based source code analysis service Better Code Hub
 (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
-project. BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
-base of BCH's evaluation. Two examples of guidelines used by BCH are the
+project. There are other well-known standards and models that have been proposed to increase software safety: 
+Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
+costs associated and poor technical evaluation; the OWASP 
+Application Security Verification Criteria~\cite{oswap:2009} which is focused only
+on web applications, and a model proposed by Xu et al. (2013)~\cite{6616351} for 
+raiting software security which we believe it was one of the first steps taken
+by SIG to introduce security on their maintainability model.
+BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
+base of BCH's evaluation. Two examples are the
 \emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
 and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
-easier to maintain and less sensitive to defects~\cite{Visser:2016:OREILLY}.
-Compliance to the guidelines is derived from the SIG's
+easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}. Compliance to the guidelines is derived from the SIG's
 industry benchmark which, according to official sources, consists of over 8
@@
 The particular subject of this paper is to explore whether there is a trade-off
-between refactoring source code to fix identified vulnerabilities while keeping
-(or even improving) the maintainability of the software system. In other words,
-we are interested in answering the following overarching question
+between refactoring source code to patch disclosed vulnerabilities while keeping
+(or even improving) the maintainability of software systems. Hedegus et al. (2018)~\cite{HEGEDUS2018313}, 
+analyzed the difference of maintainability between
+groups of refactored elements and non-refactored and conclude that the source
+code elements subjected to refactorings had significantly lower maintainability
+than the elements not affected.
 
-\begin{framed}
-\noindent
-Does improving security have a negative impact on code
-maintainability?
-\end{framed}
-
-\noindent
 In this paper, we present the results of our analysis on the maintainability of
 607 security changes collected from open-source software, from different
-programming languages.
-
-The main contributions of the present work is:
+programming languages. The main contributions of the present work is:
 \begin{itemize}
@@
 	maintainability and what patterns need more attention.
-
-  % \item An analysis of how likely security-related changes need to be
-  % maintained, in comparison to regular commits.
 
@@
 address \emph{Broken Authentication and Session Management} issues, \emph{Memory
-Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to
-highlight the need for tools and documentation to help developers adopt security
-best practices.
+Leaks} and \emph{Denial-of-Service} attacks.
+There are already several tools that suggest security improvements using static analysis, dynamic analysis, and even artificial intelligence.
+With this study, we intend to highlight the need for tools to predict the impact of patches on maintainability and documentation
+to help developers adopt security best practices~\cite{6311252, 7927935, MESQUIDA201519}.
 
@@
 proposed by BCH's authors for delivering software that is easy to maintain
-\cite{Visser:2016:OREILLY}.
-
-All guidelines are computed in the same evaluation by BCH. During each guideline
-evaluation, the tool determines the compliance towards one guideline by
-establishing limits for the percentage of code allowed to be in each of the $4$
-risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} and
-\emph{very high risk}). If the project does not violate the thresholds, then it
+\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by 
+BCH. During each guideline evaluation, the tool determines the compliance towards 
+one guideline by establishing limits for the percentage of code allowed to be in 
+each of the $4$ risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk} 
+and \emph{very high risk}). If the project does not violate the thresholds, then it
 is compliant with the guideline. These thresholds are calibrated by BCH using
@@
 guideline in the example since the bars do not reach the threshold points. This
-example only complies with $\frac{1}{10}$ guidelines (\emph{Write Clean Code}).}
+example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
 	\label{fig:bchrep}
@@
 (\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
-compliance thresholds for each severity level. \Sof{@todo: Adress thresholds confusion}
+compliance thresholds for each severity level. If the green mark is somewhere
+in the green bar it means it is compliant with a low level of severity.
 
@@
 \begin{itemize}
-	\item \textbf{Security Frameworks} need to be secure. Unfortunately, there are several
-	vulnerabilities that stay undisclosed. Thus, developers should not trust them blindly and
-	check defaults.
+	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately, 
+	there are several vulnerabilities that stay undisclosed. Thus, developers should not 
+	trust them blindly and check defaults.
 	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
@@
 In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation.
-Thus, answering these questions to closed commercial applications may be a future work opportunity.
+However, BCH thresholds are calculated from a set of both open and closed sotfware.
+Different programming languages may require different coding practices to address software safety.
+The dataset
+comprises more commits in C and PHP, i.e., the dataset may not be properly representative of the population regarding to programming languages.
 
@@
 
-\subsection{Security Maintainability Models}
+Many studies have investigated the relationship between refactorings and software quality.
 
-~\cite{6616351,oswap:2009,common:2009}
+Previous work focused on coupling metrics has evaluated the impact of refactorings and
+exhibited proof that quantifying effectively the impact of refactoring on maintainability
+may help to choose the appropriate refactoring type~\cite{1167822}. In contrast to
+this work, Hegedus et al.~\cite{HEGEDUS2018313} did not select particular metrics to
+assess the effect of refactorings. Instead, statistical tests were used to find
+the metrics that have the potential to change significantly after refactorings. 
+The differences of maintainability between groups of refactored elements and 
+non-refactored were analyzed. And, the authors conclude that the source
+code elements subjected to refactorings had significantly lower maintainability
+than the elements not affected. Studying the evolution of maintaninability issues during the development of
+Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability 
+decreases over time. The present work is based on the 10 guidelines proposed
+by the Software Improvement Group (SIG) and focuses solely on evaluating
+the impact of patching vulnerabilities on software maintainability.
+
+Researchers investigated the relationship between design patterns and
+software maintainability~\cite{10.1007/978-3-642-35267-6-18}. 300 revisions
+of \texttt{JHotDraw} were analyzed to conclude that every introduced pattern
+instance improved software maintainability. However, other studies show that the 
+use of design patterns may introduce maintainability issues into software~\cite{4493325}.
+The present work studies how security patterns influence maintainability
+for open-source software.
+
+%
+% Previous work studied the effect of programming languages in the quality of
+% code found~\cite{ray2014large}. It was found that language design has a
+% significant yet moderate impact on software quality. The authors have used the
+% number of defects as a construct of software quality. Our work analyzes
+% software quality in terms of code maintainability to study how it is affected
+% by energy efficiency-code changes.
 
@@
 overall maintainability increases (50.16\%). However, there is still a very considerable amout of
-cases that harm maintaniability (41.35\%). This study also exhibits evidence that some patterns
+cases that harm maintaniability (41.35\%).
+
+This study also exhibits evidence that some patterns
 need more attention than others. Broken Autentication issues, Denial-of-Service attacks 
-and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively.
-
-Following the implications of our work, we draw attention for the need of secure security frameworks,
+and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following the implications of our work, we draw attention for the need of secure security frameworks,
 code best practices documentation to produce safe and maintainable code, proper authorization 
@@
 software maintainability and projects popularity; expand our methodology with other software quality
-properties; and, validate these findings with closed or private software.
+properties; validate these findings with closed or private software; and, expand this analysis
+to other quality standards.
 

### 2019-02-02 paper/ICPC19/icpc19.tex
rw

@@
 by SIG to introduce security on their maintainability model.
+
 BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
@@
 address \emph{Broken Authentication and Session Management} issues, \emph{Memory
-Leaks} and \emph{Denial-of-Service} attacks.
-There are already several tools that suggest security improvements using static analysis, dynamic analysis, and even artificial intelligence.
-With this study, we intend to highlight the need for tools to predict the impact of patches on maintainability and documentation
+Leaks} and \emph{Denial-of-Service} attacks. There are already several tools that 
+suggest security improvements using static analysis, dynamic analysis, and even 
+artificial intelligence. With this study, we intend to highlight the need for tools 
+to predict the impact of patches on maintainability~\cite{4724577} and documentation
 to help developers adopt security best practices~\cite{6311252, 7927935, MESQUIDA201519}.
@@
 
+\subsection{Software Maintainability}
+
 Previous work focused on coupling metrics has evaluated the impact of refactorings and
@@
 Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability 
-decreases over time. The present work is based on the 10 guidelines proposed
+decreases over time. Palomba et al. (2018)\cite{Palomba:2018:DIM:3231288.3231337} exhibits
+proof that code smells should be carefully monitored by programmers, since there is
+a high correlation between maintainability aspects and proneness to changes and faults.
+The present work is based on the 10 guidelines proposed
 by the Software Improvement Group (SIG) and focuses solely on evaluating
@@
 
-%
-% Previous work studied the effect of programming languages in the quality of
-% code found~\cite{ray2014large}. It was found that language design has a
-% significant yet moderate impact on software quality. The authors have used the
-% number of defects as a construct of software quality. Our work analyzes
-% software quality in terms of code maintainability to study how it is affected
-% by energy efficiency-code changes.
+Other studies investigated the effect of programming languages in software quality~\cite{Ray:2017:LSP:3144574.3126905,
+2019arXiv190110220B}. The first study shows that some programming languages are more buggy than others, i.e., 
+language design has impact on software quality. However, the authors of the second
+study were not able to reproduce the first one and did not obtain any evidence about the 
+impact of language design. Both have used the number of defects as an indicator of
+software quality. The present work studies how security refactorings affect software quality
+based on the code maintainability analysis.
 
@@
 need more attention than others. Broken Autentication issues, Denial-of-Service attacks 
-and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following the implications of our work, we draw attention for the need of secure security frameworks,
+and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following 
+the implications of our work, we draw attention for the need of secure security frameworks,
 code best practices documentation to produce safe and maintainable code, proper authorization 
@@
 
+\pagebreak
+
 \balance

### 2019-02-02 paper/ICPC19/icpc19.tex
pagebreak out

@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
- 	\caption{Security Refactorings Language Distribution}
+ 	\caption{Distribution of security refactorings per language}
 	\label{fig:lang}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
- 	\caption{Projects Domain Distribution}
+ 	\centering 	\includegraphics[width=0.50\textwidth]{figures/type_dist.pdf}
+ 	\caption{Distribution of project domain}
 	\label{fig:domain}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
- 	\caption{Maintainability report of OpenSSL CVE-2014-3506 vulnerability
+ 	\centering 	\includegraphics[width=0.50\textwidth]{figures/bch_report.png}
+ 	\caption{Maintainability report of OpenSSL's CVE-2014-3506 vulnerability
 refactoring for the guideline \emph{Write Simple Units of Code} provided by
@@
 
-\pagebreak
-
 \balance

### 2019-02-02 paper/ICPC19/icpc19.tex
fixes

--- 2019-02-02 paper/ICPC19/icpc19.tex
@@
     Anonymou(s) Author(s)
+
 %     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
@@
 contains open-source implementations of the SSL and TLS cryptographic
-protocols.} to address the fact that hackers could cause a Denial-of-Service
+protocols.} to address a vulnerability where an hacker could cause a Denial-of-Service
 (DoS) attack by crafting DTLS handshake messages to trigger memory allocations
@@
 
-- unsigned long frag_len = msg_hdr->frag_len, max_len; /*!\annotation{lst:func3}!*/
+- unsigned long frag_len = msg_hdr->frag_len, max_len; 
 - if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
@@
 +    goto err; /*!\annotation{lst:func4}!*/
-+
+
   frag = dtls1_hm_fragment_new(frag_len, 0);
@@
 
-\ref{lst:func3} Replaced by attribution in line $23$.
-
 \ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message
@@
 maintainability. Even though the developer tries to simplify the codebase with
-the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len}, it
+the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it
 also disrupts $2$ of the guidelines proposed by the Software Improvement Group
@@
 address a security flaw because they tend to choose the easiest path to solve
-it. To address it, we compute the maintainability value of $607$ commits using
+it. To address it, we compute the maintainability of $607$ commits using
 \emph{Better Code Hub}. The same approach is applied to a randomly generated
-dataset of regular commits - baseline - that we use to understand how
+dataset of regular commits --- baseline --- that we use to understand how
 maintainability evolves when security refactorings are performed versus when
@@
 There are security flaws that are more difficult to refactor than others. For
-instance, introducing a mechanism to secure authentication is not so easy to
-implement as fixing a cross-site scripting vulnerability with a \texttt{escape}
+instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with an \texttt{escape}
 function. Knowing which patterns are more likely to increase maintainability
@@
 patterns need more attention. The taxonomy of security patterns used to answer
-this question is the same as the one provided by the authors of the dataset.
+this question is the same as the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
 Maintainability is measured separately for each pattern.
@@
 The methodology used to measure the impact of security refactorings on the
-maintainability of open-source software is discussed in the following sections
-and illustrated in Figure \ref{fig:met}. Aiming to answer the research questions
-outlined in the previous section, in this paper, we use a dataset containing
+maintainability of open-source software is discussed this section. The methodology 
+comprehends the following steps, as illustrated in Figure \ref{fig:met}:
+\begin{enumerate}
+	\item Extract relevant data from a dataset containing
 $716$ security refactorings collected from open-source software available on
-GitHub~\cite{Reis:2017:IJSSE}. Moreover, a baseline of regular commits was
-randomly collected from the list of projects of the main dataset to evaluate
-what is the impact of regular commits on the maintainability of open-source
-software. The Software Improvement Group's\footnote{SIG's website is available
+GitHub~\cite{Reis:2017:IJSSE}.
+	\item A baseline of regular commits was
+randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
+software.
+\item Use the Software Improvement Group's\footnote{SIG's website is available
 at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
 service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
-https://bettercodehub.com/ (Accessed on \today{})} was used to quantify
+https://bettercodehub.com/ (Accessed on \today{})} to quantify
 maintainability for both security and regular commits.
+\end{enumerate}
+
 %
@@
 of the capabilities of the tools \cite{just2014mutants}. The study yielded a
-dataset with $716$ test case for $16$ security patterns. Each test case of the
+dataset with $716$ test cases for $16$ security patterns. Each test case of the
 dataset is a triplet: the commit before the refactoring, the commit responsible
-for the refactoring and the snippets of code that differ from one version to
-another (typically, called \textit{diff}) --- where one can easily identify the
+for the refactoring, and the snippets of code that differ from one version to
+another (typically, called \textit{diffs}) --- where one can easily review the
 code used to fix the security flaw. In this study, we focus on computing the
@@
 The test cases in the dataset are classified with one of the $16$ following
-patterns, as based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top
-10 of 2017~\cite{oswap:2017}: \textit{Injection}, \textit{Broken Authentication
+patterns: \textit{Injection}, \textit{Broken Authentication
 and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access
@@
 \textit{Path Traversal}, \textit{Miscellaneous}, \textit{Resource Leak} and
-\textit{SHA-1 Hash Function}.
+\textit{SHA-1 Hash Function}. Some of the patterns are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 2017~\cite{oswap:2017}
 
@@
 	\caption{Descriptive statistics of the dataset projects}
-\begin{tabular}{@{}lllllllllll@{}}
+\begin{tabular}{@{}rrrrrrrrrrr@{}}
 \toprule
@@
 std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
-min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
-25\%  & 391.50  & 1581    & 117.25   & 49           & 1440.50  & 4        & 19       & 8466.75   & 313.75    & 143.25  \\
-median  & 838.50  & 2836.50 & 248      & 99           & 5504.50  & 9        & 59       & 37372.50  & 1792.50   & 650     \\
-75\%  & 2155    & 6828.50 & 459.50   & 261          & 18579.25 & 20       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
-max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
-Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
+min   & 1.00       & 3.00       & 1.00        & 0.00            & 103.00      & 1.00        & 0.00        & 108.00       & 0.00         & 0.00       \\
+25\%  & 391.50  & 1581.00    & 117.25   & 49.00           & 1440.50  & 4.00        & 19.00       & 8466.75   & 313.75    & 143.25  \\
+median  & 838.50  & 2836.50 & 248.00      & 99.00           & 5504.50  & 9.00        & 59.00       & 37372.50  & 1792.50   & 650.00     \\
+75\%  & 2155.00    & 6828.50 & 459.50   & 261.00          & 18579.25 & 20.00       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
+max   & 16366.00   & 31841.00   & 3446.00     & 413.00          & 114378.00   & 1227.00     & 1114.00     & 995790.00    & 33970.00     & 19329.00   \\
+Total & 165771.00  & 512182.00  & 37729.00    & 14413.00        & 1394412.00  & 4246.00     & 12168.00    & 11559485.00  & 354283.00    & 182511.00  \\ \bottomrule
 \end{tabular}
@@
 To classify the applications in the dataset, we use a taxonomy for open-source
-software published in an older study~\cite{7816479}. Figure \ref{fig:domain}
+software from previous work~\cite{7816479}. Figure \ref{fig:domain}
 presents the projects domain distribution: \textit{Aplication Software} ($25$),
@@
 Table \ref{tab:dataset} presents the descriptive statistics of the $94$
-open-source projects involved in this study, icluding number of forks, number of
-stars, number of watchers, number of contributors, number of commits, number of
-branches, number of releases,  size (cf. GitHub it is the size of the whole
-repository, including all of its history, in kilobytes), number of issues and pull
-requests.
+open-source projects involved in this study, including the number of forks, the number of
+stars, the number of watchers, the number of contributors, the number of commits, the number of
+branches, the number of releases,  size (cf. GitHub it is the size of the whole
+repository, including all of its history, in kilobytes), the number of issues, and the
+number of pull requests.
 
@@
 %
-Previous studies attempted to measure the impact of code refactorings on
-open-source software maintainability~\cite{HEGEDUS2018313} before. But, to the
-best of our knowledge, no previous work has used the BCH model and focused on
-evaluating the impact of security refactorings and compare with regular
-refactorings as in this study. We analyze the maintainability of regular commits
+Previous studies attempted to measure the impact of regular refactorings on
+open-source software maintainability~\cite{HEGEDUS2018313}. However, 
+there is no previous work focused on comparing the impact of security 
+refactorings with regular refactorings on maintainability. 
+We analyze the maintainability of regular commits
 (i.e., commits not related with security fixes) and use them as a baseline.
@@
 
-Figure \ref{fig:bchrep} shows an example of the  guidelines report
-provided by BCH for a project after finishing its evaluation. These are the
-results after performing the OpenSSL CVE-2014-3506 vulnerability refactoring -
-motivation example discussed, previously, in section \ref{sec:motivation}. This
-version of OpenSSL only complies with 1 out of 10 guidelines, the \emph{Write
-Clean Code} guideline.
+Figure \ref{fig:bchrep} shows an example of the report
+provided by BCH for a project after finishing its evaluation. The examples
+refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
+as described as motivating example in Section \ref{sec:motivation}. This
+version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
+Clean Code}.
 
@@
 mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
-maintenance and new or improved programming languages to allow developers to easily implement safety in
-their software without endangering maintainability.
+maintenance and new or improved programming languages to allow developers to easily implement safety in their software without endangering maintainability.
 

### 2019-02-03 paper/ICPC19/icpc19.tex
rw A out
icpc19
pagebreak out; other small fixes
fix
fixed conflict
fixing sentence

@@
 \usepackage{xcolor}
-
+\usepackage[hidelinks]{hyperref}
 \usepackage{xfrac}
@@
 	\item A replication package with all scripts and data created to perform the
-	empirical evaluation, for reproducibility. Available here: $<url>$.
+	empirical evaluation, for reproducibility. Available online:
+  \url{https://figshare.com/s/3e17518222df6a11a8d6}.
 
@@
 
+\medskip
+
 \setcounter{lstannotation}{0}
@@
 \\CVE-2014-3506 vulnerability},label={lst:vuln}]
+
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
@@
 (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
-(\texttt{max\_len}). If yes, it return the certificate list size, if not it
-return the maximum length.
+(\texttt{max\_len}). If that condition holds, the function returns the size of
+the certificate list, otherwise it returns the maximum length.
 
@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
- 	\caption{Distribution of security refactorings per language}
+ 	\caption{Distribution of Security Refactorings per Language}
 	\label{fig:lang}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.50\textwidth]{figures/type_dist.pdf}
- 	\caption{Distribution of project domain}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
+ 	\caption{Distribution of Projects Domain}
 	\label{fig:domain}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.50\textwidth]{figures/bch_report.png}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-2014-3506 vulnerability
@@
 software projects that is under the calculation of the thresholds used in the computation
-of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
+of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY,
+Baggen2012}.
 
@@
 Many studies have investigated the relationship between refactorings and software quality.
-
-\subsection{Software Maintainability}
-
 Previous work focused on coupling metrics has evaluated the impact of refactorings and
@@
 code elements subjected to refactorings had significantly lower maintainability
-than the elements not affected. Studying the evolution of maintaninability issues during the development of
-Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability
+than the elements not affected. Studying the evolution of maintaninability issues during the development of Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability
 decreases over time. Palomba et al. (2018)\cite{Palomba:2018:DIM:3231288.3231337} exhibits

### 2019-02-03 paper/ICPC19/icpc19.tex
comments macros
lista boring + refs and cites

@@
 \newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{red} #1}\textbf{]]}}
+\newcommand{\Luis}[1]{\textbf{[Luis:[}{\color{green} #1}\textbf{]]}}
+\newcommand{\Rui}[1]{\textbf{[Rui:[}{\color{blue} #1}\textbf{]]}}
 
@@
 measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
-dataset including $607$ security commits. Results show that fixing software
+dataset including 607 security commits. Results show that fixing software
 vulnerabilities hinders the maintainability of open-source software. This is
@@
 
-This paper is structured as follows: section \ref{sec:motivation} introduces an
+This paper is structured as follows: section~\ref{sec:motivation} introduces an
 example of a security refactoring of a known vulnerability found in
 OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
-(Accessed on \today{})}; section \ref{sec:methodology} describes the
-methodology used to answer the research questions; section \ref{sec:results}
-presents the results and section \ref{sec:discussion} discusses their
-implications; section \ref{sec:threats}, threats to the validity of the study
-are enumerated; section \ref{sec:rw} describres the different work and existing
-literature in the field of study; and, finally section \ref{sec:conclusions}
+(Accessed on \today{})}; section~\ref{sec:methodology} describes the
+methodology used to answer the research questions; section~\ref{sec:results}
+presents the results and section~\ref{sec:discussion} discusses their
+implications; section~\ref{sec:threats}, threats to the validity of the study
+are enumerated; section~\ref{sec:rw} describres the different work and existing
+literature in the field of study; and, finally section~\ref{sec:conclusions}
 concludes the main findings and elaborates on future work.
@@
 
-\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
+~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
 inserted to limit the number of bytes allowed in a DTLS handshake message for
@@
 
-\ref{lst:func2} Checks if the maximum certificate list size
+~\ref{lst:func2} Checks if the maximum certificate list size
 (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
@@
 
-\ref{lst:func5} and \ref{lst:func4} add the check to ensure that the message
+~\ref{lst:func5} and~\ref{lst:func4} add the check to ensure that the message
 length is never higher than the maxiumum allowed by the DTSL protocol in
@@
 maintainability. Even though the developer tries to simplify the codebase with
-the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it
+the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (~\ref{lst:func1}), it
 also disrupts $2$ of the guidelines proposed by the Software Improvement Group
@@
 new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
-(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
+(~\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
 \emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
@@
 maintainability of open-source software is discussed this section. The methodology
-comprehends the following steps, as illustrated in Figure \ref{fig:met}:
+comprehends the following steps, as illustrated in Figure~\ref{fig:met}:
 \begin{enumerate}
@@
 using hand-seeded test cases or mutations could lead to misleading assessments
-of the capabilities of the tools \cite{just2014mutants}. The study yielded a
-dataset with $716$ test cases for $16$ security patterns. Each test case of the
+of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
+dataset with $716$ test cases for $16$ security patterns. Some of the patterns 
+are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 
+2017~\cite{oswap:2017}. Each test case of the
 dataset is a triplet: the commit before the refactoring, the commit responsible
@@
 \end{figure}
-
-The test cases in the dataset are classified with one of the $16$ following
-patterns: \textit{Injection}, \textit{Broken Authentication
-and Session Management}, \textit{Cross-Site Scripting}, \textit{Broken Access
-Control}, \textit{Security Misconfiguration}, \textit{Sensitive Data Exposure},
-\textit{Insufficient Attack Protection}, \textit{Cross-Site Request Forgery},
-\textit{Using Components With Known Vulnerabilities}, \textit{Unprotected APIs},
-\textit{Memory Leak}, \textit{Overflow}, \textit{Denial-of-Service},
-\textit{Path Traversal}, \textit{Miscellaneous}, \textit{Resource Leak} and
-\textit{SHA-1 Hash Function}. Some of the patterns are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of 2017~\cite{oswap:2017}
 
@@
 	example, flaws allowing crackers to trigger memory allocations corresponding
-	to large length values (Listing \ref{lst:vuln}).
+	to large length values (Listing~\ref{lst:vuln}).
 
@@
 	pattern yet. The patterns not satisfying the size requirement of more than 20
-	refactorings are contemplated here.
+	refactorings are contemplated here: \textit{Resource Leak},
+	\textit{SHA-1 Hash Function}, \textit{Broken Access Control}, \textit{Sensitive Data Exposure},
+	\textit{Insufficient Attack Protection}, \textit{Overflow}, \textit{Path Traversal}.
 
@@
 The dataset contemplates security flaws for more than 13 different languages
-being PHP (39\%) and C (24\%) the most prevalent ones; see Figure \ref{fig:lang}.
+being PHP (39\%) and C (24\%) the most prevalent ones; see Figure~\ref{fig:lang}.
 To classify the applications in the dataset, we use a taxonomy for open-source
-software from previous work~\cite{7816479}. Figure \ref{fig:domain}
+software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
 presents the projects domain distribution: \textit{Aplication Software} ($25$),
@@
 %
-Table \ref{tab:dataset} presents the descriptive statistics of the $94$
+Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
 open-source projects involved in this study, including the number of forks, the number of
@@
 Hub} (BCH) is used to collect the maintanability reports of the refactorings of
-each project. Table \ref{tab:guidelines} presents the 10 guidelines
-proposed by BCH's authors for delivering software that is easy to maintain
-\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by
+each project. Table~\ref{tab:guidelines} presents the 10 guidelines
+proposed by BCH's authors for delivering software that is easy to maintain~\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by
 BCH. During each guideline evaluation, the tool determines the compliance towards
@@
 
-Figure \ref{fig:bchrep} shows an example of the report
+Figure~\ref{fig:bchrep} shows an example of the report
 provided by BCH for a project after finishing its evaluation. The examples
 refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
-as described as motivating example in Section \ref{sec:motivation}. This
+as described as motivating example in Section~\ref{sec:motivation}. This
 version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
@@
 \emph{Units} are the smallest groups of code that can be maintained and executed
-independently \cite{Visser:2016:OREILLY}. McCabe \cite{1702388} is a software
+independently~\cite{Visser:2016:OREILLY}. McCabe~\cite{1702388} is a software
 metric that calculates the number of linearly independent paths of the source
@@
 branch points per unit (\emph{at most 5}), i.e., the units are compliant with
-ISO 25010 \cite{iso:2011}. Yellow, orange and red bars represent units that do
+ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
 not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
@@
 Aiming to analyze the impact of security refactorings, we use BCH to compute
-maintainability of two different versions of the project (Figure
-\ref{fig:commit}):
+maintainability of two different versions of the project (Figure~\ref{fig:commit}):
 \begin{itemize}
@@
 equation to capture the distance between the current state of the project and
-the standard thresholds based on previous work \cite{Olivari:2018}.
+the standard thresholds based on previous work\cite{Olivari:2018}.
 \begin{itemize}
@@
 \noindent
-where $G$ is the group of selected maintainability guidelines from BCH (Table
-\ref{tab:guidelines}) and $v$ is the version of the software under evaluation.
+where $G$ is the group of selected maintainability guidelines from BCH (Table~\ref{tab:guidelines}) and $v$ is the version of the software under evaluation.
 The maintenance $M$ based on the guideline $g$ for a given version of a project
@@
 is a standard value defined by BCH. In other words, the factor $w$ is used in
-Equation \ref{eq:3} to highlight the lines of code that are not complying with
+Equation~\ref{eq:3} to highlight the lines of code that are not complying with
 the guideline. Then, we compute the difference of maintainability between the
 security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
-Figure \ref{fig:commit}.
+Figure~\ref{fig:commit}.
 
@@
 (e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
-test with the significance level $\alpha = 0.05$ \cite{10.2307/3001968} . In
+test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968} . In
 other words, we test the null hypothesis that the maintainability difference
@@
 provides an alternative that incorporates the cases where the maintainability
-difference is zero \cite{10.2307/2282543}. To understand the effect-size, as
-advocated by the Common-language effect sizes \cite{graw:1992}, we compute the
+difference is zero~\cite{10.2307/2282543}. To understand the effect-size, as
+advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
 mean difference, the median of the difference, and the percentage of cases that
@@
 different open-source projects distributed among 5 different main types of
-software, Figure \ref{fig:domain}. The following section presents the results
+software, Figure~\ref{fig:domain}. The following section presents the results
 obtained for each research question.
@@
 The impact of security and baseline commits in the maintainability of
-open-source software is presented in Figure \ref{fig:secvsreg}. For each type
+open-source software is presented in Figure~\ref{fig:secvsreg}. For each type
 of change, three types of impact of refactorings in the software maintainability
@@
 In this study, we also evaluated the impact of each programming language on the
-maintainability of open-source software, Figure \ref{fig:lang_main}. Although the
+maintainability of open-source software, Figure~\ref{fig:lang_main}. Although the
 dataset contains security refactorings for more than $13$ languages, we only
@@
 The impact of each type of security refactoring on the maintainability of
-open-source software is presented in Figure \ref{fig:pat}. The results of the
+open-source software is presented in Figure~\ref{fig:pat}. The results of the
 Wilcoxon signed-rank test by pattern returned statistical evidence which
@@
 40.2\% of the security refactoring have a negative impact on the maintainability
-of open-source software, Figure \ref{fig:secvsreg}. This raises a new tradeoff when
+of open-source software, Figure~\ref{fig:secvsreg}. This raises a new tradeoff when
 developers need to fix vulnerabilities on their projects because they might be affecting
@@
 more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}.
-However, Figure \ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on
+However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on
 software maintainability. We suspect that these values are the result of projects
@@
 Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively.
-Whereas, Broken Authentication and vulnerabilities such as the one presented in Section \ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
+Whereas, Broken Authentication and vulnerabilities such as the one presented in Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
 Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user could login as root with an empty password.
@@
 software projects that is under the calculation of the thresholds used in the computation
-of each maintainability guideline (Table \ref{tab:guidelines})~\cite{Visser:2016:OREILLY,
+of each maintainability guideline (Table~\ref{tab:guidelines})~\cite{Visser:2016:OREILLY,
 Baggen2012}.
@@
 than the elements not affected. Studying the evolution of maintaninability issues during the development of Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability
-decreases over time. Palomba et al. (2018)\cite{Palomba:2018:DIM:3231288.3231337} exhibits
+decreases over time. Palomba et al. (2018)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits
 proof that code smells should be carefully monitored by programmers, since there is

### 2019-02-03 paper/ICPC19/icpc19.tex
not starting sentence with numbers

@@
 
-40.2\% of the security refactoring have a negative impact on the maintainability
-of open-source software, Figure~\ref{fig:secvsreg}. This raises a new tradeoff when
-developers need to fix vulnerabilities on their projects because they might be affecting
-the software maintainability or even introducing new vulnerabilities. This study exhibits
-proof that developers may have to reduce maintainability for the sake of safety. We understand
-that developers should be able to produce secure code without affecting the maintainability of
+In particular, 40.2\% of the security refactoring have a negative impact on the
+maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
+a new tradeoff when developers need to fix vulnerabilities on their projects
+because they might be affecting the software maintainability or even introducing
+new vulnerabilities. This study exhibits evidence that developers may have to
+reduce maintainability for the sake of safety. We understand that developers
+should be able to produce secure code without affecting the maintainability of
 their projects. But there are still some concerns that need to be addressed:

### 2019-02-03 paper/ICPC19/icpc19.tex
review

--- 2019-02-03 paper/ICPC19/icpc19.tex
@@
 \textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
+\Luis{Nao uses numeros no inico da frase. --> We found that 40.2\% of the security refactorings}
 
@@
 
-The impact of a refactoring depends on its complexity, i.e., if the refactoring adds complexity to the codebase it is probably affecting the software maintainability.
-One evidence of that is the fact of \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings endure more cases without any impact in the maintainability.
-Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively.
-Whereas, Broken Authentication and vulnerabilities such as the one presented in Section \ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
-Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user could login as root with an empty password.
-Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
-In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability.
-Real efforts have been made in creating and incorporating coding standards and best practices in software production.
-However, security is still far from perfection and fixing vulnerabilities might hinder software maintainability.
-Thus, it is important to provide tool support to developers to help them applying these patterns without harming software maintainability.
+The impact of a refactoring depends on its complexity, i.e., if the refactoring
+adds complexity to the codebase it is probably affecting the software
+maintainability. One evidence of that is the fact of \emph{Cross-Site
+Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings
+endure more cases without any impact in the maintainability. Typically, these
+type of vulnerabilities are one line fixes --- adding an \texttt{escape}
+function or changing the version of the component, respectively. Whereas,
+Broken Authentication and vulnerabilities such as the one presented in Section
+\ref{sec:motivation} responsible by DoS attacks are more difficult to fix
+(several code lines deleted and added). Another example is the Apple
+authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872
+details available at https://support.apple.com/en-us/HT208315 (Accessed on
+\today{})}, where any user could login as root with an empty password. Patrick
+Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You
+Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on
+\today{})} and conclude that the flaw was due to the introduction of high
+cyclomatic complexity in a method used to verify the password. In sum,
+\emph{Cross-Site Scripting} and \emph{Using Components with Known
+Vulnerabilities} should take part of the developers preoccupations along with
+the other patterns that explicitly hinder maintainability because despite their
+low complexity these are patterns that still produce a considerable negative
+impact on the maintainability. Real efforts have been made in creating and
+incorporating coding standards and best practices in software production.
+However, security is still far from perfection and fixing vulnerabilities might
+hinder software maintainability. Thus, it is important to provide tool support
+to developers to help them applying these patterns without harming software
+maintainability.
 
@@
 %
+
+\Luis{Better Code Hub -> BCH}
 The maintainability formula was inferred based on the reports of the Better Code Hub.
-The high amount of different projects and backgrounds may require different maintainablity
+The high amount of different projects and backgrounds may require different maintainability
 standards. However, BCH does have a representative benchmark of closed and open-source
@@
 
-Baseline commits are retrieved randomly by selecting one commit from the same project of a given security refactoring. This approach softens the differences that may result from the
-characteristics of each project.
-However, maintainability still may be affected by the developers experience, coding style and software contribution polices which it is not evaluated in this study.
-Furthermore, this evaluation considers that 607 regular commits are enough to alleviate random irregularities in the maintainability differences of the baseline.
+Baseline commits are retrieved randomly by selecting one commit from the same
+project of a given security refactoring. This approach softens the differences
+that may result from the characteristics of each project. However,
+maintainability still may be affected by the developers experience, coding
+style and software contribution polices which it is not evaluated in this
+study. Furthermore, this evaluation considers that 607 regular commits are
+enough to alleviate random irregularities in the maintainability differences of
+the baseline.
+
+\Luis{maintainability still may be affected -> maintainability may sitll be affected}
 
@@
 
-In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation.
-However, BCH thresholds are calculated from a set of both open and closed sotfware.
-Different programming languages may require different coding practices to address software safety.
-The dataset
-comprises more commits in C and PHP, i.e., the dataset may not be properly representative of the population regarding to programming languages.
-
-All the security commits are described in English. This approach discards security refactorings provided in other languages.
+\Luis{A primeira threat que falas é que os resultados podem nao generalizar
+para software fechado. Eu acho que deviamos ser mais diretos. Para alem disso
+tirava o "In contrast to the BCH benchmark,"}
+
+In contrast to the BCH benchmark, this study uses a dataset that only covers
+open-source software to perform the maintainability evaluation. However, BCH
+thresholds are calculated from a set of both open and closed software.
+Different programming languages may require different coding practices to
+address software safety. The dataset comprises more commits in C and PHP, i.e.,
+the dataset may not be properly representative of the population regarding to
+programming languages.
+
+All the security commits are described in English. This approach discards
+security refactorings provided in other languages.
 
 \section{Related Work}\label{sec:rw}
+
+\Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
+fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
 
@@
 
+\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
+Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
+diferenca é que estamos a estudar maintainability issues noutro contexto.}
+
 Researchers investigated the relationship between design patterns and
@@
 based on the code maintainability analysis.
+
+\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
 
@@
 
-This study also exhibits evidence that some patterns
-need more attention than others. Broken Autentication issues, Denial-of-Service attacks
-and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
-the implications of our work, we draw attention for the need of secure security frameworks,
-code best practices documentation to produce safe and maintainable code, proper authorization
-mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
-maintenance and new or improved programming languages to allow developers to easily implement safety in their software without endangering maintainability.
+\luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
+
+This study also exhibits evidence that some patterns need more attention than
+others. Broken Autentication issues, Denial-of-Service attacks and Memory Leaks
+endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
+the implications of our work, we draw attention for the need of secure security
+frameworks, code best practices documentation to produce safe and maintainable
+code, proper authorization mechanisms design and implementation, better tools
+to introduce in the CI/CD pipeline to help software maintenance and new or
+improved programming languages to allow developers to easily implement safety
+in their software without endangering maintainability.
+
+\luis{many directions such as -> many directions:}
 

### 2019-02-03 paper/ICPC19/icpc19.tex
review

--- 2019-02-03 paper/ICPC19/icpc19.tex
@@
 
-The impact of a refactoring depends on its complexity, i.e., if the refactoring adds complexity to the codebase it is probably affecting the software maintainability.
-One evidence of that is the fact of \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings endure more cases without any impact in the maintainability.
-Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively.
-Whereas, Broken Authentication and vulnerabilities such as the one presented in Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
-Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user could login as root with an empty password.
-Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
-In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability.
-Real efforts have been made in creating and incorporating coding standards and best practices in software production.
-However, security is still far from perfection and fixing vulnerabilities might hinder software maintainability.
-Thus, it is important to provide tool support to developers to help them applying these patterns without harming software maintainability.
+The impact of a refactoring depends on its complexity, i.e., if the refactoring
+adds complexity to the codebase it is probably affecting the software
+maintainability. One evidence of that is the fact of \emph{Cross-Site
+Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings
+endure more cases without any impact in the maintainability. Typically, these
+type of vulnerabilities are one line fixes --- adding an \texttt{escape}
+function or changing the version of the component, respectively. Whereas,
+Broken Authentication and vulnerabilities such as the one presented in
+Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to
+fix (several code lines deleted and added). Another example is the Apple
+authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872
+details available at https://support.apple.com/en-us/HT208315 (Accessed on
+\today{})}, where any user could login as root with an empty password. Patrick
+Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You
+Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on
+\today{})} and conclude that the flaw was due to the introduction of high
+cyclomatic complexity in a method used to verify the password. In sum,
+\emph{Cross-Site Scripting} and \emph{Using Components with Known
+Vulnerabilities} should take part of the developers preoccupations along with
+the other patterns that explicitly hinder maintainability because despite their
+low complexity these are patterns that still produce a considerable negative
+impact on the maintainability. Real efforts have been made in creating and
+incorporating coding standards and best practices in software production.
+However, security is still far from perfection and fixing vulnerabilities might
+hinder software maintainability. Thus, it is important to provide tool support
+to developers to help them applying these patterns without harming software
+maintainability.
 
@@
 
-This study also exhibits evidence that some patterns
-need more attention than others. Broken Autentication issues, Denial-of-Service attacks
-and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
-the implications of our work, we draw attention for the need of secure security frameworks,
-code best practices documentation to produce safe and maintainable code, proper authorization
-mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
-maintenance and new or improved programming languages to allow developers to easily implement safety in their software without endangering maintainability.
+\luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
+
+This study also exhibits evidence that some patterns need more attention than
+others. Broken Autentication issues, Denial-of-Service attacks and Memory Leaks
+endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
+the implications of our work, we draw attention for the need of secure security
+frameworks, code best practices documentation to produce safe and maintainable
+code, proper authorization mechanisms design and implementation, better tools
+to introduce in the CI/CD pipeline to help software maintenance and new or
+improved programming languages to allow developers to easily implement safety
+in their software without endangering maintainability.
+
+\luis{many directions such as -> many directions:}
+\luis{evaluate the impact of programming languages of the
+maintainability of each refactoring pattern; - > evaluate the impact of each refactoring pattern on the maintainbility of each  }
 

### 2019-02-03 paper/ICPC19/icpc19.tex
fix luis

@@
 
-\luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
+\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
 
@@
 
-\luis{many directions such as -> many directions:}
-\luis{evaluate the impact of programming languages of the
+\Luis{many directions such as -> many directions:}
+\Luis{evaluate the impact of programming languages of the
 maintainability of each refactoring pattern; - > evaluate the impact of each refactoring pattern on the maintainbility of each  }

### 2019-02-03 paper/ICPC19/icpc19.tex
example

@@
 maintainability. Even though the developer tries to simplify the codebase with
-the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (~\ref{lst:func1}), it
+the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it
 also disrupts $2$ of the guidelines proposed by the Software Improvement Group
@@
 new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
-(~\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
+(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
 \emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
@@
 There are security flaws that are more difficult to refactor than others. For
-instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with an \texttt{escape}
-function. Knowing which patterns are more likely to increase maintainability
+instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with simple function to sanitize data.
+Knowing which patterns are more likely to increase maintainability
 issues is one step forward to bring awareness to security engineers of what
@@
 Maintainability is measured separately for each pattern.
+
+\setcounter{lstannotation}{0}
+\begin{lstlisting}[caption={Fix provided by nextcloud/server developers to a \\Cross-Site Scripting vulnerability},label={lst:fix},language=php]
+ <ul>
+    <li class='error'>
+       <?php echo $l->t( 'Cloud not found' ); ?><br/>
+-      <p class='hint'><?php if(isset($_['file'])) echo $_['file']?></p>
++      <p class='hint'><?php if(isset($_['file'])) echo htmlentities($_['file'])?></p>
+    </li>
+ </ul>
+
+\end{lstlisting}
 %
@@
 \textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
-\Luis{Nao uses numeros no inico da frase. --> We found that 40.2\% of the security refactorings}
-
-In particular, 40.2\% of the security refactoring have a negative impact on the
-maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
-a new tradeoff when developers need to fix vulnerabilities on their projects
-because they might be affecting the software maintainability or even introducing
-new vulnerabilities. This study exhibits evidence that developers may have to
-reduce maintainability for the sake of safety. We understand that developers
-should be able to produce secure code without affecting the maintainability of
+
+40.2\% of the security refactoring have a negative impact on the maintainability
+of open-source software, Figure~\ref{fig:secvsreg}. This raises a new tradeoff when
+developers need to fix vulnerabilities on their projects because they might be affecting
+the software maintainability or even introducing new vulnerabilities. This study exhibits
+proof that developers may have to reduce maintainability for the sake of safety. We understand
+that developers should be able to produce secure code without affecting the maintainability of
 their projects. But there are still some concerns that need to be addressed:
@@
 
-The impact of a refactoring depends on its complexity, i.e., if the refactoring
-adds complexity to the codebase it is probably affecting the software
-maintainability. One evidence of that is the fact of \emph{Cross-Site
-Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings
-endure more cases without any impact in the maintainability. Typically, these
-type of vulnerabilities are one line fixes --- adding an \texttt{escape}
-function or changing the version of the component, respectively. Whereas,
-Broken Authentication and vulnerabilities such as the one presented in
-Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to
-fix (several code lines deleted and added). Another example is the Apple
-authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872
-details available at https://support.apple.com/en-us/HT208315 (Accessed on
-\today{})}, where any user could login as root with an empty password. Patrick
-Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You
-Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on
-\today{})} and conclude that the flaw was due to the introduction of high
-cyclomatic complexity in a method used to verify the password. In sum,
-\emph{Cross-Site Scripting} and \emph{Using Components with Known
-Vulnerabilities} should take part of the developers preoccupations along with
-the other patterns that explicitly hinder maintainability because despite their
-low complexity these are patterns that still produce a considerable negative
-impact on the maintainability. Real efforts have been made in creating and
-incorporating coding standards and best practices in software production.
-However, security is still far from perfection and fixing vulnerabilities might
-hinder software maintainability. Thus, it is important to provide tool support
-to developers to help them applying these patterns without harming software
-maintainability.
+The impact of a refactoring depends on its complexity, i.e., if the refactoring adds complexity to the codebase it is probably affecting the software maintainability.
+One evidence of that is the fact of \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings endure more cases without any impact in the maintainability.
+Typically, these type of vulnerabilities are one line fixes --- adding an \texttt{escape} function or changing the version of the component, respectively.
+Whereas, Broken Authentication and vulnerabilities such as the one presented in Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to fix (several code lines deleted and added).
+Another example is the Apple authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872 details available at https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user could login as root with an empty password.
+Patrick Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and conclude that the flaw was due to the introduction of high cyclomatic complexity in a method used to verify the password.
+In sum, \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities} should take part of the developers preocupations along with the other patterns that explicitly hinder maintainability because despite their low complexity these are patterns that still produce a considerable negative impact on the maintainability.
+Real efforts have been made in creating and incorporating coding standards and best practices in software production.
+However, security is still far from perfection and fixing vulnerabilities might hinder software maintainability.
+Thus, it is important to provide tool support to developers to help them applying these patterns without harming software maintainability.
 
@@
 %
-
-\Luis{Better Code Hub -> BCH}
 The maintainability formula was inferred based on the reports of the Better Code Hub.
-The high amount of different projects and backgrounds may require different maintainability
+The high amount of different projects and backgrounds may require different maintainablity
 standards. However, BCH does have a representative benchmark of closed and open-source
@@
 
-Baseline commits are retrieved randomly by selecting one commit from the same
-project of a given security refactoring. This approach softens the differences
-that may result from the characteristics of each project. However,
-maintainability still may be affected by the developers experience, coding
-style and software contribution polices which it is not evaluated in this
-study. Furthermore, this evaluation considers that 607 regular commits are
-enough to alleviate random irregularities in the maintainability differences of
-the baseline.
-
-\Luis{maintainability still may be affected -> maintainability may sitll be affected}
+Baseline commits are retrieved randomly by selecting one commit from the same project of a given security refactoring. This approach softens the differences that may result from the
+characteristics of each project.
+However, maintainability still may be affected by the developers experience, coding style and software contribution polices which it is not evaluated in this study.
+Furthermore, this evaluation considers that 607 regular commits are enough to alleviate random irregularities in the maintainability differences of the baseline.
 
@@
 
-\Luis{A primeira threat que falas é que os resultados podem nao generalizar
-para software fechado. Eu acho que deviamos ser mais diretos. Para alem disso
-tirava o "In contrast to the BCH benchmark,"}
-
-In contrast to the BCH benchmark, this study uses a dataset that only covers
-open-source software to perform the maintainability evaluation. However, BCH
-thresholds are calculated from a set of both open and closed software.
-Different programming languages may require different coding practices to
-address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be properly representative of the population regarding to
-programming languages.
-
-All the security commits are described in English. This approach discards
-security refactorings provided in other languages.
+In contrast to the BCH benchmark, this study uses a dataset that only covers open-source software to perform the maintanibility evaluation.
+However, BCH thresholds are calculated from a set of both open and closed sotfware.
+Different programming languages may require different coding practices to address software safety.
+The dataset
+comprises more commits in C and PHP, i.e., the dataset may not be properly representative of the population regarding to programming languages.
+
+All the security commits are described in English. This approach discards security refactorings provided in other languages.
 
 \section{Related Work}\label{sec:rw}
-
-\Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
-fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
 
@@
 
-\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
-Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
-diferenca é que estamos a estudar maintainability issues noutro contexto.}
-
 Researchers investigated the relationship between design patterns and
@@
 based on the code maintainability analysis.
-
-\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
 
@@
 
-\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
-
-This study also exhibits evidence that some patterns need more attention than
-others. Broken Autentication issues, Denial-of-Service attacks and Memory Leaks
-endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
-the implications of our work, we draw attention for the need of secure security
-frameworks, code best practices documentation to produce safe and maintainable
-code, proper authorization mechanisms design and implementation, better tools
-to introduce in the CI/CD pipeline to help software maintenance and new or
-improved programming languages to allow developers to easily implement safety
-in their software without endangering maintainability.
-
-\Luis{many directions such as -> many directions:}
-\Luis{evaluate the impact of programming languages of the
-maintainability of each refactoring pattern; - > evaluate the impact of each refactoring pattern on the maintainbility of each  }
+This study also exhibits evidence that some patterns
+need more attention than others. Broken Autentication issues, Denial-of-Service attacks
+and Memory Leaks endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
+the implications of our work, we draw attention for the need of secure security frameworks,
+code best practices documentation to produce safe and maintainable code, proper authorization
+mechanisms design and implementation, better tools to introduce in the CI/CD pipeline to help software
+maintenance and new or improved programming languages to allow developers to easily implement safety in their software without endangering maintainability.
 

### 2019-02-03 paper/ICPC19/icpc19.tex
fixed one comment
example

@@
 \newcommand{\lstbg}[3][0pt]{{\fboxsep#1\colorbox{#2}{\strut #3}}}
-\lstset{ %
+
+\lstdefinestyle{CStyle} {
     language=C,
+    backgroundcolor=\color{white},   % choose the background color
+    basicstyle=\ttfamily\scriptsize,        % size of fonts used for the code
+    breaklines=true,                 % automatic line breaking only at whitespace
+    captionpos=b,                    % sets the caption-position to bottom
+    commentstyle=\color{mygray}\bfseries,    % comment style
+    escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+    keywordstyle=\color{blue},       % keyword style
+    stringstyle=\color{mymauve},     % string literal style
+    frame=single,
+    numbers=left,
+    stepnumber=1,
+    xleftmargin=2em,
+    escapeinside={/*!}{!*/},
+    moredelim=**[l][\color{mygreen}]{+\ },
+    moredelim=*[l][\color{red}]{-\ }
+}
+
+\lstdefinestyle{PHPStyle} {
+    language=PHP,
     backgroundcolor=\color{white},   % choose the background color
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[caption={Fix provided by OpenSSL developers to the
+\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
 \\CVE-2014-3506 vulnerability},label={lst:vuln}]
-
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
 +  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
-+  if(max_len < (unsigned long)s->max_cert_list) /*!\annotation{lst:func2}!*/
++  if(max_len < (unsigned long)s->max_cert_list) 
 +     return s->max_cert_list;
@@
 
+In the example (Listing~\ref{lst:vuln}), all functions receive as input the \textbf{handshake} represented by \texttt{s}.
+Every SSL/TLS connection begins with a handshake which is reponsible by the negotiation between two parties.
+The handshake determines what cryptographic algorithms will be used to encrypt communications, verify the server and establish a secure connection before beginning data transfer.
 In the \texttt{dtls1\_reassemble\_fragment} function the value of
@@
 message commit, $10$ handshake messages are allowed, so the attacker could
-consume approximately $180MB$ per DTLS connection.
+consume until $180MB$ per DTLS connection.
 
@@
 
-~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
-inserted to limit the number of bytes allowed in a DTLS handshake message for
-\texttt{s}. This function replaces lines $18$---$21$.
-
-~\ref{lst:func2} Checks if the maximum certificate list size
+~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to limit the number of bytes allowed in a DTLS handshake message,
+\texttt{s}. Checks if the maximum certificate list size
 (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
 (\texttt{max\_len}). If that condition holds, the function returns the size of
-the certificate list, otherwise it returns the maximum length.
+the certificate list, otherwise it returns the maximum length. This function replaces lines $18$-$21$.
 
@@
 There are security flaws that are more difficult to refactor than others. For
-instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with simple function to sanitize data.
+instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with simple function to sanitize data\footnote{More information about \texttt{htmlentities()} at http://php.net/manual/en/function.htmlentities.php (Accessed on \today{})}.
 Knowing which patterns are more likely to increase maintainability
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[caption={Fix provided by nextcloud/server developers to a \\Cross-Site Scripting vulnerability},label={lst:fix},language=php]
+\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix},language=php]
  <ul>
     <li class='error'>
-       <?php echo $l->t( 'Cloud not found' ); ?><br/>
+       <?php echo $l->t('Cloud not found');?><br/>
 -      <p class='hint'><?php if(isset($_['file'])) echo $_['file']?></p>

### 2019-02-03 paper/ICPC19/icpc19.tex
some fixes

@@
 \textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
-\Luis{Nao uses numeros no inico da frase. --> We found that 40.2\% of the security refactorings}
-
-In particular, 40.2\% of the security refactoring have a negative impact on the
+
+We found that 40.2\% of the security refactoring have a negative impact on the
 maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
@@
 
-\Luis{Better Code Hub -> BCH}
-The maintainability formula was inferred based on the reports of the Better Code Hub.
+The maintainability formula was inferred based on the BCH's reports.
 The high amount of different projects and backgrounds may require different maintainability
@@
 that may result from the characteristics of each project. However,
-maintainability still may be affected by the developers experience, coding
+maintainability may sitll be affected by the developers experience, coding
 style and software contribution polices which it is not evaluated in this
@@
 
-\Luis{maintainability still may be affected -> maintainability may sitll be affected}
-
 \subsection{External}
 
-\Luis{A primeira threat que falas é que os resultados podem nao generalizar
-para software fechado. Eu acho que deviamos ser mais diretos. Para alem disso
-tirava o "In contrast to the BCH benchmark,"}
-
-In contrast to the BCH benchmark, this study uses a dataset that only covers
+% \Luis{A primeira threat que falas é que os resultados podem nao generalizar
+% para software fechado. Eu acho que deviamos ser mais diretos.}\Sof{Diretos como?}
+
+This study uses a dataset that only covers
 open-source software to perform the maintainability evaluation. However, BCH
@@
 address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be properly representative of the population regarding to
-programming languages.
+the dataset may not be properly representative of the population regarding to programming languages.
 
@@
 
-\Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
-fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
+% \Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
+% fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
 
@@
 
-\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
-Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
-diferenca é que estamos a estudar maintainability issues noutro contexto.}
+% \Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
+% Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
+% diferenca é que estamos a estudar maintainability issues noutro contexto.}
 
@@
 
-\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
+% \Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
 
@@
 
-\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
+% \Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
 
@@
 
-\Luis{many directions such as -> many directions:}
-\Luis{evaluate the impact of programming languages of the
-maintainability of each refactoring pattern; - > evaluate the impact of each refactoring pattern on the maintainbility of each  }
-
-As future work, this study may be extended in many directions such as analyze which guidelines are
-more likely to affect security refactorings; evaluate the impact of programming languages of the
-maintainability of each refactoring pattern; investigate if there is any correlation between
+As future work, this study may be extended in many directions: analyze which guidelines are
+more likely to affect security refactorings; evaluate the impact of each refactoring pattern on software maintainbility; investigate if there is any correlation between
 software maintainability and projects popularity; expand our methodology with other software quality

### 2019-02-03 paper/ICPC19/icpc19.tex
joining two paragraphs
luis comments up

@@
 address \emph{Broken Authentication and Session Management} issues, \emph{Memory
-Leaks} and \emph{Denial-of-Service} attacks. There are already several tools that
-suggest security improvements using static analysis, dynamic analysis, and even
-artificial intelligence. With this study, we intend to highlight the need for tools
+Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight the need for tools
 to predict the impact of patches on maintainability~\cite{4724577} and documentation
@@
 
-% \Luis{A primeira threat que falas é que os resultados podem nao generalizar
-% para software fechado. Eu acho que deviamos ser mais diretos.}\Sof{Diretos como?}
+\Luis{A primeira threat que falas é que os resultados podem nao generalizar
+para software fechado. Eu acho que deviamos ser mais diretos.}\Sof{Diretos como?}
 
@@
 
-% \Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
-% fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
+\Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
+fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
 
@@
 
-% \Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
-% Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
-% diferenca é que estamos a estudar maintainability issues noutro contexto.}
+\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
+Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
+diferenca é que estamos a estudar maintainability issues noutro contexto.}
 
@@
 
-% \Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
+\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
 
@@
 
-% \Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
+\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
 

### 2019-02-03 paper/ICPC19/icpc19.tex
polishing ack
another issue fixed

@@
 buffer in line $37$. Thus, if the fragment was within the allowed size, the
-pending handshake message could consume $18MB$ ($16MB$+$2MB$). According to the
+pending handshake message could could use up to $18MB$ ($16MB$+$2MB$). According to the
 message commit, $10$ handshake messages are allowed, so the attacker could
@@
 supports that the maintainability of open-source software decreases for 4
-different patterns: Broken Autentication (p = 0.018), Denial-of-Service
-(p = 0.029), Memory Leak (p = 0.009) and Miscellaneous (p = $3.07$x$10^{-4}$).
+different patterns: \emph{Broken Autentication} (p = 0.018), \emph{Denial-of-Service}
+(p = 0.029), \emph{Memory Leak} (p = 0.009) and \emph{Miscellaneous} (p = $3.07$x$10^{-4}$).
 The results also returned statistical significance that evidence that maintainability
 is more likely to not suffer differences for open-source software after removing
-Components with Known Vulnerabilities (p = $7.96$x$10^{-4}$) and fixing Cross-Site
-Scripting Vulnerabilities (p = 0.001). Injection and Cross-Site Request Forgery
+\emph{Components with Known Vulnerabilities} (p = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site
+Scripting Vulnerabilities} (p = 0.001). \emph{Injection} and \emph{Cross-Site Request Forgery}
 produced more cases where the refactorings had a negative impact in the maintainability.
@@
 
-Projects with Broken Authentication flaws are the most affected ones by security
+Projects with \emph{Broken Authentication} flaws are the most affected ones by security
 refactorings. 55.3\% (21) of the refactorings had a negative impact in the maintainability
 while 35\% (11) had a positive impact and only 13.8\% (6) did not suffer any
-change. The results also show that fixing memory leaks affects the maintainability
+change. The results also show that fixing \emph{Memory Leak}s affects the maintainability
 of 51.3\% (41) of the cases and leaves only 13.8\% (11) of the cases unaffected.
 In 35\% (28) of the cases, maintainability increased. Refactoring vulnerabilities
-that allow denial-of-service attacks (e.g., CVE-2014-3506) has a negative impact
+that allow \emph{Denial-of-Dervice} attacks (e.g., CVE-2014-3506) has a negative impact
 of 47.5\% (19) in the maintainability of open-source software. 30.0\% (12) of
 the cases exhibit maintainability improvement while 22.5\% (9) show that overall
-maintainability does not suffer any change. Miscellaneous refactorings pattern
+maintainability does not suffer any change. \emph{Miscellaneous} refactorings pattern
 integrates several types of security flaws whose impact is of 43.1\% (84) negative
@@
 It is also important to notice that overall the maintainability of open-source
-software is not affected by Cross-Site Scripting vulnerabilities and by Using
-Components with Know Vulnerabilities. Cross-Site Scripting does not suffer any
+software is not affected by \emph{Cross-Site Scripting} vulnerabilities and by \emph{Using
+Components with Know Vulnerabilities}. \emph{Cross-Site Scripting} does not suffer any
 decrease in the maintainability of 55.3\% (63) of cases and 21.1\% (24) have a
 positive effect on it. Only 23.7\% (27) of the cases hinder the software
-maintainability. From 21 cases evaluated for Using Componentes with Known
-Vulnerabilities, only 4.8\% (1) has a positive effect in the maintainability
+maintainability. From 21 cases evaluated for \emph{Using Componentes with Known
+Vulnerabilities}, only 4.8\% (1) has a positive effect in the maintainability
 of software. 14.3\% (3) harms maintainability software and 81\% has no effect
@@
 \textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
-
+%
 We found that 40.2\% of the security refactoring have a negative impact on the
@@
 more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}.
-However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on
-software maintainability. We suspect that these values are the result of projects
+However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on software maintainability. We suspect that these values are the result of projects
 contributions policies. 9 out of 10 projects --- the equivalent to 31.8\% of the commits ---
@@
 \textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}} Although there are
-a few refactorings that harm software maintainability in the \emph{Cross-Site Scripting} and \emph{Using
-Components with Know Vulnerabilities}, overall maintainability does not suffer any changes for these patterns.
-For the remaining patterns, \emph{Injection} and \emph{Cross-Site Request Forgery}, results are not statistically
-significant.
+a few refactorings that harm software maintainability in the \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities}, overall maintainability does not suffer any changes for these patterns.
+For the remaining patterns, \emph{Injection} and \emph{Cross-Site Request Forgery}, results are not statistically significant.
 
@@
 function or changing the version of the component, respectively. Whereas,
-Broken Authentication and vulnerabilities such as the one presented in
+\emph{Broken Authentication} and vulnerabilities such as the one presented in
 Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to
@@
 This study also exhibits evidence that some patterns need more attention than
-others. Broken Autentication issues, Denial-of-Service attacks and Memory Leaks
+others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}
 endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following

### 2019-02-03 paper/ICPC19/icpc19.tex
toggle comments
fixing consume

@@
 
-\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{red} #1}\textbf{]]}}
-\newcommand{\Luis}[1]{\textbf{[Luis:[}{\color{green} #1}\textbf{]]}}
-\newcommand{\Rui}[1]{\textbf{[Rui:[}{\color{blue} #1}\textbf{]]}}
+%
+% Add comments in the text
+%
+\newboolean{showcomments}
+\setboolean{showcomments}{true}
+%\setboolean{showcomments}{false}
+
+\ifthenelse{\boolean{showcomments}}
+  {\newcommand{\nb}[3]{
+  {\color{#2}\small\fbox{\bfseries\sffamily\scriptsize#1}}
+  {\color{#2}\sffamily\small$\triangleright~$\textit{\small #3}$~\triangleleft$}
+  }
+  }
+  {\newcommand{\nb}[3]{}
+  }
+
+\newcommand\Sof[1]{\nb{Sofia}{red}{#1}}
+\newcommand\Luis[1]{\nb{Luis}{green}{#1}}
+\newcommand\Rui[1]{\nb{Rui}{blue}{#1}}
 
@@
 
-In the example (Listing~\ref{lst:vuln}), all functions receive as input the \textbf{handshake} represented by \texttt{s}.
-Every SSL/TLS connection begins with a handshake which is reponsible by the negotiation between two parties.
-The handshake determines what cryptographic algorithms will be used to encrypt communications, verify the server and establish a secure connection before beginning data transfer.
-In the \texttt{dtls1\_reassemble\_fragment} function the value of
-\texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the maximum
-handshake message size, but \texttt{msg\_len} is allocated to the fragment
-buffer in line $37$. Thus, if the fragment was within the allowed size, the
-pending handshake message could could use up to $18MB$ ($16MB$+$2MB$). According to the
-message commit, $10$ handshake messages are allowed, so the attacker could
-consume until $180MB$ per DTLS connection.
+In the example (Listing~\ref{lst:vuln}), all functions receive as input the
+\textbf{handshake} represented by \texttt{s}. Every SSL/TLS connection begins
+with a handshake which is reponsible by the negotiation between two parties. The
+handshake determines what cryptographic algorithms will be used to encrypt
+communications, verify the server and establish a secure connection before
+beginning data transfer. In the \texttt{dtls1\_reassemble\_fragment} function
+the value of \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the
+maximum handshake message size, but \texttt{msg\_len} is allocated to the
+fragment buffer in line $37$. Thus, if the fragment was within the allowed size,
+the pending handshake message could use up to $18$MB ($16$MB+$2$MB).
+According to the message commit, $10$ handshake messages are allowed. Hence,
+the attacker could potentially use $180$MB per DTLS connection.
 
@@
 \noindent
-We thank to the Software Improvement Group (SIG) team for all the support and
+We thank the Software Improvement Group (SIG) team for all the support and
 help in validating the methodology.

### 2019-02-03 paper/ICPC19/icpc19.tex
more fixes

--- 2019-02-03 paper/ICPC19/icpc19.tex
@@
 
-\newcommand{\Sof}[1]{\textbf{[Sofia:[}{\color{red} #1}\textbf{]]}}
-\newcommand{\Luis}[1]{\textbf{[Luis:[}{\color{green} #1}\textbf{]]}}
-\newcommand{\Rui}[1]{\textbf{[Rui:[}{\color{blue} #1}\textbf{]]}}
+%
+% Add comments in the text
+%
+\newboolean{showcomments}
+\setboolean{showcomments}{false}
+%\setboolean{showcomments}{false}
+
+\ifthenelse{\boolean{showcomments}}
+  {\newcommand{\nb}[3]{
+  {\color{#2}\small\fbox{\bfseries\sffamily\scriptsize#1}}
+  {\color{#2}\sffamily\small$\triangleright~$\textit{\small #3}$~\triangleleft$}
+  }
+  }
+  {\newcommand{\nb}[3]{}
+  }
+
+\newcommand\Sof[1]{\nb{Sofia}{red}{#1}}
+\newcommand\Luis[1]{\nb{Luis}{green}{#1}}
+\newcommand\Rui[1]{\nb{Rui}{blue}{#1}}
 
@@
  	\centering
- 	\includegraphics[width=0.55\textwidth]{figures/language.pdf}
+ 	\includegraphics[width=0.56\textwidth]{figures/language.pdf}
  	\caption{Maintainability difference by language after security refactorings}
@@
 also provide statistical significance for when security refactorings are performed
-in \emph{Java} (p = $3.61$x$10^{-2}$). \emph{Python} and the other languages
+in \emph{Java} (p = $3.61$x$10^{-2}$). \emph{Python} and the languages
 that integrate the \emph{Others} set contain more cases where the maintainability
@@
 
-Projects with \emph{Broken Authentication} flaws are the most affected ones by security
+Software with \emph{Broken Authentication} flaws are the most affected ones by security
 refactorings. 55.3\% (21) of the refactorings had a negative impact in the maintainability
-while 35\% (11) had a positive impact and only 13.8\% (6) did not suffer any
+while 35\% (11) yielded a positive impact and only 13.8\% (6) did not suffer any
 change. The results also show that fixing \emph{Memory Leak}s affects the maintainability
@@
 software is not affected by \emph{Cross-Site Scripting} vulnerabilities and by \emph{Using
-Components with Know Vulnerabilities}. \emph{Cross-Site Scripting} does not suffer any
+Components with Known Vulnerabilities}. \emph{Cross-Site Scripting} does not suffer any
 decrease in the maintainability of 55.3\% (63) of cases and 21.1\% (24) have a
 positive effect on it. Only 23.7\% (27) of the cases hinder the software
-maintainability. From 21 cases evaluated for \emph{Using Componentes with Known
+maintainability. From the 21 cases evaluated for \emph{Using Componentes with Known
 Vulnerabilities}, only 4.8\% (1) has a positive effect in the maintainability
@@
 \textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}} Although there are
-a few refactorings that harm software maintainability in the \emph{Cross-Site Scripting} and \emph{Using Components with Know Vulnerabilities}, overall maintainability does not suffer any changes for these patterns.
+a few refactorings that harm software maintainability in the \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities}, overall maintainability does not suffer any changes for these patterns.
 For the remaining patterns, \emph{Injection} and \emph{Cross-Site Request Forgery}, results are not statistically significant.
@@
 maintainability. One evidence of that is the fact of \emph{Cross-Site
-Scripting} and \emph{Using Components with Know Vulnerabilities} refactorings
+Scripting} and \emph{Using Components with Known Vulnerabilities} refactorings
 endure more cases without any impact in the maintainability. Typically, these

### 2019-02-03 paper/ICPC19/icpc19.tex
charts fixed
more fixes

@@
 of their software application. To answer the following two
-research questions, we use a dataset of security refactorings to measure the
+research questions, we use a dataset of security refactorings~\cite{Reis:2017:IJSSE} to measure the
 impact of the changes in the maintainability of open-source software.
@@
 
-Software with \emph{Broken Authentication} flaws are the most affected ones by security
-refactorings. 55.3\% (21) of the refactorings had a negative impact in the maintainability
-while 35\% (11) yielded a positive impact and only 13.8\% (6) did not suffer any
+Projects with \emph{Broken Authentication} flaws are the most affected ones by security
+refactorings: 55.3\% (21) of the refactorings had a negative impact in the maintainability
+while 35\% (11) yielded a positive impact, and only 13.8\% (6) did not suffer any
 change. The results also show that fixing \emph{Memory Leak}s affects the maintainability
@@
 In 35\% (28) of the cases, maintainability increased. Refactoring vulnerabilities
-that allow \emph{Denial-of-Dervice} attacks (e.g., CVE-2014-3506) has a negative impact
-of 47.5\% (19) in the maintainability of open-source software. 30.0\% (12) of
-the cases exhibit maintainability improvement while 22.5\% (9) show that overall
-maintainability does not suffer any change. \emph{Miscellaneous} refactorings pattern
-integrates several types of security flaws whose impact is of 43.1\% (84) negative
-in the maintainability of open-source sofware. 31.8\% (62) of the cases do not
-exhibit any change in maintainability and 25.1\% (49) of cases have a positive
-impact on the maintainability of open-source software.
+that allow \emph{Denial-of-Service} attacks (e.g., CVE-2014-3506) have a negative impact
+of 47.5\% (19) in the maintainability of open-source software while 30.0\% (12) of
+the cases exhibit maintainability improvement and 22.5\% (9) show that
+maintainability does not suffer any change. The \emph{Miscellaneous} refactorings pattern
+integrates several types of security flaws.
+These type of refactorings harm maintainability in 43.1\% (84) of the cases.
+In 31.8\% (62), there is no affect in maintainability and for the other 25.1\% (49) maintainability increases.
 
@@
 maintainability. From the 21 cases evaluated for \emph{Using Componentes with Known
-Vulnerabilities}, only 4.8\% (1) has a positive effect in the maintainability
-of software. 14.3\% (3) harms maintainability software and 81\% has no effect
+Vulnerabilities}, only 4.8\% (1) have a positive effect in the maintainability
+of software while 14.3\% (3) harms maintainability software, and 81\%(17) have no effect
 in the maintainability.
@@
  	\centering
- 	\includegraphics[width=0.55\textwidth]{figures/category.pdf}
+ 	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
  	\caption{Maintainability difference by pattern after security refactorings}
@@
 However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on software maintainability. We suspect that these values are the result of projects
-contributions policies. 9 out of 10 projects --- the equivalent to 31.8\% of the commits ---
+contributions policies. Nine out of ten projects --- the equivalent to 31.8\% of the commits ---
 of the top-10 of dataset projects with more contributors have restrict contribution

### 2019-02-03 paper/ICPC19/icpc19.tex
figura 5 fixed + example corrected

@@
 and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
-easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}. Compliance to the guidelines is derived from the SIG's
-industry benchmark which, according to official sources, consists of over 8
-billion lines of code in more than 180 different technologies. SIG analyzes
-around 15 million lines of code, per week.
+easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}. 
+% Compliance to the guidelines is derived from the SIG's
+% industry benchmark which, according to official sources, consists of over 8
+% billion lines of code in more than 180 different technologies. SIG analyzes
+% around 15 million lines of code, per week.
 
@@
 contains open-source implementations of the SSL and TLS cryptographic
-protocols.} to address a vulnerability where an hacker could cause a Denial-of-Service
+protocols.} to address a vulnerability where an attacker could cause a Denial-of-Service
 (DoS) attack by crafting DTLS handshake messages to trigger memory allocations
@@
 https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
-(Accessed on \today{})} by the OpenSLL developers to fix the vulnerability.
+(Accessed on \today{})} by the OpenSLL developers to patch the vulnerability.
 
@@
 
-In the example (Listing~\ref{lst:vuln}), all functions receive as input the
-\textbf{handshake} represented by \texttt{s}. Every SSL/TLS connection begins
-with a handshake which is reponsible by the negotiation between two parties. The
-handshake determines what cryptographic algorithms will be used to encrypt
-communications, verify the server and establish a secure connection before
-beginning data transfer. In the \texttt{dtls1\_reassemble\_fragment} function
-the value of \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the
-maximum handshake message size, but \texttt{msg\_len} is allocated to the
-fragment buffer in line $37$. Thus, if the fragment was within the allowed size,
-the pending handshake message could use up to $18$MB ($16$MB+$2$MB).
-According to the message commit, $10$ handshake messages are allowed. Hence,
-the attacker could potentially use $180$MB per DTLS connection.
-
-The code changes performed to fix the vulnerability were the following:
-
-~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to limit the number of bytes allowed in a DTLS handshake message,
-\texttt{s}. Checks if the maximum certificate list size
+Every SSL/TLS connection begins with a handshake which is responsible by the negotiation between two parties. 
+DTLS provides a mechanism for fragmenting a handshake message over a number of packets, each of which can be 
+transmitted separately. In the example (Listing~\ref{lst:vuln}), all functions receive as input the
+SSL session, \texttt{s}.
+In the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake message size \texttt{msg\_hdr->msg\_len}
+(line 13) and then against the maximum handshake message size allowed by DTLS1 for \texttt{s} (line 16-21). 
+However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line 37,
+when \texttt{msg\_len} is allocated to the fragment buffer, more memory than the allowed might be allocated.
+According to the message commit, $10$ handshake messages are allowed per DTLS connection.
+Hence, a remote attacker could leverage this issue to trigger
+memory allocations corresponding to large length values and lead the system to memory exhaustion that could end
+in a DoS attack. The code changes performed to fix the vulnerability were the following:
+
+~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to return the number of 
+maximum number of bytes allowed in a DTLS handshake message for \texttt{s}. The mininum is $16$KB, but may be greater 
+if the maximum certificate list size requires it which is set by the user. More specifically, it checks if the 
+maximum certificate list size
 (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
@@
 
-~\ref{lst:func5} and~\ref{lst:func4} add the check to ensure that the message
-length is never higher than the maxiumum allowed by the DTSL protocol in
-different functions.
-
-Although limiting the size of a handshake message seems an elementary problem to
+~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to ensure that the message 
+and the fragment length are never higher than the maximum allowed by DTSL. However, \ref{lst:func5} is
+the responsible for patching the vulnerability.
+
+Although checking the maximum size of a handshake message seems an elementary problem to
 repair, the example shows that there was a significant amount of changes
@@
 security, developers are also negatively impacting the maintainability
-of their software application. To answer the following two
+of their software applications. To answer the following two
 research questions, we use a dataset of security refactorings~\cite{Reis:2017:IJSSE} to measure the
@@
 	implemented functionalities related to authentication and session management,
-	allowing crackers to gain access to session tokens, passwords, keys and other
+	allowing attacker to gain access to session tokens, passwords, keys and other
 	sensitive data. \textit{Based on the OWASP Top 10 of 2013.}
@@
 	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and
-	management usually allow crackers to send forged HTTP requests including
+	management usually allow attackers to send forged HTTP requests including
 	authentication information from the victim to vulnerable web application.
@@
 
-	\item \textbf{Denial-of-Service} Security flaws that can allow a cracker to
+	\item \textbf{Denial-of-Service} Security flaws that can allow a attacker to
 	flood or crashing services. This attacks usually occur when the system
 	receives much traffic causing it to slow down and eventually stop. For
-	example, flaws allowing crackers to trigger memory allocations corresponding
+	example, flaws allowing attackers to trigger memory allocations corresponding
 	to large length values (Listing~\ref{lst:vuln}).
@@
 	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
-	allow crackers to submit untrusted data to web browsers through malicious
+	allow attackers to submit untrusted data to web browsers through malicious
 	scripts that can hijack the user sessions or redirect the user to malicious
@@
 	\item \textbf{Injection.} When developers do not keep untrusted data separate
-	from commands and queries. If a cracker sends a string that exploits the
+	from commands and queries. If a attacker sends a string that exploits the
 	syntax of the interpreter, then an injection attack is possible (e.g., SQL and

### 2019-02-03 paper/ICPC19/icpc19.tex
fixes

@@
 Hub} (BCH) is used to collect the maintanability reports of the refactorings of
-each project. Table~\ref{tab:guidelines} presents the 10 guidelines
-proposed by BCH's authors for delivering software that is easy to maintain~\cite{Visser:2016:OREILLY}. All guidelines are computed in the same evaluation by
-BCH. During each guideline evaluation, the tool determines the compliance towards
-one guideline by establishing limits for the percentage of code allowed to be in
-each of the $4$ risk severity levels (\emph{low risk}, \emph{medium risk}, \emph{high risk}
-and \emph{very high risk}). If the project does not violate the thresholds, then it
-is compliant with the guideline. These thresholds are calibrated by BCH using
-their own data/experience --- open-source and closed software systems. If a
-project is compliant with a guideline, it means that it is at least $65\%$
-better than the software used by BCH to calculate the thresholds\footnote{Check
-the answer to \emph{How can I adjust the threshold for passing/not passing a
-guideline?} at https://bettercodehub.com/docs/faq (Accessed on \today{})}.
+each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
+by BCH's authors for delivering software that is not difficult to
+maintain~\cite{Visser:2016:OREILLY}. During each guideline evaluation, the tool
+determines the compliance towards one guideline by establishing limits for the
+percentage of code allowed to be in each of the $4$ risk severity levels
+(\emph{low risk}, \emph{medium risk}, \emph{high risk} and \emph{very high
+risk}). If the project does not violate the thresholds, then it is compliant
+with the guideline. These thresholds are calibrated by BCH using their own
+data/experience --- open-source and closed software systems. If a project is
+compliant with a guideline, it means that it is at least $65\%$ better than the
+software used by BCH to calculate the thresholds\footnote{Check the answer to
+\emph{How can I adjust the threshold for passing/not passing a guideline?} at
+https://bettercodehub.com/docs/faq (Accessed on \today{})}.
 
@@
 code. BCH uses this metric to measure the number of branch points in the
-\emph{Write Simple Units of Code} guideline. The bar represents the top 30 of
-units that violates the guideline, sorted by severity, which is indicated by the
-colors of the checkboxes. The green bar represents the number of compliant
+\emph{Write Simple Units of Code} guideline. The bar at the bottom of the figure
+represents the top $30$ units that violates the guideline, sorted by severity,
+The violation severities are indicated using colors, and there is a legend to
+help interpreting the colors. The green bar represents the number of compliant
 branch points per unit (\emph{at most 5}), i.e., the units are compliant with

### 2019-02-03 paper/ICPC19/icpc19.tex
fix code

--- 2019-02-03 paper/ICPC19/icpc19.tex
@@
 \newboolean{showcomments}
-\setboolean{showcomments}{false}
+\setboolean{showcomments}{true}
 %\setboolean{showcomments}{false}
@@
 There are security flaws that are more difficult to refactor than others. For
-instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability with simple function to sanitize data\footnote{More information about \texttt{htmlentities()} at http://php.net/manual/en/function.htmlentities.php (Accessed on \today{})}.
+instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability
+since usually the last one is fixed within a line or two of code, as seen in Figure~\ref{lst:fix}. 
 Knowing which patterns are more likely to increase maintainability
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix},language=php]
+\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix}]
  <ul>

### 2019-02-03 paper/ICPC19/icpc19.tex
adding .out to gitignore
another fix

@@
 There are security flaws that are more difficult to refactor than others. For
-instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability
-since usually the last one is fixed within a line or two of code, as seen in Figure~\ref{lst:fix}. 
+instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability since usually the last one is can be fixed without adding new lines of code, as seen in Figure~\ref{lst:fix}.
+The developer added the function \emph{htmlentities} to escape the data inserted by the variable \emph{\$\_['file']}.
 Knowing which patterns are more likely to increase maintainability

### 2019-02-03 paper/ICPC19/icpc19.tex
luis comments added + fixzinho na discussao

@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
- 	\caption{Distribution of Security Refactorings per Language}
+ 	\caption{Distribution of Security Refactorings per Language\Luis{Consegues ordenar estas linguagens da mesma forma que estão ordenadas na figura7}}
 	\label{fig:lang}
@@
 is more likely to not suffer differences for open-source software after removing
-\emph{Components with Known Vulnerabilities} (p = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site
+\emph{Components with Known Vulnerabilities}\Luis{o resultado e bem diferente dos regular commits.
+este refactoring tem alguma particularidade para deixar a maintainability intacta?
+convém falar sobre isso na discussão} (p = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site
 Scripting Vulnerabilities} (p = 0.001). \emph{Injection} and \emph{Cross-Site Request Forgery}
@@
 endure more cases without any impact in the maintainability. Typically, these
-type of vulnerabilities are one line fixes --- adding an \texttt{escape}
-function or changing the version of the component, respectively. Whereas,
-\emph{Broken Authentication} and vulnerabilities such as the one presented in
+types of vulnerabilities do not need extra lines to be fixed, as showed in
+Listing~\ref{lst:fix}. 
+Whereas, \emph{Broken Authentication} and vulnerabilities such as the one presented in
 Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to

### 2019-02-03 paper/ICPC19/icpc19.tex
luis fixes

@@
 \Luis{A primeira threat que falas é que os resultados podem nao generalizar
-para software fechado. Eu acho que deviamos ser mais diretos.}\Sof{Diretos como?}
-
-This study uses a dataset that only covers
-open-source software to perform the maintainability evaluation. However, BCH
-thresholds are calculated from a set of both open and closed software.
+para software fechado. Eu acho que deviamos ser mais diretos.}
+\Sof{Melhor?}
+
+We use a dataset that only comprises refactorings of open-source software.
+However, the methodology requires to access data that is not publicly available.
+Thus, our findings may not extent to non-open source software.
+
 Different programming languages may require different coding practices to
@@
 \section{Related Work}\label{sec:rw}
-
-\Luis{pode ser uma cena só minha mas acho que fica melhor em vez de comecar
-fraseds com "And," usar outros conetores tipo "Besides," "In addition,", etc. œ}
 
@@
 The differences of maintainability between groups of refactored elements and
-non-refactored were analyzed. And, the authors conclude that the source
+non-refactored were analyzed. In addition, the authors conclude that the source
 code elements subjected to refactorings had significantly lower maintainability
@@
 a high correlation between maintainability aspects and proneness to changes and faults.
-The present work is based on the 10 guidelines proposed
-by the Software Improvement Group (SIG) and focuses solely on evaluating
+The present work uses a model proposed by BCH and focuses solely on evaluating
 the impact of patching vulnerabilities on software maintainability.
@@
 Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
-diferenca é que estamos a estudar maintainability issues noutro contexto.}
+diferenca é que estamos a estudar maintainability issues noutro contexto.}\Sof{Melhor?}
 
@@
 software quality. The present work studies how security refactorings affect software quality
-based on the code maintainability analysis.
-
-\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}
+based on the code maintainability analysis and provides some evidence that the programming language has impact on maintainability.
+
+\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}\Sof{Acho que esta!}
 
@@
 \Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
+\Sof{Melhor?}
 
@@
 endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
-the implications of our work, we draw attention for the need of secure security
+the implications of our work, we draw attention for the need of trustfull security
 frameworks, code best practices documentation to produce safe and maintainable

### 2019-02-03 paper/ICPC19/icpc19.tex
393 and grammarly pass

@@
 security on the maintainability of open-source software. Maintainability is
-measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
-dataset including 607 security commits. Results show that fixing software
+measured based on the \emph{Better Code Hub}'s model of $10$ guidelines using a
+dataset including $607$ security commits. Results show that fixing software
 vulnerabilities hinders the maintainability of open-source software. This is
@@
 Software quality is important because it is ultimately related to the overall
-cost for developing, extending and maintaing software applications. Software
+cost of developing, extending and maintaining software applications. Software
 quality characteristics include, but are not limited to, functional correctness,
@@
 systems and exploit their resources.
-As an example, in the last week of January 2019, a severe vulnerability in Apple
+As an example, in the last week of January $2019$, a severe vulnerability in Apple
 FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
@@
 
-In 2011, the International Organization for Standardization (ISO) issued an
+In $2011$, the International Organization for Standardization (ISO) issued an
 update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
 \emph{Security} as one of the main software product quality characteristics.
-Unmaintainable code is hard to test, analyse and re-use, and thus more difficult
+Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
 to find vulnerabilities and defects. Therefore, writing maintainable code
@@
 maintainability when trying to improve software safety? We suspect that some of
-the patches may have a negative impact in the software maintainability and
+the patches may have a negative impact on the software maintainability and
 possibly even be the cause of the introduction of new vulnerabilities.
@@
 Application Security Verification Criteria~\cite{oswap:2009} which is focused only
-on web applications, and a model proposed by Xu et al. (2013)~\cite{6616351} for
-raiting software security which we believe it was one of the first steps taken
+on web applications, and a model proposed by Xu et al. ($2013$)~\cite{6616351} for
+rating software security which we believe it was one of the first steps taken
 by SIG to introduce security on their maintainability model.
 
-BCH compiles the ISO/IEC 25010 into a set of 10 guidelines that are the
+BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
 base of BCH's evaluation. Two examples are the
@@
 easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
-% Compliance to the guidelines is derived from the SIG's
-% industry benchmark which, according to official sources, consists of over 8
-% billion lines of code in more than 180 different technologies. SIG analyzes
-% around 15 million lines of code, per week.
 
@@
 between refactoring source code to patch disclosed vulnerabilities while keeping
-(or even improving) the maintainability of software systems. Hedegus et al. (2018)~\cite{HEGEDUS2018313},
+(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313},
 analyzed the difference of maintainability between
@@
 In this paper, we present the results of our analysis on the maintainability of
-607 security changes collected from open-source software, from different
-programming languages. The main contributions of the present work is:
+$607$ security changes collected from open-source software, from different
+programming languages. The main contributions of the present work are:
 \begin{itemize}
@@
 As an example, consider the refactoring of the Datagram Transport Layer Security
-(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a tookit that
+(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a toolkit that
 contains open-source implementations of the SSL and TLS cryptographic
@@
 corresponding to large length values. This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-2014-3506\footnote{CVE-2014-3506
+Vulnerabilities and Exposures dictionary as CVE-$2014$-$3506$\footnote{CVE-$2014$-$3506$
 details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-evaluation. The following snippet presents the changes performend on the
-\emph{ssl/d1\_both.c} file\footnote{CVE-2014-3506 fix available  at
+evaluation. The following snippet presents the changes performed on the
+\emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
 https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
-(Accessed on \today{})} by the OpenSLL developers to patch the vulnerability.
+(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability.
 
@@
 \begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-\\CVE-2014-3506 vulnerability},label={lst:vuln}]
+\\CVE-$2014$-$3506$ vulnerability},label={lst:vuln}]
 + static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
@@
 In the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake message size \texttt{msg\_hdr->msg\_len}
-(line 13) and then against the maximum handshake message size allowed by DTLS1 for \texttt{s} (line 16-21).
-However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line 37,
+(line $13$) and then against the maximum handshake message size allowed by DTLS1 for \texttt{s} (line $16$-$21$).
+However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line $37$,
 when \texttt{msg\_len} is allocated to the fragment buffer, more memory than the allowed might be allocated.
@@
 ~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to return the number of
-maximum number of bytes allowed in a DTLS handshake message for \texttt{s}. The mininum is $16$KB, but may be greater
+maximum number of bytes allowed in a DTLS handshake message for \texttt{s}. The minimum is $16$KB but may be greater
 if the maximum certificate list size requires it which is set by the user. More specifically, it checks if the
@@
 (\texttt{max\_len}). If that condition holds, the function returns the size of
-the certificate list, otherwise it returns the maximum length. This function replaces lines $18$-$21$.
+the certificate list, otherwise, it returns the maximum length. This function replaces lines $18$-$21$.
 
 ~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to ensure that the message
-and the fragment length are never higher than the maximum allowed by DTSL. However, \ref{lst:func5} is
-the responsible for patching the vulnerability.
+and the fragment length are never higher than the maximum allowed by DTSL. However, \ref{lst:func5} is responsible for patching the vulnerability.
 
@@
 maintainability of open-source software. Our suspicion is that developers tend
-to introduce technical debt in their software when refactoring the codebose to
+to introduce technical debt in their software when refactoring the codebase to
 address a security flaw because they tend to choose the easiest path to solve
@@
 The methodology used to measure the impact of security refactorings on the
-maintainability of open-source software is discussed this section. The methodology
+maintainability of open-source software is discussed in this section. The methodology
 comprehends the following steps, as illustrated in Figure~\ref{fig:met}:
@@
 open-source software, we use a dataset of security flaws which is the outcome of
-mining $94$ GitHub projects. Reis and Abreu (2017) mined open-source
+mining $94$ GitHub projects. Reis and Abreu ($2017$) mined open-source
 software aiming at the extraction of real examples --- created by real
@@
 dataset with $716$ test cases for $16$ security patterns. Some of the patterns
-are based in the OWASP Top 10 of 2013~\cite{oswap:2013} and OWASP Top 10 of
-2017~\cite{oswap:2017}. Each test case of the
+are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
+$2017$~\cite{oswap:2017}. Each test case of the
 dataset is a triplet: the commit before the refactoring, the commit responsible
@@
 language support and project size. The final dataset used in this paper contains
-$607$ security refactorings, labelled as one of the following categories (we
+$607$ security refactorings labeled as one of the following categories (we
 have decided to change the label to \textit{Miscellaneous} of the patterns in
@@
     versions - usually known and disclosed somewhere on the Internet.
-	\textit{Based on the OWASP Top 10 of 2017.}
-
-	\item \textbf{Broken Authentication \& Session Management.} Uncorrectly
-	implemented functionalities related to authentication and session management,
-	allowing attacker to gain access to session tokens, passwords, keys and other
-	sensitive data. \textit{Based on the OWASP Top 10 of 2013.}
-
-	\item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and
-	management usually allow attackers to send forged HTTP requests including
-	authentication information from the victim to vulnerable web application.
-	\textit{Based on the OWASP Top 10 of 2013.}
-
-	\item \textbf{Denial-of-Service} Security flaws that can allow a attacker to
-	flood or crashing services. This attacks usually occur when the system
-	receives much traffic causing it to slow down and eventually stop. For
+    \textit{Based on the OWASP Top $10$ of $2017$.}
+
+    \item \textbf{Broken Authentication \& Session Management.} Incorrectly implemented functionalities related to authentication and session management,
+    allowing an attacker to gain access to session tokens, passwords, keys, and other sensitive data. \textit{Based on the OWASP Top $10$ of $2013$.}
+
+    \item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow attackers to send forged HTTP requests including authentication information from the victim to the vulnerable web application.
+    \textit{Based on the OWASP Top $10$ of $2013$.}
+
+    \item \textbf{Denial-of-Service} Security flaws that can allow an attacker to flood or crashing services.
+This attacks usually occur when the system receives much traffic causing it to slow down and eventually stop.
+For
     example, flaws allowing attackers to trigger memory allocations corresponding
@@
 
-	\item \textbf{Miscellaneous} This pattern comprises several others categories
+    \item \textbf{Miscellaneous} This pattern comprises several other categories
     of security refactorings (e.g., \textit{path traversal and buffer overflow}).
     It contains several other security refactorings that do not have their own
-	pattern yet. The patterns not satisfying the size requirement of more than 20
+    pattern yet. The patterns not satisfying the size requirement of more than $20$
     refactorings are contemplated here: \textit{Resource Leak},
@@
 
-	\item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
-	allow attackers to submit untrusted data to web browsers through malicious
-	scripts that can hijack the user sessions or redirect the user to malicious
-	sites. \textit{Based on the OWASP Top 10 of 2017.}
-
-	\item \textbf{Injection.} When developers do not keep untrusted data separate
-	from commands and queries. If a attacker sends a string that exploits the
+    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow attackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites.
+\textit{Based on the OWASP Top $10$ of $2017$.}
+
+    \item \textbf{Injection.} When developers do not keep untrusted data separate from commands and queries. If an attacker sends a string that exploits the
     syntax of the interpreter, then an injection attack is possible (e.g., SQL and
-	LDAP injection). \textit{Based on the OWASP Top 10 of 2017.}
+    LDAP injection). \textit{Based on the OWASP Top $10$ of $2017$.}
 
@@
 
-The dataset contemplates security flaws for more than 13 different languages
-being PHP (39\%) and C (24\%) the most prevalent ones; see Figure~\ref{fig:lang}.
+The dataset contemplates security flaws for more than $13$ different languages
+being PHP ($39\%$) and C ($24\%$) the most prevalent ones; see Figure~\ref{fig:lang}.
 To classify the applications in the dataset, we use a taxonomy for open-source
@@
 and frameworks} ($20$); \textit{System Software} ($18$), software that provides
-services and infrustructures (e.g., operating systems, servers and databases);
+services and infrastructures (e.g., operating systems, servers and databases);
 \textit{Software Tool} ($13$), software that supports development (e.g.,
@@
 We analyze the maintainability of regular commits
-(i.e., commits not related with security fixes) and use them as a baseline.
+(i.e., commits not related to security fixes) and use them as a baseline.
 
@@
 that particular project. We originate the regular commits from the security
-commits to ensure that differences in maintainability are not consequence of
+commits to ensure that differences in maintainability are not a consequence of
 characteristics of different projects.
@@
 As said before, the web-based source code analysis service \emph{Better Code
-Hub} (BCH) is used to collect the maintanability reports of the refactorings of
+Hub} (BCH) is used to collect the maintainability reports of the refactorings of
 each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
@@
 Figure~\ref{fig:bchrep} shows an example of the report
-provided by BCH for a project after finishing its evaluation. The examples
-refers to the OpenSSL CVE-2014-3506 vulnerability refactoring ---
+provided by BCH for a project after finishing its evaluation. The example
+refers to the OpenSSL CVE-$2014$-$3506$ vulnerability refactoring ---
 as described as motivating example in Section~\ref{sec:motivation}. This
-version of OpenSSL only complies with 1 out of 10 guidelines: \emph{Write
+version of OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write
 Clean Code}.
@@
 \midrule
-\textbf{Write Short Units of Code} & Limit code units to 15 LOCs because smaller
+\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
  units are easier to understand, reuse and test them\\\midrule
-\textbf{Write Simple Units of Code} & Limit branch points to 4 per unit because
+\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
 it makes units easier to test and modify \\\midrule
@@
 \textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
-4 because it makes units easier to understand and reuse\\\midrule
+$4$ because it makes units easier to understand and reuse\\\midrule
 \textbf{Separate Concerns in Modules} & Avoid large modules because changes in
@@
 \textbf{Couple Architecture Components Loosely} & Minimize the amount of code
-within modules that is exposed to modules in other components\\\midrule
+within modules that are exposed to modules in other components\\\midrule
 \textbf{Keep Architecture Components Balanced} & Balancing the number of
@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/bch_report.png}
- 	\caption{Maintainability report of OpenSSL's CVE-2014-3506 vulnerability
+ 	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
 refactoring for the guideline \emph{Write Simple Units of Code} provided by
@@
 \emph{Write Simple Units of Code} guideline. The bar at the bottom of the figure
-represents the top $30$ units that violates the guideline, sorted by severity,
+represents the top $30$ units that violate the guideline, sorted by severity,
 The violation severities are indicated using colors, and there is a legend to
-help interpreting the colors. The green bar represents the number of compliant
-branch points per unit (\emph{at most 5}), i.e., the units are compliant with
+help to interpret the colors. The green bar represents the number of compliant
+branch points per unit (\emph{at most $5$}), i.e., the units are compliant with
 ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
-not comply with medium (\emph{above 5}), high (\emph{above 10}) and very high
-(\emph{above 25}) severity levels. In the bar, there are marks that pinpoint the
+not comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
+(\emph{above $25$}) severity levels. In the bar, there are marks that pinpoint the
 compliance thresholds for each severity level. If the green mark is somewhere
@@
 that do not comply with the guideline at the given severity level $l$ and $w(l)$
-is the weight factor to boost the impact of non-compliant lines in comparison to
+is the weight factor to heighten the impact of non-compliant lines in comparison to
 compliant lines. Finally, the term $w(l)$ is calculated as follows:
@@
 difference between groups is zero, the observations are discarded. Thus, we use
-a more robust approach of the wilcoxon test, the Wilcoxon test by Pratt which
+a more robust approach to the Wilcoxon test, the Wilcoxon test by Pratt which
 provides an alternative that incorporates the cases where the maintainability
@@
 
-In total, this study evaluates $607$ security and baseline commits from 94
-different open-source projects distributed among 5 different main types of
+In total, this study evaluates $607$ security and baseline commits from $94$
+different open-source projects distributed among $5$ different main types of
 software, Figure~\ref{fig:domain}. The following section presents the results
@@
 Regarding security commits results, we observe that maintainability decreases
-in 40.20\% (244), stays equal in 30.5\% (185) and increases in 29.33\% (178)
+in $40.20\%$ ($244$), stays equal in $30.5\%$ ($185$) and increases in $29.33\%$ ($178$)
 of the cases after security refactorings being applied. The resulting p-value
@@
 
-For regular commits, we also observe that maintainability decreases in 41.35\%
-(251) of the cases after regular refactorings being applyed. But in contrast
-to security commits, it increases in 58.16\% (353) and stays equal in 0.5\%
-(3) of the cases. The resulting p-value of the Wilcoxon signed-rank test is
+For regular commits, we also observe that maintainability decreases in $41.35\%$
+($251$) of the cases after regular refactorings being applyed. But in contrast
+to security commits, it increases in $58.16\%$ ($353$) and stays equal in $0.5\%$
+($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank test is
 $1.54$x$10^{-6}$. Since the p-value is below the significance level of $0.05$,
@@
 one of the requirements of the Wilcoxon signed-rank test is that the input sample
-has to have more than $20$ measurements. The other languages were integrated in the
+has to have more than $20$ measurements. The other languages were integrated into the
 \emph{Others} set. The results of the Wilcoxon signed-rank test by pattern returned
 statistical evidence which supports that the maintainability of open-source software
-decreases for security refactorings in 3 different languages: \emph{C}
-(p = $5.72$x$10^{-9}$), Ruby (p = 0.057) and PHP (p = $3.54$x$10^{-6}$). The results
+decreases for security refactorings in $3$ different languages: \emph{C}
+(p = $5.72$x$10^{-9}$), Ruby (p = $0.057$) and PHP (p = $3.54$x$10^{-6}$). The results
 also provide statistical significance for when security refactorings are performed
@@
 Wilcoxon signed-rank test by pattern returned statistical evidence which
-supports that the maintainability of open-source software decreases for 4
-different patterns: \emph{Broken Autentication} (p = 0.018), \emph{Denial-of-Service}
-(p = 0.029), \emph{Memory Leak} (p = 0.009) and \emph{Miscellaneous} (p = $3.07$x$10^{-4}$).
+supports that the maintainability of open-source software decreases for $4$
+different patterns: \emph{Broken Autentication} (p = $0.018$), \emph{Denial-of-Service}
+(p = $0.029$), \emph{Memory Leak} (p = $0.009$) and \emph{Miscellaneous} (p = $3.07$x$10^{-4}$).
 The results also returned statistical significance that evidence that maintainability
@@
 convém falar sobre isso na discussão} (p = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site
-Scripting Vulnerabilities} (p = 0.001). \emph{Injection} and \emph{Cross-Site Request Forgery}
+Scripting Vulnerabilities} (p = $0.001$). \emph{Injection} and \emph{Cross-Site Request Forgery}
 produced more cases where the refactorings had a negative impact in the maintainability.
@@
 Projects with \emph{Broken Authentication} flaws are the most affected ones by security
-refactorings: 55.3\% (21) of the refactorings had a negative impact in the maintainability
-while 35\% (11) yielded a positive impact, and only 13.8\% (6) did not suffer any
+refactorings: $55.3\%$ ($21$) of the refactorings had a negative impact in the maintainability
+while $35.0\%$ ($11$) yielded a positive impact, and only $13.8\%$ ($6$) did not suffer any
 change. The results also show that fixing \emph{Memory Leak}s affects the maintainability
-of 51.3\% (41) of the cases and leaves only 13.8\% (11) of the cases unaffected.
-In 35\% (28) of the cases, maintainability increased. Refactoring vulnerabilities
-that allow \emph{Denial-of-Service} attacks (e.g., CVE-2014-3506) have a negative impact
-of 47.5\% (19) in the maintainability of open-source software while 30.0\% (12) of
-the cases exhibit maintainability improvement and 22.5\% (9) show that
+of $51.3\%$ ($41$) of the cases and leaves only $13.8\%$ ($11$) of the cases unaffected.
+In $35\%$ ($28$) of the cases, maintainability increased. Refactoring vulnerabilities
+that allow \emph{Denial-of-Service} attacks (e.g., CVE-$2014$-$3506$) have a negative impact
+of $47.5\%$ ($19$) in the maintainability of open-source software while $30.0\%$ ($12$) of
+the cases exhibit maintainability improvement and $22.5\%$ ($9$) show that
 maintainability does not suffer any change. The \emph{Miscellaneous} refactorings pattern
 integrates several types of security flaws.
-These type of refactorings harm maintainability in 43.1\% (84) of the cases.
-In 31.8\% (62), there is no affect in maintainability and for the other 25.1\% (49) maintainability increases.
+These type of refactorings harm maintainability in $43.1\%$ ($84$) of the cases.
+In $31.8\%$ ($62$), there is no effect in maintainability and for the other $25.1\%$ ($49$) maintainability increases.
 
@@
 Components with Known Vulnerabilities}. \emph{Cross-Site Scripting} does not suffer any
-decrease in the maintainability of 55.3\% (63) of cases and 21.1\% (24) have a
-positive effect on it. Only 23.7\% (27) of the cases hinder the software
-maintainability. From the 21 cases evaluated for \emph{Using Componentes with Known
-Vulnerabilities}, only 4.8\% (1) have a positive effect in the maintainability
-of software while 14.3\% (3) harms maintainability software, and 81\%(17) have no effect
+decrease in the maintainability of $55.3\%$ ($63$) of cases and $21.1\%$ ($24$) have a
+positive effect on it. Only $23.7\%$ ($27$) of the cases hinder the software
+maintainability. From the $21$ cases evaluated for \emph{Using Componentes with Known
+Vulnerabilities}, only $4.8\%$ ($1$) have a positive effect in the maintainability
+of software while $14.3\%$ ($3$) harms maintainability software, and $81\%$ ($17$) have no effect
 in the maintainability.
@@
 %
-We found that 40.2\% of the security refactoring have a negative impact on the
+We found that $40.2\%$ of the security refactoring have a negative impact on the
 maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
@@
 more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}.
-However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than 50\% on software maintainability. We suspect that these values are the result of projects
-contributions policies. Nine out of ten projects --- the equivalent to 31.8\% of the commits ---
-of the top-10 of dataset projects with more contributors have restrict contribution
+However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than $50\%$ on software maintainability. We suspect that these values are the result of projects
+contributions policies. Nine out of ten projects --- the equivalent to $31.8\%$ of the commits ---
+of the top-$10$ of dataset projects with more contributors have restrict contribution
 policies regarding code standards. Some examples are the \texttt{laravel/framework},
@@
 endure more cases without any impact in the maintainability. Typically, these
-types of vulnerabilities do not need extra lines to be fixed, as showed in
+types of vulnerabilities do not need extra lines to be fixed, as shown in
 Listing~\ref{lst:fix}. 
@@
 fix (several code lines deleted and added). Another example is the Apple
-authentication flaw discovered in 2017 (CVE-2017-13872)\footnote{CVE-2017-13872
+authentication flaw discovered in $2017$ (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$
 details available at https://support.apple.com/en-us/HT208315 (Accessed on
-\today{})}, where any user could login as root with an empty password. Patrick
+\today{})}, where any user could log in as root with an empty password. Patrick
 Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You
@@
 \emph{Cross-Site Scripting} and \emph{Using Components with Known
-Vulnerabilities} should take part of the developers preoccupations along with
+Vulnerabilities} should take part of the developers' preoccupations along with
 the other patterns that explicitly hinder maintainability because despite their
@@
 hinder software maintainability. Thus, it is important to provide tool support
-to developers to help them applying these patterns without harming software
+to developers to help them apply these patterns without harming software
 maintainability.
@@
 %
-The following section presents the potential threats to validity of this study.
+The following section presents the potential threats to the validity of this study.
 %
@@
 standards. However, BCH does have a representative benchmark of closed and open-source
-software projects that is under the calculation of the thresholds used in the computation
-of each maintainability guideline (Table~\ref{tab:guidelines})~\cite{Visser:2016:OREILLY,
-Baggen2012}.
+software projects that are under the calculation of the thresholds used in the computation
+of each maintainability guideline (Table~\ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
 
@@
 was collected based on the messages of GitHub commits produced by project developers
-to classify the changes perfomed by them to fix security flaws. This approach
+to classify the changes performed by them to fix security flaws. This approach
 discards security refactorings that were not explicit in commits messages.
@@
 that may result from the characteristics of each project. However,
-maintainability may sitll be affected by the developers experience, coding
-style and software contribution polices which it is not evaluated in this
-study. Furthermore, this evaluation considers that 607 regular commits are
+maintainability may still be affected by the developers' experience, coding
+style and software contribution policies which is not evaluated in this
+study. Furthermore, this evaluation considers that $607$ regular commits are
 enough to alleviate random irregularities in the maintainability differences of
@@
 However, the methodology requires to access data that is not publicly available.
-Thus, our findings may not extent to non-open source software.
+Thus, our findings may not extend to non-open source software.
 
@@
 address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be properly representative of the population regarding to programming languages.
+the dataset may not be properly representative of the population regarding programming languages.
 
@@
 code elements subjected to refactorings had significantly lower maintainability
-than the elements not affected. Studying the evolution of maintaninability issues during the development of Android apps, Malavolta et al. (2018)~\cite{8530041} discovered that maintainability
-decreases over time. Palomba et al. (2018)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits
-proof that code smells should be carefully monitored by programmers, since there is
+than the elements not affected. Studying the evolution of maintainability issues during the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041} discovered that maintainability
+decreases over time. Palomba et al. ($2018$)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits
+proof that code smells should be carefully monitored by programmers since there is
 a high correlation between maintainability aspects and proneness to changes and faults.
@@
 
-\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
+\Luis{eu nao acho que usar $10$ guidelines proposed by the Software Improvement
 Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
@@
 Researchers investigated the relationship between design patterns and
-software maintainability~\cite{10.1007/978-3-642-35267-6-18}. 300 revisions
+software maintainability~\cite{10.1007/978-3-642-35267-6-18}. $300$ revisions
 of \texttt{JHotDraw} were analyzed to conclude that every introduced pattern
@@
 2019arXiv190110220B}. The first study shows that some programming languages are more buggy than others, i.e.,
-language design has impact on software quality. However, the authors of the second
+language design has an impact on software quality. However, the authors of the second
 study were not able to reproduce the first one and did not obtain any evidence about the
@@
 software quality. The present work studies how security refactorings affect software quality
-based on the code maintainability analysis and provides some evidence that the programming language has impact on maintainability.
+based on the code maintainability analysis and provides some evidence that programming language have impact on maintainability.
 
@@
 
-This work presents an empirical study on the impact of 607 security refactorings on
-the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
+This work presents an empirical study on the impact of $607$ security refactorings on
+the maintainability of $94$ open-source projects. We leveraged Better Code Hub reports
 of each project version to calculate maintainability based on a model proposed in
@@
 software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
-this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
-overall maintainability increases (50.16\%). However, there is still a very considerable amout of
-cases that harm maintaniability (41.35\%).
+this evidence in $40.2\%$ of the studied cases. Regarding regular commits, it seems that
+overall maintainability increases ($50.16\%$). However, there is still a very considerable amount of
+cases that harm maintainability ($41.35\%$).
 
@@
 others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}
-endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
-the implications of our work, we draw attention for the need of trustfull security
+endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$, respectively. Following
+the implications of our work, we draw attention for the need of trustful security
 frameworks, code best practices documentation to produce safe and maintainable
@@
 As future work, this study may be extended in many directions: analyze which guidelines are
-more likely to affect security refactorings; evaluate the impact of each refactoring pattern on software maintainbility; investigate if there is any correlation between
+more likely to affect security refactorings; evaluate the impact of each refactoring pattern on software maintainability; investigate if there is any correlation between
 software maintainability and projects popularity; expand our methodology with other software quality

### 2019-02-03 paper/ICPC19/icpc19.tex
fixing commits; safe --> secure

--- 2019-02-03 paper/ICPC19/icpc19.tex
@@
 open-source software maintainability. Are developers affecting software
-maintainability when trying to improve software safety? We suspect that some of
+maintainability when trying to improve software security? We suspect that some of
 the patches may have a negative impact in the software maintainability and
@@
 (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
-project. There are other well-known standards and models that have been proposed to increase software safety:
+project. There are other well-known standards and models that have been proposed to increase software security:
 Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
@@
 new vulnerabilities. This study exhibits evidence that developers may have to
-reduce maintainability for the sake of safety. We understand that developers
+reduce maintainability for the sake of security. We understand that developers
 should be able to produce secure code without affecting the maintainability of
@@
 	trust them blindly and check defaults.
-	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
+	\item \textbf{Documentation} of software should provide developers with the best practices to implement secure and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
 	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
@@
 
-\Luis{A primeira threat que falas é que os resultados podem nao generalizar
-para software fechado. Eu acho que deviamos ser mais diretos.}
-\Sof{Melhor?}
-
-We use a dataset that only comprises refactorings of open-source software.
-However, the methodology requires to access data that is not publicly available.
-Thus, our findings may not extent to non-open source software.
+We use a dataset that only comprises refactorings of open-source software. As
+such, our findings may not generalize to non-open source software.
 
 Different programming languages may require different coding practices to
-address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be properly representative of the population regarding to programming languages.
+address software security. The dataset comprises more commits in C and PHP, i.e.,
+the dataset may not be properly representative of the population regarding to
+programming languages.
 
@@
 
-\Luis{eu nao acho que usar 10 guidelines proposed by the Software Improvement
-Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
-diferenca é que estamos a estudar maintainability issues noutro contexto.}\Sof{Melhor?}
-
 Researchers investigated the relationship between design patterns and
@@
 
-Other studies investigated the effect of programming languages in software quality~\cite{Ray:2017:LSP:3144574.3126905,
-2019arXiv190110220B}. The first study shows that some programming languages are more buggy than others, i.e.,
-language design has impact on software quality. However, the authors of the second
-study were not able to reproduce the first one and did not obtain any evidence about the
-impact of language design. Both have used the number of defects as an indicator of
-software quality. The present work studies how security refactorings affect software quality
-based on the code maintainability analysis and provides some evidence that the programming language has impact on maintainability.
-
-\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}\Sof{Acho que esta!}
+Other studies investigated the effect of programming languages in software
+quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
+study shows that there are programming languages that are more prone to bugs
+than others. However, the authors of the second study~\cite{2019arXiv190110220B}
+were not able to reproduce the first one and did not obtain any evidence about
+the impact of language design. Both have used the number of defects as an
+indicator of software quality. The present work studies how security
+refactorings affect software quality based on the code maintainability analysis
+and provides some evidence that the programming language has impact on
+maintainability.
 
@@
 
-This work presents an empirical study on the impact of 607 security refactorings on
-the maintainability of 94 open-source projects. We leveraged Better Code Hub reports
-of each project version to calculate maintainability based on a model proposed in
-previous work. Security refactorings significantly affect the software maintainability of open-source
-software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
-this evidence in 40.2\% of the studied cases. Regarding regular commits, it seems that
-overall maintainability increases (50.16\%). However, there is still a very considerable amout of
-cases that harm maintaniability (41.35\%).
-
-\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
-\Sof{Melhor?}
-
-This study also exhibits evidence that some patterns need more attention than
-others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}
-endure maintainability in 55.3\%, 51.3\% and 47.5\%, respectively. Following
-the implications of our work, we draw attention for the need of trustfull security
-frameworks, code best practices documentation to produce safe and maintainable
-code, proper authorization mechanisms design and implementation, better tools
-to introduce in the CI/CD pipeline to help software maintenance and new or
-improved programming languages to allow developers to easily implement safety
-in their software without endangering maintainability.
-
-As future work, this study may be extended in many directions: analyze which guidelines are
-more likely to affect security refactorings; evaluate the impact of each refactoring pattern on software maintainbility; investigate if there is any correlation between
-software maintainability and projects popularity; expand our methodology with other software quality
-properties; validate these findings with closed or private software; and, expand this analysis
-to other quality standards.
+This work presents an empirical study on the impact of 607 security refactorings
+on the maintainability of 94 open-source projects. We leveraged Better Code Hub
+reports of each project version to calculate maintainability based on a model
+proposed in previous work. Security refactorings significantly affect the
+software maintainability of open-source software, i.e., developers hinder
+maintainability when fixing vulnerabilities. We found this evidence in 40.2\% of
+the studied cases. Regarding regular commits, we observed that the overall
+maintainability increases (50.16\%). However, there is still a very considerable
+amout of cases that harm maintaniability (41.35\%).
+
+Our study also shows evidence that there are patterns that need more attention
+than others. \emph{Broken Autentication}, \emph{Denial-of-Service}, and
+\emph{Memory Leaks} endure maintainability in 55.3\%, 51.3\% and 47.5\%,
+respectively. Following the implications of our work, we draw attention for the
+need of code best practices  documentation to produce secure and maintainable
+code, proper authorization mechanisms design and implementation, better tools to
+introduce in the CI/CD pipeline to help software maintenance and new or improved
+programming languages to allow developers to easily implement security in their
+software without endangering maintainability.
+
+As future work, this study may be extended in many directions: analyze which
+guidelines are more likely to affect security refactorings; evaluate the impact
+of each refactoring pattern on software maintainbility; investigate if there is
+any correlation between software maintainability and projects popularity; expand
+our methodology with other software quality properties; validate these findings
+with closed or private software; and, expand this analysis to other quality
+standards.
 

### 2019-02-03 paper/ICPC19/icpc19.tex

--- 2019-02-03 paper/ICPC19/icpc19.tex
@@
 open-source software maintainability. Are developers affecting software
-maintainability when trying to improve software safety? We suspect that some of
-the patches may have a negative impact on the software maintainability and
+maintainability when trying to improve software security? We suspect that some
+patches may have a negative impact on the software maintainability and
 possibly even be the cause of the introduction of new vulnerabilities.
@@
 
-Every SSL/TLS connection begins with a handshake which is responsible by the negotiation between two parties.
-DTLS provides a mechanism for fragmenting a handshake message over a number of packets, each of which can be
-transmitted separately. In the example (Listing~\ref{lst:vuln}), all functions receive as input the
-SSL session, \texttt{s}.
-In the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake message size \texttt{msg\_hdr->msg\_len}
-(line $13$) and then against the maximum handshake message size allowed by DTLS1 for \texttt{s} (line $16$-$21$).
-However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line $37$,
-when \texttt{msg\_len} is allocated to the fragment buffer, more memory than the allowed might be allocated.
-According to the message commit, $10$ handshake messages are allowed per DTLS connection.
-Hence, a remote attacker could leverage this issue to trigger
-memory allocations corresponding to large length values and lead the system to memory exhaustion that could end
-in a DoS attack. The code changes performed to fix the vulnerability were the following:
-
-~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was inserted to return the number of
-maximum number of bytes allowed in a DTLS handshake message for \texttt{s}. The minimum is $16$KB but may be greater
-if the maximum certificate list size requires it which is set by the user. More specifically, it checks if the
-maximum certificate list size
+Every SSL/TLS connection begins with a handshake which is responsible by the
+negotiation between two parties. DTLS provides a mechanism for fragmenting a
+handshake message over a number of packets, each of which can be transmitted
+separately. In the example (Listing~\ref{lst:vuln}), all functions receive as
+input the SSL session, \texttt{s}. In the \texttt{dtls1\_reassemble\_fragment}
+function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is
+checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
+$13$) and then against the maximum handshake message size allowed by DTLS1 for
+\texttt{s} (line $16$-$21$). However, it is not verified if
+\texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line
+$37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
+than the allowed might be allocated. According to the message commit, $10$
+handshake messages are allowed per DTLS connection. Hence, a remote attacker
+could leverage this issue to trigger memory allocations corresponding to large
+length values and lead the system to memory exhaustion that could end in a DoS
+attack. The code changes performed to fix the vulnerability were the following:
+
+~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
+inserted to return the number of maximum number of bytes allowed in a DTLS
+handshake message for \texttt{s}. The minimum is $16$KB but may be greater if
+the maximum certificate list size requires it which is set by the user. More
+specifically, it checks if the maximum certificate list size
 (\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
 (\texttt{max\_len}). If that condition holds, the function returns the size of
-the certificate list, otherwise, it returns the maximum length. This function replaces lines $18$-$21$.
-
-~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to ensure that the message
-and the fragment length are never higher than the maximum allowed by DTSL. However, \ref{lst:func5} is responsible for patching the vulnerability.
-
-Although checking the maximum size of a handshake message seems an elementary problem to
-repair, the example shows that there was a significant amount of changes
-performed in the codebase which produced a negative impact in the project
-maintainability. Even though the developer tries to simplify the codebase with
-the creation of a smaller unit \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it
-also disrupts $2$ of the guidelines proposed by the Software Improvement Group
-(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}. It adds a
-new branch point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
+the certificate list, otherwise, it returns the maximum length. This function
+replaces lines $18$-$21$.
+
+~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to
+ensure that the message and the fragment length are never higher than the
+maximum allowed by DTSL. However, \ref{lst:func5} is responsible for patching
+the vulnerability.
+
+Although checking the maximum size of a handshake message seems an elementary
+problem to repair, the example shows that there was a significant amount of
+changes performed in the codebase which produced a negative impact in the
+project maintainability. Even though the developer tries to simplify the
+codebase with the creation of a smaller unit
+\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it also disrupts
+$2$ of the guidelines proposed by the Software Improvement Group (SIG) for
+building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
+point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
 (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
 \emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
-size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which contributes
-to a potential violation of the
-\emph{Keep Your Codebase Small} guideline. Hence, we have observed that the
-impact of this refactoring on the OpenSSL maintainability was negative.
+size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
+contributes to a potential violation of the \emph{Keep Your Codebase Small}
+guideline. Hence, we have observed that the impact of this refactoring on the
+OpenSSL maintainability was negative.
 
 In this paper, our concern is to study whether, while improving software
-security, developers are also negatively impacting the maintainability
-of their software applications. To answer the following two
-research questions, we use a dataset of security refactorings~\cite{Reis:2017:IJSSE} to measure the
-impact of the changes in the maintainability of open-source software.
+security, developers are also negatively impacting the maintainability of their
+software applications. To answer the following two research questions, we use a
+dataset of security refactorings~\cite{Reis:2017:IJSSE} to measure the impact of
+the changes in the maintainability of open-source software.
 
@@
 There are security flaws that are more difficult to refactor than others. For
-instance, implementing secure authentication is not as easy as fixing a cross-site scripting vulnerability since usually the last one is can be fixed without adding new lines of code, as seen in Figure~\ref{lst:fix}.
-The developer added the function \emph{htmlentities} to escape the data inserted by the variable \emph{\$\_['file']}.
-Knowing which patterns are more likely to increase maintainability
-issues is one step forward to bring awareness to security engineers of what
-patterns need more attention. The taxonomy of security patterns used to answer
-this question is the same as the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
-Maintainability is measured separately for each pattern.
+instance, implementing secure authentication is not as easy as fixing a
+cross-site scripting vulnerability since usually the last one is can be fixed
+without adding new lines of code, as seen in Figure~\ref{lst:fix}. The developer
+added the function \emph{htmlentities} to escape the data inserted by the
+variable \emph{\$\_['file']}. Knowing which patterns are more likely to increase
+maintainability issues is one step forward to bring awareness to security
+engineers of what patterns need more attention. The taxonomy of security
+patterns used to answer this question is the same as the one provided by the
+authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
+separately for each pattern.
 
@@
 The methodology used to measure the impact of security refactorings on the
-maintainability of open-source software is discussed in this section. The methodology
-comprehends the following steps, as illustrated in Figure~\ref{fig:met}:
+maintainability of open-source software is discussed in this section. The
+methodology comprehends the following steps, as illustrated in
+Figure~\ref{fig:met}:
+
 \begin{enumerate}
-	\item Extract relevant data from a dataset containing
-$716$ security refactorings collected from open-source software available on
+	\item Extract relevant data from a dataset containing $716$ security
+	refactorings collected from open-source software available on
 	GitHub~\cite{Reis:2017:IJSSE}.
-	\item A baseline of regular commits was
-randomly collected from the list of projects of the main dataset to evaluate the impact of regular commits on the maintainability of open-source
-software.
+	\item A baseline of regular commits was randomly collected from the list of
+	projects of the main dataset to evaluate the impact of regular commits on the
+	maintainability of open-source software.
   \item Use the Software Improvement Group's\footnote{SIG's website is available
@@
   service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
-https://bettercodehub.com/ (Accessed on \today{})} to quantify
-maintainability for both security and regular commits.
+  https://bettercodehub.com/ (Accessed on \today{})} to quantify maintainability
+  for both security and regular commits.
 \end{enumerate}
@@
 
-    \item \textbf{Broken Authentication \& Session Management.} Incorrectly implemented functionalities related to authentication and session management,
-    allowing an attacker to gain access to session tokens, passwords, keys, and other sensitive data. \textit{Based on the OWASP Top $10$ of $2013$.}
-
-    \item \textbf{Cross-Site Request Forgery.} Poor session tokens generation and management usually allow attackers to send forged HTTP requests including authentication information from the victim to the vulnerable web application.
-    \textit{Based on the OWASP Top $10$ of $2013$.}
-
-    \item \textbf{Denial-of-Service} Security flaws that can allow an attacker to flood or crashing services.
-This attacks usually occur when the system receives much traffic causing it to slow down and eventually stop.
-For
-    example, flaws allowing attackers to trigger memory allocations corresponding
-    to large length values (Listing~\ref{lst:vuln}).
+    \item \textbf{Broken Authentication \& Session Management.} Incorrectly
+    implemented functionalities related to authentication and session
+    management, allowing an attacker to gain access to session tokens,
+    passwords, keys, and other sensitive data. \textit{Based on the OWASP Top
+    $10$ of $2013$.}
+
+    \item \textbf{Cross-Site Request Forgery.} Poor session tokens generation
+    and management usually allow attackers to send forged HTTP requests
+    including authentication information from the victim to the vulnerable web
+    application. \textit{Based on the OWASP Top $10$ of $2013$.}
+
+    \item \textbf{Denial-of-Service} Security flaws that can allow an attacker
+    to flood or crashing services. This attacks usually occur when the system
+    receives much traffic causing it to slow down and eventually stop. For
+    example, flaws allowing attackers to trigger memory allocations
+    corresponding to large length values (Listing~\ref{lst:vuln}).
 
     \item \textbf{Miscellaneous} This pattern comprises several other categories
-    of security refactorings (e.g., \textit{path traversal and buffer overflow}).
-    It contains several other security refactorings that do not have their own
-    pattern yet. The patterns not satisfying the size requirement of more than $20$
-    refactorings are contemplated here: \textit{Resource Leak},
-    \textit{SHA-1 Hash Function}, \textit{Broken Access Control}, \textit{Sensitive Data Exposure},
-    \textit{Insufficient Attack Protection}, \textit{Overflow}, \textit{Path Traversal}.
-
-    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping allow attackers to submit untrusted data to web browsers through malicious scripts that can hijack the user sessions or redirect the user to malicious sites.
-\textit{Based on the OWASP Top $10$ of $2017$.}
-
-    \item \textbf{Injection.} When developers do not keep untrusted data separate from commands and queries. If an attacker sends a string that exploits the
-    syntax of the interpreter, then an injection attack is possible (e.g., SQL and
-    LDAP injection). \textit{Based on the OWASP Top $10$ of $2017$.}
+    of security refactorings (e.g., \textit{path traversal and buffer
+    overflow}). It contains several other security refactorings that do not have
+    their own pattern yet. The patterns not satisfying the size requirement of
+    more than $20$ refactorings are contemplated here: \textit{Resource Leak},
+    \textit{SHA-1 Hash Function}, \textit{Broken Access Control},
+    \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection},
+    \textit{Overflow}, \textit{Path Traversal}.
+
+    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
+    allow attackers to submit untrusted data to web browsers through malicious
+    scripts that can hijack the user sessions or redirect the user to malicious
+    sites. \textit{Based on the OWASP Top $10$ of $2017$.}
+
+    \item \textbf{Injection.} When developers do not keep untrusted data
+    separate from commands and queries. If an attacker sends a string that
+    exploits the syntax of the interpreter, then an injection attack is possible
+    (e.g., SQL and LDAP injection). \textit{Based on the OWASP Top $10$ of
+    $2017$.}
 
     \item \textbf{Memory Leak.} Memory management issue found more frequently in
-    programming languages that do not manage memory automatically (e.g., C/C++ and
-    Objective-C), i.e., where instead developers are responsible for handling it.
-    One of the main causes of DoS attacks.
+    programming languages that do not manage memory automatically (e.g., C/C++
+    and Objective-C), i.e., where instead developers are responsible for
+    handling it. One of the main causes of DoS attacks.
 \end{itemize}
@@
 Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
-open-source projects involved in this study, including the number of forks, the number of
-stars, the number of watchers, the number of contributors, the number of commits, the number of
-branches, the number of releases,  size (cf. GitHub it is the size of the whole
-repository, including all of its history, in kilobytes), the number of issues, and the
-number of pull requests.
+open-source projects involved in this study, including the number of forks, the
+number of stars, the number of watchers, the number of contributors, the number
+of commits, the number of branches, the number of releases,  size (cf. GitHub it
+is the size of the whole repository, including all of its history, in
+kilobytes), the number of issues, and the number of pull requests.
 
@@
 Previous studies attempted to measure the impact of regular refactorings on
-open-source software maintainability~\cite{HEGEDUS2018313}. However,
-there is no previous work focused on comparing the impact of security
-refactorings with regular refactorings on maintainability.
-We analyze the maintainability of regular commits
-(i.e., commits not related to security fixes) and use them as a baseline.
+open-source software maintainability~\cite{HEGEDUS2018313}. However, there is no
+previous work focused on comparing the impact of security refactorings with
+regular refactorings on maintainability. We analyze the maintainability of
+regular commits (i.e., commits not related to security fixes) and use them as a
+baseline.
 
@@
 \noindent
-where $G$ is the group of selected maintainability guidelines from BCH (Table~\ref{tab:guidelines}) and $v$ is the version of the software under evaluation.
-The maintenance $M$ based on the guideline $g$ for a given version of a project
-is computed with the following equation:
+where $G$ is the group of selected maintainability guidelines from BCH
+(Table~\ref{tab:guidelines}) and $v$ is the version of the software under
+evaluation. The maintenance $M$ based on the guideline $g$ for a given version
+of a project is computed with the following equation:
 
@@
 The impact of security and baseline commits in the maintainability of
-open-source software is presented in Figure~\ref{fig:secvsreg}. For each type
-of change, three types of impact of refactorings in the software maintainability
-are presented: \emph{negative}, maintainability decreases (\emph{red}); \emph{none},
-maintainability stays the same (\emph{yellow}); and, \emph{positive}, maintainability
-increases (\emph{green}). In front of each group, it is presented the mean ($\overline{x}$)
-and median ({$\widetilde{x}$) of the maintainability difference and the p-value
-resulting from the Wilcoxon signed-rank test.
+open-source software is presented in Figure~\ref{fig:secvsreg}. For each type of
+change, three types of impact of refactorings in the software maintainability
+are presented: \emph{negative}, maintainability decreases (\emph{red});
+\emph{none}, maintainability stays the same (\emph{yellow}); and,
+\emph{positive}, maintainability increases (\emph{green}). In front of each
+group, it is presented the mean ($\overline{x}$) and median ({$\widetilde{x}$)
+of the maintainability difference and the p-value resulting from the Wilcoxon
+signed-rank test.
 
@@
 
-Regarding security commits results, we observe that maintainability decreases
-in $40.20\%$ ($244$), stays equal in $30.5\%$ ($185$) and increases in $29.33\%$ ($178$)
-of the cases after security refactorings being applied. The resulting p-value
-of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$. Since the p-value is below
-the significance level of $0.05$, it is reasonable to affirm that \emph{security
-refactorings have a negative impact in the maintainability of open-source software}.
+Regarding security commits results, we observe that maintainability decreases in
+$40.20\%$ ($244$), stays equal in $30.5\%$ ($185$) and increases in $29.33\%$
+($178$) of the cases after security refactorings being applied. The resulting
+p-value of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$. Since the p-value
+is below the significance level of $0.05$, it is reasonable to affirm that
+\emph{security refactorings have a negative impact in the maintainability of
+open-source software}.
 
@@
 ($251$) of the cases after regular refactorings being applyed. But in contrast
-to security commits, it increases in $58.16\%$ ($353$) and stays equal in $0.5\%$
-($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank test is
-$1.54$x$10^{-6}$. Since the p-value is below the significance level of $0.05$,
-it is reasonable to affirm that \emph{regular refactorings have a positive
-impact in the maintainability of open-source software}.
+to security commits, it increases in $58.16\%$ ($353$) and stays equal in
+$0.5\%$ ($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank
+test is $1.54$x$10^{-6}$. Since the p-value is below the significance level of
+$0.05$, it is reasonable to affirm that \emph{regular refactorings have a
+positive impact in the maintainability of open-source software}.
 
@@
 In this study, we also evaluated the impact of each programming language on the
-maintainability of open-source software, Figure~\ref{fig:lang_main}. Although the
-dataset contains security refactorings for more than $13$ languages, we only
-present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP} because
-one of the requirements of the Wilcoxon signed-rank test is that the input sample
-has to have more than $20$ measurements. The other languages were integrated into the
-\emph{Others} set. The results of the Wilcoxon signed-rank test by pattern returned
-statistical evidence which supports that the maintainability of open-source software
-decreases for security refactorings in $3$ different languages: \emph{C}
-(p = $5.72$x$10^{-9}$), Ruby (p = $0.057$) and PHP (p = $3.54$x$10^{-6}$). The results
-also provide statistical significance for when security refactorings are performed
-in \emph{Java} (p = $3.61$x$10^{-2}$). \emph{Python} and the languages
-that integrate the \emph{Others} set contain more cases where the maintainability
-increases. However, the results obtained for these patterns are not statistically
+maintainability of open-source software, Figure~\ref{fig:lang_main}. Although
+the dataset contains security refactorings for more than $13$ languages, we only
+present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP}
+because one of the requirements of the Wilcoxon signed-rank test is that the
+input sample has to have more than $20$ measurements. The other languages were
+integrated into the \emph{Others} set. The results of the Wilcoxon signed-rank
+test by pattern returned statistical evidence which supports that the
+maintainability of open-source software decreases for security refactorings in
+$3$ different languages: \emph{C} (p = $5.72$x$10^{-9}$), Ruby (p = $0.057$) and
+PHP (p = $3.54$x$10^{-6}$). The results also provide statistical significance
+for when security refactorings are performed in \emph{Java} (p =
+$3.61$x$10^{-2}$). \emph{Python} and the languages that integrate the
+\emph{Others} set contain more cases where the maintainability increases.
+However, the results obtained for these patterns are not statistically
 significant.
@@
 supports that the maintainability of open-source software decreases for $4$
-different patterns: \emph{Broken Autentication} (p = $0.018$), \emph{Denial-of-Service}
-(p = $0.029$), \emph{Memory Leak} (p = $0.009$) and \emph{Miscellaneous} (p = $3.07$x$10^{-4}$).
-The results also returned statistical significance that evidence that maintainability
-is more likely to not suffer differences for open-source software after removing
-\emph{Components with Known Vulnerabilities}\Luis{o resultado e bem diferente dos regular commits.
-este refactoring tem alguma particularidade para deixar a maintainability intacta?
-convém falar sobre isso na discussão} (p = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site
-Scripting Vulnerabilities} (p = $0.001$). \emph{Injection} and \emph{Cross-Site Request Forgery}
-produced more cases where the refactorings had a negative impact in the maintainability.
-However, the results obtained for these patterns are not statistically significant.
-
-Projects with \emph{Broken Authentication} flaws are the most affected ones by security
-refactorings: $55.3\%$ ($21$) of the refactorings had a negative impact in the maintainability
-while $35.0\%$ ($11$) yielded a positive impact, and only $13.8\%$ ($6$) did not suffer any
-change. The results also show that fixing \emph{Memory Leak}s affects the maintainability
-of $51.3\%$ ($41$) of the cases and leaves only $13.8\%$ ($11$) of the cases unaffected.
-In $35\%$ ($28$) of the cases, maintainability increased. Refactoring vulnerabilities
-that allow \emph{Denial-of-Service} attacks (e.g., CVE-$2014$-$3506$) have a negative impact
-of $47.5\%$ ($19$) in the maintainability of open-source software while $30.0\%$ ($12$) of
-the cases exhibit maintainability improvement and $22.5\%$ ($9$) show that
-maintainability does not suffer any change. The \emph{Miscellaneous} refactorings pattern
-integrates several types of security flaws.
-These type of refactorings harm maintainability in $43.1\%$ ($84$) of the cases.
-In $31.8\%$ ($62$), there is no effect in maintainability and for the other $25.1\%$ ($49$) maintainability increases.
+different patterns: \emph{Broken Autentication} (p = $0.018$),
+\emph{Denial-of-Service} (p = $0.029$), \emph{Memory Leak} (p = $0.009$) and
+\emph{Miscellaneous} (p = $3.07$x$10^{-4}$). The results also returned
+statistical significance that evidence that maintainability is more likely to
+not suffer differences for open-source software after removing \emph{Components
+with Known Vulnerabilities}\Luis{o resultado e bem diferente dos regular
+commits. este refactoring tem alguma particularidade para deixar a
+maintainability intacta? convém falar sobre isso na discussão} (p =
+$7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting Vulnerabilities} (p =
+$0.001$). \emph{Injection} and \emph{Cross-Site Request Forgery} produced more
+cases where the refactorings had a negative impact in the maintainability.
+However, the results obtained for these patterns are not statistically
+significant.
+
+Projects with \emph{Broken Authentication} flaws are the most affected ones by
+security refactorings: $55.3\%$ ($21$) of the refactorings had a negative impact
+in the maintainability while $35.0\%$ ($11$) yielded a positive impact, and only
+$13.8\%$ ($6$) did not suffer any change. The results also show that fixing
+\emph{Memory Leak}s affects the maintainability of $51.3\%$ ($41$) of the cases
+and leaves only $13.8\%$ ($11$) of the cases unaffected. In $35\%$ ($28$) of the
+cases, maintainability increased. Refactoring vulnerabilities that allow
+\emph{Denial-of-Service} attacks (e.g., CVE-$2014$-$3506$) have a negative
+impact of $47.5\%$ ($19$) in the maintainability of open-source software while
+$30.0\%$ ($12$) of the cases exhibit maintainability improvement and $22.5\%$
+($9$) show that maintainability does not suffer any change. The
+\emph{Miscellaneous} refactorings pattern integrates several types of security
+flaws. These type of refactorings harm maintainability in $43.1\%$ ($84$) of the
+cases. In $31.8\%$ ($62$), there is no effect in maintainability and for the
+other $25.1\%$ ($49$) maintainability increases.
 
 It is also important to notice that overall the maintainability of open-source
-software is not affected by \emph{Cross-Site Scripting} vulnerabilities and by \emph{Using
-Components with Known Vulnerabilities}. \emph{Cross-Site Scripting} does not suffer any
-decrease in the maintainability of $55.3\%$ ($63$) of cases and $21.1\%$ ($24$) have a
-positive effect on it. Only $23.7\%$ ($27$) of the cases hinder the software
-maintainability. From the $21$ cases evaluated for \emph{Using Componentes with Known
-Vulnerabilities}, only $4.8\%$ ($1$) have a positive effect in the maintainability
-of software while $14.3\%$ ($3$) harms maintainability software, and $81\%$ ($17$) have no effect
-in the maintainability.
+software is not affected by \emph{Cross-Site Scripting} vulnerabilities and by
+\emph{Using Components with Known Vulnerabilities}. \emph{Cross-Site Scripting}
+does not suffer any decrease in the maintainability of $55.3\%$ ($63$) of cases
+and $21.1\%$ ($24$) have a positive effect on it. Only $23.7\%$ ($27$) of the
+cases hinder the software maintainability. From the $21$ cases evaluated for
+\emph{Using Componentes with Known Vulnerabilities}, only $4.8\%$ ($1$) have a
+positive effect in the maintainability of software while $14.3\%$ ($3$) harms
+maintainability software, and $81\%$ ($17$) have no effect in the
+maintainability.
 
@@
 \begin{itemize}
-	\item \textbf{Security Frameworks} need to be secure and user-friendly~\cite{5287006, 7676144}. Unfortunately,
-	there are several vulnerabilities that stay undisclosed. Thus, developers should not
-	trust them blindly and check defaults.
-	\item \textbf{Documentation} of software should provide developers with the best practices to implement safe and maintainable software. Unmaintainable code is difficult to test, analyze and re-use.
-	\item \textbf{Proper Authorization Mechanisms} should be designed and implemented. This is
-	one of the most complex security features to implement.
-Some mechanisms might assist authorization maintenance in long run, such as the combination of \texttt{HTTP} filters with \texttt{Spring Security} annotations.
-	\item \textbf{Deprecated code} should be avoided by developers.
-They should integrate tools such as BCH on their projects CI/CD pipelines to help them be more aware of the maintainability of their code and issues (e.g., dead and duplicated code).
-	\item\textbf{Programming Languages} should provide new mechanisms to easily implement security patterns without endangering maintainability.
-
+	\item \textbf{Security Frameworks} need to be secure and
+	user-friendly~\cite{5287006, 7676144}. Unfortunately, there are several
+	vulnerabilities that stay undisclosed. Thus, developers should not trust them
+	blindly and check defaults.
+
+	\item \textbf{Documentation} of software should provide developers with the
+	best practices to implement secure and maintainable software. Unmaintainable
+	code is difficult to test, analyze and re-use.
+
+	\item \textbf{Proper Authorization Mechanisms} should be designed and
+	implemented. This is one of the most complex security features to implement.
+	Some mechanisms might assist authorization maintenance in long run, such as
+	the combination of \texttt{HTTP} filters with \texttt{Spring Security}
+	annotations.
+
+	\item \textbf{Deprecated code} should be avoided by developers. They should
+	integrate tools such as BCH on their projects CI/CD pipelines to help them be
+	more aware of the maintainability of their code and issues (e.g., dead and
+	duplicated code).
+
+	\item\textbf{Programming Languages} should provide new mechanisms to easily
+	implement security patterns without endangering maintainability.
 \end{itemize}
 
-We would expect that the negative impact of programming languages on maintainability
-would be more intense due to the buzz of the last years surrounding the idea that
-poor design of programming languages and lack of using the best practices lead to
-more buggy and vulnerable code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}.
-However, Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of more than $50\%$ on software maintainability. We suspect that these values are the result of projects
-contributions policies. Nine out of ten projects --- the equivalent to $31.8\%$ of the commits ---
-of the top-$10$ of dataset projects with more contributors have restrict contribution
-policies regarding code standards. Some examples are the \texttt{laravel/framework},
-\texttt{php/php-src}, \texttt{openssl/openssl} and
-\texttt{cakephp/cakephp}. For instance, \texttt{laravel/framework} requires the
-\texttt{PSR-2 Standard}\footnote{PSR-2 Standard available at
+We would expect that the negative impact of programming languages on
+maintainability would be more intense due to the buzz of the last years
+surrounding the idea that poor design of programming languages and lack of using
+the best practices lead to more buggy and vulnerable
+code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
+Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of
+more than $50\%$ on software maintainability. We suspect that these values are
+the result of projects contributions policies. Nine out of ten projects --- the
+equivalent to $31.8\%$ of the commits --- of the top-$10$ of dataset projects
+with more contributors have restrict contribution policies regarding code
+standards. Some examples are the \texttt{laravel/framework},
+\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
+instance, \texttt{laravel/framework} requires the \texttt{PSR-2
+Standard}\footnote{PSR-2 Standard available at
 https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.
@@
 \begin{framed}
-\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to affect open-source software maintainability?}
+\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
+affect open-source software maintainability?}
 \end{framed}
 
-\textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service attacks and Memory Leaks and Miscellaneous patterns are more likely to affect open-source software maintainability.}} Although there are
-a few refactorings that harm software maintainability in the \emph{Cross-Site Scripting} and \emph{Using Components with Known Vulnerabilities}, overall maintainability does not suffer any changes for these patterns.
-For the remaining patterns, \emph{Injection} and \emph{Cross-Site Request Forgery}, results are not statistically significant.
+\textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service
+attacks and Memory Leaks and Miscellaneous patterns are more likely to affect
+open-source software maintainability.}} Although there are a few refactorings
+that harm software maintainability in the \emph{Cross-Site Scripting} and
+\emph{Using Components with Known Vulnerabilities}, overall maintainability does
+not suffer any changes for these patterns. For the remaining patterns,
+\emph{Injection} and \emph{Cross-Site Request Forgery}, results are not
+statistically significant.
 
@@
 adds complexity to the codebase it is probably affecting the software
-maintainability. One evidence of that is the fact of \emph{Cross-Site
-Scripting} and \emph{Using Components with Known Vulnerabilities} refactorings
-endure more cases without any impact in the maintainability. Typically, these
-types of vulnerabilities do not need extra lines to be fixed, as shown in
-Listing~\ref{lst:fix}. 
-Whereas, \emph{Broken Authentication} and vulnerabilities such as the one presented in
-Section~\ref{sec:motivation} responsible by DoS attacks are more difficult to
-fix (several code lines deleted and added). Another example is the Apple
-authentication flaw discovered in $2017$ (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$
-details available at https://support.apple.com/en-us/HT208315 (Accessed on
-\today{})}, where any user could log in as root with an empty password. Patrick
-Wardle examined the cause of the issue\footnote{\emph{Why $<$blank$>$ Gets You
-Root} available at https://objective-see.com/blog/blog\_0x24.html (Accessed on
-\today{})} and conclude that the flaw was due to the introduction of high
-cyclomatic complexity in a method used to verify the password. In sum,
-\emph{Cross-Site Scripting} and \emph{Using Components with Known
-Vulnerabilities} should take part of the developers' preoccupations along with
-the other patterns that explicitly hinder maintainability because despite their
-low complexity these are patterns that still produce a considerable negative
-impact on the maintainability. Real efforts have been made in creating and
-incorporating coding standards and best practices in software production.
-However, security is still far from perfection and fixing vulnerabilities might
-hinder software maintainability. Thus, it is important to provide tool support
-to developers to help them apply these patterns without harming software
-maintainability.
+maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting}
+and \emph{Using Components with Known Vulnerabilities} refactorings endure more
+cases without any impact in the maintainability. Typically, these types of
+vulnerabilities do not need extra lines to be fixed, as shown in
+Listing~\ref{lst:fix}. Whereas, \emph{Broken Authentication} and vulnerabilities
+such as the one presented in Section~\ref{sec:motivation} responsible by DoS
+attacks are more difficult to fix (several code lines deleted and added).
+Another example is the Apple authentication flaw discovered in $2017$
+(CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
+https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user
+could log in as root with an empty password. Patrick Wardle examined the cause
+of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
+https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and
+conclude that the flaw was due to the introduction of high cyclomatic complexity
+in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and
+\emph{Using Components with Known Vulnerabilities} should take part of the
+developers' preoccupations along with the other patterns that explicitly hinder
+maintainability because despite their low complexity these are patterns that
+still produce a considerable negative impact on the maintainability. Real
+efforts have been made in creating and incorporating coding standards and best
+practices in software production. However, security is still far from perfection
+and fixing vulnerabilities might hinder software maintainability. Thus, it is
+important to provide tool support to developers to help them apply these
+patterns without harming software maintainability.
 
@@
 %
-The following section presents the potential threats to the validity of this study.
+The following section presents the potential threats to the validity of this
+study.
 %
@@
 %
-
-The maintainability formula was inferred based on the BCH's reports.
-The high amount of different projects and backgrounds may require different maintainability
-standards. However, BCH does have a representative benchmark of closed and open-source
-software projects that are under the calculation of the thresholds used in the computation
-of each maintainability guideline (Table~\ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
+The maintainability formula was inferred based on the BCH's reports. The high
+amount of different projects and backgrounds may require different
+maintainability standards. However, BCH does have a representative benchmark of
+closed and open-source software projects that are under the calculation of the
+thresholds used in the computation of each maintainability guideline
+(Table~\ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
 
@@
 
-The security refactorings dataset provided by the previous work~\cite{Reis:2017:IJSSE}
-was collected based on the messages of GitHub commits produced by project developers
-to classify the changes performed by them to fix security flaws. This approach
-discards security refactorings that were not explicit in commits messages.
+The security refactorings dataset provided by the previous
+work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
+commits produced by project developers to classify the changes performed by them
+to fix security flaws. This approach discards security refactorings that were
+not explicit in commits messages.
 
@@
 maintainability may still be affected by the developers' experience, coding
-style and software contribution policies which is not evaluated in this
-study. Furthermore, this evaluation considers that $607$ regular commits are
-enough to alleviate random irregularities in the maintainability differences of
-the baseline.
+style and software contribution policies which is not evaluated in this study.
+Furthermore, this evaluation considers that $607$ regular commits are enough to
+alleviate random irregularities in the maintainability differences of the
+baseline.
 
 \subsection{External}
-
-\Luis{A primeira threat que falas é que os resultados podem nao generalizar
-para software fechado. Eu acho que deviamos ser mais diretos.}
-\Sof{Melhor?}
 
@@
 address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be properly representative of the population regarding programming languages.
+the dataset may not be properly representative of the population regarding
+programming languages.
 
@@
 
-Many studies have investigated the relationship between refactorings and software quality.
-Previous work focused on coupling metrics has evaluated the impact of refactorings and
-exhibited proof that quantifying effectively the impact of refactoring on maintainability
-may help to choose the appropriate refactoring type~\cite{1167822}. In contrast to
-this work, Hegedus et al.~\cite{HEGEDUS2018313} did not select particular metrics to
-assess the effect of refactorings. Instead, statistical tests were used to find
-the metrics that have the potential to change significantly after refactorings.
-The differences of maintainability between groups of refactored elements and
-non-refactored were analyzed. In addition, the authors conclude that the source
-code elements subjected to refactorings had significantly lower maintainability
-than the elements not affected. Studying the evolution of maintainability issues during the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041} discovered that maintainability
-decreases over time. Palomba et al. ($2018$)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits
-proof that code smells should be carefully monitored by programmers since there is
-a high correlation between maintainability aspects and proneness to changes and faults.
-The present work uses a model proposed by BCH and focuses solely on evaluating
-the impact of patching vulnerabilities on software maintainability.
-
-\Luis{eu nao acho que usar $10$ guidelines proposed by the Software Improvement
-Group (SIG) seja algo que nos esteja a diferenciar. Acho que o a prinicpal
-diferenca é que estamos a estudar maintainability issues noutro contexto.}\Sof{Melhor?}
-
-Researchers investigated the relationship between design patterns and
-software maintainability~\cite{10.1007/978-3-642-35267-6-18}. $300$ revisions
-of \texttt{JHotDraw} were analyzed to conclude that every introduced pattern
+Many studies have investigated the relationship between refactorings and
+software quality. Previous work focused on coupling metrics has evaluated the
+impact of refactorings and exhibited proof that quantifying effectively the
+impact of refactoring on maintainability may help to choose the appropriate
+refactoring type~\cite{1167822}. In contrast to this work, Hegedus et
+al.~\cite{HEGEDUS2018313} did not select particular metrics to assess the effect
+of refactorings. Instead, statistical tests were used to find the metrics that
+have the potential to change significantly after refactorings. The differences
+of maintainability between groups of refactored elements and non-refactored were
+analyzed. In addition, the authors conclude that the source code elements
+subjected to refactorings had significantly lower maintainability than the
+elements not affected. Studying the evolution of maintainability issues during
+the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
+discovered that maintainability decreases over time. Palomba et al.
+($2018$)~\cite{Palomba:2018:DIM:3231288.3231337} exhibits proof that code smells
+should be carefully monitored by programmers since there is a high correlation
+between maintainability aspects and proneness to changes and faults. The present
+work uses a model proposed by BCH and focuses solely on evaluating the impact of
+patching vulnerabilities on software maintainability.
+
+Researchers investigated the relationship between design patterns and software
+maintainability~\cite{10.1007/978-3-642-35267-6-18}. $300$ revisions of
+\texttt{JHotDraw} were analyzed to conclude that every introduced pattern
 instance improved software maintainability. However, other studies show that the
-use of design patterns may introduce maintainability issues into software~\cite{4493325}.
-The present work studies how security patterns influence maintainability
-for open-source software.
-
-Other studies investigated the effect of programming languages in software quality~\cite{Ray:2017:LSP:3144574.3126905,
-2019arXiv190110220B}. The first study shows that some programming languages are more buggy than others, i.e.,
-language design has an impact on software quality. However, the authors of the second
-study were not able to reproduce the first one and did not obtain any evidence about the
-impact of language design. Both have used the number of defects as an indicator of
-software quality. The present work studies how security refactorings affect software quality
-based on the code maintainability analysis and provides some evidence that programming language have impact on maintainability.
-
-\Luis{ tambem fizemos analise por linguagem, nesta ultima citacao deviamos falar sobre isso tb.}\Sof{Acho que esta!}
+use of design patterns may introduce maintainability issues into
+software~\cite{4493325}. The present work studies how security patterns
+influence maintainability for open-source software.
+
+Other studies investigated the effect of programming languages in software
+quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
+study shows that some programming languages are more buggy than others, i.e.,
+language design has an impact on software quality. However, the authors of the
+second study were not able to reproduce the first one and did not obtain any
+evidence about the impact of language design. Both have used the number of
+defects as an indicator of software quality. The present work studies how
+security refactorings affect software quality based on the code maintainability
+analysis and provides some evidence that programming language have impact on
+maintainability.
 
@@
 
-This work presents an empirical study on the impact of $607$ security refactorings on
-the maintainability of $94$ open-source projects. We leveraged Better Code Hub reports
-of each project version to calculate maintainability based on a model proposed in
-previous work. Security refactorings significantly affect the software maintainability of open-source
-software, i.e., developers hinder maintainability when fixing vulnerabilities. We found
-this evidence in $40.2\%$ of the studied cases. Regarding regular commits, it seems that
-overall maintainability increases ($50.16\%$). However, there is still a very considerable amount of
-cases that harm maintainability ($41.35\%$).
-
-\Luis{secure security frameworks, nao percebi bem o que se quer dizer aí.}
-\Sof{Melhor?}
+This work presents an empirical study on the impact of $607$ security
+refactorings on the maintainability of $94$ open-source projects. We leveraged
+Better Code Hub reports of each project version to calculate maintainability
+based on a model proposed in previous work. Security refactorings significantly
+affect the software maintainability of open-source software, i.e., developers
+hinder maintainability when fixing vulnerabilities. We found this evidence in
+$40.2\%$ of the studied cases. Regarding regular commits, it seems that overall
+maintainability increases ($50.16\%$). However, there is still a very
+considerable amount of cases that harm maintainability ($41.35\%$).
 
 This study also exhibits evidence that some patterns need more attention than
-others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and \emph{Memory Leaks}
-endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$, respectively. Following
-the implications of our work, we draw attention for the need of trustful security
-frameworks, code best practices documentation to produce safe and maintainable
-code, proper authorization mechanisms design and implementation, better tools
-to introduce in the CI/CD pipeline to help software maintenance and new or
-improved programming languages to allow developers to easily implement safety
-in their software without endangering maintainability.
-
-As future work, this study may be extended in many directions: analyze which guidelines are
-more likely to affect security refactorings; evaluate the impact of each refactoring pattern on software maintainability; investigate if there is any correlation between
-software maintainability and projects popularity; expand our methodology with other software quality
-properties; validate these findings with closed or private software; and, expand this analysis
-to other quality standards.
+others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and
+\emph{Memory Leaks} endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$,
+respectively. Following the implications of our work, we draw attention for the
+need of trustful security frameworks, code best practices documentation to
+produce safe and maintainable code, proper authorization mechanisms design and
+implementation, better tools to introduce in the CI/CD pipeline to help software
+maintenance and new or improved programming languages to allow developers to
+easily implement safety in their software without endangering maintainability.
+
+As future work, this study may be extended in many directions: analyze which
+guidelines are more likely to affect security refactorings; evaluate the impact
+of each refactoring pattern on software maintainability; investigate if there is
+any correlation between software maintainability and projects popularity; expand
+our methodology with other software quality properties; validate these findings
+with closed or private software; and, expand this analysis to other quality
+standards.
 

### 2019-02-03 paper/ICPC19/icpc19.tex
fixed sentence

@@
 \emph{Denial-of-Service} (p = $0.029$), \emph{Memory Leak} (p = $0.009$) and
-\emph{Miscellaneous} (p = $3.07$x$10^{-4}$). The results also returned
-statistical significance that evidence that maintainability is more likely to
-not suffer differences for open-source software after removing \emph{Components
-with Known Vulnerabilities}\Luis{o resultado e bem diferente dos regular
-commits. este refactoring tem alguma particularidade para deixar a
-maintainability intacta? convém falar sobre isso na discussão} (p =
-$7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting Vulnerabilities} (p =
-$0.001$). \emph{Injection} and \emph{Cross-Site Request Forgery} produced more
+\emph{Miscellaneous} (p = $3.07$x$10^{-4}$). With statistical significance,
+results demonstrate evidence that maintainability is not negatively impacted
+when removing \emph{Components with Known Vulnerabilities} (p = $7.96$x$10^{-4}$)
+and fixing \emph{Cross-Site Scripting Vulnerabilities} (p = $0.001$).
+\emph{Injection} and \emph{Cross-Site Request Forgery} produced more
 cases where the refactorings had a negative impact in the maintainability.
@@
 
-This study also exhibits evidence that some patterns need more attention than
+This study also exhibits evidence that some security patterns need more attention than
 others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and

### 2019-02-04 paper/ICPC19/icpc19.tex
fix caption
fixing language order comment
adjusting columns 🙃

@@
 As an example, in the last week of January $2019$, a severe vulnerability in Apple
-FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html
-(Accessed on \today{})} --- the outcome of a design/logical flaw that allowed callers to see and
-hear others without them picking up the call.
+FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-
+privacy-hack.html (Accessed on \today{})} --- the outcome of a design/logical 
+flaw that allowed callers to see and hear others without them picking up the call.
 
@@
 to find vulnerabilities and defects. Therefore, writing maintainable code
-prevents the introduction of vulnerabilities in the codebases. In this work, we are
-concerned about the impact that patching vulnerabilities may have in the
+prevents the introduction of vulnerabilities in the codebases. In this work, we 
+are concerned about the impact that patching vulnerabilities may have in the
 open-source software maintainability. Are developers affecting software
@@
 it, correct it or adapt it to changes in environment, and in requirements'' on
-software quality ISO/IEC 25010. As ISO does not provide any specific guidelines or
-formula to calculate maintainability, we resort on Software Improvement
+software quality ISO/IEC 25010. As ISO does not provide any specific guidelines 
+or formula to calculate maintainability, we resort on Software Improvement
 Group (SIG)'s web-based source code analysis service Better Code Hub
 (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
-project. There are other well-known standards and models that have been proposed to increase software security:
-Common Criteria~\cite{common:2009} which suffered a lot of criticism regarding the
-costs associated and poor technical evaluation; the OWASP
-Application Security Verification Criteria~\cite{oswap:2009} which is focused only
-on web applications, and a model proposed by Xu et al. ($2013$)~\cite{6616351} for
-rating software security which we believe it was one of the first steps taken
-by SIG to introduce security on their maintainability model.
+project. There are other well-known standards and models that have been proposed
+to increase software security: Common Criteria~\cite{common:2009} which suffered
+a lot of criticism regarding the costs associated and poor technical evaluation; 
+the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is 
+focused only on web applications, and a model proposed by Xu et al. ($2013$)
+~\cite{6616351} for rating software security which we believe it was one of the
+first steps taken by SIG to introduce security on their maintainability model.
 
 BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
-base of BCH's evaluation. Two examples are the
-\emph{Write Simple Units of Code} based on the McCabe Complexity~\cite{1702388}
-and \emph{Keep your Codebase Small} based on the idea that smaller codebases are
-easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
+base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code} 
+based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
+based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
 
@@
 between refactoring source code to patch disclosed vulnerabilities while keeping
-(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313},
-analyzed the difference of maintainability between
+(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
 groups of refactored elements and non-refactored and conclude that the source
@@
 programming languages. The main contributions of the present work are:
+
 \begin{itemize}
@@
 	maintainability and what patterns need more attention.
-
 	\item A replication package with all scripts and data created to perform the
@@
   \url{https://figshare.com/s/3e17518222df6a11a8d6}.
-
 \end{itemize}
@@
 address \emph{Broken Authentication and Session Management} issues, \emph{Memory
-Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight the need for tools
-to predict the impact of patches on maintainability~\cite{4724577} and documentation
-to help developers adopt security best practices~\cite{6311252, 7927935, MESQUIDA201519}.
+Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight 
+the need for tools to predict the impact of patches on maintainability~\cite{4724577} 
+and documentation to help developers adopt security best practices~\cite{6311252, 
+7927935, MESQUIDA201519}.
 
@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
- 	\caption{Distribution of Security Refactorings per Language\Luis{Consegues ordenar estas linguagens da mesma forma que estão ordenadas na figura7}}
+ 	\caption{Distribution of Security Refactorings per Language}
 	\label{fig:lang}
@@
  	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
- 	\caption{Distribution of Projects Domain}
+ 	\caption{Distribution of the Domain of Projects}
 	\label{fig:domain}

### 2019-02-04 paper/ICPC19/icpc19.tex
S1 and S2

@@
 project. There are other well-known standards and models that have been proposed
-to increase software security: Common Criteria~\cite{common:2009} which suffered
-a lot of criticism regarding the costs associated and poor technical evaluation; 
+to increase software security: Common Criteria~\cite{common:2009} which received
+negative criticism regarding the costs associated and poor technical evaluation;
 the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
 focused only on web applications, and a model proposed by Xu et al. ($2013$)
-~\cite{6616351} for rating software security which we believe it was one of the
-first steps taken by SIG to introduce security on their maintainability model.
+~\cite{6616351} for rating software security (arguably, it was one of the
+first steps taken by SIG to introduce security on their maintainability model).
 
@@
 The particular subject of this paper is to explore whether there is a trade-off
-between refactoring source code to patch disclosed vulnerabilities while keeping
+between refactoring source code to patch disclosed vulnerabilities and keeping
 (or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
@@
 code elements subjected to refactorings had significantly lower maintainability
-than the elements not affected.
+than the elements not modified.
 
@@
 $37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
-than the allowed might be allocated. According to the message commit, $10$
+than the allowed might be allocated. According to the commit message, $10$
 handshake messages are allowed per DTLS connection. Hence, a remote attacker
@@
 ~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
-inserted to return the number of maximum number of bytes allowed in a DTLS
+inserted to return the maximum number of bytes allowed in a DTLS
 handshake message for \texttt{s}. The minimum is $16$KB but may be greater if
@@
 
-~\ref{lst:func5} and~\ref{lst:func4} add the check to different functions to
-ensure that the message and the fragment length are never higher than the
-maximum allowed by DTSL. However, \ref{lst:func5} is responsible for patching
+~\ref{lst:func5} and~\ref{lst:func4} are guards added to different functions to
+ascertain that the message and the fragment length are never higher than the
+maximum allowed by DTSL. In particular, \ref{lst:func5} is responsible for patching
 the vulnerability.
@@
 changes performed in the codebase which produced a negative impact in the
-project maintainability. Even though the developer tries to simplify the
-codebase with the creation of a smaller unit
-\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), it also disrupts
-$2$ of the guidelines proposed by the Software Improvement Group (SIG) for
+maintainability. Even though the developer tries to simplify the
+codebase by introducing a smaller unit
+\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
+disrupts $2$ of the guidelines proposed by the Software Improvement Group (SIG) for
 building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
@@
 without adding new lines of code, as seen in Figure~\ref{lst:fix}. The developer
-added the function \emph{htmlentities} to escape the data inserted by the
+added the function \emph{htmlentities} to \textit{escape} the data given by the
 variable \emph{\$\_['file']}. Knowing which patterns are more likely to increase

### 2019-02-04 paper/ICPC19/icpc19.tex
changing comprehends
swap misc
comtemplates

@@
 maintainability of open-source software is discussed in this section. The
-methodology comprehends the following steps, as illustrated in
+methodology is comprised of the following steps, as illustrated in
 Figure~\ref{fig:met}:
@@
 
+    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
+    allow attackers to submit untrusted data to web browsers through malicious
+    scripts that can hijack the user sessions or redirect the user to malicious
+    sites. \textit{Based on the OWASP Top $10$ of $2017$.}
+
+    \item \textbf{Injection.} When developers do not keep untrusted data
+    separate from commands and queries. If an attacker sends a string that
+    exploits the syntax of the interpreter, then an injection attack is possible
+    (e.g., SQL and LDAP injection). \textit{Based on the OWASP Top $10$ of
+    $2017$.}
+
+    \item \textbf{Memory Leak.} Memory management issue found more frequently in
+    programming languages that do not manage memory automatically (e.g., C/C++
+    and Objective-C), i.e., where instead developers are responsible for
+    handling it. One of the main causes of DoS attacks.
+
     \item \textbf{Miscellaneous} This pattern comprises several other categories
@@
     \textit{Overflow}, \textit{Path Traversal}.
-
-    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
-    allow attackers to submit untrusted data to web browsers through malicious
-    scripts that can hijack the user sessions or redirect the user to malicious
-    sites. \textit{Based on the OWASP Top $10$ of $2017$.}
-
-    \item \textbf{Injection.} When developers do not keep untrusted data
-    separate from commands and queries. If an attacker sends a string that
-    exploits the syntax of the interpreter, then an injection attack is possible
-    (e.g., SQL and LDAP injection). \textit{Based on the OWASP Top $10$ of
-    $2017$.}
-
-    \item \textbf{Memory Leak.} Memory management issue found more frequently in
-    programming languages that do not manage memory automatically (e.g., C/C++
-    and Objective-C), i.e., where instead developers are responsible for
-    handling it. One of the main causes of DoS attacks.
 \end{itemize}
@@
 
-The dataset contemplates security flaws for more than $13$ different languages
+The dataset contains security flaws for more than $13$ different languages
 being PHP ($39\%$) and C ($24\%$) the most prevalent ones; see Figure~\ref{fig:lang}.

### 2019-02-04 paper/ICPC19/icpc19.tex
more fixes

@@
 
-The dataset contains security flaws for more than $13$ different languages
-being PHP ($39\%$) and C ($24\%$) the most prevalent ones; see Figure~\ref{fig:lang}.
+The dataset contains security flaws for more than $13$ different languages,
+being PHP ($39\%$) and C ($24\%$) the most prevalent ones (see Figure~\ref{fig:lang}).
 To classify the applications in the dataset, we use a taxonomy for open-source

### 2019-02-04 paper/ICPC19/icpc19.tex
more fixes

@@
 number of stars, the number of watchers, the number of contributors, the number
-of commits, the number of branches, the number of releases,  size (cf. GitHub it
+of commits, the number of branches, the number of releases,  size (cf. GitHub, it
 is the size of the whole repository, including all of its history, in
@@
 with the guideline. These thresholds are calibrated by BCH using their own
-data/experience --- open-source and closed software systems. If a project is
+data/experience --- using open-source and closed software systems. If a project is
 compliant with a guideline, it means that it is at least $65\%$ better than the

### 2019-02-04 paper/ICPC19/icpc19.tex
pass over paper

@@
 
-In total, this study evaluates $607$ security and baseline commits from $94$
-different open-source projects distributed among $5$ different main types of
-software, Figure~\ref{fig:domain}. The following section presents the results
-obtained for each research question.
+As mentioned before, this study evaluates $607$ security and baseline commits
+from $94$ distinct open-source projects, distributed amongst $5$ main
+types of software applications. The following section presents the
+results obtained for each research question.
 
@@
 are presented: \emph{negative}, maintainability decreases (\emph{red});
-\emph{none}, maintainability stays the same (\emph{yellow}); and,
-\emph{positive}, maintainability increases (\emph{green}). In front of each
+\emph{none}, maintainability remains the same (\emph{yellow}); and,
+\emph{positive}, maintainability increases (\emph{green}). Next to each
 group, it is presented the mean ($\overline{x}$) and median ({$\widetilde{x}$)
+\Rui{notation is not consistent... use M}
 of the maintainability difference and the p-value resulting from the Wilcoxon
-signed-rank test.
+signed-rank test. \Rui{this is not consistent to the statistical section...}
 
@@
 Regarding security commits results, we observe that maintainability decreases in
-$40.20\%$ ($244$), stays equal in $30.5\%$ ($185$) and increases in $29.33\%$
-($178$) of the cases after security refactorings being applied. The resulting
-p-value of the Wilcoxon signed-rank test is $1.21$x$10^{-10}$. Since the p-value
-is below the significance level of $0.05$, it is reasonable to affirm that
+$40.20\%$ ($244$), remains the same in $30.5\%$ ($185$) and increases in $29.33\%$
+($178$) of the cases after security refactorings are applied. The resulting
+p-value of the Wilcoxon signed-rank \Rui{consistentcy...} test is $1.21$x$10^{-10}$.
+Since the p-value is below the significance level of $0.05$, we argue that
 \emph{security refactorings have a negative impact in the maintainability of
@@
 
-For regular commits, we also observe that maintainability decreases in $41.35\%$
-($251$) of the cases after regular refactorings being applyed. But in contrast
-to security commits, it increases in $58.16\%$ ($353$) and stays equal in
-$0.5\%$ ($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank
+For regular commits, we observe that the maintainability decreases in $41.35\%$
+($251$) of the cases after regular refactorings are applied. But in contrast
+to security commits, it increases in $58.16\%$ ($353$) and remains the same in
+$0.5\%$ ($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank\Rui{consistency..}
 test is $1.54$x$10^{-6}$. Since the p-value is below the significance level of
-$0.05$, it is reasonable to affirm that \emph{regular refactorings have a
+$0.05$, we argue that \emph{regular refactorings have a
 positive impact in the maintainability of open-source software}.
@@
 
-In this study, we also evaluated the impact of each programming language on the
-maintainability of open-source software, Figure~\ref{fig:lang_main}. Although
+We have also evaluated the impact in the maintainability per programming language
+of open-source software, Figure~\ref{fig:lang_main}. Although
 the dataset contains security refactorings for more than $13$ languages, we only
 present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP}
-because one of the requirements of the Wilcoxon signed-rank test is that the
-input sample has to have more than $20$ measurements. The other languages were
-integrated into the \emph{Others} set. The results of the Wilcoxon signed-rank
-test by pattern returned statistical evidence which supports that the
+because one of the requirements of the Wilcoxon signed-rank test\Rui{consistency} is that the
+input sample has to have more than $20$ samples. The other languages were
+integrated into \emph{Others}. The results of the Wilcoxon signed-rank\Rui{consistency}
+test by pattern indicate that there is statistical evidence to support that the
 maintainability of open-source software decreases for security refactorings in
-$3$ different languages: \emph{C} (p = $5.72$x$10^{-9}$), Ruby (p = $0.057$) and
-PHP (p = $3.54$x$10^{-6}$). The results also provide statistical significance
-for when security refactorings are performed in \emph{Java} (p =
+$3$ different languages: \emph{C} (p-value = $5.72$x$10^{-9}$), Ruby (p-value = $0.057$) and
+PHP (p-value = $3.54$x$10^{-6}$). The results also provide statistical significance
+for when security refactorings are performed in \emph{Java} (p-value =
 $3.61$x$10^{-2}$). \emph{Python} and the languages that integrate the
@@
 open-source software is presented in Figure~\ref{fig:pat}. The results of the
-Wilcoxon signed-rank test by pattern returned statistical evidence which
-supports that the maintainability of open-source software decreases for $4$
-different patterns: \emph{Broken Autentication} (p = $0.018$),
-\emph{Denial-of-Service} (p = $0.029$), \emph{Memory Leak} (p = $0.009$) and
-\emph{Miscellaneous} (p = $3.07$x$10^{-4}$). With statistical significance,
+Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical evidence
+supporting that the maintainability of open-source software decreases for $4$
+different patterns: \emph{Broken Autentication} (p-value = $0.018$),
+\emph{Denial-of-Service} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
+\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical significance,
 results demonstrate evidence that maintainability is not negatively impacted
-when removing \emph{Components with Known Vulnerabilities} (p = $7.96$x$10^{-4}$)
-and fixing \emph{Cross-Site Scripting Vulnerabilities} (p = $0.001$).
+when removing \emph{Components with Known Vulnerabilities} (p-value = $7.96$x$10^{-4}$)
+and fixing \emph{Cross-Site Scripting Vulnerabilities} (p-value = $0.001$).
 \emph{Injection} and \emph{Cross-Site Request Forgery} produced more
@@
 We found that $40.2\%$ of the security refactoring have a negative impact on the
-maintainability of open-source software, Figure~\ref{fig:secvsreg}. This raises
+maintainability of open-source software (Figure~\ref{fig:secvsreg}). This raises
 a new tradeoff when developers need to fix vulnerabilities on their projects
@@
 	vulnerabilities that stay undisclosed. Thus, developers should not trust them
-	blindly and check defaults.
+	blindly and check defaults. \Rui{I don't understand the point.}
 
@@
 	the combination of \texttt{HTTP} filters with \texttt{Spring Security}
-	annotations.
+	annotations. \Rui{not crystal clear -- do you mean programming languages
+  should offer these mechanisms? What is that combination that you're talking
+  about?}
 
@@
 	more aware of the maintainability of their code and issues (e.g., dead and
-	duplicated code).
+	duplicated code). \Rui{not clear. deprecated code vs. BCH??}
 
 	\item\textbf{Programming Languages} should provide new mechanisms to easily
-	implement security patterns without endangering maintainability.
+	implement security patterns without endangering maintainability. \Rui{merge with
+  3rd bullet?}
 \end{itemize}
 
+\Rui{Intense, buzz --- this is not a blog. Don't understand the point very
+well too.}
 We would expect that the negative impact of programming languages on
@@
 important to provide tool support to developers to help them apply these
-patterns without harming software maintainability.
+patterns without harming software maintainability. \Rui{and redesign programming
+languages?}
 
@@
 maintainability standards. However, BCH does have a representative benchmark of
-closed and open-source software projects that are under the calculation of the
-thresholds used in the computation of each maintainability guideline
-(Table~\ref{tab:guidelines})~\cite{Visser:2016:OREILLY, Baggen2012}.
+closed and open-source software projects that are used to compute the
+thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
 
@@
 All the security commits are described in English. This approach discards
-security refactorings provided in other languages.
+security refactorings provided in other languages.\Rui{ehmm, isn't this
+related to the first paragraph in the internal threat?}
 
@@
 Many studies have investigated the relationship between refactorings and
-software quality. Previous work focused on coupling metrics has evaluated the
+software quality. Previous work focused on coupling metrics \Rui{?} has evaluated the
 impact of refactorings and exhibited proof that quantifying effectively the
@@
 Other studies investigated the effect of programming languages in software
-quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
+quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}\Rui{quality or maintainability?}. The first
 study shows that some programming languages are more buggy than others, i.e.,
@@
 evidence about the impact of language design. Both have used the number of
-defects as an indicator of software quality. The present work studies how
+defects as an indicator of software quality. Our works studies how
 security refactorings affect software quality based on the code maintainability
-analysis and provides some evidence that programming language have impact on
+analysis and provides evidence that programming language have impact on
 maintainability.
@@
 refactorings on the maintainability of $94$ open-source projects. We leveraged
-Better Code Hub reports of each project version to calculate maintainability
+Better Code Hub reports to calculate maintainability
 based on a model proposed in previous work. Security refactorings significantly
@@
 
-This study also exhibits evidence that some security patterns need more attention than
+This study also shows evidence that there are security patterns that need more attention than
 others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and
@@
 respectively. Following the implications of our work, we draw attention for the
-need of trustful security frameworks, code best practices documentation to
+need of trustful security frameworks \Rui{?}, code best practices documentation to
 produce safe and maintainable code, proper authorization mechanisms design and
@@
 As future work, this study may be extended in many directions: analyze which
-guidelines are more likely to affect security refactorings; evaluate the impact
-of each refactoring pattern on software maintainability; investigate if there is
-any correlation between software maintainability and projects popularity; expand
+guidelines are more likely to affect security refactorings \Rui{? }; evaluate the impact
+of each refactoring pattern on software maintainability \Rui{? }; investigate if there is
+any correlation between software maintainability and popularity of the project; expand
 our methodology with other software quality properties; validate these findings

### 2019-02-04 paper/ICPC19/icpc19.tex
pass over paper
some fixes

@@
 checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
-$13$) and then against the maximum handshake message size allowed by DTLS1 for
-\texttt{s} (line $16$-$21$). However, it is not verified if
-\texttt{msg\_hdr->msg\_len} is lower than the allowed by DTLS. Thus, in line
+$13$) and then against the maximum size allowed in a DTLS handshake message for \texttt{s} (line $16$-$21$)\Rui{DTLS1 for s?}.
+However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the size allowed by DTLS.
+\Rui{$<$- does not parse} Thus, in line
 $37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
@@
 inserted to return the maximum number of bytes allowed in a DTLS
-handshake message for \texttt{s}. The minimum is $16$KB but may be greater if
-the maximum certificate list size requires it which is set by the user. More
+handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if the maximum certificate list size requires it. \Rui{$<$- does not parse} More
 specifically, it checks if the maximum certificate list size
@@
 \begin{itemize}
-    \item \textbf{Using Components with Known Vulnerabilities.} The majority of
+    \item \textbf{Using Components with Known Vulnerabilities (21 cases).} The majority of
     software produced today integrates several components, such as libraries and
@@
 
-    \item \textbf{Broken Authentication \& Session Management.} Incorrectly
+    \item \textbf{Broken Authentication \& Session Management (38 cases).} Incorrectly
     implemented functionalities related to authentication and session
@@
 
-    \item \textbf{Cross-Site Request Forgery.} Poor session tokens generation
+    \item \textbf{Cross-Site Request Forgery (30 cases).} Poor session tokens generation
     and management usually allow attackers to send forged HTTP requests
@@
 
-    \item \textbf{Denial-of-Service} Security flaws that can allow an attacker
+    \item \textbf{Denial-of-Service (40 cases).} Security flaws that can allow an attacker
     to flood or crashing services. This attacks usually occur when the system
@@
 
-    \item \textbf{Cross-Site Scripting.} Lack of proper validation or escaping
+    \item \textbf{Cross-Site Scripting (114 cases).} Lack of proper validation or escaping
     allow attackers to submit untrusted data to web browsers through malicious
@@
 
-    \item \textbf{Injection.} When developers do not keep untrusted data
+    \item \textbf{Injection (89 cases).} When developers do not keep untrusted data
     separate from commands and queries. If an attacker sends a string that
@@
 
-    \item \textbf{Memory Leak.} Memory management issue found more frequently in
+    \item \textbf{Memory Leak (80 cases).} Memory management issue found more frequently in
     programming languages that do not manage memory automatically (e.g., C/C++
@@
 
-    \item \textbf{Miscellaneous} This pattern comprises several other categories
+    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several other categories
     of security refactorings (e.g., \textit{path traversal and buffer
@@
     their own pattern yet. The patterns not satisfying the size requirement of
-    more than $20$ refactorings are contemplated here: \textit{Resource Leak},
-    \textit{SHA-1 Hash Function}, \textit{Broken Access Control},
-    \textit{Sensitive Data Exposure}, \textit{Insufficient Attack Protection},
-    \textit{Overflow}, \textit{Path Traversal}.
+    more than $20$ refactorings are the following ones: \textit{Resource Leak} (19),
+    \textit{SHA-1 Hash Function} (1), \textit{Broken Access Control} (2),
+    \textit{Sensitive Data Exposure} (18), \textit{Insufficient Attack Protection} (4),
+    \textit{Overflow} (18), \textit{Path Traversal} (19).
 \end{itemize}
@@
 
-\emph{Units} are the smallest groups of code that can be maintained and executed
-independently~\cite{Visser:2016:OREILLY}. McCabe~\cite{1702388} is a software
+\emph{Units} are the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY}.
+
+McCabe~\cite{1702388} is a software
 metric that calculates the number of linearly independent paths of the source
@@
 \emph{positive}, maintainability increases (\emph{green}). Next to each
-group, it is presented the mean ($\overline{x}$) and median ({$\widetilde{x}$)
-\Rui{notation is not consistent... use M}
+group, it is presented the mean ($\overline{x}$) and median (M)
 of the maintainability difference and the p-value resulting from the Wilcoxon
@@
 Other studies investigated the effect of programming languages in software
-quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}\Rui{quality or maintainability?}. The first
+quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
 study shows that some programming languages are more buggy than others, i.e.,
@@
 As future work, this study may be extended in many directions: analyze which
-guidelines are more likely to affect security refactorings \Rui{? }; evaluate the impact
-of each refactoring pattern on software maintainability \Rui{? }; investigate if there is
+guidelines are more likely to be affected by security refactorings \Rui{?}\Sof{melhor?}; evaluate the impact
+of each refactoring pattern on software maintainability \Rui{?}\Sof{}; investigate if there is
 any correlation between software maintainability and popularity of the project; expand
@@
 help in validating the methodology.
+%ack pedro adao when accepted.
 

### 2019-02-04 paper/ICPC19/icpc19.tex
conclusion fix

@@
 As future work, this study may be extended in many directions: analyze which
-guidelines are more likely to be affected by security refactorings \Rui{?}\Sof{melhor?}; evaluate the impact
-of each refactoring pattern on software maintainability \Rui{?}\Sof{}; investigate if there is
-any correlation between software maintainability and popularity of the project; expand
-our methodology with other software quality properties; validate these findings
-with closed or private software; and, expand this analysis to other quality
-standards.
+guidelines are more likely to be affected by security-related refactorings;
+evaluate the impact of each refactoring pattern on software maintainability
+\Rui{?}\Sof{}; investigate if there is any correlation between software
+maintainability and popularity of the project; expand our methodology with other
+software quality properties; validate these findings with closed or private
+software; and, expand this analysis to other quality standards.
 

### 2019-02-04 paper/ICPC19/icpc19.tex
issues fixed until S3 and the end of subsection B

@@
 checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
-$13$) and then against the maximum size allowed in a DTLS handshake message for \texttt{s} (line $16$-$21$)\Rui{DTLS1 for s?}.
+$13$) and then against the maximum size allowed in a DTLS handshake message for \texttt{s} (line $16$-$21$)\Rui{DTLS1 for s?}\Sof{Better?}.
 However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the size allowed by DTLS.
-\Rui{$<$- does not parse} Thus, in line
+\Rui{$<$- does not parse}\Sof{Better?} Thus, in line
 $37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
@@
 inserted to return the maximum number of bytes allowed in a DTLS
-handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if the maximum certificate list size requires it. \Rui{$<$- does not parse} More
+handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if the maximum certificate list size requires it. \Rui{$<$- does not parse}\Sof{Better?} More
 specifically, it checks if the maximum certificate list size
@@
 \begin{itemize}
-    \item \textbf{Using Components with Known Vulnerabilities (21 cases).} The majority of
+    \item \Sof{Cases added, check it out}\textbf{Using Components with Known Vulnerabilities (21 cases).} The majority of
     software produced today integrates several components, such as libraries and
@@
 
-\emph{Units} are the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY}.
+One of the guidelines with which the project does not comply is the one presented in the report (Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}.
+SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java).
 
@@
 code. BCH uses this metric to measure the number of branch points in the
-\emph{Write Simple Units of Code} guideline. The bar at the bottom of the figure
+\emph{Write Simple Units of Code} guideline.
+
+\emph{Units} are the smallest sets of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY}. For instance, in Java
+units are methods and contructors.
+
+The bar at the bottom of the figure
 represents the top $30$ units that violate the guideline, sorted by severity,
@@
 \begin{itemize}
-	\item \textbf{Security Frameworks} need to be secure and
-	user-friendly~\cite{5287006, 7676144}. Unfortunately, there are several
-	vulnerabilities that stay undisclosed. Thus, developers should not trust them
-	blindly and check defaults. \Rui{I don't understand the point.}
+	% \item \textbf{Security Frameworks} need to be secure and
+	% user-friendly~\cite{5287006, 7676144}. Unfortunately, there are several
+	% vulnerabilities that stay undisclosed. Thus, developers should not trust them
+	% blindly and check defaults. \Rui{I don't understand the point.}
 
@@
 	code is difficult to test, analyze and re-use.
-
-	\item \textbf{Proper Authorization Mechanisms} should be designed and
+	\item\textbf{Programming Languages} should provide new mechanisms to easily
+	implement security patterns without endangering maintainability. \Rui{merge with
+  3rd bullet?}
+	Proper Authorization Mechanisms should be designed and
 	implemented. This is one of the most complex security features to implement.
@@
 
-	\item \textbf{Deprecated code} should be avoided by developers. They should
+	\item \textbf{Software Developers} should be avoided by developers. They should
 	integrate tools such as BCH on their projects CI/CD pipelines to help them be
-	more aware of the maintainability of their code and issues (e.g., dead and
-	duplicated code). \Rui{not clear. deprecated code vs. BCH??}
-
-	\item\textbf{Programming Languages} should provide new mechanisms to easily
-	implement security patterns without endangering maintainability. \Rui{merge with
-  3rd bullet?}
+	more aware of the maintainability of their code and issues.
+
 \end{itemize}
@@
 well too.}
-We would expect that the negative impact of programming languages on
-maintainability would be more intense due to the buzz of the last years
-surrounding the idea that poor design of programming languages and lack of using
-the best practices lead to more buggy and vulnerable
-code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
+
+% Ray et al. (2017)~\cite{Ray:2017:LSP:3144574.3126905} and
+% Berger et al. (2019)~\cite{2019arXiv190110220B}
+%
+% It would be expected that programming languages would have a more significant impact on the software maintainability according with previous work.
+%
+% According with the implications of previous work~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}, we would expect that the negative impact of programming languages on
+% maintainability would be more significant
+% surrounding the idea that poor design of programming languages and lack of using
+% the best practices lead to more buggy and vulnerable
+% code.
+
+However,
 Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of
@@
 important to provide tool support to developers to help them apply these
-patterns without harming software maintainability. \Rui{and redesign programming
+patterns without harming software maintainability. It is also important to redesign programming languages. \Rui{and redesign programming
 languages?}
@@
 security refactorings provided in other languages.\Rui{ehmm, isn't this
-related to the first paragraph in the internal threat?}
+related to the first paragraph in the internal threat?}\Sof{I think we can close this one}
 
@@
 Many studies have investigated the relationship between refactorings and
-software quality. Previous work focused on coupling metrics \Rui{?} has evaluated the
+software quality. Previous work focused on object-oriented metrics \Rui{?}\Sof{Better?} has evaluated the
 impact of refactorings and exhibited proof that quantifying effectively the
@@
 \emph{Memory Leaks} endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$,
-respectively. Following the implications of our work, we draw attention for the
-need of trustful security frameworks \Rui{?}, code best practices documentation to
+respectively. \Sof{Check this par again}Following the implications of our work, we draw attention for the
+need of code best practices documentation to
 produce safe and maintainable code, proper authorization mechanisms design and
@@
 guidelines are more likely to be affected by security-related refactorings;
-evaluate the impact of each refactoring pattern on software maintainability
-\Rui{?}\Sof{}; investigate if there is any correlation between software
+investigate what are the most frequent vulnerabilities in each programming language\Rui{?}\Sof{Better?}; study if there is any correlation between software
 maintainability and popularity of the project; expand our methodology with other

### 2019-02-04 paper/ICPC19/icpc19.tex
cf figure

@@
 
-One of the guidelines with which the project does not comply is the one presented in the report (Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}.
-SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java).
+\Luis{(Figure -- (cf. Figure [mesmo para outras refs semelhantes que estao por
+ai espalhadas]}
+One of the guidelines with which the project does not comply is
+the one presented in the report (Figure~\ref{fig:bchrep}): \emph{Write Simple
+Units of Code}. SIG defines \emph{Units} as the smallest groups of code that
+can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g.,
+methods and constructors in Java).
 

### 2019-02-04 paper/ICPC19/icpc19.tex
median of differences is zero

@@
   \includegraphics[width=0.56\textwidth]{figures/language.pdf}
- 	\caption{Maintainability difference by language after security refactorings}
+  \caption{Maintainability difference by language after security refactorings.}
   \label{fig:lang_main}
@@
  	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
- 	\caption{Maintainability difference by pattern after security refactorings}
+ 	\caption{Maintainability difference by pattern after security refactorings.
+  \Luis{a median e' zero em muitos casos -- como interpretam isso? se nao houver
+boa/simples explicacao talvez seja melhor usar o common language.}}
 	\label{fig:pat}

### 2019-02-04 paper/ICPC19/icpc19.tex
update

@@
 
-\Luis{(Figure -- (cf. Figure [mesmo para outras refs semelhantes que estao por
-ai espalhadas]}
-One of the guidelines with which the project does not comply is
-the one presented in the report (Figure~\ref{fig:bchrep}): \emph{Write Simple
-Units of Code}. SIG defines \emph{Units} as the smallest groups of code that
-can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g.,
-methods and constructors in Java).
+One of the guidelines with which the project does not comply is the one presented in the report (cf.
+Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}.
+SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java).
+BCH reports
 
@@
 \emph{Write Simple Units of Code} guideline.
-
-\emph{Units} are the smallest sets of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY}. For instance, in Java
-units are methods and contructors.
 

### 2019-02-04 paper/ICPC19/icpc19.tex
cleaning fixes

@@
 checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
-$13$) and then against the maximum size allowed in a DTLS handshake message for \texttt{s} (line $16$-$21$)\Rui{DTLS1 for s?}\Sof{Better?}.
+$13$) and then against the maximum size allowed in a DTLS handshake message for \texttt{s} (line $16$-$21$).
 However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the size allowed by DTLS.
-\Rui{$<$- does not parse}\Sof{Better?} Thus, in line
+Thus, in line
 $37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
@@
 inserted to return the maximum number of bytes allowed in a DTLS
-handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if the maximum certificate list size requires it. \Rui{$<$- does not parse}\Sof{Better?} More
+handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if the maximum certificate list size requires it. More
 specifically, it checks if the maximum certificate list size
@@
 \begin{itemize}
-    \item \Sof{Cases added, check it out}\textbf{Using Components with Known Vulnerabilities (21 cases).} The majority of
+    \item \textbf{Using Components with Known Vulnerabilities (21 cases).} The majority of
     software produced today integrates several components, such as libraries and
@@
 SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java).
-BCH reports
+BCH evaluates this guideline based on the number
 

### 2019-02-04 paper/ICPC19/icpc19.tex
comentarios listing

@@
 cross-site scripting vulnerability since usually the last one is can be fixed
-without adding new lines of code, as seen in Figure~\ref{lst:fix}. The developer
-added the function \emph{htmlentities} to \textit{escape} the data given by the
-variable \emph{\$\_['file']}. Knowing which patterns are more likely to increase
-maintainability issues is one step forward to bring awareness to security
-engineers of what patterns need more attention. The taxonomy of security
-patterns used to answer this question is the same as the one provided by the
-authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
-separately for each pattern.
+without adding new lines of code, as seen in Figure~\ref{lst:fix}.\Luis{as seen
+in Listing?} The developer added the function \emph{htmlentities} to
+\textit{escape} the data given by the variable \emph{\$\_['file']}.\Luis{usa
+sempre o mesmo tipo de letra para nomes de funcoes. \\texttt?} Knowing which
+patterns are more likely to increase maintainability issues is one step forward
+to bring awareness to security engineers of what patterns need more attention.
+The taxonomy of security patterns used to answer this question is the same as
+the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
+Maintainability is measured separately for each pattern.
 
@@
 -      <p class='hint'><?php if(isset($_['file'])) echo $_['file']?></p>
-+      <p class='hint'><?php if(isset($_['file'])) echo htmlentities($_['file'])?></p>
++      <p class='hint'><?php if(isset($_['file'])) echo htmlentities($_['file'])?><?php 
     </li>
@@
 
-\end{lstlisting}
+\end{lstlisting} \Luis{no listing lst:fix o que achas de fazer linebreak para
+cada uma das tags? assim nao havi o problema de ter a mesma linha dividida e ficava mais facil de ler:\\
+\texttt{<p class='hint'>}\\
+\texttt{~~<?php }\\
+\texttt{</p>}\\
+Also: reduz numero de espacos para cada tab -- 2 é suficiente.
+}
 %

### 2019-02-04 paper/ICPC19/icpc19.tex
more issues solved

@@
 cross-site scripting vulnerability since usually the last one is can be fixed
-without adding new lines of code, as seen in Figure~\ref{lst:fix}.\Luis{as seen
-in Listing?} The developer added the function \emph{htmlentities} to
-\textit{escape} the data given by the variable \emph{\$\_['file']}.\Luis{usa
-sempre o mesmo tipo de letra para nomes de funcoes. \\texttt?} Knowing which
+without adding new lines of code, as seen in Listing~\ref{lst:fix}. The developer added the function \emph{htmlentities} to
+\textit{escape} the data given by the variable \texttt{\$\_['file']}. Knowing which
 patterns are more likely to increase maintainability issues is one step forward
@@
 All the security commits are described in English. This approach discards
-security refactorings provided in other languages.\Rui{ehmm, isn't this
-related to the first paragraph in the internal threat?}\Sof{I think we can close this one}
+security refactorings provided in other languages.
 
@@
 Many studies have investigated the relationship between refactorings and
-software quality. Previous work focused on object-oriented metrics \Rui{?}\Sof{Better?} has evaluated the
+software quality. Previous work focused on object-oriented metrics has evaluated the
 impact of refactorings and exhibited proof that quantifying effectively the
@@
 guidelines are more likely to be affected by security-related refactorings;
-investigate what are the most frequent vulnerabilities in each programming language\Rui{?}\Sof{Better?}; study if there is any correlation between software
+investigate what are the most frequent vulnerabilities in each programming language; study if there is any correlation between software
 maintainability and popularity of the project; expand our methodology with other

### 2019-02-04 paper/ICPC19/icpc19.tex
review partial

--- 2019-02-04 paper/ICPC19/icpc19.tex
@@
 \newcommand\Sof[1]{\nb{Sofia}{red}{#1}}
-\newcommand\Luis[1]{\nb{Luis}{green}{#1}}
+\newcommand\Luis[1]{\nb{Luis}{mygreen}{#1}}
 \newcommand\Rui[1]{\nb{Rui}{blue}{#1}}
@@
 methodology is comprised of the following steps, as illustrated in
-Figure~\ref{fig:met}:
+Figure~\ref{fig:met}\Luis{como usamos numeros na enumeracao que descreve a
+figura, esses numeros tb deveriam estar na figura.}:
 
@@
 
-    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several other categories
-    of security refactorings (e.g., \textit{path traversal and buffer
-    overflow}). It contains several other security refactorings that do not have
-    their own pattern yet. The patterns not satisfying the size requirement of
-    more than $20$ refactorings are the following ones: \textit{Resource Leak} (19),
-    \textit{SHA-1 Hash Function} (1), \textit{Broken Access Control} (2),
-    \textit{Sensitive Data Exposure} (18), \textit{Insufficient Attack Protection} (4),
-    \textit{Overflow} (18), \textit{Path Traversal} (19).
-\end{itemize}
+    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several
+    other categories of security refactorings (e.g., \textit{path traversal and
+    buffer overflow})\Luis{exemplos ja nao precisa estar aqui porque sao
+    enumerados logo a seguir}. It contains several other security refactorings
+    that do not have their own pattern yet. The patterns not satisfying the
+    size requirement of more than $20$ refactorings are the following ones:
+    \textit{Resource Leak} (19), \textit{SHA-1 Hash Function} (1),
+    \textit{Broken Access Control} (2), \textit{Sensitive Data Exposure} (18),
+    \textit{Insufficient Attack Protection} (4), \textit{Overflow} (18),
+    \textit{Path Traversal} (19).\Luis{Acho que este item pode ser reescrito de forma mais clara. sugestao:
+    This category comprises the security refactorings that (1) do not relate to
+    any particular pattern or (2) relate to patterns that do not occur in at
+    least 20 refactorings: \textit{Resource Leak} (19), \dots} \end{itemize}
 
@@
 software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
-presents the projects domain distribution: \textit{Aplication Software} ($25$),
+presents the projects domain distribution: \textit{Application Software} ($25$),
 software that provides end-users with functional systems; \textit{Web libraries

### 2019-02-04 paper/ICPC19/icpc19.tex
more reviews

@@
 BCH evaluates this guideline based on the number
+\Luis{this seems unfinished. i think we ve already mention this Mccabe metric somewhere else.}
 
@@
 open-source software is presented in Figure~\ref{fig:pat}. The results of the
-Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical evidence
-supporting that the maintainability of open-source software decreases for $4$
-different patterns: \emph{Broken Autentication} (p-value = $0.018$),
-\emph{Denial-of-Service} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
-\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical significance,
-results demonstrate evidence that maintainability is not negatively impacted
-when removing \emph{Components with Known Vulnerabilities} (p-value = $7.96$x$10^{-4}$)
-and fixing \emph{Cross-Site Scripting Vulnerabilities} (p-value = $0.001$).
-\emph{Injection} and \emph{Cross-Site Request Forgery} produced more
-cases where the refactorings had a negative impact in the maintainability.
-However, the results obtained for these patterns are not statistically
-significant.
+Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical
+evidence supporting that the maintainability of open-source software decreases
+for $4$ different patterns: \emph{Broken Autentication} (p-value = $0.018$),
+\emph{Denial-of-Service}\Luis{em varios pontos deste paper este pattern era
+referido com a sigla DoS. Por uma questao de consistencia com os outros
+patterns sugiro que se escreva sempre por extenso (ou entao sempre com a
+sigla).} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
+\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical
+significance, results demonstrate evidence that maintainability is not
+negatively impacted when removing \emph{Components with Known Vulnerabilities}
+(p-value = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting
+Vulnerabilities} (p-value = $0.001$). \emph{Injection} and \emph{Cross-Site
+Request Forgery} produced more cases where the refactorings had a negative
+impact in the maintainability. However, the results obtained for these patterns
+are not statistically significant.
 

### 2019-02-04 paper/ICPC19/icpc19.tex
more fixes

@@
     frame=single,
-    numbers=left,
     stepnumber=1,
-    xleftmargin=2em,
     escapeinside={/*!}{!*/},
@@
 \newcommand\Sof[1]{\nb{Sofia}{red}{#1}}
-\newcommand\Luis[1]{\nb{Luis}{mygreen}{#1}}
+\newcommand\Luis[1]{\nb{Luis}{green}{#1}}
 \newcommand\Rui[1]{\nb{Rui}{blue}{#1}}
@@
   <li class='error'>
-      <?php echo $l->t('Cloud not found');?><br/>
--     <p class='hint'><?php if(isset($_['file'])) echo $_['file']?></p>
-+     <p class='hint'><?php if(isset($_['file'])) echo htmlentities($_['file'])?><?php 
+   <?php echo $l->t('Cloud not found');?>
+   <br/>
+   <p class='hint'>
+-    <?php if(isset($_['file'])) echo $_['file']?>
+   </p>
+   <p class='hint'>
++    <?php if(isset($_['file'])) echo htmlentities($_['file'])?>
+   </p> 
   </li>
@@
 
-\end{lstlisting} \Luis{no listing lst:fix o que achas de fazer linebreak para
-cada uma das tags? assim nao havi o problema de ter a mesma linha dividida e ficava mais facil de ler:\\
-\texttt{<p class='hint'>}\\
-\texttt{~~<?php }\\
-\texttt{</p>}\\
-Also: reduz numero de espacos para cada tab -- 2 é suficiente.
-}
+\end{lstlisting} % \Luis{no listing lst:fix o que achas de fazer linebreak para
+% cada uma das tags? assim nao havi o problema de ter a mesma linha dividida e ficava mais facil de ler:\\
+% \texttt{<p class='hint'>}\\
+% \texttt{~~<?php }\\
+% \texttt{</p>}\\
+% Also: reduz numero de espacos para cada tab -- 2 é suficiente.
+% }
 %
@@
 methodology is comprised of the following steps, as illustrated in
-Figure~\ref{fig:met}\Luis{como usamos numeros na enumeracao que descreve a
-figura, esses numeros tb deveriam estar na figura.}:
+Figure~\ref{fig:met}:
 
@@
 
-    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several
-    other categories of security refactorings (e.g., \textit{path traversal and
-    buffer overflow})\Luis{exemplos ja nao precisa estar aqui porque sao
-    enumerados logo a seguir}. It contains several other security refactorings
-    that do not have their own pattern yet. The patterns not satisfying the
-    size requirement of more than $20$ refactorings are the following ones:
-    \textit{Resource Leak} (19), \textit{SHA-1 Hash Function} (1),
-    \textit{Broken Access Control} (2), \textit{Sensitive Data Exposure} (18),
-    \textit{Insufficient Attack Protection} (4), \textit{Overflow} (18),
-    \textit{Path Traversal} (19).\Luis{Acho que este item pode ser reescrito de forma mais clara. sugestao:
-    This category comprises the security refactorings that (1) do not relate to
-    any particular pattern or (2) relate to patterns that do not occur in at
-    least 20 refactorings: \textit{Resource Leak} (19), \dots} \end{itemize}
+    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several other categories
+    of security refactorings (e.g., \textit{path traversal and buffer
+    overflow}). It contains several other security refactorings that do not have
+    their own pattern yet. The patterns not satisfying the size requirement of
+    more than $20$ refactorings are the following ones: \textit{Resource Leak} (19),
+    \textit{SHA-1 Hash Function} (1), \textit{Broken Access Control} (2),
+    \textit{Sensitive Data Exposure} (18), \textit{Insufficient Attack Protection} (4),
+    \textit{Overflow} (18), \textit{Path Traversal} (19).
+\end{itemize}
 
@@
 software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
-presents the projects domain distribution: \textit{Application Software} ($25$),
+presents the projects domain distribution: \textit{Aplication Software} ($25$),
 software that provides end-users with functional systems; \textit{Web libraries
@@
 BCH evaluates this guideline based on the number
-\Luis{this seems unfinished. i think we ve already mention this Mccabe metric somewhere else.}
 
@@
 open-source software is presented in Figure~\ref{fig:pat}. The results of the
-Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical
-evidence supporting that the maintainability of open-source software decreases
-for $4$ different patterns: \emph{Broken Autentication} (p-value = $0.018$),
-\emph{Denial-of-Service}\Luis{em varios pontos deste paper este pattern era
-referido com a sigla DoS. Por uma questao de consistencia com os outros
-patterns sugiro que se escreva sempre por extenso (ou entao sempre com a
-sigla).} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
-\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical
-significance, results demonstrate evidence that maintainability is not
-negatively impacted when removing \emph{Components with Known Vulnerabilities}
-(p-value = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting
-Vulnerabilities} (p-value = $0.001$). \emph{Injection} and \emph{Cross-Site
-Request Forgery} produced more cases where the refactorings had a negative
-impact in the maintainability. However, the results obtained for these patterns
-are not statistically significant.
+Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical evidence
+supporting that the maintainability of open-source software decreases for $4$
+different patterns: \emph{Broken Autentication} (p-value = $0.018$),
+\emph{Denial-of-Service} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
+\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical significance,
+results demonstrate evidence that maintainability is not negatively impacted
+when removing \emph{Components with Known Vulnerabilities} (p-value = $7.96$x$10^{-4}$)
+and fixing \emph{Cross-Site Scripting Vulnerabilities} (p-value = $0.001$).
+\emph{Injection} and \emph{Cross-Site Request Forgery} produced more
+cases where the refactorings had a negative impact in the maintainability.
+However, the results obtained for these patterns are not statistically
+significant.
 

### 2019-02-05 paper/ICPC19/icpc19.tex
more fixes down

@@
 first steps taken by SIG to introduce security on their maintainability model). 
-
 BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
@@
 between refactoring source code to patch disclosed vulnerabilities and keeping
-(or even improving) the maintainability of software systems. Hedegus et al. ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
+(or even improving) the maintainability of software systems. Hedegus et al. 
+($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
 groups of refactored elements and non-refactored and conclude that the source
@@
 checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
-$13$) and then against the maximum size allowed in a DTLS handshake message for \texttt{s} (line $16$-$21$).
-However, it is not verified if \texttt{msg\_hdr->msg\_len} is lower than the size allowed by DTLS.
-Thus, in line
+$13$) and then against the maximum size allowed in a DTLS handshake message for 
+\texttt{s} (line $16$-$21$). However, it is not verified if \texttt{msg\_hdr->msg\_len} 
+is lower than the size allowed by DTLS. Thus, in line
 $37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
@@
 inserted to return the maximum number of bytes allowed in a DTLS
-handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if the maximum certificate list size requires it. More
+handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if 
+the maximum certificate list size requires it. More
 specifically, it checks if the maximum certificate list size
@@
    <?php echo $l->t('Cloud not found');?>
-   <br/>
    <p class='hint'>
@@
 
-\end{lstlisting} % \Luis{no listing lst:fix o que achas de fazer linebreak para
-% cada uma das tags? assim nao havi o problema de ter a mesma linha dividida e ficava mais facil de ler:\\
-% \texttt{<p class='hint'>}\\
-% \texttt{~~<?php }\\
-% \texttt{</p>}\\
-% Also: reduz numero de espacos para cada tab -- 2 é suficiente.
-% }
-%
+\end{lstlisting}
+
 \section{Methodology}\label{sec:methodology}
@@
 
-    \item \textbf{Miscellaneous (195 cases)} This pattern comprises several
-    other categories of security refactorings (e.g., \textit{path traversal and
-    buffer overflow})\Luis{exemplos ja nao precisa estar aqui porque sao
-    enumerados logo a seguir}. It contains several other security refactorings
-    that do not have their own pattern yet. The patterns not satisfying the
-    size requirement of more than $20$ refactorings are the following ones:
+    \item \textbf{Miscellaneous (195 cases)} This pattern comprises 
+	several other security refactorings that do not have their own pattern yet and patterns that
+	do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings:
     \textit{Resource Leak} (19), \textit{SHA-1 Hash Function} (1),
@@
     \textit{Insufficient Attack Protection} (4), \textit{Overflow} (18),
-    \textit{Path Traversal} (19).\Luis{Acho que este item pode ser reescrito de forma mais clara. sugestao:
-    This category comprises the security refactorings that (1) do not relate to
-    any particular pattern or (2) relate to patterns that do not occur in at
-    least 20 refactorings: \textit{Resource Leak} (19), \dots} \end{itemize}
+    \textit{Path Traversal} (19). \end{itemize}
 
@@
 
-One of the guidelines with which the project does not comply is the one presented in the report (cf.
-Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}.
-SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java).
-BCH evaluates this guideline based on the number
-\Luis{this seems unfinished. i think we ve already mention this Mccabe metric somewhere else.}
-
-McCabe~\cite{1702388} is a software
-metric that calculates the number of linearly independent paths of the source
-code. BCH uses this metric to measure the number of branch points in the
-\emph{Write Simple Units of Code} guideline.
-
-The bar at the bottom of the figure
-represents the top $30$ units that violate the guideline, sorted by severity,
+SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed 
+independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java). 
+One of the guidelines with which the project does not comply is the one presented in 
+the report (cf. Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. BCH analyzes 
+this guideline based on the McCabe Complexity~\cite{1702388} to calculate the number of branch 
+points of a method. The bar at the bottom of the figure
+represents the top $30$ units that violate the guideline, sorted by severity.
 The violation severities are indicated using colors, and there is a legend to
-help to interpret the colors. The green bar represents the number of compliant
-branch points per unit (\emph{at most $5$}), i.e., the units are compliant with
+help to interpret them. The green bar represents the number of compliant
+branch points per unit (\emph{at most $5$}), i.e., the number of units are compliant with
 ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
@@
 Aiming to analyze the impact of security refactorings, we use BCH to compute
-maintainability of two different versions of the project (Figure~\ref{fig:commit}):
+maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
 \begin{itemize}
@@
 (e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
-test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968} . In
+test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968}. In
 other words, we test the null hypothesis that the maintainability difference
 between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
-security-commit) follows a symmetric distribution around $0$. However, when the
-difference between groups is zero, the observations are discarded. Thus, we use
-a more robust approach to the Wilcoxon test, the Wilcoxon test by Pratt which
-provides an alternative that incorporates the cases where the maintainability
-difference is zero~\cite{10.2307/2282543}. To understand the effect-size, as
+security-commit) come from the same distribution. Nevertheless, this test has a 
+limitation: it does not consider the groups of commits with a zero-difference 
+maintainability. In $1959$, Pratt provided an improvement to the test to solve 
+this issue making the test more robust. Thus, we use a version of the Wilcoxon 
+test that incorporates the cases where maintainability is different from zero~\cite{10.2307/2282543}. 
+To understand the effect-size, as
 advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
@@
 group, it is presented the mean ($\overline{x}$) and median (M)
-of the maintainability difference and the p-value resulting from the Wilcoxon
-signed-rank test. \Rui{this is not consistent to the statistical section...}
+of the maintainability difference and the p-value resulting from the Paired Wilcoxon signed-rank test.
 
@@
 ($178$) of the cases after security refactorings are applied. The resulting
-p-value of the Wilcoxon signed-rank \Rui{consistentcy...} test is $1.21$x$10^{-10}$.
+p-value of the Paired Wilcoxon signed-rank test is $1.21$x$10^{-10}$.
 Since the p-value is below the significance level of $0.05$, we argue that
@@
 to security commits, it increases in $58.16\%$ ($353$) and remains the same in
-$0.5\%$ ($3$) of the cases. The resulting p-value of the Wilcoxon signed-rank\Rui{consistency..}
+$0.5\%$ ($3$) of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
 test is $1.54$x$10^{-6}$. Since the p-value is below the significance level of
@@
 present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP}
-because one of the requirements of the Wilcoxon signed-rank test\Rui{consistency} is that the
+because one of the requirements of the Paired Wilcoxon signed-rank is that the
 input sample has to have more than $20$ samples. The other languages were
-integrated into \emph{Others}. The results of the Wilcoxon signed-rank\Rui{consistency}
+integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
 test by pattern indicate that there is statistical evidence to support that the
@@
 open-source software is presented in Figure~\ref{fig:pat}. The results of the
-Wilcoxon signed-rank test \Rui{again} per pattern indicated statistical
+Paired Wilcoxon signed-rank per pattern indicated statistical
 evidence supporting that the maintainability of open-source software decreases
 for $4$ different patterns: \emph{Broken Autentication} (p-value = $0.018$),
-\emph{Denial-of-Service}\Luis{em varios pontos deste paper este pattern era
-referido com a sigla DoS. Por uma questao de consistencia com os outros
-patterns sugiro que se escreva sempre por extenso (ou entao sempre com a
-sigla).} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
+\emph{Denial-of-Service} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
 \emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical
@@
   \Luis{a median e' zero em muitos casos -- como interpretam isso? se nao houver
-boa/simples explicacao talvez seja melhor usar o common language.}}
+boa/simples explicacao talvez seja melhor usar o common language.}\Sof{Preciso de pensar melhor nisto}}
 	\label{fig:pat}
@@
 We found that $40.2\%$ of the security refactoring have a negative impact on the
-maintainability of open-source software (Figure~\ref{fig:secvsreg}). This raises
+maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
 a new tradeoff when developers need to fix vulnerabilities on their projects
@@
 \begin{itemize}
-	% \item \textbf{Security Frameworks} need to be secure and
-	% user-friendly~\cite{5287006, 7676144}. Unfortunately, there are several
-	% vulnerabilities that stay undisclosed. Thus, developers should not trust them
-	% blindly and check defaults. \Rui{I don't understand the point.}
-
-	\item \textbf{Documentation} of software should provide developers with the
+	\item \textbf{Documentation} of software must provide developers with the
 	best practices to implement secure and maintainable software. Unmaintainable
 	code is difficult to test, analyze and re-use.
-	\item\textbf{Programming Languages} should provide new mechanisms to easily
-	implement security patterns without endangering maintainability. \Rui{merge with
-  3rd bullet?}
-	Proper Authorization Mechanisms should be designed and
-	implemented. This is one of the most complex security features to implement.
+	\item\textbf{Programming Languages} should provide new design patterns to easily
+	implement security patterns without endangering maintainability. One example is 
+	the need for designing new authorization mechanisms since these are one of the most 
+	complex security features to implement.
 	Some mechanisms might assist authorization maintenance in long run, such as
 	the combination of \texttt{HTTP} filters with \texttt{Spring Security}
-	annotations. \Rui{not crystal clear -- do you mean programming languages
-  should offer these mechanisms? What is that combination that you're talking
-  about?}
-
-	\item \textbf{Software Developers} should be avoided by developers. They should
-	integrate tools such as BCH on their projects CI/CD pipelines to help them be
-	more aware of the maintainability of their code and issues.
+	annotations.
+
+	\item \textbf{Software Developers} lack the knowledge to create secure and 
+	maintainable code. Mainly because academic curriculums are not yet prepared 
+	to educate these matters to them. Universities should do their part on making 
+	this possible. Online services such as BCH have also a great potential to help 
+	developers to be more aware of the maintainability issues introduced by their 
+	changes. Thus, developers can put more work into improving their code maintainability 
+	and avoid common issues such as code duplication.
 
@@
 
-\Rui{Intense, buzz --- this is not a blog. Don't understand the point very
-well too.}
-
-% Ray et al. (2017)~\cite{Ray:2017:LSP:3144574.3126905} and
-% Berger et al. (2019)~\cite{2019arXiv190110220B}
-%
-% It would be expected that programming languages would have a more significant impact on the software maintainability according with previous work.
-%
-% According with the implications of previous work~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}, we would expect that the negative impact of programming languages on
-% maintainability would be more significant
-% surrounding the idea that poor design of programming languages and lack of using
-% the best practices lead to more buggy and vulnerable
-% code.
-
-However,
+We would expect that the negative impact of programming languages on
+maintainability would be more severe given the ideas that poor design of programming languages and lack of using
+the best practices lead to more buggy and vulnerable
+code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
 Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of
@@
 Listing~\ref{lst:fix}. Whereas, \emph{Broken Authentication} and vulnerabilities
-such as the one presented in Section~\ref{sec:motivation} responsible by DoS
+such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
 attacks are more difficult to fix (several code lines deleted and added).
@@
 and fixing vulnerabilities might hinder software maintainability. Thus, it is
-important to provide tool support to developers to help them apply these
-patterns without harming software maintainability. It is also important to redesign programming languages. \Rui{and redesign programming
-languages?}
+not only important to provide tool support to developers to help them apply these
+patterns without harming software maintainability, but also to create and redesign 
+programming languages to facilitate developers improving software maintainability.
 
@@
 impact of refactorings and exhibited proof that quantifying effectively the
-impact of refactoring on maintainability may help to choose the appropriate
+impact of refactorings on maintainability may help to choose the appropriate
 refactoring type~\cite{1167822}. In contrast to this work, Hegedus et
@@
 Other studies investigated the effect of programming languages in software
-quality~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. The first
+quality~\cite{Ray:2014:LSS:2635868.2635922,Ray:2017:LSP:3144574.3126905}. The first
 study shows that some programming languages are more buggy than others, i.e.,
@@
 evidence about the impact of language design. Both have used the number of
-defects as an indicator of software quality. Our works studies how
+defects as an indicator of software quality. In addition, 
+Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922, 
+Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the previously demonstrated 
+correlation between programming language and software defects. Our works studies how
 security refactorings affect software quality based on the code maintainability
@@
 \emph{Memory Leaks} endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$,
-respectively. \Sof{Check this par again}Following the implications of our work, we draw attention for the
-need of code best practices documentation to
-produce safe and maintainable code, proper authorization mechanisms design and
-implementation, better tools to introduce in the CI/CD pipeline to help software
-maintenance and new or improved programming languages to allow developers to
-easily implement safety in their software without endangering maintainability.
+respectively. Following the implications of our work, we draw attention for the
+need of better documentation explaining how to use best practices to produce secure 
+and maintainable code; new and better programming languages; new academic curriculums 
+in the fields of security and software quality; and, tools such as BCH to help developers 
+predit the effect of their patches.
 
@@
 guidelines are more likely to be affected by security-related refactorings;
-investigate what are the most frequent vulnerabilities in each programming language; study if there is any correlation between software
+investigate what are the most frequent vulnerabilities in each programming language; 
+study if there is any correlation between software
 maintainability and popularity of the project; expand our methodology with other

### 2019-02-05 paper/ICPC19/icpc19.tex
a few fixes

@@
 \newboolean{showcomments}
-\setboolean{showcomments}{true}
+\setboolean{showcomments}{false}
 %\setboolean{showcomments}{false}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/bch_report.png}
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
@@
 
-All the security commits are described in English. This approach discards
-security refactorings provided in other languages.
+Our approach does not consider security refactorings in any other languages but
+English.
 
@@
 
-This study also shows evidence that there are security patterns that need more attention than
-others. \emph{Broken Autentication} issues, \emph{Denial-of-Service} attacks and
-\emph{Memory Leaks} endure maintainability in $55.3\%$, $51.3\%$, and $47.5\%$,
-respectively. Following the implications of our work, we draw attention for the
-need of better documentation explaining how to use best practices to produce secure 
-and maintainable code; new and better programming languages; new academic curriculums 
-in the fields of security and software quality; and, tools such as BCH to help developers 
-predit the effect of their patches.
+This study also shows evidence that there are security patterns that need more
+attention than others. \emph{Broken Autentication} issues,
+\emph{Denial-of-Service} attacks and \emph{Memory Leaks} endure maintainability
+in $55.3\%$, $51.3\%$, and $47.5\%$, respectively. Therefore, we draw attention
+for the need of better documentation explaining how to use best practices to
+produce secure and maintainable code; new and better programming languages; new
+academic curriculums in the fields of security and software quality; and, tools
+such as BCH to help developers predit the effect of their patches.
 

### 2019-02-05 paper/ICPC19/icpc19.tex
some more fixes

@@
 
-Other studies investigated the effect of programming languages in software
+There are studies that investigated the impact of programming languages on software
 quality~\cite{Ray:2014:LSS:2635868.2635922,Ray:2017:LSP:3144574.3126905}. The first
-study shows that some programming languages are more buggy than others, i.e.,
-language design has an impact on software quality. However, the authors of the
-second study were not able to reproduce the first one and did not obtain any
+study shows that some programming languages are more buggy-prone than others. However,
+the authors of the second study were not able to reproduce it, not obtaining any
 evidence about the impact of language design. Both have used the number of

### 2019-02-05 paper/ICPC19/icpc19.tex
discussion fixes

@@
 	code is difficult to test, analyze and re-use.
-	\item\textbf{Programming Languages} should provide new design patterns to easily
-	implement security patterns without endangering maintainability. One example is
-	the need for designing new authorization mechanisms since these are one of the most
-	complex security features to implement.
-	Some mechanisms might assist authorization maintenance in long run, such as
-	the combination of \texttt{HTTP} filters with \texttt{Spring Security}
-	annotations.
+
+	\item\textbf{Programming Languages} should provide new design patterns to
+	easily implement security patterns without endangering maintainability.
+	Ultimately, new programming languages, both secure and maintainable by design,
+	may be designed to help developers be highly productive when writing and
+	maintaing  secure applications. One example is the need for designing new
+	authorization mechanisms since these are one of the most complex security
+	features to implement.
 
 	\item \textbf{Software Developers} lack the knowledge to create secure and
-	maintainable code. Mainly because academic curriculums are not yet prepared
-	to educate these matters to them. Universities should do their part on making
+	maintainable code. Mainly because academic curricula are not yet prepared
+	to properly educate them. Universities should do their part on making
 	this possible. Online services such as BCH have also a great potential to help

### 2019-02-05 paper/ICPC19/icpc19.tex
curricula

@@
 produce secure and maintainable code; new and better programming languages; new
-academic curriculums in the fields of security and software quality; and, tools
+academic curricula in the fields of security and software quality; and, tools
 such as BCH to help developers predit the effect of their patches.

### 2019-02-05 paper/ICPC19/icpc19.tex
fixes

@@
 \begin{itemize}
-	\item \textbf{Size project changes do not affect the maintainability
+	\item \textbf{Size of project changes does not affect the maintainability
 	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we
 	aim to evaluate the alike security patterns occurring in different projects.
-	Thus, normalization based on its size cannot be used - we convert relative
+	Thus, normalization based on its size cannot be used --- we convert relative
 	data to the respective number of lines of code.
@@
 \emph{positive}, maintainability increases (\emph{green}). Next to each
-group, it is presented the mean ($\overline{x}$) and median (M)
-of the maintainability difference and the p-value resulting from the Paired Wilcoxon signed-rank test.
+type of change, it is shown the mean ($\overline{x}$) and median (M)
+of the maintainability difference and the p-value resulting from the
+Paired Wilcoxon signed-rank test.
 

### 2019-02-05 paper/ICPC19/icpc19.tex
fixing stuff

@@
 equation to capture the distance between the current state of the project and
-the standard thresholds based on previous work\cite{Olivari:2018}.
+the standard thresholds based on previous work~\cite{Olivari:2018}. The equation
+considers the following:
 \begin{itemize}
 	\item \textbf{Size of project changes does not affect the maintainability
-	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} In this work, we
-	aim to evaluate the alike security patterns occurring in different projects.
+	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} We
+	aim at evaluating security patterns occurring in different projects similarly.
 	Thus, normalization based on its size cannot be used --- we convert relative

### 2019-02-05 paper/ICPC19/icpc19.tex
polishing
Pedro Adao's version

@@
 	aim at evaluating security patterns occurring in different projects similarly.
-	Thus, normalization based on its size cannot be used --- we convert relative
-	data to the respective number of lines of code.
+	Thus, the divised metric uses the \textit{raw} number of lines of code rather
+  percentages for normalization purposes.
 	\item \textbf{Distance to lower severity levels are less penalized than to the
@@
 
-We compute the mean average of the maintainability score $M(v)$ for all the
-selected guidelines, as follows:
+We compute the mean average of the maintainability score $M(v)$ as follows:
 
@@
 \noindent
-where $G$ is the group of selected maintainability guidelines from BCH
+where $G$ is the group of maintainability guidelines from BCH
 (Table~\ref{tab:guidelines}) and $v$ is the version of the software under
@@
   \centering
-  \includegraphics[width=0.56\textwidth]{figures/language.pdf}
+  \includegraphics[width=0.5\textwidth]{figures/language.pdf}
   \caption{Maintainability difference by language after security refactorings.}

### 2019-02-05 paper/ICPC19/icpc19.tex
polishing
fix
meaning of M(v)

@@
 
-SIG defines \emph{Units} as the smallest groups of code that can be maintained and executed
-independently~\cite{Visser:2016:OREILLY} (e.g., methods and constructors in Java).
-One of the guidelines with which the project does not comply is the one presented in
-the report (cf. Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. BCH analyzes
-this guideline based on the McCabe Complexity~\cite{1702388} to calculate the number of branch
-points of a method. The bar at the bottom of the figure
-represents the top $30$ units that violate the guideline, sorted by severity.
-The violation severities are indicated using colors, and there is a legend to
-help to interpret them. The green bar represents the number of compliant
-branch points per unit (\emph{at most $5$}), i.e., the number of units are compliant with
-ISO 25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do
-not comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
-(\emph{above $25$}) severity levels. In the bar, there are marks that pinpoint the
-compliance thresholds for each severity level. If the green mark is somewhere
-in the green bar it means it is compliant with a low level of severity.
+SIG defines \emph{Units} as the smallest groups of code that can be maintained
+and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and
+constructors in Java). One of the guidelines with which the project does not
+comply is the one presented in the report (cf. Figure~\ref{fig:bchrep}):
+\emph{Write Simple Units of Code}. BCH analyzes this guideline based on the
+McCabe Complexity~\cite{1702388} to calculate the number of branch points of a
+method. The bar at the bottom of the figure represents the top $30$ units that
+violate the guideline, sorted by severity. The violation severities are
+indicated using colors, and there is a legend to help to interpret them. The
+green bar represents the number of compliant branch points per unit (\emph{at
+most $5$}), i.e., the number of units are compliant with ISO
+25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do not
+comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
+(\emph{above $25$}) severity levels. In the bar, there are marks that pinpoint
+the compliance thresholds for each severity level. If the green mark is
+somewhere in the green bar it means it is compliant with a low level of
+severity.
 
@@
 
-We compute the mean average of the maintainability score $M(v)$ as follows:
+Given the violations for the BCH guidelines, we compute the maintainability score
+$M(v)$ as follows:
 
@@
 (Table~\ref{tab:guidelines}) and $v$ is the version of the software under
-evaluation. The maintenance $M$ based on the guideline $g$ for a given version
+evaluation. $M(v) < 0$ indicates that version $v$ violates the guidelines,
+while $M(v) > 0$ indicates that version $v$ follows the BCH guideliens.
+
+The maintenance for the guideline $g$, $M_g$ for a given version
 of a project is computed with the following equation:

### 2019-02-05 paper/ICPC19/icpc19.tex
fix
fixing word in rw

@@
 (Table~\ref{tab:guidelines}) and $v$ is the version of the software under
-evaluation. $M(v) < 0$ indicates that version $v$ violates the guidelines,
-while $M(v) > 0$ indicates that version $v$ follows the BCH guideliens.
+evaluation. $M(v) < 0$ indicates that version $v$ is violating (some of) the
+guidelines, while $M(v) > 0$ indicates that version $v$ is following
+the BCH guideliens.
 
@@
 Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the previously demonstrated
-correlation between programming language and software defects. Our works studies how
+correlation between programming language and software defects. Our work studies how
 security refactorings affect software quality based on the code maintainability

### 2019-02-05 paper/ICPC19/icpc19.tex
php listing

@@
     language=PHP,
+    alsolanguage=HTML,
     backgroundcolor=\color{white},   % choose the background color
@@
    <p class='hint'>
--   <?php if(isset($_['file'])) echo $_['file']?>
+    <?php
+-   if(isset($_['file'])) echo $_['file']
+    ?>
    </p>
    <p class='hint'>
-+   <?php if(isset($_['file'])) echo htmlentities($_['file'])?>
+    <?php
++   if(isset($_['file'])) echo htmlentities($_['file'])
+    ?>
    </p>

### 2019-02-05 paper/ICPC19/icpc19.tex
icp19

@@
 We would expect that the negative impact of programming languages on
-maintainability would be more severe given the ideas that poor design of programming languages and lack of using
-the best practices lead to more buggy and vulnerable
+maintainability would be more severe given the ideas that poor design of programming 
+languages and lack of using the best practices lead to more buggy and vulnerable
 code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
@@
 
+We measured the effect sizes between both distribution - before and after patches - per 
+language and security flaw using the Common Language Effect Size~\cite{cliff:1993} method
+to understand the magnitude of the differences between distributions. Overall, the effect 
+sizes were of a low-medium level. Thus, considering the median equal 0 and the low size effects
+it is reasonable to affirm that there is a significant real effect but it might only be 
+recognized after a thoughtful data analysis.
+
 \section{Threats to Validity}\label{sec:threats}

### 2019-02-05 paper/ICPC19/icpc19.tex
back to the page limits

--- 2019-02-05 paper/ICPC19/icpc19.tex
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/methodology.pdf}
  	\caption{Study Methodology}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/language_dist.pdf}
  	\caption{Distribution of Security Refactorings per Language}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/type_dist.pdf}
  	\caption{Distribution of the Domain of Projects}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/language.pdf}
+  \includegraphics[width=0.45\textwidth]{figures/language.pdf}
   \caption{Maintainability difference by language after security refactorings.}
@@
 flaws. These type of refactorings harm maintainability in $43.1\%$ ($84$) of the
-cases. In $31.8\%$ ($62$), there is no effect in maintainability and for the
+cases. In $31.8\%$ ($62$), there is no impact and for the
 other $25.1\%$ ($49$) maintainability increases.
@@
  	\centering
- 	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
+ 	\includegraphics[width=0.50\textwidth]{figures/category.pdf}
  	\caption{Maintainability difference by pattern after security refactorings.
@@
 We would expect that the negative impact of programming languages on
-maintainability would be more severe given the ideas that poor design of programming languages and lack of using
-the best practices lead to more buggy and vulnerable
+maintainability to be more severe, as arguably poor design of programming
+languages and lack of using the best practices lead to more buggy and vulnerable
 code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
@@
 
+We measured the effect sizes between both distribution -- before and after
+patches --- per language and security patterns using the Common Language Effect
+Size~\cite{cliff:1993} method. Effect sizes are of a low-medium level.
+Thus, considering the $0$ median and the low size effects, it is reasonable
+to affirm that there is a significant real effect but it might only be
+recognized after a thoughtful data analysis.
+
 \section{Threats to Validity}\label{sec:threats}
@@
 The maintainability formula was inferred based on the BCH's reports. The high
-amount of different projects and backgrounds may require different
-maintainability standards. However, BCH does have a representative benchmark of
-closed and open-source software projects that are used to compute the
+amount of different projects and backgrounds may require other
+maintainability standards. However, BCH does use a representative benchmark of
+closed and open-source software projects to compute the
 thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.

### 2019-02-06 paper/ICPC19/icpc19.tex
small fixes

@@
     frame=single,
+	numbers=left,
     stepnumber=1,
+	xleftmargin=2em,
     escapeinside={/*!}{!*/},

### 2019-02-06 paper/ICPC19/icpc19.tex
review sec II

@@
 separately. In the example (Listing~\ref{lst:vuln}), all functions receive as
-input the SSL session, \texttt{s}. In the \texttt{dtls1\_reassemble\_fragment}
-function the size of the fragment \texttt{msg\_hdr-$>$frag\_off+frag\_len} is
-checked against the handshake message size \texttt{msg\_hdr->msg\_len} (line
-$13$) and then against the maximum size allowed in a DTLS handshake message for
-\texttt{s} (line $16$-$21$). However, it is not verified if \texttt{msg\_hdr->msg\_len}
-is lower than the size allowed by DTLS. Thus, in line
-$37$, when \texttt{msg\_len} is allocated to the fragment buffer, more memory
-than the allowed might be allocated. According to the commit message, $10$
-handshake messages are allowed per DTLS connection. Hence, a remote attacker
-could leverage this issue to trigger memory allocations corresponding to large
-length values and lead the system to memory exhaustion that could end in a DoS
-attack. The code changes performed to fix the vulnerability were the following:
+input the SSL session, \texttt{s}.\Luis{is the previous sentence relevant?.} In
+the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment
+\texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake
+message size \texttt{msg\_hdr->msg\_len} (line $13$) and then against the
+maximum size allowed in a DTLS handshake message for \texttt{s} (line
+$16$-$21$).\Luis{I'm not sure about my comment: I think it could be easier if
+we explain the previous sentence in pure natural language. These kind of
+statements \texttt{msg\_hdr-$>$frag\_off+frag\_len} are not easy to process in
+this context.} However, it is not verified if \texttt{msg\_hdr->msg\_len} is
+lower than the size allowed by DTLS\Luis{did not understand the problem of not
+checking the minimum.}. Thus, in line $37$, when \texttt{msg\_len} is allocated
+to the fragment buffer, more memory than the allowed might be allocated.
+According to the commit message, $10$ handshake messages are allowed per DTLS
+connection. Hence, a remote attacker could leverage this issue to trigger
+memory allocations corresponding to large length values and lead the system to
+memory exhaustion that could end in a DoS attack.\Luis{I'm not happy with this
+paragraph. It's difficult to process and I dont think a reviewer will actually
+spend a lot of energy to understand this. (Ignore this if you find it clear.)}
+The code changes performed to fix the vulnerability were the following:
 

### 2019-02-07 scripts/histogram.py
update

@@ -4,6 +4,9 @@ import csv
 import collections
 import pandas as pd
 import numpy as np
+import matplotlib
+matplotlib.rcParams['font.family'] = 'serif'
+matplotlib.rcParams['mathtext.fontset'] = 'cm'
 import matplotlib.pyplot as plt
 
 
@@ -12,7 +15,7 @@ def language_dist(commits):
     df = pd.read_csv(commits)
     
     lang_trans = {'objc':'Objective-C', 'php':'PHP', 'ruby':'Ruby', 'c':'C', 'c++':'C++', 'groovy':'Groovy', 'javascript':'JavaScript',
-                    'python':'Python', 'java':'Java', 'objc++':'Objective-C++', 'scala':'scala', 'swift':'Swift', 'smarty':'Smarty'}
+                    'python':'Python', 'java':'Java', 'objc++':'Objective-C++', 'scala':'Scala', 'swift':'Swift', 'smarty':'Smarty'}
     
     dic = {}
     for i in df['language']:
@@ -27,21 +30,17 @@ def language_dist(commits):
     dic_lang['no'].append(dic['others']/len(df['language']))
     
     for i in dic:
-        if i != 'others':
-        #     no = dic['others']
-        # else:
+        if i != 'others': 
             dic_lang['lang'].append(lang_trans[i])
             dic_lang['no'].append(dic[i]/len(df['language']))
     
-    
     lang = pd.DataFrame(dic_lang)
-    print(lang)
     
     ax = lang['no'].plot(kind='barh', figsize=(10, 11), fontsize=12, color="#3F86DB")
     vals = ax.get_xticks()
     ax.set_xticklabels(['{:,.0%}'.format(x) for x in vals])
     ax.set_yticklabels(dic_lang['lang'], minor=False)
-    plt.subplots_adjust(left=0.2, right=0.95, top=0.9, bottom=0.05)
+    plt.subplots_adjust(left=0.15, right=0.95, top=0.9, bottom=0.05)
     plt.gca().xaxis.grid(True, linestyle='--')
     
     plt.savefig('../paper/ICPC19/figures/language_dist.pdf')
@@ -64,14 +63,12 @@ def type_dist(projects):
         dic_type['no'].append(dic[i]) 
                 
     app_type = pd.DataFrame(dic_type)
-    print(app_type)
     
     ax = app_type['no'].plot(kind='barh', figsize=(10, 7), fontsize=12, color="#3F86DB")
-    #ax.set_xlabel("Count", fontsize=11)
[…]

### 2019-02-07 scripts/histogram.py
tight layout

@@ -43,7 +43,7 @@ def language_dist(commits):
     ax.set_yticklabels(dic_lang['lang'], minor=False)
     plt.subplots_adjust(left=0.2, right=0.95, top=0.9, bottom=0.05)
     plt.gca().xaxis.grid(True, linestyle='--')
-    
+    plt.tight_layout()
     plt.savefig('../paper/ICPC19/figures/language_dist.pdf')
 
 def type_dist(projects):
@@ -73,7 +73,7 @@ def type_dist(projects):
     ax.set_yticklabels(app_type['type'], minor=False)
     plt.subplots_adjust(left=0.3, right=0.9, top=0.9, bottom=0.05)
     plt.gca().xaxis.grid(True, linestyle='--')
-    
+    plt.tight_layout()
     plt.savefig('../paper/ICPC19/figures/type_dist.pdf')
 
 def main(projects, commits):    

### 2019-02-07 scripts/get_maintainability.py
maintainability script update

@@ -4,10 +4,14 @@ import csv
 import collections
 import pandas as pd
 import numpy as np
+import matplotlib
+matplotlib.rcParams['font.family'] = 'serif'
+matplotlib.rcParams['mathtext.fontset'] = 'cm'
 import matplotlib.pyplot as plt
 from scipy.stats import wilcoxon, shapiro
 from statistics import mean, stdev
 from math import sqrt
+from cles import cles_paired
 
 def _print_hypothesis_test(differences):
     """Paired Wilcoxon signed-rank test (N should be > 20)"""
@@ -83,7 +87,7 @@ def total_main_barchart(total_sec, total_reg, graphics, result):
             p = '<0.001'
         else:
             p = '={:.{}f}'.format(l[1]['pvalue'], 3)
-        box_text = '$\overline{x}$='+ '{:.{}f}'.format(l[1]['mean'], 2) + '\n$\widetilde{x}$=' + '{:.{}f}'.format(l[1]['med'], 2) + '\np' + p
+        box_text = '$\overline{x}$='+ '{:.{}f}'.format(l[1]['mean'], 2) + '\nM=' + '{:.{}f}'.format(l[1]['med'], 2) + '\np' + p
         ax.text(0.46, y, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':4}, fontsize=9)
         y -= 1.1
     
@@ -208,14 +212,10 @@ def sec_vs_rg_commits(CACHE, graphics, dataset):
         
     fil = df[(df['main(diff)_reg'] != 'error') & (df['main(diff)_sec'] != 'error')]
     
-    cohens_d = (fil['main(diff)_reg'].astype('float64').mean() - fil['main(diff)_sec'].astype('float64').mean()) / (sqrt((stdev(fil['main(diff)_reg'].astype('float64')) ** 2 + stdev(fil['main(diff)_sec'].astype('float64')) ** 2) / 2))
-    print('coehns', cohens_d)
-    
     stats1 = _print_hypothesis_test(fil['main(diff)_sec'].astype('float64'))
     stats = _print_hypothesis_test(fil['main(diff)_reg'].astype('float64'))
 
     result = stats.append([stats1])
-    print(result)
     result.to_csv('../results/statistical_test.csv')
     
     total_sec, total_reg = {'neg': 0, 'pos': 0, 'nul': 0}, {'neg': 0, 'pos': 0, 'nul': 0}
[…]

### 2019-02-07 paper/ICPC19/icpc19.tex

@@
 implications; section~\ref{sec:threats}, threats to the validity of the study
-are enumerated; section~\ref{sec:rw} describres the different work and existing
+are enumerated; section~\ref{sec:rw} describes the different work and existing
 literature in the field of study; and, finally section~\ref{sec:conclusions}
@@
 $16$-$21$).\Luis{I'm not sure about my comment: I think it could be easier if
-we explain the previous sentence in pure natural language. These kind of
+we explain the previous sentence in pure natural language. This kind of
 statements \texttt{msg\_hdr-$>$frag\_off+frag\_len} are not easy to process in
@@
  	\centering 	\includegraphics[width=0.4\textwidth]{figures/language_dist.pdf}
- 	\caption{Distribution of Security Refactorings per Language}
+ 	\caption{Distribution of Security Refactorings per Language\Luis{esta figure pode ser maior nao pode?}}
 	\label{fig:lang}
@@
     \item \textbf{Cross-Site Scripting (114 cases).} Lack of proper validation or escaping
-    allow attackers to submit untrusted data to web browsers through malicious
+    allows attackers to submit untrusted data to web browsers through malicious
     scripts that can hijack the user sessions or redirect the user to malicious
@@
 percentage of code allowed to be in each of the $4$ risk severity levels
-(\emph{low risk}, \emph{medium risk}, \emph{high risk} and \emph{very high
+(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
 risk}). If the project does not violate the thresholds, then it is compliant
@@
 
-Aiming to analyze the impact of security refactorings, we use BCH to compute
+Aiming to analyze the impact of security refactorings, we use BCH to compute the
 maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
@@
 	aim at evaluating security patterns occurring in different projects similarly.
-	Thus, the divised metric uses the \textit{raw} number of lines of code rather
+	Thus, the derived metric uses the \textit{raw} number of lines of code rather
   percentages for normalization purposes.
-	\item \textbf{Distance to lower severity levels are less penalized than to the
+	\item \textbf{Distance to lower severity levels is less penalized than to the
 	thresholds in high severity levels.} Severity level weights based on the
@@
 guidelines, while $M(v) > 0$ indicates that version $v$ is following
-the BCH guideliens.
+the BCH guidelines.
 
@@
 the dataset contains security refactorings for more than $13$ languages, we only
-present the results for \emph{Java}, \emph{C}, \emph{Ruby} and \emph{PHP}
+present the results for \emph{Java}, \emph{C}, \emph{Ruby}, and \emph{PHP}
 because one of the requirements of the Paired Wilcoxon signed-rank is that the
@@
 	may be designed to help developers be highly productive when writing and
-	maintaing  secure applications. One example is the need for designing new
+	maintaining  secure applications. One example is the need for designing new
 	authorization mechanisms since these are one of the most complex security
@@
 \textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service
-attacks and Memory Leaks and Miscellaneous patterns are more likely to affect
+attacks, and Memory Leaks, and Miscellaneous patterns are more likely to affect
 open-source software maintainability.}} Although there are a few refactorings
@@
 Size~\cite{cliff:1993} method. Effect sizes are of a low-medium level.
-Thus, considering the $0$ median and the low size effects, it is reasonable
-to affirm that there is a significant real effect but it might only be
+Thus, considering the $0$ median and the low effect sizes, it is reasonable
+to affirm that there is a significant effect but it might only be
 recognized after a thoughtful data analysis.
@@
 security refactorings affect software quality based on the code maintainability
-analysis and provides evidence that programming language have impact on
+analysis and provides evidence that programming languages have an impact on
 maintainability.
@@
 academic curricula in the fields of security and software quality; and, tools
-such as BCH to help developers predit the effect of their patches.
+such as BCH to help developers predict the effect of their patches.
 
@@
 guidelines are more likely to be affected by security-related refactorings;
-investigate what are the most frequent vulnerabilities in each programming language;
-study if there is any correlation between software
-maintainability and popularity of the project; expand our methodology with other
+investigate what are the most frequent vulnerabilities in each programming
+language; study if there is any correlation between code maintainability and
+the popularity of a software project; expand our methodology with other
 software quality properties; validate these findings with closed or private

### 2019-02-07 scripts/get_maintainability.py
histogram size adjustment

@@ -89,7 +89,7 @@ def total_main_barchart(total_sec, total_reg, graphics, result):
     
     plt.gca().xaxis.grid(True, linestyle='--')
     
-
+    plt.tight_layout()
     plt.savefig('../paper/ICPC19/figures/maintainability.pdf')
 
 def dic_barchart(dic, graphics, title, file_name):
@@ -108,6 +108,7 @@ def dic_barchart(dic, graphics, title, file_name):
     plt.xticks(index + bar_width, df['type'], fontsize=8, rotation='vertical')
     plt.legend()
     plt.subplots_adjust(bottom=0.2)
+    plt.tight_layout()
     plt.savefig(graphics+file_name)
 
 def check_if_in(key, dic):
@@ -297,7 +298,7 @@ def main_by_type_barchart(patterns, diff):
         y -= 1.02
         
     plt.gca().xaxis.grid(True, linestyle='--')
-
+    plt.tight_layout()
     plt.savefig('../paper/ICPC19/figures/category.pdf')
 
 def sec_by_type(CACHE, fil):

### 2019-02-07 scripts/histogram.py

@@ -35,15 +35,15 @@ def language_dist(commits):
             dic_lang['no'].append(dic[i]/len(df['language']))
     
     lang = pd.DataFrame(dic_lang)
-    
-    ax = lang['no'].plot(kind='barh', figsize=(10, 11), fontsize=12, color="#3F86DB")
+
+    fig, ax = plt.subplots(figsize=(10, 8))
+    ax = lang['no'].plot(kind='barh', fontsize=16, color="#3F86DB", ax=ax)
     vals = ax.get_xticks()
     ax.set_xticklabels(['{:,.0%}'.format(x) for x in vals])
     ax.set_yticklabels(dic_lang['lang'], minor=False)
-    plt.subplots_adjust(left=0.15, right=0.95, top=0.9, bottom=0.05)
-    plt.gca().xaxis.grid(True, linestyle='--')
-    plt.tight_layout()
-    plt.savefig('../paper/ICPC19/figures/language_dist.pdf')
+    ax.xaxis.grid(True, linestyle='--')
+    fig.tight_layout()
+    fig.savefig('../paper/ICPC19/figures/language_dist.pdf')
 
 def type_dist(projects):
     
@@ -64,14 +64,14 @@ def type_dist(projects):
                 
     app_type = pd.DataFrame(dic_type)
     
-    ax = app_type['no'].plot(kind='barh', figsize=(10, 7), fontsize=12, color="#3F86DB")
+    fig, ax = plt.subplots()
+    app_type['no'].plot(kind='barh', figsize=(10, 4), fontsize=16, color="#3F86DB", ax=ax)
     vals = ax.get_xticks()
     ax.set_xticklabels(['{}'.format(int(x)) for x in vals])
     ax.set_yticklabels(app_type['type'], minor=False)
-    plt.subplots_adjust(left=0.31, right=0.9, top=0.9, bottom=0.05)
-    plt.gca().xaxis.grid(True, linestyle='--')
-    plt.tight_layout()
-    plt.savefig('../paper/ICPC19/figures/type_dist.pdf')
+    ax.xaxis.grid(True, linestyle='--')
+    fig.tight_layout()
+    fig.savefig('../paper/ICPC19/figures/type_dist.pdf')
 
 def main(projects, commits):    
     

### 2019-02-07 paper/ICPC19/icpc19.tex
histogram size adjustment
table descriptive stats projects with zero decimals

@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/language_dist.pdf}
+ 	\centering 	\includegraphics[width=\linewidth]{figures/language_dist.pdf}
  	\caption{Distribution of Security Refactorings per Language\Luis{esta figure pode ser maior nao pode?}}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/type_dist.pdf}
+ 	\centering 	\includegraphics[width=\linewidth]{figures/type_dist.pdf}
  	\caption{Distribution of the Domain of Projects}
@@
 	\centering
-	\caption{Descriptive statistics of the dataset projects}
+	\caption{Descriptive statistics of the projects in the dataset}
 \begin{tabular}{@{}rrrrrrrrrrr@{}}
@@
       & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
-mean  & 1763.52 & 5448.74 & 401.37   & 153.33       & 14834.17 & 45.17    & 129.45   & 122973.24 & 3768.97   & 1941.61 \\
-std   & 2434.03 & 6215.09 & 486.60   & 123.68       & 22234.46 & 150.15   & 189.93   & 209732.51 & 5933.16   & 3603.31 \\
-min   & 1.00       & 3.00       & 1.00        & 0.00            & 103.00      & 1.00        & 0.00        & 108.00       & 0.00         & 0.00       \\
-25\%  & 391.50  & 1581.00    & 117.25   & 49.00           & 1440.50  & 4.00        & 19.00       & 8466.75   & 313.75    & 143.25  \\
-median  & 838.50  & 2836.50 & 248.00      & 99.00           & 5504.50  & 9.00        & 59.00       & 37372.50  & 1792.50   & 650.00     \\
-75\%  & 2155.00    & 6828.50 & 459.50   & 261.00          & 18579.25 & 20.00       & 142.75   & 117699.50 & 4087.75   & 1907.25 \\
-max   & 16366.00   & 31841.00   & 3446.00     & 413.00          & 114378.00   & 1227.00     & 1114.00     & 995790.00    & 33970.00     & 19329.00   \\
-Total & 165771.00  & 512182.00  & 37729.00    & 14413.00        & 1394412.00  & 4246.00     & 12168.00    & 11559485.00  & 354283.00    & 182511.00  \\ \bottomrule
+mean  & 1763 & 5448 & 401   & 153       & 14834 & 45    & 129   & 122973 & 3768   & 1941 \\
+std   & 2434 & 6215 & 486   & 123       & 22234 & 150   & 189   & 209732 & 5933   & 3603 \\
+min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
+25\%  & 391  & 1581    & 117   & 49           & 1440  & 4        & 19       & 8466   & 313    & 143  \\
+median  & 838  & 2836 & 248      & 99           & 5504  & 9        & 59       & 37372  & 1792   & 650     \\
+75\%  & 2155    & 6828 & 459   & 261          & 18579 & 20       & 142   & 117699 & 4087   & 1907 \\
+max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
+Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511.00  \\ \bottomrule
 \end{tabular}

### 2019-02-07 paper/ICPC19/icpc19.tex
luis fixes solved + typos + small problems raised by Pedro Adão

@@
 presents the results and section~\ref{sec:discussion} discusses their
-implications; section~\ref{sec:threats}, threats to the validity of the study
-are enumerated; section~\ref{sec:rw} describes the different work and existing
+implications; section~\ref{sec:threats}, enumerates the threats to the validity of 
+this study; section~\ref{sec:rw} describes the different work and existing
 literature in the field of study; and, finally section~\ref{sec:conclusions}
@@
 
+Every SSL/TLS connection, \texttt{s}, begins with a handshake which is responsible by the
+negotiation between two parties. DTLS provides a mechanism for fragmenting a
+handshake message over a number of packets, each of which can be transmitted
+separately over the Internet from one party to another. The problem in CVE-$2014$-$3506$ 
+is that the size of the handshake message collected by the receiver is not verified against 
+the maximum size allowed by DTLS and latter in the code it is allocated to a new fragment (line $37$). 
+Thus, if an attacker has access to one of these packages it can change the handshake messsage size and
+trigger unnecessary memory allocations of large length which can lead to memory exhaustion that could
+could end in a DoS attack.
+
 \medskip
@@
 
-Every SSL/TLS connection begins with a handshake which is responsible by the
-negotiation between two parties. DTLS provides a mechanism for fragmenting a
-handshake message over a number of packets, each of which can be transmitted
-separately. In the example (Listing~\ref{lst:vuln}), all functions receive as
-input the SSL session, \texttt{s}.\Luis{is the previous sentence relevant?.} In
-the \texttt{dtls1\_reassemble\_fragment} function the size of the fragment
-\texttt{msg\_hdr-$>$frag\_off+frag\_len} is checked against the handshake
-message size \texttt{msg\_hdr->msg\_len} (line $13$) and then against the
-maximum size allowed in a DTLS handshake message for \texttt{s} (line
-$16$-$21$).\Luis{I'm not sure about my comment: I think it could be easier if
-we explain the previous sentence in pure natural language. This kind of
-statements \texttt{msg\_hdr-$>$frag\_off+frag\_len} are not easy to process in
-this context.} However, it is not verified if \texttt{msg\_hdr->msg\_len} is
-lower than the size allowed by DTLS\Luis{did not understand the problem of not
-checking the minimum.}. Thus, in line $37$, when \texttt{msg\_len} is allocated
-to the fragment buffer, more memory than the allowed might be allocated.
-According to the commit message, $10$ handshake messages are allowed per DTLS
-connection. Hence, a remote attacker could leverage this issue to trigger
-memory allocations corresponding to large length values and lead the system to
-memory exhaustion that could end in a DoS attack.\Luis{I'm not happy with this
-paragraph. It's difficult to process and I dont think a reviewer will actually
-spend a lot of energy to understand this. (Ignore this if you find it clear.)}
 The code changes performed to fix the vulnerability were the following:
@@
 handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if
-the maximum certificate list size requires it. More
-specifically, it checks if the maximum certificate list size
-(\texttt{s-$>$max\_cert\_list}) requires more length than the maximum length
+the maximum certificate list size (\texttt{s-$>$max\_cert\_list}) requires it. More
+specifically, it checks if the maximum certificate list size requires more length than the maximum length
 (\texttt{max\_len}). If that condition holds, the function returns the size of
@@
 ascertain that the message and the fragment length are never higher than the
-maximum allowed by DTSL. In particular, \ref{lst:func5} is responsible for patching
+maximum allowed by DTLS. In particular, \ref{lst:func5} is responsible for patching
 the vulnerability.
@@
 calculate their maintainability reports. However, there are refactorings
-discarded from tour study due to limitations of BCH: in particular, lack of
+discarded from our study due to limitations of BCH: in particular, lack of
 language support and project size. The final dataset used in this paper contains
@@
 other words, we test the null hypothesis that the maintainability difference
-between pairs of versions $v_{v-1}$, $v_v$ (i.e., before and after a
+between pairs of versions $v_{s-1}$, $v_v$ (i.e., before and after a
 security-commit) come from the same distribution. Nevertheless, this test has a
@@
 https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and
-conclude that the flaw was due to the introduction of high cyclomatic complexity
+concluded that the flaw was due to the introduction of high cyclomatic complexity
 in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and
@@
 of maintainability between groups of refactored elements and non-refactored were
-analyzed. In addition, the authors conclude that the source code elements
+analyzed. In addition, the authors concluded that the source code elements
 subjected to refactorings had significantly lower maintainability than the
@@
 Better Code Hub reports to calculate maintainability
-based on a model proposed in previous work. Security refactorings significantly
+based on a model proposed in previous work~\cite{Olivari:2018}. Security refactorings significantly
 affect the software maintainability of open-source software, i.e., developers

### 2019-02-07 paper/ICPC19/icpc19.tex
fixes

@@
 Every SSL/TLS connection, \texttt{s}, begins with a handshake which is responsible by the
-negotiation between two parties. DTLS provides a mechanism for fragmenting a
+negotiation between the two parties. DTLS provides a mechanism for fragmenting a
 handshake message over a number of packets, each of which can be transmitted
-separately over the Internet from one party to another. The problem in CVE-$2014$-$3506$ 
+separately over the internet from one party to another. The problem in CVE-$2014$-$3506$ 
 is that the size of the handshake message collected by the receiver is not verified against 
-the maximum size allowed by DTLS and latter in the code it is allocated to a new fragment (line $37$). 
-Thus, if an attacker has access to one of these packages it can change the handshake messsage size and
+the maximum size allowed by DTLS. It is then allocated to a new fragment (line $37$). 
+Thus, if an attacker has access to one of these packages, the attacker can change the handshake messsage size and
 trigger unnecessary memory allocations of large length which can lead to memory exhaustion that could

### 2019-02-08 paper/ICPC19/icpc19.tex
update

@@
 (BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
-project. There are other well-known standards and models that have been proposed
+project. BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
+base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
+based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
+based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}. 
+There are other well-known standards and models that have been proposed
 to increase software security: Common Criteria~\cite{common:2009} which received
@@
 first steps taken by SIG to introduce security on their maintainability model).
-BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
-base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
-based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
-based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
 
@@
 previous work focused on comparing the impact of security refactorings with
-regular refactorings on maintainability. We analyze the maintainability of
-regular commits (i.e., commits not related to security fixes) and use them as a
-baseline.
-
-The baseline dataset uses the security commits dataset as input. For each
+regular refactorings on maintainability.
+We analyze the maintainability of regular commits (i.e., commits not related 
+to security fixes) and use them as a baseline.
+
+The baseline dataset uses the security commits dataset as input. 
+Both datasets have the same size. For each
 security commit, one random commit is selected from the list of all commits of
@@
  	\includegraphics[width=0.50\textwidth]{figures/category.pdf}
- 	\caption{Maintainability difference by pattern after security refactorings.
-  \Luis{a median e' zero em muitos casos -- como interpretam isso? se nao houver
-boa/simples explicacao talvez seja melhor usar o common language.}\Sof{Preciso de pensar melhor nisto}}
+ 	\caption{Maintainability difference by pattern after security refactorings.}
 	\label{fig:pat}
@@
 more than $50\%$ on software maintainability. We suspect that these values are
-the result of projects contributions policies. Nine out of ten projects --- the
-equivalent to $31.8\%$ of the commits --- of the top-$10$ of dataset projects
+the result of projects contributions policies. Nine of the top-10 of dataset projects  --- the
+equivalent to $31.8\%$ of the commits --- 
 with more contributors have restrict contribution policies regarding code
@@
 instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-Standard}\footnote{PSR-2 Standard available at
+Standard} that is a coding style guide for \texttt{PHP}\footnote{PSR-2 Standard available at
 https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.
@@
 investigate what are the most frequent vulnerabilities in each programming
-language; study if there is any correlation between code maintainability and
+language; investigate how refactorings impact maintainability for each
+vulnerability and programming language; 
+study if there is any correlation between code maintainability and
 the popularity of a software project; expand our methodology with other

### 2019-02-09 paper/ICPC19/icpc19.tex
icpc19
last revision

@@
 security on the maintainability of open-source software. Maintainability is
-measured based on the \emph{Better Code Hub}'s model of $10$ guidelines using a
-dataset including $607$ security commits. Results show that fixing software
+measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
+dataset including 607 security commits. Results show that fixing software
 vulnerabilities hinders the maintainability of open-source software. This is
@@
 hackers exploit them. It turns out that fixing these issues is as simple as
-modifying the software codebase. However, these refactorings may have a negative
-impact on software maintenance, mostly because developers follow the quickest
-solution to fix the problem and not necessarily the most elegant and performant
-one. Thus, we want to understand how security refactorings impact the
-maintainability of software.
+modifying the software codebase. However, these refactorings may have a
+negative impact on software maintenance --- e.g., developers might follow the
+quickest solution to fix the problem and not necessarily the most elegant and
+performant one. Thus, we want to understand how security refactorings impact
+the maintainability of software.
 
@@
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-evaluation. The following snippet presents the changes performed on the
+evaluation. The snippet in Listing~\ref{lst:vuln} presents the changes performed on the
 \emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
@@
 
-Every SSL/TLS connection, \texttt{s}, begins with a handshake which is responsible by the
-negotiation between the two parties. DTLS provides a mechanism for fragmenting a
-handshake message over a number of packets, each of which can be transmitted
-separately over the internet from one party to another. The problem in CVE-$2014$-$3506$ 
-is that the size of the handshake message collected by the receiver is not verified against 
-the maximum size allowed by DTLS. It is then allocated to a new fragment (line $37$). 
-Thus, if an attacker has access to one of these packages, the attacker can change the handshake messsage size and
-trigger unnecessary memory allocations of large length which can lead to memory exhaustion that could
-could end in a DoS attack.
+Every SSL/TLS connection begins with a handshake which is responsible by the
+negotiation between the two parties. DTLS provides a mechanism for fragmenting
+a handshake message over a number of packets, each of which can be transmitted
+separately over the internet from one party to another. The problem in
+CVE-$2014$-$3506$ is that the size of the handshake message collected by the
+receiver is not verified against the maximum size allowed by DTLS. It is then
+allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~37). Thus, if
+an attacker has access to one of these packages, the attacker can change the
+handshake messsage size and trigger unnecessary memory allocations of large
+length which can lead to memory exhaustion that could could end in a DoS attack.
 
@@
 the certificate list, otherwise, it returns the maximum length. This function
-replaces lines $18$-$21$.
+replaces lines $18$--$21$.
 
@@
 \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
-disrupts $2$ of the guidelines proposed by the Software Improvement Group (SIG) for
+disrupts two of the guidelines proposed by the Software Improvement Group (SIG) for
 building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
@@
 instance, implementing secure authentication is not as easy as fixing a
-cross-site scripting vulnerability since usually the last one is can be fixed
-without adding new lines of code, as seen in Listing~\ref{lst:fix}. The developer added the function \emph{htmlentities} to
-\textit{escape} the data given by the variable \texttt{\$\_['file']}. Knowing which
-patterns are more likely to increase maintainability issues is one step forward
-to bring awareness to security engineers of what patterns need more attention.
-The taxonomy of security patterns used to answer this question is the same as
-the one provided by the authors of the dataset~\cite{Reis:2017:IJSSE}.
-Maintainability is measured separately for each pattern.
+cross-site scripting vulnerability since usually the last one can be fixed
+without adding new lines of code. A typical fix for the cross-site scripting
+vulnerability is presented in Listing~\ref{lst:fix}. The developer added the
+function \texttt{htmlentities} to \textit{escape} the data given by the variable
+\texttt{\$\_['file']}. Knowing which patterns are more likely to increase
+maintainability issues is one step forward to bring awareness to security
+engineers of what patterns need more attention. The taxonomy of security
+patterns used to answer this question is the same as the one provided by the
+authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
+separately for each pattern.
 
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=\linewidth]{figures/language_dist.pdf}
- 	\caption{Distribution of Security Refactorings per Language\Luis{esta figure pode ser maior nao pode?}}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
+ 	\caption{Distribution of Security Refactorings per Language}
 	\label{fig:lang}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=\linewidth]{figures/type_dist.pdf}
+ 	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
  	\caption{Distribution of the Domain of Projects}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/bch_report.png}
+ 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.49\textwidth]{figures/commit.pdf}
+ 	\centering 	\includegraphics[width=0.8\linewidth]{figures/commit.pdf}
  	\caption{Maintainability difference for security commits}
@@
 other words, we test the null hypothesis that the maintainability difference
-between pairs of versions $v_{s-1}$, $v_v$ (i.e., before and after a
+between pairs of versions $v_{s-1}$, $v_s$ (i.e., before and after a
 security-commit) come from the same distribution. Nevertheless, this test has a
@@
   \centering
-  \includegraphics[width=0.45\textwidth]{figures/language.pdf}
+  \includegraphics[width=0.5\textwidth]{figures/language.pdf}
   \caption{Maintainability difference by language after security refactorings.}
@@
  	\centering
- 	\includegraphics[width=0.50\textwidth]{figures/category.pdf}
+ 	\includegraphics[width=0.56\textwidth]{figures/category.pdf}
  	\caption{Maintainability difference by pattern after security refactorings.}
@@
 address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be properly representative of the population regarding
+the dataset may not be representative of the population regarding
 programming languages.

### 2019-02-09 paper/ICPC19/icpc19.tex
fixes

@@
 website\footnote{Zero Day Initiative website available at
-https://www.zerodayinitiative.com/advisories/published/ (Accessed on \today{})}.
+\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
 These vulnerabilities are gateways for attackers to penetrate
@@
 As an example, in the last week of January $2019$, a severe vulnerability in Apple
-FaceTime was revealed\footnote{https://thehackernews.com/2019/01/apple-facetime-
-privacy-hack.html (Accessed on \today{})} --- the outcome of a design/logical
+FaceTime was revealed\footnote{FaceTime vulnerability details available at \url{https://thehackernews.com/2019/01/apple-facetime-
+privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
 flaw that allowed callers to see and hear others without them picking up the call.
@@
 example of a security refactoring of a known vulnerability found in
-OpenSSL\footnote{Repository available at https://github.com/openssl/openssl
+OpenSSL\footnote{Repository available at \url{https://github.com/openssl/openssl}
 (Accessed on \today{})}; section~\ref{sec:methodology} describes the
@@
 this study; section~\ref{sec:rw} describes the different work and existing
-literature in the field of study; and, finally section~\ref{sec:conclusions}
+literature in the field of study; and, finally, section~\ref{sec:conclusions}
 concludes the main findings and elaborates on future work.
@@
 Vulnerabilities and Exposures dictionary as CVE-$2014$-$3506$\footnote{CVE-$2014$-$3506$
-details available at http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506
+details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506}
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
@@
 \emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
-https://github.com/openssl/openssl/commit/\\1250f12613b61758675848f6600ebd914ccd7636
+\url{https://github.com/openssl/openssl/commit/1250f12613b61758675848f6600ebd914ccd7636}
 (Accessed on \today{})} by the OpenSSL developers to patch the vulnerability.
@@
 receiver is not verified against the maximum size allowed by DTLS. It is then
-allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~37). Thus, if
-an attacker has access to one of these packages, the attacker can change the
-handshake messsage size and trigger unnecessary memory allocations of large
-length which can lead to memory exhaustion that could could end in a DoS attack.
+allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~37). Thus, if an 
+attacker has access to one of these packages, the attacker can change the 
+handshake message size and trigger unnecessary memory allocations of large 
+length which can lead to memory exhaustion that could end in a DoS attack.
 
@@
   \item Use the Software Improvement Group's\footnote{SIG's website is available
-  at https://www.sig.eu/ (Accessed on \today{})} web-based source code analysis
+  at \url{https://www.sig.eu/} (Accessed on \today{})} web-based source code analysis
   service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
-  https://bettercodehub.com/ (Accessed on \today{})} to quantify maintainability
+  \url{https://bettercodehub.com/} (Accessed on \today{})} to quantify maintainability
   for both security and regular commits.
@@
 \emph{How can I adjust the threshold for passing/not passing a guideline?} at
-https://bettercodehub.com/docs/faq (Accessed on \today{})}.
+\url{https://bettercodehub.com/docs/faq} (Accessed on \today{})}.
 
@@
 Standard} that is a coding style guide for \texttt{PHP}\footnote{PSR-2 Standard available at
-https://www.php-fig.org/psr/psr-2/ (Accessed on \today{})}.
+\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})}.
 
@@
 (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
-https://support.apple.com/en-us/HT208315 (Accessed on \today{})}, where any user
+\url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
 could log in as root with an empty password. Patrick Wardle examined the cause
 of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
-https://objective-see.com/blog/blog\_0x24.html (Accessed on \today{})} and
+\url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
 concluded that the flaw was due to the introduction of high cyclomatic complexity

### 2019-02-09 paper/ICPC19/icpc19.tex
more fixes

@@
 address a security flaw because they tend to choose the easiest path to solve
-it. To address it, we compute the maintainability of $607$ commits using
-\emph{Better Code Hub}. The same approach is applied to a randomly generated
+it. To address it, we compute the maintainability of $607$ commits using the
+\emph{Better Code Hub} tool. The same approach is applied to a randomly generated
 dataset of regular commits --- baseline --- that we use to understand how
@@
 \end{framed}
-
+% here
 There are security flaws that are more difficult to refactor than others. For
@@
 The methodology used to measure the impact of security refactorings on the
-maintainability of open-source software is discussed in this section. The
-methodology is comprised of the following steps, as illustrated in
+maintainability of open-source software is discussed in this section. 
+The methodology comprises the following steps, as illustrated in
 Figure~\ref{fig:met}:
@@
 discarded from our study due to limitations of BCH: in particular, lack of
-language support and project size. The final dataset used in this paper contains
+language support and project size. The final dataset used in this paper comprises
 $607$ security refactorings labeled as one of the following categories (we
@@
     to flood or crashing services. This attacks usually occur when the system
-    receives much traffic causing it to slow down and eventually stop. For
+    receives much traffic causing it to slow down and eventually crash. For
     example, flaws allowing attackers to trigger memory allocations
@@
 
-    \item \textbf{Memory Leak (80 cases).} Memory management issue found more frequently in
-    programming languages that do not manage memory automatically (e.g., C/C++
-    and Objective-C), i.e., where instead developers are responsible for
-    handling it. One of the main causes of DoS attacks.
+    \item \textbf{Memory Leak (80 cases).} Memory management issues are found more frequently 
+	in programming languages that do not manage memory automatically (e.g., C/C++
+    and Objective-C), i.e., where developers are responsible for
+    handling it instead. It is one of the main causes of DoS attacks.
 
@@
 max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
-Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511.00  \\ \bottomrule
+Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
 \end{tabular}
@@
 The dataset contains security flaws for more than $13$ different languages,
-being PHP ($39\%$) and C ($24\%$) the most prevalent ones (see Figure~\ref{fig:lang}).
+being PHP ($39\%$) and C ($24\%$) the most prevalent ones (cf. Figure~\ref{fig:lang}).
 To classify the applications in the dataset, we use a taxonomy for open-source
@@
 	refactoring;
-	\item $v_{s}$, the version after fixing the security flaw, i.e., after the
+	\item $v_{s}$, the version free of the security flaw, i.e., after the
 	refactoring;

### 2019-02-09 paper/ICPC19/icpc19.tex
more fixes

@@
 
-We have also evaluated the impact in the maintainability per programming language
+We have also evaluated the impact on the maintainability per programming language
 of open-source software, Figure~\ref{fig:lang_main}. Although
@@
 present the results for \emph{Java}, \emph{C}, \emph{Ruby}, and \emph{PHP}
-because one of the requirements of the Paired Wilcoxon signed-rank is that the
-input sample has to have more than $20$ samples. The other languages were
+because the Paired Wilcoxon signed-rank requirement: input sample has to have more 
+than $20$ samples. The other languages were
 integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
@@
 open-source software is presented in Figure~\ref{fig:pat}. The results of the
-Paired Wilcoxon signed-rank per pattern indicated statistical
+Paired Wilcoxon signed-rank per pattern exhibited statistical
 evidence supporting that the maintainability of open-source software decreases
@@
 Vulnerabilities} (p-value = $0.001$). \emph{Injection} and \emph{Cross-Site
-Request Forgery} produced more cases where the refactorings had a negative
+Request Forgery} produced more cases where refactorings had a negative
 impact in the maintainability. However, the results obtained for these patterns
@@
  	\centering
- 	\includegraphics[width=0.56\textwidth]{figures/category.pdf}
+ 	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
  	\caption{Maintainability difference by pattern after security refactorings.}
@@
 instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-Standard} that is a coding style guide for \texttt{PHP}\footnote{PSR-2 Standard available at
-\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})}.
+Standard}\footnote{PSR-2 Standard available at
+\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
 

### 2019-02-10 scripts/histogram.py
major updates

@@ -10,7 +10,7 @@ matplotlib.rcParams['mathtext.fontset'] = 'cm'
 import matplotlib.pyplot as plt
 
 
-def language_dist(commits):
+def language_dist(commits, output):
     
     df = pd.read_csv(commits)
     
@@ -30,11 +30,17 @@ def language_dist(commits):
     dic_lang['no'].append(dic['others']/len(df['language']))
     
     for i in dic:
-        if i != 'others': 
+        if i not in ('others','php','ruby','c','python','java'): 
             dic_lang['lang'].append(lang_trans[i])
             dic_lang['no'].append(dic[i]/len(df['language']))
     
+    for i in ('php','ruby','c','python','java'):
+        dic_lang['lang'].append(lang_trans[i])
+        dic_lang['no'].append(dic[i]/len(df['language']))
+    
     lang = pd.DataFrame(dic_lang)
+    print(lang)
+    
 
     fig, ax = plt.subplots(figsize=(10, 8))
     ax = lang['no'].plot(kind='barh', fontsize=16, color="#3F86DB", ax=ax)
@@ -43,9 +49,9 @@ def language_dist(commits):
     ax.set_yticklabels(dic_lang['lang'], minor=False)
     ax.xaxis.grid(True, linestyle='--')
     fig.tight_layout()
-    fig.savefig('../paper/ICPC19/figures/language_dist.pdf')
+    fig.savefig('{}/language_dist.pdf'.format(output))
 
-def type_dist(projects):
+def type_dist(projects, output):
     
     df = pd.read_csv(projects)
     
@@ -63,6 +69,7 @@ def type_dist(projects):
         dic_type['no'].append(dic[i]) 
                 
     app_type = pd.DataFrame(dic_type)
+    print(app_type)
     
     fig, ax = plt.subplots()
     app_type['no'].plot(kind='barh', figsize=(10, 4), fontsize=16, color="#3F86DB", ax=ax)
@@ -71,20 +78,20 @@ def type_dist(projects):
     ax.set_yticklabels(app_type['type'], minor=False)
     ax.xaxis.grid(True, linestyle='--')
     fig.tight_layout()
-    fig.savefig('../paper/ICPC19/figures/type_dist.pdf')
+    fig.savefig('{}/type_dist.pdf'.format(output))
 
-def main(projects, commits):    
+def main(projects, commits, output):    
     
-    language_dist(commits)
+    language_dist(commits, output)
     plt.clf()
[…]

### 2019-02-10 paper/ICPC19/icpc19.tex
major updates
fix..?
icpc19 2nd submission version

@@
 \IEEEoverridecommandlockouts
-% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
 \usepackage{cite}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.8\linewidth]{figures/commit.pdf}
+ 	\centering 	\includegraphics[width=0.75\linewidth]{figures/commit.pdf}
  	\caption{Maintainability difference for security commits}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/maintainability.pdf}
+ 	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability.pdf}
  	\caption{Maintainability difference for security and baseline refactorings}
@@
 the dataset contains security refactorings for more than $13$ languages, we only
-present the results for \emph{Java}, \emph{C}, \emph{Ruby}, and \emph{PHP}
-because the Paired Wilcoxon signed-rank requirement: input sample has to have more 
+present the results for \emph{Java}, \emph{Python}, \emph{C}, \emph{Ruby}, and \emph{PHP}
+because of the Paired Wilcoxon signed-rank test requirement: input sample has to have more 
 than $20$ samples. The other languages were
@@
 $3$ different languages: \emph{C} (p-value = $5.72$x$10^{-9}$), Ruby (p-value = $0.057$) and
-PHP (p-value = $3.54$x$10^{-6}$). The results also provide statistical significance
+PHP (p-value = $3.54$x$10^{-6}$). Following the \textit{American Statistical Association} recommendations~\cite{doi:10.1080/00031305.2016.1154108}, 
+we consider the results for Ruby as statistically significant because its p-value is on the borderline.
+The results also exhibit statistical significance
 for when security refactorings are performed in \emph{Java} (p-value =

### 2019-02-11 paper/ICPC19/icpc19.tex
submission version

@@
 \begin{enumerate}
-	\item Extract relevant data from a dataset containing $716$ security
+	\item Extract relevant data from a dataset containing $711$ security
 	refactorings collected from open-source software available on
@@
 of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
-dataset with $716$ test cases for $16$ security patterns. Some of the patterns
+dataset with $711$ test cases for $16$ security patterns. Some of the patterns
 are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
@@
 
-The $716$ refactorings in the dataset were run against the BCH toolset to
+The $711$ refactorings in the dataset were run against the BCH toolset to
 calculate their maintainability reports. However, there are refactorings
@@
 {
- \bibliographystyle{IEEEtran}
+  \bibliographystyle{unsrt}
   \bibliography{icpc19}

### 2019-02-11 scripts/connect.py

@@ -1,19 +1,0 @@
-from utils import *
-import redis
-
-# connect to the database
-def connect_to_db(cf_filename):
-    try:
-        redis_cf = load_cf_file(cf_filename)
-        # redis connection
-        r = redis.Redis(
-            host=redis_cf['redis']['host'],
-            port=redis_cf['redis']['port'],
-            password=redis_cf['redis']['password'],
-            charset='utf-8',
-            decode_responses=True)
-        print('Succesfully connected to the database!')
-        return r
-    except Exception as ex:
-        print('Error:', ex)
-        exit('Failed to connect, Exiting...')

### 2019-02-11 scripts/dataset_statistics.py

@@ -1,0 +1,19 @@
+import maintainability.better_code_hub as bch
+import argparse
+import csv
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+
+def main(projects):    
+    df = pd.read_csv(projects)
+    stats = df[['forks','stars', 'watchers', 'contributors', 'commits', 'branches', 'releases', 'size', 't_issues', 't_prs']].describe()
+    stats.to_csv('dataset_statistics.csv')    
+
+if __name__ == "__main__":
+    
+    parser = argparse.ArgumentParser()
+    parser.add_argument('--projects-csv',metavar='projects-csv',required=True,help='the projects filename')
+
+    args = parser.parse_args()
+    main(projects=args.projects_csv)

### 2019-02-11 scripts/get_maintainability.py

@@ -21,7 +21,7 @@ def _print_hypothesis_test(differences):
     stats = {'test': [test], 'med':[pd.DataFrame({'diff': differences})['diff'].median()],'pvalue': [pvalue], 
                 'mean':[pd.DataFrame({'diff': differences})['diff'].mean()]}
     df = pd.DataFrame(stats)
-    print("Wilcoxon signed-rank test {}, p-value: {}".format(test,pvalue))
+    #print("Wilcoxon signed-rank test {}, p-value: {}".format(test,pvalue))
     return df
 
 def _hypothesis_test(differences):
@@ -49,11 +49,12 @@ def pair_info(file):
 def clean_plot():
     plt.gcf().clear()
 
-# TODO: IMPROVE
 def count(id, pattern, p):
     pattern[p][id] += 1
 
-def total_main_barchart(total_sec, total_reg, graphics, result):
+def total_main_barchart(total_sec, total_reg, reports, result):
+    
+    print(len(total_sec))
     
     sec = sum([value for key,value in total_sec.items()])
     reg = sum([value for key,value in total_reg.items()])
@@ -66,7 +67,6 @@ def total_main_barchart(total_sec, total_reg, graphics, result):
     index = np.arange(len(d['type']))
     
     df = pd.DataFrame(d)
-    print(df[['neg','pos','none']])
     
     fig = plt.figure()
     ax = fig.add_subplot(111)
@@ -88,32 +88,13 @@ def total_main_barchart(total_sec, total_reg, graphics, result):
         else:
             p = '={:.{}f}'.format(l[1]['pvalue'], 3)
         box_text = '$\overline{x}$='+ '{:.{}f}'.format(l[1]['mean'], 2) + '\nM=' + '{:.{}f}'.format(l[1]['med'], 2) + '\np' + p
-        ax.text(0.46, y, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':4}, fontsize=9)
+        ax.text(0.49, y, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':4}, fontsize=9)
         y -= 1.1
     
     plt.gca().xaxis.grid(True, linestyle='--')
     
     plt.tight_layout()
-    plt.savefig('../paper/ICPC19/figures/maintainability.pdf')
-
-def dic_barchart(dic, graphics, title, file_name):
-    d = {'neg' : pd.Series([dic[key][0] for key in dic]),
-        'pos' : pd.Series([dic[key][1] for key in dic]),
[…]

### 2019-02-11 scripts/get_vulns_commits_info.py

@@ -1,77 +1,0 @@
-from connect import connect_to_db
-import argparse
-import csv
-
-def get_patterns(conn, key):
-    pip = conn.pipeline()
-    pip.lrange(key, 0, -1)
-    return pip.execute()[0]
-
-def get_keys(conn, key):
-    pip = conn.pipeline()
-    pip.keys(pattern=key)
-    return pip.execute()
-
-def main(config, filepath):
-    
-    r = connect_to_db(config)
-    
-    vulns = get_patterns(r, 'vulns_class')
-    commits = get_keys(r,'commit:*:*:*:*')
-
-    with open(filepath,'w') as out:
-        
-        fields = ['owner', 'project', 'sha', 'sha-p', 'language', 'pattern', 'year','reported']
-        writer = csv.DictWriter(out, fieldnames=fields)    
-        writer.writeheader()
-        count = 0
-        for c in commits[0]:
-
-            _, owner, project, _, _ = c.split(':')
-            
-            # improve this
-            pip = r.pipeline()
-            pip.hget(c,'vuln?')
-            pip.hget(c,'sha')
-            pip.hget(c,'sha-p')
-            pip.hget(c,'repo_owner')
-            pip.hget(c,'repo_name')
-            pip.hget('repo:%s:%s:n'%(owner, project), 'report')
-            pip.hget(c,'lang')
-            pip.hget(c,'class')
-            pip.hget(c,'change_to')
-            pip.hget(c,'year')
-
-            ver, sha, sha_p, repo_owner, repo_name, rep, lang, pattern, change_to, year = pip.execute()
-
-            # samples verified
-            if ver == 'Y':
-                # samples reported in IJSSE
-                if rep == 'Y':
-                    reported = 1
-                else:
-                    reported = 0
-                if change_to is not None and change_to in vulns:
-                    pattern = change_to
-
-                writer.writerow({'owner' : repo_owner,
-                                'project': repo_name,
-                                'sha': sha,
-                                'sha-p': sha_p,
-                                'language': lang,
-                                'pattern': pattern,
-                                'year': year,
-                                'reported': reported
-                                })
-
-                count+=1
-
-    print(count, ' vulnerabilities verified and reported.')
-
-if __name__ == "__main__":
-    parser = argparse.ArgumentParser()
-    parser.add_argument('--redis-config',metavar='config',required=True,help='the config to redis')
[…]

### 2019-02-11 scripts/maintainability/better_code_hub.py

@@ -180,20 +180,15 @@ def _compute_maintainability_for_guideline(guideline, total_loc):
         ]
 
     bad_thresholds = guideline['qualityProfileComplianceThresholds'][1:]
-    print(bad_thresholds)
     return _compute_distance_to_thresholds(volumes, bad_thresholds)
 
 def _compute_distance_to_thresholds(volumes, bad_thresholds):
     results = []
     for index, threshold in enumerate(bad_thresholds):
-        print('threshold=', threshold)
         good_code_loc = sum(volumes[:index+1])
-        print('good_code_loc', good_code_loc)
         bad_code_loc = sum(volumes[index+1:])
-        print('bad_code_loc', bad_code_loc)
         granularity = 0.01 # thresholds have a minimum of 0.01
         bad_code_factor = ((1-threshold)/(threshold+granularity))
-        print('bad_code_factor', bad_code_factor)
         results.append(good_code_loc - bad_code_factor*bad_code_loc)
     return mean(results)
 

### 2019-02-11 scripts/utils.py

@@ -1,7 +1,0 @@
-import json
-
-# load json config file
-def load_cf_file(cf_filename):
-    with open(cf_filename) as cf:
-        data = json.load(cf)
-    return data

### 2019-05-07 paper/ICPC19/icpc19.tex
table structure

@@
 	\caption{Guidelines to produce maintainable code}
-\begin{tabular}{L{2.5cm}L{5.5cm}}
+\begin{tabular}{L{2cm}L{4cm}L{1cm}}
 
 \toprule
-\textbf{10 Guidelines} & \textbf{Description}\\
+\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
 \midrule
 \textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
- units are easier to understand, reuse and test them\\\midrule
+ units are easier to understand, reuse and test them & \\\midrule
 \textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
-it makes units easier to test and modify \\\midrule
+it makes units easier to test and modify & \\\midrule
 \textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
-multiple places (inefficient and error-prone)\\\midrule
+multiple places (inefficient and error-prone) &\\\midrule
 \textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
-$4$ because it makes units easier to understand and reuse\\\midrule
+$4$ because it makes units easier to understand and reuse & \\\midrule
 \textbf{Separate Concerns in Modules} & Avoid large modules because changes in
-loosely coupled databases are easier to oversee and execute\\\midrule
+loosely coupled databases are easier to oversee and execute & \\\midrule
 \textbf{Couple Architecture Components Loosely} & Minimize the amount of code
-within modules that are exposed to modules in other components\\\midrule
+within modules that are exposed to modules in other components & \\\midrule
 \textbf{Keep Architecture Components Balanced} & Balancing the number of
-components ease locating code and allow for isolated maintenance\\\midrule
+components ease locating code and allow for isolated maintenance & \\\midrule
 \textbf{Keep your Codebase Small} & Reduce and avoid the system size because
-small products are easier to manage and maintain\\\midrule
+small products are easier to manage and maintain & \\\midrule
 \textbf{Automate Tests} & Test your codebase because it makes development
-predictable and less risky\\\midrule
+predictable and less risky & \\\midrule
 \textbf{Write Clean Code} & Avoid producing software with code smells because
-it is more likely to be maintainable in the future\\
+it is more likely to be maintainable in the future & \\
 \bottomrule

### 2019-05-09 paper/ICPC19/icpc19.tex
paper issues addressed

@@
 \IEEEoverridecommandlockouts
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
 \usepackage{cite}
@@
 \usepackage{xcolor}
+\usepackage{footnote}
 \usepackage[hidelinks]{hyperref}
@@
 
-\title{Fixing Vulnerabilities Hinders Maintainability}
+\title{Fixing Vulnerabilities Hinders Maintainability (REG)}
 
@@
 vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
-developers address these vulnerabilities on their codebases, they actually
-affect the maintainability of software, potentially introducing other
+developers address these vulnerabilities on their code bases, they actually
+affect the maintainability and, consequently, the reliability of software $-$ potentially introducing other
 vulnerabilities. This paper evaluates the impact of refactorings to improve
@@
 and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
-attacks. Furthermore, we conclude that changes to codebases to fix vulnerabilities
-need to be performed with extra care.
+attacks. Furthermore, we conclude that changes to code bases to fix vulnerabilities
+need to be performed with extra care in open-source software.
 \end{abstract}
@@
 Software quality is important because it is ultimately related to the overall
-cost of developing, extending and maintaining software applications. Software
-quality characteristics include, but are not limited to, functional correctness,
-reliability, usability, maintainability, and security. Security is an important
+cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. Software
+quality characteristics include, but are not limited to functional correctnes,
+reliability, usability, maintainability, and security~\cite{iso:2011}. Security is an important
 non-functional requirement during the development of software systems.  However,
@@
 update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
-\emph{Security} as one of the main software product quality characteristics.
-Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
+\emph{Security} as one of the main software product quality characteristics. Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
 to find vulnerabilities and defects. Therefore, writing maintainable code
-prevents the introduction of vulnerabilities in the codebases. In this work, we
-are concerned about the impact that patching vulnerabilities may have in the
-open-source software maintainability. Are developers affecting software
-maintainability when trying to improve software security? We suspect that some
-patches may have a negative impact on the software maintainability and
-possibly even be the cause of the introduction of new vulnerabilities.
-
+supports the production of more reliable software preventing the introduction of faults/vulnerabilities in the code bases\Sof{cite smth}.
 ISO describes software maintainability as ``the degree of effectiveness and
@@
 it, correct it or adapt it to changes in environment, and in requirements'' on
-software quality ISO/IEC 25010. As ISO does not provide any specific guidelines
+software quality ISO/IEC 25010. Software maintainbility can also be defined as the property of
+software that reveals how easily the software system can be maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure.
+
+There is still little information about the impact of patching vulnerabilities
+in open-source software maintainability. 
+% Hedegus et al.
+% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
+% groups of refactored elements and non-refactored and conclude that the source
+% code elements subjected to refactorings had significantly lower maintainability
+% than the elements not modified.
+\textit{Are developers affecting software
+maintainability when trying to improve software security?}, it is what we want to investigate. Our suspiciousness is
+that some of the patches may have a negative impact on the software maintainability and possibly even be the cause of the introduction
+of new vulnerabilities $-$ harming software reliability. Therefore, in this paper we present an empirical study on the evolution of patches of vulnerabilities accross open-source software.
+
+As ISO does not provide any specific guidelines
 or formula to calculate maintainability, we resort on Software Improvement
 Group (SIG)'s web-based source code analysis service Better Code Hub
-(BCH)~\cite{Visser:2016:OREILLY} to compute the maintainability of a given
-project. BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
-base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
-based on the McCabe Complexity~\cite{1702388} and \emph{Keep your Codebase Small}
-based on the idea that smaller codebases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
-There are other well-known standards and models that have been proposed
-to increase software security: Common Criteria~\cite{common:2009} which received
-negative criticism regarding the costs associated and poor technical evaluation;
-the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
-focused only on web applications, and a model proposed by Xu et al. ($2013$)
-~\cite{6616351} for rating software security (arguably, it was one of the
-first steps taken by SIG to introduce security on their maintainability model).
-
-The particular subject of this paper is to explore whether there is a trade-off
-between refactoring source code to patch disclosed vulnerabilities and keeping
-(or even improving) the maintainability of software systems. Hedegus et al.
-($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
-groups of refactored elements and non-refactored and conclude that the source
-code elements subjected to refactorings had significantly lower maintainability
-than the elements not modified.
-
-In this paper, we present the results of our analysis on the maintainability of
-$607$ security changes collected from open-source software, from different
-programming languages. The main contributions of the present work are:
+(BCH) to compute the software compliance with a set of $10$ guidelines to
+produce quality software~\cite{Visser:2016:OREILLY} based in ISO/IEC 25010.
+The results obtained for each guideline are merged into a mathematical formula that calculates software maintainability.
+From a methodological perspective, we leveraged a dataset of $607$ security changes collected from open-source software.
+Each change provides two different versions of the software - the vulnerable one and the patched one.
+For each version, we calculate software maintainability, using the proposed formula, after BCH computes the compliance with each guideline.
+Thus, we assess the impact of security changes comparing the maintainbility value of both versions.
+We also compare the impact of security changes with regular changes on software
+maintainability.
+For that, we also collect a dataset of regular changes and do the same experiments as for security changes.
+This empirical study exhibits proof that changes applied in the code bases to fix
+vulnerabilities affect code maintainability. Especially, for refactorings that
+address \emph{Broken Authentication and Session Management} issues, \emph{Memory
+Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight
+the need for tools to predict the impact of patches on maintainability~\cite{4724577}
+and documentation to help developers adopt security best practices~\cite{6311252,
+7927935, MESQUIDA201519}.
+
+%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
+%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
+%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
+%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
+
+% There are other well-known standards and models that have been proposed
+% to increase software security: Common Criteria~\cite{common:2009} which received
+% negative criticism regarding the costs associated and poor technical evaluation;
+% the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
+% focused only on web applications, and a model proposed by Xu et al. ($2013$)
+% ~\cite{6616351} for rating software security (arguably, it was one of the
+% first steps taken by SIG to introduce security on their maintainability model).
+
+The main contributions of the present work are:
 
@@
 \end{itemize}
-
-This empirical study exhibits proof that changes applied in the codebase to fix
-vulnerabilities affect code maintainability. Especially, for refactorings that
-address \emph{Broken Authentication and Session Management} issues, \emph{Memory
-Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight
-the need for tools to predict the impact of patches on maintainability~\cite{4724577}
-and documentation to help developers adopt security best practices~\cite{6311252,
-7927935, MESQUIDA201519}.
 
@@
 hackers exploit them. It turns out that fixing these issues is as simple as
-modifying the software codebase. However, these refactorings may have a
+modifying the software code base. However, these refactorings may have a
 negative impact on software maintenance --- e.g., developers might follow the
@@
 
-  static int dtls1_reassemble_fragment(SSL *s, struct hm_header_st* msg_hdr, int *ok){
+  static int dtls1_reassemble_fragment(
+     SSL *s, struct hm_header_st* msg_hdr, int *ok){
 
@@
 - if((msg_hdr->frag_off+frag_len) > max_len)
-
 + if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
@@
   item = pqueue_find(s->d1->buffered_messages, seq64be);
-  if(item == NULL)
-  {
+  if(item == NULL){
      frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);
@@
 
- static int
- dtls1_process_out_of_seq_message(SSL *s, struct hm_header_st* msg_hdr, int *ok)
- {
+ static int dtls1_process_out_of_seq_message(
+     SSL *s, struct hm_header_st* msg_hdr, int *ok){
 
 // [snip]
-
-  if(frag_len && frag_len < msg_hdr->msg_len)
-     return dtls1_reassemble_fragment(s, msg_hdr, ok);
 
@@
 problem to repair, the example shows that there was a significant amount of
-changes performed in the codebase which produced a negative impact in the
+changes performed in the code base which produced a negative impact in the
 maintainability. Even though the developer tries to simplify the
-codebase by introducing a smaller unit
+code base by introducing a smaller unit
 \texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
@@
 (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
-\emph{Write Simple Units of Code} guideline. Moreover, it increases the codebase
+\emph{Write Simple Units of Code} guideline. Moreover, it increases the code base
 size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
-contributes to a potential violation of the \emph{Keep Your Codebase Small}
+contributes to a potential violation of the \emph{Keep Your code base Small}
 guideline. Hence, we have observed that the impact of this refactoring on the
@@
 maintainability of open-source software?}
+\vspace{-0.1cm}
 \end{framed}
+\vspace{-0.2cm}
 
@@
 maintainability of open-source software. Our suspicion is that developers tend
-to introduce technical debt in their software when refactoring the codebase to
+to introduce technical debt in their software when refactoring the code base to
 address a security flaw because they tend to choose the easiest path to solve
@@
 affect open-source software maintainability?}
+\vspace{-0.1cm}
 \end{framed}
+\vspace{-0.2cm}
 % here
@@
 \begin{enumerate}
-	\item Extract relevant data from a dataset containing $711$ security
+	\item Extract relevant data (e.g., sha, owner of the project, name of the repository) and code bases from a dataset containing $711$ security
 	refactorings collected from open-source software available on
@@
 	maintainability of open-source software.
-  \item Use the Software Improvement Group's\footnote{SIG's website is available
-  at \url{https://www.sig.eu/} (Accessed on \today{})} web-based source code analysis
-  service \emph{Better Code Hub} (BCH)\footnote{BCH's website is available at
+  \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} (Accessed on \today{})} web-based source code analysis
+  service \emph{Better Code Hub} (BCH)\footnote{BCH's website:
   \url{https://bettercodehub.com/} (Accessed on \today{})} to quantify maintainability
-  for both security and regular commits.
+  for both security and regular commits. The code bases that resemble to the commits
+  of our datasets are collected before the analysis performed by BCH.
+
 \end{enumerate}
+\vspace{-0.3cm}
 
@@
 	\label{fig:met}
+
 \end{figure}
+	\vspace{-0.2cm}
+
 %
@@
 evaluate if the impact was positive, negative or none.
+\vspace{-0.2cm}
 
@@
 \end{figure}
+\vspace{-0.2cm}
 
@@
 \end{tabular}
+\vspace{-0.3cm}
 \label{tab:dataset}
@@
 Hub} (BCH) is used to collect the maintainability reports of the refactorings of
-each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
+each project. Table~\ref{tab:guidelines} presents the $10$ guidelines $-$ proposed
 by BCH's authors for delivering software that is not difficult to
-maintain~\cite{Visser:2016:OREILLY}. During each guideline evaluation, the tool
+maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric calculated by BCH.
+Some of theses guidelines are calculated using the metrics presented in~\cite{criteria:2017}.
+During each guideline evaluation, the tool
 determines the compliance towards one guideline by establishing limits for the
@@
 
-Figure~\ref{fig:bchrep} shows an example of the report
-provided by BCH for a project after finishing its evaluation. The example
-refers to the OpenSSL CVE-$2014$-$3506$ vulnerability refactoring ---
-as described as motivating example in Section~\ref{sec:motivation}. This
-version of OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write
-Clean Code}.
-
-\begin{table}[h]
-	\caption{Guidelines to produce maintainable code}
-\begin{tabular}{L{2cm}L{4cm}L{1cm}}
-
-\toprule
-\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
-\midrule
-\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
- units are easier to understand, reuse and test them & \\\midrule
-\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
-it makes units easier to test and modify & \\\midrule
-\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
-multiple places (inefficient and error-prone) &\\\midrule
-\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
-$4$ because it makes units easier to understand and reuse & \\\midrule
-\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
-loosely coupled databases are easier to oversee and execute & \\\midrule
-\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
-within modules that are exposed to modules in other components & \\\midrule
-\textbf{Keep Architecture Components Balanced} & Balancing the number of
-components ease locating code and allow for isolated maintenance & \\\midrule
-\textbf{Keep your Codebase Small} & Reduce and avoid the system size because
-small products are easier to manage and maintain & \\\midrule
-\textbf{Automate Tests} & Test your codebase because it makes development
-predictable and less risky & \\\midrule
-\textbf{Write Clean Code} & Avoid producing software with code smells because
-it is more likely to be maintainable in the future & \\
-\bottomrule
-\end{tabular}
-\label{tab:guidelines}
-\end{table}
-
 \begin{figure}[h]
@@
 \end{figure}
+
+Figure~\ref{fig:bchrep} shows an example of the report
+provided by BCH for a project after finishing its evaluation. The example
+refers to the OpenSSL CVE-$2014$-$3506$ vulnerability refactoring ---
+as described as motivating example in Section~\ref{sec:motivation}. This
+version of OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write
+Clean Code}.
+
+\begin{table}[h]
+	\caption{Guidelines to produce maintainable code}
+\begin{tabular}{L{1.66cm}L{3.25cm}L{2.65cm}}
+
+\toprule
+\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
+\midrule
+
+\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
+ units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of LOCs within each unit~\cite{criteria:2017} \\\midrule
+
+\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
+it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision points~\cite{1702388,criteria:2017}\\\midrule
+
+\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
+multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant LOCs~\cite{criteria:2017}\\\midrule
+
+\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
+$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} \# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
+
+\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
+loosely coupled databases are easier to oversee and execute & \textbf{Module Coupling:} \# of
+incoming dependencies~\cite{criteria:2017} \\\midrule
+
+\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
+within modules that are exposed to modules in other components & \textbf{Component Independence:} \% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
+
+\textbf{Keep Architecture Components Balanced} & Balancing the number of
+components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
+
+\textbf{Keep your code base Small} & Reduce and avoid the system size because
+small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted to man-month/man-year~\cite{criteria:2017} \\\midrule
+
+\textbf{Automate Tests} & Test your code base because it makes development
+predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit complexity, component independence and volume~\cite{Visser:2016:OREILLY}
+ \\\midrule
+
+\textbf{Write Clean Code} & Avoid producing software with code smells because
+it is more likely to be maintainable in the future & Help ? \\
+\bottomrule
+\end{tabular}
+\vspace{-0.5cm}
+\label{tab:guidelines}
+\end{table}
 
@@
 \begin{equation}
-    w(l) = \frac{1 - T(l)}{T(l)}
+    w(l) = \frac{1 - \theta(l)}{\theta(l)}
 \end{equation}
@@
 \noindent
-where $T(l)$ is the threshold in percentage of the lines of code that are
+where $\theta(l)$ is the threshold in percentage of the lines of code that are
 accepted to be non-compliant with the guideline for the severity level $l$. This
@@
 
+\vspace{-0.3cm}
 \begin{figure}[h]
@@
 affect open-source software maintainability?}
+\vspace{-0.1cm}
 \end{framed}
+\vspace{-0.2cm}
 
@@
 of open-source software?}
+\vspace{-0.1cm}
 \end{framed}
+\vspace{-0.2cm}
 
@@
 The impact of a refactoring depends on its complexity, i.e., if the refactoring
-adds complexity to the codebase it is probably affecting the software
+adds complexity to the code base it is probably affecting the software
 maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting}
@@
 However, the methodology requires to access data that is not publicly available.
-Thus, our findings may not extend to non-open source software.
-
-Different programming languages may require different coding practices to
+Thus, our findings may not extend to non-open source software. Different programming languages may require different coding practices to
 address software safety. The dataset comprises more commits in C and PHP, i.e.,
 the dataset may not be representative of the population regarding
-programming languages.
-
-Our approach does not consider security refactorings in any other languages but
+programming languages. Our approach does not consider security refactorings in any other languages but
 English.

### 2019-05-09 paper/ICPC19/icpc19.tex
mapping finished

@@
 \textbf{Write Clean Code} & Avoid producing software with code smells because
-it is more likely to be maintainable in the future & Help ? \\
+it is more likely to be maintainable in the future & \textbf{Code Smells:} \# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long identifier names) \\
 \bottomrule

### 2019-05-17 paper/issre19.tex
changing path
abstract pass

--- 2019-05-09 paper/ICPC19/icpc19.tex
@@
 \begin{abstract}
-Security is a requirement of utmost importance to produce software with high
+Security is a requirement of utmost importance to produce software of high
 quality. However, there is still a considerable amount of zero-day
-vulnerabilities being discovered, and fixed, almost weekly. We suspect that when
-developers address these vulnerabilities on their code bases, they actually
-affect the maintainability and, consequently, the reliability of software $-$ potentially introducing other
-vulnerabilities. This paper evaluates the impact of refactorings to improve
-security on the maintainability of open-source software. Maintainability is
-measured based on the \emph{Better Code Hub}'s model of 10 guidelines using a
-dataset including 607 security commits. Results show that fixing software
-vulnerabilities hinders the maintainability of open-source software. This is
-particularly the case for refactorings that address \emph{Broken Authentication
+vulnerabilities being discovered, and fixed, almost weekly. We suspect that 
+when developers address these vulnerabilities on their code bases, they actually
+affect the maintainability and, pottentially, the reliability of software through the
+introduction of other vulnerabilities. This paper evaluates the impact 
+of refactorings to improve security on the maintainability of open-source software. 
+Maintainability is measured based on the \emph{Better Code Hub}'s model of 10 guidelines 
+using a dataset including 607 security commits. Results show that fixing software
+vulnerabilities hinders the maintainability of open-source software. This is,
+particularly, the case for refactorings that address \emph{Broken Authentication
 and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
@@
 Software quality is important because it is ultimately related to the overall
-cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. Software
-quality characteristics include, but are not limited to functional correctnes,
-reliability, usability, maintainability, and security~\cite{iso:2011}. Security is an important
-non-functional requirement during the development of software systems.  However,
-there is still a considerable amount of zero-day vulnerabilities being
+cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
+Software quality characteristics include, but are not limited to functional correctnes,
+reliability, usability, maintainability, and security~\cite{iso:2011}. Security is an 
+important non-functional requirement during the development of software systems. 
+However, there is still a considerable amount of zero-day vulnerabilities being
 discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
@@
 \url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
-These vulnerabilities are gateways for attackers to penetrate
-systems and exploit their resources.
-As an example, in the last week of January $2019$, a severe vulnerability in Apple
-FaceTime was revealed\footnote{FaceTime vulnerability details available at \url{https://thehackernews.com/2019/01/apple-facetime-
+These vulnerabilities are gateways for attackers to penetrate systems and 
+exploit their resources. As an example, in the last week of January $2019$, 
+a severe vulnerability in Apple FaceTime was revealed\footnote{FaceTime vulnerability 
+details available at \url{https://thehackernews.com/2019/01/apple-facetime-
 privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
@@
 update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
-\emph{Security} as one of the main software product quality characteristics. Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
+\emph{Security} as one of the main software product quality characteristics. 
+Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
 to find vulnerabilities and defects. Therefore, writing maintainable code
-supports the production of more reliable software preventing the introduction of faults/vulnerabilities in the code bases\Sof{cite smth}.
-ISO describes software maintainability as ``the degree of effectiveness and
+supports the production of more reliable software preventing the introduction of 
+faults/vulnerabilities in the code bases\Sof{cite smth}. ISO describes software 
+maintainability as ``the degree of effectiveness and
 efficiency with which a software product or system can be modified to improve
 it, correct it or adapt it to changes in environment, and in requirements'' on
-software quality ISO/IEC 25010. Software maintainbility can also be defined as the property of
-software that reveals how easily the software system can be maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure.
-
-There is still little information about the impact of patching vulnerabilities
-in open-source software maintainability. 
+software quality ISO/IEC 25010. Software maintainbility can also be defined as 
+the property of software that reveals how easily the software system can be 
+maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure.
+
 % Hedegus et al.
@@
 % than the elements not modified.
-\textit{Are developers affecting software
-maintainability when trying to improve software security?}, it is what we want to investigate. Our suspiciousness is
-that some of the patches may have a negative impact on the software maintainability and possibly even be the cause of the introduction
-of new vulnerabilities $-$ harming software reliability. Therefore, in this paper we present an empirical study on the evolution of patches of vulnerabilities accross open-source software.
-
-As ISO does not provide any specific guidelines
-or formula to calculate maintainability, we resort on Software Improvement
-Group (SIG)'s web-based source code analysis service Better Code Hub
-(BCH) to compute the software compliance with a set of $10$ guidelines to
-produce quality software~\cite{Visser:2016:OREILLY} based in ISO/IEC 25010.
-The results obtained for each guideline are merged into a mathematical formula that calculates software maintainability.
-From a methodological perspective, we leveraged a dataset of $607$ security changes collected from open-source software.
-Each change provides two different versions of the software - the vulnerable one and the patched one.
-For each version, we calculate software maintainability, using the proposed formula, after BCH computes the compliance with each guideline.
-Thus, we assess the impact of security changes comparing the maintainbility value of both versions.
+
+There is still little information about the impact of patching vulnerabilities
+in open-source software maintainability. \textit{Are developers affecting software
+maintainability when trying to improve software security?}, it is what we want to 
+investigate. Our suspiciousness is that some of the patches may have a negative 
+impact on the software maintainability and possibly even be the cause of the 
+introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
+in this paper we present an empirical study on the evolution of patches of 
+vulnerabilities accross open-source software.
+
+As ISO does not provide any specific guidelines or formula to calculate 
+maintainability, we resort on Software Improvement Group (SIG)'s web-based source 
+code analysis service Better Code Hub (BCH) to compute the software compliance 
+with a set of $10$ guidelines to produce quality software~\cite{Visser:2016:OREILLY} 
+based in ISO/IEC 25010. The results obtained for each guideline are merged into a 
+mathematical formula that calculates software maintainability. From a methodological 
+perspective, we leveraged a dataset of $607$ security changes collected from open-source 
+software. Each change provides two different versions of the software - the vulnerable 
+one and the patched one. For each version, we calculate software maintainability, using 
+the proposed formula, after BCH computes the compliance with each guideline. Thus, we 
+assess the impact of security changes comparing the maintainbility value of both versions. 
 We also compare the impact of security changes with regular changes on software
-maintainability.
-For that, we also collect a dataset of regular changes and do the same experiments as for security changes.
-This empirical study exhibits proof that changes applied in the code bases to fix
-vulnerabilities affect code maintainability. Especially, for refactorings that
-address \emph{Broken Authentication and Session Management} issues, \emph{Memory
-Leaks} and \emph{Denial-of-Service} attacks. With this study, we intend to highlight
-the need for tools to predict the impact of patches on maintainability~\cite{4724577}
-and documentation to help developers adopt security best practices~\cite{6311252,
-7927935, MESQUIDA201519}.
+maintainability. For that, we also collect a dataset of regular changes and do the same 
+experiments as for security changes. This empirical study exhibits proof that changes 
+applied in the code bases to fix vulnerabilities affect code maintainability. 
+Especially, for refactorings that address \emph{Broken Authentication and Session Management} 
+issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks. With this study, 
+we intend to highlight the need for tools to predict the impact of patches on 
+maintainability~\cite{4724577} and documentation to help developers adopt security 
+best practices~\cite{6311252, 7927935, MESQUIDA201519}.
 
@@
 the maximum certificate list size (\texttt{s-$>$max\_cert\_list}) requires it. More
-specifically, it checks if the maximum certificate list size requires more length than the maximum length
-(\texttt{max\_len}). If that condition holds, the function returns the size of
-the certificate list, otherwise, it returns the maximum length. This function
-replaces lines $18$--$21$.
+specifically, it checks if the maximum certificate list size requires more length 
+than the maximum length (\texttt{max\_len}). If that condition holds, the function 
+returns the size of the certificate list, otherwise, it returns the maximum length. 
+This function replaces lines $18$--$21$.
 
@@
 \begin{enumerate}
-	\item Extract relevant data (e.g., sha, owner of the project, name of the repository) and code bases from a dataset containing $711$ security
+	\item Extract relevant data (e.g., sha, owner of the project, name of 
+	the repository) and code bases from a dataset containing $711$ security
 	refactorings collected from open-source software available on
@@
 	maintainability of open-source software.
-  \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} (Accessed on \today{})} web-based source code analysis
+  \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} 
+  (Accessed on \today{})} web-based source code analysis
   service \emph{Better Code Hub} (BCH)\footnote{BCH's website:
@@
 by BCH's authors for delivering software that is not difficult to
-maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric calculated by BCH.
-Some of theses guidelines are calculated using the metrics presented in~\cite{criteria:2017}.
-During each guideline evaluation, the tool
+maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric 
+calculated by BCH. Some of theses guidelines are calculated using the metrics 
+presented in~\cite{criteria:2017}. During each guideline evaluation, the tool
 determines the compliance towards one guideline by establishing limits for the
@@
 \textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
- units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of LOCs within each unit~\cite{criteria:2017} \\\midrule
+ units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of 
+ LOCs within each unit~\cite{criteria:2017} \\\midrule
 
 \textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
-it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision points~\cite{1702388,criteria:2017}\\\midrule
+it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision 
+points~\cite{1702388,criteria:2017}\\\midrule
 
 \textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
-multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant LOCs~\cite{criteria:2017}\\\midrule
+multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant 
+LOCs~\cite{criteria:2017}\\\midrule
 
 \textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
-$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} \# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
+$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} 
+\# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
 
@@
 \textbf{Couple Architecture Components Loosely} & Minimize the amount of code
-within modules that are exposed to modules in other components & \textbf{Component Independence:} \% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
+within modules that are exposed to modules in other components & \textbf{Component Independence:} 
+\% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
 
 \textbf{Keep Architecture Components Balanced} & Balancing the number of
-components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
+components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} 
+Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
 
 \textbf{Keep your code base Small} & Reduce and avoid the system size because
-small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted to man-month/man-year~\cite{criteria:2017} \\\midrule
+small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted 
+to man-month/man-year~\cite{criteria:2017} \\\midrule
 
 \textbf{Automate Tests} & Test your code base because it makes development
-predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit complexity, component independence and volume~\cite{Visser:2016:OREILLY}
+predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit 
+complexity, component independence and volume~\cite{Visser:2016:OREILLY}
  \\\midrule
@@
 \textbf{Write Clean Code} & Avoid producing software with code smells because
-it is more likely to be maintainable in the future & Help ? \\
+it is more likely to be maintainable in the future & \textbf{Code Smells:} 
+\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
+identifier names) \\
 \bottomrule
@@
 $3$ different languages: \emph{C} (p-value = $5.72$x$10^{-9}$), Ruby (p-value = $0.057$) and
-PHP (p-value = $3.54$x$10^{-6}$). Following the \textit{American Statistical Association} recommendations~\cite{doi:10.1080/00031305.2016.1154108}, 
-we consider the results for Ruby as statistically significant because its p-value is on the borderline.
-The results also exhibit statistical significance
-for when security refactorings are performed in \emph{Java} (p-value =
-$3.61$x$10^{-2}$). \emph{Python} and the languages that integrate the
-\emph{Others} set contain more cases where the maintainability increases.
-However, the results obtained for these patterns are not statistically
-significant.
+PHP (p-value = $3.54$x$10^{-6}$). Following the \textit{American Statistical Association} 
+recommendations~\cite{doi:10.1080/00031305.2016.1154108}, we consider the results for 
+Ruby as statistically significant because its p-value is on the borderline.
+The results also exhibit statistical significance for when security refactorings 
+are performed in \emph{Java} (p-value = $3.61$x$10^{-2}$). \emph{Python} and the 
+languages that integrate the \emph{Others} set contain more cases where the 
+maintainability increases. However, the results obtained for these patterns are 
+not statistically significant.
 
@@
 However, the methodology requires to access data that is not publicly available.
-Thus, our findings may not extend to non-open source software. Different programming languages may require different coding practices to
-address software safety. The dataset comprises more commits in C and PHP, i.e.,
-the dataset may not be representative of the population regarding
-programming languages. Our approach does not consider security refactorings in any other languages but
-English.
+Thus, our findings may not extend to non-open source software. Different programming 
+languages may require different coding practices to address software safety. The 
+dataset comprises more commits in C and PHP, i.e., the dataset may not be representative 
+of the population regarding programming languages. Our approach does not consider 
+security refactorings in any other languages but English.
 
@@
 Better Code Hub reports to calculate maintainability
-based on a model proposed in previous work~\cite{Olivari:2018}. Security refactorings significantly
-affect the software maintainability of open-source software, i.e., developers
-hinder maintainability when fixing vulnerabilities. We found this evidence in
+based on a model proposed in previous work~\cite{Olivari:2018}. Security refactorings 
+significantly affect the software maintainability of open-source software, i.e., 
+developers hinder maintainability when fixing vulnerabilities. We found this evidence in
 $40.2\%$ of the studied cases. Regarding regular commits, it seems that overall
@@
   \bibliographystyle{unsrt}
-  \bibliography{icpc19}
+  \bibliography{issre19}
 }

### 2019-05-23 paper/issre19.tex

@@
 
+\makeatletter
+\newcommand\footnoteref[1]{\protected@xdef\@thefnmark{\ref{#1}}\@footnotemark}
+\makeatother
+
 \begin{document}
@@
 quality. However, there is still a considerable amount of zero-day
-vulnerabilities being discovered, and fixed, almost weekly. We suspect that 
+vulnerabilities being discovered and fixed, almost weekly. We suspect that 
 when developers address these vulnerabilities on their code bases, they actually
-affect the maintainability and, pottentially, the reliability of software through the
-introduction of other vulnerabilities. This paper evaluates the impact 
+affect the maintainability and, potentially, the reliability of software through the
+introduction of new vulnerabilities. This paper evaluates the impact 
 of refactorings to improve security on the maintainability of open-source software. 
@@
 Software quality characteristics include, but are not limited to functional correctnes,
-reliability, usability, maintainability, and security~\cite{iso:2011}. Security is an 
+reliability, usability, maintainability, and security. Security is an 
 important non-functional requirement during the development of software systems. 
@@
 In $2011$, the International Organization for Standardization (ISO) issued an
-update for software product quality ISO/IEC 25010~\cite{iso:2011} considering
-\emph{Security} as one of the main software product quality characteristics. 
+update for software product quality ISO/IEC 25010 considering
+\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
+However, many developers still lack the knowledge about best practices to deliver secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
 Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
@@
 supports the production of more reliable software preventing the introduction of 
-faults/vulnerabilities in the code bases\Sof{cite smth}. ISO describes software 
+faults/vulnerabilities in the code bases. ISO describes software 
 maintainability as ``the degree of effectiveness and
@@
 it, correct it or adapt it to changes in environment, and in requirements'' on
-software quality ISO/IEC 25010. Software maintainbility can also be defined as 
-the property of software that reveals how easily the software system can be 
-maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure.
+software quality ISO/IEC 25010~\cite{iso:2011}. Novel tools have been built to 
+detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
+Despite these efforts, improving software security is not a trivial task and requires 
+implementing refactorings that might affect software maintainability. 
+Our suspiciousness is that some of the these patches may have a negative 
+impact on the software maintainability and, possibly, even be the cause of the 
+introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
+in this paper, we present an empirical study on the impact of patches of 
+vulnerabilities in software maintenance accross open-source software.
+
+% Software maintainability can also be defined as 
+% the property of software that reveals how easily the software system can be 
+% maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure.
 
@@
 
-There is still little information about the impact of patching vulnerabilities
-in open-source software maintainability. \textit{Are developers affecting software
-maintainability when trying to improve software security?}, it is what we want to 
-investigate. Our suspiciousness is that some of the patches may have a negative 
-impact on the software maintainability and possibly even be the cause of the 
-introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
-in this paper we present an empirical study on the evolution of patches of 
-vulnerabilities accross open-source software.
-
-As ISO does not provide any specific guidelines or formula to calculate 
+% There is still little information about the impact of patching vulnerabilities
+% in open-source software maintainability. \textit{Are developers affecting software
+% maintainability when trying to improve software security?}, it is what we want to
+% investigate.
+
+As ISO does not provide any specific guidelines or formulas to calculate 
 maintainability, we resort on Software Improvement Group (SIG)'s web-based source 
 code analysis service Better Code Hub (BCH) to compute the software compliance 
-with a set of $10$ guidelines to produce quality software~\cite{Visser:2016:OREILLY} 
-based in ISO/IEC 25010. The results obtained for each guideline are merged into a 
-mathematical formula that calculates software maintainability. From a methodological 
-perspective, we leveraged a dataset of $607$ security changes collected from open-source 
-software. Each change provides two different versions of the software - the vulnerable 
-one and the patched one. For each version, we calculate software maintainability, using 
-the proposed formula, after BCH computes the compliance with each guideline. Thus, we 
-assess the impact of security changes comparing the maintainbility value of both versions. 
-We also compare the impact of security changes with regular changes on software
-maintainability. For that, we also collect a dataset of regular changes and do the same 
-experiments as for security changes. This empirical study exhibits proof that changes 
+with a set of $10$ guidelines/metrics to produce quality software 
+based in ISO/IEC 25010~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
+were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041}. 
+There are other well-known standards and models that have been proposed
+to increase software security: Common Criteria~\cite{common:2009} which received
+negative criticism regarding the costs associated and poor technical evaluation;
+the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
+focused only on web applications, and a model proposed by Xu et al. ($2013$)
+~\cite{6616351} for rating software security (arguably, it was one of the
+first steps taken by SIG to introduce security on their maintainability model). 
+We use BCH's toolset to provide an assessment of maintainability in 
+software mainly due to the broad number of metrics used.
+
+From a methodological 
+perspective, we leveraged a dataset of $607$ security refactorings collected from open-source 
+software. We calculate software maintainability before and after the refactoring to measure its 
+impact. This empirical study exhibits proof that changes 
 applied in the code bases to fix vulnerabilities affect code maintainability. 
 Especially, for refactorings that address \emph{Broken Authentication and Session Management} 
-issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks. With this study, 
+issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks. This is important work
+because there is still little information about the impact of security patches on the software
+maintainability. With this study, 
 we intend to highlight the need for tools to predict the impact of patches on 
@@
 %based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
-
-% There are other well-known standards and models that have been proposed
-% to increase software security: Common Criteria~\cite{common:2009} which received
-% negative criticism regarding the costs associated and poor technical evaluation;
-% the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
-% focused only on web applications, and a model proposed by Xu et al. ($2013$)
-% ~\cite{6616351} for rating software security (arguably, it was one of the
-% first steps taken by SIG to introduce security on their maintainability model).
 
@@
 example of a security refactoring of a known vulnerability found in
-OpenSSL\footnote{Repository available at \url{https://github.com/openssl/openssl}
+OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
+contains open-source implementations of the SSL and TLS cryptographic
+protocols. Repository available at \url{https://github.com/openssl/openssl}
 (Accessed on \today{})}; section~\ref{sec:methodology} describes the
@@
 %
-Due to time-to-market pressure and the lack of security expertise, code-related
+Due to time-to-market pressure and the lack of security expertise~\cite{8077802}, code-related
 security flaws are generally detected \textit{after the fact}, i.e., when
@@
 As an example, consider the refactoring of the Datagram Transport Layer Security
-(DTLS) protocol implementation in OpenSSL\footnote{OpenSSL is a toolkit that
-contains open-source implementations of the SSL and TLS cryptographic
-protocols.} to address a vulnerability where an attacker could cause a Denial-of-Service
+(DTLS) protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
 (DoS) attack by crafting DTLS handshake messages to trigger memory allocations
@@
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-evaluation. The snippet in Listing~\ref{lst:vuln} presents the changes performed on the
+evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
 \emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
 \url{https://github.com/openssl/openssl/commit/1250f12613b61758675848f6600ebd914ccd7636}
-(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability.
-
-Every SSL/TLS connection begins with a handshake which is responsible by the
+(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
 negotiation between the two parties. DTLS provides a mechanism for fragmenting
@@
 receiver is not verified against the maximum size allowed by DTLS. It is then
-allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~37). Thus, if an
-attacker has access to one of these packages, the attacker can change the
+allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~36). Thus, if an
+attacker has access to one of these packages, he can change the
 handshake message size and trigger unnecessary memory allocations of large
@@
 
-The code changes performed to fix the vulnerability were the following:
+The code changes performed to fix the vulnerability are:
 
@@
 returns the size of the certificate list, otherwise, it returns the maximum length. 
-This function replaces lines $18$--$21$.
+This function replaces lines $17$--$21$.
 
@@
 code base by introducing a smaller unit
-\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change also
-disrupts two of the guidelines proposed by the Software Improvement Group (SIG) for
+\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change disrupts
+one of the guidelines proposed by the Software Improvement Group (SIG) for
 building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
@@
 (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
-\emph{Write Simple Units of Code} guideline. Moreover, it increases the code base
-size in $4$ lines of code ($13$ lines added, $9$ lines deleted) which
-contributes to a potential violation of the \emph{Keep Your code base Small}
-guideline. Hence, we have observed that the impact of this refactoring on the
-OpenSSL maintainability was negative.
+\emph{Write Simple Units of Code} guideline which resulted in a negative impact on the
+OpenSSL maintainability.
 
@@
 maintainability of open-source software?}
+\end{framed}
 \vspace{-0.1cm}
-\end{framed}
-\vspace{-0.2cm}
 
@@
 affect open-source software maintainability?}
+\end{framed}
 \vspace{-0.1cm}
-\end{framed}
-\vspace{-0.2cm}
 % here
@@
 authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
-separately for each pattern.
+separately for each pattern.\\
 
@@
 maintainability of open-source software?}
+\vspace{-0.05cm}
+\end{framed}
 \vspace{-0.1cm}
-\end{framed}
-\vspace{-0.2cm}
 
@@
 affect open-source software maintainability?}
+\vspace{-0.05cm}
+\end{framed}
 \vspace{-0.1cm}
-\end{framed}
-\vspace{-0.2cm}
 
@@
 of open-source software?}
+\vspace{-0.05cm}
+\end{framed}
 \vspace{-0.1cm}
-\end{framed}
-\vspace{-0.2cm}
 
@@
 affect open-source software maintainability?}
-\vspace{-0.1cm}
+\vspace{-0.05cm}
 \end{framed}
-\vspace{-0.2cm}
+\vspace{-0.05cm}
 
@@
 to fix security flaws. This approach discards security refactorings that were
-not explicit in commits messages.
-
-Baseline commits are retrieved randomly by selecting one commit from the same
+not explicit in commits messages. Baseline commits are retrieved randomly by selecting one commit from the same
 project of a given security refactoring. This approach softens the differences
@@
 use of design patterns may introduce maintainability issues into
-software~\cite{4493325}. The present work studies how security patterns
-influence maintainability for open-source software.
+software~\cite{4493325}.
+Yskout et.
+al were not able to detect if the usage of design patterns has a positive impact, but concludes that developers prefer to work with the support of security patterns~\cite{8077802}.
+The present work studies how security patterns influence maintainability for open-source software.
 
@@
 refactorings on the maintainability of $94$ open-source projects. We leveraged
-Better Code Hub reports to calculate maintainability
-based on a model proposed in previous work~\cite{Olivari:2018}. Security refactorings 
+Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018}. Security refactorings 
 significantly affect the software maintainability of open-source software, i.e., 
@@
 %ack pedro adao when accepted.
-
+\pagebreak
 \balance

### 2020-01-11 paper/icpc20.tex
organizing papers folders

--- 2019-05-23 paper/issre19.tex
@@
-\documentclass[10pt,conference]{IEEEtran}
-\IEEEoverridecommandlockouts
+\documentclass[sigconf,review]{acmart}
+
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
+
 \usepackage[utf8]{inputenc}
 \usepackage[T1]{fontenc}
-\usepackage{cite}
 \usepackage{amsmath,amssymb,amsfonts}
@@
 \usepackage{booktabs}
-\usepackage[table,xcdraw]{xcolor}
 \usepackage{framed}
@@
 \usepackage{footnote}
-\usepackage[hidelinks]{hyperref}
 \usepackage{xfrac}
-
 \usepackage{balance}
@@
 
-\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
-    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
-
 %
@@
 
+\setcopyright{acmcopyright}
+\copyrightyear{2020}
+\acmYear{2020}
+%\acmDOI{}
+
+\acmConference[ICPC'20]{ICPC '20: International Conference in Program Comprehension}{May 23--24, 2020}{Seoul, South Korea}
+% \acmBooktitle{}
+% \acmPrice{}
+% \acmISBN{}
+
 \begin{document}
 
-\title{Fixing Vulnerabilities Hinders Maintainability (REG)}
+\title{Fixing Vulnerabilities Hinders Maintainability}
 
@@
     Anonymou(s) Author(s)
-
-%     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-% \textit{name of organization (of Aff.)}\\
-% City, Country \\
-% email address}
-% \and
-% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-% \textit{name of organization (of Aff.)}\\
-% City, Country \\
-% email address}
-% \and
-% \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-% \textit{name of organization (of Aff.)}\\
-% City, Country \\
-% email address}
 }
 
-\maketitle
-
+\renewcommand{\shortauthors}{Reis et al.}
+
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
 \begin{abstract}
@@
 
-\begin{IEEEkeywords}
-Security, Software Maintenance, Open-Source Software
-\end{IEEEkeywords}
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+\begin{CCSXML}
+<ccs2012>
+ <concept>
+  <concept_id>10010520.10010553.10010562</concept_id>
+  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+  <concept_significance>500</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010575.10010755</concept_id>
+  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+  <concept_significance>300</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10010520.10010553.10010554</concept_id>
+  <concept_desc>Computer systems organization~Robotics</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>10003033.10003083.10003095</concept_id>
+  <concept_desc>Networks~Network reliability</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Computer systems organization~Embedded systems}
+\ccsdesc[300]{Computer systems organization~Redundancy}
+\ccsdesc{Computer systems organization~Robotics}
+\ccsdesc[100]{Networks~Network reliability}
+
+\keywords{Security, Software Maintenance, Open-Source Software}
+
+\maketitle
 
 \section{Introduction}
-
+%
 Software quality is important because it is ultimately related to the overall
@@
 flaw that allowed callers to see and hear others without them picking up the call.
-
+%
 In $2011$, the International Organization for Standardization (ISO) issued an
@@
 detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
+%
 Despite these efforts, improving software security is not a trivial task and requires 
@@
 vulnerabilities in software maintenance accross open-source software.
-
+%
 % Software maintainability can also be defined as 
@@
 % maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure. 
-
 % Hedegus et al.
@@
 % than the elements not modified.
-
 % There is still little information about the impact of patching vulnerabilities
@@
 % investigate.
-
+%
 As ISO does not provide any specific guidelines or formulas to calculate 
@@
 first steps taken by SIG to introduce security on their maintainability model). 
+%
 We use BCH's toolset to provide an assessment of maintainability in 
 software mainly due to the broad number of metrics used.
-
 From a methodological 
@@
 best practices~\cite{6311252, 7927935, MESQUIDA201519}.
-
+%
 %BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
@@
 %based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
-
+%
 The main contributions of the present work are:
-
+%
 \begin{itemize}
@@
 \end{itemize}
-
+%
 This paper is structured as follows: section~\ref{sec:motivation} introduces an
@@
 the maintainability of software. 
-
 As an example, consider the refactoring of the Datagram Transport Layer Security
@@
 length which can lead to memory exhaustion that could end in a DoS attack.
-
 \medskip
-
 \setcounter{lstannotation}{0}
@@
 maintainability of open-source software?}
-\vspace{-0.05cm}
+\vspace{-0.1cm}
 \end{framed}
@@
 
-\vspace{-0.3cm}
 \begin{figure}[h]
+	\vspace{-0.2cm}
+
  	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability.pdf}
@@
 	\label{fig:secvsreg}
+	\vspace{-0.2cm}
+
 \end{figure}
@@
 
-\vspace{-0.3cm}
 \begin{figure}[h]
+\vspace{-0.2cm}
   \centering
@@
   \label{fig:lang_main}
+  \vspace{-0.2cm}
+
 \end{figure}
@@
 affect open-source software maintainability?}
-\vspace{-0.05cm}
+\vspace{-0.1cm}
 \end{framed}
@@
 
-\vspace{-0.3cm}
 \begin{figure}[h]
+	\vspace{-0.2cm}
  	\centering
@@
 	\label{fig:pat}
+	\vspace{-0.2cm}
+
 \end{figure}
@@
 of open-source software?}
-\vspace{-0.05cm}
+\vspace{-0.1cm}
 \end{framed}
@@
 affect open-source software maintainability?}
-\vspace{-0.05cm}
+\vspace{-0.1cm}
 \end{framed}
-\vspace{-0.05cm}
+\vspace{-0.1cm}
 
@@
 
-\section*{Acknowledgment}\label{sec:ack}
-
-\noindent
+\begin{acks}
 We thank the Software Improvement Group (SIG) team for all the support and
 help in validating the methodology.
-%ack pedro adao when accepted.
-\pagebreak
-\balance
-
-{
-  \bibliographystyle{unsrt}
-  \bibliography{issre19}
-}
+\end{acks}
+
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{icpc20}
 
 \end{document}
+\endinput
+%%
+%% End of file `sample-sigconf.tex'.

### 2020-01-24 paper/icpc20.tex
methodology 1st version

@@
 \begin{abstract}
-Security is a requirement of utmost importance to produce software of high
-quality. However, there is still a considerable amount of zero-day
-vulnerabilities being discovered and fixed, almost weekly. We suspect that 
-when developers address these vulnerabilities on their code bases, they actually
-affect the maintainability and, potentially, the reliability of software through the
-introduction of new vulnerabilities. This paper evaluates the impact 
-of refactorings to improve security on the maintainability of open-source software. 
-Maintainability is measured based on the \emph{Better Code Hub}'s model of 10 guidelines 
-using a dataset including 607 security commits. Results show that fixing software
-vulnerabilities hinders the maintainability of open-source software. This is,
-particularly, the case for refactorings that address \emph{Broken Authentication
-and Session Management} issues, \emph{Memory Leaks} and \emph{Denial-of-Service}
-attacks. Furthermore, we conclude that changes to code bases to fix vulnerabilities
-need to be performed with extra care in open-source software.
+
+Security is a requirement of utmost importance to produce software with 
+high quality. However, there is still a considerable amount of 
+vulnerabilities being discovered and fixed, almost weekly. We suspect 
+that when developers address these vulnerabilities on their codebases, 
+they actually affect the maintainability of software, potentially 
+introducing other vulnerabilities. This paper evaluates the impact of 
+patches to improve security on the maintainability of open-source software. 
+Maintainability is measured based on the Better Code Hub’s model of 10 
+guidelines using a dataset including 1020 security commits. Results show 
+that fixing software vulnerabilities hinders the maintainability of 
+open-source software. We conclude that changes to codebases 
+to fix vulnerabilities need to be performed with extra care.
+
 \end{abstract}
@@
 %
-The methodology used to measure the impact of security refactorings on the
+The methodology used to measure the impact of security patches on the
 maintainability of open-source software is discussed in this section.
@@
 \begin{enumerate}
-	\item Extract relevant data (e.g., sha, owner of the project, name of 
-	the repository) and code bases from a dataset containing $711$ security
+	\item Combine the datasets from related work that classifies
+the activities of developers addressing security-oriented patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064}.
+	\item
+	Extract relevant data (e.g., sha, sha of the parent commit, owner of the project, name of 
+	the repository) and code bases from the combined dataset containing $1325$ security
 	refactorings collected from open-source software available on
-	GitHub~\cite{Reis:2017:IJSSE}.
+	GitHub. We consider the previous version of the patch as a vulnerable version
+	of the software.
 	\item A baseline of regular commits was randomly collected from the list of
-	projects of the main dataset to evaluate the impact of regular commits on the
+	projects of the final dataset to evaluate the impact of regular commits on the
 	maintainability of open-source software.
@@
  	\centering 	\includegraphics[width=0.4\textwidth]{figures/methodology.pdf}
- 	\caption{Study Methodology}
+ 	\caption{Study Methodology\Sof{Needs to be updated. Add Vulas.}}
 	\label{fig:met}
@@
 %
-\subsection{Dataset}
+\subsection{Datasets}
 %
 To evaluate the impact of security refactorings on the maintainability of
-open-source software, we use a dataset of security flaws which is the outcome of
-mining $94$ GitHub projects. Reis and Abreu ($2017$) mined open-source
+open-source software, we use a combined dataset of security flaws which is 
+the outcome of mining/manually inspect $301$ GitHub projects. Reis and Abreu 
+($2017$) mined open-source
 software aiming at the extraction of real examples --- created by real
@@
 evaluate if the impact was positive, negative or none.
-\vspace{-0.2cm}
-
-\begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
- 	\caption{Distribution of Security Refactorings per Language}
-	\label{fig:lang}
-\end{figure}
-\vspace{-0.2cm}
-
-\begin{figure}[h]
- 	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
- 	\caption{Distribution of the Domain of Projects}
-	\label{fig:domain}
-\end{figure}
-
-The $711$ refactorings in the dataset were run against the BCH toolset to
+
+Ponta et al. ($2019$) tracked the \url{pivotal.io} website for vulnerabilities 
+from $2014$ to $2019$. For each new vulnerability, the authors manually searched 
+for references to commits involved in the patch on the NVD's website. However, $70\%$
+of the vulnerabilities did not have any references to commits. Thus, the authors
+used their expertise to locate the commits in the repositories. This technique 
+yielded a database of $624$ patches. The original dataset contained $1282$ commits. 
+However, for some vulnerabilities, one commit was not enough to patch the vulnerability. 
+Thus, many of the entries were parts of the patch. For our purpose, we shrank the dataset 
+and only considered the first and last commit used to patch the vulnerability. The $1325$ 
+patches in the dataset were run against the BCH toolset to
 calculate their maintainability reports. However, there are refactorings
@@
 language support and project size. The final dataset used in this paper comprises
-$607$ security refactorings labeled as one of the following categories (we
-have decided to change the label to \textit{Miscellaneous} of the patterns in
-the dataset containing less than $20$ instances):
+$1020$ security patches. $547$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE). Some of the most prevailing categories are the following (categories
+with less than $20$ instances were merged in a bigger category named \textit{Miscellaneous}):
+
+% \begin{figure}[h]
+%  	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
+%  	\caption{Distribution of Security Refactorings per Language}
+% 	\label{fig:lang}
+% \end{figure}
+% \vspace{-0.2cm}
+%
+% \begin{figure}[h]
+%  	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
+%  	\caption{Distribution of the Domain of Projects}
+% 	\label{fig:domain}
+% \end{figure}
+
+The $1325$ patches in the dataset were run against the BCH toolset to
+calculate their maintainability reports. However, there are refactorings
+discarded from our study due to limitations of BCH: in particular, lack of
+language support and project size. The final dataset used in this paper comprises
+$1020$ security patches. $547$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE). Some of the most prevailing categories are the following (categories
+with less than $20$ instances were merged in a bigger category named \textit{Miscellaneous}):
 
 \begin{itemize}
-    \item \textbf{Using Components with Known Vulnerabilities (21 cases).} The majority of
-    software produced today integrates several components, such as libraries and
-    frameworks, which may affect the software if developers use vulnerable
-    versions - usually known and disclosed somewhere on the Internet.
-    \textit{Based on the OWASP Top $10$ of $2017$.}
-
-    \item \textbf{Broken Authentication \& Session Management (38 cases).} Incorrectly
+    \item \textbf{CW-89: Improper Neutralization of Special Elements used in an SQL Command/
+	SQL Injection (23 cases).} When developers do not keep untrusted data
+    separate from SQL queries. If an attacker sends a command that
+    exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
+
+    \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).} 
+	XML documents are exchanged through the web containing entities with URIs
+	that resolve to local/external files. Thus, when XML parsers are not well configured, 
+	attackers have allowed to directly those files.
+
+    \item \textbf{CWE-20: Improper Input Validation (75 cases).} When developers
+	forget to validate the inputs of an application, attackers may have control
+	of the control/data flow of the program.
+
+    \item \textbf{CWE-200: Information Exposure (38 cases).}
+
+	Application's logs are one way of intentional/unintentional disclosure information
+	to attackers. Many times attackers get access to logs when sould not be authorized to.
+
+	\item \textbf{CWE-352: Cross-Site Request Forgery (CSRF) (20 cases).} Poor session tokens generation
+    and management usually allow attackers to send forged HTTP requests
+    including authentication information from the victim to the vulnerable web
+    application.
+
+	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (36 cases).}
+
+	Incorrectly
 	    implemented functionalities related to authentication and session
 	    management, allowing an attacker to gain access to session tokens,
-    passwords, keys, and other sensitive data. \textit{Based on the OWASP Top
-    $10$ of $2013$.}
-
-    \item \textbf{Cross-Site Request Forgery (30 cases).} Poor session tokens generation
-    and management usually allow attackers to send forged HTTP requests
-    including authentication information from the victim to the vulnerable web
-    application. \textit{Based on the OWASP Top $10$ of $2013$.}
-
-    \item \textbf{Denial-of-Service (40 cases).} Security flaws that can allow an attacker
-    to flood or crashing services. This attacks usually occur when the system
-    receives much traffic causing it to slow down and eventually crash. For
-    example, flaws allowing attackers to trigger memory allocations
-    corresponding to large length values (Listing~\ref{lst:vuln}).
-
-    \item \textbf{Cross-Site Scripting (114 cases).} Lack of proper validation or escaping
-    allows attackers to submit untrusted data to web browsers through malicious
-    scripts that can hijack the user sessions or redirect the user to malicious
-    sites. \textit{Based on the OWASP Top $10$ of $2017$.}
-
-    \item \textbf{Injection (89 cases).} When developers do not keep untrusted data
-    separate from commands and queries. If an attacker sends a string that
-    exploits the syntax of the interpreter, then an injection attack is possible
-    (e.g., SQL and LDAP injection). \textit{Based on the OWASP Top $10$ of
-    $2017$.}
-
-    \item \textbf{Memory Leak (80 cases).} Memory management issues are found more frequently
+	    passwords, keys, and other sensitive data.
+
+	\item \textbf{CWE-399: Resource Management Errors (20 cases).} 
+	Resource management issues are found more frequently
 		in programming languages that do not manage memory automatically (e.g., C/C++
@@
 
-    \item \textbf{Miscellaneous (195 cases)} This pattern comprises
-	several other security refactorings that do not have their own pattern yet and patterns that
-	do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings:
-    \textit{Resource Leak} (19), \textit{SHA-1 Hash Function} (1),
-    \textit{Broken Access Control} (2), \textit{Sensitive Data Exposure} (18),
-    \textit{Insufficient Attack Protection} (4), \textit{Overflow} (18),
-    \textit{Path Traversal} (19). \end{itemize}
-
-\begin{table*}[h]
-	\centering
-	\caption{Descriptive statistics of the projects in the dataset}
-\begin{tabular}{@{}rrrrrrrrrrr@{}}
-\toprule
-      & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
-mean  & 1763 & 5448 & 401   & 153       & 14834 & 45    & 129   & 122973 & 3768   & 1941 \\
-std   & 2434 & 6215 & 486   & 123       & 22234 & 150   & 189   & 209732 & 5933   & 3603 \\
-min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
-25\%  & 391  & 1581    & 117   & 49           & 1440  & 4        & 19       & 8466   & 313    & 143  \\
-median  & 838  & 2836 & 248      & 99           & 5504  & 9        & 59       & 37372  & 1792   & 650     \\
-75\%  & 2155    & 6828 & 459   & 261          & 18579 & 20       & 142   & 117699 & 4087   & 1907 \\
-max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
-Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
-\end{tabular}
-\vspace{-0.3cm}
-\label{tab:dataset}
-\end{table*}
-
-The dataset contains security flaws for more than $13$ different languages,
-being PHP ($39\%$) and C ($24\%$) the most prevalent ones (cf. Figure~\ref{fig:lang}).
-To classify the applications in the dataset, we use a taxonomy for open-source
-software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
-presents the projects domain distribution: \textit{Application Software} ($25$),
-software that provides end-users with functional systems; \textit{Web libraries
-and frameworks} ($20$); \textit{System Software} ($18$), software that provides
-services and infrastructures (e.g., operating systems, servers and databases);
-\textit{Software Tool} ($13$), software that supports development (e.g.,
-programming languages, compilers, package managers, IDEs); and, \textit{Non-web
-libraries and frameworks} ($18$), for desktop and mobile software.
-%
-Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
-open-source projects involved in this study, including the number of forks, the
-number of stars, the number of watchers, the number of contributors, the number
-of commits, the number of branches, the number of releases, size (cf. GitHub, it
-is the size of the whole repository, including all of its history, in
-kilobytes), the number of issues, and the number of pull requests.
+	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)}
+	Lack of proper validation or escaping
+	    allows attackers to submit untrusted data to web browsers through malicious
+	    scripts that can hijack the user sessions or redirect the user to malicious
+	    sites.
+
+	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (36 cases).} 
+	When developers forget to neutralize special elements within the pathnames to files/directories, attackers 
+	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
+	or entire repositories.
+
+	\item \textbf{Miscellaneous (473 cases)} This pattern comprises
+		several other security refactorings that do not have a CWE assigned and patterns that
+		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
+
+\end{itemize}
+
+% \begin{table*}[h]
+% 	\centering
+% 	\caption{Descriptive statistics of the projects in the dataset}
+% \begin{tabular}{@{}rrrrrrrrrrr@{}}
+% \toprule
+%       & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
+% mean  & 1763 & 5448 & 401   & 153       & 14834 & 45    & 129   & 122973 & 3768   & 1941 \\
+% std   & 2434 & 6215 & 486   & 123       & 22234 & 150   & 189   & 209732 & 5933   & 3603 \\
+% min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
+% 25\%  & 391  & 1581    & 117   & 49           & 1440  & 4        & 19       & 8466   & 313    & 143  \\
+% median  & 838  & 2836 & 248      & 99           & 5504  & 9        & 59       & 37372  & 1792   & 650     \\
+% 75\%  & 2155    & 6828 & 459   & 261          & 18579 & 20       & 142   & 117699 & 4087   & 1907 \\
+% max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
+% Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
+% \end{tabular}
+% \vspace{-0.3cm}
+% \label{tab:dataset}
+% \end{table*}
+
+% The dataset contains security flaws for more than $13$ different languages,
+% being PHP ($39\%$) and C ($24\%$) the most prevalent ones (cf. Figure~\ref{fig:lang}).
+% To classify the applications in the dataset, we use a taxonomy for open-source
+% software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
+% presents the projects domain distribution: \textit{Application Software} ($25$),
+% software that provides end-users with functional systems; \textit{Web libraries
+% and frameworks} ($20$); \textit{System Software} ($18$), software that provides
+% services and infrastructures (e.g., operating systems, servers and databases);
+% \textit{Software Tool} ($13$), software that supports development (e.g.,
+% programming languages, compilers, package managers, IDEs); and, \textit{Non-web
+% libraries and frameworks} ($18$), for desktop and mobile software.
+% %
+% Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
+% open-source projects involved in this study, including the number of forks, the
+% number of stars, the number of watchers, the number of contributors, the number
+% of commits, the number of branches, the number of releases, size (cf. GitHub, it
+% is the size of the whole repository, including all of its history, in
+% kilobytes), the number of issues, and the number of pull requests.
 
@@
 Both datasets have the same size. For each
-security commit, one random commit is selected from the list of all commits of
-that particular project. We originate the regular commits from the security
-commits to ensure that differences in maintainability are not a consequence of
+security commit, a commit with an approximate size is searched (due to the complexity 
+of some patches, it was not possible to find patches with the same amount of added and 
+deleted sizes). Our search uses the patch diff to look for a new similar diff. The 
+algorithm follows the following steps:
+
+\begin{enumerate}
+\item Diff comparison between the patch commit and its parent.
+\item Selects a random commit from the security patch project.
+\item Diff comparison between the random commit and its parent.
+\item Check if the diff from the regular change has the same/approximated size to 
+the security patch.
+\end{enumerate}
+
+Our approximation was calculated using a step of 1. Every 10 attempts to search for 
+a change with the same diff the step increased. Thus, if a change suffered 20 deletions 
+and 10 additions. Then, in the first 10 attempts, the algorithm searches for a diff with 
+deletions between 19 and 21 and additions between 9 and 11. After the 10 first attempts, 
+the algorithm searches for a diff with deletions between 18 and 22 and additions between 
+8 and 10. We originate the regular commits from the security
+commits to ensuring that differences in maintainability are not a consequence of
 characteristics of different projects.
-%
+
 \subsection{Maintainability Analysis}
@@
 
+\Sof{wip - made the connection to Luis work, dropped two of the formulas. But at the
+same time I think we should present the first one. - conflicted}
+
 The BCH tool does not compute the final score that our study needs to compare
-maintainability amongst different project versions. Thus, our work proposes an
-equation to capture the distance between the current state of the project and
-the standard thresholds based on previous work~\cite{Olivari:2018}. The equation
-considers the following:
+maintainability amongst different project versions. For this, we
+follow previous work on measuring the impact of energy-oriented refactorings
+~\cite{cruz2019energyoriented}. Luis et al. ($2019$) proposed an equation
+to capture the distance between the current state of the project and
+the standard thresholds based on the insights provided in ~\cite{Olivari:2018}. 
+The equation considers the following:
 \begin{itemize}
@@
 
-Given the violations for the BCH guidelines, we compute the maintainability score
+Given the violations for the BCH guidelines, the maintainability score is computed
 $M(v)$ as follows:
@@
 severity level (medium, high, and very high) and $L$ is the group of severity
-levels of maintainability infractions. The compliance $C$ for a given severity
-level $l$ is derived by:
-
-\begin{equation}\label{eq:3}
-    C(l) = LOC_{compliant}(l) - w(l) * LOC_{\neg compliant}(l)
-\end{equation}
-
-\noindent
-where $LOC_{compliant}(l)$ are the lines of code that comply with the guideline
-at the given severity level $l$, $LOC_{\neg compliant}(l)$ are the lines of code
-that do not comply with the guideline at the given severity level $l$ and $w(l)$
-is the weight factor to heighten the impact of non-compliant lines in comparison to
-compliant lines. Finally, the term $w(l)$ is calculated as follows:
-
-\begin{equation}
-    w(l) = \frac{1 - \theta(l)}{\theta(l)}
-\end{equation}
-
-\noindent
-where $\theta(l)$ is the threshold in percentage of the lines of code that are
-accepted to be non-compliant with the guideline for the severity level $l$. This
-is a standard value defined by BCH. In other words, the factor $w$ is used in
-Equation~\ref{eq:3} to highlight the lines of code that are not complying with
-the guideline. Then, we compute the difference of maintainability between the
+levels of maintainability infractions. The compliance $C$ for a severity level
+is calculated based on the number of lines of code that comply and not comply
+with the guideline at a given severity level $l$ and $w(l)$ as explained in more
+detail in previous work~\cite{cruz2019energyoriented}.
+
+In our analysis, we compute the difference of maintainability between the
 security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
-Figure~\ref{fig:commit}.
+Figure~\ref{fig:commit}. Thus, we can determine which patches had a positive, negative
+or null impact on the project maintainability.
+
+% The compliance $C$ for a given severity
+% level $l$ is derived by:
+%
+% \begin{equation}\label{eq:3}
+%     C(l) = LOC_{compliant}(l) - w(l) * LOC_{\neg compliant}(l)
+% \end{equation}
+%
+% \noindent
+% where $LOC_{compliant}(l)$ are the lines of code that comply with the guideline
+% at the given severity level $l$, $LOC_{\neg compliant}(l)$ are the lines of code
+% that do not comply with the guideline at the given severity level $l$ and $w(l)$
+% is the weight factor to heighten the impact of non-compliant lines in comparison to
+% compliant lines. Finally, the term $w(l)$ is calculated as follows:
+%
+% \begin{equation}
+%     w(l) = \frac{1 - \theta(l)}{\theta(l)}
+% \end{equation}
+%
+% \noindent
+% where $\theta(l)$ is the threshold in percentage of the lines of code that are
+% accepted to be non-compliant with the guideline for the severity level $l$. This
+% is a standard value defined by BCH. In other words, the factor $w$ is used in
+% Equation~\ref{eq:3} to highlight the lines of code that are not complying with
+% the guideline. Then, we compute the difference of maintainability between the
+% security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
+% Figure~\ref{fig:commit}.
 

### 2020-01-27 paper/icpc20.tex
charts updated

@@
 
-\renewcommand{\shortauthors}{Reis et al.}
+\renewcommand{\shortauthors}{Anon. et al.}
 
@@
 \begin{itemize}
-    \item \textbf{CW-89: Improper Neutralization of Special Elements used in an SQL Command/
-	SQL Injection (23 cases).} When developers do not keep untrusted data
+    \item \textbf{CW-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
     separate from SQL queries. If an attacker sends a command that
@@
 
-    \item \textbf{CWE-20: Improper Input Validation (75 cases).} When developers
+    \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
 	forget to validate the inputs of an application, attackers may have control
@@
 
-    \item \textbf{CWE-200: Information Exposure (38 cases).}
-
-	Application's logs are one way of intentional/unintentional disclosure information
+    \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
 	to attackers. Many times attackers get access to logs when sould not be authorized to.
 
-	\item \textbf{CWE-352: Cross-Site Request Forgery (CSRF) (20 cases).} Poor session tokens generation
+	\item \textbf{CWE-352: Cross-Site Request Forgery (CSRF) (22 cases).} Poor session tokens generation
     and management usually allow attackers to send forged HTTP requests
@@
 
-	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (36 cases).}
-
-	Incorrectly
+	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
 	    implemented functionalities related to authentication and session
@@
 
-	\item \textbf{CWE-399: Resource Management Errors (20 cases).} 
-	Resource management issues are found more frequently
+	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
 		in programming languages that do not manage memory automatically (e.g., C/C++
@@
 
-	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)}
-	Lack of proper validation or escaping
+	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
 	    allows attackers to submit untrusted data to web browsers through malicious
@@
 
-	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (36 cases).} 
-	When developers forget to neutralize special elements within the pathnames to files/directories, attackers 
+	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
@@
 \begin{figure}[h]	
-	\vspace{-0.2cm}
-
- 	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability.pdf}
- 	\caption{Maintainability difference for security and baseline refactorings}
+ 	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability_sec_vs_reg.pdf}
+ 	\caption{Maintainability difference for security and baseline patches}
 	\label{fig:secvsreg}
-	\vspace{-0.2cm}
+
+\end{figure}
+
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.57\textwidth]{figures/impact_per_guideline.pdf}
+ 	\caption{Impact of a security patch in 8 out of the 10 guidelines.}
+	\label{fig:guidelines}
+
+\end{figure}
+
+\begin{figure}[h]
+ 	\centering
+ 	\includegraphics[width=0.5\textwidth]{figures/maintainability_severity.pdf}
+ 	\caption{Maintainability difference by the vulnerabilities severity.}
+	\label{fig:severity}
 
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/language.pdf}
+  \includegraphics[width=0.5\textwidth]{figures/maintainability_language.pdf}
   \caption{Maintainability difference by language after security refactorings.}
   \label{fig:lang_main}
-  \vspace{-0.2cm}
 
@@
 \begin{figure}[h]
-	\vspace{-0.2cm}
  	\centering
- 	\includegraphics[width=0.57\textwidth]{figures/category.pdf}
- 	\caption{Maintainability difference by pattern after security refactorings.}
+ 	\includegraphics[width=0.5\textwidth]{figures/maintainability_cwe.pdf}
+ 	\caption{Maintainability difference by pattern after security patches.}
 	\label{fig:pat}
-	\vspace{-0.2cm}
 

### 2020-01-27 paper/icpc20.tex
new pass in the methodology

@@
 	Extract relevant data (e.g., sha, sha of the parent commit, owner of the project, name of 
-	the repository) and code bases from the combined dataset containing $1325$ security
+	the repository) and code bases from the combined dataset containing $1335$ security
 	refactorings collected from open-source software available on
-	GitHub. We consider the previous version of the patch as a vulnerable version
-	of the software.
+	GitHub.
 	\item A baseline of regular commits was randomly collected from the list of
@@
 To evaluate the impact of security refactorings on the maintainability of
-open-source software, we use a combined dataset of security flaws which is 
-the outcome of mining/manually inspect $301$ GitHub projects. Reis and Abreu 
+open-source software, we use a combined dataset of security vulnerabilities which is 
+the outcome of mining/manually inspect $267$ GitHub projects. Reis and Abreu 
 ($2017$) mined open-source
 software aiming at the extraction of real examples --- created by real
-developers --- to test and assess the performance of static analysis tools since
+developers --- to test and assess the performance of static analysis tools~\cite{Reis:2017:IJSSE} since
 using hand-seeded test cases or mutations could lead to misleading assessments
 of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
-dataset with $711$ test cases for $16$ security patterns. Some of the patterns
+dataset of $711$ test cases for $16$ security patterns. Some of the patterns
 are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
 $2017$~\cite{oswap:2017}. Each test case of the
-dataset is a triplet: the commit before the refactoring, the commit responsible
-for the refactoring, and the snippets of code that differ from one version to
+dataset is a triplet: the commit before the patching, the commit responsible
+for the patching, and the snippets of code that differ from one version to
 another (typically, called \textit{diffs}) --- where one can easily review the
 code used to fix the security flaw. In this study, we focus on computing the
-maintainability of the commits before and after the security refactoring to
-evaluate if the impact was positive, negative or none.
+maintainability of the commits before and after the security patching to
+evaluate if its impact was positive, negative or none.
 
@@
 used their expertise to locate the commits in the repositories. This technique 
-yielded a database of $624$ patches. The original dataset contained $1282$ commits. 
-However, for some vulnerabilities, one commit was not enough to patch the vulnerability. 
-Thus, many of the entries were parts of the patch. For our purpose, we shrank the dataset 
-and only considered the first and last commit used to patch the vulnerability. The $1325$ 
-patches in the dataset were run against the BCH toolset to
+yielded a database of $624$ patches~\cite{10.1109/MSR.2019.00064}. The original dataset contains $1282$ commits. One patch can have multiple commits assigned.
+To fit the dataset in our methodology, we search for the first and last commits
+used to patch the vulnerability. We assumed that the last commit of the patch as the fix (\emph{sha}) and the parent of the first commit as the vulnerable version (\emph{sha-p}).
+
+The $1335$ patches in the dataset were run against the BCH toolset to
 calculate their maintainability reports. However, there are refactorings
@@
 language support and project size. The final dataset used in this paper comprises
-$1020$ security patches. $547$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE). Some of the most prevailing categories are the following (categories
+$1038$ security patches. $556$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE). The most prevailing categories are the following (categories
 with less than $20$ instances were merged in a bigger category named \textit{Miscellaneous}):
@@
 
-The $1325$ patches in the dataset were run against the BCH toolset to
-calculate their maintainability reports. However, there are refactorings
-discarded from our study due to limitations of BCH: in particular, lack of
-language support and project size. The final dataset used in this paper comprises
-$1020$ security patches. $547$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE). Some of the most prevailing categories are the following (categories
-with less than $20$ instances were merged in a bigger category named \textit{Miscellaneous}):
-
 \begin{itemize}
@@
 
-	\item \textbf{CWE-352: Cross-Site Request Forgery (CSRF) (22 cases).} Poor session tokens generation
+	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
     and management usually allow attackers to send forged HTTP requests
@@
 
-	\item \textbf{Miscellaneous (473 cases)} This pattern comprises
+	\item \textbf{Miscellaneous (674 cases)} This pattern comprises
 		several other security refactorings that do not have a CWE assigned and patterns that
@@
 Both datasets have the same size. For each
-security commit, a commit with an approximate size is searched (due to the complexity 
-of some patches, it was not possible to find patches with the same amount of added and 
-deleted sizes). Our search uses the patch diff to look for a new similar diff. The 
-algorithm follows the following steps:
+security commit, a commit with an approximate size is searched. Due to the complexity 
+of some patches, it was not possible to find patches with the exact number of added and 
+deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. The algorithm follows the following steps:
 
 \begin{enumerate}
-\item Diff comparison between the patch commit and its parent.
-\item Selects a random commit from the security patch project.
-\item Diff comparison between the random commit and its parent.
-\item Check if the diff from the regular change has the same/approximated size to 
+\item Diff comparison between the security patch commit (\emph{sha}) and its parent(\emph{sha-p}).
+\item Selects a random commit from the security patch project (\emph{sha-reg}).
+\item Diff comparison between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
+\item Check if the size from the regular change has the same/approximated size to 
 the security patch.
@@
 
-Our approximation was calculated using a step of 1. Every 10 attempts to search for 
-a change with the same diff the step increased. Thus, if a change suffered 20 deletions 
-and 10 additions. Then, in the first 10 attempts, the algorithm searches for a diff with 
-deletions between 19 and 21 and additions between 9 and 11. After the 10 first attempts, 
-the algorithm searches for a diff with deletions between 18 and 22 and additions between 
-8 and 10. We originate the regular commits from the security
-commits to ensuring that differences in maintainability are not a consequence of
+Our approximation was calculated using an initial step of $1$. Every $10$ attempts to search for 
+a change with the same \emph{diff} the step increased. Thus, if a change suffered $20$ deletions 
+and $10$ additions. Then, in the first $10$ attempts, the algorithm searches for a \emph{diff} with 
+deletions between $19$ and $21$ and additions between $9$ and $11$. After the $10$ first attempts, 
+the algorithm increase the step for $2$ and searches for a \emph{diff} with deletions between $18$ and $22$ and additions between 
+$8$ and $10$. We originate the regular commits from the security
+commits to ensure that differences in maintainability are not a consequence of
 characteristics of different projects.
@@
 risk}). If the project does not violate the thresholds, then it is compliant
-with the guideline. These thresholds are calibrated by BCH using their own
-data/experience --- using open-source and closed software systems. If a project is
+with the guideline. \textbf{These thresholds are calibrated by BCH using their own
+data/experience --- using open-source and closed software systems}. If a project is
 compliant with a guideline, it means that it is at least $65\%$ better than the
@@
 
+\begin{table}[H]
+	\caption{Guidelines to produce maintainable code}
+\begin{tabular}{L{1.66cm}L{3.25cm}L{2.65cm}}
+
+\toprule
+\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
+\midrule
+
+\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
+ units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of 
+ LOCs within each unit~\cite{criteria:2017} \\\midrule
+
+\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
+it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision 
+points~\cite{1702388,criteria:2017}\\\midrule
+
+\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
+multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant 
+LOCs~\cite{criteria:2017}\\\midrule
+
+\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
+$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} 
+\# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
+
+\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
+loosely coupled databases are easier to oversee and execute & \textbf{Module Coupling:} \# of
+incoming dependencies~\cite{criteria:2017} \\\midrule
+
+\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
+within modules that are exposed to modules in other components & \textbf{Component Independence:} 
+\% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
+
+\textbf{Keep Architecture Components Balanced} & Balancing the number of
+components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} 
+Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
+
+\textbf{Keep your code base Small} & Reduce and avoid the system size because
+small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted 
+to man-month/man-year~\cite{criteria:2017} \\\midrule
+
+\textbf{Automate Tests} & Test your code base because it makes development
+predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit 
+complexity, component independence and volume~\cite{Visser:2016:OREILLY}
+ \\\midrule
+
+\textbf{Write Clean Code} & Avoid producing software with code smells because
+it is more likely to be maintainable in the future & \textbf{Code Smells:} 
+\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
+identifier names) \\
+\bottomrule
+\end{tabular}
+\label{tab:guidelines}
+\end{table}
+
 Figure~\ref{fig:bchrep} shows an example of the report
@@
 Clean Code}.
-
-\begin{table}[h]
-	\caption{Guidelines to produce maintainable code}
-\begin{tabular}{L{1.66cm}L{3.25cm}L{2.65cm}}
-
-\toprule
-\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
-\midrule
-
-\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
- units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of 
- LOCs within each unit~\cite{criteria:2017} \\\midrule
-
-\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
-it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision 
-points~\cite{1702388,criteria:2017}\\\midrule
-
-\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
-multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant 
-LOCs~\cite{criteria:2017}\\\midrule
-
-\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
-$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} 
-\# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
-
-\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
-loosely coupled databases are easier to oversee and execute & \textbf{Module Coupling:} \# of
-incoming dependencies~\cite{criteria:2017} \\\midrule
-
-\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
-within modules that are exposed to modules in other components & \textbf{Component Independence:} 
-\% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
-
-\textbf{Keep Architecture Components Balanced} & Balancing the number of
-components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} 
-Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
-
-\textbf{Keep your code base Small} & Reduce and avoid the system size because
-small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted 
-to man-month/man-year~\cite{criteria:2017} \\\midrule
-
-\textbf{Automate Tests} & Test your code base because it makes development
-predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit 
-complexity, component independence and volume~\cite{Visser:2016:OREILLY}
- \\\midrule
-
-\textbf{Write Clean Code} & Avoid producing software with code smells because
-it is more likely to be maintainable in the future & \textbf{Code Smells:} 
-\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
-identifier names) \\
-\bottomrule
-\end{tabular}
-\vspace{-0.5cm}
-\label{tab:guidelines}
-\end{table}
 
@@
 most $5$}), i.e., the number of units are compliant with ISO
-25010~\cite{iso:2011}. Yellow, orange and red bars represent units that do not
+$25010$~\cite{iso:2011}. Yellow, orange and red bars represent units that do not
 comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
@@
 
-Aiming to analyze the impact of security refactorings, we use BCH to compute the
+Aiming to analyze the impact of security patches, we use BCH to compute the
 maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
@@
 	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
-	refactoring;
+	patch (\emph{sha-p});
 	\item $v_{s}$, the version free of the security flaw, i.e., after the
-	refactoring;
+	patch (\emph{sha});
 \end{itemize}
-
-\Sof{wip - made the connection to Luis work, dropped two of the formulas. But at the
-same time I think we should present the first one. - conflicted}
 
@@
 to capture the distance between the current state of the project and
-the standard thresholds based on the insights provided in ~\cite{Olivari:2018}. 
+the standard thresholds calculated by the BCH based on the insights provided in ~\cite{Olivari:2018}. 
 The equation considers the following:
@@
 	\item \textbf{Size of project changes does not affect the maintainability
-	difference between two versions, $\Delta M (v_{s-1},v_{s})$.} We
+	difference between two versions, $\Delta M (v_{s-1},v_{s}) = M(v_{s}) - M(v_{s-1})$.} We
 	aim at evaluating security patterns occurring in different projects similarly.
@@
 guidelines, while $M(v) > 0$ indicates that version $v$ is following
-the BCH guidelines.
-
-The maintenance for the guideline $g$, $M_g$ for a given version
-of a project is computed with the following equation:
-
-\begin{equation}
-    M_{g} = \frac{1}{|L|} \sum_{l \in L}^{} C(l) , L = \{medium, high, veryHigh\}
-\end{equation}
-
-\noindent
-where $C$ is the compliance with the maintainability guideline for the given
-severity level (medium, high, and very high) and $L$ is the group of severity
-levels of maintainability infractions. The compliance $C$ for a severity level
+the BCH guidelines. The maintenance for the guideline $g$, $M_g$ for a given version
+of a project is computed as the summation of the compliance with the 
+maintainability guideline for the given severity level (medium, high, and very high).
+The compliance for a severity level
 is calculated based on the number of lines of code that comply and not comply
-with the guideline at a given severity level $l$ and $w(l)$ as explained in more
+with the guideline at a given severity level as explained in more
 detail in previous work~\cite{cruz2019energyoriented}.
@@
 or null impact on the project maintainability. 
-
 % The compliance $C$ for a given severity
@@
 
-As mentioned before, this study evaluates $607$ security and baseline commits
-from $94$ distinct open-source projects, distributed amongst $5$ main
-types of software applications. The following section presents the
+As mentioned before, this study evaluates $1038$ security and baseline commits
+from $267$ distinct open-source projects. The following section presents the
 results obtained for each research question.

### 2020-01-27 paper/icpc20.tex
methodology + rq1

@@
 patches to improve security on the maintainability of open-source software. 
-Maintainability is measured based on the Better Code Hub’s model of 10 
-guidelines using a dataset including 1020 security commits. Results show 
+Maintainability is measured based on the Better Code Hub’s model of $10$ 
+guidelines using a dataset including $1037$ security commits. Results show 
 that fixing software vulnerabilities hinders the maintainability of 
@@
 authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
-separately for each pattern.\\
+separately for each pattern.
 
@@
 \begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix}]
- <ul>
   <li class='error'>
@@
   </li>
- </ul>
-
 \end{lstlisting}
@@
 \end{enumerate}
-\vspace{-0.3cm}
-
 %
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/methodology.pdf}
- 	\caption{Study Methodology\Sof{Needs to be updated. Add Vulas.}}
+ 	\centering 	\includegraphics[width=0.45\textwidth]{figures/methodology.pdf}
+ 	\caption{Study Methodology}
 	\label{fig:met}
-
 \end{figure}
-	\vspace{-0.2cm}
-
 %
@@
 language support and project size. The final dataset used in this paper comprises
-$1038$ security patches. $556$ of these patches are identified with Common Weakness
+$1037$ security patches. $556$ of these patches are identified with Common Weakness
 Enumeration IDs (CWE). The most prevailing categories are the following (categories
@@
 to capture the distance between the current state of the project and
-the standard thresholds calculated by the BCH based on the insights provided in ~\cite{Olivari:2018}. 
+the standard thresholds calculated by the BCH based on the insights provided in~\cite{Olivari:2018}. 
 The equation considers the following:
@@
 with the guideline at a given severity level as explained in more
-detail in previous work~\cite{cruz2019energyoriented}.
-
-In our analysis, we compute the difference of maintainability between the
-security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
-Figure~\ref{fig:commit}. Thus, we can determine which patches had a positive, negative
-or null impact on the project maintainability. 
+detail in previous work~\cite{cruz2019energyoriented}. In our analysis, we compute 
+the difference of maintainability between the security commit ($v_{s}$) and its parent 
+commit ($v_{s-1}$), as illustrated in Figure~\ref{fig:commit}. Thus, we can determine 
+which patches had a positive, negative or null impact on the project maintainability. 
 % The compliance $C$ for a given severity
@@
 
-As mentioned before, this study evaluates $1038$ security and baseline commits
+As mentioned before, this study evaluates $1037$ security and baseline commits
 from $267$ distinct open-source projects. The following section presents the
@@
 
-\begin{figure}[h]	
+\begin{figure}[H]	
  	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability_sec_vs_reg.pdf}
@@
 
-\begin{figure}[h]
+Regarding security commits results, we observe that maintainability decreases in
+$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
+($400$) of the cases after security refactorings are applied. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
+Since the p-value is below the significance level of $0.05$, we argue that
+\emph{security pacthes have a negative impact in the maintainability of
+open-source software}.
+
+\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
+of the cases after regular refactorings are applied. But in contrast
+to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
+and remains the same in \Sof{<todo>}
+%$0.5\%$ ($3$) 
+of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
+test is \Sof{<todo>}
+%$1.54$x$10^{-6}$
+. Since the p-value is below the significance level of $0.05$
+, we argue that \emph{regular refactorings have a
+positive impact in the maintainability of open-source software}.
+
+\begin{figure}[H]
  	\centering
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
  	\centering
@@
 
-Regarding security commits results, we observe that maintainability decreases in
-$40.20\%$ ($244$), remains the same in $30.5\%$ ($185$) and increases in $29.33\%$
-($178$) of the cases after security refactorings are applied. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $1.21$x$10^{-10}$.
-Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security refactorings have a negative impact in the maintainability of
-open-source software}.
-
-For regular commits, we observe that the maintainability decreases in $41.35\%$
-($251$) of the cases after regular refactorings are applied. But in contrast
-to security commits, it increases in $58.16\%$ ($353$) and remains the same in
-$0.5\%$ ($3$) of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
-test is $1.54$x$10^{-6}$. Since the p-value is below the significance level of
-$0.05$, we argue that \emph{regular refactorings have a
-positive impact in the maintainability of open-source software}.
-
-\begin{figure}[h]
-\vspace{-0.2cm}
+\begin{figure}[H]
   \centering
   \includegraphics[width=0.5\textwidth]{figures/maintainability_language.pdf}
-  \caption{Maintainability difference by language after security refactorings.}
+  \caption{Maintainability difference by language after security patches.}
   \label{fig:lang_main}  
-
 \end{figure}
@@
 of open-source software, Figure~\ref{fig:lang_main}. Although
-the dataset contains security refactorings for more than $13$ languages, we only
-present the results for \emph{Java}, \emph{Python}, \emph{C}, \emph{Ruby}, and \emph{PHP}
+the dataset contains security patches for more languages, we only
+present the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
+\emph{Objective-C/C++}, and \emph{Java}
 because of the Paired Wilcoxon signed-rank test requirement: input sample has to have more 
@@
 integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
-test by pattern indicate that there is statistical evidence to support that the
-maintainability of open-source software decreases for security refactorings in
-$3$ different languages: \emph{C} (p-value = $5.72$x$10^{-9}$), Ruby (p-value = $0.057$) and
-PHP (p-value = $3.54$x$10^{-6}$). Following the \textit{American Statistical Association} 
-recommendations~\cite{doi:10.1080/00031305.2016.1154108}, we consider the results for 
-Ruby as statistically significant because its p-value is on the borderline.
-The results also exhibit statistical significance for when security refactorings 
-are performed in \emph{Java} (p-value = $3.61$x$10^{-2}$). \emph{Python} and the 
-languages that integrate the \emph{Others} set contain more cases where the 
-maintainability increases. However, the results obtained for these patterns are 
-not statistically significant.
+test by language indicate that there is statistical evidence to support that the
+maintainability of open-source software decreases for security patches for \emph{C/C++} 
+(p-value = $2.5$x$10^{-5}$).
+% Following the \textit{American Statistical Association}
+% recommendations~\cite{doi:10.1080/00031305.2016.1154108}, we consider the results for
+% Ruby as statistically significant because its p-value is on the borderline.
+If we considered a significance level of $0.1$, we could also conclude that security 
+patches in \emph{PHP} (p-value = $0.072$) also have a negative impact
+on the maintainability of software.
+However, to the rest of the languages presented in Figure~\ref{fig:lang_main}, it is
+not possible to make conclusions with statistical significance. But it is important
+to notice that overall there is a very significant number of patches that
+hinder the maintainability of software (e.g., \emph{Python}, \emph{Ruby}, 
+\emph{Objective-C/C++}, \emph{Java}), even if the number of negative cases does not 
+surpass the number of positive or none.
 
@@
 
-\begin{acks}
-We thank the Software Improvement Group (SIG) team for all the support and
-help in validating the methodology.
-\end{acks}
+% \begin{acks}
+% This work was supported by national funds through FCT, Fundação para a Ciência 
+% e a Tecnologia, under project UIDB/50021/2020 and…
+% We thank the Software Improvement Group (SIG) team for all the support and
+% help in validating the methodology.
+% \end{acks}
 

### 2020-01-27 paper/icpc20.tex
results

@@
 
-	\item \textbf{Miscellaneous (674 cases)} This pattern comprises
+	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
 		several other security refactorings that do not have a CWE assigned and patterns that
@@
 \begin{enumerate}
-\item Diff comparison between the security patch commit (\emph{sha}) and its parent(\emph{sha-p}).
+\item Diff comparison between the security patch commit (\emph{sha}) and its parent (\emph{sha-p}).
 \item Selects a random commit from the security patch project (\emph{sha-reg}).
@@
 
-\begin{figure}[H]	
+\begin{figure}[h]	
  	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability_sec_vs_reg.pdf}
- 	\caption{Maintainability difference for security and baseline patches}
+ 	\caption{Maintainability difference for security and baseline patches.}
 	\label{fig:secvsreg}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
  	\centering
- 	\includegraphics[width=0.57\textwidth]{figures/impact_per_guideline.pdf}
- 	\caption{Impact of a security patch in 8 out of the 10 guidelines.}
+ 	\includegraphics[width=0.5\textwidth]{figures/impact_per_guideline.pdf}
+ 	\caption{Impact of the security patches for each guideline.}
 	\label{fig:guidelines}
-
+	\scalebox{0.5}{
+	\begin{tabular}{r@{: }l r@{: }l}
+	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
+	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns in Modules \\
+	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
+	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of Code
+	\end{tabular}
+	}
 \end{figure}
 
-\begin{figure}[H]
+Maintainability is being calculated as the mean of the guidelines computed
+by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
+guideline individually. For projects with large 
+codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
+viable due to limitations of BCH. Thus, we decided to not consider this guideline 
+in our analysis.
+Regarding the impact of security patches in each guideline, we observe
+that overall our patches have a significant positive impact in $5$ guidelines: 
+\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
+Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
+Components Loosely} (p-value = $2.43$x$10^{-07}$), \emph{Keep Unit Interfaces 
+Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
+$7.9$x$10^{-35}$). However, each of the previous guidelines also suffer a negative 
+impact from $10\%$ to $35\%$.
+\emph{Separate Concerns in Module} guideline also follows the same results but without any statistical significance since the p-value is higher than $0.05$.
+According with the results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and $38.3\%$/$397$ of negative cases) are the guidelines which suffered more impact from security patches.
+
+\begin{figure}[h]
  	\centering
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \centering

### 2020-01-27 paper/icpc20.tex
more writing on results + cleaning

@@
 Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
-$7.9$x$10^{-35}$). However, each of the previous guidelines also suffer a negative 
-impact from $10\%$ to $35\%$.
-\emph{Separate Concerns in Module} guideline also follows the same results but without any statistical significance since the p-value is higher than $0.05$.
-According with the results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and $38.3\%$/$397$ of negative cases) are the guidelines which suffered more impact from security patches.
+$7.9$x$10^{-35}$). However, each of the previous guidelines also suffer a 
+negative impact from $10\%$ to $35\%$. \emph{Separate Concerns in Module} 
+guideline also follows the same results but without any statistical 
+significance since the p-value is higher than $0.05$. According with the 
+results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 
+of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and 
+$38.3\%$/$397$ of negative cases) are the guidelines which suffered more 
+impact from security patches.
 
 \begin{figure}[h]
- 	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/maintainability_severity.pdf}
+ 	\centering 	\includegraphics[width=0.45\textwidth]{figures/maintainability_severity.pdf}
  	\caption{Maintainability difference by the vulnerabilities severity.}
@@
 \end{figure}
+
+For the vulnerabilities identified with CVE(s), we scraped NVD to retrieve 
+their severity levels. In total, we obtained severity scores for $574$ 
+vulnerabilities: $121$ \emph{HIGH}, $423$ \emph{MEDIUM} and $30$ \emph{LOW}. 
+Figure~\ref{fig:severity} presents the impact
+of security patches per severity level on the maintainability of open-source 
+software. Although with no statistical significance, we observe that 
+vulnerabilities with higher severity levels (\emph{HIGH} and \emph{MEDIUM}) have more negative impact than the \emph{LOW} severity level on the
+maintainability of software.
 

### 2020-01-27 paper/icpc20.tex
more polising on the research questions'

@@
 hackers exploit them. It turns out that fixing these issues is as simple as
-modifying the software code base. However, these refactorings may have a
+modifying the software code base. However, these patches may have a
 negative impact on software maintenance --- e.g., developers might follow the
 quickest solution to fix the problem and not necessarily the most elegant and
-performant one. Thus, we want to understand how security refactorings impact
+performant one. Thus, we want to understand how security patches impact
 the maintainability of software. 
@@
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact of security refactorings on the
+\textit{\textbf{RQ1} What is the impact of security patches on the
 maintainability of open-source software?}
@@
 
-Often security flaws require refactoring code to make software more secure.
-However, there is no evidence yet of how security refactorings impact the
-maintainability of open-source software. Our suspicion is that developers tend
-to introduce technical debt in their software when refactoring the code base to
+Often security flaws require patching code to make software more secure.
+However, \textbf{there is no evidence yet of how security patches impact the
+maintainability of open-source software}. Our suspicion is that developers tend
+to introduce technical debt in their software when patching the code base to
 address a security flaw because they tend to choose the easiest path to solve
-it. To address it, we compute the maintainability of $607$ commits using the
+it. To address it, we compute the maintainability of $1037$ commits using the
 \emph{Better Code Hub} tool. The same approach is applied to a randomly generated
 dataset of regular commits --- baseline --- that we use to understand how
-maintainability evolves when security refactorings are performed versus when
+maintainability evolves when security patches are performed versus when
 they are not.
@@
 \begin{framed}
-\textit{\textbf{RQ2} Which patterns of security refactorings are more likely to
+\textit{\textbf{RQ2} Which patterns of security patches are more likely to
 affect open-source software maintainability?}
@@
 % here
-There are security flaws that are more difficult to refactor than others. For
+There are security flaws that are more difficult to patch than others. For
 instance, implementing secure authentication is not as easy as fixing a
 cross-site scripting vulnerability since usually the last one can be fixed
-without adding new lines of code. A typical fix for the cross-site scripting
-vulnerability is presented in Listing~\ref{lst:fix}. The developer added the
-function \texttt{htmlentities} to \textit{escape} the data given by the variable
-\texttt{\$\_['file']}. Knowing which patterns are more likely to increase
+without adding new lines of code/complexity to the code. A typical fix for 
+the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to \textit{escape} the data given by the variable
+\texttt{\$\_['file']}. \textbf{Knowing which patterns are more likely to increase
 maintainability issues is one step forward to bring awareness to security
-engineers of what patterns need more attention. The taxonomy of security
-patterns used to answer this question is the same as the one provided by the
-authors of the dataset~\cite{Reis:2017:IJSSE}. Maintainability is measured
+engineers of what patterns need more attention}. The taxonomy of security
+patterns used to answer this question is the one provided by the Common Weakness Enumeration
+(CWE). Maintainability is measured
 separately for each pattern.
@@
 \begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix}]
-  <li class='error'>
-   <?php echo $l->t('Cloud not found');?>
    <p class='hint'>
@@
    </p>
-  </li>
 \end{lstlisting}

### 2020-01-28 paper/icpc20.tex
discussion draft'

@@
 the maintainability of software. 
-As an example, consider the refactoring of the Datagram Transport Layer Security
+%%%%% NEEDS REVIEW
+\Sof{After dropping Keep Your Codebase Small the example changed to a positive impact - the example needs to be changed} As an example, consider the refactoring of the Datagram Transport Layer Security
 (DTLS) protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
@@
 the vulnerability.
-
+%
 Although checking the maximum size of a handshake message seems an elementary
@@
 OpenSSL maintainability.
-
+%%%%%%% 
 In this paper, our concern is to study whether, while improving software
 security, developers are also negatively impacting the maintainability of their
-software applications. To answer the following two research questions, we use a
-dataset of security refactorings~\cite{Reis:2017:IJSSE} to measure the impact of
-the changes in the maintainability of open-source software.
-
+software applications. To answer the following two research questions, we use two datasets
+of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
+security patches on the maintainability of open-source software.
+%
 \begin{framed}
@@
 \vspace{-0.1cm}
-
+%
 Often security flaws require patching code to make software more secure.
@@
 they are not.
-
+%
 \begin{framed}
@@
 \vspace{-0.1cm}
-% here
+% 
 There are security flaws that are more difficult to patch than others. For
@@
 separately for each pattern.
-
+%
 \setcounter{lstannotation}{0}
@@
 \end{lstlisting}
-
+%
 \section{Methodology}\label{sec:methodology}
@@
 Figure~\ref{fig:met}:
-
+%
 \begin{enumerate}
@@
 the activities of developers addressing security-oriented patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064}.
+%
 	\item
@@
 	GitHub.
+%
 	\item A baseline of regular commits was randomly collected from the list of
@@
 	maintainability of open-source software.
+%
   \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} 
@@
   for both security and regular commits. The code bases that resemble to the commits
-  of our datasets are collected before the analysis performed by BCH.
-
+  of our datasets are collected before the analysis performed by BCH.s
 \end{enumerate}
@@
 evaluate if its impact was positive, negative or none.
-
+%
 Ponta et al. ($2019$) tracked the \url{pivotal.io} website for vulnerabilities 
@@
 used to patch the vulnerability. We assumed that the last commit of the patch as the fix (\emph{sha}) and the parent of the first commit as the vulnerable version (\emph{sha-p}).
-
+%
 The $1335$ patches in the dataset were run against the BCH toolset to
@@
 with less than $20$ instances were merged in a bigger category named \textit{Miscellaneous}):
-
-% \begin{figure}[h]
-%  	\centering 	\includegraphics[width=0.5\textwidth]{figures/language_dist.pdf}
-%  	\caption{Distribution of Security Refactorings per Language}
-% 	\label{fig:lang}
-% \end{figure}
-% \vspace{-0.2cm}
-%
-% \begin{figure}[h]
-%  	\centering 	\includegraphics[width=0.5\textwidth]{figures/type_dist.pdf}
-%  	\caption{Distribution of the Domain of Projects}
-% 	\label{fig:domain}
-% \end{figure}
-
+%
 \begin{itemize}
@@
     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
-
+%
     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).} 
@@
 	attackers have allowed to directly those files.
-
+%	
     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
@@
 	of the control/data flow of the program.
-
+%	
     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
 	to attackers. Many times attackers get access to logs when sould not be authorized to.
-
+%	
 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
@@
     application.
-
+%
 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
@@
 	    passwords, keys, and other sensitive data.
-
+%	
 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
@@
 	    handling it instead. It is one of the main causes of DoS attacks.
-
+%	
 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
@@
 	    sites.
-
+%	
 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
@@
 	or entire repositories.
-
+%
 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
@@
 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
-
 \end{itemize}
-
-% \begin{table*}[h]
-% 	\centering
-% 	\caption{Descriptive statistics of the projects in the dataset}
-% \begin{tabular}{@{}rrrrrrrrrrr@{}}
-% \toprule
-%       & Forks   & Stars   & Watchers & Contributors & Commits  & Branches & Releases & Size      & Issues & Pull Requests  \\ \midrule
-% mean  & 1763 & 5448 & 401   & 153       & 14834 & 45    & 129   & 122973 & 3768   & 1941 \\
-% std   & 2434 & 6215 & 486   & 123       & 22234 & 150   & 189   & 209732 & 5933   & 3603 \\
-% min   & 1       & 3       & 1        & 0            & 103      & 1        & 0        & 108       & 0         & 0       \\
-% 25\%  & 391  & 1581    & 117   & 49           & 1440  & 4        & 19       & 8466   & 313    & 143  \\
-% median  & 838  & 2836 & 248      & 99           & 5504  & 9        & 59       & 37372  & 1792   & 650     \\
-% 75\%  & 2155    & 6828 & 459   & 261          & 18579 & 20       & 142   & 117699 & 4087   & 1907 \\
-% max   & 16366   & 31841   & 3446     & 413          & 114378   & 1227     & 1114     & 995790    & 33970     & 19329   \\
-% Total & 165771  & 512182  & 37729    & 14413        & 1394412  & 4246     & 12168    & 11559485  & 354283    & 182511  \\ \bottomrule
-% \end{tabular}
-% \vspace{-0.3cm}
-% \label{tab:dataset}
-% \end{table*}
-
-% The dataset contains security flaws for more than $13$ different languages,
-% being PHP ($39\%$) and C ($24\%$) the most prevalent ones (cf. Figure~\ref{fig:lang}).
-% To classify the applications in the dataset, we use a taxonomy for open-source
-% software from previous work~\cite{7816479}. Figure~\ref{fig:domain}
-% presents the projects domain distribution: \textit{Application Software} ($25$),
-% software that provides end-users with functional systems; \textit{Web libraries
-% and frameworks} ($20$); \textit{System Software} ($18$), software that provides
-% services and infrastructures (e.g., operating systems, servers and databases);
-% \textit{Software Tool} ($13$), software that supports development (e.g.,
-% programming languages, compilers, package managers, IDEs); and, \textit{Non-web
-% libraries and frameworks} ($18$), for desktop and mobile software.
-% %
-% Table~\ref{tab:dataset} presents the descriptive statistics of the $94$
-% open-source projects involved in this study, including the number of forks, the
-% number of stars, the number of watchers, the number of contributors, the number
-% of commits, the number of branches, the number of releases, size (cf. GitHub, it
-% is the size of the whole repository, including all of its history, in
-% kilobytes), the number of issues, and the number of pull requests.
-
 %
@@
 %$1.54$x$10^{-6}$
-. Since the p-value is below the significance level of $0.05$
-, we argue that \emph{regular refactorings have a
+. Since the p-value is below the significance level of $0.05$, 
+we argue that \emph{regular refactorings have a
 positive impact in the maintainability of open-source software}.
@@
 
-The impact of each type of security refactoring on the maintainability of
+The impact of each type of security patches on the maintainability of
 open-source software is presented in Figure~\ref{fig:pat}. The results of the
 Paired Wilcoxon signed-rank per pattern exhibited statistical
-evidence supporting that the maintainability of open-source software decreases
-for $4$ different patterns: \emph{Broken Autentication} (p-value = $0.018$),
-\emph{Denial-of-Service} (p-value = $0.029$), \emph{Memory Leak} (p-value = $0.009$) and
-\emph{Miscellaneous} (p-value = $3.07$x$10^{-4}$). With statistical
-significance, results demonstrate evidence that maintainability is not
-negatively impacted when removing \emph{Components with Known Vulnerabilities}
-(p-value = $7.96$x$10^{-4}$) and fixing \emph{Cross-Site Scripting
-Vulnerabilities} (p-value = $0.001$). \emph{Injection} and \emph{Cross-Site
-Request Forgery} produced more cases where refactorings had a negative
-impact in the maintainability. However, the results obtained for these patterns
-are not statistically significant.
-
-Projects with \emph{Broken Authentication} flaws are the most affected ones by
-security refactorings: $55.3\%$ ($21$) of the refactorings had a negative impact
-in the maintainability while $35.0\%$ ($11$) yielded a positive impact, and only
-$13.8\%$ ($6$) did not suffer any change. The results also show that fixing
-\emph{Memory Leak}s affects the maintainability of $51.3\%$ ($41$) of the cases
-and leaves only $13.8\%$ ($11$) of the cases unaffected. In $35\%$ ($28$) of the
-cases, maintainability increased. Refactoring vulnerabilities that allow
-\emph{Denial-of-Service} attacks (e.g., CVE-$2014$-$3506$) have a negative
-impact of $47.5\%$ ($19$) in the maintainability of open-source software while
-$30.0\%$ ($12$) of the cases exhibit maintainability improvement and $22.5\%$
-($9$) show that maintainability does not suffer any change. The
+evidence supporting that the maintainability of open-source software decreases for
+only one category: \emph{Miscellaneous} (p-value = $0.033$). Although without 
+statistical significance, we also observe that the maintainability of open-source software decreases $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.
+
+Also with no statistical
+significance, results demonstrate evidence that maintainability is less
+negatively impacted when patching \emph{SQL Injection (CWE-89)}, \emph{Improper Restriction of XML External Entity Reference (CWE-611)}, \emph{Improper Input Validation (CWE-20)}, \emph{Cross-Site Request Forgery (CWE-352)}, \emph{Permissions, Privileges, and Access Controls (CWE-264)}.
+However,
+for all the previous patterns there is a percentage of security patches that hinder maintainability between $30\%$ and $50\%$ which is very significant.
+\emph{Cross-Site Scripting (CWE-79)} is the pattern that suffers less positive/negative impact from security patches.
+As the mean shows, the security
+patches of this type of vulnerability usually have no impact on the maintainability of open-source software.
+
+Software with \emph{Resource Management Errors (CWE-399)} flaws are the most affected ones by security patches: $54.5\%$ ($12$) of the patches had a negative impact
+in the maintainability while $36.4\%$ ($8$) yielded a positive impact, and only
+$9.1\%$ ($2$) did not suffer any change. The results also show that fixing
+\emph{Information Exposure (CWE-200)} vulnerabilities affects the maintainability of $51.2\%$ ($20$) of the cases
+and leaves only $2.6\%$ ($1$) of the cases unaffected. In $46.2\%$ ($18$) of the
+cases, maintainability increased. 
+Patching \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)} vulnerabilities has a negative
+impact of $48.6\%$ ($18$) in the maintainability of open-source software while
+$46.0\%$ ($17$) of the cases exhibit maintainability improvement and $5.4\%$
+($2$) show that maintainability does not suffer any change. The
 \emph{Miscellaneous} refactorings pattern integrates several types of security
-flaws. These type of refactorings harm maintainability in $43.1\%$ ($84$) of the
-cases. In $31.8\%$ ($62$), there is no impact and for the
-other $25.1\%$ ($49$) maintainability increases.
+flaws. These type of refactorings harm maintainability in $41.3\%$ ($673$) of the
+cases. In $22.7\%$ ($153$), there is no impact and for the
+other $36.0\%$ ($242$) maintainability increases.
 
 It is also important to notice that overall the maintainability of open-source
-software is not affected by \emph{Cross-Site Scripting} vulnerabilities and by
-\emph{Using Components with Known Vulnerabilities}. \emph{Cross-Site Scripting}
-does not suffer any decrease in the maintainability of $55.3\%$ ($63$) of cases
-and $21.1\%$ ($24$) have a positive effect on it. Only $23.7\%$ ($27$) of the
-cases hinder the software maintainability. From the $21$ cases evaluated for
-\emph{Using Componentes with Known Vulnerabilities}, only $4.8\%$ ($1$) have a
-positive effect in the maintainability of software while $14.3\%$ ($3$) harms
-maintainability software, and $81\%$ ($17$) have no effect in the
-maintainability.
+software is not affected by \emph{Cross-Site Scripting (CWE-79)} vulnerabilities. \emph{Cross-Site Scripting (CWE-79)}
+does not suffer any decrease in the maintainability of $38.7\%$ ($24$) of cases
+and $35.5\%$ ($22$) have a negative effect on it. Only $25.8\%$ ($16$) of the
+cases improve the software maintainability.
 
@@
  	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/maintainability_cwe.pdf}
+ 	\includegraphics[width=0.45\textwidth]{figures/maintainability_cwe.pdf}
  	\caption{Maintainability difference by pattern after security patches.}
 	\label{fig:pat}	
-
 \end{figure}
@@
 
-\textbf{\textit{Security refactorings harm the maintainability of open-source software.}}
-%
-We found that $40.2\%$ of the security refactoring have a negative impact on the
+\textbf{\textit{Security patches harm the maintainability of open-source software.}}
+%
+We found that $41.76\%$ of the security patches have a negative impact on the
 maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
@@
 code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
-Figure~\ref{fig:lang_main} shows that only \emph{C} has a negative impact of
-more than $50\%$ on software maintainability. We suspect that these values are
-the result of projects contributions policies. Nine of the top-10 of dataset projects  --- the
-equivalent to $31.8\%$ of the commits ---
+Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
+the result of projects contributions policies. Nine of the top-10 of dataset projects 
 with more contributors have restrict contribution policies regarding code
@@
 
-\textbf{\textit{Broken Authentication \& Session Management, Denial-of-Service
-attacks, and Memory Leaks, and Miscellaneous patterns are more likely to affect
-open-source software maintainability.}} Although there are a few refactorings
-that harm software maintainability in the \emph{Cross-Site Scripting} and
-\emph{Using Components with Known Vulnerabilities}, overall maintainability does
-not suffer any changes for these patterns. For the remaining patterns,
-\emph{Injection} and \emph{Cross-Site Request Forgery}, results are not
-statistically significant.
+\textbf{\textit{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource Management Errors (CWE-399), and Miscellaneous patterns are more likely to affect
+open-source software maintainability.}}\Sof{Posso responder isto sem significancia estatistica?} Although there are a few refactorings
+that harm software maintainability in the \emph{Cross-Site Scripting}, overall maintainability does
+not suffer any changes for these patterns.
 

### 2020-01-28 paper/icpc20.tex
fixes

@@
 Maintainability is measured based on the Better Code Hub’s model of $10$ 
-guidelines using a dataset including $1037$ security commits. Results show 
+guidelines using a dataset including $1037$ security commits. \Rui{Add numbers} Results show 
 that fixing software vulnerabilities hinders the maintainability of 
@@
 impact. This empirical study exhibits proof that changes 
-applied in the code bases to fix vulnerabilities affect code maintainability. 
+applied in the code bases to patch vulnerabilities affect code maintainability. 
 Especially, for refactorings that address \emph{Broken Authentication and Session Management} 
@@
 building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
-point to the \texttt{dtls1\_process\_out\_of\_seq\_message} unit
+point to the \sloppy\texttt{dtls1\_process\_out\_of\_seq\_message} unit
 (\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
@@
 \begin{itemize}
-    \item \textbf{CW-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
+    \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
     separate from SQL queries. If an attacker sends a command that
@@
 maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric 
-calculated by BCH. Some of theses guidelines are calculated using the metrics 
+calculated by BCH. Theses guidelines are calculated using the metrics 
 presented in~\cite{criteria:2017}. During each guideline evaluation, the tool

### 2020-01-28 paper/icpc20.tex
patterns table

@@
   for both security and regular commits. The code bases that resemble to the commits
-  of our datasets are collected before the analysis performed by BCH.s
+  of our datasets are collected before the analysis performed by BCH.
 \end{enumerate}
@@
 used their expertise to locate the commits in the repositories. This technique 
-yielded a database of $624$ patches~\cite{10.1109/MSR.2019.00064}. The original dataset contains $1282$ commits. One patch can have multiple commits assigned.
+yielded a database of $624$ patches~\cite{10.1109/MSR.2019.00064}. The original 
+dataset contains $1282$ commits. One patch can have multiple commits assigned.
 To fit the dataset in our methodology, we search for the first and last commits
-used to patch the vulnerability. We assumed that the last commit of the patch as the fix (\emph{sha}) and the parent of the first commit as the vulnerable version (\emph{sha-p}).
+used to patch the vulnerability. We assumed that the last commit of the patch as 
+the fix (\emph{sha}) and the parent of the first commit as the vulnerable version 
+(\emph{sha-p}).
 %
@@
 $1037$ security patches. $556$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE). The most prevailing categories are the following (categories
-with less than $20$ instances were merged in a bigger category named \textit{Miscellaneous}):
-%
-\begin{itemize}
-    \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
+Enumeration IDs (CWE). In Table~\ref{tab:patterns}, the most prevailing categories are presented. Patterns
+with less than $20$ instances were merged in a bigger pattern named \textit{Miscellaneous}.
+
+\begin{table}[h]
+\footnotesize
+	\caption{Guidelines to produce maintainable code}
+\begin{tabular}{L{1.1cm}L{2.1cm}L{4.3cm}}
+
+\toprule
+\textbf{Pattern} & \textbf{Name} & \textbf{Description}\\
+\midrule
+
+\textbf{CWE-89} & 
+ Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases) & 
+ When developers do not keep untrusted data
      separate from SQL queries. If an attacker sends a command that
-    exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
-%
-    \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).} 
+     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.\\\midrule
+
+\textbf{CWE-611} & 
+ Improper Restriction of XML External Entity Reference (24 cases) & 
 	XML documents are exchanged through the web containing entities with URIs
  	that resolve to local/external files. Thus, when XML parsers are not well configured,
-	attackers have allowed to directly those files.
-%	
-    \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
+ 	attackers have allowed to directly those files.\\\midrule
+
+ \textbf{CWE-20} & 
+ Improper Input Validation (76 cases) & 
+	When developers
  	forget to validate the inputs of an application, attackers may have control
-	of the control/data flow of the program.
-%	
-    \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
-	to attackers. Many times attackers get access to logs when sould not be authorized to.
-%	
-	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
+ 	of the control/data flow of the program.\\\midrule
+
+  \textbf{CWE-200} & 
+Information Exposure (39 cases)& 
+Application's logs are one way of intentional/unintentional disclosure information
+ 	to attackers. Many times attackers get access to logs when sould not be authorized to.\\\midrule
+
+   \textbf{CWE-352} & 
+Cross-Site Request Forgery/CSRF (22 cases)& 
+Poor session tokens generation
      and management usually allow attackers to send forged HTTP requests
      including authentication information from the victim to the vulnerable web
-    application.
-%
-	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
+     application.\\\midrule
+
+       \textbf{CWE-264} & 
+Permissions, Privileges, and Access Controls (59 cases)& 
+Incorrectly
  	    implemented functionalities related to authentication and session
  	    management, allowing an attacker to gain access to session tokens,
-	    passwords, keys, and other sensitive data.
-%	
-	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
+ 	    passwords, keys, and other sensitive data.\\\midrule
+
+ 	           \textbf{CWE-399} & 
+Resource Management Errors (22 cases)& 
+Resource management issues are found more frequently
  		in programming languages that do not manage memory automatically (e.g., C/C++
  	    and Objective-C), i.e., where developers are responsible for
-	    handling it instead. It is one of the main causes of DoS attacks.
-%	
-	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
+ 	    handling it instead. It is one of the main causes of Denial-of-Service attacks.\\\midrule
+
+ 	     	           \textbf{CWE-79} & 
+Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)& 
+Lack of proper validation or escaping
  	    allows attackers to submit untrusted data to web browsers through malicious
  	    scripts that can hijack the user sessions or redirect the user to malicious
-	    sites.
-%	
-	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
+ 	    sites.\\\midrule
+
+ 	     	           \textbf{CWE-22} & 
+Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases) & 
+When developers forget to neutralize special elements within the path names to files/directories, attackers
  	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
-	or entire repositories.
-%
-	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
+ 	or entire repositories.\\\midrule
+
+ 	 	     	           \textbf{MISC} & 
+Miscellaneous (673 cases) & 
+This pattern comprises
  		several other security refactorings that do not have a CWE assigned and patterns that
-		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
-\end{itemize}
+ 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
+\bottomrule
+\end{tabular}
+\label{tab:patterns}
+\end{table}
+
+%
+% \begin{itemize}
+%     \item \textbf{CWE-89: Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases).} When developers do not keep untrusted data
+%     separate from SQL queries. If an attacker sends a command that
+%     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.
+% %
+%     \item \textbf{CWE-611: Improper Restriction of XML External Entity Reference (24 cases).}
+% 	XML documents are exchanged through the web containing entities with URIs
+% 	that resolve to local/external files. Thus, when XML parsers are not well configured,
+% 	attackers have allowed to directly those files.
+% %
+%     \item \textbf{CWE-20: Improper Input Validation (76 cases).} When developers
+% 	forget to validate the inputs of an application, attackers may have control
+% 	of the control/data flow of the program.
+% %
+%     \item \textbf{CWE-200: Information Exposure (39 cases).} Application's logs are one way of intentional/unintentional disclosure information
+% 	to attackers. Many times attackers get access to logs when sould not be authorized to.
+% %
+% 	\item \textbf{CWE-352: Cross-Site Request Forgery/CSRF (22 cases).} Poor session tokens generation
+%     and management usually allow attackers to send forged HTTP requests
+%     including authentication information from the victim to the vulnerable web
+%     application.
+% %
+% 	\item \textbf{CWE-264: Permissions, Privileges, and Access Controls (59 cases).} Incorrectly
+% 	    implemented functionalities related to authentication and session
+% 	    management, allowing an attacker to gain access to session tokens,
+% 	    passwords, keys, and other sensitive data.
+% %
+% 	\item \textbf{CWE-399: Resource Management Errors (22 cases).} Resource management issues are found more frequently
+% 		in programming languages that do not manage memory automatically (e.g., C/C++
+% 	    and Objective-C), i.e., where developers are responsible for
+% 	    handling it instead. It is one of the main causes of DoS attacks.
+% %
+% 	\item \textbf{CWE-79: Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)} Lack of proper validation or escaping
+% 	    allows attackers to submit untrusted data to web browsers through malicious
+% 	    scripts that can hijack the user sessions or redirect the user to malicious
+% 	    sites.
+% %
+% 	\item \textbf{CWE-22: Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases).}  When developers forget to neutralize special elements within the pathnames to files/directories, attackers
+% 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
+% 	or entire repositories.
+% %
+% 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
+% 		several other security refactorings that do not have a CWE assigned and patterns that
+% 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
+% \end{itemize}
 %
@@
 
-\begin{table}[H]
+\begin{table}[h]
+\footnotesize
 	\caption{Guidelines to produce maintainable code}
-\begin{tabular}{L{1.66cm}L{3.25cm}L{2.65cm}}
+\begin{tabular}{L{1.66cm}L{3.3cm}L{2.7cm}}
 

### 2020-01-30 paper/icpc20.tex
new rq

@@
 %%
-\begin{CCSXML}
-<ccs2012>
- <concept>
-  <concept_id>10010520.10010553.10010562</concept_id>
-  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-  <concept_significance>500</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010575.10010755</concept_id>
-  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-  <concept_significance>300</concept_significance>
- </concept>
- <concept>
-  <concept_id>10010520.10010553.10010554</concept_id>
-  <concept_desc>Computer systems organization~Robotics</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>10003033.10003083.10003095</concept_id>
-  <concept_desc>Networks~Network reliability</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Computer systems organization~Embedded systems}
-\ccsdesc[300]{Computer systems organization~Redundancy}
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
+%\begin{CCSXML}
+%<ccs2012>
+% <concept>
+%  <concept_id>10010520.10010553.10010562</concept_id>
+%  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
+%  <concept_significance>500</concept_significance>
+% </concept>
+% <concept>
+%  <concept_id>10010520.10010575.10010755</concept_id>
+%  <concept_desc>Computer systems organization~Redundancy</concept_desc>
+%  <concept_significance>300</concept_significance>
+% </concept>
+% <concept>
+%  <concept_id>10010520.10010553.10010554</concept_id>
+%  <concept_desc>Computer systems organization~Robotics</concept_desc>
+%  <concept_significance>100</concept_significance>
+% </concept>
+% <concept>
+%  <concept_id>10003033.10003083.10003095</concept_id>
+%  <concept_desc>Networks~Network reliability</concept_desc>
+%  <concept_significance>100</concept_significance>
+% </concept>
+%</ccs2012>
+%\end{CCSXML}
+%
+%\ccsdesc[500]{Computer systems organization~Embedded systems}
+%\ccsdesc[300]{Computer systems organization~Redundancy}
+%\ccsdesc{Computer systems organization~Robotics}
+%\ccsdesc[100]{Networks~Network reliability}
 
@@
 	empirical evaluation, for reproducibility. Available online:
-  \url{https://figshare.com/s/3e17518222df6a11a8d6}.
+  \textit{<anonymized for double blind review>}.
 \end{itemize}
@@
 length which can lead to memory exhaustion that could end in a DoS attack.
+
 \medskip
@@
 it. To address it, we compute the maintainability of $1037$ commits using the
-\emph{Better Code Hub} tool. The same approach is applied to a randomly generated
-dataset of regular commits --- baseline --- that we use to understand how
+\emph{Better Code Hub} tool. 
+%
+\begin{framed}
+\textit{\textbf{RQ2} What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}
+\end{framed}
+\vspace{-0.1cm}
+We also compute the maintainability of $1037$ random regular commits using the
+\emph{Better Code Hub} tool --- baseline.  We use them to understand \textbf{how
 maintainability evolves when security patches are performed versus when
-they are not.
+they are not}.
 %
 \begin{framed}
-\textit{\textbf{RQ2} Which patterns of security patches are more likely to
+\textit{\textbf{RQ3} Which patterns of security patches are more likely to
 affect open-source software maintainability?}
@@
 -   if(isset($_['file'])) echo $_['file']
-    ?>
-   </p>
-   <p class='hint'>
-    <?php
 +   if(isset($_['file'])) echo htmlentities($_['file'])
@@
 \begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.46\textwidth]{figures/maintainability_sec_vs_reg.pdf}
+ 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
  	\caption{Maintainability difference for security and baseline patches.}
@@
 %
-\subsection{Construct}
-%
-The maintainability formula was inferred based on the BCH's reports. The high
+\\\textit{Construct:} The maintainability formula was inferred based on the BCH's reports. The high
 amount of different projects and backgrounds may require other
@@
 thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
-
-\subsection{Internal}
-
-The security refactorings dataset provided by the previous
+%
+\\\textit{Internal:} The security refactorings dataset provided by the previous
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
@@
 baseline.
-
-\subsection{External}
-
-We use a dataset that only comprises refactorings of open-source software.
+%
+\\\textit{External:} We use a dataset that only comprises refactorings of open-source software.
 However, the methodology requires to access data that is not publicly available.

### 2020-01-30 paper/icpc20.tex
changing example to CVE-2014-3488

@@
     language=C,
+    backgroundcolor=\color{white},   % choose the background color
+    basicstyle=\ttfamily\scriptsize,        % size of fonts used for the code
+    breaklines=true,                 % automatic line breaking only at whitespace
+    captionpos=b,                    % sets the caption-position to bottom
+    commentstyle=\color{mygray}\bfseries,    % comment style
+    escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
+    keywordstyle=\color{blue},       % keyword style
+    stringstyle=\color{mymauve},     % string literal style
+    frame=single,
+    numbers=left,
+    stepnumber=1,
+    xleftmargin=2em,
+    escapeinside={/*!}{!*/},
+    moredelim=**[l][\color{mygreen}]{+\ },
+    moredelim=*[l][\color{red}]{-\ }
+}
+
+\lstdefinestyle{JavaStyle} {
+    language=Java,
     backgroundcolor=\color{white},   % choose the background color
@@
 cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
-Software quality characteristics include, but are not limited to functional correctnes,
+Software quality characteristics include, but are not limited to functional correctness,
 reliability, usability, maintainability, and security. Security is an 
 important non-functional requirement during the development of software systems. 
-However, there is still a considerable amount of zero-day vulnerabilities being
+However, there is still a considerable amount of vulnerabilities being
 discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
@@
 privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
-flaw that allowed callers to see and hear others without them picking up the call.
+flaw that allowed callers to see and hear others without them picking up the call.%\cite{cacheOut}.
 %
@@
 %%%%% NEEDS REVIEW
-\Sof{After dropping Keep Your Codebase Small the example changed to a positive impact - the example needs to be changed} As an example, consider the refactoring of the Datagram Transport Layer Security
+As an example, consider the refactoring of the Datagram Transport Layer Security
 (DTLS) protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-\\CVE-$2014$-$3506$ vulnerability},label={lst:vuln}]
-+ static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
-+  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
-+  if(max_len < (unsigned long)s->max_cert_list)
-+     return s->max_cert_list;
-+  return max_len;
+\begin{lstlisting}[style={JavaStyle}, caption={Fix provided by Netty developers to the
+\\CVE-$2014$-$3488$ vulnerability},label={lst:vuln}]
+
+// [snip]
+   for (;;) {
++       final int outAppBufSize = engine.getSession().getApplicationBufferSize();
++       final ByteBuffer outAppBuf;
++       if (nioOutAppBuf.capacity() < outAppBufSize) {
++              outAppBuf = ByteBuffer.allocate(outAppBufSize);
++       } else {
++              outAppBuf = nioOutAppBuf;
 +       }
 
-  static int dtls1_reassemble_fragment(
-     SSL *s, struct hm_header_st* msg_hdr, int *ok){
-
+         try {
+-               result = engine.unwrap(nioInNetBuf, nioOutAppBuf);
+-               result = engine.unwrap(nioInNetBuf, outAppBuf);
 // [snip]
-
-- unsigned long frag_len = msg_hdr->frag_len, max_len;
-- if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
--    goto err;
-
-- if(DTLS1_HM_HEADER_LENGTH +
--    	SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
--    max_len = s->max_cert_list;
-- else
--    max_len = DTLS1_HM_HEADER_LENGTH +
--   		SSL3_RT_MAX_ENCRYPTED_LENGTH;
-
-+ unsigned long frag_len = msg_hdr->frag_len;
-
-- if((msg_hdr->frag_off+frag_len) > max_len)
-+ if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
-+ msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
-     goto err; /*!\annotation{lst:func5}!*/
-
-  memset(seq64be,0,sizeof(seq64be));
-  seq64be[6] = (unsigned char) (msg_hdr->seq>>8);
-  seq64be[7] = (unsigned char) msg_hdr->seq;
-  item = pqueue_find(s->d1->buffered_messages, seq64be);
-  if(item == NULL){
-     frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);
-
-// [snip]
-
- static int dtls1_process_out_of_seq_message(
-     SSL *s, struct hm_header_st* msg_hdr, int *ok){
-
-// [snip]
-
-+ if(frag_len > dtls1_max_handshake_message_len(s))
-+    goto err; /*!\annotation{lst:func4}!*/
-
-  frag = dtls1_hm_fragment_new(frag_len, 0);
+                 break;
+         } finally {
+-               nioOutAppBuf.flip();
++              outAppBuf.flip();
+
+                nettyInNetBuf.readerIndex(nettyInNetBufStartOffset + nioInNetBuf.position() - nioInNetBufStartOffset);
+
+-               if (nioOutAppBuf.hasRemaining()) {
++              if (outAppBuf.hasRemaining()) {
+                       if (nettyOutAppBuf == null) {
+                              ChannelBufferFactory factory = ctx.getChannel().getConfig().getBufferFactory();
+                              nettyOutAppBuf = factory.getBuffer(initialNettyOutAppBufCapacity);
+                       }
+-                    nettyOutAppBuf.writeBytes(nioOutAppBuf);
++                    nettyOutAppBuf.writeBytes(outAppBuf);
+                 }
+-              nioOutAppBuf.clear();
++              outAppBuf.clear();
+          }
+     }
 
 \end{lstlisting}
+
+%\medskip
+%\setcounter{lstannotation}{0}
+%\begin{lstlisting}[style={JavaStyle}, caption={Fix provided by OpenSSL developers to the
+%\\CVE-$2014$-$3506$ vulnerability},label={lst:vuln}]
+%+ static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
+%+  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
+%+  if(max_len < (unsigned long)s->max_cert_list)
+%+     return s->max_cert_list;
+%+  return max_len;
+%+ }
+%
+%  static int dtls1_reassemble_fragment(
+%     SSL *s, struct hm_header_st* msg_hdr, int *ok){
+%
+%// [snip]
+%
+%- unsigned long frag_len = msg_hdr->frag_len, max_len;
+%- if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
+%-    goto err;
+%
+%- if(DTLS1_HM_HEADER_LENGTH +
+%-    	SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
+%-    max_len = s->max_cert_list;
+%- else
+%-    max_len = DTLS1_HM_HEADER_LENGTH +
+%-   		SSL3_RT_MAX_ENCRYPTED_LENGTH;
+%
+%+ unsigned long frag_len = msg_hdr->frag_len;
+%
+%- if((msg_hdr->frag_off+frag_len) > max_len)
+%+ if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
+%+ msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
+%     goto err; /*!\annotation{lst:func5}!*/
+%
+%  memset(seq64be,0,sizeof(seq64be));
+%  seq64be[6] = (unsigned char) (msg_hdr->seq>>8);
+%  seq64be[7] = (unsigned char) msg_hdr->seq;
+%  item = pqueue_find(s->d1->buffered_messages, seq64be);
+%  if(item == NULL){
+%     frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);
+%
+%// [snip]
+%
+% static int dtls1_process_out_of_seq_message(
+%     SSL *s, struct hm_header_st* msg_hdr, int *ok){
+%
+%// [snip]
+%
+%+ if(frag_len > dtls1_max_handshake_message_len(s))
+%+    goto err; /*!\annotation{lst:func4}!*/
+%
+%  frag = dtls1_hm_fragment_new(frag_len, 0);
+%
+%\end{lstlisting}
 
@@
 
-This work presents an empirical study on the impact of $607$ security
-refactorings on the maintainability of $94$ open-source projects. We leveraged
-Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018}. Security refactorings 
+This work presents an empirical study on the impact of $1037$ security
+patches on the maintainability of $267$ open-source projects. We leveraged
+Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018}. Security patches 
 significantly affect the software maintainability of open-source software, i.e., 
-developers hinder maintainability when fixing vulnerabilities. We found this evidence in
-$40.2\%$ of the studied cases. Regarding regular commits, it seems that overall
-maintainability increases ($50.16\%$). However, there is still a very
-considerable amount of cases that harm maintainability ($41.35\%$).
+developers hinder maintainability when patching vulnerabilities. We found this evidence in
+$41.76\%$ of the studied cases. Regarding regular commits, it seems that overall
+maintainability increases (\Sof{Complete}). However, there is still a very
+considerable amount of cases that harm maintainability (\Sof{Complete}).
 

### 2020-01-31 paper/icpc20.tex
new version

@@
 Maintainability is measured based on the Better Code Hub’s model of $10$ 
-guidelines using a dataset including $1037$ security commits. \Rui{Add numbers} Results show 
+guidelines using a dataset including $1037$ security commits. Results show 
 that fixing software vulnerabilities hinders the maintainability of 
-open-source software. We conclude that changes to codebases 
+open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399) . We conclude that changes to code bases 
 to fix vulnerabilities need to be performed with extra care.
@@
 Despite these efforts, improving software security is not a trivial task and requires 
-implementing refactorings that might affect software maintainability. 
+implementing patches that might affect software maintainability. 
 Our suspiciousness is that some of the these patches may have a negative 
@@
 From a methodological 
-perspective, we leveraged a dataset of $607$ security refactorings collected from open-source 
-software. We calculate software maintainability before and after the refactoring to measure its 
-impact. This empirical study exhibits proof that changes 
+perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
+software. We calculate software maintainability before and after the patch to measure its 
+impact. This empirical study presents evidence that changes 
 applied in the code bases to patch vulnerabilities affect code maintainability. 
-Especially, for refactorings that address \emph{Broken Authentication and Session Management} 
+Especially, for patches that address \emph{Broken Authentication and Session Management} 
 issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks. This is important work
@@
 \\CVE-$2014$-$3488$ vulnerability},label={lst:vuln}]
-
 // [snip]
@@
 %%%%%%% 
-In this paper, our concern is to study whether, while improving software
+Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software
 security, developers are also negatively impacting the maintainability of their
-software applications. To answer the following two research questions, we use two datasets
+software applications.
+This is important because software maintainability is approximately $75\%$ of the cost related to a project.
+To answer the following three research questions, we use two datasets
 of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
@@
 \begin{framed}
-\textit{\textbf{RQ2} What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}
-\end{framed}
-\vspace{-0.1cm}
-We also compute the maintainability of $1037$ random regular commits using the
-\emph{Better Code Hub} tool --- baseline.  We use them to understand \textbf{how
-maintainability evolves when security patches are performed versus when
-they are not}.
-%
-\begin{framed}
-\textit{\textbf{RQ3} Which patterns of security patches are more likely to
+\textit{\textbf{RQ2} Which patterns of security patches are more likely to
 affect open-source software maintainability?}
@@
 \vspace{-0.1cm}
-% 
 There are security flaws that are more difficult to patch than others. For
@@
 \end{lstlisting}
+%
+\begin{framed}
+\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}
+\end{framed}
+\vspace{-0.1cm}
+%
+We also compute the maintainability of $1037$ random regular commits using the
+\emph{Better Code Hub} tool --- baseline.  We use them to understand \textbf{how
+maintainability evolves when security patches are performed versus when
+they are not}.
 %
@@
 
-As said before, the web-based source code analysis service \emph{Better Code
+In this research, we follow a very similar methodology to the one presented in previous work on the maintainability 
+of energy-oriented fixes~\cite{cruz2019energyoriented}. As said before, the web-based source code analysis service \emph{Better Code
 Hub} (BCH) is used to collect the maintainability reports of the refactorings of
@@
 \vspace{-0.1cm}
-
-The impact of security and baseline commits in the maintainability of
-open-source software is presented in Figure~\ref{fig:secvsreg}. For each type of
-change, three types of impact of refactorings in the software maintainability
-are presented: \emph{negative}, maintainability decreases (\emph{red});
-\emph{none}, maintainability remains the same (\emph{yellow}); and,
-\emph{positive}, maintainability increases (\emph{green}). Next to each
-type of change, it is shown the mean ($\overline{x}$) and median (M)
-of the maintainability difference and the p-value resulting from the
-Paired Wilcoxon signed-rank test.
-
-\begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
-	\label{fig:secvsreg}
-
-\end{figure}
-
-Regarding security commits results, we observe that maintainability decreases in
-$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
-($400$) of the cases after security refactorings are applied. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
-Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security pacthes have a negative impact in the maintainability of
-open-source software}.
-
-\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
-of the cases after regular refactorings are applied. But in contrast
-to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
-and remains the same in \Sof{<todo>}
-%$0.5\%$ ($3$) 
-of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
-test is \Sof{<todo>}
-%$1.54$x$10^{-6}$
-. Since the p-value is below the significance level of $0.05$, 
-we argue that \emph{regular refactorings have a
-positive impact in the maintainability of open-source software}.
 
@@
 only one category: \emph{Miscellaneous} (p-value = $0.033$). Although without 
-statistical significance, we also observe that the maintainability of open-source software decreases $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.
+statistical significance, we also observe that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.
 
@@
 
+\begin{framed}
+\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}
+\vspace{-0.1cm}
+\end{framed}
+\vspace{-0.1cm}
+
+The impact of security and baseline commits in the maintainability of
+open-source software is presented in Figure~\ref{fig:secvsreg}. For each type of
+change, three types of impact of refactorings in the software maintainability
+are presented: \emph{negative}, maintainability decreases (\emph{red});
+\emph{none}, maintainability remains the same (\emph{yellow}); and,
+\emph{positive}, maintainability increases (\emph{green}). Next to each
+type of change, it is shown the mean ($\overline{x}$) and median (M)
+of the maintainability difference and the p-value resulting from the
+Paired Wilcoxon signed-rank test.
+
+\begin{figure}[h]	
+ 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
+ 	\caption{Maintainability difference for security and baseline patches.}
+	\label{fig:secvsreg}
+\end{figure}
+
+Regarding security commits results, we observe that maintainability decreases in
+$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
+($400$) of the cases after security patches are applied. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
+Since the p-value is below the significance level of $0.05$, we argue that
+\emph{security patches have a negative impact in the maintainability of
+open-source software}.
+
+\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
+of the cases after regular refactorings are applied. But in contrast
+to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
+and remains the same in \Sof{<todo>}
+%$0.5\%$ ($3$) 
+of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
+test is \Sof{<todo>}
+%$1.54$x$10^{-6}$
+. The baseline is still being computed by BCH but we can already see
+that regular changes usually have no impact on the maintainability of
+software.
+
+%Since the p-value is below the significance level of $0.05$, 
+%we argue that \emph{regular refactorings have a
+%positive impact in the maintainability of open-source software}.
+
 \section{Discussion}\label{sec:discussion}
@@
 \begin{framed}
-\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability
+\textit{\textbf{RQ1} What is the impact of security patches on the maintainability
 of open-source software?}
@@
 %
-We found that $41.76\%$ of the security patches have a negative impact on the
-maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
-a new tradeoff when developers need to fix vulnerabilities on their projects
-because they might be affecting the software maintainability or even introducing
-new vulnerabilities. This study exhibits evidence that developers may have to
-reduce maintainability for the sake of security. We understand that developers
-should be able to produce secure code without affecting the maintainability of
-their projects. But there are still some concerns that need to be addressed:
-\begin{itemize}
-	\item \textbf{Documentation} of software must provide developers with the
-	best practices to implement secure and maintainable software. Unmaintainable
-	code is difficult to test, analyze and re-use.
-
-	\item\textbf{Programming Languages} should provide new design patterns to
-	easily implement security patterns without endangering maintainability.
-	Ultimately, new programming languages, both secure and maintainable by design,
-	may be designed to help developers be highly productive when writing and
-	maintaining  secure applications. One example is the need for designing new
-	authorization mechanisms since these are one of the most complex security
-	features to implement.
-
-	\item \textbf{Software Developers} lack the knowledge to create secure and
-	maintainable code. Mainly because academic curricula are not yet prepared
-	to properly educate them. Universities should do their part on making
-	this possible. Online services such as BCH have also a great potential to help
-	developers to be more aware of the maintainability issues introduced by their
-	changes. Thus, developers can put more work into improving their code maintainability
-	and avoid common issues such as code duplication.
-
-\end{itemize}
 
@@
 
-\textbf{\textit{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource Management Errors (CWE-399), and Miscellaneous patterns are more likely to affect
-open-source software maintainability.}}\Sof{Posso responder isto sem significancia estatistica?} Although there are a few refactorings
-that harm software maintainability in the \emph{Cross-Site Scripting}, overall maintainability does
-not suffer any changes for these patterns.
-
-The impact of a refactoring depends on its complexity, i.e., if the refactoring
+\textbf{\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
+open-source software maintainability.}}
+
+Overall, the results obtained for the maintainability per CWE are no statistical significance
+
+The impact of a patch depends on its complexity, i.e., if the patchs
 adds complexity to the code base it is probably affecting the software
-maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting}
-and \emph{Using Components with Known Vulnerabilities} refactorings endure more
+maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} (CWE-79)
+patches endure more
 cases without any impact in the maintainability. Typically, these types of
 vulnerabilities do not need extra lines to be fixed, as shown in
-Listing~\ref{lst:fix}. Whereas, \emph{Broken Authentication} and vulnerabilities
+Listing~\ref{lst:fix}. Whereas, \emph{Resource Management Errors} (CWE-399) and vulnerabilities
 such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
@@
 concluded that the flaw was due to the introduction of high cyclomatic complexity
-in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and
-\emph{Using Components with Known Vulnerabilities} should take part of the
+in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
 developers' preoccupations along with the other patterns that explicitly hinder
@@
 
+\begin{framed}
+\textit{\textbf{RQ3} What is the impact of security patches versus regular changes
+on the maintainability of open-source software?}
+\vspace{-0.1cm}
+\end{framed}
+\vspace{-0.1cm}
+
+We found that $41.76\%$ of the security patches have a negative impact on the
+maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
+a new trade-off when developers need to fix vulnerabilities on their projects
+because they might be affecting the software maintainability or even introducing
+new vulnerabilities. This study exhibits evidence that developers may have to
+reduce maintainability for the sake of security. We understand that developers
+should be able to produce secure code without affecting the maintainability of
+their projects. But there are still some concerns that need to be addressed:
+\begin{itemize}
+	\item \textbf{Documentation} of software must provide developers with the
+	best practices to implement secure and maintainable software. Unmaintainable
+	code is difficult to test, analyze and re-use.
+
+	\item\textbf{Programming Languages} should provide new design patterns to
+	easily implement security patterns without endangering maintainability.
+	Ultimately, new programming languages, both secure and maintainable by design,
+	may be designed to help developers be highly productive when writing and
+	maintaining  secure applications. One example is the need for designing new
+	authorization mechanisms since these are one of the most complex security
+	features to implement.
+
+	\item \textbf{Software Developers} lack the knowledge to create secure and
+	maintainable code. Mainly because academic curricula are not yet prepared
+	to properly educate them. Universities should do their part on making
+	this possible. Online services such as BCH have also a great potential to help
+	developers to be more aware of the maintainability issues introduced by their
+	changes. Thus, developers can put more work into improving their code maintainability
+	and avoid common issues such as code duplication.
+
+\end{itemize}
+
 \section{Threats to Validity}\label{sec:threats}
@@
 %
-\\\textit{Construct:} The maintainability formula was inferred based on the BCH's reports. The high
+\\\textit{\textbf{Construct:}} The maintainability formula was inferred based on the BCH's reports. The high
 amount of different projects and backgrounds may require other
@@
 %
-\\\textit{Internal:} The security refactorings dataset provided by the previous
+\\\textit{\textbf{Internal:}} The security patches dataset provided by the previous
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
 commits produced by project developers to classify the changes performed by them
-to fix security flaws. This approach discards security refactorings that were
+to fix security flaws. This approach discards security patches that were
 not explicit in commits messages. Baseline commits are retrieved randomly by selecting one commit from the same
-project of a given security refactoring. This approach softens the differences
+project of a given security patch. This approach softens the differences
 that may result from the characteristics of each project. However,
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $607$ regular commits are enough to
+Furthermore, this evaluation considers that $1037$ regular commits are enough to
 alleviate random irregularities in the maintainability differences of the
@@
 %
-\\\textit{External:} We use a dataset that only comprises refactorings of open-source software.
+\\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.
 However, the methodology requires to access data that is not publicly available.
@@
 languages may require different coding practices to address software safety. The 
-dataset comprises more commits in C and PHP, i.e., the dataset may not be representative 
+dataset comprises more commits in Java, i.e., the dataset may not be representative 
 of the population regarding programming languages. Our approach does not consider 
-security refactorings in any other languages but English.
+security patches in any other languages but English.
 
@@
 This study also shows evidence that there are security patterns that need more
-attention than others. \emph{Broken Autentication} issues,
-\emph{Denial-of-Service} attacks and \emph{Memory Leaks} endure maintainability
-in $55.3\%$, $51.3\%$, and $47.5\%$, respectively. Therefore, we draw attention
+attention than others. \emph{Path Traversal} (CWE-22),
+\emph{Information exposure} (CWE-200) and \emph{Resource Management Errors} (CWE-399) vulnerabilities endure maintainability
+in $48.6\%$, $51.2\%$, and $54.5\%$, respectively. Therefore, we draw attention
 for the need of better documentation explaining how to use best practices to
@@
 
-As future work, this study may be extended in many directions: analyze which
-guidelines are more likely to be affected by security-related refactorings;
+As future work, this study may be extended in many directions: 
 investigate what are the most frequent vulnerabilities in each programming
-language; investigate how refactorings impact maintainability for each
+language; investigate how patches impact maintainability for each
 vulnerability and programming language;

### 2020-01-31 paper/icpc20.tex
new version

@@
 viable due to limitations of BCH. Thus, we decided to not consider this guideline 
-in our analysis.
+in our analysis. For each type of
+guideline, three types of impact of patches in the software maintainability
+are presented: \emph{negative}, maintainability decreases (\emph{red});
+\emph{none}, maintainability remains the same (\emph{yellow}); and,
+\emph{positive}, maintainability increases (\emph{green}). Next to each
+type of change, it is shown the mean ($\overline{x}$) and median (M)
+of the maintainability difference and the p-value resulting from the
+Paired Wilcoxon signed-rank test.
 Regarding the impact of security patches in each guideline, we observe
@@
 ($2$) show that maintainability does not suffer any change. The
-\emph{Miscellaneous} refactorings pattern integrates several types of security
-flaws. These type of refactorings harm maintainability in $41.3\%$ ($673$) of the
+\emph{Miscellaneous} patches pattern integrates several types of security
+flaws. These type of patches harm maintainability in $41.3\%$ ($673$) of the
 cases. In $22.7\%$ ($153$), there is no impact and for the
 other $36.0\%$ ($242$) maintainability increases.
-
 It is also important to notice that overall the maintainability of open-source
@@
 The impact of security and baseline commits in the maintainability of
-open-source software is presented in Figure~\ref{fig:secvsreg}. For each type of
-change, three types of impact of refactorings in the software maintainability
-are presented: \emph{negative}, maintainability decreases (\emph{red});
-\emph{none}, maintainability remains the same (\emph{yellow}); and,
-\emph{positive}, maintainability increases (\emph{green}). Next to each
-type of change, it is shown the mean ($\overline{x}$) and median (M)
-of the maintainability difference and the p-value resulting from the
-Paired Wilcoxon signed-rank test.
+open-source software is presented in Figure~\ref{fig:secvsreg}.
 
@@
 %$1.54$x$10^{-6}$
-. The baseline is still being computed by BCH but we can already see
+. The baseline is still being computed by BCH but we can see already 
 that regular changes usually have no impact on the maintainability of
-software.
+software which seems to contradict research performed in the past. However,
+we inspected a considerable amount of cases where this happens and
+found that considering the size of the \emph{diff} to generate the random
+regular commit may limit the type of commit obtained. There are several
+cases which size is small due to the high amount of
 
@@
 %
-
 We would expect that the negative impact of programming languages on
@@
 Overall, the results obtained for the maintainability per CWE are no statistical significance 
-
 The impact of a patch depends on its complexity, i.e., if the patchs

### 2020-01-31 paper/icpc20.tex
new version

@@
 Information Exposure (CWE-200) and Resource Management Errors (CWE-399) . We conclude that changes to code bases 
-to fix vulnerabilities need to be performed with extra care.
+to fix vulnerabilities need to be performed with extra care and that security-related code evolution should be subject in computer curricula.
 
@@
 in this paper, we present an empirical study on the impact of patches of 
-vulnerabilities in software maintenance accross open-source software.
+vulnerabilities in software maintenance across open-source software.
 %
@@
 codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
-viable due to limitations of BCH. Thus, we decided to not consider this guideline 
+viable due to limitations of BCH.
+For this guideline, the tool sets a limit of 20 Person-years.
+However, a large amount of projects of our database surpass this limit.
+Thus, we decided to not consider this guideline
 in our analysis. For each type of
@@
 \emph{positive}, maintainability increases (\emph{green}). Next to each
-type of change, it is shown the mean ($\overline{x}$) and median (M)
+type of guideline, it is shown the mean ($\overline{x}$) and median (M)
 of the maintainability difference and the p-value resulting from the
@@
 Regarding the impact of security patches in each guideline, we observe
-that overall our patches have a significant positive impact in $5$ guidelines: 
-\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
+that overall patches have a significant positive impact in $5$ guidelines: 
+\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Arschitecture 
 Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
@@
 Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
-$7.9$x$10^{-35}$). However, each of the previous guidelines also suffer a 
-negative impact from $10\%$ to $35\%$. \emph{Separate Concerns in Module} 
+$7.9$x$10^{-35}$).
+However, for each of these guidelines we also observe a substantial number of cases where the maintainability suffers a negative impact $-$ from $10\%$ to $35\%$ $-$ which highlights the need of better practices/techniques/tools to produce software.
+\emph{Separate Concerns in Module}
 guideline also follows the same results but without any statistical 
@@
 % Ruby as statistically significant because its p-value is on the borderline.
-If we considered a significance level of $0.1$, we could also conclude that security 
+We could also conclude that security 
 patches in \emph{PHP} (p-value = $0.072$) also have a negative impact
-on the maintainability of software.
-However, to the rest of the languages presented in Figure~\ref{fig:lang_main}, it is
-not possible to make conclusions with statistical significance. But it is important
-to notice that overall there is a very significant number of patches that
-hinder the maintainability of software (e.g., \emph{Python}, \emph{Ruby}, 
-\emph{Objective-C/C++}, \emph{Java}), even if the number of negative cases does not 
-surpass the number of positive or none.
+on the maintainability of software, if we considered a significance level of $0.1$. 
+Another programming language which impacts negatively the maintainability of software
+is \emph{Ruby} (30 in 71 vulnerabilities, $42.3\%$).
+
+Overall, it is important to notice that there is a very significant number of patches that
+hinder the maintainability of software $-$ even if the number of negative cases does not 
+surpass the number of positive or none $-$ for  $3$ different languages: \emph{Python},
+\emph{Objective-C/C++} and \emph{Java}. For those languages, the negative impact is between $35\%$ and $50\%$ which highlights the need for better and secure programming languages. 
+Results should be read with care for the languages where statistical significance was not found. However, 
+data reports very interesting hints about the impact of programming languages in 
+security patches.
 
@@
 The impact of each type of security patches on the maintainability of
-open-source software is presented in Figure~\ref{fig:pat}. The results of the
-Paired Wilcoxon signed-rank per pattern exhibited statistical
-evidence supporting that the maintainability of open-source software decreases for
-only one category: \emph{Miscellaneous} (p-value = $0.033$). Although without 
-statistical significance, we also observe that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.
-
-Also with no statistical
-significance, results demonstrate evidence that maintainability is less
+open-source software is presented in Figure~\ref{fig:pat}.
+Results demonstrate evidence that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.
+
+We also observe that maintainability is less
 negatively impacted when patching \emph{SQL Injection (CWE-89)}, \emph{Improper Restriction of XML External Entity Reference (CWE-611)}, \emph{Improper Input Validation (CWE-20)}, \emph{Cross-Site Request Forgery (CWE-352)}, \emph{Permissions, Privileges, and Access Controls (CWE-264)}.
@@
 for all the previous patterns there is a percentage of security patches that hinder maintainability between $30\%$ and $50\%$ which is very significant.
-\emph{Cross-Site Scripting (CWE-79)} is the pattern that suffers less positive/negative impact from security patches.
-As the mean shows, the security
-patches of this type of vulnerability usually have no impact on the maintainability of open-source software.
+\emph{Cross-Site Scripting (CWE-79)} is the pattern that suffers less positive/negative impact from security patches, i.e., security
+patches of this type of vulnerability typically have no impact on the maintainability of open-source software.
 
@@
 
+The results of the
+Paired Wilcoxon signed-rank per pattern exhibited statistical
+evidence supporting that the maintainability of open-source software decreases for
+only one category: \emph{Miscellaneous} (p-value = $0.033$). However, no statistical
+significance was found to the other patterns. Thus, results should be read with care.
+
 \begin{figure}[h]
@@
 
-\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
-of the cases after regular refactorings are applied. But in contrast
-to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
-and remains the same in \Sof{<todo>}
-%$0.5\%$ ($3$) 
+For regular commits, we observe that the maintainability decreases in $21.04\%$ ($133$) 
+of the cases after regular changes are applied. But in contrast
+to security commits, it increases in $26.90\%$ ($170$) 
+and remains the same in $52.06\%$ ($329$) 
 of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
-test is \Sof{<todo>}
-%$1.54$x$10^{-6}$
-. The baseline is still being computed by BCH but we can see already 
-that regular changes usually have no impact on the maintainability of
-software which seems to contradict research performed in the past. However,
-we inspected a considerable amount of cases where this happens and
-found that considering the size of the \emph{diff} to generate the random
-regular commit may limit the type of commit obtained. There are several
-cases which size is small due to the high amount of
+test is $5.74x10^{-78}$. The baseline is still being computed by BCH but it is possible to observe 
+that regular changes typically have little impact on the maintainability which contradicts research performed in the past. However,
+we manually inspected a considerable amount of cases where this happens and
+found that identifying regular commits as similar in size as the security-related commit is limiting the type of regular commits (e.g, input refactorings, renaming variables or functions)\footnote{Example of a random regular commit generated with our methodology (Accessed on \today{}):
+\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad322448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}}.
 

### 2020-01-31 paper/icpc20.tex
new version

@@
 \textbf{\textit{Security patches harm the maintainability of open-source software.}}
-%
+
+The guideline analysis suggests that all guidelines are affected by security patches. Being the
+\emph{Write Short Units of Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns in Modules} the guidelines with higher negative impact on the maintainability ($>30\%$ of cases).
+\emph{Keep Architecture Components Balanced} is the guideline with lower negative impact on
+the maintainability (approx. $10\%$). This might imply that developers should give different
+levels of attention to the guidelines when patching vulnerabilities.
+%
+As expected, vulnerabilities with higher severity are more likely to impact negatively
+the maintainability of software. This suggests that high/medium severity vulnerabilities may
+need more attention than low severity when patching vulnerabilities.
+%
+
 We would expect that the negative impact of programming languages on

### 2020-01-31 paper/icpc20.tex
new version

@@
 \vspace{-0.1cm}
+\textbf{\textit{Patching security vulnerabilities has a larger negative and positive impact
+than when fixing regular commits.}}
 
 We found that $41.76\%$ of the security patches have a negative impact on the
-maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
-a new trade-off when developers need to fix vulnerabilities on their projects
+maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). While $21.04\%$ of regular commits have a negative impact on the maintainability. This raises
+a new trade-off when developers need to patch vulnerabilities on their projects
 because they might be affecting the software maintainability or even introducing
@@
 %
-\\\textit{\textbf{Internal:}} The security patches dataset provided by the previous
+\\\textit{\textbf{Internal:}} The security patches dataset provided by previous
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
@@
 to fix security flaws. This approach discards security patches that were
-not explicit in commits messages. Baseline commits are retrieved randomly by selecting one commit from the same
+not explicit in commits messages. Pontas et. al\cite{10.1109/MSR.2019.00064} dataset
+has patches that were manually mined in software repositories. Being a manual technique,
+the search might led to
+wrong associations between vulnerabilities and code snippets.
+Baseline commits are retrieved randomly by selecting one commit from the same
 project of a given security patch. This approach softens the differences
@@
 should be carefully monitored by programmers since there is a high correlation
-between maintainability aspects and proneness to changes and faults. The present
-work uses a model proposed by BCH and focuses solely on evaluating the impact of
+between maintainability aspects and proneness to changes and faults.
+In 2019, Luis et al.~\cite{cruz2019energyoriented} proposed a formula to calculate maintainability based on the BCH's guidelines and measured the impact of energy-oriented fixes on the maintainability of software.
+The present
+work uses the same model (BCH) and focuses solely on evaluating the impact of
 patching vulnerabilities on software maintainability.
@@
 $41.76\%$ of the studied cases. Regarding regular commits, it seems that overall
-maintainability increases (\Sof{Complete}). However, there is still a very
-considerable amount of cases that harm maintainability (\Sof{Complete}).
+maintainability stays the same ($52.06\%$). However, there is still a 
+considerable amount of cases that harm maintainability ($21.04\%$).
 

### 2020-01-31 paper/icpc20.tex
new version

@@
 This paper is structured as follows: section~\ref{sec:motivation} introduces an
-example of a security refactoring of a known vulnerability found in
-OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
+example of a security refactoring of a known vulnerability found in the
+PHP interpreter\footnote{\label{openssl}OpenSSL is a toolkit that
 contains open-source implementations of the SSL and TLS cryptographic
@@
 the maintainability of software.
+
 %%%%% NEEDS REVIEW
-As an example, consider the refactoring of the Datagram Transport Layer Security
-(DTLS) protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
-(DoS) attack by crafting DTLS handshake messages to trigger memory allocations
+As an example, consider the patch of the state-machine implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
+(DoS) attack by crafting TLS messages to trigger memory allocations
 corresponding to large length values. This vulnerability is listed at the Common
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[style={JavaStyle}, caption={Fix provided by Netty developers to the
-\\CVE-$2014$-$3488$ vulnerability},label={lst:vuln}]
+\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
+\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
+ static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
  // [snip]
-   for (;;) {
-+       final int outAppBufSize = engine.getSession().getApplicationBufferSize();
-+       final ByteBuffer outAppBuf;
-+       if (nioOutAppBuf.capacity() < outAppBufSize) {
-+              outAppBuf = ByteBuffer.allocate(outAppBufSize);
-+       } else {
-+              outAppBuf = nioOutAppBuf;
-+       }
-
-         try {
--               result = engine.unwrap(nioInNetBuf, nioOutAppBuf);
--               result = engine.unwrap(nioInNetBuf, outAppBuf);
++   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
++        && !BUF_MEM_grow_clean(s->init_buf, 
++                                (int)s->s3->tmp.message_size
++                                + SSL3_HM_HEADER_LENGTH)) {
++         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
++         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
++          return SUB_STATE_ERROR;
  // [snip]
-                 break;
-         } finally {
--               nioOutAppBuf.flip();
-+              outAppBuf.flip();
-
-                nettyInNetBuf.readerIndex(nettyInNetBufStartOffset + nioInNetBuf.position() - nioInNetBufStartOffset);
-
--               if (nioOutAppBuf.hasRemaining()) {
-+              if (outAppBuf.hasRemaining()) {
-                       if (nettyOutAppBuf == null) {
-                              ChannelBufferFactory factory = ctx.getChannel().getConfig().getBufferFactory();
-                              nettyOutAppBuf = factory.getBuffer(initialNettyOutAppBufCapacity);
  }
--                    nettyOutAppBuf.writeBytes(nioOutAppBuf);
-+                    nettyOutAppBuf.writeBytes(outAppBuf);
-                 }
--              nioOutAppBuf.clear();
-+              outAppBuf.clear();
-          }
-     }
-
 \end{lstlisting}
-
+%
 %\medskip
 %\setcounter{lstannotation}{0}
-%\begin{lstlisting}[style={JavaStyle}, caption={Fix provided by OpenSSL developers to the
-%\\CVE-$2014$-$3506$ vulnerability},label={lst:vuln}]
-%+ static unsigned long dtls1_max_handshake_message_len(const SSL *s){ /*!\annotation{lst:func1}!*/
-%+  unsigned long max_len = DTLS1_HM_HEADER_LENGTH + SSL3_RT_MAX_ENCRYPTED_LENGTH;
-%+  if(max_len < (unsigned long)s->max_cert_list)
-%+     return s->max_cert_list;
-%+  return max_len;
-%+ }
-%
-%  static int dtls1_reassemble_fragment(
-%     SSL *s, struct hm_header_st* msg_hdr, int *ok){
-%
-%// [snip]
-%
-%- unsigned long frag_len = msg_hdr->frag_len, max_len;
-%- if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len)
-%-    goto err;
-%
-%- if(DTLS1_HM_HEADER_LENGTH +
-%-    	SSL3_RT_MAX_ENCRYPTED_LENGTH < s->max_cert_list)
-%-    max_len = s->max_cert_list;
-%- else
-%-    max_len = DTLS1_HM_HEADER_LENGTH +
-%-   		SSL3_RT_MAX_ENCRYPTED_LENGTH;
-%
-%+ unsigned long frag_len = msg_hdr->frag_len;
-%
-%- if((msg_hdr->frag_off+frag_len) > max_len)
-%+ if((msg_hdr->frag_off+frag_len) > msg_hdr->msg_len ||
-%+ msg_hdr->msg_len > dtls1_max_handshake_message_len(s))
-%     goto err; /*!\annotation{lst:func5}!*/
-%
-%  memset(seq64be,0,sizeof(seq64be));
-%  seq64be[6] = (unsigned char) (msg_hdr->seq>>8);
-%  seq64be[7] = (unsigned char) msg_hdr->seq;
-%  item = pqueue_find(s->d1->buffered_messages, seq64be);
-%  if(item == NULL){
-%     frag = dtls1_hm_fragment_new(msg_hdr->msg_len, 1);
-%
-%// [snip]
-%
-% static int dtls1_process_out_of_seq_message(
-%     SSL *s, struct hm_header_st* msg_hdr, int *ok){
-%
-%// [snip]
-%
-%+ if(frag_len > dtls1_max_handshake_message_len(s))
-%+    goto err; /*!\annotation{lst:func4}!*/
-%
-%  frag = dtls1_hm_fragment_new(frag_len, 0);
-%
+%\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
+%\\CVE-2016-6307 vulnerability for file ssl/statem/statem_lib.c },label={lst:vuln}]
+% int tls_get_message_header(SSL *s, int *mt){
+% // [snip]
+%+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+%+        && !BUF_MEM_grow_clean(s->init_buf, 
+%+                                (int)s->s3->tmp.message_size
+%+                                + SSL3_HM_HEADER_LENGTH)) {
+%+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+%+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+%+          return SUB_STATE_ERROR;
+% // [snip]
+% }
 %\end{lstlisting}
 
-The code changes performed to fix the vulnerability are:
+%The code changes performed to fix the vulnerability are:
 

### 2020-01-31 paper/icpc20.tex
new version

@@
 %%%%% NEEDS REVIEW
-As an example, consider the patch of the state-machine implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
+As an example, consider the patch of the TLS state-machine protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
 (DoS) attack by crafting TLS messages to trigger memory allocations
 corresponding to large length values. This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-$2014$-$3506$\footnote{CVE-$2014$-$3506$
-details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2014-3506}
+Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$ \footnote{CVE-$2016$-$6307$
+details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
 evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
-\emph{ssl/d1\_both.c} file\footnote{CVE-$2014$-$3506$ fix available  at
-\url{https://github.com/openssl/openssl/commit/1250f12613b61758675848f6600ebd914ccd7636}
+\emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
+\url{https://github.com/openssl/openssl/commit/c1ef7c971d0bbf117c3c80f65b5875e2e7b024b1}
 (Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
-negotiation between the two parties. DTLS provides a mechanism for fragmenting
-a handshake message over a number of packets, each of which can be transmitted
-separately over the internet from one party to another. The problem in
-CVE-$2014$-$3506$ is that the size of the handshake message collected by the
-receiver is not verified against the maximum size allowed by DTLS. It is then
-allocated to a new fragment (cf. Listing~\ref{lst:vuln}, line~36). Thus, if an
-attacker has access to one of these packages, he can change the
-handshake message size and trigger unnecessary memory allocations of large
-length which can lead to memory exhaustion that could end in a DoS attack.
+negotiation between the two parties. The problem in
+CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
+memory for a message is allocated too early, prior to the excessive
+message length check which can lead to a Denial of Service through memory
+exhaustion if an attacker triggers up a certain message length to be allocated to service a
+connection. In this case, there is only a security impact if the application does not call  \texttt{SSL\_free()} in a timely manner in the
+event that the connection fails;
+or, the application is working in a constrained environment where there
+is very little free memory;
+or, the attacker initiates multiple connection attempts such that there
+are multiple connections in a state where memory has been allocated for
+the connection, \texttt{SSL\_free()} has not yet been called, and there is
+insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
+be transitory because as soon as the connection fails the memory is
+subsequently freed again in the \texttt{SSL\_free()} call. However there is an
+increased risk during this period of application crashes due to the lack
+of memory - which would then mean a more serious Denial of Service.
 
@@
 +                                (int)s->s3->tmp.message_size
-+                                + SSL3_HM_HEADER_LENGTH)) {
++                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
 +         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
@@
  }
+\end{lstlisting}
+
+\medskip
+\setcounter{lstannotation}{0}
+\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
+\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
+ int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
+ // [snip]
+  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
+        SSL3_HM_HEADER_LENGTH;
+- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
+-      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+-      goto err;
+   }
+ // [snip]
+
+- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func5}!*/  
+-     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+-     goto err;
+- }
+
 \end{lstlisting}
@@
 
-~\ref{lst:func1} \texttt{dtls1\_max\_handshake\_message\_len} function was
-inserted to return the maximum number of bytes allowed in a DTLS
-handshake message for \texttt{s}. The minimum is $16$KB but it might be greater if
-the maximum certificate list size (\texttt{s-$>$max\_cert\_list}) requires it. More
-specifically, it checks if the maximum certificate list size requires more length 
-than the maximum length (\texttt{max\_len}). If that condition holds, the function 
-returns the size of the certificate list, otherwise, it returns the maximum length. 
-This function replaces lines $17$--$21$.
-
-~\ref{lst:func5} and~\ref{lst:func4} are guards added to different functions to
-ascertain that the message and the fragment length are never higher than the
-maximum allowed by DTLS. In particular, \ref{lst:func5} is responsible for patching
-the vulnerability.
-%
-Although checking the maximum size of a handshake message seems an elementary
+~\ref{lst:func1} \texttt{read\_state\_machine} function abstracts a sub-state-machine
+to handle the messages exchanged between peers. When messages are larger than
+what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
+of this resource was being verified very early in the application's flow.  Listing ~\ref{lst:vuln2} presents those verifications (lines $5$-$7$ and line $11$-$13$).
+These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}.
+
+%
+Although checking the maximum size of a message seems an elementary
 problem to repair, the example shows that there was a significant amount of
 changes performed in the code base which produced a negative impact in the
-maintainability. Even though the developer tries to simplify the
-code base by introducing a smaller unit
-\texttt{dtls1\_max\_handshake\_message\_len} (\ref{lst:func1}), this change disrupts
+maintainability. The developer tries to patch the
+code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
+and disrupts
 one of the guidelines proposed by the Software Improvement Group (SIG) for
-building maintainable software~\cite{Visser:2016:OREILLY}. It adds a new branch
-point to the \sloppy\texttt{dtls1\_process\_out\_of\_seq\_message} unit
-(\ref{lst:func5}) increasing the cyclomatic complexity and breaking the
-\emph{Write Simple Units of Code} guideline which resulted in a negative impact on the
+building maintainable software~\cite{Visser:2016:OREILLY} $-$
+the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
 OpenSSL maintainability.

### 2020-01-31 paper/icpc20.tex
new version

@@
 \end{lstlisting}
-
+ \vspace{-0.5cm}
 \medskip
@@
 %
-\begin{framed}
-\textit{\textbf{RQ1} What is the impact of security patches on the
-maintainability of open-source software?}
-\end{framed}
-\vspace{-0.1cm}
-%
+
+\subsubsection*{\textit{\textbf{RQ1} What is the impact of security patches on the
+maintainability of open-source software?}}
+
 Often security flaws require patching code to make software more secure.
@@
 %
-\begin{framed}
-\textit{\textbf{RQ2} Which patterns of security patches are more likely to
-affect open-source software maintainability?}
-\end{framed}
-\vspace{-0.1cm}
+\subsubsection*{\textit{\textbf{RQ2}  Which patterns of security patches are more likely to
+affect open-source software maintainability?}}
+
 There are security flaws that are more difficult to patch than others. For
@@
 \end{lstlisting}
-%
-\begin{framed}
-\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}
-\end{framed}
-\vspace{-0.1cm}
+\vspace{-0.3cm}
+%
+\subsubsection*{\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}}
 %
@@
 	\label{fig:met}
+	\vspace{-0.5cm}
 \end{figure}
@@
 \label{tab:patterns}
+\vspace{-0.5cm}
 \end{table}
@@
 	\label{fig:bchrep}
+	\vspace{-0.8cm}
 \end{figure}
@@
 \label{tab:guidelines}
+\vspace{-0.3cm}
 \end{table}
@@
 	\label{fig:commit}
+	\vspace{-0.5cm}
 \end{figure}
@@
 
-\begin{framed}
-\textit{\textbf{RQ1} What is the impact of security patches on the
-maintainability of open-source software?}
-\vspace{-0.1cm}
-\end{framed}
-\vspace{-0.1cm}
+\subsubsection*{\textit{\textbf{RQ1} What is the impact of security patches on the
+maintainability of open-source software?}}
 
@@
 	}
+	\vspace{-0.3cm}
 \end{figure}
@@
 	\label{fig:severity}
-
+	\vspace{-0.4cm}
 \end{figure}
@@
   \label{fig:lang_main}  
+  \vspace{-0.5cm}
 \end{figure}
@@
 
-\begin{framed}
-\textit{\textbf{RQ2} Which patterns of security patches are more likely to
-affect open-source software maintainability?}
-\vspace{-0.1cm}
-\end{framed}
-\vspace{-0.1cm}
+\subsubsection*{\textit{\textbf{RQ2} Which patterns of security patches are more likely to
+affect open-source software maintainability?}}
 
@@
 	\label{fig:pat}	
+	\vspace{-0.5cm}
 \end{figure}
 
-\begin{framed}
-\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}
-\vspace{-0.1cm}
-\end{framed}
-\vspace{-0.1cm}
+\subsubsection*{\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}}
 
@@
 	\label{fig:secvsreg}
+	\vspace{-0.5cm}
 \end{figure}
@@
 
-\begin{framed}
-\textit{\textbf{RQ1} What is the impact of security patches on the maintainability
-of open-source software?}
-\vspace{-0.1cm}
-\end{framed}
-\vspace{-0.1cm}
-
-\textbf{\textit{Security patches harm the maintainability of open-source software.}}
+\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the maintainability
+of open-source software?}}}
+
+\textit{Security patches harm the maintainability of open-source software.}
 
@@
 
-\begin{framed}
-\textit{\textbf{RQ2} Which patterns of security patches are more likely to
-affect open-source software maintainability?}
-\vspace{-0.1cm}
-\end{framed}
-\vspace{-0.1cm}
-
-\textbf{\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
-open-source software maintainability.}}
+\subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
+affect open-source software maintainability?}}}
+
+\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
+open-source software maintainability.}
 
@@
 
-\begin{framed}
-\textit{\textbf{RQ3} What is the impact of security patches versus regular changes
-on the maintainability of open-source software?}
-\vspace{-0.1cm}
-\end{framed}
-\vspace{-0.1cm}
-\textbf{\textit{Patching security vulnerabilities has a larger negative and positive impact
-than when fixing regular commits.}}
+\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes
+on the maintainability of open-source software?}}}
+\textit{Patching security vulnerabilities has a larger negative and positive impact
+than when fixing regular commits.}
 

### 2020-01-31 paper/icpc20.tex
new version

@@
 
-\subsubsection*{\textit{\textbf{RQ1} What is the impact of security patches on the
-maintainability of open-source software?}}
+\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the
+maintainability of open-source software?}}}
 
@@
 %
-\subsubsection*{\textit{\textbf{RQ2}  Which patterns of security patches are more likely to
-affect open-source software maintainability?}}
+\subsubsection*{\textit{\textbf{RQ2}  \textbf{Which patterns of security patches are more likely to
+affect open-source software maintainability?}}}
 
@@
 %
-\subsubsection*{\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}}
+\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}}}
 %
@@
 
-\subsubsection*{\textit{\textbf{RQ2} Which patterns of security patches are more likely to
-affect open-source software maintainability?}}
+\subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
+affect open-source software maintainability?}}}
 
@@
 
-\subsubsection*{\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}}
+\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}}}
 

### 2020-01-31 paper/icpc20.tex
new version

@@
 risk}). If the project does not violate the thresholds, then it is compliant
-with the guideline. \textbf{These thresholds are calibrated by BCH using their own
-data/experience --- using open-source and closed software systems}. If a project is
+with the guideline. These thresholds are calibrated by BCH using their own
+data/experience --- using open-source and closed software systems. If a project is
 compliant with a guideline, it means that it is at least $65\%$ better than the
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
+ 	\centering 	\includegraphics[width=0.44\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
@@
 	\label{fig:bchrep}
-	\vspace{-0.8cm}
+	\vspace{-0.6cm}
 \end{figure}
@@
 \label{tab:guidelines}
-\vspace{-0.3cm}
+\vspace{-0.2cm}
 \end{table}
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.75\linewidth]{figures/commit.pdf}
+ 	\centering 	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
  	\caption{Maintainability difference for security commits}
@@
 Paired Wilcoxon signed-rank test.
-Regarding the impact of security patches in each guideline, we observe
+Regarding security commits results, we observe that maintainability decreases in
+$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
+($400$) of the cases after security patches are applied. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
+Since the p-value is below the significance level of $0.05$, we argue that
+\emph{security patches have a negative impact in the maintainability of
+open-source software}.
+Regarding the impact of security patches for each guideline, we observe
 that overall patches have a significant positive impact in $5$ guidelines: 
@@
  	\centering
- 	\includegraphics[width=0.45\textwidth]{figures/maintainability_cwe.pdf}
+ 	\includegraphics[width=0.43\textwidth]{figures/maintainability_cwe.pdf}
  	\caption{Maintainability difference by pattern after security patches.}
@@
 
-Regarding security commits results, we observe that maintainability decreases in
-$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
-($400$) of the cases after security patches are applied. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
-Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security patches have a negative impact in the maintainability of
-open-source software}.
-
+We have seen previously a deterioration in maintainability when patching security vulnerabilities. 
 For regular commits, we observe that the maintainability decreases in $21.04\%$ ($133$) 
@@
 of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
-test is $5.74x10^{-78}$. The baseline is still being computed by BCH but it is possible to observe 
+test is $5.74x10^{-78}$. The baseline is still being computed at the time of submission (it is computationally very expensive, and will be concluded soon) but it is possible to observe 
 that regular changes typically have little impact on the maintainability which contradicts research performed in the past. However,
@@
 
-As future work, this study may be extended in many directions: 
+As future work, this study may be extended in several directions: 
 investigate what are the most frequent vulnerabilities in each programming

### 2020-01-31 paper/icpc20.tex
new version

@@
 
-As mentioned before, this study evaluates $1037$ security and baseline commits
+This study evaluates $1037$ security and baseline commits
 from $267$ distinct open-source projects. The following section presents the
@@
 Paired Wilcoxon signed-rank test.
+
 Regarding security commits results, we observe that maintainability decreases in
@@
 that overall patches have a significant positive impact in $5$ guidelines: 
-\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Arschitecture 
+\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
 Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
@@
 guideline also follows the same results but without any statistical 
-significance since the p-value is higher than $0.05$. According with the 
+significance since the p-value is higher than $0.05$. According to the 
 results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 
 of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and 
-$38.3\%$/$397$ of negative cases) are the guidelines which suffered more 
-impact from security patches.
+$38.3\%$/$397$ of negative cases) are the most affected affected guidelines.
 
@@
 
-For the vulnerabilities identified with CVE(s), we scraped NVD to retrieve 
+For the vulnerabilities identified with CVE(s), we scraped the National Vulnerability Database (NVD) Website to retrieve 
 their severity levels. In total, we obtained severity scores for $574$ 
@@
 We have also evaluated the impact on the maintainability per programming language
-of open-source software, Figure~\ref{fig:lang_main}. Although
-the dataset contains security patches for more languages, we only
-present the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
-\emph{Objective-C/C++}, and \emph{Java}
-because of the Paired Wilcoxon signed-rank test requirement: input sample has to have more 
-than $20$ samples. The other languages were
+of open-source software, Figure~\ref{fig:lang_main}.
+We restrict this analysis to programming languages with at least $20$ data points, as this is a requirement for our hypothesis tests.
+Thus, we compare the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP},
+\emph{Objective-C/C++}, and \emph{Java}. The other languages were
 integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
@@
 % Ruby as statistically significant because its p-value is on the borderline.
-We could also conclude that security 
-patches in \emph{PHP} (p-value = $0.072$) also have a negative impact
-on the maintainability of software, if we considered a significance level of $0.1$. 
-Another programming language which impacts negatively the maintainability of software
-is \emph{Ruby} (30 in 71 vulnerabilities, $42.3\%$).
+\emph{Ruby} and \emph{PHP} languages also impact negatively the maintainability of software.
+Despite not having statistical significance, there are a considerable number of \emph{Ruby} and \emph{PHP} patches hindering maintainability that should not be disregarded
 
@@
 software is not affected by \emph{Cross-Site Scripting (CWE-79)} vulnerabilities. \emph{Cross-Site Scripting (CWE-79)}
-does not suffer any decrease in the maintainability of $38.7\%$ ($24$) of cases
+does not suffer any decrease in the maintainability of $38.7\%$ ($24$) of the cases
 and $35.5\%$ ($22$) have a negative effect on it. Only $25.8\%$ ($16$) of the
@@
  	\centering
- 	\includegraphics[width=0.43\textwidth]{figures/maintainability_cwe.pdf}
+ 	\includegraphics[width=0.4\textwidth]{figures/maintainability_cwe.pdf}
  	\caption{Maintainability difference by pattern after security patches.}

### 2020-02-06 paper/icpc20.tex

@@
 open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
-Information Exposure (CWE-200) and Resource Management Errors (CWE-399) . We conclude that changes to code bases 
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399). We conclude that changes to code bases 
 to fix vulnerabilities need to be performed with extra care and that security-related code evolution should be subject in computer curricula.
@@
 These vulnerabilities are gateways for attackers to penetrate systems and 
-exploit their resources. As an example, in the last week of January $2019$, 
+exploit their resources. \Sof{Improve example - outdated}As an example, in the last week of January $2019$, 
 a severe vulnerability in Apple FaceTime was revealed\footnote{FaceTime vulnerability 
@@
 based in ISO/IEC 25010~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
-were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041}. 
+were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
 There are other well-known standards and models that have been proposed
@@
 applied in the code bases to patch vulnerabilities affect code maintainability. 
-Especially, for patches that address \emph{Broken Authentication and Session Management} 
-issues, \emph{Memory Leaks} and \emph{Denial-of-Service} attacks. This is important work
+Especially, for patches that address Path Traversal (CWE-22),
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399). This is important work
 because there is still little information about the impact of security patches on the software
@@
 example of a security refactoring of a known vulnerability found in the
-PHP interpreter\footnote{\label{openssl}OpenSSL is a toolkit that
+protocol implementation of OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
 contains open-source implementations of the SSL and TLS cryptographic

### 2020-02-07 scripts/maintainability/create_reg_dataset.py
small fixes to reg. generation commits script

@@ -22,7 +22,7 @@ CLONAGE_DIR = "./tmp"
 def tool(security_dataset, regular_dataset):
     """CLI to collect random commits and analyze maintainability."""
     df = pd.read_csv(security_dataset)
-    rows = add_random_regular_commits(df)
+    rows = add_random_regular_commits(df, regular_dataset)
 
 @timer()
 def add_random_regular_commits(rows, regular_dataset):

### 2020-04-01 scripts/maintainability/create_reg_dataset.py
adapting scripts for regular and security + exception fix

@@ -15,9 +15,9 @@ from maintainability import ghutils, gitutils, log
 CLONAGE_DIR = "./tmp"
 
 @click.command()
-@click.option('--security-dataset', prompt=True, default="../dataset/db_icpc20_release_reg_final.csv",
+@click.option('--security-dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
               help="Input dataset of security commits.")
-@click.option('--regular-dataset', prompt=True, default="../dataset/db_icpc20_release_reg_final.csv",
+@click.option('--regular-dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
               help="Output dataset with regular commits.")
 def tool(security_dataset, regular_dataset):
     """CLI to collect random commits and analyze maintainability."""

### 2020-04-01 scripts/maintainability/eval_maintainability.py

@@ -5,6 +5,7 @@ import logging
 
 import click
 from contexttimer import timer
+import pandas as pd
 
 from maintainability import log
 import maintainability.better_code_hub as bch
@@ -12,37 +13,45 @@ import maintainability.better_code_hub as bch
 RETRY_ATTEMPTS = 10
 
 @click.command()
-@click.option('--dataset', prompt=True, default="../dataset/db_icpc20_multi_fix.csv",
+@click.option('--dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
               help="Dataset with commits for comparison.")
-def tool(dataset):
+@click.option('--commit', prompt=True, default="security",
+              help="type of dataset")
+def tool(dataset, commit):
     """CLI to evaluate maintainability."""
     log.info('Starting dataset analysis!')
     logging.basicConfig(level='INFO')
-    with open(dataset, 'r', newline='') as csvfile:
-        csv_reader = csv.DictReader(csvfile)
-        rows = list(csv_reader)
-        collect_maintainability(rows)
+    df = pd.read_csv(dataset)
+    collect_maintainability(df, commit, dataset)
 
 @timer()
-def collect_maintainability(rows):
+def collect_maintainability(rows, commit, dataset):
     """Collect maintainability of regular/random commits"""
-    for row in rows:
+    for i, row in rows.iterrows():        
         user = row['owner']
         project = row['project']
-        # if user == 'torvalds' and project == 'linux':
-        #     continue
-        commit_sha = row['sha']
-        parent_commit_sha = row['sha-p']
+        
+        if commit == 'regular':            
+            commit_sha = row['sha-reg']
+            parent_commit_sha = row['sha-reg-p']
+            if row['reg'] != 'FOUND':
+                continue
+        else:
+            commit_sha = row['sha']
+            parent_commit_sha = row['sha-p']
+               
         try:
             bch.robust_analyze_commit(user, project, commit_sha)
             bch.robust_analyze_commit(user, project, parent_commit_sha)
         except Exception as error:
             log.error(error)
+            rows.at[i, 'ERROR'] = 'YES'
             log.error("Exception not expected")
             if not isinstance(error, bch.BetterCodeHubException):
                 import pdb; pdb.set_trace()
             log.error(f"Skipping {user}/{project}.")
-
+        
+        rows.to_csv(dataset, index=False)
 
 if __name__ == '__main__':
     tool() # pylint: disable=no-value-for-parameter

### 2020-04-01 scripts/maintainability/gitutils.py

@@ -122,7 +122,8 @@ def resolve_commit(repo_dir, commit_sha_expression):
     try:
         hex_s = repo.commit(commit_sha_expression).hexsha
         return hex_s
-    except IndexError as idx_err:
-        log.error("Hexsha not resolved {}, repo: {}".format(commit_sha, repo))
+    except:
+        log.error("Hexsha not resolved {}, repo: {}".format(commit_sha_expression, repo))
         return None
+        
     

### 2020-04-01 paper/icsme20.tex
adpating paper

--- 2020-02-06 paper/icpc20.tex
@@
-\documentclass[sigconf,review]{acmart}
-
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
+\documentclass[10pt,conference]{IEEEtran}
+\IEEEoverridecommandlockouts
 
@@
 \usepackage{balance}
+\usepackage{url}
 
@@
 
-\setcopyright{acmcopyright}
-\copyrightyear{2020}
-\acmYear{2020}
-%\acmDOI{}
-
-\acmConference[ICPC'20]{ICPC '20: International Conference in Program Comprehension}{May 23--24, 2020}{Seoul, South Korea}
-% \acmBooktitle{}
-% \acmPrice{}
-% \acmISBN{}
-
 \begin{document}
@@
     Anonymou(s) Author(s)
+
+%     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
+% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+% \textit{name of organization (of Aff.)}\\
+% City, Country \\
+% email address}
+% \and
+% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
+% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+% \textit{name of organization (of Aff.)}\\
+% City, Country \\
+% email address}
+% \and
+% \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
+% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
+% \textit{name of organization (of Aff.)}\\
+% City, Country \\
+% email address}
 }
 
-\renewcommand{\shortauthors}{Anon. et al.}
-
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
+\maketitle
+
 \begin{abstract}
-
 Security is a requirement of utmost importance to produce software with 
@@
 open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
-Information Exposure (CWE-200) and Resource Management Errors (CWE-399). We conclude that changes to code bases 
-to fix vulnerabilities need to be performed with extra care and that security-related code evolution should be subject in computer curricula.
-
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399). 
+We conclude that changes to code bases to fix vulnerabilities need to be 
+performed with extra care and that security-related code evolution should 
+be subject in computer curricula.
 \end{abstract}
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
-%% Please copy and paste the code instead of the example below.
-%%
-%\begin{CCSXML}
-%<ccs2012>
-% <concept>
-%  <concept_id>10010520.10010553.10010562</concept_id>
-%  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
-%  <concept_significance>500</concept_significance>
-% </concept>
-% <concept>
-%  <concept_id>10010520.10010575.10010755</concept_id>
-%  <concept_desc>Computer systems organization~Redundancy</concept_desc>
-%  <concept_significance>300</concept_significance>
-% </concept>
-% <concept>
-%  <concept_id>10010520.10010553.10010554</concept_id>
-%  <concept_desc>Computer systems organization~Robotics</concept_desc>
-%  <concept_significance>100</concept_significance>
-% </concept>
-% <concept>
-%  <concept_id>10003033.10003083.10003095</concept_id>
-%  <concept_desc>Networks~Network reliability</concept_desc>
-%  <concept_significance>100</concept_significance>
-% </concept>
-%</ccs2012>
-%\end{CCSXML}
-%
-%\ccsdesc[500]{Computer systems organization~Embedded systems}
-%\ccsdesc[300]{Computer systems organization~Redundancy}
-%\ccsdesc{Computer systems organization~Robotics}
-%\ccsdesc[100]{Networks~Network reliability}
-
-\keywords{Security, Software Maintenance, Open-Source Software}
-
-\maketitle
+\begin{IEEEkeywords}
+Security, Software Maintenance, Open-Source Software
+\end{IEEEkeywords}
 
@@
 \end{lstlisting}
- \vspace{-0.5cm}
+
 \medskip
@@
 \end{lstlisting}
-\vspace{-0.3cm}
 %
@@
 	\label{fig:met}
-	\vspace{-0.5cm}
 \end{figure}
@@
 \label{tab:patterns}
-\vspace{-0.5cm}
 \end{table}
@@
 	\label{fig:bchrep}
-	\vspace{-0.6cm}
 \end{figure}
@@
 \label{tab:guidelines}
-\vspace{-0.2cm}
 \end{table}
@@
 	\label{fig:commit}
-	\vspace{-0.5cm}
 \end{figure}
@@
 	}
-	\vspace{-0.3cm}
 \end{figure}
@@
 	\label{fig:severity}
-	\vspace{-0.4cm}
 \end{figure}
@@
   \label{fig:lang_main}  
-  \vspace{-0.5cm}
 \end{figure}
@@
 	\label{fig:pat}	
-	\vspace{-0.5cm}
 \end{figure}
@@
 	\label{fig:secvsreg}
-	\vspace{-0.5cm}
 \end{figure}
@@
 
-% \begin{acks}
-% This work was supported by national funds through FCT, Fundação para a Ciência 
-% e a Tecnologia, under project UIDB/50021/2020 and…
+%
+% \section*{Acknowledgment}\label{sec:ack}
+%
+% \noindent
 % We thank the Software Improvement Group (SIG) team for all the support and
 % help in validating the methodology.
-% \end{acks}
-
-\bibliographystyle{ACM-Reference-Format}
-\bibliography{icpc20}
+% %ack pedro adao when accepted.
+% \pagebreak
+\balance
+
+{
+  \bibliographystyle{unsrt}
+  \bibliography{icsme20}
+}
 
 \end{document}
-\endinput
-%%
-%% End of file `sample-sigconf.tex'.

### 2020-04-06 scripts/report.py
language report refactored

@@ -21,6 +21,8 @@ from collections import OrderedDict
 guidelines = {'Write Short Units of Code':'WShortUC', 'Write Simple Units of Code':'WSimpleUC', 'Write Code Once':'WCO', 'Keep Unit Interfaces Small':'KUIS', 'Separate Concerns in Modules':'SCM',
                 'Couple Architecture Components Loosely':'CACL', 'Keep Architecture Components Balanced':'KACB', 'Write Clean Code':'WCC'}
 
+language_map = {'Java': ['java', 'scala'], 'Python': ['py'], 'Groovy': ['groovy'], 'JavaScript': ['js'] , 'PHP': ['ctp', 'php', 'inc', 'tpl'], 'Objective-C/C++': ['m', 'mm'], 'Ruby': ['rb'], 'C/C++': ['cpp', 'cc', 'h', 'c'], 'Config. Files': ['template', 'gemspec', 'VERSION', 'Gemfile', 'classpath', 'gradle', 'json', 'xml', 'bash', 'lock']}
+
 def readBCHCache(path):
     return bch.BCHCache(path)
     
@@ -199,84 +201,6 @@ def check_if_belongs_to_cwe(key):
         if key in composites[i]:
             return i
     return None
-
-def get_language(key):
-    language_map = {'Java': ['java', 'scala'], 'Python': ['py'], 'Groovy': ['groovy'], 'JavaScript': ['js'] , 'PHP': ['ctp', 'php', 'inc', 'tpl'], 'Objective-C/C++': ['m', 'mm'], 'Ruby': ['rb'], 'C/C++': ['cpp', 'cc', 'h', 'c'], 'Config. Files': ['template', 'gemspec', 'VERSION', 'Gemfile', 'classpath', 'gradle', 'json', 'xml', 'bash', 'lock']}
-    for k, v in language_map.items():
-        if key in language_map[k]:
-            return k
-    return None
-
-def report_maintainability_lang(reports, df, wilcoxon=True, boxes_start=6, text_start=0.67, delta=1, left_padding=0.25, f_size=9):
-                
-    for i, r in df.iterrows():
-        lang = get_language(r['Language'])
-        if lang != None:
-            df.at[i, 'Language'] = lang
-        
-    langs = [i for i in df['Language'].unique() if len(df[df['Language'] == i]) > 19]
-    
-    for i, r in df.iterrows():
-        if r['Language'] not in langs:
-            df.at[i, 'Language'] = 'Others'
-            
[…]

### 2020-04-09 scripts/report.py
new refactoring

@@ -1,4 +1,5 @@
 import maintainability.better_code_hub as bch
+import stats.chart as chart
 import argparse
 import csv
 import collections
@@ -11,17 +12,13 @@ matplotlib.rcParams['font.family'] = 'serif'
 matplotlib.rcParams['mathtext.fontset'] = 'cm'
 
 import matplotlib.pyplot as plt
-from scipy.stats import wilcoxon, shapiro
+from scipy.stats import wilcoxon
 from statistics import mean, stdev
 from math import sqrt
 import pandas as pd
 from collections import OrderedDict
 
 
-guidelines = {'Write Short Units of Code':'WShortUC', 'Write Simple Units of Code':'WSimpleUC', 'Write Code Once':'WCO', 'Keep Unit Interfaces Small':'KUIS', 'Separate Concerns in Modules':'SCM',
-                'Couple Architecture Components Loosely':'CACL', 'Keep Architecture Components Balanced':'KACB', 'Write Clean Code':'WCC'}
-
-language_map = {'Java': ['java', 'scala'], 'Python': ['py'], 'Groovy': ['groovy'], 'JavaScript': ['js'] , 'PHP': ['ctp', 'php', 'inc', 'tpl'], 'Objective-C/C++': ['m', 'mm'], 'Ruby': ['rb'], 'C/C++': ['cpp', 'cc', 'h', 'c'], 'Config. Files': ['template', 'gemspec', 'VERSION', 'Gemfile', 'classpath', 'gradle', 'json', 'xml', 'bash', 'lock']}
 
 def readBCHCache(path):
     return bch.BCHCache(path)
@@ -104,84 +101,7 @@ def report_guidelines_violin_plot(reports, filename):
     fig.tight_layout()
     fig.savefig('{}/{}'.format(reports, "maintainability_per_guideline_sns.pdf"))
 
-def report_maintainability_severity(reports, df, wilcoxon = True, boxes_start=3, text_start=0.55, delta=1, left_padding=0.05, f_size=7.5):
-    
-    results = {};
-    severity = ['LOW', 'MEDIUM', 'HIGH']
-    
-    for i, r in df.iterrows():
-        if r['Severity'] not in severity:
-            df.at[i, 'Severity'] = 'Other'
-    severity = ['Other'] + severity
-    
-    for s in severity:
-        impact = [0, 0, 0]
-        impact[0] = df[(df['diff'] < 0) & (df['Severity'] == s)].shape[0]
[…]

### 2020-04-09 scripts/stats/chart.py

@@ -1,0 +1,172 @@
+import matplotlib.pyplot as plt
+import pandas as pd
+
+from scipy.stats import wilcoxon
+import numpy as np
+
+from . import enum
+from . import tests
+from . import data
+
+
+_BAR_WIDTH = 0.25
+
+
+def set_bars(df, idx):
+    plt.barh(idx, df['pos'], _BAR_WIDTH, alpha=0.7, align='center', color='green', label='Positive')
+    plt.barh(idx + _BAR_WIDTH, df['nul'], _BAR_WIDTH, alpha=0.7, align='center', color='orange', label='None')
+    plt.barh(idx + 2*_BAR_WIDTH, df['neg'], _BAR_WIDTH, alpha=0.7, align='center', color='red', label='Negative')
+
+def format_p_value(p):
+    return '<0.001' if float('{:.{}f}'.format(p, 3)) <= 0.000 else '={:.{}f}'.format(p, 3)
+
+def set_ticks(df, idx, fontsize=7):
+    plt.yticks(idx + _BAR_WIDTH, df['type'], fontsize=fontsize)
+    plt.xticks(fontsize=fontsize)
+    plt.gca().set_xticklabels(['{:.0f}%'.format(x*100) for x in plt.gca().get_xticks()])
+
+def config_report(data, yaxs_size, x, y, plt_config):
+    fig = plt.figure(figsize=(x,y))
+    return pd.DataFrame(data), np.arange(yaxs_size), fig, fig.add_subplot(plt_config)
+
+def save_report(path, name):
+    plt.gca().xaxis.grid(True, linestyle='--')
+    plt.tight_layout()
+    plt.savefig('{}/{}'.format(path, name))
+
+# comparison chart
+def main_comparison_chart(reports, df_sec, df_reg):    
+
+        test = (tests.hypothesis_test(df_sec['diff']), tests.hypothesis_test(df_reg['diff']))
+        total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)
+        
+        db_size = len(df_sec)        
+
+        d = {'neg' : pd.Series([total_reg['neg']/db_size, total_sec['neg']/db_size]),
+            'pos' : pd.Series([total_reg['pos']/db_size, total_sec['pos']/db_size]),
+            'nul' : pd.Series([total_reg['nul']/db_size, total_sec['nul']/db_size]),
+            'type' : pd.Series(['Regular Change','Security Change'])}
+        
[…]

### 2020-04-09 scripts/stats/data.py

@@ -1,0 +1,26 @@
+
+
+def filter_results_per_field(df, t, field):
+    return [df[(df['diff'] < 0) & (df[field] == t)].shape[0],
+                df[(df['diff'] > 0) & (df[field] == t)].shape[0],
+                df[(df['diff'] == 0) & (df[field] == t)].shape[0]]
+
+def filter_results(df):
+    return {'neg':df[df['diff'] < 0].shape[0],
+                'pos':df[df['diff'] > 0].shape[0],
+                'nul':df[df['diff'] == 0].shape[0]}
+
+def filter_results_per_guideline(df, g):
+    return [df[df[g+'-diff'] < 0].shape[0],
+                df[df[g+'-diff'] > 0].shape[0],
+                df[df[g+'-diff'] == 0].shape[0]]
+
+def add_others_group(df, group, key):
+    for i, r in df.iterrows():
+        if r[key] not in group:
+            df.at[i, key] = 'Other'
+    return ['Other'] + group
+    
+
+    
+    

### 2020-04-09 scripts/stats/enum.py

@@ -1,0 +1,28 @@
+# data.py 
+
+guidelines = {'Write Short Units of Code':'WShortUC', 
+                'Write Simple Units of Code':'WSimpleUC', 
+                'Write Code Once':'WCO', 
+                'Keep Unit Interfaces Small':'KUIS', 
+                'Separate Concerns in Modules':'SCM',
+                'Couple Architecture Components Loosely':'CACL', 
+                'Keep Architecture Components Balanced':'KACB', 
+                'Write Clean Code':'WCC'}
+
+languages = {'Java': ['java', 'scala'], 
+                'Python': ['py'], 
+                'Groovy': ['groovy'], 
+                'JavaScript': ['js'] , 
+                'PHP': ['ctp', 'php', 'inc', 'tpl'], 
+                'Objective-C/C++': ['m', 'mm'], 
+                'Ruby': ['rb'], 
+                'C/C++': ['cpp', 'cc', 'h', 'c'], 
+                'Config. Files': ['template', 'gemspec', 'VERSION', 'Gemfile', 'classpath', 'gradle', 'json', 'xml', 'bash', 'lock']}
+
+severity = ['LOW', 'MEDIUM', 'HIGH']
+
+def get_language(key):
+    for k, v in languages.items():
+        if key in languages[k]:
+            return k
+    return None

### 2020-04-09 scripts/stats/tests.py

@@ -1,0 +1,12 @@
+import pandas as pd
+from scipy.stats import wilcoxon
+
+def hypothesis_test(diff):
+    test, pvalue = wilcoxon(x=diff, zero_method="pratt")
+    return pd.DataFrame({'test': [test], 
+                        'med':[pd.DataFrame({'diff': diff})['diff'].median()],
+                        'pvalue': [pvalue], 
+                        'mean':[pd.DataFrame({'diff': diff})['diff'].mean()]})
+
+def filter_small_sample_groups(df, key):
+    return [i for i in df[key].unique() if len(df[df[key] == i]) > 19]

### 2020-04-09 scripts/report.py
report update and organization

@@ -18,61 +18,6 @@ from math import sqrt
 import pandas as pd
 from collections import OrderedDict
 
-
-
-def readBCHCache(path):
-    return bch.BCHCache(path)
-    
-def main_calculation(df, cache, dataset):
-    
-    none = 0; error = 0    
-
-    if dataset == 'regular':
-        sha_key = 'sha-reg'
-        sha_p_key = 'sha-reg-p'
-    else:
-        sha_key = 'sha'
-        sha_p_key = 'sha-p'
-    
-    for i, r in df.iterrows():
-
-        if r[sha_key] is np.nan or r[sha_p_key] is np.nan:
-            continue
-            
-        info_f = cache.get_stored_commit_analysis(r['owner'], r['project'], r[sha_key])
-        info_p = cache.get_stored_commit_analysis(r['owner'], r['project'], r[sha_p_key])
-            
-        if info_f is None or info_p is None:
-            none+=1
-            df.at[i, 'diff'] = np.nan
-            continue
-            
-        if info_f.get('error') or info_p.get('error'):
-            error+=1
-            df.at[i, 'diff'] = np.nan
-            continue
-        
-        try:
-            main_f = bch.compute_maintainability_score_per_guideline(info_f)
-            main_p = bch.compute_maintainability_score_per_guideline(info_p)
-            
-            score_f = bch.compute_maintainability_score(info_f)
-            score_p = bch.compute_maintainability_score(info_p)
-
-            df.at[i, 'main_fix'] = score_f
-            df.at[i, 'main_prev'] = score_p
-            df.at[i, 'diff'] = score_f - score_p
-            
-            for k in main_f.keys():
-                df.at[i, k+'-fix'] = main_f[k]
-                df.at[i, k+'-prev'] = main_p[k]
-                df.at[i, k+'-diff'] = main_f[k] - main_p[k]
-                
-        except ZeroDivisionError as e:
-            df.at[i, 'diff'] = np.nan
-            error+=1
-    
-    return df    
     
 def report_guidelines_violin_plot(reports, filename):
     
@@ -200,6 +145,7 @@ def report_maintainability_cwe(reports, df, wilcoxon=True, boxes_start=9, text_s
 def merge_cache(cache_1_filename, cache_2_filename):
     cache_1 = readBCHCache(cache_1_filename)
     print('Length cache 1: {}'.format(len(cache_1.get_data())))
+    
[…]

### 2020-04-09 scripts/report.py
more refactoring

@@ -12,157 +12,9 @@ matplotlib.rcParams['font.family'] = 'serif'
 matplotlib.rcParams['mathtext.fontset'] = 'cm'
 
 import matplotlib.pyplot as plt
-from scipy.stats import wilcoxon
-from statistics import mean, stdev
 from math import sqrt
 import pandas as pd
 from collections import OrderedDict
-
-    
-def report_guidelines_violin_plot(reports, filename):
-    
-    df = pd.read_csv(filename)
-    
-    stats = {'value':[], 'fix': [], 'guideline':[]}
-    for i, r in df.iterrows():
-        for g in guidelines:
-            stats['guideline'].append(g)
-            stats['guideline'].append(g)
-            stats['value'].append(r[g+'-prev'])
-            stats['fix'].append('before')
-            stats['value'].append(r[g+'-fix'])
-            stats['fix'].append('after')
-    
-    f = pd.DataFrame.from_dict(stats)
-    print(f)
-    
-    g = sns.violinplot(x="guideline", y="value", hue="fix", data=f, palette="Set2", split=True)
-    #g.set_yscale("symlog")
-    plt.ylim(-9000000,9000000) 
-    plt.xticks(rotation=90)
-    
-    fig = matplotlib.pyplot.gcf()
-    fig.set_size_inches(18.5, 10.5, forward=True)
-    fig.tight_layout()
-    fig.savefig('{}/{}'.format(reports, "maintainability_per_guideline_sns.pdf"))
-
-
-
-def read_cwe_composites(file):
-    composites = dict()
-    with open(file, 'r') as i:
-        f = i.readlines()
-        for group in f:
-            cwes = group.strip().split('\t')
-            if cwes[0] not in composites:
-                composites[cwes[0]] = cwes[1::]
-        return composites
-
-def check_if_belongs_to_cwe(key):
-    composites = read_cwe_composites('cwe')
-    if key in composites.keys():
-        return key
-    for i in composites:
-        if key in composites[i]:
-            return i
-    return None
-
-def report_maintainability_cwe(reports, df, wilcoxon=True, boxes_start=9, text_start=0.55, delta=1, left_padding=0.25, f_size=9):
-                
-    for i, r in df.iterrows():
-        if r['CWE'] is not np.nan:
[…]

### 2020-04-09 scripts/stats/chart.py

@@ -34,7 +34,6 @@ def save_report(path, name):
     plt.tight_layout()
     plt.savefig('{}/{}'.format(path, name))
 
-# comparison chart
 def main_comparison_chart(reports, df_sec, df_reg):    
 
         test = (tests.hypothesis_test(df_sec['diff']), tests.hypothesis_test(df_reg['diff']))
@@ -63,6 +62,43 @@ def main_comparison_chart(reports, df_sec, df_reg):
             y -= 1.1
         
         save_report(reports, 'main_comparison.pdf')
+
+def main_per_cwe_chart(reports, df, wilcoxon = True):
+                
+    for i, r in df.iterrows():
+        if r['CWE'] is not np.nan:
+            cwe = enum.check_if_belongs_to_cwe(r['CWE'])
+            if cwe != None:
+                df.at[i, 'CWE'] = cwe
+    
+    cwes = data.add_others_group(df, tests.filter_small_cwe_groups(df), 'CWE', 'MISC')    
+    results = {c:data.filter_results_per_field(df, c, 'CWE') for c in cwes}      
+    test_cwe = [tests.hypothesis_test(df[df['CWE'] == c]['diff']) for c in cwes]
+    
+    d = {'neg': pd.Series([results[i][0]/sum(results[i]) for i in cwes]),
+        'pos': pd.Series([results[i][1]/sum(results[i]) for i in cwes]),
+        'nul': pd.Series([results[i][2]/sum(results[i]) for i in cwes]),
+        'N': pd.Series([sum(results[i]) for i in cwes if wilcoxon]),
+        'mean': pd.Series([i['mean'][0] for i in test_cwe if wilcoxon]),
+        'med': pd.Series([i['med'][0] for i in test_cwe if wilcoxon]),
+        'p': pd.Series([i['pvalue'][0] for i in test_cwe if wilcoxon]),
+        'type': pd.Series([i for i in cwes])}
+
+    df, idx, fig, ax = config_report(d, len(d['type']), 5, 10, 111)
+    
+    set_bars(df, idx)
+    set_ticks(df, idx, 7)
+    plt.subplots_adjust(left=0.25, right=0.85, top=0.9, bottom=0.06)
+    plt.legend(loc='upper center', bbox_to_anchor=(0.5, 1.04), fancybox=True, ncol=3, fontsize=9)
+
+    boxes_start=9
+    for l in df[::-1].iterrows():
+        p = format_p_value(l[1]['p'])
[…]

### 2020-04-09 scripts/stats/data.py

@@ -15,11 +15,11 @@ def filter_results_per_guideline(df, g):
                 df[df[g+'-diff'] > 0].shape[0],
                 df[df[g+'-diff'] == 0].shape[0]]
 
-def add_others_group(df, group, key):
+def add_others_group(df, group, key, label):
     for i, r in df.iterrows():
         if r[key] not in group:
-            df.at[i, key] = 'Other'
-    return ['Other'] + group
+            df.at[i, key] = label
+    return [label] + group
     
 
     

### 2020-04-09 scripts/stats/enum.py

@@ -26,3 +26,22 @@ def get_language(key):
         if key in languages[k]:
             return k
     return None
+
+def read_cwe_composites(file):
+    composites = {}
+    with open(file, 'r') as i:
+        f = i.readlines()
+        for group in f:
+            cwes = group.strip().split('\t')
+            if cwes[0] not in composites:
+                composites[cwes[0]] = cwes[1::]
+    return composites
+
+def check_if_belongs_to_cwe(key):
+    composites = read_cwe_composites('stats/cwe')
+    if key in composites.keys():
+        return key
+    for i in composites:
+        if key in composites[i]:
+            return i
+    return None

### 2020-04-09 scripts/stats/tests.py

@@ -10,3 +10,6 @@ def hypothesis_test(diff):
 
 def filter_small_sample_groups(df, key):
     return [i for i in df[key].unique() if len(df[df[key] == i]) > 19]
+
+def filter_small_cwe_groups(df):
+    return [i for i in df['CWE'].unique() if str(i) != 'nan' and (len(df[df['CWE'] == i]) > 19) and 'CWE' in i]

### 2020-04-09 scripts/stats/chart.py
1st version of the package

@@ -52,13 +52,13 @@ def main_comparison_chart(reports, df_sec, df_reg):
         set_ticks(df, idx, 9)
         
         plt.subplots_adjust(left=0.2, right=0.85, top=0.9, bottom=0.1)
-        plt.legend(loc='upper center', bbox_to_anchor=(0.5, 1.1), fancybox=True, ncol=3, fontsize=9)
+        plt.legend(loc='upper center', bbox_to_anchor=(0.5, 1.07), fancybox=True, ncol=3, fontsize=8)
 
         y = 1.2
         for l in test:
             p = format_p_value(l['pvalue'][0])
             box_text = '$\overline{x}$='+ '{:.{}f}'.format(l['mean'][0], 2) + '\nM=' + '{:.{}f}'.format(l['med'][0], 2) + '\np' + p
-            ax.text(0.43, y, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':4}, fontsize=9)
+            ax.text(0.4, y, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':4}, fontsize=10)
             y -= 1.1
         
         save_report(reports, 'main_comparison.pdf')
@@ -87,15 +87,15 @@ def main_per_cwe_chart(reports, df, wilcoxon = True):
     df, idx, fig, ax = config_report(d, len(d['type']), 5, 10, 111)
     
     set_bars(df, idx)
-    set_ticks(df, idx, 7)
+    set_ticks(df, idx, 9)
     plt.subplots_adjust(left=0.25, right=0.85, top=0.9, bottom=0.06)
-    plt.legend(loc='upper center', bbox_to_anchor=(0.5, 1.04), fancybox=True, ncol=3, fontsize=9)
+    plt.legend(loc='upper center', bbox_to_anchor=(0.5, 1.04), fancybox=True, ncol=3, fontsize=8)
 
     boxes_start=9
     for l in df[::-1].iterrows():
         p = format_p_value(l[1]['p'])
         box_text = 'N='+str(l[1]['N'])+'\n$\overline{x}$='+ '{:.{}f}'.format(l[1]['mean'], 2) + '\nM=' + '{:.{}f}'.format(l[1]['med'], 2) + '\np' + p
-        ax.text(0.55, boxes_start, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':3}, fontsize=8)
[…]

### 2020-04-09 paper/icsme20.tex

@@
 in this paper, we present an empirical study on the impact of patches of 
-vulnerabilities in software maintenance across open-source software.
+vulnerabilities on software maintenance across open-source software.
 %
@@
 As ISO does not provide any specific guidelines or formulas to calculate 
-maintainability, we resort on Software Improvement Group (SIG)'s web-based source 
+maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
 code analysis service Better Code Hub (BCH) to compute the software compliance 
@@
 dataset of $711$ test cases for $16$ security patterns. Some of the patterns
-are based in the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
+are based on the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
 $2017$~\cite{oswap:2017}. Each test case of the
@@
 a change with the same \emph{diff} the step increased. Thus, if a change suffered $20$ deletions 
-and $10$ additions. Then, in the first $10$ attempts, the algorithm searches for a \emph{diff} with 
+and $10$ additions, then, in the first $10$ attempts, the algorithm searches for a \emph{diff} with 
 deletions between $19$ and $21$ and additions between $9$ and $11$. After the $10$ first attempts, 
@@
 refers to the OpenSSL CVE-$2014$-$3506$ vulnerability refactoring ---
-as described as motivating example in Section~\ref{sec:motivation}. This
+as described by Section~\ref{sec:motivation}. This
 version of OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write
@@
 Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security patches have a negative impact in the maintainability of
+\emph{security patches have a negative impact on the maintainability of
 open-source software}.

### 2020-04-15 paper/icsme20.tex
paper update

@@
 \usepackage{balance}
-\usepackage{url}
+\usepackage{hyperref}
 
@@
 
-\title{Fixing Vulnerabilities Hinders Maintainability}
+% \title{Fixing Vulnerabilities Hinders Maintainability}
+\title{\Sof{TBA - New Titles in the comments}}
+%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
+%\title{The reason why you should be using coding practices: safer software.}
+%\title{What is the impact of security patches in software evolution?}
+%\title{Hey, security engineers! You should pay more attention to your patches.}
 
@@
 Maintainability is measured based on the Better Code Hub’s model of $10$ 
-guidelines using a dataset including $1037$ security commits. Results show 
-that fixing software vulnerabilities hinders the maintainability of 
-open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
-Information Exposure (CWE-200) and Resource Management Errors (CWE-399). 
-We conclude that changes to code bases to fix vulnerabilities need to be 
-performed with extra care and that security-related code evolution should 
-be subject in computer curricula.
+guidelines using a dataset including $1335$ security commits
+\Sof{These are the initial values, should I use the final? In the end, 
+we studied 1031 commits and 8 guidelines}. 
+Results show 
+that fixing software vulnerabilities may hinder the maintainability of 
+open-source software. This is particularly the case for patches that 
+address Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource 
+Management Errors (CWE-399). We conclude that changes to code bases to patch 
+vulnerabilities need to be performed with extra care and that security-related 
+code evolution should be subject in computer curricula.
 \end{abstract}
@@
 These vulnerabilities are gateways for attackers to penetrate systems and 
-exploit their resources. \Sof{Improve example - outdated}As an example, in the last week of January $2019$, 
+exploit their resources. \Sof{Update example for 2020 - outdated}As an example, in the last week of January $2019$, 
 a severe vulnerability in Apple FaceTime was revealed\footnote{FaceTime vulnerability 
-details available at \url{https://thehackernews.com/2019/01/apple-facetime-
-privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
-flaw that allowed callers to see and hear others without them picking up the call.%\cite{cacheOut}.
+details available at \url{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
+flaw that allowed callers to see and hear others without them picking up the call.\Sof{---}%\cite{cacheOut}.
 %
@@
 we intend to highlight the need for tools to predict the impact of patches on 
-maintainability~\cite{4724577} and documentation to help developers adopt security 
-best practices~\cite{6311252, 7927935, MESQUIDA201519}.
+maintainability~\cite{4724577}, documentation to help developers adopt security 
+best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
+security-related code evolution in computer curricula.
 %
@@
 	empirical evaluation, for reproducibility. Available online:
-  \textit{<anonymized for double blind review>}.
+  \url{https://figshare.com/s/4861207064900dfb3372}.
 \end{itemize}
@@
 presents the results and section~\ref{sec:discussion} discusses their
-implications; section~\ref{sec:threats}, enumerates the threats to the validity of
+implications; section~\ref{sec:threats} enumerates the threats to the validity of
 this study; section~\ref{sec:rw} describes the different work and existing
@@
 %%%%% NEEDS REVIEW
-As an example, consider the patch of the TLS state-machine protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
+\Sof{Improve for 2nd version}
+As an example, consider the patch of the TLS state-machine protocol implementation in 
+OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
 (DoS) attack by crafting TLS messages to trigger memory allocations
@@
 OpenSSL maintainability.
+\Sof{-- END --}
 %%%%%%% 
-Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software
+Software maintainability is designed as the degree to which an application is understood, 
+repaired or enhanced. In this paper, our concern is to study whether, while improving software
 security, developers are also negatively impacting the maintainability of their
-software applications.
-This is important because software maintainability is approximately $75\%$ of the cost related to a project.
-To answer the following three research questions, we use two datasets
-of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
+software applications. This is important because software maintainability is 
+approximately $75\%$ of the cost related to a project. To answer the following three 
+research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
+to measure the impact of
 security patches on the maintainability of open-source software. 
@@
 
-\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the
-maintainability of open-source software?}}}
-
-Often security flaws require patching code to make software more secure.
+\textit{\textbf{RQ1: What is the impact of security patches on the
+maintainability of open-source software?}} Often security flaws require patching 
+code to make software more secure.
 However, \textbf{there is no evidence yet of how security patches impact the
@@
 address a security flaw because they tend to choose the easiest path to solve
-it. To address it, we compute the maintainability of $1037$ commits using the
+it. To address it, we compute the maintainability of $1335$ commits using the
 \emph{Better Code Hub} tool. 
 %
-\subsubsection*{\textit{\textbf{RQ2}  \textbf{Which patterns of security patches are more likely to
-affect open-source software maintainability?}}}
-
+
+\textit{\textbf{RQ2: Which patterns of security patches are more likely to
+affect open-source software maintainability?}}
 There are security flaws that are more difficult to patch than others. For
@@
 without adding new lines of code/complexity to the code. A typical fix for 
-the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to \textit{escape} the data given by the variable
+the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to escape the data given by the variable
 \texttt{\$\_['file']}. \textbf{Knowing which patterns are more likely to increase
@@
 %
-\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}}}
-%
-We also compute the maintainability of $1037$ random regular commits using the
+\textit{\textbf{RQ3: What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}}
+%
+We also compute the maintainability of random regular commits using the
 \emph{Better Code Hub} tool --- baseline.  We use them to understand \textbf{how
@@
 	\item
-	Extract relevant data (e.g., sha, sha of the parent commit, owner of the project, name of 
-	the repository) and code bases from the combined dataset containing $1335$ security
+	Extract relevant data (e.g., owner and name of
+	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and code bases from the combined dataset containing $1335$ security
 	refactorings collected from open-source software available on
@@
 open-source software, we use a combined dataset of security vulnerabilities which is 
-the outcome of mining/manually inspect $267$ GitHub projects. Reis and Abreu 
+the outcome of mining/manually inspect $265$ GitHub projects. Reis and Abreu 
 ($2017$) mined open-source
@@
 of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
-dataset of $711$ test cases for $16$ security patterns. Some of the patterns
+dataset of $711$ test cases for $16$ different security vulnerability types. Some of the vulnerability types
 are based on the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
@@
 another (typically, called \textit{diffs}) --- where one can easily review the
-code used to fix the security flaw. In this study, we focus on computing the
-maintainability of the commits before and after the security patching to
-evaluate if its impact was positive, negative or none.
+code used to fix the security flaw. 
 %
@@
 from $2014$ to $2019$. For each new vulnerability, the authors manually searched 
-for references to commits involved in the patch on the NVD's website. However, $70\%$
+for references to commits involved in the patch on the National Vulnerability Database (NVD) website. However, $70\%$
 of the vulnerabilities did not have any references to commits. Thus, the authors
@@
 To fit the dataset in our methodology, we search for the first and last commits
-used to patch the vulnerability. We assumed that the last commit of the patch as 
+used to patch the vulnerability. We assumed the last commit of the patch as 
 the fix (\emph{sha}) and the parent of the first commit as the vulnerable version 
@@
 %
+In this study, we focus on computing the
+maintainability of the commits before and after the security patching to
+evaluate if its impact was positive, negative or none.
 The $1335$ patches in the dataset were run against the BCH toolset to
@@
 language support and project size. The final dataset used in this paper comprises
-$1037$ security patches. $556$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE). In Table~\ref{tab:patterns}, the most prevailing categories are presented. Patterns
-with less than $20$ instances were merged in a bigger pattern named \textit{Miscellaneous}.
+$1031$ security patches. $539$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE). In Table~\ref{tab:patterns}, the most prevailing categories are presented. To satisfy the Wilcoxon test size requirement, types of vulnerabilities
+with less than $20$ instances were merged in a bigger group named \textit{Miscellaneous}.
 
 \begin{table}[h]
-\footnotesize
+\scriptsize
 	\caption{Guidelines to produce maintainable code}
@@
  	     	           \textbf{CWE-79} & 
-Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)& 
+Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (63 cases)& 
 Lack of proper validation or escaping
@@
  	 	     	           \textbf{MISC} & 
-Miscellaneous (673 cases) & 
+Miscellaneous (666 cases) & 
 This pattern comprises
@@
 of some patches, it was not possible to find patches with the exact number of added and 
-deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. The algorithm follows the following steps:
+deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. The algorithm integrates the following steps:
 
 \begin{enumerate}
-\item Diff comparison between the security patch commit (\emph{sha}) and its parent (\emph{sha-p}).
-\item Selects a random commit from the security patch project (\emph{sha-reg}).
-\item Diff comparison between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
-\item Check if the size from the regular change has the same/approximated size to 
+\item Calculates the \emph{diff} between the security patch commit (\emph{sha}) and its parent (\emph{sha-p}).
+\item Selects a random commit from the project (\emph{sha-reg}).
+\item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
+\item Checks if the size from the regular change has the same/approximated size to 
 the security patch.
@@
 
-Our approximation was calculated using an initial step of $1$. Every $10$ attempts to search for 
-a change with the same \emph{diff} the step increased. Thus, if a change suffered $20$ deletions 
-and $10$ additions, then, in the first $10$ attempts, the algorithm searches for a \emph{diff} with 
-deletions between $19$ and $21$ and additions between $9$ and $11$. After the $10$ first attempts, 
-the algorithm increase the step for $2$ and searches for a \emph{diff} with deletions between $18$ and $22$ and additions between 
-$8$ and $10$. We originate the regular commits from the security
-commits to ensure that differences in maintainability are not a consequence of
-characteristics of different projects.
+The pair of the random commit and its parent is accepted if the \emph{diff} 
+size fits in the range size considered acceptable at the moment. The algorithm 
+starts with a step of $1$. Every $10$ attempts to search for a change with the 
+same \emph{diff} the step increases. Thus, if a change suffered $20$ deletions 
+and $10$ additions, then, in the first $10$ attempts, the algorithm searches 
+for a \emph{diff} with deletions between $19$ and $21$ and additions between 
+$9$ and $11$. After the $10$ first attempts, the algorithm increase the step 
+for $2$ and searches for a \emph{diff} with deletions between $18$ and $22$ 
+and additions between $8$ and $10$. We originate the regular commits from the 
+security commits to ensure that differences in maintainability are not a 
+consequence of characteristics of different projects.
 
@@
 
-In this research, we follow a very similar methodology to the one presented in previous work on the maintainability 
-of energy-oriented fixes~\cite{cruz2019energyoriented}. As said before, the web-based source code analysis service \emph{Better Code
+In this research, we follow a very similar methodology to the one presented 
+in previous work on the maintainability of energy-oriented fixes~\cite{cruz2019energyoriented}. 
+As said before, the web-based source code analysis service \emph{Better Code
 Hub} (BCH) is used to collect the maintainability reports of the refactorings of
@@
 calculated by BCH. Theses guidelines are calculated using the metrics 
-presented in~\cite{criteria:2017}. During each guideline evaluation, the tool
-determines the compliance towards one guideline by establishing limits for the
-percentage of code allowed to be in each of the $4$ risk severity levels
+presented in~\cite{criteria:2017} and also briefly explained on Table~\ref{tab:guidelines}. 
+During each guideline evaluation, the tool determines the compliance towards 
+one guideline by establishing limits for the percentage of code allowed to be 
+in each of the $4$ risk severity levels
 (\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
@@
 \begin{table}[h]
-\footnotesize
+\scriptsize
 	\caption{Guidelines to produce maintainable code}
@@
 
-SIG defines \emph{Units} as the smallest groups of code that can be maintained
+\Sof{Needs improvement}SIG defines \emph{Units} as the smallest groups of code that can be maintained
 and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and
@@
 somewhere in the green bar it means it is compliant with a low level of
-severity.
+severity.\Sof{--END--}
 
@@
 to capture the distance between the current state of the project and
-the standard thresholds calculated by the BCH based on the insights provided in~\cite{Olivari:2018}. 
-The equation considers the following:
-\begin{itemize}
-	\item \textbf{Size of project changes does not affect the maintainability
-	difference between two versions, $\Delta M (v_{s-1},v_{s}) = M(v_{s}) - M(v_{s-1})$.} We
-	aim at evaluating security patterns occurring in different projects similarly.
-	Thus, the derived metric uses the \textit{raw} number of lines of code rather
-  percentages for normalization purposes.
-	\item \textbf{Distance to lower severity levels is less penalized than to the
-	thresholds in high severity levels.} Severity level weights based on the
-	severity level to count lines of code that violate maintainability guidelines.
-\end{itemize}
+the standard thresholds calculated by the BCH based on the insights 
+provided in~\cite{Olivari:2018}. The equation provided in~\cite{cruz2019energyoriented} 
+considers that the size of project changes does not affect the maintainability and
+that the distance to lower severity levels is less penalized than to the
+thresholds in high severity levels.
+% 	thresholds in high severity levels
+% 	difference between two versions
+% The equation considers the following:
+% \begin{itemize}
+% 	\item \textbf{Size of project changes does not affect the maintainability
+% 	difference between two versions, $\Delta M (v_{s-1},v_{s}) = M(v_{s}) - M(v_{s-1})$.} We
+% 	aim at evaluating security patterns occurring in different projects similarly.
+% 	Thus, the derived metric uses the \textit{raw} number of lines of code rather
+%   percentages for normalization purposes.
+% 	\item \textbf{Distance to lower severity levels is less penalized than to the
+% 	thresholds in high severity levels.} Severity level weights based on the
+% 	severity level to count lines of code that violate maintainability guidelines.
+% \end{itemize}
 
@@
 
-This study evaluates $1037$ security and baseline commits
-from $267$ distinct open-source projects. The following section presents the
+This study evaluates $1031$ security and baseline commits
+from $265$ distinct open-source projects. The following section presents the
 results obtained for each research question.
 
-\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the
-maintainability of open-source software?}}}
+\textit{\textbf{RQ1: What is the impact of security patches on the
+maintainability of open-source software?}}
 
@@
  	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/impact_per_guideline.pdf}
- 	\caption{Impact of the security patches for each guideline.}
+ 	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
+ 	\caption{Impact of the security patches per guideline.}
 	\label{fig:guidelines}
@@
 by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
-guideline individually. For projects with large 
-codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
-viable due to limitations of BCH.
-For this guideline, the tool sets a limit of 20 Person-years.
-However, a large amount of projects of our database surpass this limit.
-Thus, we decided to not consider this guideline
-in our analysis. For each type of
+guideline individually. 
+For each type of
 guideline, three types of impact of patches in the software maintainability
@@
 
-Regarding security commits results, we observe that maintainability decreases in
-$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
-($400$) of the cases after security patches are applied. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
-Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security patches have a negative impact on the maintainability of
-open-source software}.
 Regarding the impact of security patches for each guideline, we observe
@@
 Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
-$7.9$x$10^{-35}$).
-However, for each of these guidelines we also observe a substantial number of cases where the maintainability suffers a negative impact $-$ from $10\%$ to $35\%$ $-$ which highlights the need of better practices/techniques/tools to produce software.
-\emph{Separate Concerns in Module}
+$7.9$x$10^{-35}$). However, for each of these guidelines we also observe a 
+substantial number of cases where the maintainability suffers a negative 
+impact $-$ from $10\%$ to $35\%$ $-$ which highlights the need of better 
+practices/techniques/tools to produce software. \emph{Separate Concerns in Module} 
 guideline also follows the same results but without any statistical 
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.45\textwidth]{figures/maintainability_severity.pdf}
+ 	\centering 	\includegraphics[width=0.45\textwidth]{figures/main_per_severity.pdf}
  	\caption{Maintainability difference by the vulnerabilities severity.}
@@
 \end{figure}
+
+For projects with large 
+codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
+viable because they retrieved values above the limit set by BCH ($20$ Person-years).
+The \emph{Automated Tests} guideline was also not considered because the BCH does not return
+the amount of lines of code that complies with the guideline -- which is one of the variables
+our formula uses to calculate the maintainability of the project regarding each guideline. 
+Thus, we decided to not consider both guidelines in our analysis.
 
@@
 software. Although with no statistical significance, we observe that 
-vulnerabilities with higher severity levels (\emph{HIGH} and \emph{MEDIUM}) have more negative impact than the \emph{LOW} severity level on the
+vulnerabilities with higher severity levels (\emph{HIGH} and \emph{MEDIUM}) have 
+more negative impact than the \emph{LOW} severity level on the
 maintainability of software.
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/maintainability_language.pdf}
+  \includegraphics[width=0.5\textwidth]{figures/main_per_language.pdf}
   \caption{Maintainability difference by language after security patches.}
@@
 We have also evaluated the impact on the maintainability per programming language
-of open-source software, Figure~\ref{fig:lang_main}.
-We restrict this analysis to programming languages with at least $20$ data points, as this is a requirement for our hypothesis tests.
-Thus, we compare the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP},
+of open-source software, Figure~\ref{fig:lang_main}. We restrict this analysis to 
+programming languages with at least $20$ data points, as this is a requirement 
+for our hypothesis tests. Thus, we compare the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
 \emph{Objective-C/C++}, and \emph{Java}. The other languages were
@@
 \emph{Ruby} and \emph{PHP} languages also impact negatively the maintainability of software. 
-Despite not having statistical significance, there are a considerable number of \emph{Ruby} and \emph{PHP} patches hindering maintainability that should not be disregarded
+Despite not having statistical significance, there are a considerable number of \emph{Ruby} 
+and \emph{PHP} patches hindering maintainability that should not be disregarded.
 
@@
 security patches.
+
+Regarding security commits results, we observe that maintainability decreases in
+$41.61\%$ ($429$), remains the same in $19.69\%$ ($203$) and increases in $38.70\%$
+($399$) of the cases after security patches are applied. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $4.9$x$10^{-2}$.
+Since the p-value is below the significance level of $0.05$, we argue that
+\emph{security patches have a negative impact on the maintainability of
+open-source software}.
 
@@
  	\centering
- 	\includegraphics[width=0.4\textwidth]{figures/maintainability_cwe.pdf}
+ 	\includegraphics[width=0.4\textwidth]{figures/main_per_cwe.pdf}
  	\caption{Maintainability difference by pattern after security patches.}
@@
 \begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
+ 	\centering 	\includegraphics[width=0.43\textwidth]{figures/main_comparison.pdf}
  	\caption{Maintainability difference for security and baseline patches.}
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $1037$ regular commits are enough to
+Furthermore, this evaluation considers that $1031$ regular commits are enough to
 alleviate random irregularities in the maintainability differences of the
@@
 
-This work presents an empirical study on the impact of $1037$ security
+This work presents an empirical study on the impact of $1031$ security
 patches on the maintainability of $267$ open-source projects. We leveraged

### 2020-04-15 scripts/maintainability/better_code_hub.py
package updated

@@ -182,6 +182,7 @@ def _compute_maintainability_for_guideline(guideline, total_loc):
         return 0
     if guideline['guideline'] == "Keep Your Codebase Small":
         return 0
+        
     volumes = guideline.get('qualityProfileVolume')
     if guideline['guideline'] == "Keep Architecture Components Balanced":
         total_compoments = sum(volumes)

### 2020-04-15 scripts/report.py

@@ -103,6 +103,10 @@ def guideline(secdb, reports):
 def cwe(secdb, reports):
     df_sec = pd.read_csv(secdb)
     chart.main_per_cwe_chart(reports, df_sec)
+    
+def cwe_spec(secdb, reports, cwe):
+    df_sec = pd.read_csv(secdb)
+    chart.main_per_cwe_spec_chart(reports, cwe, df_sec)
 
 def comparison(secdb, regdb, reports):
     df_sec = pd.read_csv(secdb)
@@ -113,14 +117,16 @@ def comparison(secdb, regdb, reports):
 if __name__ == "__main__":
     
     parser = argparse.ArgumentParser(description='Report results')    
-    parser.add_argument('--report', dest='goal', choices=['export', 'comparison', 'severity', 'guideline', 'language', 'cwe'],
+    parser.add_argument('--report', dest='goal', choices=['export', 'comparison', 'severity', 'guideline', 'language', 'cwe', 'cwe-spec'],
                         help='choose a report goal')
                         
     parser.add_argument('-results', type=str, metavar='folder path', help='results folder path')  
     parser.add_argument('-reports', type=str, metavar='folder path', help='reports folder path')      
     parser.add_argument('-secdb', type=str, metavar='file path', help='security dataset path')    
     parser.add_argument('-regdb', type=str, metavar='file path', help='regular dataset path')    
-    parser.add_argument('-cache', type=str, metavar='file path', help='cache path')    
+    parser.add_argument('-cache', type=str, metavar='file path', help='cache path')   
+    parser.add_argument('-cwe', type=str, metavar='file path', help='cache path')    
+     
     
     args = parser.parse_args()
 
@@ -142,5 +148,8 @@ if __name__ == "__main__":
     elif args.goal == 'cwe':
         if args.secdb != None and args.reports != None:
             cwe(secdb=args.secdb, reports=args.reports)
+    elif args.goal == 'cwe-spec':
+        if args.secdb != None and args.reports != None and args.cwe != None:
+            cwe_spec(secdb=args.secdb, reports=args.reports, cwe=args.cwe)
     else:
         print('Something is wrong. Verify your parameters')

### 2020-04-15 scripts/stats/chart.py

@@ -9,19 +9,18 @@ from . import tests
 from . import data
 
 
-_BAR_WIDTH = 0.25
-
+BAR_WIDTH = 0.25
 
 def set_bars(df, idx):
-    plt.barh(idx, df['pos'], _BAR_WIDTH, alpha=0.7, align='center', color='green', label='Positive')
-    plt.barh(idx + _BAR_WIDTH, df['nul'], _BAR_WIDTH, alpha=0.7, align='center', color='orange', label='None')
-    plt.barh(idx + 2*_BAR_WIDTH, df['neg'], _BAR_WIDTH, alpha=0.7, align='center', color='red', label='Negative')
+    plt.barh(idx, df['pos'], BAR_WIDTH, alpha=0.7, align='center', color='green', label='Positive')
+    plt.barh(idx + BAR_WIDTH, df['nul'], BAR_WIDTH, alpha=0.7, align='center', color='orange', label='None')
+    plt.barh(idx + 2*BAR_WIDTH, df['neg'], BAR_WIDTH, alpha=0.7, align='center', color='red', label='Negative')
 
 def format_p_value(p):
     return '<0.001' if float('{:.{}f}'.format(p, 3)) <= 0.000 else '={:.{}f}'.format(p, 3)
 
 def set_ticks(df, idx, fontsize=7):
-    plt.yticks(idx + _BAR_WIDTH, df['type'], fontsize=fontsize)
+    plt.yticks(idx + BAR_WIDTH, df['type'], fontsize=fontsize)
     plt.xticks(fontsize=fontsize)
     plt.gca().set_xticklabels(['{:.0f}%'.format(x*100) for x in plt.gca().get_xticks()])
 
@@ -37,7 +36,10 @@ def save_report(path, name):
 def main_comparison_chart(reports, df_sec, df_reg):    
 
         test = (tests.hypothesis_test(df_sec['diff']), tests.hypothesis_test(df_reg['diff']))
+        print(test)
         total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)
+        print(total_sec)
+        print(total_reg)
         
         db_size = len(df_sec)        
 
@@ -63,16 +65,61 @@ def main_comparison_chart(reports, df_sec, df_reg):
         
         save_report(reports, 'main_comparison.pdf')
 
+
+def main_per_cwe_spec_chart(reports, cwe, df, wilcoxon = True):
+    
+    composites = enum.read_cwe_composites('stats/{}'.format(cwe))
[…]

### 2020-04-15 scripts/stats/enum.py

@@ -37,8 +37,7 @@ def read_cwe_composites(file):
                 composites[cwes[0]] = cwes[1::]
     return composites
 
-def check_if_belongs_to_cwe(key):
-    composites = read_cwe_composites('stats/cwe')
+def check_if_belongs_to_cwe(composites, key):
     if key in composites.keys():
         return key
     for i in composites:

### 2020-04-15 scripts/stats/tests.py

@@ -13,3 +13,6 @@ def filter_small_sample_groups(df, key):
 
 def filter_small_cwe_groups(df):
     return [i for i in df['CWE'].unique() if str(i) != 'nan' and (len(df[df['CWE'] == i]) > 19) and 'CWE' in i]
+
+def filter_cwe_groups(df):
+    return [i for i in df['CWE'].unique() if str(i) != 'nan' and 'CWE' in i]

### 2020-04-20 paper/icsme20.tex
paper update

@@
 language support and project size. The final dataset used in this paper comprises
-$1031$ security patches. $539$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE). In Table~\ref{tab:patterns}, the most prevailing categories are presented. To satisfy the Wilcoxon test size requirement, types of vulnerabilities
-with less than $20$ instances were merged in a bigger group named \textit{Miscellaneous}.
-
-\begin{table}[h]
-\scriptsize
-	\caption{Guidelines to produce maintainable code}
-\begin{tabular}{L{1.1cm}L{2.1cm}L{4.3cm}}
-
-\toprule
-\textbf{Pattern} & \textbf{Name} & \textbf{Description}\\
-\midrule
-
-\textbf{CWE-89} & 
- Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases) & 
- When developers do not keep untrusted data
-     separate from SQL queries. If an attacker sends a command that
-     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.\\\midrule
-
-\textbf{CWE-611} & 
- Improper Restriction of XML External Entity Reference (24 cases) & 
-	XML documents are exchanged through the web containing entities with URIs
- 	that resolve to local/external files. Thus, when XML parsers are not well configured,
- 	attackers have allowed to directly those files.\\\midrule
-
- \textbf{CWE-20} & 
- Improper Input Validation (76 cases) & 
-	When developers
- 	forget to validate the inputs of an application, attackers may have control
- 	of the control/data flow of the program.\\\midrule
-
-  \textbf{CWE-200} & 
-Information Exposure (39 cases)& 
-Application's logs are one way of intentional/unintentional disclosure information
- 	to attackers. Many times attackers get access to logs when sould not be authorized to.\\\midrule
-
-   \textbf{CWE-352} & 
-Cross-Site Request Forgery/CSRF (22 cases)& 
-Poor session tokens generation
-     and management usually allow attackers to send forged HTTP requests
-     including authentication information from the victim to the vulnerable web
-     application.\\\midrule
-
-       \textbf{CWE-264} & 
-Permissions, Privileges, and Access Controls (59 cases)& 
-Incorrectly
- 	    implemented functionalities related to authentication and session
- 	    management, allowing an attacker to gain access to session tokens,
- 	    passwords, keys, and other sensitive data.\\\midrule
-
- 	           \textbf{CWE-399} & 
-Resource Management Errors (22 cases)& 
-Resource management issues are found more frequently
- 		in programming languages that do not manage memory automatically (e.g., C/C++
- 	    and Objective-C), i.e., where developers are responsible for
- 	    handling it instead. It is one of the main causes of Denial-of-Service attacks.\\\midrule
-
- 	     	           \textbf{CWE-79} & 
-Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (63 cases)& 
-Lack of proper validation or escaping
- 	    allows attackers to submit untrusted data to web browsers through malicious
- 	    scripts that can hijack the user sessions or redirect the user to malicious
- 	    sites.\\\midrule
-
- 	     	           \textbf{CWE-22} & 
-Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases) & 
-When developers forget to neutralize special elements within the path names to files/directories, attackers
- 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
- 	or entire repositories.\\\midrule
-
- 	 	     	           \textbf{MISC} & 
-Miscellaneous (666 cases) & 
-This pattern comprises
- 		several other security refactorings that do not have a CWE assigned and patterns that
- 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
-\bottomrule
-\end{tabular}
-\label{tab:patterns}
-\end{table}
+$1031$ security patches. $912$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE) where $59.1\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $40.9\%$ were manually classified following the \texttt{Research Concepts} view map from the CWE's list.
+
+% In Table~\ref{tab:patterns}, the most prevailing categories are presented. To satisfy the Wilcoxon test size requirement, types of vulnerabilities
+% with less than $20$ instances were merged in a bigger group named \textit{Miscellaneous}.
+
+% \begin{table}[h]
+% \scriptsize
+% 	\caption{Guidelines to produce maintainable code}
+% \begin{tabular}{L{1.1cm}L{2.1cm}L{4.3cm}}
+%
+% \toprule
+% \textbf{Pattern} & \textbf{Name} & \textbf{Description}\\
+% \midrule
+%
+% \textbf{CWE-89} &
+%  Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases) &
+%  When developers do not keep untrusted data
+%      separate from SQL queries. If an attacker sends a command that
+%      exploits the syntax of the SQL interpreter, then SQL injection attack is possible.\\\midrule
+%
+% \textbf{CWE-611} &
+%  Improper Restriction of XML External Entity Reference (24 cases) &
+% 	XML documents are exchanged through the web containing entities with URIs
+%  	that resolve to local/external files. Thus, when XML parsers are not well configured,
+%  	attackers have allowed to directly those files.\\\midrule
+%
+%  \textbf{CWE-20} &
+%  Improper Input Validation (76 cases) &
+% 	When developers
+%  	forget to validate the inputs of an application, attackers may have control
+%  	of the control/data flow of the program.\\\midrule
+%
+%   \textbf{CWE-200} &
+% Information Exposure (39 cases)&
+% Application's logs are one way of intentional/unintentional disclosure information
+%  	to attackers. Many times attackers get access to logs when sould not be authorized to.\\\midrule
+%
+%    \textbf{CWE-352} &
+% Cross-Site Request Forgery/CSRF (22 cases)&
+% Poor session tokens generation
+%      and management usually allow attackers to send forged HTTP requests
+%      including authentication information from the victim to the vulnerable web
+%      application.\\\midrule
+%
+%        \textbf{CWE-264} &
+% Permissions, Privileges, and Access Controls (59 cases)&
+% Incorrectly
+%  	    implemented functionalities related to authentication and session
+%  	    management, allowing an attacker to gain access to session tokens,
+%  	    passwords, keys, and other sensitive data.\\\midrule
+%
+%  	           \textbf{CWE-399} &
+% Resource Management Errors (22 cases)&
+% Resource management issues are found more frequently
+%  		in programming languages that do not manage memory automatically (e.g., C/C++
+%  	    and Objective-C), i.e., where developers are responsible for
+%  	    handling it instead. It is one of the main causes of Denial-of-Service attacks.\\\midrule
+%
+%  	     	           \textbf{CWE-79} &
+% Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (63 cases)&
+% Lack of proper validation or escaping
+%  	    allows attackers to submit untrusted data to web browsers through malicious
+%  	    scripts that can hijack the user sessions or redirect the user to malicious
+%  	    sites.\\\midrule
+%
+%  	     	           \textbf{CWE-22} &
+% Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases) &
+% When developers forget to neutralize special elements within the path names to files/directories, attackers
+%  	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
+%  	or entire repositories.\\\midrule
+%
+%  	 	     	           \textbf{MISC} &
+% Miscellaneous (666 cases) &
+% This pattern comprises
+%  		several other security refactorings that do not have a CWE assigned and patterns that
+%  		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
+% \bottomrule
+% \end{tabular}
+% \label{tab:patterns}
+% \end{table}
 
@@
 test that incorporates the cases where maintainability is different from zero~\cite{10.2307/2282543}.
+The Wilcoxon test requires a distribution size of at leat $20$ instances.
 To understand the effect-size, as
@@
  	\centering
- 	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
+	\vspace{-0.3 cm}
+ 	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
+	\vspace{-0.3 cm}
  	\caption{Impact of the security patches per guideline.}
 	\label{fig:guidelines}
+	\vspace{0.2 cm}
 	\scalebox{0.5}{
@@
 by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
-guideline individually. 
-For each type of
+guideline individually. For each type of
 guideline, three types of impact of patches in the software maintainability
@@
 \emph{positive}, maintainability increases (\emph{green}). Next to each
-type of guideline, it is shown the mean ($\overline{x}$) and median (M)
+type of guideline, it is presented the mean ($\overline{x}$) and median (M)
 of the maintainability difference and the p-value resulting from the
@@
 Regarding the impact of security patches for each guideline, we observe
-that overall patches have a significant positive impact in $5$ guidelines: 
+that overall patches have a significant positive impact in $6$ guidelines: 
 \emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
@@
 $38.3\%$/$397$ of negative cases) are the most affected affected guidelines.
+
+ %%%%%%%%%%% HERE
 

### 2020-04-29 scripts/maintainability/better_code_hub.py
2nd version of the package

@@ -64,6 +64,7 @@ def external_analyze_commit(user, project, commit_sha):
     """Analyze project commit without write privileges."""
     try:
         forked_repo = ghutils.git_fork(user, project)
+        log.info(forked_repo)
         log.info(f"Repo {user}/{project} Forked")
         result = analyze_commit(forked_repo.owner.login, project, commit_sha)
     except ProjectNotSupported:
@@ -150,14 +151,6 @@ def collect_report(user, project):
         raise StillProcessing
     raise BetterCodeHubException
 
-def compute_maintainability_score_old(report):
-    """Compute maintainability score."""
-    guidelines_scores = [
-        1 - guideline.get('percentageOfNonCompliantCode')
-        for guideline in report.get("analysisResults")
-    ]
-    return sum(guidelines_scores)/len(guidelines_scores)
-
 def compute_maintainability_score(report):
     """Compute maintainability score."""
     total_loc = get_project_loc(report)
@@ -182,7 +175,6 @@ def _compute_maintainability_for_guideline(guideline, total_loc):
         return 0
     if guideline['guideline'] == "Keep Your Codebase Small":
         return 0
-        
     volumes = guideline.get('qualityProfileVolume')
     if guideline['guideline'] == "Keep Architecture Components Balanced":
         total_compoments = sum(volumes)

### 2020-04-29 scripts/maintainability/eval_maintainability.py

@@ -13,7 +13,7 @@ import maintainability.better_code_hub as bch
 RETRY_ATTEMPTS = 10
 
 @click.command()
-@click.option('--dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
+@click.option('--dataset', prompt=True, default="../dataset/reg.csv",
               help="Dataset with commits for comparison.")
 @click.option('--commit', prompt=True, default="security",
               help="type of dataset")
@@ -27,15 +27,13 @@ def tool(dataset, commit):
 @timer()
 def collect_maintainability(rows, commit, dataset):
     """Collect maintainability of regular/random commits"""
-    for i, row in rows.iterrows():        
+    for i, row in rows[::-1].iterrows():        
         user = row['owner']
         project = row['project']
         
         if commit == 'regular':            
             commit_sha = row['sha-reg']
             parent_commit_sha = row['sha-reg-p']
-            if row['reg'] != 'FOUND':
-                continue
         else:
             commit_sha = row['sha']
             parent_commit_sha = row['sha-p']

### 2020-04-29 scripts/report.py

@@ -78,14 +78,9 @@ def export(secdb, regdb, results, cache_path):
     cache = bch.BCHCache(cache_path)
     
     df_sec, sec_res_path = main_calculation_by_db(secdb, results, cache, 'security')
-    df_reg, reg_res_path = main_calculation_by_db(regdb, results, cache, 'regular')
-    
-    ids = df_reg[~df_reg['diff'].notnull()].index
-    
-    df_reg = df_reg.drop(ids) 
-    df_sec = df_sec.drop(ids)
-        
     df_sec.to_csv(sec_res_path, index=False)
+    
+    df_reg, reg_res_path = main_calculation_by_db(regdb, results, cache, 'regular')
     df_reg.to_csv(reg_res_path, index=False)
         
 def language(secdb, reports):
@@ -127,7 +122,6 @@ if __name__ == "__main__":
     parser.add_argument('-cache', type=str, metavar='file path', help='cache path')   
     parser.add_argument('-cwe', type=str, metavar='file path', help='cache path')    
      
-    
     args = parser.parse_args()
 
     if args.goal == 'export':  

### 2020-04-29 scripts/stats/chart.py

@@ -35,36 +35,38 @@ def save_report(path, name):
 
 def main_comparison_chart(reports, df_sec, df_reg):    
 
-        test = (tests.hypothesis_test(df_sec['diff']), tests.hypothesis_test(df_reg['diff']))
-        print(test)
-        total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)
-        print(total_sec)
-        print(total_reg)
-        
-        db_size = len(df_sec)        
+        test = pd.concat((tests.hypothesis_test(df_reg['diff'], 'reg'), 
+                            tests.hypothesis_test(df_sec['diff'], 'sec')),
+                            ignore_index=True)
+                        
+        total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)        
+        db_size = len(df_sec)  
 
-        d = {'neg' : pd.Series([total_reg['neg']/db_size, total_sec['neg']/db_size]),
-            'pos' : pd.Series([total_reg['pos']/db_size, total_sec['pos']/db_size]),
-            'nul' : pd.Series([total_reg['nul']/db_size, total_sec['nul']/db_size]),
-            'type' : pd.Series(['Regular Change','Security Change'])}
+        rep = {'neg' : pd.Series([total_reg['neg']/db_size, total_sec['neg']/db_size]),
+                'neg_abs' : pd.Series([total_reg['neg'], total_sec['neg']]),
+                'pos' : pd.Series([total_reg['pos']/db_size, total_sec['pos']/db_size]),
+                'pos_abs' : pd.Series([total_reg['pos'], total_sec['pos']]),
+                'nul' : pd.Series([total_reg['nul']/db_size, total_sec['nul']/db_size]),
+                'nul_abs' : pd.Series([total_reg['nul'], total_sec['nul']]),
+                'type' : pd.Series(['Regular Change','Security Change'])}
         
-        df, idx, fig, ax = config_report(d, len(d['type']), 7, 6, 111)
+        df, idx, fig, ax = config_report(rep, len(rep['type']), 7, 6, 111)
         
         set_bars(df, idx)
-        set_ticks(df, idx, 9)
+        set_ticks(df, idx, 9) 
         
[…]

### 2020-04-29 scripts/stats/tests.py

@@ -1,9 +1,9 @@
 import pandas as pd
 from scipy.stats import wilcoxon
 
-def hypothesis_test(diff):
+def hypothesis_test(diff, i):
     test, pvalue = wilcoxon(x=diff, zero_method="pratt")
-    return pd.DataFrame({'test': [test], 
+    return pd.DataFrame({'guideline':[i], 'test': [test], 
                         'med':[pd.DataFrame({'diff': diff})['diff'].median()],
                         'pvalue': [pvalue], 
                         'mean':[pd.DataFrame({'diff': diff})['diff'].mean()]})

### 2020-04-29 paper/icsme20.tex
intro

@@
 %\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
-%\title{The reason why you should be using coding practices: safer software.}
+%\title{The reason why you should be applying coding practices: safer software.}
 %\title{What is the impact of security patches in software evolution?}
-%\title{Hey, security engineers! You should pay more attention to your patches.}
+%\title{Hey, developers! You should pay more attention to your patches.}
 
@@
 These vulnerabilities are gateways for attackers to penetrate systems and 
-exploit their resources. \Sof{Update example for 2020 - outdated}As an example, in the last week of January $2019$, 
-a severe vulnerability in Apple FaceTime was revealed\footnote{FaceTime vulnerability 
-details available at \url{https://thehackernews.com/2019/01/apple-facetime-privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
-flaw that allowed callers to see and hear others without them picking up the call.\Sof{---}%\cite{cacheOut}.
-%
+exploit their resources. As an example, at the beginning of $2020$, academics published a 
+new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
+security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
+address in previous generations and that makes every user using a CPU released before Q$4$ $2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
+at the cost of features and/or performance.
+%
+
 In $2011$, the International Organization for Standardization (ISO) issued an
@@
 detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
-%
 Despite these efforts, improving software security is not a trivial task and requires 
@@
 %
-As ISO does not provide any specific guidelines or formulas to calculate 
+
+As ISO does not provide any specific guidelines/formulas to calculate 
 maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
@@
 with a set of $10$ guidelines/metrics to produce quality software 
-based in ISO/IEC 25010~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
+based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
 were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
@@
 software mainly due to the broad number of metrics used.
+
 From a methodological 
@@
 impact. This empirical study presents evidence that changes 
-applied in the code bases to patch vulnerabilities affect code maintainability. 
-Especially, for patches that address Path Traversal (CWE-22),
-Information Exposure (CWE-200) and Resource Management Errors (CWE-399). This is important work
+applied in the code bases to patch vulnerabilities affect code maintainability.\Sof{fix}Especially, for patches that address Path Traversal (CWE-22),
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399). \Sof{-END-}This is important work
 because there is still little information about the impact of security patches on the software
@@
 %
+
 The main contributions of the present work are:
@@
 %%%%% NEEDS REVIEW
-\Sof{Improve for 2nd version}
 As an example, consider the patch of the TLS state-machine protocol implementation in 
@@
 corresponding to large length values. This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$ \footnote{CVE-$2016$-$6307$
+Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$~\footnote{CVE-$2016$-$6307$
 details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
+research. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
 \emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
@@
 CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
-memory for a message is allocated too early, prior to the excessive
+the memory for a message is allocated too early, prior to the excessive
 message length check which can lead to a Denial of Service through memory
@@
 are multiple connections in a state where memory has been allocated for
-the connection, \texttt{SSL\_free()} has not yet been called, and there is
+the connection, when \texttt{SSL\_free()} has not yet been called, and there is
 insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
 be transitory because as soon as the connection fails the memory is
-subsequently freed again in the \texttt{SSL\_free()} call. However there is an
+subsequently freed again in the \texttt{SSL\_free()} call. However, there is an
 increased risk during this period of application crashes due to the lack
@@
 \\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
- static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
+ static SUB_STATE_RETURN read_state_machine(SSL *s){ 
  // [snip]
@@
 \\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
- int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
+ int tls_get_message_header(SSL *s, int *mt){   
  // [snip]
@@
         SSL3_HM_HEADER_LENGTH;
-- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
+- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func3}!*/
 -      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
@@
 
-- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func5}!*/  
+- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func4}!*/
 -     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
@@
 
-~\ref{lst:func1} \texttt{read\_state\_machine} function abstracts a sub-state-machine
+The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
 to handle the messages exchanged between peers. When messages are larger than
 what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
-of this resource was being verified very early in the application's flow.  Listing ~\ref{lst:vuln2} presents those verifications (lines $5$-$7$ and line $11$-$13$).
-These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}.
+of this resource was being verified very early in the application's flow, using the \texttt{tls\_get\_message\_header} function presented in Listing ~\ref{lst:vuln2} (\ref{lst:func3} and \ref{lst:func4}).
+These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).
 
@@
 between pairs of versions $v_{s-1}$, $v_s$ (i.e., before and after a
-security-commit) come from the same distribution. Nevertheless, this test has a
-limitation: it does not consider the groups of commits with a zero-difference
-maintainability. In $1959$, Pratt provided an improvement to the test to solve
-this issue making the test more robust. Thus, we use a version of the Wilcoxon
-test that incorporates the cases where maintainability is different from zero~\cite{10.2307/2282543}.
+security-commit) come from the same distribution. Nevertheless, this test 
+has a limitation: it does not consider the groups of commits with a 
+zero-difference maintainability. In $1959$, Pratt provided an improvement to 
+the test to solve this issue making the test more robust. Thus, we use a 
+version of the Wilcoxon test that incorporates the cases where 
+maintainability is different from zero~\cite{10.2307/2282543}.
 The Wilcoxon test requires a distribution size of at leat $20$ instances.
-To understand the effect-size, as
-advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
-mean difference, the median of the difference, and the percentage of cases that
-reduce maintainability.
+To understand the effect-size, as advocated by the Common-language effect 
+sizes\cite{graw:1992}, we compute the mean difference, the median of the 
+difference, and the percentage of cases that reduce maintainability.
 %
@@
 
-This study evaluates $1031$ security and baseline commits
-from $265$ distinct open-source projects. The following section presents the
-results obtained for each research question.
+This study evaluates $1031$ security and baseline commits from $265$ 
+distinct open-source projects. The following section presents the results 
+obtained for each research question.
 
@@
 Maintainability is being calculated as the mean of the guidelines computed
-by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
-guideline individually. For each type of
-guideline, three types of impact of patches in the software maintainability
-are presented: \emph{negative}, maintainability decreases (\emph{red});
+by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on 
+each guideline individually. For each type of guideline, three types of 
+impact of patches in the software maintainability are presented: 
+\emph{negative}, maintainability decreases (\emph{red});
 \emph{none}, maintainability remains the same (\emph{yellow}); and,
@@
 that overall patches have a significant positive impact in $6$ guidelines: 
-\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
-Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
-Components Loosely} (p-value = $2.43$x$10^{-07}$), \emph{Keep Unit Interfaces 
-Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
-$7.9$x$10^{-35}$). However, for each of these guidelines we also observe a 
+\emph{Write Clean Code} (p-value = $5.49$x$10^{-54}$), \emph{Keep 
+Architecture Components Balanced} (p-value = $5.30$x$10^{-71}$), 
+\emph{Couple Architecture Components Loosely} (p-value = $2.24$x$10^{-07}$), 
+\emph{Separate Concerns in Modules} (p-value = $0.399$), \emph{Keep Unit 
+Interfaces Small} (p-value = $1.94$x$10^{-13}$), \emph{Write Code Once} 
+(p-value = $1.03$x$10^{-34}$).
+
+However, for each guideline we also observe a 
 substantial number of cases where the maintainability suffers a negative 
-impact $-$ from $10\%$ to $35\%$ $-$ which highlights the need of better 
-practices/techniques/tools to produce software. \emph{Separate Concerns in Module} 
-guideline also follows the same results but without any statistical 
-significance since the p-value is higher than $0.05$. According to the 
+impact $-$ from $10\%$ to $40\%$ $-$ which highlights the need of better 
+practices/techniques/tools to produce software. \emph{Separate Concerns in 
+Module} guideline does not the p-value is higher than $0.05$.
+
+According to the 
 results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 

### 2020-04-29 scripts/stats/chart.py
chart update

@@ -132,7 +132,7 @@ def main_per_cwe_chart(reports, df, wilcoxon = True):
         'N': pd.Series([sum(results[i]) for i in cwes]),
         'type': pd.Series(cwes)}
 
-    df, idx, fig, ax = config_report(rep, len(rep['type']), 5, 10, 111)
+    df, idx, fig, ax = config_report(rep, len(rep['type']), 5, 8, 111)
     
     set_bars(df, idx)
     set_ticks(df, idx, 9)

### 2020-04-29 paper/icsme20.tex
results

@@
 \usepackage{hyperref}
-
 \usepackage{array}
+\usepackage{subfigure}
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
@@
 Maintainability is measured based on the Better Code Hub’s model of $10$ 
-guidelines using a dataset including $1335$ security commits
-\Sof{These are the initial values, should I use the final? In the end, 
-we studied 1031 commits and 8 guidelines}. 
-Results show 
-that fixing software vulnerabilities may hinder the maintainability of 
+guidelines using a dataset including $1335$ security commits.
+Results show that fixing software vulnerabilities may hinder the maintainability of 
 open-source software. This is particularly the case for patches that 
@@
 %
-\section{Results}\label{sec:results}
-
-This study evaluates $1031$ security and baseline commits from $265$ 
+\section{Results \& Discussion}\label{sec:results}
+
+This study evaluates a final total of $998$ security and baseline commits from $262$ 
 distinct open-source projects. The following section presents the results 
@@
 maintainability of open-source software?}}
+
+Maintainability is being calculated as the mean of the guidelines computed
+by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on 
+each guideline individually. For each type of guideline, three types of 
+impact of patches in the software maintainability are presented: 
+\emph{negative}, maintainability decreases (\emph{red});
+\emph{none}, maintainability remains the same (\emph{yellow}); and,
+\emph{positive}, maintainability increases (\emph{green}). Next to each
+type of guideline, it is presented the mean ($\overline{x}$) and median (M)
+of the maintainability difference and the p-value resulting from the
+Paired Wilcoxon signed-rank test.
+
+Regarding the impact of security patches for each guideline, we observe
+that overall patches have a significant positive impact in $6$ guidelines: 
+\emph{Write Clean Code} (p-value = $3.66$x$10^{-51}$), \emph{Keep 
+Architecture Components Balanced} (p-value = $5.07$x$10^{-73}$), 
+\emph{Couple Architecture Components Loosely} (p-value = $1.61$x$10^{-08}$), 
+\emph{Separate Concerns in Modules} (p-value = $0.090$), \emph{Keep Unit 
+Interfaces Small} (p-value = $8.76$x$10^{-15}$), \emph{Write Code Once} 
+(p-value = $4.97$x$10^{-35}$). Overall, for each of the $8$ guidelines we also 
+observe a substantial number of cases where the maintainability suffers a 
+negative impact $-$ from $10\%$ to $40\%$ $-$ which highlights the need of 
+better practices/techniques/tools to produce software. However, the results do 
+not provide statistical significance to the \emph{Separate 
+Concerns in Modules} guideline - results should be read carefully.
+
+According to the results, the \emph{Write Simple Units of Code} 
+(p-value=$0.037$ and $372$ ($37.3\%$) of negative cases) and \emph{Write Short 
+Units of Code} (p-value=$0.028$ and $375$ ($37.6\%$) of negative cases) are 
+the most affected guidelines.
+
+ For projects with large 
+ codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
+ viable because they retrieved values above the limit set by BCH ($20$ Person-years).
+ The \emph{Automated Tests} guideline was also not considered because the BCH does not return
+ the amount of lines of code that complies with the guideline -- which is one of the variables
+ our formula uses to calculate the maintainability of the project regarding each guideline. 
+ Thus, we decided to not consider both guidelines in our analysis.
 
@@
  	\vspace{-0.3 cm}
- 	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
+  	\includegraphics[width=0.35\textwidth]{figures/main_per_guideline.pdf}
  	\vspace{-0.3 cm}
@@
 
-Maintainability is being calculated as the mean of the guidelines computed
-by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on 
-each guideline individually. For each type of guideline, three types of 
-impact of patches in the software maintainability are presented: 
-\emph{negative}, maintainability decreases (\emph{red});
-\emph{none}, maintainability remains the same (\emph{yellow}); and,
-\emph{positive}, maintainability increases (\emph{green}). Next to each
-type of guideline, it is presented the mean ($\overline{x}$) and median (M)
-of the maintainability difference and the p-value resulting from the
-Paired Wilcoxon signed-rank test.
-
-Regarding the impact of security patches for each guideline, we observe
-that overall patches have a significant positive impact in $6$ guidelines: 
-\emph{Write Clean Code} (p-value = $5.49$x$10^{-54}$), \emph{Keep 
-Architecture Components Balanced} (p-value = $5.30$x$10^{-71}$), 
-\emph{Couple Architecture Components Loosely} (p-value = $2.24$x$10^{-07}$), 
-\emph{Separate Concerns in Modules} (p-value = $0.399$), \emph{Keep Unit 
-Interfaces Small} (p-value = $1.94$x$10^{-13}$), \emph{Write Code Once} 
-(p-value = $1.03$x$10^{-34}$).
-
-However, for each guideline we also observe a 
-substantial number of cases where the maintainability suffers a negative 
-impact $-$ from $10\%$ to $40\%$ $-$ which highlights the need of better 
-practices/techniques/tools to produce software. \emph{Separate Concerns in 
-Module} guideline does not the p-value is higher than $0.05$.
-
-According to the 
-results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 
-of negative cases) and \emph{Write Short Units of Code} (p-value=$0.029$ and 
-$38.3\%$/$397$ of negative cases) are the most affected affected guidelines.
-
- %%%%%%%%%%% HERE
+Some of the vulnerabilities are identified with \emph{Common Vulnerabilities 
+and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
+Database} (NVD) website to obtain their severity levels. In total, we 
+retrieved severity scores for $551$ vulnerabilities: $117$ (\emph{HIGH}), 
+$404$ (\emph{MEDIUM}) and $30$ (\emph{LOW}). Figure~\ref{fig:severity} 
+presents the impact of security patches per severity level on the 
+maintainability of open-source software. We observe that patches for 
+\emph{HIGH} ($47.9\%$) and \emph{MEDIUM} ($44.1\%$) severity vulnerabilities 
+hinder more the maintainability of software than \emph{LOW} ($30.0\%$) 
+severity vulnerabilities. Again, patches have a considerable 
+negative impact on software maintainability - between $30\%$ 
+and $50\%$. None statistical significance was retrieved. However, 
+results should not be disregarded because they seem to confirm the 
+assumption that higher severity vulnerabilities patches may have a 
+more negative impact on maintainability.
 
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.45\textwidth]{figures/main_per_severity.pdf}
- 	\caption{Maintainability difference by the vulnerabilities severity.}
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
+ 	\caption{Maintainability difference by vulnerability severity.}
 	\label{fig:severity}
@@
 
-For projects with large 
-codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
-viable because they retrieved values above the limit set by BCH ($20$ Person-years).
-The \emph{Automated Tests} guideline was also not considered because the BCH does not return
-the amount of lines of code that complies with the guideline -- which is one of the variables
-our formula uses to calculate the maintainability of the project regarding each guideline. 
-Thus, we decided to not consider both guidelines in our analysis.
-
-For the vulnerabilities identified with CVE(s), we scraped the National Vulnerability Database (NVD) Website to retrieve 
-their severity levels. In total, we obtained severity scores for $574$ 
-vulnerabilities: $121$ \emph{HIGH}, $423$ \emph{MEDIUM} and $30$ \emph{LOW}. 
-Figure~\ref{fig:severity} presents the impact
-of security patches per severity level on the maintainability of open-source 
-software. Although with no statistical significance, we observe that 
-vulnerabilities with higher severity levels (\emph{HIGH} and \emph{MEDIUM}) have 
-more negative impact than the \emph{LOW} severity level on the
-maintainability of software.
+We have also evaluated the impact on the maintainability per programming 
+language of open-source software (Figure~\ref{fig:lang_main}). We restrict 
+this analysis to programming languages with at least $20$ data points, as this 
+is a requirement for our hypothesis tests. Thus, we compare the results for \texttt{C/C++}, 
+\texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, \texttt{Python} and \texttt{PHP}. 
+The remaining languages are included into \texttt{Other}.
+
+\texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and \texttt{PHP} ($38.27\%$) are
+the programming languages with a more negative impact on the maintainability 
+of software. \texttt{Java} seems to be the language less affected
+by patching, more than $50\%$ of patches had a positive impact on maintainbility. 
+But overall patches in all languages have a considerable
+negative impact on maintainability - from $30\%$ to $50\%$ - which confirms
+the need for better and more secure programming languages. 
+Statistical significance was only retrieved for the \texttt{C/C++} 
+(p-value = $2.24$x$10^{-05}$) and \texttt{Ruby} (p-value = $0.041$) languages.
+Results should be read with care where statistical significance was not found.
+Yet, data reports very interesting hints on the impact of programming languages on 
+security patches.
 
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/main_per_language.pdf}
-  \caption{Maintainability difference by language after security patches.}
+  \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
+  \caption{Maintainability difference by programming language.}
   \label{fig:lang_main}  
@@
 
-We have also evaluated the impact on the maintainability per programming language
-of open-source software, Figure~\ref{fig:lang_main}. We restrict this analysis to 
-programming languages with at least $20$ data points, as this is a requirement 
-for our hypothesis tests. Thus, we compare the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
-\emph{Objective-C/C++}, and \emph{Java}. The other languages were
-integrated into \emph{Others}. The results of the Paired Wilcoxon signed-rank
-test by language indicate that there is statistical evidence to support that the
-maintainability of open-source software decreases for security patches for \emph{C/C++} 
-(p-value = $2.5$x$10^{-5}$). 
 % Following the \textit{American Statistical Association}
@@
 % Ruby as statistically significant because its p-value is on the borderline.
-\emph{Ruby} and \emph{PHP} languages also impact negatively the maintainability of software. 
-Despite not having statistical significance, there are a considerable number of \emph{Ruby} 
-and \emph{PHP} patches hindering maintainability that should not be disregarded.
-
-Overall, it is important to notice that there is a very significant number of patches that
-hinder the maintainability of software $-$ even if the number of negative cases does not 
-surpass the number of positive or none $-$ for  $3$ different languages: \emph{Python},
-\emph{Objective-C/C++} and \emph{Java}. For those languages, the negative impact is between $35\%$ and $50\%$ which highlights the need for better and secure programming languages. 
-Results should be read with care for the languages where statistical significance was not found. However, 
-data reports very interesting hints about the impact of programming languages in 
-security patches.
-
-Regarding security commits results, we observe that maintainability decreases in
-$41.61\%$ ($429$), remains the same in $19.69\%$ ($203$) and increases in $38.70\%$
-($399$) of the cases after security patches are applied. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $4.9$x$10^{-2}$.
+
+Regarding security commits results (Figure~\ref{fig:secvsreg}), we observe that
+maintainability decreases in $41.28\%$ ($412$), remains the same in $20.64\%$ ($206$)
+and increases in $38.08\%$ ($380$) of the cases. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $0.044$.
 Since the p-value is below the significance level of $0.05$, we argue that
@@
 open-source software}.
-
-\subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
-affect open-source software maintainability?}}}
-
-The impact of each type of security patches on the maintainability of
-open-source software is presented in Figure~\ref{fig:pat}.
-Results demonstrate evidence that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.
-
-We also observe that maintainability is less
-negatively impacted when patching \emph{SQL Injection (CWE-89)}, \emph{Improper Restriction of XML External Entity Reference (CWE-611)}, \emph{Improper Input Validation (CWE-20)}, \emph{Cross-Site Request Forgery (CWE-352)}, \emph{Permissions, Privileges, and Access Controls (CWE-264)}.
-However,
-for all the previous patterns there is a percentage of security patches that hinder maintainability between $30\%$ and $50\%$ which is very significant.
-\emph{Cross-Site Scripting (CWE-79)} is the pattern that suffers less positive/negative impact from security patches, i.e., security
-patches of this type of vulnerability typically have no impact on the maintainability of open-source software.
-
-Software with \emph{Resource Management Errors (CWE-399)} flaws are the most affected ones by security patches: $54.5\%$ ($12$) of the patches had a negative impact
-in the maintainability while $36.4\%$ ($8$) yielded a positive impact, and only
-$9.1\%$ ($2$) did not suffer any change. The results also show that fixing
-\emph{Information Exposure (CWE-200)} vulnerabilities affects the maintainability of $51.2\%$ ($20$) of the cases
-and leaves only $2.6\%$ ($1$) of the cases unaffected. In $46.2\%$ ($18$) of the
-cases, maintainability increased. 
-Patching \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)} vulnerabilities has a negative
-impact of $48.6\%$ ($18$) in the maintainability of open-source software while
-$46.0\%$ ($17$) of the cases exhibit maintainability improvement and $5.4\%$
-($2$) show that maintainability does not suffer any change. The
-\emph{Miscellaneous} patches pattern integrates several types of security
-flaws. These type of patches harm maintainability in $41.3\%$ ($673$) of the
-cases. In $22.7\%$ ($153$), there is no impact and for the
-other $36.0\%$ ($242$) maintainability increases.
-It is also important to notice that overall the maintainability of open-source
-software is not affected by \emph{Cross-Site Scripting (CWE-79)} vulnerabilities. \emph{Cross-Site Scripting (CWE-79)}
-does not suffer any decrease in the maintainability of $38.7\%$ ($24$) of the cases
-and $35.5\%$ ($22$) have a negative effect on it. Only $25.8\%$ ($16$) of the
-cases improve the software maintainability.
-
-The results of the
-Paired Wilcoxon signed-rank per pattern exhibited statistical
-evidence supporting that the maintainability of open-source software decreases for
-only one category: \emph{Miscellaneous} (p-value = $0.033$). However, no statistical
-significance was found to the other patterns. Thus, results should be read with care.
+%
+
+\textit{\textbf{RQ2: Which weaknesses are more likely to
+affect open-source software maintainability?}}
+
+\begin{figure*}[htp]
+  \centering
+  \subfigure[Maintainability difference by first-level weaknesses from the 
+  \textit{Research Concepts} list on Common Weakness Enumeration (CWE)]{\includegraphics[scale=0.4]{figures/main_per_cwe.pdf}}\quad\quad
+  \subfigure[Maintainability difference by sub-weaknesses of the 
+  \textit{Improper Neutralization} Weakness (CWE-707)]{\includegraphics[scale=0.4]{figures/main_per_cwe_spec_707.pdf}}\quad\quad
+    \subfigure[Maintainability difference by sub-weaknesses of the 
+	\textit{Improper Control of a Resource Through its Lifetime} Weakness (CWE-664)]{\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
+	\label{fig:pat}
+\end{figure*}
+
+\emph{Weakness}, according to Common Weakness Enumeration (CWE) glossary, 
+is a type of mistake that, in proper conditions, could contribute to the 
+introduction of vulnerabilities within that product. This enumeration
+system was used to classify the vulnerabilities included in our study.
+Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
+the impact of all patches grouped using the first-level weaknesses from
+the \emph{Research Concepts} list on CWE; and, the remaining charts
+show two of those first-level weaknesses more in depth reporting the impact 
+on maintainability of their sub-weaknesses (\emph{(b) and (c)}).
+
+Considering the first-level weaknesses from the \emph{Research Concepts}
+list (Figure~\ref{fig:pat}-\emph{a}), we observe that patching vulnerabilities
+have negative impact in $5$ weaknesses: Insufficient Control Flow Management 
+(CWE-691), Protection Mechanism Failure (CWE-693), Improper Access Control (CWE-284), 
+Improper Neutralization (CWE-707) and Improper Control of a Resource Through its Lifetime
+(CWE-664). The CWE-707 ($301$) and CWE-664 ($321$) weaknesses integrate a 
+larger amount of cases compared to the remaining weaknesses.
+Thus, we present an analysis of their sub-weaknesses on Figure~\ref{fig:pat}-\emph{b} and
+Figure~\ref{fig:pat}-\emph{c}, respectively. Results demonstrate evidence that patching vulnerabilities hinders 
+the maintainability of open-source software for $5$ different weaknesses: \emph{Improper Input Validation 
+(CWE-20)}, \emph{Information Exposure (CWE-200)}, \emph{Improper Restriction of Operations within the Bounds of 
+a Memory Buffer (CWE-119)}, \emph{Missing Release of Memory after Effective Lifetime 
+(CWE-401)} and \emph{Path Traversal (CWE-22)}. Results also show that software maintainability
+is less negatively impacted when patching \emph{Improper Restriction of XML External Entity Reference 
+(CWE-611)}.
+
+\emph{Cross-Site Scripting (CWE-79)} is the weakness that suffers less 
+positive/negative impact from security patches, i.e., patching usually 
+has no impact on the maintainability of open-source software. This can 
+be explained by the fact that patches to this weakness usually are easily 
+done by adding an \texttt{escape} function. Patches of \emph{SQL Injection
+(CWE-89)} weaknesses equally hinder and improve the software maintainability. 
+But overall a very considerable percentage of patches hinder the open-source 
+software maintainability - between $30\%$ and $60\%$.
+
+Statistical significance was only found to the CWE-119 weakness (p-value = 
+$0.041$). However, results obtained to the other weaknesses should not be 
+disregarded because they offer an interesting view on how the maintainability 
+is impacted per weakness which can provide insights to developers/researchers
+on the weaknesses that need more attention.
+
+\textit{\textbf{RQ3: What is the impact of security patches versus regular patches on the
+maintainability of open-source software?}}
+
+The impact of security and baseline changes in the maintainability of
+open-source software is presented in Figure~\ref{fig:secvsreg}.  
+We have seen, previously, a deterioration on software maintainability 
+when patching vulnerabilities: $41.28\%$ ($412$) suffered a 
+negative impact, $20.64\%$ ($206$) remained the same and $38.08\%$ 
+($380$) increased the software maintainability. For regular changes, 
+we observe that the maintainability decreases in $27.05\%$ ($270$) 
+and increases in $30.26\%$ ($302$) of the cases. But in contrast to 
+security patches, the maintainability of regular changes remains the 
+same in $42.69\%$ ($426$) of the cases and performing regular
+changes has a more positive impact than negative on the software 
+maintainbility.
+
+Results show that regular changes typically have no impact on the 
+software maintainability which contradicts previous research.
+However, we manually inspected a considerable amount of cases from
+the regular changes distribution where this happens and found that 
+identifying regular commits as similar in size as the security-related 
+commit is limiting the type of regular commits (e.g, input 
+refactorings, variables or functions, type conversion, etc).  
+One example is the regular commit\footnote{Regular change: \url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad322448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated
+by our methodology and using as reference the following security commit
+\footnote{Security change: \url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
+overflow vulnerability patch, our methodology chose a similiar
+type of change. We strongly believe that this phenomenon lead
+that increased significantly the number of cases with no impact
+on the software maintainability.
+
+The resulting p-value of the Paired Wilcoxon signed-rank test 
+is $0.331$, i.e., conclusions regarding the regular changes 
+are not statistically significant. Yet, it is important to
+note that \emph{patching security vulnerabilities seem to have a 
+more negative impact on the maintainability of open-source 
+software than performing regular changes}.
 
 \begin{figure}[h]	
- 	\centering
- 	\includegraphics[width=0.4\textwidth]{figures/main_per_cwe.pdf}
- 	\caption{Maintainability difference by pattern after security patches.}
-	\label{fig:pat}	
-\end{figure}
-
-\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}}}
-
-The impact of security and baseline commits in the maintainability of
-open-source software is presented in Figure~\ref{fig:secvsreg}.
-
-\begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.43\textwidth]{figures/main_comparison.pdf}
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
  	\caption{Maintainability difference for security and baseline patches.}
@@
 
-We have seen previously a deterioration in maintainability when patching security vulnerabilities. 
-For regular commits, we observe that the maintainability decreases in $21.04\%$ ($133$) 
-of the cases after regular changes are applied. But in contrast
-to security commits, it increases in $26.90\%$ ($170$) 
-and remains the same in $52.06\%$ ($329$) 
-of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
-test is $5.74x10^{-78}$. The baseline is still being computed at the time of submission (it is computationally very expensive, and will be concluded soon) but it is possible to observe 
-that regular changes typically have little impact on the maintainability which contradicts research performed in the past. However,
-we manually inspected a considerable amount of cases where this happens and
-found that identifying regular commits as similar in size as the security-related commit is limiting the type of regular commits (e.g, input refactorings, renaming variables or functions)\footnote{Example of a random regular commit generated with our methodology (Accessed on \today{}):
-\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad322448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}}.
-
-%Since the p-value is below the significance level of $0.05$, 
-%we argue that \emph{regular refactorings have a
-%positive impact in the maintainability of open-source software}.
-
 \section{Discussion}\label{sec:discussion}
@@
 
-\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the maintainability
-of open-source software?}}}
-
-\textit{Security patches harm the maintainability of open-source software.}
+\textit{\textbf{RQ1: What is the impact of security patches on the maintainability
+of open-source software?}}
+\textit{Patching vulnerabilities may harm the maintainability of open-source software.}
 

### 2020-04-30 paper/icsme20.tex
more updates

@@
 patches to improve security on the maintainability of open-source software. 
-Maintainability is measured based on the Better Code Hub’s model of $10$ 
-guidelines using a dataset including $1335$ security commits.
+Maintainability is measured based on the Better Code Hub’s model of 10 
+guidelines using a dataset including 1335 security commits.
 Results show that fixing software vulnerabilities may hinder the maintainability of 
 open-source software. This is particularly the case for patches that 
-address Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource 
-Management Errors (CWE-399). We conclude that changes to code bases to patch 
+address Improper Input Validation (CWE-20), Information 
+Exposure (CWE-200), Improper Restriction of Operations within the Bounds of a 
+Memory Buffer (CWE-119), Missing Release of Memory after Effective Lifetime
+(CWE-401) and Path Traversal (CWE-22). We conclude that changes to code bases to patch 
 vulnerabilities need to be performed with extra care and that security-related 
@@
 impact. This empirical study presents evidence that changes 
-applied in the code bases to patch vulnerabilities affect code maintainability.\Sof{fix}Especially, for patches that address Path Traversal (CWE-22),
-Information Exposure (CWE-200) and Resource Management Errors (CWE-399). \Sof{-END-}This is important work
+applied in the code bases to patch vulnerabilities affect code maintainability. Especially, for patches that address \emph{Improper Input Validation} (CWE-20), \emph{Information 
+Exposure} (CWE-200), \emph{Improper Restriction of Operations within the Bounds of a 
+Memory Buffer} (CWE-119), \emph{Missing Release of Memory after Effective Lifetime} 
+(CWE-401) and \emph{Path Traversal} (CWE-22). This is important work
 because there is still little information about the impact of security patches on the software
@@
 %%%%% NEEDS REVIEW
-As an example, consider the patch of the TLS state-machine protocol implementation in 
+\Sof{As an example, consider the patch of the TLS state-machine protocol implementation in 
 OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
@@
 increased risk during this period of application crashes due to the lack
-of memory - which would then mean a more serious Denial of Service.
+of memory - which would then mean a more serious Denial of Service.}
 
@@
 
-The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
+\Sof{The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
 to handle the messages exchanged between peers. When messages are larger than
@@
 These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).
-
 %
@@
 the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
-OpenSSL maintainability.
-\Sof{-- END --}
+OpenSSL maintainability.}
+
 %%%%%%% 
@@
 language support and project size. The final dataset used in this paper comprises
-$1031$ security patches. $912$ of these patches are identified with Common Weakness
+$998$ security patches. $912$ of these patches are identified with Common Weakness
 Enumeration IDs (CWE) where $59.1\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $40.9\%$ were manually classified following the \texttt{Research Concepts} view map from the CWE's list.
@@
 
-The pair of the random commit and its parent is accepted if the \emph{diff} 
+\Sof{The pair of the random commit and its parent is accepted if the \emph{diff} 
 size fits in the range size considered acceptable at the moment. The algorithm 
@@
 security commits to ensure that differences in maintainability are not a 
-consequence of characteristics of different projects.
+consequence of characteristics of different projects.}
 
@@
 
-\Sof{Needs improvement}SIG defines \emph{Units} as the smallest groups of code that can be maintained
+\Sof{SIG defines \emph{Units} as the smallest groups of code that can be maintained
 and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and
@@
 somewhere in the green bar it means it is compliant with a low level of
-severity.\Sof{--END--}
+severity.}
 
@@
 maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
+
 \begin{itemize}
@@
 maintainability guideline for the given severity level (medium, high, and very high).
-The compliance for a severity level
+\Sof{The compliance for a severity level
 is calculated based on the number of lines of code that comply and not comply
 with the guideline at a given severity level as explained in more
-detail in previous work~\cite{cruz2019energyoriented}. In our analysis, we compute 
+detail in previous work~\cite{cruz2019energyoriented}.} In our analysis, we compute 
 the difference of maintainability between the security commit ($v_{s}$) and its parent 
@@
 %
-\\\textit{\textbf{Construct:}} The maintainability formula was inferred based on the BCH's reports. The high
-amount of different projects and backgrounds may require other
-maintainability standards. However, BCH does use a representative benchmark of
-closed and open-source software projects to compute the
-thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
+\\\textit{\textbf{Construct:}} The maintainability formula was inferred based 
+on the BCH's reports. The high amount of different projects and backgrounds 
+may require other maintainability standards. However, BCH does use a 
+representative benchmark of closed and open-source software projects to compute 
+the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
 %
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $1031$ regular commits are enough to
+Furthermore, this evaluation considers that $998$ regular commits are enough to
 alleviate random irregularities in the maintainability differences of the
@@
 should be carefully monitored by programmers since there is a high correlation
-between maintainability aspects and proneness to changes and faults.
-In 2019, Luis et al.~\cite{cruz2019energyoriented} proposed a formula to calculate maintainability based on the BCH's guidelines and measured the impact of energy-oriented fixes on the maintainability of software.
-The present
-work uses the same model (BCH) and focuses solely on evaluating the impact of
-patching vulnerabilities on software maintainability.
+between maintainability aspects and proneness to changes and faults. In 2019, Luis et 
+al.~\cite{cruz2019energyoriented} proposed a formula to calculate maintainability 
+based on the BCH's guidelines and measured the impact of energy-oriented fixes 
+on the maintainability of software.
+The present work uses the same model (BCH) and focuses solely on evaluating 
+the impact of patching vulnerabilities on software maintainability.
 
@@
 use of design patterns may introduce maintainability issues into
-software~\cite{4493325}.
-Yskout et.
-al were not able to detect if the usage of design patterns has a positive impact, but concludes that developers prefer to work with the support of security patterns~\cite{8077802}.
-The present work studies how security patterns influence maintainability for open-source software.
+software~\cite{4493325}. Yskout et. al were not able to detect if the usage of 
+design patterns has a positive impact, but concludes that developers prefer to 
+work with the support of security patterns~\cite{8077802}. The present work 
+studies how security patterns influence maintainability for open-source software.
 
@@
 Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922,
-Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the previously demonstrated
-correlation between programming language and software defects. Our work studies how
-security refactorings affect software quality based on the code maintainability
-analysis and provides evidence that programming languages have an impact on
-maintainability.
+Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the 
+previously demonstrated correlation between programming language and software 
+defects. Our work studies how security refactorings affect software quality based 
+on the code maintainability analysis and provides evidence that programming languages 
+have an impact on maintainability.
 
@@
 
-This work presents an empirical study on the impact of $1031$ security
-patches on the maintainability of $267$ open-source projects. We leveraged
-Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018}. Security patches 
+This work presents an empirical study on the impact of $998$ security
+patches on the maintainability of $262$ open-source projects. We leveraged
+Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018, cruz2019energyoriented}. Security patches 
 significantly affect the software maintainability of open-source software, i.e., 
 developers hinder maintainability when patching vulnerabilities. We found this evidence in
-$41.76\%$ of the studied cases. Regarding regular commits, it seems that overall
-maintainability stays the same ($52.06\%$). However, there is still a 
-considerable amount of cases that harm maintainability ($21.04\%$).
+$41.28\%$ of the studied cases. Regarding regular commits, it seems that overall
+maintainability stays the same ($42.69\%$). However, there is still a 
+considerable amount of cases that harm maintainability ($27.05\%$).
 
 This study also shows evidence that there are security patterns that need more
-attention than others. \emph{Path Traversal} (CWE-22),
-\emph{Information exposure} (CWE-200) and \emph{Resource Management Errors} (CWE-399) vulnerabilities endure maintainability
-in $48.6\%$, $51.2\%$, and $54.5\%$, respectively. Therefore, we draw attention
-for the need of better documentation explaining how to use best practices to
+attention than others. \emph{Improper Input Validation} (CWE-20), \emph{Information 
+Exposure} (CWE-200), \emph{Improper Restriction of Operations within the Bounds of a 
+Memory Buffer} (CWE-119), \emph{Missing Release of Memory after Effective Lifetime} 
+(CWE-401) and \emph{Path Traversal} (CWE-22) endure maintainability
+in $41.9\%$, $52.5\%$, $42.9\%$, $49.4\%$, and $41.7\%$, respectively. Therefore, 
+we draw attention for the need of better documentation explaining how to use best practices to
 produce secure and maintainable code; new and better programming languages; new

### 2020-05-04 paper/icsme20.tex
review

--- 2020-04-29 paper/icsme20.tex
@@
 \usepackage{balance}
-\usepackage{hyperref}
+\usepackage{url}
 
@@
 
-% \title{Fixing Vulnerabilities Hinders Maintainability}
-\title{\Sof{TBA - New Titles in the comments}}
-%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
-%\title{The reason why you should be applying coding practices: safer software.}
-%\title{What is the impact of security patches in software evolution?}
-%\title{Hey, developers! You should pay more attention to your patches.}
+\title{Fixing Vulnerabilities Hinders Maintainability}
 
@@
 Maintainability is measured based on the Better Code Hub’s model of $10$ 
-guidelines using a dataset including $1335$ security commits
-\Sof{These are the initial values, should I use the final? In the end, 
-we studied 1031 commits and 8 guidelines}. 
-Results show 
-that fixing software vulnerabilities may hinder the maintainability of 
-open-source software. This is particularly the case for patches that 
-address Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource 
-Management Errors (CWE-399). We conclude that changes to code bases to patch 
-vulnerabilities need to be performed with extra care and that security-related 
-code evolution should be subject in computer curricula.
+guidelines using a dataset including $1037$ security commits. Results show 
+that fixing software vulnerabilities hinders the maintainability of 
+open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399). 
+We conclude that changes to code bases to fix vulnerabilities need to be 
+performed with extra care and that security-related code evolution should 
+be subject in computer curricula.
 \end{abstract}
@@
 These vulnerabilities are gateways for attackers to penetrate systems and 
-exploit their resources. As an example, at the beginning of $2020$, academics published a 
-new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
-security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
-address in previous generations and that makes every user using a CPU released before Q$4$ $2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
-at the cost of features and/or performance.
-%
-
+exploit their resources. \Sof{Improve example - outdated}\Luis{I think it is okay. Just say it is from 2019. It's too precise atm.}As an example, in the last week of January $2019$, 
+a severe vulnerability in Apple FaceTime was revealed\footnote{FaceTime vulnerability 
+details available at \url{https://thehackernews.com/2019/01/apple-facetime-
+privacy-hack.html} (Accessed on \today{})} --- the outcome of a design/logical
+flaw that allowed callers to see and hear others without them picking up the call.%\cite{cacheOut}.
+%
 In $2011$, the International Organization for Standardization (ISO) issued an
@@
 %
-
-As ISO does not provide any specific guidelines/formulas to calculate 
+As ISO does not provide any specific guidelines or formulas to calculate 
 maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
@@
 with a set of $10$ guidelines/metrics to produce quality software 
-based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
+based in ISO/IEC 25010~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
 were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
@@
 software mainly due to the broad number of metrics used.
-
 From a methodological 
@@
 impact. This empirical study presents evidence that changes 
-applied in the code bases to patch vulnerabilities affect code maintainability.\Sof{fix}Especially, for patches that address Path Traversal (CWE-22),
-Information Exposure (CWE-200) and Resource Management Errors (CWE-399). \Sof{-END-}This is important work
+applied in the code bases to patch vulnerabilities affect code maintainability. 
+Especially, for patches that address Path Traversal (CWE-22),
+Information Exposure (CWE-200) and Resource Management Errors (CWE-399). \Luis{Little information is known about..}This is important work
 because there is still little information about the impact of security patches on the software
@@
 we intend to highlight the need for tools to predict the impact of patches on 
-maintainability~\cite{4724577}, documentation to help developers adopt security 
-best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
-security-related code evolution in computer curricula.
+maintainability~\cite{4724577} and documentation to help developers adopt security 
+best practices~\cite{6311252, 7927935, MESQUIDA201519}.
 %
@@
 %
-
 The main contributions of the present work are:
@@
 	empirical evaluation, for reproducibility. Available online:
-  \url{https://figshare.com/s/4861207064900dfb3372}.
+  \textit{<anonymized for double blind review>}.
 \end{itemize}
@@
 presents the results and section~\ref{sec:discussion} discusses their
-implications; section~\ref{sec:threats} enumerates the threats to the validity of
+implications; section~\ref{sec:threats}, enumerates the threats to the validity of
 this study; section~\ref{sec:rw} describes the different work and existing
@@
 %%%%% NEEDS REVIEW
-As an example, consider the patch of the TLS state-machine protocol implementation in 
-OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
+As an example, consider the patch of the TLS state-machine protocol implementation in OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
 (DoS) attack by crafting TLS messages to trigger memory allocations
 corresponding to large length values. This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$~\footnote{CVE-$2016$-$6307$
+Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$ \footnote{CVE-$2016$-$6307$
 details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-research. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
+evaluation. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
 \emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
@@
 CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
-the memory for a message is allocated too early, prior to the excessive
+memory for a message is allocated too early, prior to the excessive
 message length check which can lead to a Denial of Service through memory
@@
 are multiple connections in a state where memory has been allocated for
-the connection, when \texttt{SSL\_free()} has not yet been called, and there is
+the connection, \texttt{SSL\_free()} has not yet been called, and there is
 insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
 be transitory because as soon as the connection fails the memory is
-subsequently freed again in the \texttt{SSL\_free()} call. However, there is an
+subsequently freed again in the \texttt{SSL\_free()} call. However there is an
 increased risk during this period of application crashes due to the lack
@@
 \\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
- static SUB_STATE_RETURN read_state_machine(SSL *s){ 
+ static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
  // [snip]
@@
 \\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
- int tls_get_message_header(SSL *s, int *mt){   
+ int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
  // [snip]
@@
         SSL3_HM_HEADER_LENGTH;
-- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func3}!*/
+- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
 -      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
@@
 
-- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func4}!*/
+- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func5}!*/  
 -     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
@@
 
-The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
+~\ref{lst:func1} \texttt{read\_state\_machine} function abstracts a sub-state-machine
 to handle the messages exchanged between peers. When messages are larger than
 what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
-of this resource was being verified very early in the application's flow, using the \texttt{tls\_get\_message\_header} function presented in Listing ~\ref{lst:vuln2} (\ref{lst:func3} and \ref{lst:func4}).
-These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).
+of this resource was being verified very early in the application's flow.  Listing ~\ref{lst:vuln2} presents those verifications (lines $5$-$7$ and line $11$-$13$).
+These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}.
 
@@
 OpenSSL maintainability.
-\Sof{-- END --}
 %%%%%%% 
-Software maintainability is designed as the degree to which an application is understood, 
-repaired or enhanced. In this paper, our concern is to study whether, while improving software
+Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software
 security, developers are also negatively impacting the maintainability of their
-software applications. This is important because software maintainability is 
-approximately $75\%$ of the cost related to a project. To answer the following three 
-research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
-to measure the impact of
+software applications.
+This is important because software maintainability is approximately $75\%$ of the cost related to a project.
+To answer the following three research questions, we use two datasets
+of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
 security patches on the maintainability of open-source software. 
@@
 
-\textit{\textbf{RQ1: What is the impact of security patches on the
-maintainability of open-source software?}} Often security flaws require patching 
-code to make software more secure.
+\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the
+maintainability of open-source software?}}}
+
+Often security flaws require patching code to make software more secure.
 However, \textbf{there is no evidence yet of how security patches impact the
@@
 address a security flaw because they tend to choose the easiest path to solve
-it. To address it, we compute the maintainability of $1335$ commits using the
+it. To address it, we compute the maintainability of $1037$ commits using the
 \emph{Better Code Hub} tool. 
 %
-
-\textit{\textbf{RQ2: Which patterns of security patches are more likely to
-affect open-source software maintainability?}}
+\subsubsection*{\textit{\textbf{RQ2}  \textbf{Which patterns of security patches are more likely to
+affect open-source software maintainability?}}}
+
 There are security flaws that are more difficult to patch than others. For
@@
 without adding new lines of code/complexity to the code. A typical fix for 
-the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to escape the data given by the variable
+the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to \textit{escape} the data given by the variable
 \texttt{\$\_['file']}. \textbf{Knowing which patterns are more likely to increase
@@
 %
-\textit{\textbf{RQ3: What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}}
-%
-We also compute the maintainability of random regular commits using the
+\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes on the
+maintainability of open-source software?}}}
+%
+We also compute the maintainability of $1037$ random regular commits using the
 \emph{Better Code Hub} tool --- baseline.  We use them to understand \textbf{how
@@
 %
+\Luis{this passive voice here sounds a bit weird (imo)}
 The methodology used to measure the impact of security patches on the
@@
 Figure~\ref{fig:met}:
+\Luis{could you make the numbers in the figure a bit larger}
 %
@@
 	\item
-	Extract relevant data (e.g., owner and name of
-	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and code bases from the combined dataset containing $1335$ security
+	Extract relevant data (e.g., sha, sha of the parent commit, owner of the project, name of 
+	the repository) and code bases from the combined dataset containing $1335$ security
 	refactorings collected from open-source software available on
@@
 %
+\Luis{We already made clear that we ``evaluate the impact..'' }
 To evaluate the impact of security refactorings on the maintainability of
 open-source software, we use a combined dataset of security vulnerabilities which is 
-the outcome of mining/manually inspect $265$ GitHub projects. Reis and Abreu 
+the outcome of mining/manually inspect $267$ GitHub projects\Luis{:mining and manually inspecting:}.
+\Luis{it is not clear the connection between the 267 projects and the works we mention below.
+There should be a sentence here clarifying and making the connection.} Reis and Abreu
 ($2017$) mined open-source
@@
 of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
-dataset of $711$ test cases for $16$ different security vulnerability types. Some of the vulnerability types
+dataset of $711$ test cases for $16$ security patterns. Some of the patterns
 are based on the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
@@
 another (typically, called \textit{diffs}) --- where one can easily review the
-code used to fix the security flaw. 
+code used to fix the security flaw. In this study, we focus on computing the
+maintainability of the commits before and after the security patching to
+evaluate if its impact was positive, negative or none.
 %
@@
 from $2014$ to $2019$. For each new vulnerability, the authors manually searched 
-for references to commits involved in the patch on the National Vulnerability Database (NVD) website. However, $70\%$
+for references to commits involved in the patch on the NVD's website. However, $70\%$
 of the vulnerabilities did not have any references to commits. Thus, the authors
@@
 To fit the dataset in our methodology, we search for the first and last commits
-used to patch the vulnerability. We assumed the last commit of the patch as 
+used to patch the vulnerability. We assumed that the last commit of the patch as 
 the fix (\emph{sha}) and the parent of the first commit as the vulnerable version 
@@
 %
-In this study, we focus on computing the
-maintainability of the commits before and after the security patching to
-evaluate if its impact was positive, negative or none.
 The $1335$ patches in the dataset were run against the BCH toolset to
@@
 language support and project size. The final dataset used in this paper comprises
-$1031$ security patches. $912$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE) where $59.1\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $40.9\%$ were manually classified following the \texttt{Research Concepts} view map from the CWE's list.
-
-% In Table~\ref{tab:patterns}, the most prevailing categories are presented. To satisfy the Wilcoxon test size requirement, types of vulnerabilities
-% with less than $20$ instances were merged in a bigger group named \textit{Miscellaneous}.
-
-% \begin{table}[h]
-% \scriptsize
-% 	\caption{Guidelines to produce maintainable code}
-% \begin{tabular}{L{1.1cm}L{2.1cm}L{4.3cm}}
-%
-% \toprule
-% \textbf{Pattern} & \textbf{Name} & \textbf{Description}\\
-% \midrule
-%
-% \textbf{CWE-89} &
-%  Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases) &
-%  When developers do not keep untrusted data
-%      separate from SQL queries. If an attacker sends a command that
-%      exploits the syntax of the SQL interpreter, then SQL injection attack is possible.\\\midrule
-%
-% \textbf{CWE-611} &
-%  Improper Restriction of XML External Entity Reference (24 cases) &
-% 	XML documents are exchanged through the web containing entities with URIs
-%  	that resolve to local/external files. Thus, when XML parsers are not well configured,
-%  	attackers have allowed to directly those files.\\\midrule
-%
-%  \textbf{CWE-20} &
-%  Improper Input Validation (76 cases) &
-% 	When developers
-%  	forget to validate the inputs of an application, attackers may have control
-%  	of the control/data flow of the program.\\\midrule
-%
-%   \textbf{CWE-200} &
-% Information Exposure (39 cases)&
-% Application's logs are one way of intentional/unintentional disclosure information
-%  	to attackers. Many times attackers get access to logs when sould not be authorized to.\\\midrule
-%
-%    \textbf{CWE-352} &
-% Cross-Site Request Forgery/CSRF (22 cases)&
-% Poor session tokens generation
-%      and management usually allow attackers to send forged HTTP requests
-%      including authentication information from the victim to the vulnerable web
-%      application.\\\midrule
-%
-%        \textbf{CWE-264} &
-% Permissions, Privileges, and Access Controls (59 cases)&
-% Incorrectly
-%  	    implemented functionalities related to authentication and session
-%  	    management, allowing an attacker to gain access to session tokens,
-%  	    passwords, keys, and other sensitive data.\\\midrule
-%
-%  	           \textbf{CWE-399} &
-% Resource Management Errors (22 cases)&
-% Resource management issues are found more frequently
-%  		in programming languages that do not manage memory automatically (e.g., C/C++
-%  	    and Objective-C), i.e., where developers are responsible for
-%  	    handling it instead. It is one of the main causes of Denial-of-Service attacks.\\\midrule
-%
-%  	     	           \textbf{CWE-79} &
-% Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (63 cases)&
-% Lack of proper validation or escaping
-%  	    allows attackers to submit untrusted data to web browsers through malicious
-%  	    scripts that can hijack the user sessions or redirect the user to malicious
-%  	    sites.\\\midrule
-%
-%  	     	           \textbf{CWE-22} &
-% Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases) &
-% When developers forget to neutralize special elements within the path names to files/directories, attackers
-%  	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
-%  	or entire repositories.\\\midrule
-%
-%  	 	     	           \textbf{MISC} &
-% Miscellaneous (666 cases) &
-% This pattern comprises
-%  		several other security refactorings that do not have a CWE assigned and patterns that
-%  		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
-% \bottomrule
-% \end{tabular}
-% \label{tab:patterns}
-% \end{table}
+$1037$ security patches. $556$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE). In Table~\ref{tab:patterns}, the most prevailing categories are presented. Patterns
+with less than $20$ instances were merged in a bigger pattern named \textit{Miscellaneous}.
+
+\begin{table}[h]
+\footnotesize
+	\caption{Guidelines to produce maintainable code}
+\begin{tabular}{L{1.1cm}L{2.1cm}L{4.3cm}}
+
+\toprule
+\textbf{Pattern} & \textbf{Name} & \textbf{Description}\\
+\midrule
+
+\textbf{CWE-89} & 
+ Improper Neutralization of Special Elements used in an SQL Command/SQL Injection (23 cases) & 
+ When developers do not keep untrusted data
+     separate from SQL queries. If an attacker sends a command that
+     exploits the syntax of the SQL interpreter, then SQL injection attack is possible.\\\midrule
+
+\textbf{CWE-611} & 
+ Improper Restriction of XML External Entity Reference (24 cases) & 
+	XML documents are exchanged through the web containing entities with URIs
+ 	that resolve to local/external files. Thus, when XML parsers are not well configured,
+ 	attackers have allowed to directly those files.\\\midrule
+
+ \textbf{CWE-20} & 
+ Improper Input Validation (76 cases) & 
+	When developers
+ 	forget to validate the inputs of an application, attackers may have control
+ 	of the control/data flow of the program.\\\midrule
+
+  \textbf{CWE-200} & 
+Information Exposure (39 cases)& 
+Application's logs are one way of intentional/unintentional disclosure information
+ 	to attackers. Many times attackers get access to logs when sould not be authorized to.\\\midrule
+
+   \textbf{CWE-352} & 
+Cross-Site Request Forgery/CSRF (22 cases)& 
+Poor session tokens generation
+     and management usually allow attackers to send forged HTTP requests
+     including authentication information from the victim to the vulnerable web
+     application.\\\midrule
+
+       \textbf{CWE-264} & 
+Permissions, Privileges, and Access Controls (59 cases)& 
+Incorrectly
+ 	    implemented functionalities related to authentication and session
+ 	    management, allowing an attacker to gain access to session tokens,
+ 	    passwords, keys, and other sensitive data.\\\midrule
+
+ 	           \textbf{CWE-399} & 
+Resource Management Errors (22 cases)& 
+Resource management issues are found more frequently
+ 		in programming languages that do not manage memory automatically (e.g., C/C++
+ 	    and Objective-C), i.e., where developers are responsible for
+ 	    handling it instead. It is one of the main causes of Denial-of-Service attacks.\\\midrule
+
+ 	     	           \textbf{CWE-79} & 
+Improper Neutralization of Input During Web Page Generation/Cross-site Scripting (62 cases)& 
+Lack of proper validation or escaping
+ 	    allows attackers to submit untrusted data to web browsers through malicious
+ 	    scripts that can hijack the user sessions or redirect the user to malicious
+ 	    sites.\\\midrule
+
+ 	     	           \textbf{CWE-22} & 
+Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (37 cases) & 
+When developers forget to neutralize special elements within the path names to files/directories, attackers
+ 	may leverage this flaw to resolve to a location outside of the restricted directory and gain access to documents
+ 	or entire repositories.\\\midrule
+
+ 	 	     	           \textbf{MISC} & 
+Miscellaneous (673 cases) & 
+This pattern comprises
+ 		several other security refactorings that do not have a CWE assigned and patterns that
+ 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
+\bottomrule
+\end{tabular}
+\label{tab:patterns}
+\end{table}
 
@@
 of some patches, it was not possible to find patches with the exact number of added and 
-deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. The algorithm integrates the following steps:
+deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. The algorithm follows the following steps:
 
 \begin{enumerate}
-\item Calculates the \emph{diff} between the security patch commit (\emph{sha}) and its parent (\emph{sha-p}).
-\item Selects a random commit from the project (\emph{sha-reg}).
-\item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
-\item Checks if the size from the regular change has the same/approximated size to 
+\item Diff comparison between the security patch commit (\emph{sha}) and its parent (\emph{sha-p}).
+\item Selects a random commit from the security patch project (\emph{sha-reg}).
+\item Diff comparison between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
+\item Check if the size from the regular change has the same/approximated size to 
 the security patch.
@@
 
-The pair of the random commit and its parent is accepted if the \emph{diff} 
-size fits in the range size considered acceptable at the moment. The algorithm 
-starts with a step of $1$. Every $10$ attempts to search for a change with the 
-same \emph{diff} the step increases. Thus, if a change suffered $20$ deletions 
-and $10$ additions, then, in the first $10$ attempts, the algorithm searches 
-for a \emph{diff} with deletions between $19$ and $21$ and additions between 
-$9$ and $11$. After the $10$ first attempts, the algorithm increase the step 
-for $2$ and searches for a \emph{diff} with deletions between $18$ and $22$ 
-and additions between $8$ and $10$. We originate the regular commits from the 
-security commits to ensure that differences in maintainability are not a 
-consequence of characteristics of different projects.
+Our approximation was calculated using an initial step of $1$. Every $10$ attempts to search for 
+a change with the same \emph{diff} the step increased. Thus, if a change suffered $20$ deletions 
+and $10$ additions, then, in the first $10$ attempts, the algorithm searches for a \emph{diff} with 
+deletions between $19$ and $21$ and additions between $9$ and $11$. After the $10$ first attempts, 
+the algorithm increase the step for $2$ and searches for a \emph{diff} with deletions between $18$ and $22$ and additions between 
+$8$ and $10$. We originate the regular commits from the security
+commits to ensure that differences in maintainability are not a consequence of
+characteristics of different projects.
 
@@
 
-In this research, we follow a very similar methodology to the one presented 
-in previous work on the maintainability of energy-oriented fixes~\cite{cruz2019energyoriented}. 
-As said before, the web-based source code analysis service \emph{Better Code
+In this research, we follow a very similar methodology to the one presented in previous work on the maintainability 
+of energy-oriented fixes~\cite{cruz2019energyoriented}. As said before, the web-based source code analysis service \emph{Better Code
 Hub} (BCH) is used to collect the maintainability reports of the refactorings of
@@
 calculated by BCH. Theses guidelines are calculated using the metrics 
-presented in~\cite{criteria:2017} and also briefly explained on Table~\ref{tab:guidelines}. 
-During each guideline evaluation, the tool determines the compliance towards 
-one guideline by establishing limits for the percentage of code allowed to be 
-in each of the $4$ risk severity levels
+presented in~\cite{criteria:2017}. During each guideline evaluation, the tool
+determines the compliance towards one guideline by establishing limits for the
+percentage of code allowed to be in each of the $4$ risk severity levels
 (\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
@@
 \begin{table}[h]
-\scriptsize
+\footnotesize
 	\caption{Guidelines to produce maintainable code}
@@
 
-\Sof{Needs improvement}SIG defines \emph{Units} as the smallest groups of code that can be maintained
+SIG defines \emph{Units} as the smallest groups of code that can be maintained
 and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and
@@
 somewhere in the green bar it means it is compliant with a low level of
-severity.\Sof{--END--}
+severity.
 
@@
 to capture the distance between the current state of the project and
-the standard thresholds calculated by the BCH based on the insights 
-provided in~\cite{Olivari:2018}. The equation provided in~\cite{cruz2019energyoriented} 
-considers that the size of project changes does not affect the maintainability and
-that the distance to lower severity levels is less penalized than to the
-thresholds in high severity levels.
-% 	thresholds in high severity levels
-% 	difference between two versions
-% The equation considers the following:
-% \begin{itemize}
-% 	\item \textbf{Size of project changes does not affect the maintainability
-% 	difference between two versions, $\Delta M (v_{s-1},v_{s}) = M(v_{s}) - M(v_{s-1})$.} We
-% 	aim at evaluating security patterns occurring in different projects similarly.
-% 	Thus, the derived metric uses the \textit{raw} number of lines of code rather
-%   percentages for normalization purposes.
-% 	\item \textbf{Distance to lower severity levels is less penalized than to the
-% 	thresholds in high severity levels.} Severity level weights based on the
-% 	severity level to count lines of code that violate maintainability guidelines.
-% \end{itemize}
+the standard thresholds calculated by the BCH based on the insights provided in~\cite{Olivari:2018}. 
+The equation considers the following:
+\begin{itemize}
+	\item \textbf{Size of project changes does not affect the maintainability
+	difference between two versions, $\Delta M (v_{s-1},v_{s}) = M(v_{s}) - M(v_{s-1})$.} We
+	aim at evaluating security patterns occurring in different projects similarly.
+	Thus, the derived metric uses the \textit{raw} number of lines of code rather
+  percentages for normalization purposes.
+	\item \textbf{Distance to lower severity levels is less penalized than to the
+	thresholds in high severity levels.} Severity level weights based on the
+	severity level to count lines of code that violate maintainability guidelines.
+\end{itemize}
 
@@
 maintainability guideline for the given severity level (medium, high, and very high).
+\Luis{The compliance for a severity level
+is calculated based on previous work, which calculates the number of lines of code that comply and not comply
+with the guideline at a given severity level~\cite{cruz2019energyoriented}.}
 The compliance for a severity level
@@
 between pairs of versions $v_{s-1}$, $v_s$ (i.e., before and after a
-security-commit) come from the same distribution. Nevertheless, this test 
-has a limitation: it does not consider the groups of commits with a 
-zero-difference maintainability. In $1959$, Pratt provided an improvement to 
-the test to solve this issue making the test more robust. Thus, we use a 
-version of the Wilcoxon test that incorporates the cases where 
-maintainability is different from zero~\cite{10.2307/2282543}.
-The Wilcoxon test requires a distribution size of at leat $20$ instances.
-To understand the effect-size, as advocated by the Common-language effect 
-sizes\cite{graw:1992}, we compute the mean difference, the median of the 
-difference, and the percentage of cases that reduce maintainability.
+security-commit) come from the same distribution. Nevertheless, this test has a
+limitation: it does not consider the groups of commits with a zero-difference
+maintainability. In $1959$, Pratt provided an improvement to the test to solve
+this issue making the test more robust. Thus, we use a version of the Wilcoxon
+test that incorporates the cases where maintainability is different from zero~\cite{10.2307/2282543}.
+To understand the effect-size, as
+advocated by the Common-language effect sizes\cite{graw:1992}, we compute the
+mean difference, the median of the difference, and the percentage of cases that
+reduce maintainability.
 %
@@
 
-This study evaluates $1031$ security and baseline commits from $265$ 
-distinct open-source projects. The following section presents the results 
-obtained for each research question.
-
-\textit{\textbf{RQ1: What is the impact of security patches on the
-maintainability of open-source software?}}
+This study evaluates $1037$ security and baseline commits
+from $267$ distinct open-source projects. The following section presents the
+results obtained for each research question.
+
+\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the
+maintainability of open-source software?}}}
 
@@
  	\centering
-	\vspace{-0.3 cm}
- 	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
-	\vspace{-0.3 cm}
- 	\caption{Impact of the security patches per guideline.}
+ 	\includegraphics[width=0.5\textwidth]{figures/impact_per_guideline.pdf}
+ 	\caption{Impact of the security patches for each guideline.}
 	\label{fig:guidelines}
-	\vspace{0.2 cm}
 	\scalebox{0.5}{
@@
 Maintainability is being calculated as the mean of the guidelines computed
-by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on 
-each guideline individually. For each type of guideline, three types of 
-impact of patches in the software maintainability are presented: 
-\emph{negative}, maintainability decreases (\emph{red});
+by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on each 
+guideline individually. For projects with large 
+codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
+viable due to limitations of BCH.
+For this guideline, the tool sets a limit of 20 Person-years.
+However, a large amount of projects of our database surpass this limit.
+Thus, we decided to not consider this guideline
+in our analysis. For each type of
+guideline, three types of impact of patches in the software maintainability
+are presented: \emph{negative}, maintainability decreases (\emph{red});
 \emph{none}, maintainability remains the same (\emph{yellow}); and,
 \emph{positive}, maintainability increases (\emph{green}). Next to each
-type of guideline, it is presented the mean ($\overline{x}$) and median (M)
+type of guideline, it is shown the mean ($\overline{x}$) and median (M)
 of the maintainability difference and the p-value resulting from the
@@
 
+Regarding security commits results, we observe that maintainability decreases in
+$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
+($400$) of the cases after security patches are applied. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
+Since the p-value is below the significance level of $0.05$, we argue that
+\emph{security patches have a negative impact on the maintainability of
+open-source software}.
 Regarding the impact of security patches for each guideline, we observe
-that overall patches have a significant positive impact in $6$ guidelines: 
-\emph{Write Clean Code} (p-value = $5.49$x$10^{-54}$), \emph{Keep 
-Architecture Components Balanced} (p-value = $5.30$x$10^{-71}$), 
-\emph{Couple Architecture Components Loosely} (p-value = $2.24$x$10^{-07}$), 
-\emph{Separate Concerns in Modules} (p-value = $0.399$), \emph{Keep Unit 
-Interfaces Small} (p-value = $1.94$x$10^{-13}$), \emph{Write Code Once} 
-(p-value = $1.03$x$10^{-34}$).
-
-However, for each guideline we also observe a 
-substantial number of cases where the maintainability suffers a negative 
-impact $-$ from $10\%$ to $40\%$ $-$ which highlights the need of better 
-practices/techniques/tools to produce software. \emph{Separate Concerns in 
-Module} guideline does not the p-value is higher than $0.05$.
-
-According to the 
+that overall patches have a significant positive impact in $5$ guidelines: 
+\emph{Write Clean Code} (p-value = $4.9$x$10^{-54}$), \emph{Keep Architecture 
+Components Balanced} (p-value = $5.26$x$10^{-71}$), \emph{Couple Architecture 
+Components Loosely} (p-value = $2.43$x$10^{-07}$), \emph{Keep Unit Interfaces 
+Small} (p-value = $2.25$x$10^{-13}$), \emph{ Write Code Once} (p-value = 
+$7.9$x$10^{-35}$).
+However, for each of these guidelines we also observe a substantial number of cases where the maintainability suffers a negative impact $-$ from $10\%$ to $35\%$ $-$ which highlights the need of better practices/techniques/tools to produce software.
+\emph{Separate Concerns in Module}
+guideline also follows the same results but without any statistical 
+significance since the p-value is higher than $0.05$. According to the 
 results, \emph{Write Simple Units of Code} (p-value=$0.015$ and $38.2\%$/$396$ 
@@
 
- %%%%%%%%%%% HERE
-
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.45\textwidth]{figures/main_per_severity.pdf}
+ 	\centering 	\includegraphics[width=0.45\textwidth]{figures/maintainability_severity.pdf}
  	\caption{Maintainability difference by the vulnerabilities severity.}
@@
 \end{figure}
-
-For projects with large 
-codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
-viable because they retrieved values above the limit set by BCH ($20$ Person-years).
-The \emph{Automated Tests} guideline was also not considered because the BCH does not return
-the amount of lines of code that complies with the guideline -- which is one of the variables
-our formula uses to calculate the maintainability of the project regarding each guideline. 
-Thus, we decided to not consider both guidelines in our analysis.
 
@@
 software. Although with no statistical significance, we observe that 
-vulnerabilities with higher severity levels (\emph{HIGH} and \emph{MEDIUM}) have 
-more negative impact than the \emph{LOW} severity level on the
+vulnerabilities with higher severity levels (\emph{HIGH} and \emph{MEDIUM}) have more negative impact than the \emph{LOW} severity level on the
 maintainability of software.
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figures/main_per_language.pdf}
+  \includegraphics[width=0.5\textwidth]{figures/maintainability_language.pdf}
   \caption{Maintainability difference by language after security patches.}
@@
 We have also evaluated the impact on the maintainability per programming language
-of open-source software, Figure~\ref{fig:lang_main}. We restrict this analysis to 
-programming languages with at least $20$ data points, as this is a requirement 
-for our hypothesis tests. Thus, we compare the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP}, 
+of open-source software, Figure~\ref{fig:lang_main}.
+We restrict this analysis to programming languages with at least $20$ data points, as this is a requirement for our hypothesis tests.
+Thus, we compare the results for \emph{Python}, \emph{C/C++}, \emph{Ruby}, \emph{PHP},
 \emph{Objective-C/C++}, and \emph{Java}. The other languages were
@@
 \emph{Ruby} and \emph{PHP} languages also impact negatively the maintainability of software.
-Despite not having statistical significance, there are a considerable number of \emph{Ruby} 
-and \emph{PHP} patches hindering maintainability that should not be disregarded.
+Despite not having statistical significance, there are a considerable number of \emph{Ruby} and \emph{PHP} patches hindering maintainability that should not be disregarded
 
@@
 security patches.
-
-Regarding security commits results, we observe that maintainability decreases in
-$41.61\%$ ($429$), remains the same in $19.69\%$ ($203$) and increases in $38.70\%$
-($399$) of the cases after security patches are applied. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $4.9$x$10^{-2}$.
-Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security patches have a negative impact on the maintainability of
-open-source software}.
 
@@
  	\centering
- 	\includegraphics[width=0.4\textwidth]{figures/main_per_cwe.pdf}
+ 	\includegraphics[width=0.4\textwidth]{figures/maintainability_cwe.pdf}
  	\caption{Maintainability difference by pattern after security patches.}
@@
 \begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.43\textwidth]{figures/main_comparison.pdf}
+ 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
  	\caption{Maintainability difference for security and baseline patches.}
@@
 
-The guideline analysis suggests that all guidelines are affected by security patches. Being the
+The guideline analysis suggests that all guidelines are affected by security patches. 
+\Luis{The guidelines ... have the most negative impact ...}
+Being the
 \emph{Write Short Units of Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns in Modules} the guidelines with higher negative impact on the maintainability ($>30\%$ of cases).
 \emph{Keep Architecture Components Balanced} is the guideline with lower negative impact on
-the maintainability (approx. $10\%$). This might imply that developers should give different
+the maintainability (approx.
+$10\%$).
+This \Luis{I dont like the ``might imply''.
+I was going to recommend replacing by ``suggests'', but I can see it is already being used below.
+Can you think of something else?}
+might imply that developers should give different
 levels of attention to the guidelines when patching vulnerabilities.
@@
 Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
-the result of projects contributions policies. Nine of the top-10 of dataset projects 
+the result of projects contributions policies. \Luis{In our dataset, nine of the ten projects with more contributors have strict...} Nine of the top-10 of dataset projects 
 with more contributors have restrict contribution policies regarding code
@@
 \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
-
+\Luis{I did not fully understand this paragraph.. We are mixing programming languages, contribution policies, and number of contributors in the same bucket.}
+\Luis{I think this section is missing more discussion.
+What do you think should be changed in the community to cope with our findings?
+How can practitioners fix security vulnerabilities wihthout hindering maintainability?
+do we need more tools, better programming languages, more strict contribution policies?
+is there a trade-off between secure software and technical debt?
+show us how these findings can change SE}
 \subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
@@
 Overall, the results obtained for the maintainability per CWE are no statistical significance 
-The impact of a patch depends on its complexity, i.e., if the patchs
+The impact of a patch depends on its complexity, i.e., if the patch
 adds complexity to the code base it is probably affecting the software
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $1031$ regular commits are enough to
+Furthermore, this evaluation considers that $1037$ regular commits are enough to
 alleviate random irregularities in the maintainability differences of the
@@
 
-This work presents an empirical study on the impact of $1031$ security
+This work presents an empirical study on the impact of $1037$ security
 patches on the maintainability of $267$ open-source projects. We leveraged

### 2020-05-04 paper/icsme20.tex
new update

--- 2020-04-30 paper/icsme20.tex
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
+ 	\centering 	
+	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
  	\caption{Maintainability difference for security commits}
@@
 
+Patches can be performed through one commit, several consecutive commits, or 
+commit(s) interleaved with more programming activities, being the later
+commonly called a floss-refactoring. This study considers mainly 1-commit 
+patches. Yet, a small percentage of data points had more than one commit 
+involved in the patch. We manually inspected these cases and $23$ 
+floss-refactorings were identified and disregarded since it would not
+be fair to measure the maintainability of these cases where other
+programming activities are involved.
+
 The BCH tool does not compute the final score that our study needs to compare
-maintainability amongst different project versions. For this, we
-follow previous work on measuring the impact of energy-oriented refactorings
-~\cite{cruz2019energyoriented}. Luis et al. ($2019$) proposed an equation
-to capture the distance between the current state of the project and
-the standard thresholds calculated by the BCH based on the insights 
-provided in~\cite{Olivari:2018}. The equation provided in~\cite{cruz2019energyoriented} 
-considers that the size of project changes does not affect the maintainability and
-that the distance to lower severity levels is less penalized than to the
+maintainability amongst different project versions. For this, we follow previous 
+work on measuring the impact of energy-oriented 
+refactorings~\cite{cruz2019energyoriented}. Luis et al. ($2019$) proposed an 
+equation to capture the distance between the current state of the project and
+the standard thresholds calculated by the BCH based on the insights provided 
+in~\cite{Olivari:2018}. The equation provided in~\cite{cruz2019energyoriented} 
+considers that the size of project changes does not affect the maintainability 
+and that the distance to lower severity levels is less penalized than to the
 thresholds in high severity levels.
+
 % 	thresholds in high severity levels
@@
 which patches had a positive, negative or null impact on the project maintainability.
+
+\Sof{@todo: Don't forget to mention the limitations of BCH and the outliers.}
+
 % The compliance $C$ for a given severity
@@
 
-This study evaluates a final total of $998$ security and baseline commits from $262$ 
-distinct open-source projects. The following section presents the results 
-obtained for each research question.
+This study evaluates a final total of $998$ security and baseline commits 
+from $262$ distinct open-source projects. The following section presents and 
+discusses the results for each question. Research questions are also answered.
 
@@
 maintainability of open-source software?}}
-
-Maintainability is being calculated as the mean of the guidelines computed
-by BCH. Figure~\ref{fig:guidelines} shows the impact of security patches on 
-each guideline individually. For each type of guideline, three types of 
-impact of patches in the software maintainability are presented: 
+\textit{Patching vulnerabilities may harm the maintainability of open-source
+software.}
+
+In this section, we report and discuss the impact of patches on open-source
+software maintainability under four different views: guidelines/metrics, 
+severity, programming language and formula proposed - the mean of all 
+guidelines/metrics.
+
+% As expected, vulnerabilities with higher severity are more likely to impact negatively
+% the maintainability of software. This suggests that high/medium severity vulnerabilities may
+% need more attention than low severity when patching vulnerabilities.
+% %
+% We would expect that the negative impact of programming languages on
+% maintainability to be more severe, as arguably poor design of programming
+% languages and lack of using the best practices lead to more buggy and vulnerable
+% code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
+% Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
+% the result of projects contributions policies. Nine of the top-10 of dataset projects
+% with more contributors have restrict contribution policies regarding code
+% standards. Some examples are the \texttt{laravel/framework},
+% \texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
+% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
+% Standard}\footnote{PSR-2 Standard available at
+% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
+
+\textbf{Guidelines/Metrics:} Each patch performs a set of changes
+on the source code of the software. These changes may have a different
+impact on the guidelines/metrics used to measure software maintainability. 
+Figure~\ref{fig:guidelines} shows the impact of security patches 
+on each guideline individually. Under each guideline is stated the 
+metric used to the calculations. Table~\ref{tab:guidelines} describes 
+in more detail the metrics behind the guidelines presented. 
+For each type of guideline, three types of impact of patches in 
+the software maintainability are presented: 
 \emph{negative}, maintainability decreases (\emph{red});
@@
 Regarding the impact of security patches for each guideline, we observe
-that overall patches have a significant positive impact in $6$ guidelines: 
-\emph{Write Clean Code} (p-value = $3.66$x$10^{-51}$), \emph{Keep 
-Architecture Components Balanced} (p-value = $5.07$x$10^{-73}$), 
-\emph{Couple Architecture Components Loosely} (p-value = $1.61$x$10^{-08}$), 
-\emph{Separate Concerns in Modules} (p-value = $0.090$), \emph{Keep Unit 
-Interfaces Small} (p-value = $8.76$x$10^{-15}$), \emph{Write Code Once} 
-(p-value = $4.97$x$10^{-35}$). Overall, for each of the $8$ guidelines we also 
-observe a substantial number of cases where the maintainability suffers a 
-negative impact $-$ from $10\%$ to $40\%$ $-$ which highlights the need of 
-better practices/techniques/tools to produce software. However, the results do 
-not provide statistical significance to the \emph{Separate 
-Concerns in Modules} guideline - results should be read carefully.
-
-According to the results, the \emph{Write Simple Units of Code} 
-(p-value=$0.037$ and $372$ ($37.3\%$) of negative cases) and \emph{Write Short 
-Units of Code} (p-value=$0.028$ and $375$ ($37.6\%$) of negative cases) are 
-the most affected guidelines.
-
- For projects with large 
- codebases, the results retrieved for \emph{Keep Your Codebase Small} were not 
- viable because they retrieved values above the limit set by BCH ($20$ Person-years).
- The \emph{Automated Tests} guideline was also not considered because the BCH does not return
- the amount of lines of code that complies with the guideline -- which is one of the variables
- our formula uses to calculate the maintainability of the project regarding each guideline. 
- Thus, we decided to not consider both guidelines in our analysis.
-
- \begin{figure}[h]
+that overall patches have a more positive impact than negative in
+all guidelines/metrics. However, patching vulnerabilities also has a very
+significant negative impact on all guidelines - between $10\%$ and $40\%$. 
+\emph{Write Simple Units of Code} ($37.3\%$), \emph{Write Short Units of 
+Code} ($37.6\%$) and \emph{Separate Concerns in Modules} ($33.2\%$) are the 
+most affected guidelines. This may imply that developers when patching 
+vulnerabilities fail to design/implement solutions that continue to respect 
+the limit bounds of branch points (/programming conditions) and function/module 
+sizes (e.g., functions should not have more than $15$ LOCs). Still, on 
+respecting bound limits, developers also seem to not consider the limit 
+of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
+guideline required by BCH, in $24.5\%$ of the cases. This guideline is usually
+violated when the patch requires to input new information to a function/class 
+and developers fail to use the \emph{Introduce Parameter Object} refactoring 
+pattern. Results do not provide statistical significance to the \emph{Separate 
+Concerns in Modules} guideline, i.e., results should be read carefully.
+
+Software architecture is also affected while patching vulnerabilities.
+Both \emph{Couple Architecture Component Loosely} and \emph{Keep
+Architecture Components Balanced} guidelines suffer a negative impact of 
+$26.7\%$ and $11.1\%$, respectively. Component independence and balance
+are important to make easier to find the source code developers
+want to patch/improve and to understand how the high-level components
+interact with others. However, results may imply that developers
+fail to use techniques such as encapsulation to hide implementation
+details and make the system more modular.
+
+Software systems typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. 
+The \emph{Write Code Once} guideline results show that duplicated code 
+increased in $174$ ($17.4\%$) of the patches. Previous work showed a 
+correlation between code smells and code duplication~\cite{7476787} 
+which may also be reflected in the \emph{Write Clean Code} guideline results. 
+BCH reported new code smells for $152$ ($15.2\%$) patches which according 
+to previous work may harm software security since some of these new code
+smells may be new vulnerabilities~\cite{8819456}. Thus, while developers 
+tend to ignore refactoring techinques such as the \emph{Replace Method 
+with Method Object} technique, they may also be introducing new software 
+vulnerabilities.
+
+For projects with large codebases, the results retrieved for \emph{Keep Your 
+Codebase Small} were not viable because they retrieved values above the limit 
+set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
+not considered because the BCH tool does not return the same type of information
+as for other guidelines. Due to BCH limitations, we detected a few data points
+that retrieved incorrect overall calculations for one version of the software. 
+Those data points were disregarded from the study.
+
+Although overall patching vulnerabilities has a less negative impact, it is 
+important to note that the design debt of one guideline can lead
+to severe impacts on the software quality~\cite{10.1145/1985362.1985366} 
+and in the market value of a company~\cite{4267025}. Thus,
+tools like Better Code Hub and static analysis should be leveraged to 
+assess the risk of patching vulnerabilities - alongside with pull
+requests and code reviews. While style-guided static code analysis
+may be daunting due to the amount of rules and different effects
+on maintainability, BCH ensures that the $10$ chosen guidelines are
+the ones with the highest effect on maintainability. 
+%
+% 
+ \begin{figure}[H]
   	\centering
  	\vspace{-0.3 cm}
-  	\includegraphics[width=0.35\textwidth]{figures/main_per_guideline.pdf}
+  	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
  	\vspace{-0.3 cm}
@@
  	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
- 	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns in Modules \\
+ 	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns 
+	in Modules \\
  	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
- 	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of Code
+ 	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of 
+	Code
  	\end{tabular}
@@
 
-Some of the vulnerabilities are identified with \emph{Common Vulnerabilities 
-and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
+\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
+Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
 Database} (NVD) website to obtain their severity levels. In total, we 
@@
 
-We have also evaluated the impact on the maintainability per programming 
+\textbf{Programming Languages:} We have also evaluated the impact on the maintainability per programming 
 language of open-source software (Figure~\ref{fig:lang_main}). We restrict 
 this analysis to programming languages with at least $20$ data points, as this 
-is a requirement for our hypothesis tests. Thus, we compare the results for \texttt{C/C++}, 
-\texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, \texttt{Python} and \texttt{PHP}. 
-The remaining languages are included into \texttt{Other}.
-
-\texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and \texttt{PHP} ($38.27\%$) are
-the programming languages with a more negative impact on the maintainability 
-of software. \texttt{Java} seems to be the language less affected
-by patching, more than $50\%$ of patches had a positive impact on maintainbility. 
-But overall patches in all languages have a considerable
-negative impact on maintainability - from $30\%$ to $50\%$ - which confirms
-the need for better and more secure programming languages. 
+is a requirement for our hypothesis tests. Thus, we compare the results for 
+\texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
+\texttt{Python} and \texttt{PHP}. The remaining languages are included into 
+\texttt{Other}. \texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and 
+\texttt{PHP} ($38.27\%$) are the programming languages with a more negative
+impact on the maintainability of software. \texttt{Java} seems to be the 
+language less affected by patching, more than $50\%$ of patches had a positive 
+impact on maintainbility. But overall patches in all languages have a 
+considerable negative impact on maintainability - from $30\%$ to $50\%$ - which 
+confirms the need for better and more secure programming languages. 
 Statistical significance was only retrieved for the \texttt{C/C++} 
@@
 Results should be read with care where statistical significance was not found.
-Yet, data reports very interesting hints on the impact of programming languages on 
-security patches.
+Yet, data reports very interesting hints on the impact of programming languages 
+on security patches.
 
@@
 
-Regarding security commits results (Figure~\ref{fig:secvsreg}), we observe that
-maintainability decreases in $41.28\%$ ($412$), remains the same in $20.64\%$ ($206$)
-and increases in $38.08\%$ ($380$) of the cases. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $0.044$.
-Since the p-value is below the significance level of $0.05$, we argue that
-\emph{security patches have a negative impact on the maintainability of
-open-source software}.
+\textbf{Formula:} Maintainability is being calculated as the mean of 
+the guidelines computed by Better Code Hub (BCH). Regarding security 
+commits results (Figure~\ref{fig:secvsreg}), we observe that
+maintainability decreases in $41.28\%$ ($412$), remains the same in $20.64\%$ 
+($206$) and increases in $38.08\%$ ($380$) of the cases. The resulting
+p-value of the Paired Wilcoxon signed-rank test is $0.044$. Since the p-value is 
+below the significance level of $0.05$, we argue that \emph{security patches 
+have a negative impact on the maintainability of open-source software}.
 %
@@
 have negative impact in $5$ weaknesses: Insufficient Control Flow Management 
-(CWE-691), Protection Mechanism Failure (CWE-693), Improper Access Control (CWE-284), 
-Improper Neutralization (CWE-707) and Improper Control of a Resource Through its Lifetime
-(CWE-664). The CWE-707 ($301$) and CWE-664 ($321$) weaknesses integrate a 
-larger amount of cases compared to the remaining weaknesses.
-Thus, we present an analysis of their sub-weaknesses on Figure~\ref{fig:pat}-\emph{b} and
-Figure~\ref{fig:pat}-\emph{c}, respectively. Results demonstrate evidence that patching vulnerabilities hinders 
-the maintainability of open-source software for $5$ different weaknesses: \emph{Improper Input Validation 
-(CWE-20)}, \emph{Information Exposure (CWE-200)}, \emph{Improper Restriction of Operations within the Bounds of 
-a Memory Buffer (CWE-119)}, \emph{Missing Release of Memory after Effective Lifetime 
-(CWE-401)} and \emph{Path Traversal (CWE-22)}. Results also show that software maintainability
-is less negatively impacted when patching \emph{Improper Restriction of XML External Entity Reference 
-(CWE-611)}.
+(CWE-691), Protection Mechanism Failure (CWE-693), Improper Access Control 
+(CWE-284), Improper Neutralization (CWE-707) and Improper Control of a Resource 
+Through its Lifetime (CWE-664). The CWE-707 ($301$) and CWE-664 ($321$) 
+weaknesses integrate a larger amount of cases compared to the remaining 
+weaknesses. Thus, we present an analysis of their sub-weaknesses on 
+Figure~\ref{fig:pat}-\emph{b} and Figure~\ref{fig:pat}-\emph{c}, respectively. 
+Results demonstrate evidence that patching vulnerabilities hinders 
+the maintainability of open-source software for $5$ different weaknesses: 
+\emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
+(CWE-200)}, \emph{Improper Restriction of Operations within the Bounds of 
+a Memory Buffer (CWE-119)}, \emph{Missing Release of Memory after Effective 
+Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)}. Results also show that 
+software maintainability is less negatively impacted when patching 
+\emph{Improper Restriction of XML External Entity Reference (CWE-611)}.
 
@@
 
-\textit{\textbf{RQ3: What is the impact of security patches versus regular patches on the
-maintainability of open-source software?}}
+\textit{\textbf{RQ3: What is the impact of security patches versus regular 
+patches on the maintainability of open-source software?}}
 
@@
 refactorings, variables or functions, type conversion, etc).  
-One example is the regular commit\footnote{Regular change: \url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad322448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated
+One example is the regular commit\footnote{Regular change: 
+\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad32
+2448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated 
 by our methodology and using as reference the following security commit
-\footnote{Security change: \url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
+\footnote{Security change: 
+\url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc
+445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
 overflow vulnerability patch, our methodology chose a similiar
@@
 
-The resulting p-value of the Paired Wilcoxon signed-rank test 
-is $0.331$, i.e., conclusions regarding the regular changes 
-are not statistically significant. Yet, it is important to
-note that \emph{patching security vulnerabilities seem to have a 
-more negative impact on the maintainability of open-source 
-software than performing regular changes}.
+The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
+conclusions regarding the regular changes are not statistically significant. 
+Yet, it is important to note that \emph{patching security vulnerabilities seem 
+to have a more negative impact on the maintainability of open-source software 
+sthan performing regular changes}.
 
@@
 
-\section{Discussion}\label{sec:discussion}
-
-In this section, we discuss the results and answer the proposed research questions.
-
-\textit{\textbf{RQ1: What is the impact of security patches on the maintainability
-of open-source software?}}
-\textit{Patching vulnerabilities may harm the maintainability of open-source software.}
-
-The guideline analysis suggests that all guidelines are affected by security patches. Being the
-\emph{Write Short Units of Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns in Modules} the guidelines with higher negative impact on the maintainability ($>30\%$ of cases).
+% \section{Discussion}\label{sec:discussion}
+%
+% In this section, we discuss the results and answer the proposed research
+% questions.
+%
+% \textit{\textbf{RQ1: What is the impact of security patches on the
+% maintainability of open-source software?}}
+% \textit{Patching vulnerabilities may harm the maintainability of open-source
+% software.}
+%
+% The guideline analysis suggests that all guidelines are affected by security patches. Being the
+% \emph{Write Short Units of Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns in Modules} the guidelines with higher negative impact on the maintainability ($>30\%$ of cases).
 \emph{Keep Architecture Components Balanced} is the guideline with lower negative impact on
-the maintainability (approx. $10\%$). This might imply that developers should give different
-levels of attention to the guidelines when patching vulnerabilities.
-%
-As expected, vulnerabilities with higher severity are more likely to impact negatively
-the maintainability of software. This suggests that high/medium severity vulnerabilities may
-need more attention than low severity when patching vulnerabilities.
-%
-
-We would expect that the negative impact of programming languages on
-maintainability to be more severe, as arguably poor design of programming
-languages and lack of using the best practices lead to more buggy and vulnerable
-code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
-Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
-the result of projects contributions policies. Nine of the top-10 of dataset projects 
-with more contributors have restrict contribution policies regarding code
-standards. Some examples are the \texttt{laravel/framework},
-\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
-instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-Standard}\footnote{PSR-2 Standard available at
-\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
-
-\subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
-affect open-source software maintainability?}}}
-
-\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
-open-source software maintainability.}
-
-Overall, the results obtained for the maintainability per CWE are no statistical significance 
-The impact of a patch depends on its complexity, i.e., if the patchs
-adds complexity to the code base it is probably affecting the software
-maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} (CWE-79)
-patches endure more
-cases without any impact in the maintainability. Typically, these types of
-vulnerabilities do not need extra lines to be fixed, as shown in
-Listing~\ref{lst:fix}. Whereas, \emph{Resource Management Errors} (CWE-399) and vulnerabilities
-such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
-attacks are more difficult to fix (several code lines deleted and added).
-Another example is the Apple authentication flaw discovered in $2017$
-(CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
-\url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
-could log in as root with an empty password. Patrick Wardle examined the cause
-of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
-\url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
-concluded that the flaw was due to the introduction of high cyclomatic complexity
-in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
-developers' preoccupations along with the other patterns that explicitly hinder
-maintainability because despite their low complexity these are patterns that
-still produce a considerable negative impact on the maintainability. Real
-efforts have been made in creating and incorporating coding standards and best
-practices in software production. However, security is still far from perfection
-and fixing vulnerabilities might hinder software maintainability. Thus, it is
-not only important to provide tool support to developers to help them apply these
-patterns without harming software maintainability, but also to create and redesign
-programming languages to facilitate developers improving software maintainability.
-
-We measured the effect sizes between both distribution -- before and after
-patches --- per language and security patterns using the Common Language Effect
-Size~\cite{cliff:1993} method. Effect sizes are of a low-medium level.
-Thus, considering the $0$ median and the low effect sizes, it is reasonable
-to affirm that there is a significant effect but it might only be
-recognized after a thoughtful data analysis.
-
-\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes
-on the maintainability of open-source software?}}}
-\textit{Patching security vulnerabilities has a larger negative and positive impact
-than when fixing regular commits.}
-
-We found that $41.76\%$ of the security patches have a negative impact on the
-maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). While $21.04\%$ of regular commits have a negative impact on the maintainability. This raises
-a new trade-off when developers need to patch vulnerabilities on their projects
-because they might be affecting the software maintainability or even introducing
-new vulnerabilities. This study exhibits evidence that developers may have to
-reduce maintainability for the sake of security. We understand that developers
-should be able to produce secure code without affecting the maintainability of
-their projects. But there are still some concerns that need to be addressed:
-\begin{itemize}
-	\item \textbf{Documentation} of software must provide developers with the
-	best practices to implement secure and maintainable software. Unmaintainable
-	code is difficult to test, analyze and re-use.
-
-	\item\textbf{Programming Languages} should provide new design patterns to
-	easily implement security patterns without endangering maintainability.
-	Ultimately, new programming languages, both secure and maintainable by design,
-	may be designed to help developers be highly productive when writing and
-	maintaining  secure applications. One example is the need for designing new
-	authorization mechanisms since these are one of the most complex security
-	features to implement.
-
-	\item \textbf{Software Developers} lack the knowledge to create secure and
-	maintainable code. Mainly because academic curricula are not yet prepared
-	to properly educate them. Universities should do their part on making
-	this possible. Online services such as BCH have also a great potential to help
-	developers to be more aware of the maintainability issues introduced by their
-	changes. Thus, developers can put more work into improving their code maintainability
-	and avoid common issues such as code duplication.
-
-\end{itemize}
+% the maintainability (approx. $10\%$). This might imply that developers should give different
+% levels of attention to the guidelines when patching vulnerabilities.
+% %
+% As expected, vulnerabilities with higher severity are more likely to impact negatively
+% the maintainability of software. This suggests that high/medium severity vulnerabilities may
+% need more attention than low severity when patching vulnerabilities.
+% %
+% We would expect that the negative impact of programming languages on
+% maintainability to be more severe, as arguably poor design of programming
+% languages and lack of using the best practices lead to more buggy and vulnerable
+% code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
+% Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
+% the result of projects contributions policies. Nine of the top-10 of dataset projects
+% with more contributors have restrict contribution policies regarding code
+% standards. Some examples are the \texttt{laravel/framework},
+% \texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
+% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
+% Standard}\footnote{PSR-2 Standard available at
+% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
+%
+% \subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
+% affect open-source software maintainability?}}}
+%
+%
+% \textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
+% open-source software maintainability.}
+%
+% Overall, the results obtained for the maintainability per CWE are no statistical significance
+% The impact of a patch depends on its complexity, i.e., if the patchs
+% adds complexity to the code base it is probably affecting the software
+% maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} (CWE-79)
+% patches endure more
+% cases without any impact in the maintainability. Typically, these types of
+% vulnerabilities do not need extra lines to be fixed, as shown in
+% Listing~\ref{lst:fix}. Whereas, \emph{Resource Management Errors} (CWE-399) and vulnerabilities
+% such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
+% attacks are more difficult to fix (several code lines deleted and added).
+% Another example is the Apple authentication flaw discovered in $2017$
+% (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
+% \url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
+% could log in as root with an empty password. Patrick Wardle examined the cause
+% of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
+% \url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
+% concluded that the flaw was due to the introduction of high cyclomatic complexity
+% in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
+% developers' preoccupations along with the other patterns that explicitly hinder
+% maintainability because despite their low complexity these are patterns that
+% still produce a considerable negative impact on the maintainability. Real
+% efforts have been made in creating and incorporating coding standards and best
+% practices in software production. However, security is still far from perfection
+% and fixing vulnerabilities might hinder software maintainability. Thus, it is
+% not only important to provide tool support to developers to help them apply these
+% patterns without harming software maintainability, but also to create and redesign
+% programming languages to facilitate developers improving software maintainability.
+%
+% We measured the effect sizes between both distribution -- before and after
+% patches --- per language and security patterns using the Common Language Effect
+% Size~\cite{cliff:1993} method. Effect sizes are of a low-medium level.
+% Thus, considering the $0$ median and the low effect sizes, it is reasonable
+% to affirm that there is a significant effect but it might only be
+% recognized after a thoughtful data analysis.
+%
+%
+% \subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes
+% on the maintainability of open-source software?}}}
+% \textit{Patching security vulnerabilities has a larger negative and positive impact
+% than when fixing regular commits.}
+%
+% We found that $41.76\%$ of the security patches have a negative impact on the
+% maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). While $21.04\%$ of regular commits have a negative impact on the maintainability. This raises
+% a new trade-off when developers need to patch vulnerabilities on their projects
+% because they might be affecting the software maintainability or even introducing
+% new vulnerabilities. This study exhibits evidence that developers may have to
+% reduce maintainability for the sake of security. We understand that developers
+% should be able to produce secure code without affecting the maintainability of
+% their projects. But there are still some concerns that need to be addressed:
+% \begin{itemize}
+% 	\item \textbf{Documentation} of software must provide developers with the
+% 	best practices to implement secure and maintainable software. Unmaintainable
+% 	code is difficult to test, analyze and re-use.
+%
+% 	\item\textbf{Programming Languages} should provide new design patterns to
+% 	easily implement security patterns without endangering maintainability.
+% 	Ultimately, new programming languages, both secure and maintainable by design,
+% 	may be designed to help developers be highly productive when writing and
+% 	maintaining  secure applications. One example is the need for designing new
+% 	authorization mechanisms since these are one of the most complex security
+% 	features to implement.
+%
+% 	\item \textbf{Software Developers} lack the knowledge to create secure and
+% 	maintainable code. Mainly because academic curricula are not yet prepared
+% 	to properly educate them. Universities should do their part on making
+% 	this possible. Online services such as BCH have also a great potential to help
+% 	developers to be more aware of the maintainability issues introduced by their
+% 	changes. Thus, developers can put more work into improving their code maintainability
+% 	and avoid common issues such as code duplication.
+%
+% \end{itemize}
 

### 2020-05-04 paper/icsme20.tex
new version

@@
 %\title{Hey, developers! You should pay more attention to your patches.}
+%\title{Are good coding pratices used when patching software vulnerabilities?}
 
@@
 impact. This empirical study presents evidence that changes 
-applied in the code bases to patch vulnerabilities affect code maintainability. Especially, for patches that address \emph{Improper Input Validation} (CWE-20), \emph{Information 
-Exposure} (CWE-200), \emph{Improper Restriction of Operations within the Bounds of a 
-Memory Buffer} (CWE-119), \emph{Missing Release of Memory after Effective Lifetime} 
-(CWE-401) and \emph{Path Traversal} (CWE-22). This is important work
+applied in the code bases to patch vulnerabilities affect code maintainability. Especially, 
+for patches that address \emph{Improper Restriction of Operations within the Bounds of 
+a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
+after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200).
+This is important work
 because there is still little information about the impact of security patches on the software
@@
 maintainability of open-source software?}}
-\textit{Patching vulnerabilities may harm the maintainability of open-source
-software.}
-
-In this section, we report and discuss the impact of patches on open-source
+
+In RQ1, we report and discuss the impact of patches on open-source
 software maintainability under four different views: guidelines/metrics, 
@@
 guidelines/metrics.
-
-% As expected, vulnerabilities with higher severity are more likely to impact negatively
-% the maintainability of software. This suggests that high/medium severity vulnerabilities may
-% need more attention than low severity when patching vulnerabilities.
-% %
-% We would expect that the negative impact of programming languages on
-% maintainability to be more severe, as arguably poor design of programming
-% languages and lack of using the best practices lead to more buggy and vulnerable
-% code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
-% Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
-% the result of projects contributions policies. Nine of the top-10 of dataset projects
-% with more contributors have restrict contribution policies regarding code
-% standards. Some examples are the \texttt{laravel/framework},
-% \texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
-% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-% Standard}\footnote{PSR-2 Standard available at
-% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
 
@@
 and in the market value of a company~\cite{4267025}. Thus,
-tools like Better Code Hub and static analysis should be leveraged to 
+tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
+ESLint, Infer, and more) should be leveraged to 
 assess the risk of patching vulnerabilities - alongside with pull
-requests and code reviews. While style-guided static code analysis
-may be daunting due to the amount of rules and different effects
-on maintainability, BCH ensures that the $10$ chosen guidelines are
-the ones with the highest effect on maintainability. 
-%
-% 
- \begin{figure}[H]
+requests and code reviews. While static code analysis
+may be daunting due to the number of rules and different effects
+on maintainability, BCH only uses the top-$10$ of the guidelines
+with the highest effect on maintainability. Static analysis 
+can be used to complement the source code analysis, introduce
+the capability of vulnerability detection, and, consequently,
+reduce the risk of hindering maintainability.
+
+Nevertheless, results show that developers may fail to follow the coding
+practice guidelines and hinder software maintainability after patching; and, 
+that different levels of attention should be paid to each guideline. For instance,
+\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
+are the most affected guidelines.
+
+%
+% 
+ \begin{figure}[h]
   	\centering
@@
 assumption that higher severity vulnerabilities patches may have a 
-more negative impact on maintainability.
+more negative impact on maintainability. This suggests that high/medium 
+severity vulnerabilities may need more attention than low severity while 
+patching vulnerabilities which was also observed in previous work but 
+regarding the companies market values~\cite{4267025}.
 
@@
 \texttt{Other}. \texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and 
-\texttt{PHP} ($38.27\%$) are the programming languages with a more negative
+\texttt{PHP} ($38.27\%$) are the programming languages with worse 
 impact on the maintainability of software. \texttt{Java} seems to be the 
@@
 
+We would expect that the negative impact of programming languages on
+maintainability to be more severe, as arguably poor design of programming
+languages and lack of using the best practices lead to more buggy and vulnerable
+code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
+Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
+a significant negative impact approximate to $50\%$ on software 
+maintainability. We suspect that these values are
+the result of projects contributions policies. Nine of the top-10 of dataset projects
+with more contributors have restrict contribution policies regarding code
+standards. Some examples are the \texttt{laravel/framework},
+\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
+instance, \texttt{laravel/framework} requires the \texttt{PSR-2
+Standard}\footnote{PSR-2 Standard available at
+\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
+
 % Following the \textit{American Statistical Association}
@@
 
-\textbf{Formula:} Maintainability is being calculated as the mean of 
+\textbf{Overall Score:} Maintainability is being calculated as the mean of 
 the guidelines computed by Better Code Hub (BCH). Regarding security 
@@
 below the significance level of $0.05$, we argue that \emph{security patches 
-have a negative impact on the maintainability of open-source software}.
+may have a negative impact on the maintainability of open-source software}; and, 
+that tools such as Better Code Hub should be integrated in the CI/CD pipeline
+to help developers evaluate the risk of patches hindering software maintainability
+alongside with Pull Requests/Code Reviews.
 %
@@
 is a type of mistake that, in proper conditions, could contribute to the 
-introduction of vulnerabilities within that product. This enumeration
+introduction of vulnerabilities within that product. The CWE enumeration
 system was used to classify the vulnerabilities included in our study.
@@
 the \emph{Research Concepts} list on CWE; and, the remaining charts
-show two of those first-level weaknesses more in depth reporting the impact 
-on maintainability of their sub-weaknesses (\emph{(b) and (c)}).
+present the impact on maintainability of the sub-weaknesses of two of
+the first-level weaknesses (\emph{(b) and (c)}).
 
@@
 
-\emph{Cross-Site Scripting (CWE-79)} is the weakness that suffers less 
-positive/negative impact from security patches, i.e., patching usually 
-has no impact on the maintainability of open-source software. This can 
-be explained by the fact that patches to this weakness usually are easily 
-done by adding an \texttt{escape} function. Patches of \emph{SQL Injection
+The impact of a patch depends on its complexity, i.e., if the patche
+adds complexity to the code base it is probably affecting the software
+maintainability. \emph{Cross-Site Scripting (CWE-79)} patches endure more
+cases with no impact on the open-source software maintainability.
+Typically, these types of vulnerabilities do not need extra lines 
+to be fixed, as shown in Listing~\ref{lst:fix} - one simple \texttt{escape} 
+function patches the issue.  Whereas, \emph{Missing Release of Memory after Effective 
+Lifetime (CWE-401)} and \emph{Resource Management Errors} (CWE-399) vulnerabilities
+are usually the root cause of Denial-of-Service attacks which are more difficult
+to patch since it usually requires to add more complexity to the program - as
+seen in the example presented on Section~\ref{sec:motivation}.
+
+Patches of \emph{SQL Injection
 (CWE-89)} weaknesses equally hinder and improve the software maintainability. 
@@
 
+\textit{Improper Restriction of Operations within the Bounds of 
+a Memory Buffer (CWE-119), Path Traversal (CWE-22), Missing Release of Memory after Effective 
+Lifetime (CWE-401) and \emph{Information Exposure 
+(CWE-200)} are more likely to affect open-source software maintainability.}
 Statistical significance was only found to the CWE-119 weakness (p-value = 
@@
 
-\textit{\textbf{RQ3: What is the impact of security patches versus regular 
-patches on the maintainability of open-source software?}}
-
-The impact of security and baseline changes in the maintainability of
-open-source software is presented in Figure~\ref{fig:secvsreg}.  
-We have seen, previously, a deterioration on software maintainability 
-when patching vulnerabilities: $41.28\%$ ($412$) suffered a 
-negative impact, $20.64\%$ ($206$) remained the same and $38.08\%$ 
-($380$) increased the software maintainability. For regular changes, 
-we observe that the maintainability decreases in $27.05\%$ ($270$) 
-and increases in $30.26\%$ ($302$) of the cases. But in contrast to 
-security patches, the maintainability of regular changes remains the 
-same in $42.69\%$ ($426$) of the cases and performing regular
-changes has a more positive impact than negative on the software 
-maintainbility.
-
-Results show that regular changes typically have no impact on the 
-software maintainability which contradicts previous research.
-However, we manually inspected a considerable amount of cases from
-the regular changes distribution where this happens and found that 
-identifying regular commits as similar in size as the security-related 
-commit is limiting the type of regular commits (e.g, input 
-refactorings, variables or functions, type conversion, etc).  
-One example is the regular commit\footnote{Regular change: 
-\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad32
-2448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated 
-by our methodology and using as reference the following security commit
-\footnote{Security change: 
-\url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc
-445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
-overflow vulnerability patch, our methodology chose a similiar
-type of change. We strongly believe that this phenomenon lead
-that increased significantly the number of cases with no impact
-on the software maintainability.
-
-The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
-conclusions regarding the regular changes are not statistically significant. 
-Yet, it is important to note that \emph{patching security vulnerabilities seem 
-to have a more negative impact on the maintainability of open-source software 
-sthan performing regular changes}.
-
-\begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
-	\label{fig:secvsreg}
-\end{figure}
-
-% \section{Discussion}\label{sec:discussion}
-%
-% In this section, we discuss the results and answer the proposed research
-% questions.
-%
-% \textit{\textbf{RQ1: What is the impact of security patches on the
-% maintainability of open-source software?}}
-% \textit{Patching vulnerabilities may harm the maintainability of open-source
-% software.}
-%
-% The guideline analysis suggests that all guidelines are affected by security patches. Being the
-% \emph{Write Short Units of Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns in Modules} the guidelines with higher negative impact on the maintainability ($>30\%$ of cases).
-\emph{Keep Architecture Components Balanced} is the guideline with lower negative impact on
-% the maintainability (approx. $10\%$). This might imply that developers should give different
-% levels of attention to the guidelines when patching vulnerabilities.
-% %
-% As expected, vulnerabilities with higher severity are more likely to impact negatively
-% the maintainability of software. This suggests that high/medium severity vulnerabilities may
-% need more attention than low severity when patching vulnerabilities.
-% %
-% We would expect that the negative impact of programming languages on
-% maintainability to be more severe, as arguably poor design of programming
-% languages and lack of using the best practices lead to more buggy and vulnerable
-% code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
-% Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Python} have a negative impact approximate to $50\%$ on software maintainability. We suspect that these values are
-% the result of projects contributions policies. Nine of the top-10 of dataset projects
-% with more contributors have restrict contribution policies regarding code
-% standards. Some examples are the \texttt{laravel/framework},
-% \texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
-% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-% Standard}\footnote{PSR-2 Standard available at
-% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
-%
-% \subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
-% affect open-source software maintainability?}}}
-%
-%
-% \textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
-% open-source software maintainability.}
-%
-% Overall, the results obtained for the maintainability per CWE are no statistical significance
-% The impact of a patch depends on its complexity, i.e., if the patchs
-% adds complexity to the code base it is probably affecting the software
-% maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} (CWE-79)
-% patches endure more
-% cases without any impact in the maintainability. Typically, these types of
-% vulnerabilities do not need extra lines to be fixed, as shown in
-% Listing~\ref{lst:fix}. Whereas, \emph{Resource Management Errors} (CWE-399) and vulnerabilities
-% such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
-% attacks are more difficult to fix (several code lines deleted and added).
 % Another example is the Apple authentication flaw discovered in $2017$
@@
 %
-% We measured the effect sizes between both distribution -- before and after
-% patches --- per language and security patterns using the Common Language Effect
-% Size~\cite{cliff:1993} method. Effect sizes are of a low-medium level.
-% Thus, considering the $0$ median and the low effect sizes, it is reasonable
-% to affirm that there is a significant effect but it might only be
-% recognized after a thoughtful data analysis.
-%
-%
-% \subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes
-% on the maintainability of open-source software?}}}
-% \textit{Patching security vulnerabilities has a larger negative and positive impact
-% than when fixing regular commits.}
-%
-% We found that $41.76\%$ of the security patches have a negative impact on the
-% maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). While $21.04\%$ of regular commits have a negative impact on the maintainability. This raises
-% a new trade-off when developers need to patch vulnerabilities on their projects
-% because they might be affecting the software maintainability or even introducing
-% new vulnerabilities. This study exhibits evidence that developers may have to
-% reduce maintainability for the sake of security. We understand that developers
-% should be able to produce secure code without affecting the maintainability of
-% their projects. But there are still some concerns that need to be addressed:
-% \begin{itemize}
-% 	\item \textbf{Documentation} of software must provide developers with the
-% 	best practices to implement secure and maintainable software. Unmaintainable
-% 	code is difficult to test, analyze and re-use.
-%
-% 	\item\textbf{Programming Languages} should provide new design patterns to
-% 	easily implement security patterns without endangering maintainability.
-% 	Ultimately, new programming languages, both secure and maintainable by design,
-% 	may be designed to help developers be highly productive when writing and
-% 	maintaining  secure applications. One example is the need for designing new
-% 	authorization mechanisms since these are one of the most complex security
-% 	features to implement.
-%
-% 	\item \textbf{Software Developers} lack the knowledge to create secure and
-% 	maintainable code. Mainly because academic curricula are not yet prepared
-% 	to properly educate them. Universities should do their part on making
-% 	this possible. Online services such as BCH have also a great potential to help
-% 	developers to be more aware of the maintainability issues introduced by their
-% 	changes. Thus, developers can put more work into improving their code maintainability
-% 	and avoid common issues such as code duplication.
-%
-% \end{itemize}
+
+\textit{\textbf{RQ3: What is the impact of security changes versus regular 
+changes on the maintainability of open-source software?}}
+
+The impact of security and baseline changes in the maintainability of
+open-source software is presented in Figure~\ref{fig:secvsreg}.  
+We have seen, previously, a deterioration on software maintainability 
+when patching vulnerabilities: $41.28\%$ ($412$) suffered a 
+negative impact, $20.64\%$ ($206$) remained the same and $38.08\%$ 
+($380$) increased the software maintainability. For regular changes, 
+we observe that the maintainability decreases in $27.05\%$ ($270$) 
+and increases in $30.26\%$ ($302$) of the cases. But in contrast to 
+security patches, the maintainability of regular changes remains the 
+same in $42.69\%$ ($426$) of the cases and performing regular
+changes has a more positive impact than negative on the software 
+maintainability.
+
+Results show that regular changes typically have no impact on the 
+software maintainability which contradicts previous research\Sof{missing citation}.
+However, we manually inspected a considerable amount of cases from
+the regular changes distribution where this happens and found that 
+identifying regular commits as similar in size as the security-related 
+commit is limiting the type of regular commits (e.g, input 
+refactorings, variables or functions, type conversion, etc).  
+One example is the regular commit\footnote{Regular change: 
+\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad32
+2448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated 
+by our methodology and using as reference the following security commit
+\footnote{Security change: 
+\url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc
+445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
+overflow vulnerability patch, our methodology chose a similiar
+type of change. We strongly believe that this phenomenon lead
+to the significant number of cases where there is no impact
+on the software maintainability.
+
+Results show that for both types of changes there is a negative impact
+on software maintainability - especially, for security changes. This raises 
+a new trade-off when developers need to patch vulnerabilities on their projects
+because they might be hindering the software maintainability or even introducing
+new vulnerabilities. This study exhibits evidence that developers may have to
+reduce maintainability for the sake of security. We understand that developers
+should be able to patch and produce secure code without hindering the maintainability of
+their projects. But there are still some concerns that need to be addressed
+and that this study tries to bring awareness for:
+
+\begin{itemize}
+
+	\item\textbf{Computer Curricula Needs to be Updated:}
+	Software Developers lack the knowledge to create secure and
+	maintainable code. Mainly because academic curricula is not yet prepared
+	to properly educate them. Universities should do their part on making
+	this possible. Online services such as BCH have also a great potential to help
+	developers to be more aware of the maintainability issues introduced by their
+	changes - alongside with pull requests and code reviews and integrated in 
+	the CI/CD pipeline. Thus, developers can 
+	put more work into improving their code maintainability and avoiding the risk
+	of introducing new vulnerabilities after patching older ones.
+
+	\item\textbf{Secure Programming Languages Wanted:} Programming Languages 
+	should provide new design patterns to easily implement security patterns 
+	without endangering maintainability.
+	Ultimately, new programming languages, both secure and maintainable by design,
+	may be designed to help developers be highly productive when writing and
+	maintaining secure applications. One example is the need for designing new
+	authorization mechanisms since these are one of the most complex security
+	features to implement.
+
+	\item\textbf{Poor Coding Practices Documentation:} 
+	Software documentation must provided to developers with the
+	best practices to implement secure and maintainable software. Unmaintainable
+	code is difficult to test, analyze and re-use. Software companies must
+	do an effort to provide these artifacts to their developers, in order,
+	to avoid the introduction of new vulnerabilities while patching older ones.
+
+\end{itemize}
+
+The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
+conclusions regarding the regular changes are not statistically significant. 
+Yet, it is important to note that \emph{patching security vulnerabilities seem 
+to have a more negative impact on the maintainability of open-source software 
+than performing regular changes}.
+
+\begin{figure}[h]	
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
+ 	\caption{Maintainability difference for security and baseline patches.}
+	\label{fig:secvsreg}
+\end{figure}
 
@@
 the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
+The way how maintainability is being calculated - mean of all guidelines - is
+a preliminar measure to show how the maintainability is overall affected. Nevertheless,
+in some cases the maintainability for the two versions might be the same but have
+different causes, i.e., being affected by different guidelines. 
 %
@@
 has patches that were manually mined in software repositories. Being a manual technique,
-the search might led to
-wrong associations between vulnerabilities and code snippets.
-Baseline commits are retrieved randomly by selecting one commit from the same
+the search might led to wrong associations between vulnerabilities and code snippets. 
+However, we manually inspected all the data points again and disregarded the ones we were
+in doubt. 
+When comparing security changes versus regular changes, our study does not take into consideration that 
+developers may patch the issue quickly and later proceed to refactor.
+We assume that vulnerabilities are patched once or in different parts consecutively.
+Baseline commits are retrieved randomly by selecting one commit with an 
+approximate size from the same
 project of a given security patch. This approach softens the differences
@@
 As future work, this study may be extended in several directions: 
-investigate what are the most frequent vulnerabilities in each programming
-language; investigate how patches impact maintainability for each
-vulnerability and programming language;
-study if there is any correlation between code maintainability and
-the popularity of a software project; expand our methodology with other
+investigate which guidelines affect most the maintainability per
+weakness; check if vulnerability patches are followed by refactoring
+commits or not and how much time does it take to do it; 
+investigate how patches impact maintainability for each
+vulnerability and programming language; expand our methodology with other
 software quality properties; validate these findings with closed or private

### 2020-05-04 scripts/report.py

@@ -7,10 +7,6 @@ import pandas as pd
 import numpy as np
 import matplotlib
 import seaborn as sns
-
-matplotlib.rcParams['font.family'] = 'serif'
-matplotlib.rcParams['mathtext.fontset'] = 'cm'
-
 import matplotlib.pyplot as plt
 from math import sqrt
 import pandas as pd

### 2020-05-04 scripts/stats/chart.py

@@ -1,5 +1,9 @@
 import matplotlib.pyplot as plt
+from matplotlib import rc
+from matplotlib import rcParams
+
 import pandas as pd
+
 
 from scipy.stats import wilcoxon
 import numpy as np
@@ -8,6 +12,9 @@ from . import enum
 from . import tests
 from . import data
 
+rc('text', usetex=True)
+rcParams['font.family'] = 'serif'
+rcParams['mathtext.fontset'] = 'cm'
 
 BAR_WIDTH = 0.25
 
@@ -17,12 +24,12 @@ def set_bars(df, idx):
     plt.barh(idx + 2*BAR_WIDTH, df['neg'], BAR_WIDTH, alpha=0.7, align='center', color='red', label='Negative')
 
 def format_p_value(p):
-    return '<0.001' if float('{:.{}f}'.format(p, 3)) <= 0.000 else '={:.{}f}'.format(p, 3)
+    return '$<$0.001' if float('{:.{}f}'.format(p, 3)) <= 0.000 else '={:.{}f}'.format(p, 3)
 
 def set_ticks(df, idx, fontsize=7):
     plt.yticks(idx + BAR_WIDTH, df['type'], fontsize=fontsize)
     plt.xticks(fontsize=fontsize)
-    plt.gca().set_xticklabels(['{:.0f}%'.format(x*100) for x in plt.gca().get_xticks()])
+    plt.gca().set_xticklabels(['{:.0f}$\%$'.format(x*100) for x in plt.gca().get_xticks()])
 
 def config_report(data, yaxs_size, x, y, plt_config):
     fig = plt.figure(figsize=(x,y))
@@ -182,7 +189,7 @@ def main_per_guideline_chart(reports, df, wilcoxon = True):
         for i, r in test.join(df)[::-1].iterrows():    
             p = format_p_value(r['pvalue'])
             box_text = 'N='+str(sum([r['pos_abs'], r['neg_abs'], r['nul_abs']]))+'\n$\overline{x}$='+ '{:.{}f}'.format(r['mean'], 2) + '\nM=' + '{:.{}f}'.format(r['med'], 2) + '\np' + p
-            ax.text(0.57, boxes_start, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':3}, fontsize=8)
+            ax.text(0.56, boxes_start, box_text , bbox={'facecolor':'white', 'alpha':0.8, 'pad':3}, fontsize=8)
[…]

### 2020-05-04 scripts/stats/enum.py

@@ -1,13 +1,13 @@
 # data.py 
 
-guidelines = {'Write Short Units of Code':'WShortUC', 
-                'Write Simple Units of Code':'WSimpleUC', 
-                'Write Code Once':'WCO', 
-                'Keep Unit Interfaces Small':'KUIS', 
-                'Separate Concerns in Modules':'SCM',
-                'Couple Architecture Components Loosely':'CACL', 
-                'Keep Architecture Components Balanced':'KACB', 
-                'Write Clean Code':'WCC'}
+guidelines = {'Write Short Units of Code':'\\textbf{WShortUC}\nUnit Size',
+                'Write Simple Units of Code':'\\textbf{WSimpleUC}\nMcCabe Complexity',
+                'Write Code Once':'\\textbf{WCO}\nDuplication',
+                'Keep Unit Interfaces Small':'\\textbf{KUIS}\nUnit Interfacing',
+                'Separate Concerns in Modules':'\\textbf{SCM}\nModule Coupling',
+                'Couple Architecture Components Loosely':'\\textbf{CACL}\nComponent Independence',
+                'Keep Architecture Components Balanced':'\\textbf{KACB}\nComponent Balance',
+                'Write Clean Code':'\\textbf{WCC}\nCode Smells'}
 
 languages = {'Java': ['java', 'scala'], 
                 'Python': ['py'], 

### 2020-05-04 paper/icsme20.tex
luis comments added

@@
 open-source software. This is particularly the case for patches that 
-address Improper Input Validation (CWE-20), Information 
-Exposure (CWE-200), Improper Restriction of Operations within the Bounds of a 
-Memory Buffer (CWE-119), Missing Release of Memory after Effective Lifetime
-(CWE-401) and Path Traversal (CWE-22). We conclude that changes to code bases to patch 
+address Improper Restriction of Operations within the Bounds of 
+a Memory Buffer (CWE-119), Path Traversal (CWE-22), Missing Release 
+of Memory after Effective Lifetime (CWE- 401) and Information Exposure 
+(CWE-200). We conclude that changes to code bases to patch 
 vulnerabilities need to be performed with extra care and that security-related 
@@
 a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
-after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200).
+after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200)\Luis{Little information is known about..}.
 This is important work
@@
 methodology used to answer the research questions; section~\ref{sec:results}
-presents the results and section~\ref{sec:discussion} discusses their
+presents the results and discusses their
 implications; section~\ref{sec:threats} enumerates the threats to the validity of
@@
 %
+
 \section{Motivation and Research Questions}\label{sec:motivation}
@@
 %%%%% NEEDS REVIEW
-\Sof{As an example, consider the patch of the TLS state-machine protocol implementation in 
+As an example, consider the patch of the TLS state-machine protocol implementation in 
 OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
-(DoS) attack by crafting TLS messages to trigger memory allocations
-corresponding to large length values. This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-$2016$-$6307$~\footnote{CVE-$2016$-$6307$
-details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6307}
+(DoS) attack by sending excessively large OCSP Status Request extensions
+
+The Online Certificate Status Protocol (OCSP) is a protocol that allows the client to verify the server certificate for revocation without messing with certificate revocation lists.
+Its drawback is that it requires the client to connect to the server CA OCSP server and request the status of the certificate.
+This extension however, enables a TLS server to include its CA OCSP server response in the handshake.
+That is an HTTPS server may periodically run ocsptool (see ocsptool Invocation) to obtain its certificate revocation status and serve it to the clients.
+That way a client avoids an additional connection to the OCSP server.
+
+This vulnerability is listed at the Common
+Vulnerabilities and Exposures dictionary as CVE-$2016$-$6304$~\footnote{CVE-$2016$-$6304$
+details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6304}
 (Accessed on \today{})}. It is amongst the vulnerabilities studied in our
 research. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
-\emph{ssl/statem/statem.c} file\footnote{CVE-$2016$-$6307$ fix available  at
-\url{https://github.com/openssl/openssl/commit/c1ef7c971d0bbf117c3c80f65b5875e2e7b024b1}
+\emph{ssl/t1\_lib.c} file\footnote{CVE-$2016$-$6304$ fix available  at
+\url{https://github.com/openssl/openssl/commit/e408c09bbf7c3057bda4b8d20bec1b3a7771c15b}
 (Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
 negotiation between the two parties. The problem in
-CVE-$2016$-$6307$ is that due to a flaw in the logic of version $1.1.0$,
+CVE-$2016$-$6304$ is that due to a flaw in the logic of version $1.1.0$,
 the memory for a message is allocated too early, prior to the excessive
@@
 increased risk during this period of application crashes due to the lack
-of memory - which would then mean a more serious Denial of Service.}
+of memory - which would then mean a more serious Denial of Service.
 
@@
 \begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
- static SUB_STATE_RETURN read_state_machine(SSL *s){ 
+\\CVE-2016-6304 vulnerability for file ssl/t1\_lib.c},label={lst:vuln}]
+static int ssl_scan_clienthello_tlsext(SSL *s, PACKET *pkt, int *al){ 
  // [snip]
-+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
-+        && !BUF_MEM_grow_clean(s->init_buf, 
-+                                (int)s->s3->tmp.message_size
-+                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
-+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
-+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-+          return SUB_STATE_ERROR;
++  sk_OCSP_RESPID_pop_free(s->tlsext_ocsp_ids, OCSP_RESPID_free);
++  if (PACKET_remaining(&responder_id_list) > 0) {
++     s->tlsext_ocsp_ids = sk_OCSP_RESPID_new_null();
++     if (s->tlsext_ocsp_ids == NULL) {
++        *al = SSL_AD_INTERNAL_ERROR;
++        return 0;
++     }
++  } else {
++     s->tlsext_ocsp_ids = NULL;
++  }
+
+   while (PACKET_remaining(&responder_id_list) > 0) {
+     OCSP_RESPID *id;
+     PACKET responder_id;
+     const unsigned char *id_data;
+     if (!PACKET_get_length_prefixed_2(&responder_id_list, &responder_id) || PACKET_remaining(&responder_id) == 0) {
+          return 0;
+     }
+
+-  if (s->tlsext_ocsp_ids == NULL 
+-      && (s->tlsext_ocsp_ids = 
+-      sk_OCSP_RESPID_new_null()) == NULL) {
+-    *al = SSL_AD_INTERNAL_ERROR;
+-    return 0;
+-  }
+
  // [snip]
@@
 
-\medskip
-\setcounter{lstannotation}{0}
-\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
- int tls_get_message_header(SSL *s, int *mt){   
- // [snip]
-  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
-        SSL3_HM_HEADER_LENGTH;
-- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func3}!*/
--      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
--      goto err;
-   }
- // [snip]
-
-- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func4}!*/
--     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
--     goto err;
-- }
-
-\end{lstlisting}
+% \medskip
+% \setcounter{lstannotation}{0}
+% \begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
+% \\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
+%  static SUB_STATE_RETURN read_state_machine(SSL *s){
+%  // [snip]
+% +   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+% +        && !BUF_MEM_grow_clean(s->init_buf,
+% +                                (int)s->s3->tmp.message_size
+% +                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
+% +         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+% +         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+% +          return SUB_STATE_ERROR;
+%  // [snip]
+%  }
+% \end{lstlisting}
+%
+% \medskip
+% \setcounter{lstannotation}{0}
+% \begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
+% \\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
+%  int tls_get_message_header(SSL *s, int *mt){
+%  // [snip]
+%   l = RECORD_LAYER_get_rrec_length(&s->rlayer) +
+%         SSL3_HM_HEADER_LENGTH;
+% - if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func3}!*/
+% -      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+% -      goto err;
+%    }
+%  // [snip]
+%
+% - if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func4}!*/
+% -     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+% -     goto err;
+% - }
+%
+% \end{lstlisting}
 %
@@
 
-\Sof{The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
-to handle the messages exchanged between peers. When messages are larger than
-what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
-of this resource was being verified very early in the application's flow, using the \texttt{tls\_get\_message\_header} function presented in Listing ~\ref{lst:vuln2} (\ref{lst:func3} and \ref{lst:func4}).
-These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).
-%
-Although checking the maximum size of a message seems an elementary
-problem to repair, the example shows that there was a significant amount of
-changes performed in the code base which produced a negative impact in the
-maintainability. The developer tries to patch the
-code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
-and disrupts
-one of the guidelines proposed by the Software Improvement Group (SIG) for
-building maintainable software~\cite{Visser:2016:OREILLY} $-$
-the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
-OpenSSL maintainability.}
+% \Sof{The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
+% to handle the messages exchanged between peers. When messages are larger than
+% what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
+% of this resource was being verified very early in the application's flow, using the \texttt{tls\_get\_message\_header} function presented in Listing ~\ref{lst:vuln2} (\ref{lst:func3} and \ref{lst:func4}).
+% These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).
+% %
+% Although checking the maximum size of a message seems an elementary
+% problem to repair, the example shows that there was a significant amount of
+% changes performed in the code base which produced a negative impact in the
+% maintainability. The developer tries to patch the
+% code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
+% and disrupts
+% one of the guidelines proposed by the Software Improvement Group (SIG) for
+% building maintainable software~\cite{Visser:2016:OREILLY} $-$
+% the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
+% OpenSSL maintainability.}
 
@@
 The methodology comprises the following steps, as illustrated in
-Figure~\ref{fig:met}:
+Figure~\ref{fig:met}\Luis{could you make the numbers in the figure a bit larger}:
 %
@@
 	Extract relevant data (e.g., owner and name of
-	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and code bases from the combined dataset containing $1335$ security
+	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and 
+	code bases from the combined dataset containing $1335$ security
 	refactorings collected from open-source software available on
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.45\textwidth]{figures/methodology.pdf}
+ 	\centering 	
+	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
  	\caption{Study Methodology}
@@
 %
+\Luis{We already made clear that we ``evaluate the impact..'' }
 To evaluate the impact of security refactorings on the maintainability of
 open-source software, we use a combined dataset of security vulnerabilities which is 
-the outcome of mining/manually inspect $265$ GitHub projects. Reis and Abreu 
+the outcome of mining/manually inspect $265$ GitHub projects\Luis{:mining and manually inspecting:}.
+\Luis{it is not clear the connection between the 267 projects and the works we mention below.
+There should be a sentence here clarifying and making the connection.}.
+Reis and Abreu
 ($2017$) mined open-source
@@
 maintainability guideline for the given severity level (medium, high, and very high).
-\Sof{The compliance for a severity level
-is calculated based on the number of lines of code that comply and not comply
-with the guideline at a given severity level as explained in more
-detail in previous work~\cite{cruz2019energyoriented}.} In our analysis, we compute 
+The compliance for a severity level
+is calculated based on previous work, which calculates the number of lines of code that comply and not comply
+with the guideline at a given severity level~\cite{cruz2019energyoriented}. In our analysis, we compute 
 the difference of maintainability between the security commit ($v_{s}$) and its parent 
@@
 maintainability. We suspect that these values are
-the result of projects contributions policies. Nine of the top-10 of dataset projects
+the result of projects contributions policies. \Luis{In our dataset, nine of the ten projects with more contributors have strict...} Nine of the top-10 of dataset projects
 with more contributors have restrict contribution policies regarding code
@@
 
+\Luis{I did not fully understand this paragraph.. We are mixing programming languages, contribution policies, and number of contributors in the same bucket.}
+\Luis{I think this section is missing more discussion.
+What do you think should be changed in the community to cope with our findings?
+How can practitioners fix security vulnerabilities wihthout hindering maintainability?
+do we need more tools, better programming languages, more strict contribution policies?
+is there a trade-off between secure software and technical debt?
+show us how these findings can change SE}
 % Following the \textit{American Statistical Association}
@@
 
-The impact of a patch depends on its complexity, i.e., if the patche
+The impact of a patch depends on its complexity, i.e., if the patch
 adds complexity to the code base it is probably affecting the software

### 2020-05-04 paper/icsme20.tex
luis comments fix

@@
 a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
-after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200)\Luis{Little information is known about..}.
+after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200)\Luis{Little information is known about..}\Sof{Do you want me to explain what is each vulnerability?}.
 This is important work
@@
 %
-\Luis{this passive voice here sounds a bit weird (imo)}
-The methodology used to measure the impact of security patches on the
-maintainability of open-source software is discussed in this section.
-The methodology comprises the following steps, as illustrated in
-Figure~\ref{fig:met}\Luis{could you make the numbers in the figure a bit larger}:
+In this section, we discuss the methodology used to measure
+the impact of security patches on the maintainbility of open-source
+software. The methodology comprises the following steps, as illustrated in
+Figure~\ref{fig:met}.
 %
@@
 %
-\Luis{We already made clear that we ``evaluate the impact..'' }
-To evaluate the impact of security refactorings on the maintainability of
-open-source software, we use a combined dataset of security vulnerabilities which is 
-the outcome of mining/manually inspect $265$ GitHub projects\Luis{:mining and manually inspecting:}.
-\Luis{it is not clear the connection between the 267 projects and the works we mention below.
-There should be a sentence here clarifying and making the connection.}.
-Reis and Abreu
+We use a combined dataset of security vulnerabilities which is 
+the outcome of mining and manually inspect a total of $265$ GitHub projects for both
+datasets. Reis and Abreu 
 ($2017$) mined open-source
@@
 language support and project size. The final dataset used in this paper comprises
-$998$ security patches. $912$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE) where $59.1\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $40.9\%$ were manually classified following the \texttt{Research Concepts} view map from the CWE's list.
+$998$ security patches. $883$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE) where $62.5\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $37.5\%$ were manually classified following the \texttt{Research Concepts} CWE's list.
 

### 2020-05-04 paper/icsme20.tex
more fixing

@@
 a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
-after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200)\Luis{Little information is known about..}\Sof{Do you want me to explain what is each vulnerability?}.
+after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200)\Luis{Little information is known about..}.
 This is important work
@@
 \end{itemize}
-%
+%-
 This paper is structured as follows: section~\ref{sec:motivation} introduces an
@@
 As an example, consider the patch of the TLS state-machine protocol implementation in 
-OpenSSL\footnoteref{openssl} to address a vulnerability where an attacker could cause a Denial-of-Service
-(DoS) attack by sending excessively large OCSP Status Request extensions
+OpenSSL\footnoteref{openssl} to address multiple memory leaks in \texttt{t1$\_$lib.c} 
+where a remote attacker could cause a Denial-of-Service
+(DoS) attack - through memory consumption - via large OCSP Status Request extensions.
 
@@
 \url{https://github.com/openssl/openssl/commit/e408c09bbf7c3057bda4b8d20bec1b3a7771c15b}
-(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake which is responsible by the
-negotiation between the two parties. The problem in
+(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake 
+which is responsible by the negotiation between the two parties. The problem in
 CVE-$2016$-$6304$ is that due to a flaw in the logic of version $1.1.0$,
@@
 exhaustion if an attacker triggers up a certain message length to be allocated to service a
-connection. In this case, there is only a security impact if the application does not call  \texttt{SSL\_free()} in a timely manner in the
-event that the connection fails;
+connection. In this case, there is only a security impact if the application does not call  
+\texttt{SSL\_free()} in a timely manner in the event that the connection fails;
 or, the application is working in a constrained environment where there
@@
 
-\Sof{SIG defines \emph{Units} as the smallest groups of code that can be maintained
+SIG defines \emph{Units} as the smallest groups of code that can be maintained
 and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and
@@
 somewhere in the green bar it means it is compliant with a low level of
-severity.}
+severity.
 
@@
 commit(s) interleaved with more programming activities, being the later
-commonly called a floss-refactoring. This study considers mainly 1-commit 
+commonly called a floss-refactoring. This study considers mainly $1$-commit 
 patches. Yet, a small percentage of data points had more than one commit 
@@
 which patches had a positive, negative or null impact on the project maintainability.
-
-\Sof{@todo: Don't forget to mention the limitations of BCH and the outliers.}
 
@@
 %
+
 \section{Results \& Discussion}\label{sec:results}
@@
 maintainability. We suspect that these values are
-the result of projects contributions policies. \Luis{In our dataset, nine of the ten projects with more contributors have strict...} Nine of the top-10 of dataset projects
-with more contributors have restrict contribution policies regarding code
-standards. Some examples are the \texttt{laravel/framework},
-\texttt{php/php-src}, \texttt{openssl/openssl} and \texttt{cakephp/cakephp}. For
+the result of projects contributions policies. In our dataset, 
+nine of the ten projects with more contributors have strict contribution policies regarding code
+standards. For
 instance, \texttt{laravel/framework} requires the \texttt{PSR-2
@@
 
-\Luis{I did not fully understand this paragraph.. We are mixing programming languages, contribution policies, and number of contributors in the same bucket.}
-\Luis{I think this section is missing more discussion.
+% \Luis{I think this section is missing more discussion.
 What do you think should be changed in the community to cope with our findings?

### 2020-05-05 paper/icsme20.tex
new update

@@
 
-%%%%% NEEDS REVIEW
 As an example, consider the patch of the TLS state-machine protocol implementation in 
-OpenSSL\footnoteref{openssl} to address multiple memory leaks in \texttt{t1$\_$lib.c} 
+OpenSSL\footnoteref{openssl} to address a memory leak flaw found in the way OpenSSL 
+handled TLS status request extension data during session renegotiation.multiple memory leaks 
 where a remote attacker could cause a Denial-of-Service
 (DoS) attack - through memory consumption - via large OCSP Status Request extensions.
-
-The Online Certificate Status Protocol (OCSP) is a protocol that allows the client to verify the server certificate for revocation without messing with certificate revocation lists.
-Its drawback is that it requires the client to connect to the server CA OCSP server and request the status of the certificate.
-This extension however, enables a TLS server to include its CA OCSP server response in the handshake.
-That is an HTTPS server may periodically run ocsptool (see ocsptool Invocation) to obtain its certificate revocation status and serve it to the clients.
-That way a client avoids an additional connection to the OCSP server.
+The Online Certificate Status Protocol (OCSP) is a protocol that allows the client to 
+verify the server certificate for revocation without messing with certificate revocation 
+lists. This extension however, enables a TLS server to include its response in the handshake.
 
@@
 research. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
-\emph{ssl/t1\_lib.c} file\footnote{CVE-$2016$-$6304$ fix available  at
+\emph{ssl/t1$\_$lib.c} file\footnote{CVE-$2016$-$6304$ fix available  at
 \url{https://github.com/openssl/openssl/commit/e408c09bbf7c3057bda4b8d20bec1b3a7771c15b}
-(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. Every SSL/TLS connection begins with a handshake 
-which is responsible by the negotiation between the two parties. The problem in
-CVE-$2016$-$6304$ is that due to a flaw in the logic of version $1.1.0$,
-the memory for a message is allocated too early, prior to the excessive
-message length check which can lead to a Denial of Service through memory
-exhaustion if an attacker triggers up a certain message length to be allocated to service a
-connection. In this case, there is only a security impact if the application does not call  
-\texttt{SSL\_free()} in a timely manner in the event that the connection fails;
-or, the application is working in a constrained environment where there
-is very little free memory;
-or, the attacker initiates multiple connection attempts such that there
-are multiple connections in a state where memory has been allocated for
-the connection, when \texttt{SSL\_free()} has not yet been called, and there is
-insufficient memory to service the multiple requests. Except in the instance of $1$ above, any Denial Of Service is likely to
-be transitory because as soon as the connection fails the memory is
-subsequently freed again in the \texttt{SSL\_free()} call. However, there is an
-increased risk during this period of application crashes due to the lack
-of memory - which would then mean a more serious Denial of Service.
+(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability.
+
+Every SSL/TLS connection begins with a handshake 
+which is responsible by the negotiation between the two parties. 
+The OSCP Status Request extension allows the client to verify
+the server certificate. The problem in CVE-$2016$-$6304$
+is that if the client sistematically requests a renegotiation,
+while sending large OCSP Status Request extensions each time,
+then there will be unbounded memory growth on the server
+and lead to a Denial-of-Service attack through memory exhaustion.
 
@@
  // [snip]
-+  sk_OCSP_RESPID_pop_free(s->tlsext_ocsp_ids, OCSP_RESPID_free);
++  sk_OCSP_RESPID_pop_free(s->tlsext_ocsp_ids, OCSP_RESPID_free); /*!\annotation{lst:func1}!*/
 +  if (PACKET_remaining(&responder_id_list) > 0) {
@@
 
-%The code changes performed to fix the vulnerability are:
+The code changes performed to fix the vulnerability are:
+
+The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any OCSP$\_$RESPIDs 
+from a previous handshake to prevent unbounded memory growth~\ref{lst:func1}.
 
@@
 research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
-to measure the impact of
-security patches on the maintainability of open-source software. 
+to measure the impact of security patches on the maintainability of open-source software. 
 %
@@
 without adding new lines of code/complexity to the code. A typical fix for 
-the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. The developer added the function \texttt{htmlentities} to escape the data given by the variable
+the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. 
+The developer added the function \texttt{htmlentities} to escape the data given by the variable
 \texttt{\$\_['file']}. \textbf{Knowing which patterns are more likely to increase
@@
 %
+
 Ponta et al. ($2019$) tracked the \url{pivotal.io} website for vulnerabilities 
 from $2014$ to $2019$. For each new vulnerability, the authors manually searched 
-for references to commits involved in the patch on the National Vulnerability Database (NVD) website. However, $70\%$
+for references to commits involved in the patch on the National Vulnerability 
+Database (NVD) website. However, $70\%$
 of the vulnerabilities did not have any references to commits. Thus, the authors
@@
 %
+
 In this study, we focus on computing the
@@
 $998$ security patches. $883$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE) where $62.5\%$ were scraped from the National Vulnerability Dataset (NVD) and the remaining $37.5\%$ were manually classified following the \texttt{Research Concepts} CWE's list.
+Enumeration IDs (CWE) where $62.5\%$ were scraped from the National Vulnerability 
+Dataset (NVD) and the remaining $37.5\%$ were manually classified following the 
+\emph{Research Concepts} CWE's list.
 
@@
 to security fixes) and use them as a baseline.
-
 The baseline dataset uses the security commits dataset as input.
@@
 of some patches, it was not possible to find patches with the exact number of added and 
-deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. The algorithm integrates the following steps:
+deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. 
+The algorithm integrates the following steps:
 
@@
 
-\Sof{The pair of the random commit and its parent is accepted if the \emph{diff} 
-size fits in the range size considered acceptable at the moment. The algorithm 
-starts with a step of $1$. Every $10$ attempts to search for a change with the 
-same \emph{diff} the step increases. Thus, if a change suffered $20$ deletions 
-and $10$ additions, then, in the first $10$ attempts, the algorithm searches 
-for a \emph{diff} with deletions between $19$ and $21$ and additions between 
-$9$ and $11$. After the $10$ first attempts, the algorithm increase the step 
-for $2$ and searches for a \emph{diff} with deletions between $18$ and $22$ 
-and additions between $8$ and $10$. We originate the regular commits from the 
+The pair of the random commit and its parent is accepted if the \emph{diff} 
+size fits in the range size. This range widens every $10$ attempts to
+search for a chanfe with the same size. We originate the regular commits from the 
 security commits to ensure that differences in maintainability are not a 
-consequence of characteristics of different projects.}
+consequence of characteristics of different projects.
 
@@
 \begin{figure}[h]
- 	\centering 	\includegraphics[width=0.44\textwidth]{figures/bch_report.png}
+ 	\centering 	
+	\includegraphics[width=0.44\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
@@
 
-For projects with large codebases, the results retrieved for \emph{Keep Your 
-Codebase Small} were not viable because they retrieved values above the limit 
-set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
-not considered because the BCH tool does not return the same type of information
-as for other guidelines. Due to BCH limitations, we detected a few data points
-that retrieved incorrect overall calculations for one version of the software. 
-Those data points were disregarded from the study.
-
-Although overall patching vulnerabilities has a less negative impact, it is 
-important to note that the design debt of one guideline can lead
-to severe impacts on the software quality~\cite{10.1145/1985362.1985366} 
-and in the market value of a company~\cite{4267025}. Thus,
-tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
-ESLint, Infer, and more) should be leveraged to 
-assess the risk of patching vulnerabilities - alongside with pull
-requests and code reviews. While static code analysis
-may be daunting due to the number of rules and different effects
-on maintainability, BCH only uses the top-$10$ of the guidelines
-with the highest effect on maintainability. Static analysis 
-can be used to complement the source code analysis, introduce
-the capability of vulnerability detection, and, consequently,
-reduce the risk of hindering maintainability.
-
-Nevertheless, results show that developers may fail to follow the coding
-practice guidelines and hinder software maintainability after patching; and, 
-that different levels of attention should be paid to each guideline. For instance,
-\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
-are the most affected guidelines.
-
-%
-% 
  \begin{figure}[h]
@@
  	\vspace{-0.3 cm}
-  	\includegraphics[width=0.45\textwidth]{figures/main_per_guideline.pdf}
+  	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
  	\vspace{-0.3 cm}
@@
  \end{figure}
+
+For projects with large codebases, the results retrieved for \emph{Keep Your 
+Codebase Small} were not viable because they retrieved values above the limit 
+set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
+not considered because the BCH tool does not return the same type of information
+as for other guidelines. Due to BCH limitations, we detected a few data points
+that retrieved incorrect overall calculations for one version of the software. 
+Those data points were disregarded from the study.
+
+Although overall patching vulnerabilities has a less negative impact, it is 
+important to note that the design debt of one guideline can lead
+to severe impacts on the software quality~\cite{10.1145/1985362.1985366} 
+and in the market value of a company~\cite{4267025}. Thus,
+tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
+ESLint, Infer, and more) should be leveraged to 
+assess the risk of patching vulnerabilities - alongside with pull
+requests and code reviews. While static code analysis
+may be daunting due to the number of rules and different effects
+on maintainability, BCH only uses the top-$10$ of the guidelines
+with the highest effect on maintainability. Static analysis 
+can be used to complement the source code analysis, introduce
+the capability of vulnerability detection, and, consequently,
+reduce the risk of hindering maintainability.
+
+Nevertheless, results show that developers may fail to follow the coding
+practice guidelines and hinder software maintainability after patching; and, 
+that different levels of attention should be paid to each guideline. For instance,
+\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
+are the most affected guidelines.
+%
+%
 
@@
 the result of projects contributions policies. In our dataset, 
-nine of the ten projects with more contributors have strict contribution policies regarding code
+nine of the ten projects with more contributors have strsict contribution policies regarding code
 standards. For
@@
   \subfigure[Maintainability difference by first-level weaknesses from the 
-  \textit{Research Concepts} list on Common Weakness Enumeration (CWE)]{\includegraphics[scale=0.4]{figures/main_per_cwe.pdf}}\quad\quad
+  \textit{Research Concepts} list on Common Weakness Enumeration (CWE)]{
+  \includegraphics[scale=0.4]{figures/main_per_cwe.pdf}}\quad\quad
   \subfigure[Maintainability difference by sub-weaknesses of the 
-  \textit{Improper Neutralization} Weakness (CWE-707)]{\includegraphics[scale=0.4]{figures/main_per_cwe_spec_707.pdf}}\quad\quad
+  \textit{Improper Neutralization} Weakness (CWE-707)]{
+  \includegraphics[scale=0.4]{figures/main_per_cwe_spec_707.pdf}}\quad\quad
     \subfigure[Maintainability difference by sub-weaknesses of the 
-	\textit{Improper Control of a Resource Through its Lifetime} Weakness (CWE-664)]{\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
+	\textit{Improper Control of a Resource Through its Lifetime} Weakness (CWE-664)]{
+	\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
 	\label{fig:pat}
@@
 
+Real
+efforts have been made in creating and incorporating coding standards and best
+practices in software production. However, security is still far from perfection
+and fixing vulnerabilities might hinder software maintainability. Thus, it is
+not only important to provide tool support to developers to help them apply these
+patterns without harming software maintainability, but also to create and redesign
+programming languages to facilitate developers improving software maintainability.
+
 % Another example is the Apple authentication flaw discovered in $2017$
@@
 refactorings, variables or functions, type conversion, etc).  
-One example is the regular commit\footnote{Regular change: 
-\url{https://github.com/AFNetworking/AFNetworking/compare/9cde4e45845486bdb46ad32
-2448f93df3b8479bf...3486a008a1fd1348d0f2fe748a5e23933d94647a}} generated 
-by our methodology and using as reference the following security commit
-\footnote{Security change: 
-\url{https://github.com/AFNetworking/AFNetworking/compare/49186d9a3032dc02da1d0cc
-445cbf69b0cfb001b...2b2558a90e7bfb2ef5169b1ca21b91851e5a5b7d}}. From a
-overflow vulnerability patch, our methodology chose a similiar
-type of change. We strongly believe that this phenomenon lead
+We strongly believe that this phenomenon lead
 to the significant number of cases where there is no impact
@@
 
-Results show that for both types of changes there is a negative impact
+For both types of changes, there is a negative impact
 on software maintainability - especially, for security changes. This raises 
@@
 
+\begin{figure}[h]	
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
+ 	\caption{Maintainability difference for security and baseline patches.}
+	\label{fig:secvsreg}
+\end{figure}
+
 The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
@@
 than performing regular changes}.
-
-\begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
-	\label{fig:secvsreg}
-\end{figure}
 
@@
 the search might led to wrong associations between vulnerabilities and code snippets. 
-However, we manually inspected all the data points again and disregarded the ones we were
-in doubt. 
 When comparing security changes versus regular changes, our study does not take into consideration that 

### 2020-05-05 paper/icsme20.tex
more comments addressed
cleaning comment

@@
 a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
-after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200)\Luis{Little information is known about..}.
-This is important work
-because there is still little information about the impact of security patches on the software
+after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200).
+Little information is known about the impact of security patches on the software
 maintainability. With this study, 

### 2020-05-07 paper/icsme20.tex
t2v

@@
 the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
-The way how maintainability is being calculated - mean of all guidelines - is
-a preliminar measure to show how the maintainability is overall affected. Nevertheless,
-in some cases the maintainability for the two versions might be the same but have
-different causes, i.e., being affected by different guidelines. 
+Maintainability is computed as the mean of all guidelines. A patch can
+have the same impact on the overall maintainability of both software versions 
+- non-vulnerable and vulnerable - but not necessarily on the same 
+guidelines.
 %
@@
 not explicit in commits messages. Pontas et. al\cite{10.1109/MSR.2019.00064} dataset
-has patches that were manually mined in software repositories. Being a manual technique,
-the search might led to wrong associations between vulnerabilities and code snippets. 
-When comparing security changes versus regular changes, our study does not take into consideration that 
-developers may patch the issue quickly and later proceed to refactor.
-We assume that vulnerabilities are patched once or in different parts consecutively.
+has patches that were manually mined in software repositories and, consequently,
+wrong associations may have been performed between vulnerabilities and code snippets. 
+The dataset also included floss-refactorings. Both cases were manually inspected
+and deleted from the dataset. Thus, this study does not consider that a
+developer may perform a patch quickly and later proceed to the refactor.
 Baseline commits are retrieved randomly by selecting one commit with an 
-approximate size from the same
-project of a given security patch. This approach softens the differences
+approximate size and from the same project of a given security patch. 
+This approach softens the differences
 that may result from the characteristics of each project. However,
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $998$ regular commits are enough to
+Furthermore, this evaluation considers that $998$ regular commits - any kind
+of commit - are enough to
 alleviate random irregularities in the maintainability differences of the
 baseline. 
-%
 \\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.

### 2020-05-07 paper/icsme20.tex
conclusion update

@@
 significantly affect the software maintainability of open-source software, i.e., 
-developers hinder maintainability when patching vulnerabilities. We found this evidence in
+developers hinder maintainability while patching vulnerabilities. We found this evidence in
 $41.28\%$ of the studied cases. Regarding regular commits, it seems that overall
@@
 considerable amount of cases that harm maintainability ($27.05\%$).
+Regarding security patches, we also observe that some guidelines are more
+affected than others: \emph{Write Simple Units of Code} ($37.3\%$) and
+\emph{Write Short Units of Code} ($37.6\%$) the most affected ones.
 
@@
 in $41.9\%$, $52.5\%$, $42.9\%$, $49.4\%$, and $41.7\%$, respectively. Therefore, 
-we draw attention for the need of better documentation explaining how to use best practices to
-produce secure and maintainable code; new and better programming languages; new
-academic curricula in the fields of security and software quality; and, tools
-such as BCH to help developers predict the effect of their patches.
+we draw attention for the need of better documentation explaining 
+how to use best practices to produce secure and maintainable code; 
+new and better programming languages; new
+academic curricula in the fields of security and software quality; and, 
+tools such as Better Code Hub and efficient static analysis to help 
+developers predict the effect of their patches on software maintainability.
 
@@
 weakness; check if vulnerability patches are followed by refactoring
-commits or not and how much time does it take to do it; 
-investigate how patches impact maintainability for each
-vulnerability and programming language; expand our methodology with other
-software quality properties; validate these findings with closed or private
+commits and how much time does it take to do it; 
+expand our methodology with other software quality properties; 
+validate these findings with closed/private
 software; and, expand this analysis to other quality standards.

### 2020-05-08 paper/icsme20.tex
motivation

@@
 
-As an example, consider the patch of the TLS state-machine protocol implementation in 
-OpenSSL\footnoteref{openssl} to address a memory leak flaw found in the way OpenSSL 
-handled TLS status request extension data during session renegotiation.multiple memory leaks 
-where a remote attacker could cause a Denial-of-Service
+As an example, consider the patch of the Online Certificate Status Procotol (OCSP) 
+implementation in OpenSSL\footnoteref{openssl} to address a memory leak flaw 
+found in the way OpenSSL handled TLS status request extension data during session 
+renegotiation and where a remote attacker could cause a Denial-of-Service
 (DoS) attack - through memory consumption - via large OCSP Status Request extensions.
-The Online Certificate Status Protocol (OCSP) is a protocol that allows the client to 
-verify the server certificate for revocation without messing with certificate revocation 
-lists. This extension however, enables a TLS server to include its response in the handshake.
+OCSP, formally known as the TLS Certificate Status Request extension, is a protocol 
+that allows the client to verify the server certificate for revocation.
 
@@
 Every SSL/TLS connection begins with a handshake 
-which is responsible by the negotiation between the two parties. 
+which is responsible for the negotiation between the two parties. 
 The OSCP Status Request extension allows the client to verify
-the server certificate. The problem in CVE-$2016$-$6304$
-is that if the client sistematically requests a renegotiation,
+the server certificate and enables a TLS server to include
+its response in the handshake. The problem in CVE-$2016$-$6304$
+is that if the client systematically requests a renegotiation,
 while sending large OCSP Status Request extensions each time,
@@
 and lead to a Denial-of-Service attack through memory exhaustion. 
+Applications not enabling OCSP stapling support are not affected.
 

### 2020-05-08 paper/icsme20.tex
polishing motivation

@@
 
-As an example, consider the patch of the Online Certificate Status Procotol (OCSP) 
-implementation in OpenSSL\footnoteref{openssl} to address a memory leak flaw 
+As an example, consider the patch of the TLS state machine protocol implementation
+in OpenSSL\footnoteref{openssl} to address a memory leak flaw 
 found in the way OpenSSL handled TLS status request extension data during session 
-renegotiation and where a remote attacker could cause a Denial-of-Service
-(DoS) attack - through memory consumption - via large OCSP Status Request extensions.
-OCSP, formally known as the TLS Certificate Status Request extension, is a protocol 
-that allows the client to verify the server certificate for revocation.
+renegotiation and where a malicious client could cause a Denial-of-Service
+(DoS) attack via large OCSP Status Request extensions - if OCSP stapling support
+enabled. Online Certificate Status Procotol (OCSP) stapling, formally known as the TLS Certificate 
+Status Request extension, is a standard for checking the revocation status of
+certificates.
 
@@
 its response in the handshake. The problem in CVE-$2016$-$6304$
-is that if the client systematically requests a renegotiation,
-while sending large OCSP Status Request extensions each time,
-then there will be unbounded memory growth on the server
-and lead to a Denial-of-Service attack through memory exhaustion. 
+was a flaw in the logic of OpenSSL, a client 
+can systematically requests renegotiation,
+while sending large OCSP Status Request extensions each time. This
+increases the amount of unbounded memory on the server
+which will eventually lead to a DoS attack through memory exhaustion. 
 Applications not enabling OCSP stapling support are not affected.

### 2020-05-15 paper/icsme20.tex
more updates to the paper

@@
 As an example, consider the patch of the TLS state machine protocol implementation
-in OpenSSL\footnoteref{openssl} to address a memory leak flaw 
-found in the way OpenSSL handled TLS status request extension data during session 
+in OpenSSL\footnoteref{openssl} to address a memory leak flaw found in the way 
+OpenSSL handled TLS status request extension data during session 
 renegotiation and where a malicious client could cause a Denial-of-Service
@@
 (Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. 
-
 Every SSL/TLS connection begins with a handshake 
@@
 its response in the handshake. The problem in CVE-$2016$-$6304$
-was a flaw in the logic of OpenSSL, a client 
-can systematically requests renegotiation,
-while sending large OCSP Status Request extensions each time. This
-increases the amount of unbounded memory on the server
-which will eventually lead to a DoS attack through memory exhaustion. 
+is a flaw in the logic of OpenSSL that does not handle memory efficiently
+when large OCSP Status Request extensions are sent each time a client 
+requests renegotiation. This was possible because the OCSP responses IDs 
+were not released between handshakes. Instead, they would be allocated 
+again and again.
+
+%We remove any OCSP_RESPIDs from a previous handshake
+%to prevent unbounded memory growth - CVE-2016-6304
+
+If 
+a malicious client does it several times it may lead to an unbounded memory
+growth on the server.
+
+Applications 
+which eventually lead to a DoS attack through memory exhaustion. 
 Applications not enabling OCSP stapling support are not affected.
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-\\CVE-2016-6304 vulnerability for file ssl/t1\_lib.c},label={lst:vuln}]
+\begin{lstlisting}[style={CStyle}, caption={Patch provided by OpenSSL developers to the
+\\CVE-2016-6304 vulnerability on file ssl/t1\_lib.c},label={lst:vuln}]
 static int ssl_scan_clienthello_tlsext(SSL *s, PACKET *pkt, int *al){ 
@@
 +  sk_OCSP_RESPID_pop_free(s->tlsext_ocsp_ids, OCSP_RESPID_free); /*!\annotation{lst:func1}!*/
-+  if (PACKET_remaining(&responder_id_list) > 0) {
++  if (PACKET_remaining(&responder_id_list) > 0) { /*!\annotation{lst:func2}!*/
 +     s->tlsext_ocsp_ids = sk_OCSP_RESPID_new_null();
@@
 -      && (s->tlsext_ocsp_ids = 
--      sk_OCSP_RESPID_new_null()) == NULL) {
+-      sk_OCSP_RESPID_new_null()) == NULL) { /*!\annotation{lst:func3}!*/
 -    *al = SSL_AD_INTERNAL_ERROR;
@@
 
-The code changes performed to fix the vulnerability are:
-
-The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any OCSP$\_$RESPIDs 
-from a previous handshake to prevent unbounded memory growth~\ref{lst:func1}.
+The code changes performed to patch the vulnerability are:
+
+~\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any OCSP$\_$RESPIDs 
+from a previous handshake to prevent unbounded memory growth.
 
@@
 work on measuring the impact of energy-oriented 
-refactorings~\cite{cruz2019energyoriented}. Luis et al. ($2019$) proposed an 
+refactorings~\cite{cruz2019energyoriented}. Cruz et al. ($2019$) proposed an 
 equation to capture the distance between the current state of the project and
@@
 set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
-not considered because the BCH tool does not return the same type of information
-as for other guidelines. Due to BCH limitations, we detected a few data points
+not considered for this study because the BCH tool does not contemplate two
+of the most important techniques to security testing: vulnerability scanning
+and pentration testing. Instead, it only contemplates unit 
+testing.
+Due to BCH limitations, we detected a few data points
 that retrieved incorrect overall calculations for one version of the software. 
@@
 should be carefully monitored by programmers since there is a high correlation
-between maintainability aspects and proneness to changes and faults. In 2019, Luis et 
+between maintainability aspects and proneness to changes and faults. In 2019, Cruz et 
 al.~\cite{cruz2019energyoriented} proposed a formula to calculate maintainability 

### 2020-05-15 paper/icsme20.tex
motivation on hold

@@
 
-As an example, consider the patch of the TLS state machine protocol implementation
+\Sof{Waiting for final results --}As an example, consider the patch of the TLS state machine protocol implementation
 in OpenSSL\footnoteref{openssl} to address a memory leak flaw found in the way 
@@
 % OpenSSL maintainability.}
+
+\Sof{-- Waiting for final results}
 

### 2020-05-20 paper/icsme20.tex
new paper version

--- 2020-05-15 paper/icsme20.tex
@@
 guidelines using a dataset including 1335 security commits.
-Results show that fixing software vulnerabilities may hinder the maintainability of 
-open-source software. This is particularly the case for patches that 
-address Improper Restriction of Operations within the Bounds of 
-a Memory Buffer (CWE-119), Path Traversal (CWE-22), Missing Release 
-of Memory after Effective Lifetime (CWE- 401) and Information Exposure 
-(CWE-200). We conclude that changes to code bases to patch 
+Results show that fixing software vulnerabilities may hinder the maintainability 
+of open-source software and that different levels of attention should be given to 
+different severity levels, programming languages and types of weaknesses. We 
+conclude that changes to code bases to patch 
 vulnerabilities need to be performed with extra care and that security-related 
@@
 impact. This empirical study presents evidence that changes 
-applied in the code bases to patch vulnerabilities affect code maintainability. Especially, 
-for patches that address \emph{Improper Restriction of Operations within the Bounds of 
-a Memory Buffer} (CWE-119), \emph{Path Traversal} (CWE-22), \emph{Missing Release of Memory 
-after Effective Lifetime} (CWE-401) and \emph{Information Exposure} (CWE-200).
+applied in the code bases to patch vulnerabilities affect code maintainability.
+Results also show that different levels of attention should be given to different
+severity levels, programming languages and types of weaknesses.
 Little information is known about the impact of security patches on the software
@@
 OpenSSL handled TLS status request extension data during session 
-renegotiation and where a malicious client could cause a Denial-of-Service
+renegotiation, and where a malicious client could cause a Denial-of-Service
 (DoS) attack via large OCSP Status Request extensions - if OCSP stapling support
-enabled. Online Certificate Status Procotol (OCSP) stapling, formally known as the TLS Certificate 
+was enabled. Online Certificate Status Procotol (OCSP) stapling, formally known as the TLS Certificate 
 Status Request extension, is a standard for checking the revocation status of
@@
 were not released between handshakes. Instead, they would be allocated 
-again and again.
+again and again. Thus, if a malicious client does it several times it may lead to 
+an unbounded memory growth on the server and, eventually, lead to a 
+DoS attack through memory exhaustion.
 
@@
 %to prevent unbounded memory growth - CVE-2016-6304
-
-If 
-a malicious client does it several times it may lead to an unbounded memory
-growth on the server.
-
-Applications 
-which eventually lead to a DoS attack through memory exhaustion. 
-Applications not enabling OCSP stapling support are not affected.
 
@@
 +  sk_OCSP_RESPID_pop_free(s->tlsext_ocsp_ids, OCSP_RESPID_free); /*!\annotation{lst:func1}!*/
-+  if (PACKET_remaining(&responder_id_list) > 0) { /*!\annotation{lst:func2}!*/
++  if (PACKET_remaining(&responder_id_list) > 0) { 
 +     s->tlsext_ocsp_ids = sk_OCSP_RESPID_new_null();
-+     if (s->tlsext_ocsp_ids == NULL) {
++     if (s->tlsext_ocsp_ids == NULL) { /*!\annotation{lst:func2}!*/
 +        *al = SSL_AD_INTERNAL_ERROR;
@@
 
-The code changes performed to patch the vulnerability are:
-
-~\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any OCSP$\_$RESPIDs 
-from a previous handshake to prevent unbounded memory growth.
+The code changes performed to patch the vulnerability were:
+
+~\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any
+memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded memory growth - which was not being perfomed before.
+After releasing the unbounded memory, the program performs the same check 
+in ~\ref{lst:func2} that used to perform in ~\ref{lst:func3} which is responsible 
+for handling the application when no OCSP response IDS were being allocated. Now, instead of performing it after packaging processing, it does it previously.
+
+Although performing this patch, seems an elementary problem to repair,
+the example shows that there was a significant amount of changes performed
+in the code base which performed a negative impact in the software maintanibility.
+While patching, the developer introduced $6$ new lines in a method already with 
+a large number of lines of code and introduced more complexity to the code
+with new branch points which disrupts two of the guidelines proposed by the Software Improvement Group (SIG) for
+building maintainable software~\cite{Visser:2016:OREILLY} - \emph{Write Short Units of Code}
+and \emph{Write Simple Units of Code}.
 
@@
 language support and project size. The final dataset used in this paper comprises
-$998$ security patches. $883$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE) where $62.5\%$ were scraped from the National Vulnerability 
-Dataset (NVD) and the remaining $37.5\%$ were manually classified following the 
+$969$ security patches. $867$ of these patches are identified with Common Weakness
+Enumeration IDs (CWE) where $62.1\%$ were scraped from the National Vulnerability 
+Dataset (NVD) and the remaining $37.9\%$ were manually classified following the 
 \emph{Research Concepts} CWE's list.
@@
 size fits in the range size. This range widens every $10$ attempts to
-search for a chanfe with the same size. We originate the regular commits from the 
+search for a change with the same size. We originate the regular commits from the 
 security commits to ensure that differences in maintainability are not a 
@@
 	\includegraphics[width=0.44\textwidth]{figures/bch_report.png}
- 	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
+ 	\caption{Maintainability report of OpenSSL's CVE-$2016$-$6304$ vulnerability
 refactoring for the guideline \emph{Write Simple Units of Code} provided by
@@
 provided by BCH for a project after finishing its evaluation. The example
-refers to the OpenSSL CVE-$2014$-$3506$ vulnerability refactoring ---
+refers to the OpenSSL CVE-$2016$-$6304$ vulnerability refactoring -
 as described by Section~\ref{sec:motivation}. This
@@
 
-This study evaluates a final total of $998$ security and baseline commits 
-from $262$ distinct open-source projects. The following section presents and 
-discusses the results for each question. Research questions are also answered.
+This study evaluates a final total of $969$ security and baseline commits 
+from $260$ distinct open-source projects. The following section presents and 
+discusses the results and answers to each research question.
 
@@
 
-In RQ1, we report and discuss the impact of patches on open-source
-software maintainability under four different views: guidelines/metrics, 
-severity, programming language and formula proposed - the mean of all 
-guidelines/metrics.
+In \emph{RQ1}, we report and discuss the impact of patches on open-source
+software maintainability under two different views: guidelines/metrics 
+and formula proposed - the mean of all 
+guidelines/metrics - $M(v)$.
 
@@
 Figure~\ref{fig:guidelines} shows the impact of security patches 
-on each guideline individually. Under each guideline is stated the 
+on each guideline individually and the average impact on all guidelines 
+together - $M(v)$. Under each guideline is stated the 
 metric used to the calculations. Table~\ref{tab:guidelines} describes 
@@
 \emph{none}, maintainability remains the same (\emph{yellow}); and,
-\emph{positive}, maintainability increases (\emph{green}). Next to each
+\emph{positive}, maintainability increases (\emph{green}). 
+For each type of guideline, a boxplot 
+and a swarmplot are presented to show the variability and dipersion of
+the results alongside the number of absolute
+and relative cases of each impact. Next to each
 type of guideline, it is presented the mean ($\overline{x}$) and median (M)
 of the maintainability difference and the p-value resulting from the
-Paired Wilcoxon signed-rank test.
+Paired Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
+average impact of all guidelines.
 
@@
 significant negative impact on all guidelines - between $10\%$ and $40\%$. 
-\emph{Write Simple Units of Code} ($37.3\%$), \emph{Write Short Units of 
-Code} ($37.6\%$) and \emph{Separate Concerns in Modules} ($33.2\%$) are the 
-most affected guidelines. This may imply that developers when patching 
+\emph{Write Short Units of 
+Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
+most affected guidelines. 
+This may imply that developers when patching 
 vulnerabilities fail to design/implement solutions that continue to respect 
 the limit bounds of branch points (/programming conditions) and function/module 
-sizes (e.g., functions should not have more than $15$ LOCs). Still, on 
+sizes (e.g., functions should not have more than $15$ LOCs) that are recommended
+for coding best practices. Still, on 
 respecting bound limits, developers also seem to not consider the limit 
 of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
-guideline required by BCH, in $24.5\%$ of the cases. This guideline is usually
+guideline required by BCH, in $24.8\%$ of the cases. This guideline is usually
 violated when the patch requires to input new information to a function/class 
@@
 Architecture Components Balanced} guidelines suffer a negative impact of 
-$26.7\%$ and $11.1\%$, respectively. Component independence and balance
-are important to make easier to find the source code developers
+$26.9\%$ and $11.0\%$, respectively. Component independence and balance
+are important to make easier to find the source code that developers
 want to patch/improve and to understand how the high-level components
@@
 The \emph{Write Code Once} guideline results show that duplicated code 
-increased in $174$ ($17.4\%$) of the patches. Previous work showed a 
+increased in $171$ ($17.7\%$) of the patches. Previous work showed a 
 correlation between code smells and code duplication~\cite{7476787} 
 which may also be reflected in the \emph{Write Clean Code} guideline results. 
-BCH reported new code smells for $152$ ($15.2\%$) patches which according 
+BCH reported new code smells for $145$ ($15.0\%$) patches which according 
 to previous work may harm software security since some of these new code
-smells may be new vulnerabilities~\cite{8819456}. Thus, while developers 
+smells may be new vulnerabilities~\cite{8819456}.
+Thus, while developers 
 tend to ignore refactoring techinques such as the \emph{Replace Method 
 with Method Object} technique, they may also be introducing new software 
-vulnerabilities.
-
- \begin{figure}[h]
-  	\centering
- 	\vspace{-0.3 cm}
-  	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
- 	\vspace{-0.3 cm}
-  	\caption{Impact of the security patches per guideline.}
- 	\label{fig:guidelines}
- 	\vspace{0.2 cm}
- 	\scalebox{0.5}{
- 	\begin{tabular}{r@{: }l r@{: }l}
- 	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
- 	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns 
-	in Modules \\
- 	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
- 	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of 
-	Code
- 	\end{tabular}
- 	}	
- \end{figure}
+vulnerabilities and, consequently, harm the market value and economy of companies
+~\cite{4267025}.
 
@@
 set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
-not considered for this study because the BCH tool does not contemplate two
+not considered since the tool does not contemplate two
 of the most important techniques to security testing: vulnerability scanning
-and pentration testing. Instead, it only contemplates unit 
-testing.
-Due to BCH limitations, we detected a few data points
-that retrieved incorrect overall calculations for one version of the software. 
+and penetration testing. Instead, it only contemplates unit 
+testing. Due to BCH limitations, we detected a few data points
+that retrieved incorrect overall calculations. 
 Those data points were disregarded from the study.
 
-Although overall patching vulnerabilities has a less negative impact, it is 
+\textbf{Overall Score ($M(v)$):} 
+Although overall patching vulnerabilities has a less negative impact
+on software maintainability guidelines, this is not reflected in the average 
+impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
+The $M(v)$ plots show that $406$ ($41.9\%$) of the cases have negative impact
+on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7$)
+have a negative impact. The larger number of negative cases may be 
+explained by guidelines with higher concentration of negative cases with higher amplitudes
+such as WShortUC, WSimpleUC and SCM. In other words, it is 
 important to note that the design debt of one guideline can lead
-to severe impacts on the software quality~\cite{10.1145/1985362.1985366} 
-and in the market value of a company~\cite{4267025}. Thus,
-tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
+to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
+Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
 ESLint, Infer, and more) should be leveraged to 
@@
 
-Nevertheless, results show that developers may fail to follow the coding
-practice guidelines and hinder software maintainability after patching; and, 
-that different levels of attention should be paid to each guideline. For instance,
-\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
-are the most affected guidelines.
-%
-%
+ \begin{figure}[H]
+  	\centering
+  	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
+ 	\vspace{-0.8 cm}
+  	\caption{Impact of the security patches per guideline.}
+ 	\label{fig:guidelines}
+ 	\vspace{0.2 cm}
+ 	\scalebox{0.5}{
+ 	\begin{tabular}{r@{: }l r@{: }l}
+ 	\textbf{WCC} & Write Clean Code & \textbf{KACB} & Keep Architecture Components Balanced\\
+ 	\textbf{CACL} & Couple Architecture Components Loosely & \textbf{SCM} & Separate Concerns
+	in Modules \\
+ 	\textbf{KUIS} & Keep Unit Interfaces Small & \textbf{WCO} & Write Code Once \\
+ 	\textbf{WSimpleUC} & Write Simple Units of Code & \textbf{WShortUC} & Write Short Units of
+	Code
+ 	\end{tabular}
+ 	}
+ 	\vspace{-0.3 cm}
+ \end{figure}
 
@@
 Database} (NVD) website to obtain their severity levels. In total, we 
-retrieved severity scores for $551$ vulnerabilities: $117$ (\emph{HIGH}), 
-$404$ (\emph{MEDIUM}) and $30$ (\emph{LOW}). Figure~\ref{fig:severity} 
+retrieved severity scores for $536$ vulnerabilities: $112$ (\emph{HIGH}), 
+$395$ (\emph{MEDIUM}) and $29$ (\emph{LOW}). Figure~\ref{fig:severity} 
 presents the impact of security patches per severity level on the 
 maintainability of open-source software. We observe that patches for 
-\emph{HIGH} ($47.9\%$) and \emph{MEDIUM} ($44.1\%$) severity vulnerabilities 
-hinder more the maintainability of software than \emph{LOW} ($30.0\%$) 
+\emph{HIGH} ($50.0\%$) and \emph{MEDIUM} ($43.8\%$) severity vulnerabilities 
+hinder more the maintainability of software than \emph{LOW} ($27.6\%$) 
 severity vulnerabilities. Again, patches have a considerable 
-negative impact on software maintainability - between $30\%$ 
-and $50\%$. None statistical significance was retrieved. However, 
+negative impact on software maintainability - between $20\%$ 
+and $50\%$. Statistical significance was retrieved only for \emph{LOW}
+severity vulnerabilitie, i.e., \emph{LOW} severity vulnerabilities
+may have less negative impact on software maintainability thant the 
+other severity levels. However, 
 results should not be disregarded because they seem to confirm the 
 assumption that higher severity vulnerabilities patches may have a 
-more negative impact on maintainability. This suggests that high/medium 
+more negative impact on maintainability and, consequently, that high/medium 
 severity vulnerabilities may need more attention than low severity while 
-patching vulnerabilities which was also observed in previous work but 
-regarding the companies market values~\cite{4267025}.
+patching vulnerabilities.
 
@@
 
-\textbf{Programming Languages:} We have also evaluated the impact on the maintainability per programming 
-language of open-source software (Figure~\ref{fig:lang_main}). We restrict 
+\textbf{Programming Languages:} The impact on software maintainability per programming 
+language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
 this analysis to programming languages with at least $20$ data points, as this 
-is a requirement for our hypothesis tests. Thus, we compare the results for 
+is a requirement for our hypothesis tests (HT). Thus, we compare the results for 
 \texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
-\texttt{Python} and \texttt{PHP}. The remaining languages are included into 
-\texttt{Other}. \texttt{C/C++} ($46.54\%$), \texttt{Ruby} ($46.48\%$) and 
-\texttt{PHP} ($38.27\%$) are the programming languages with worse 
-impact on the maintainability of software. \texttt{Java} seems to be the 
-language less affected by patching, more than $50\%$ of patches had a positive 
-impact on maintainbility. But overall patches in all languages have a 
-considerable negative impact on maintainability - from $30\%$ to $50\%$ - which 
+\texttt{Python} and \texttt{PHP}. \texttt{Groovy} samples are not presented
+because the number of cases did not fulfill the HT requirement. 
+\texttt{C/C++} ($46.5\%$), \texttt{Ruby} ($46.5\%$) and 
+\texttt{PHP} ($38.6\%$) are the programming languages with worse 
+impact on software maintainability. \texttt{Java} ($50.9\%$) and \texttt{Python} 
+($46.5\%$) seem to be the 
+languages less affected by patching, i.e., more than $45\%$ of patches had a positive 
+impact on maintainability. But overall patches in all languages have a 
+considerable negative impact on maintainability - between $35\%$ to $50\%$ - which 
 confirms the need for better and more secure programming languages. 
 Statistical significance was only retrieved for the \texttt{C/C++} 
-(p-value = $2.24$x$10^{-05}$) and \texttt{Ruby} (p-value = $0.041$) languages.
+(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
 Results should be read with care where statistical significance was not found.
@@
 on security patches.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
-  \caption{Maintainability difference by programming language.}
-  \label{fig:lang_main}  
-\end{figure}
 
@@
 maintainability. We suspect that these values are
-the result of projects contributions policies. In our dataset, 
-nine of the ten projects with more contributors have strsict contribution policies regarding code
-standards. For
+the result of projects contributions policies (e.g., coding standards). In our dataset, 
+$9$/$10$ projects with more contributors have strict contribution 
+policies regarding code standards. For
 instance, \texttt{laravel/framework} requires the \texttt{PSR-2
@@
 \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
+  \caption{Maintainability difference by programming language.}
+  \label{fig:lang_main}  
+\end{figure}
+
+\textbf{Conclusion:} Results show that developers may fail to follow the coding
+practice guidelines and hinder software maintainability after patching; and, 
+that different levels of attention should be paid to each guideline. For instance,
+\emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
+are the most affected ones. No statistical significance was retrieved
+for \emph{Separate Concerns in Modules}. Different severity vulnerabilities
+may need different levels of attention and better and more secure programming languages
+are needed.
+The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
+is $0.044$. Since the p-value is 
+below the significance level of $0.05$, we argue that \emph{security patches 
+may have a negative impact on the maintainability of open-source software}; and, 
+that tools such as Better Code Hub should be integrated in the CI/CD pipeline
+to help developers evaluate the risk of patches hindering software maintainability
+alongside with Pull Requests/Code Reviews.
 
@@
 % Ruby as statistically significant because its p-value is on the borderline.
-
-\textbf{Overall Score:} Maintainability is being calculated as the mean of 
-the guidelines computed by Better Code Hub (BCH). Regarding security 
-commits results (Figure~\ref{fig:secvsreg}), we observe that
-maintainability decreases in $41.28\%$ ($412$), remains the same in $20.64\%$ 
-($206$) and increases in $38.08\%$ ($380$) of the cases. The resulting
-p-value of the Paired Wilcoxon signed-rank test is $0.044$. Since the p-value is 
-below the significance level of $0.05$, we argue that \emph{security patches 
-may have a negative impact on the maintainability of open-source software}; and, 
-that tools such as Better Code Hub should be integrated in the CI/CD pipeline
-to help developers evaluate the risk of patches hindering software maintainability
-alongside with Pull Requests/Code Reviews.
 %
@@
 
-\emph{Weakness}, according to Common Weakness Enumeration (CWE) glossary, 
+\emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
 is a type of mistake that, in proper conditions, could contribute to the 
@@
 the \emph{Research Concepts} list on CWE; and, the remaining charts
-present the impact on maintainability of the sub-weaknesses of two of
-the first-level weaknesses (\emph{(b) and (c)}).
-
-Considering the first-level weaknesses from the \emph{Research Concepts}
-list (Figure~\ref{fig:pat}-\emph{a}), we observe that patching vulnerabilities
-have negative impact in $5$ weaknesses: Insufficient Control Flow Management 
+present the impact on maintainability of the sub-weaknesses from two of
+the weaknesses presented in \emph{(a)}.
+
+In Figure~\ref{fig:pat}-\emph{a}, we observe that patching vulnerabilities
+such as Insufficient Control Flow Management 
 (CWE-691), Protection Mechanism Failure (CWE-693), Improper Access Control 
 (CWE-284), Improper Neutralization (CWE-707) and Improper Control of a Resource 
-Through its Lifetime (CWE-664). The CWE-707 ($301$) and CWE-664 ($321$) 
+Through its Lifetime (CWE-664) may have a negative impact on software 
+maintainability. The CWE-707 ($295$) and CWE-664 ($318$) 
 weaknesses integrate a larger amount of cases compared to the remaining 
@@
 Results demonstrate evidence that patching vulnerabilities hinders 
-the maintainability of open-source software for $5$ different weaknesses: 
+the maintainability of open-source software for $4$ different weaknesses: 
 \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
-(CWE-200)}, \emph{Improper Restriction of Operations within the Bounds of 
-a Memory Buffer (CWE-119)}, \emph{Missing Release of Memory after Effective 
+(CWE-200)}, \emph{Missing Release of Memory after Effective 
 Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)}. Results also show that 
@@
 adds complexity to the code base it is probably affecting the software
-maintainability. \emph{Cross-Site Scripting (CWE-79)} patches endure more
+maintainability. \emph{Cross-Site Scripting (CWE-79)} and 
+\emph{Improper Restriction of Operations within the Bounds of 
+a Memory Buffer (CWE-119)} patches endure more
 cases with no impact on the open-source software maintainability.
-Typically, these types of vulnerabilities do not need extra lines 
-to be fixed, as shown in Listing~\ref{lst:fix} - one simple \texttt{escape} 
-function patches the issue. Whereas, \emph{Missing Release of Memory after Effective 
-Lifetime (CWE-401)} and \emph{Resource Management Errors} (CWE-399) vulnerabilities
-are usually the root cause of Denial-of-Service attacks which are more difficult
-to patch since it usually requires to add more complexity to the program - as
-seen in the example presented on Section~\ref{sec:motivation}.
-
+Typically, CWE-79 vulnerabilities do not need extra lines 
+to be fixed, as shown in Listing~\ref{lst:fix} - one simple 
+\texttt{escape} function patches the issue. On the same type of fix,
+CWE-199 vulnerabilities may also be fixed without adding new source code
+(e.g., replacing the \texttt{strncpy} function by a more secure one 
+\texttt{strlcpy} that checks if the buffer is null terminated). However,
+some buffer overflows may be harder to fix and lead to more complex 
+solutions such as the one performed for 
+CVE-$2016$-$0799$\footnote{CVE-$2016$-$0799$ patch details available at 
+\url{https://github.com/openssl/openssl/commit/9cb177301fdab492e4cfef376b28339afe3ef663} (Accessed on \today{})}.
+As CWE-199 
+vulnerabilities, \emph{Missing Release of Memory after Effective 
+Lifetime (CWE-401)}, or also called \emph{Resource Management Errors (CWE-399)}, can also be the root cause of Denial-of-Service attacks and  
+difficult to patch since it usually requires to add more complexity to the program 
+- as seen in the example presented on Section~\ref{sec:motivation}. 
 Patches of \emph{SQL Injection
 (CWE-89)} weaknesses equally hinder and improve the software maintainability. 
-But overall a very considerable percentage of patches hinder the open-source 
-software maintainability - between $30\%$ and $60\%$.
-
-\textit{Improper Restriction of Operations within the Bounds of 
-a Memory Buffer (CWE-119), Path Traversal (CWE-22), Missing Release of Memory after Effective 
-Lifetime (CWE-401) and \emph{Information Exposure 
-(CWE-200)} are more likely to affect open-source software maintainability.}
-Statistical significance was only found to the CWE-119 weakness (p-value = 
-$0.041$). However, results obtained to the other weaknesses should not be 
-disregarded because they offer an interesting view on how the maintainability 
-is impacted per weakness which can provide insights to developers/researchers
-on the weaknesses that need more attention.
+But a very considerable percentage of patches hinder the open-source 
+software maintainability ($41.1\%$).
+
+\emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
+(CWE-200)}, \emph{Missing Release of Memory after Effective 
+Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the most likely 
+weaknesses to affect open-source software maintainability.
+No statistical evidence was retrieved mostly due to the difficult
+of finding a higher numbers of cases of open-source patches per weakness. However, 
+results should not be disregarded because they offer an interesting view on how 
+the maintainability is impacted per weakness which can provide insights to 
+developers/researchers on the weaknesses that need more attention. Further
+research should be perfomed to retrieve conclusions.
 
 Real
-efforts have been made in creating and incorporating coding standards and best
+efforts have been performed in creating and incorporating coding standards and best
 practices in software production. However, security is still far from perfection
@@
 not only important to provide tool support to developers to help them apply these
-patterns without harming software maintainability, but also to create and redesign
+patches without harming software maintainability, but also to create and redesign
 programming languages to facilitate developers improving software maintainability.
-
-% Another example is the Apple authentication flaw discovered in $2017$
-% (CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
-% \url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
-% could log in as root with an empty password. Patrick Wardle examined the cause
-% of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
-% \url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
-% concluded that the flaw was due to the introduction of high cyclomatic complexity
-% in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
-% developers' preoccupations along with the other patterns that explicitly hinder
-% maintainability because despite their low complexity these are patterns that
-% still produce a considerable negative impact on the maintainability. Real
-% efforts have been made in creating and incorporating coding standards and best
-% practices in software production. However, security is still far from perfection
-% and fixing vulnerabilities might hinder software maintainability. Thus, it is
-% not only important to provide tool support to developers to help them apply these
-% patterns without harming software maintainability, but also to create and redesign
-% programming languages to facilitate developers improving software maintainability.
-%
 
@@
 We have seen, previously, a deterioration on software maintainability 
-when patching vulnerabilities: $41.28\%$ ($412$) suffered a 
-negative impact, $20.64\%$ ($206$) remained the same and $38.08\%$ 
-($380$) increased the software maintainability. For regular changes, 
-we observe that the maintainability decreases in $27.05\%$ ($270$) 
-and increases in $30.26\%$ ($302$) of the cases. But in contrast to 
+when patching vulnerabilities: $41.9\%$ ($406$) suffered a 
+negative impact, $38.7\%$ ($375$) remained the same and $19.4\%$ 
+($188$) increased the software maintainability. For regular changes, 
+we observe that the maintainability decreases in $27.0\%$ ($262$) 
+and increases in $30.5\%$ ($295$) of the cases. But in contrast to 
 security patches, the maintainability of regular changes remains the 
-same in $42.69\%$ ($426$) of the cases and performing regular
+same in $42.5\%$ ($412$) of the cases and performing regular
 changes has a more positive impact than negative on the software 
@@
 
-Results show that regular changes typically have no impact on the 
-software maintainability which contradicts previous research\Sof{missing citation}.
+Results show that regular changes typically have less negative impact
+on the software maintainability of open-source software.
+%no impact on the software maintainability which contradicts previous research\Sof{missing citation}.
 However, we manually inspected a considerable amount of cases from
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $998$ regular commits - any kind
+Furthermore, this evaluation considers that $969$ regular commits - any kind
 of commit - are enough to
@@
 
-This work presents an empirical study on the impact of $998$ security
-patches on the maintainability of $262$ open-source projects. We leveraged
+This work presents an empirical study on the impact of $969$ security
+patches on the maintainability of $260$ open-source projects. We leveraged
 Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018, cruz2019energyoriented}. Security patches 
@@
 developers hinder maintainability while patching vulnerabilities. We found this evidence in
-$41.28\%$ of the studied cases. Regarding regular commits, it seems that overall
-maintainability stays the same ($42.69\%$). However, there is still a 
-considerable amount of cases that harm maintainability ($27.05\%$).
+$41.9\%$ of the studied cases. When comparing with the impact of regular changes, security patches seem to have a higher negative impact than regular commits.
 Regarding security patches, we also observe that some guidelines are more
-affected than others: \emph{Write Simple Units of Code} ($37.3\%$) and
-\emph{Write Short Units of Code} ($37.6\%$) the most affected ones.
-
-This study also shows evidence that there are security patterns that need more
-attention than others. \emph{Improper Input Validation} (CWE-20), \emph{Information 
-Exposure} (CWE-200), \emph{Improper Restriction of Operations within the Bounds of a 
-Memory Buffer} (CWE-119), \emph{Missing Release of Memory after Effective Lifetime} 
-(CWE-401) and \emph{Path Traversal} (CWE-22) endure maintainability
-in $41.9\%$, $52.5\%$, $42.9\%$, $49.4\%$, and $41.7\%$, respectively. Therefore, 
+affected than others: \emph{Write Simple Units of Code} ($37.9\%$) and
+\emph{Write Short Units of Code} ($38.3\%$).
+
+The current study also hints that more severe vulnerabilities may have
+a more negative impact on the software maintainability and that patches
+is some programming languages may needed more attention than others (e.g., \texttt{C/C++}, \texttt{Ruby} and \texttt{PHP}. Therefore, 
 we draw attention for the need of better documentation explaining 
@@
 academic curricula in the fields of security and software quality; and, 
-tools such as Better Code Hub and efficient static analysis to help 
+tools such as Better Code Hub and efficient static analysis to integrate
+in the CI/CD pipelines to help 
 developers predict the effect of their patches on software maintainability.

### 2020-05-25 paper/icsme20.tex
intro/abstract

@@
 However, there is still a considerable amount of vulnerabilities being
-discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
+discovered and fixed, almost weekly, as disclosed by the Zero Day Initiative
 website\footnote{Zero Day Initiative website available at
 \url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
-These vulnerabilities are gateways for attackers to penetrate systems and 
-exploit their resources. As an example, at the beginning of $2020$, academics published a 
-new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
-security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
-address in previous generations and that makes every user using a CPU released before Q$4$ $2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
-at the cost of features and/or performance.
+%These vulnerabilities are gateways for attackers to penetrate systems and 
+%exploit their resources. As an example, at the beginning of $2020$, academics published a 
+%new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
+%security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
+%address in previous generations and that makes every user using a CPU released before Q$4$ 
+%$2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
+%at the cost of features and/or performance.
 %
@@
 \emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
-However, many developers still lack the knowledge about best practices to deliver secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
+However, many developers still lack the knowledge about best practices to deliver secure and 
+high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
+\Rui{link between security and maintainability is still obscure.}
 Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
@@
 implementing patches that might affect software maintainability. 
-Our suspiciousness is that some of the these patches may have a negative 
+Our \Rui{wrong word -->} suspiciousness is that some of the these patches may have a negative 
 impact on the software maintainability and, possibly, even be the cause of the 
@@
 We use BCH's toolset to provide an assessment of maintainability in 
-software mainly due to the broad number of metrics used.
+software mainly due to the broad number of metrics used. \Rui{this paragraph is boring and in cicles. revise.}
 
@@
 applied in the code bases to patch vulnerabilities affect code maintainability.
-Results also show that different levels of attention should be given to different
+\Rui{Sentence is too criptic -->} Results also show that different levels of attention should be given to different
 severity levels, programming languages and types of weaknesses.

### 2020-05-25 paper/icsme20.tex
intro/abstract

@@
   maintainability and what patterns need more attention.
+  \item \Rui{results, conclusions}
 	\item A replication package with all scripts and data created to perform the
@@
 security flaws are generally detected \textit{after the fact}, i.e., when
-hackers exploit them. It turns out that fixing these issues is as simple as
+hackers exploit them. \Rui{strange sentence -->} It turns out that fixing these issues is as simple as
 modifying the software code base. However, these patches may have a
@@
 %%%%%%% 
-Software maintainability is designed as the degree to which an application is understood, 
+Software maintainability is \Rui{designed??} designed as the degree to which an application is understood, 
 repaired or enhanced. In this paper, our concern is to study whether, while improving software
@@
 to security fixes) and use them as a baseline.
-The baseline dataset uses the security commits dataset as input.
+\Rui{strange sentence:} The baseline dataset uses the security commits dataset as input.
 Both datasets have the same size. For each
-security commit, a commit with an approximate size is searched. Due to the complexity 
+security commit, a commit with an approximate size is fetched. Due to the complexity 
 of some patches, it was not possible to find patches with the exact number of added and 
@@
 on each guideline individually and the average impact on all guidelines 
-together - $M(v)$. Under each guideline is stated the 
+together - $M(v)$. \Rui{doesn't parse -->} Under each guideline is stated the 
 metric used to the calculations. Table~\ref{tab:guidelines} describes 
-in more detail the metrics behind the guidelines presented. 
+in more detail the metrics behind the guidelines. 
 For each type of guideline, three types of impact of patches in 
@@
 
+\Rui{the next couple of sentences read contradictory...}
 Regarding the impact of security patches for each guideline, we observe
@@
 \emph{Write Short Units of 
-Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
+Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and 
+\emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
 most affected guidelines. 
+\Rui{The next sentence should be in a morer  abstract way.. too much detail, no?}
 This may imply that developers when patching 
@@
 explained by guidelines with higher concentration of negative cases with higher amplitudes
-such as WShortUC, WSimpleUC and SCM. In other words, it is 
+such as \Rui{these acronyms came out of the blue.} WShortUC, WSimpleUC and SCM. In other words, it is 
 important to note that the design debt of one guideline can lead
@@
 \emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
-is a type of mistake that, in proper conditions, could contribute to the 
+is a type of mistake that, \Rui{meaning?} in proper conditions, could contribute to the 
 introduction of vulnerabilities within that product. The CWE enumeration

### 2020-05-25 paper/icsme20.tex
small update
new comments adressed + luis comments added

@@
 \usepackage{subfigure}
+\usepackage{changepage}
+
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
@@
 
-% \title{Fixing Vulnerabilities Hinders Maintainability}
-\title{\Sof{TBA - New Titles in the comments}}
+\title{Fixing Vulnerabilities Potentially Hinders Maintainability}
 %\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
 %\title{The reason why you should be applying coding practices: safer software.}
-%\title{What is the impact of security patches in software evolution?}
+%\title{What is the impact of  corollarysecurity patches in software evolution?}
 %\title{Hey, developers! You should pay more attention to your patches.}
@@
 vulnerabilities need to be performed with extra care and that security-related 
-code evolution should be subject in computer curricula.
+code evolution should be subject in computer curricula.\Rui{Needs improvement}
 \end{abstract}
@@
 on each guideline individually and the average impact on all guidelines 
-together - $M(v)$. \Rui{doesn't parse -->} Under each guideline is stated the 
+together - $M(v)$. \Rui{doesn't parse -->} Under each guideline, it is stated the 
 metric used to the calculations. Table~\ref{tab:guidelines} describes 
 in more detail the metrics behind the guidelines. 
-For each type of guideline, three types of impact of patches in 
-the software maintainability are presented: 
-\emph{negative}, maintainability decreases (\emph{red});
-\emph{none}, maintainability remains the same (\emph{yellow}); and,
-\emph{positive}, maintainability increases (\emph{green}). 
 For each type of guideline, a boxplot 
@@
 impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
-The $M(v)$ plots show that $406$ ($41.9\%$) of the cases have negative impact
-on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7$)
+The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
+on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7\%$) cases
 have a negative impact. The larger number of negative cases may be 
-explained by guidelines with higher concentration of negative cases with higher amplitudes
-such as \Rui{these acronyms came out of the blue.} WShortUC, WSimpleUC and SCM. In other words, it is 
-important to note that the design debt of one guideline can lead
-to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
-Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
-ESLint, Infer, and more) should be leveraged to 
-assess the risk of patching vulnerabilities - alongside with pull
-requests and code reviews. While static code analysis
-may be daunting due to the number of rules and different effects
-on maintainability, BCH only uses the top-$10$ of the guidelines
-with the highest effect on maintainability. Static analysis 
-can be used to complement the source code analysis, introduce
-the capability of vulnerability detection, and, consequently,
-reduce the risk of hindering maintainability.
-
- \begin{figure}[H]
+explained by guidelines with higher concentration of negative cases with higher amplitudes, 
+such as, \emph{Write Short Units of 
+Code}, \emph{Write Simple Units of 
+Code} and \emph{Separate Concerns in Modules}.
+
+% For each type of guideline, three types of impact of patches in
+% the software maintainability are presented:
+% \emph{negative}, maintainability decreases (\emph{red});
+% \emph{none}, maintainability remains the same (\emph{yellow}); and,
+% \emph{positive}, maintainability increases (\emph{green}).
+
+ \begin{figure}[htp]
+     \begin{adjustwidth}{-2.5cm}{-1cm}
+
   	\centering
-  	\includegraphics[width=0.5\textwidth]{figures/main_per_guideline.pdf}
- 	\vspace{-0.8 cm}
+  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
   	\caption{Impact of the security patches per guideline.}
  	\label{fig:guidelines}
- 	\vspace{0.2 cm}
- 	\scalebox{0.5}{
- 	\begin{tabular}{r@{: }l r@{: }l}
- 	\textbf{WCC} & Write Clean Code & \textbf{KACB} & Keep Architecture Components Balanced\\
- 	\textbf{CACL} & Couple Architecture Components Loosely & \textbf{SCM} & Separate Concerns
-	in Modules \\
- 	\textbf{KUIS} & Keep Unit Interfaces Small & \textbf{WCO} & Write Code Once \\
- 	\textbf{WSimpleUC} & Write Simple Units of Code & \textbf{WShortUC} & Write Short Units of
-	Code
- 	\end{tabular}
- 	}
- 	\vspace{-0.3 cm}
+    \end{adjustwidth}
+ 	\vspace{-0.5 cm}
+
  \end{figure}
@@
 Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
-Database} (NVD) website to obtain their severity levels. In total, we 
+Database} (NVD) website to collect their severity levels. In total, we 
 retrieved severity scores for $536$ vulnerabilities: $112$ (\emph{HIGH}), 
@@
 this analysis to programming languages with at least $20$ data points, as this 
-is a requirement for our hypothesis tests (HT). Thus, we compare the results for 
+is a requirement for our hypothesis tests. Thus, we compare the results for 
 \texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
 \texttt{Python} and \texttt{PHP}. \texttt{Groovy} samples are not presented
-because the number of cases did not fulfill the HT requirement. 
+because the number of cases did not fulfill the statistical test requirement. 
 \texttt{C/C++} ($46.5\%$), \texttt{Ruby} ($46.5\%$) and 
@@
 (p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
-Results should be read with care where statistical significance was not found.
+The remaining programming languages did not yield statistical significance, 
+thus results should be read with care.
 Yet, data reports very interesting hints on the impact of programming languages 
@@
 
-\textbf{Conclusion:} Results show that developers may fail to follow the coding
+\textbf{Conclusion:} Results show that developers fail to follow the coding
 practice guidelines and hinder software maintainability after patching; and, 
@@
 
+\Luis{
+Can we say somehting on why some vulbnerabilities are more affect than others? Complexity? Hard to understand? Good  Documentation resources?
+this is valid for RQ1 as well. What are the implciations of results in these different vulneratbilities?}
+
 \textit{\textbf{RQ3: What is the impact of security changes versus regular 
@@
 For both types of changes, there is a negative impact
-on software maintainability - especially, for security changes. This raises 
+on software maintainability $-$ especially, for security changes. This raises 
 a new trade-off when developers need to patch vulnerabilities on their projects
 because they might be hindering the software maintainability or even introducing
-new vulnerabilities. This study exhibits evidence that developers may have to
+new vulnerabilities.
+
+\begin{figure}[h]	
+ 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
+ 	\caption{Maintainability difference for security and baseline patches.}
+	\label{fig:secvsreg}
+\end{figure}
+
+The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
+conclusions regarding the regular changes are not statistically significant. 
+Yet, it is important to note that \emph{patching security vulnerabilities seem 
+to have a more negative impact on the maintainability of open-source software 
+than performing regular changes}.
+
+\section{Study Implications}\label{sec:implications}
+
+ In other words, it is 
+important to note that the design debt of one guideline can lead
+to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
+Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
+ESLint, Infer, and more) should be leveraged to 
+assess the risk of patching vulnerabilities - alongside with pull
+requests and code reviews. While static code analysis
+may be daunting due to the number of rules and different effects
+on maintainability, BCH only uses the top-$10$ of the guidelines
+with the highest effect on maintainability. Static analysis 
+can be used to complement the source code analysis, introduce
+the capability of vulnerability detection, and, consequently,
+reduce the risk of hindering maintainability.
+
+This study exhibits evidence that developers may have to
 reduce maintainability for the sake of security. We understand that developers
@@
 \end{itemize}
-
-\begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
-	\label{fig:secvsreg}
-\end{figure}
-
-The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
-conclusions regarding the regular changes are not statistically significant. 
-Yet, it is important to note that \emph{patching security vulnerabilities seem 
-to have a more negative impact on the maintainability of open-source software 
-than performing regular changes}.
 

### 2020-05-27 paper/icsme20.tex
comments addressed

@@
 In $2011$, the International Organization for Standardization (ISO) issued an
-update for software product quality ISO/IEC 25010 considering
-\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
-However, many developers still lack the knowledge about best practices to deliver secure and 
-high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
-\Rui{link between security and maintainability is still obscure.}
-Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
-to find vulnerabilities and defects. Therefore, writing maintainable code
-supports the production of more reliable software preventing the introduction of 
-faults/vulnerabilities in the code bases. ISO describes software 
-maintainability as ``the degree of effectiveness and
+update for software product quality ISO/IEC 25010 considering \emph{Security} 
+as one of the main software product quality characteristics~\cite{iso:2011}. 
+However, many developers still lack the knowledge about best practices to deliver 
+secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
+ISO describes software maintainability as ``the degree of effectiveness and
 efficiency with which a software product or system can be modified to improve
 it, correct it or adapt it to changes in environment, and in requirements'' on
-software quality ISO/IEC 25010~\cite{iso:2011}. Novel tools have been built to 
+software quality ISO/IEC 25010~\cite{iso:2011}. 
+Unmaintainable code is hard to test, analyze and re-use. Therefore, 
+it is important to write maintainable code to support the production 
+of more secure software and prevent 
+the introduction of new vulnerabilities in the code bases.
+
+Novel tools have been built to 
 detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
@@
 implementing patches that might affect software maintainability. 
-Our \Rui{wrong word -->} suspiciousness is that some of the these patches may have a negative 
+Our suspicion is that some of these patches may have a negative 
 impact on the software maintainability and, possibly, even be the cause of the 
@@
 with a set of $10$ guidelines/metrics to produce quality software 
-based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
-were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
+based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. 
 There are other well-known standards and models that have been proposed
@@
 first steps taken by SIG to introduce security on their maintainability model). 
-%
-We use BCH's toolset to provide an assessment of maintainability in 
-software mainly due to the broad number of metrics used. \Rui{this paragraph is boring and in cicles. revise.}
+Nevertheless, our study uses BCH's toolset to provide an assessment of maintainability in 
+software for the following reasons: 1) BCH considers a broad number of code metrics; and, 2)
+code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}.
 
 From a methodological 
-perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
+perspective, we leveraged a dataset of $1335$ security patches/refactorings collected from open-source 
 software. We calculate software maintainability before and after the patch to measure its 
@@
 applied in the code bases to patch vulnerabilities affect code maintainability.
-\Rui{Sentence is too criptic -->} Results also show that different levels of attention should be given to different
-severity levels, programming languages and types of weaknesses.
+Results also hint that developers should pay different levels of attention 
+to different severity levels and classes of weaknesses when patching vulnerabilities. We also
+show that patches in programming languages such as, \emph{C/C++}, \emph{Ruby} and \emph{PHP}, 
+may have a more negative impact in software maintainability.
 Little information is known about the impact of security patches on the software
@@
 \begin{itemize}
+  \item Evidence that supports a new trade-off between security and maintainability:
+  when developers need to patch vulnerabilities, they might be hindering the software
+  maintainability or even introducing new vulnerabilities.
 	\item An empirical study on the impact of security changes on software
-  maintainability and what patterns need more attention.
-  \item \Rui{results, conclusions}
+  maintainability and what weaknesses, severity levels and programming languages need more attention.
 	\item A replication package with all scripts and data created to perform the
@@
 security flaws are generally detected \textit{after the fact}, i.e., when
-hackers exploit them. \Rui{strange sentence -->} It turns out that fixing these issues is as simple as
-modifying the software code base. However, these patches may have a
+hackers exploit them. Patching these issues is usually done by modifying the code base. 
+However, these patches may have a
 negative impact on software maintenance --- e.g., developers might follow the
@@
 %%%%%%% 
-Software maintainability is \Rui{designed??} designed as the degree to which an application is understood, 
+Software maintainability is designated as the degree to which an application is understood, 
 repaired or enhanced. In this paper, our concern is to study whether, while improving software
@@
 
-\textit{\textbf{RQ2: Which patterns of security patches are more likely to
+\textit{\textbf{RQ2: Which weaknesses are more likely to
 affect open-source software maintainability?}}
 There are security flaws that are more difficult to patch than others. For
-instance, implementing secure authentication is not as easy as fixing a
+instance, implementing secure authentication is not as easy as patching a
 cross-site scripting vulnerability since usually the last one can be fixed
@@
 The developer added the function \texttt{htmlentities} to escape the data given by the variable
-\texttt{\$\_['file']}. \textbf{Knowing which patterns are more likely to increase
+\texttt{\$\_['file']}. \textbf{Knowing which weaknesses are more likely to increase
 maintainability issues is one step forward to bring awareness to security
-engineers of what patterns need more attention}. The taxonomy of security
+engineers of what weaknesses need more attention}. The taxonomy of security
 patterns used to answer this question is the one provided by the Common Weakness Enumeration
-(CWE). Maintainability is measured
-separately for each pattern.
+(CWE). \emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
+is a type of code-flaw that could contribute to the 
+introduction of vulnerabilities within that product. Maintainability is measured
+separately for each weakness.
 %
@@
 to security fixes) and use them as a baseline.
-\Rui{strange sentence:} The baseline dataset uses the security commits dataset as input.
-Both datasets have the same size. For each
+The baseline dataset is generated from the security commits dataset.
+For each
 security commit, a commit with an approximate size is fetched. Due to the complexity 
@@
 on each guideline individually and the average impact on all guidelines 
-together - $M(v)$. \Rui{doesn't parse -->} Under each guideline, it is stated the 
+together - $M(v)$. Under each guideline, it is stated the 
 metric used to the calculations. Table~\ref{tab:guidelines} describes 
@@
 
-\Rui{the next couple of sentences read contradictory...}
 Regarding the impact of security patches for each guideline, we observe
-that overall patches have a more positive impact than negative in
-all guidelines/metrics. However, patching vulnerabilities also has a very
+that overall patches have more cases where the impact is positive. However, 
+we also see that patching vulnerabilities also has a very
 significant negative impact on all guidelines - between $10\%$ and $40\%$. 
@@
 \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
-most affected guidelines. 
-\Rui{The next sentence should be in a morer  abstract way.. too much detail, no?}
+most negativelly affected guidelines. 
 This may imply that developers when patching 
 vulnerabilities fail to design/implement solutions that continue to respect 
-the limit bounds of branch points (/programming conditions) and function/module 
-sizes (e.g., functions should not have more than $15$ LOCs) that are recommended
-for coding best practices. Still, on 
+the limit bounds of branch points and function/module 
+sizes that are recommended by coding best practices. Still, on 
 respecting bound limits, developers also seem to not consider the limit 
@@
 
-% For each type of guideline, three types of impact of patches in
-% the software maintainability are presented:
-% \emph{negative}, maintainability decreases (\emph{red});
-% \emph{none}, maintainability remains the same (\emph{yellow}); and,
-% \emph{positive}, maintainability increases (\emph{green}).
-
  \begin{figure}[htp]
      \begin{adjustwidth}{-2.5cm}{-1cm}  
-
   	\centering
@@
     \end{adjustwidth}	
- 	\vspace{-0.5 cm}
-
+	\vspace{-0.5cm}
  \end{figure}
@@
 are the most affected ones. No statistical significance was retrieved
-for \emph{Separate Concerns in Modules}. Different severity vulnerabilities
-may need different levels of attention and better and more secure programming languages
-are needed.
+for \emph{Separate Concerns in Modules}. 
 The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
@@
 to help developers evaluate the risk of patches hindering software maintainability
-alongside with Pull Requests/Code Reviews.
+alongside with Pull Requests/Code Reviews. We also conclude that different severity 
+vulnerabilities may need different levels of attention and better and more secure 
+programming languages are needed.
 
@@
 
-\emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
-is a type of mistake that, \Rui{meaning?} in proper conditions, could contribute to the 
-introduction of vulnerabilities within that product. The CWE enumeration
-system was used to classify the vulnerabilities included in our study.
+In \emph{RQ2}, we report and discuss the impact of security patches on open-source
+software maintainability per weakness. In this study, we use weakness definition 
+mentioned in Section~\ref{sec:motivation} and the CWE enumeration
+system to classify the vulnerabilities.
 Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
@@
 
-Real
-efforts have been performed in creating and incorporating coding standards and best
-practices in software production. However, security is still far from perfection
-and fixing vulnerabilities might hinder software maintainability. Thus, it is
-not only important to provide tool support to developers to help them apply these
-patches without harming software maintainability, but also to create and redesign
-programming languages to facilitate developers improving software maintainability.
-
-\Luis{
-Can we say somehting on why some vulbnerabilities are more affect than others? Complexity? Hard to understand? Good  Documentation resources?
-this is valid for RQ1 as well. What are the implciations of results in these different vulneratbilities?}
-
 \textit{\textbf{RQ3: What is the impact of security changes versus regular 
@@
 on the software maintainability of open-source software.
-%no impact on the software maintainability which contradicts previous research\Sof{missing citation}.
 However, we manually inspected a considerable amount of cases from
@@
 For both types of changes, there is a negative impact
-on software maintainability $-$ especially, for security changes. This raises 
+on software maintainability $-$ especially, for security changes. 
+The resulting p-value of the Paired Wilcoxon signed-rank test is $0.300$, i.e., 
+conclusions regarding the regular changes are not statistically significant. 
+Yet, it is important to note that \emph{patching security vulnerabilities seem 
+to have a more negative impact on the maintainability of open-source software 
+than performing regular changes}.
+This raises 
 a new trade-off when developers need to patch vulnerabilities on their projects
@@
 
-The resulting p-value of the Paired Wilcoxon signed-rank test is $0.331$, i.e., 
-conclusions regarding the regular changes are not statistically significant. 
-Yet, it is important to note that \emph{patching security vulnerabilities seem 
-to have a more negative impact on the maintainability of open-source software 
-than performing regular changes}.
-
 \section{Study Implications}\label{sec:implications}
 
- In other words, it is 
-important to note that the design debt of one guideline can lead
-to severe impacts on the software quality~\cite{10.1145/1985362.1985366}.  
-Thus, tools like Better Code Hub and static analysis tools (e.g., SonarQube, Codacy,
-ESLint, Infer, and more) should be leveraged to 
-assess the risk of patching vulnerabilities - alongside with pull
-requests and code reviews. While static code analysis
-may be daunting due to the number of rules and different effects
-on maintainability, BCH only uses the top-$10$ of the guidelines
-with the highest effect on maintainability. Static analysis 
-can be used to complement the source code analysis, introduce
-the capability of vulnerability detection, and, consequently,
-reduce the risk of hindering maintainability.
-
-This study exhibits evidence that developers may have to
+This section presents the implications that we retrieved from this
+study and think are important to consider in the future by developers.
+Our results exhibit evidence that developers may have to
 reduce maintainability for the sake of security. We understand that developers
@@
 
+	\item\textbf{Tools for Patch Risk Detection:}
+	Design debt of one guideline can lead
+	to severe impacts on the software quality~\cite{10.1145/1985362.1985366}. 
+	Thus, tools like Better Code Hub and static analysis (e.g., SonarQube, Codacy,
+	ESLint, Infer, and more) should be leveraged to 
+	assess the risk of patching vulnerabilities - alongside pull
+	requests/code reviews. While static code analysis
+	may be daunting due to the number of rules and different effects
+	on maintainability, BCH uses the top-$10$ of the guidelines
+	with the highest effect on maintainability. Yet, static analysis 
+	can be used to complement the source code analysis, introduce
+	the capability of vulnerability detection, and, predict and
+	reduce the risk of hindering maintainability.
+
 	\item\textbf{Computer Curricula Needs to be Updated:}
@@
 	to properly educate them. Universities should do their part on making
-	this possible. Online services such as BCH have also a great potential to help
+	this possible. Online services such as BCH have a great potential to help
 	developers to be more aware of the maintainability issues introduced by their
-	changes - alongside with pull requests and code reviews and integrated in 
-	the CI/CD pipeline. Thus, developers can 
+	changes. Thus, developers can 
 	put more work into improving their code maintainability and avoiding the risk
-	of introducing new vulnerabilities after patching older ones.
+	of introducing new vulnerabilities while patching older ones.
 
 	\item\textbf{Secure Programming Languages Wanted:} Programming Languages 
-	should provide new design patterns to easily implement security patterns 
+	should provide new design patterns to easily implement security weaknesses 
 	without endangering maintainability.

### 2020-05-27 paper/icsme20.tex
small fixes

@@
  \begin{figure}[htp]
+ 	\vspace{-0.9cm}
+
      \begin{adjustwidth}{-2.5cm}{-1cm}  
@@
   	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
-  	\caption{Impact of the security patches per guideline.}
+
+    \end{adjustwidth}
+  	\caption{Impact of the security patches per guideline.
+For each guideline, $3$ types of impact in the software maintainability are presented: negative, maintainability decreases (red); none, maintainability remains the same (yellow); and, positive, maintainability increases (green).}
  	\label{fig:guidelines}	
-    \end{adjustwidth}	
 	\vspace{-0.5cm}
@@
 and that this study tries to bring awareness for:
-
-\begin{itemize}
-
-	\item\textbf{Tools for Patch Risk Detection:}
+%
+\\\textit{\textbf{Tools for Patch Risk Detection:}}
 	Design debt of one guideline can lead
@@
 	reduce the risk of hindering maintainability.
-
-	\item\textbf{Computer Curricula Needs to be Updated:}
+%
+\\\textit{\textbf{Computer Curricula Needs to be Updated:}}
 	Software Developers lack the knowledge to create secure and
@@
 	of introducing new vulnerabilities while patching older ones.
-
-	\item\textbf{Secure Programming Languages Wanted:} Programming Languages 
+%	
+\\\textit{\textbf{Secure Programming Languages Wanted:}} Programming Languages 
 	should provide new design patterns to easily implement security weaknesses 
@@
 	features to implement.
-
-	\item\textbf{Poor Coding Practices Documentation:} 
+%
+\\\textit{\textbf{Poor Coding Practices Documentation:}}
 	Software documentation must provided to developers with the
@@
 	to avoid the introduction of new vulnerabilities while patching older ones.
-
-\end{itemize}
 

### 2020-05-27 paper/icsme20.tex
-, -- and ---

@@
 impact on the software maintainability and, possibly, even be the cause of the 
-introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
+introduction of new vulnerabilities---harming software reliability. Therefore, 
 in this paper, we present an empirical study on the impact of patches of 
@@
 \end{itemize}
-%-
+%
 This paper is structured as follows: section~\ref{sec:motivation} introduces an
@@
 However, these patches may have a
-negative impact on software maintenance --- e.g., developers might follow the
-quickest solution to fix the problem and not necessarily the most elegant and
-performant one. Thus, we want to understand how security patches impact
+negative impact on software maintenance (e.g., developers might follow the
+quickest solution to fix the problem but not necessarily the most elegant and
+performant one). Thus, we want to understand how security patches impact
 the maintainability of software.
@@
 renegotiation, and where a malicious client could cause a Denial-of-Service
-(DoS) attack via large OCSP Status Request extensions - if OCSP stapling support
+(DoS) attack via large OCSP Status Request extensions when OCSP stapling support
 was enabled. Online Certificate Status Procotol (OCSP) stapling, formally known as the TLS Certificate 
@@
 
-~\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any
+\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any
 memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded memory growth - which was not being perfomed before.
 After releasing the unbounded memory, the program performs the same check 
-in ~\ref{lst:func2} that used to perform in ~\ref{lst:func3} which is responsible 
+in \ref{lst:func2} that used to perform in \ref{lst:func3} which is responsible 
 for handling the application when no OCSP response IDS were being allocated. Now, instead of performing it after packaging processing, it does it previously.
 
-Although performing this patch, seems an elementary problem to repair,
-the example shows that there was a significant amount of changes performed
-in the code base which performed a negative impact in the software maintanibility.
+Patching this vulnerability seems a rudimentary task. Yet,
+a considerable amount of changes were performed
+in the codebase which yielded to a negative impact on software maintainability.
 While patching, the developer introduced $6$ new lines in a method already with 
 a large number of lines of code and introduced more complexity to the code
-with new branch points which disrupts two of the guidelines proposed by the Software Improvement Group (SIG) for
-building maintainable software~\cite{Visser:2016:OREILLY} - \emph{Write Short Units of Code}
+with new branch points which disrupts two of the guidelines proposed by the Software Improvement Group 
+(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}: \emph{Write Short Units of Code}
 and \emph{Write Simple Units of Code}.
@@
 We also compute the maintainability of random regular commits using the
-\emph{Better Code Hub} tool --- baseline.  We use them to understand \textbf{how
+\emph{Better Code Hub} tool---baseline.  We use them to understand \textbf{how
 maintainability evolves when security patches are performed versus when
@@
 ($2017$) mined open-source
-software aiming at the extraction of real examples --- created by real
-developers --- to test and assess the performance of static analysis tools~\cite{Reis:2017:IJSSE} since
+software aiming at the extraction of real examples---created by real
+developers---to test and assess the performance of static analysis tools~\cite{Reis:2017:IJSSE} since
 using hand-seeded test cases or mutations could lead to misleading assessments
@@
 for the patching, and the snippets of code that differ from one version to
-another (typically, called \textit{diffs}) --- where one can easily review the
+another (typically, called \textit{diffs})---where one can easily review the
 code used to fix the security flaw. 
@@
 Hub} (BCH) is used to collect the maintainability reports of the refactorings of
-each project. Table~\ref{tab:guidelines} presents the $10$ guidelines $-$ proposed
+each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
 by BCH's authors for delivering software that is not difficult to
-maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric 
+maintain~\cite{Visser:2016:OREILLY} and, maps each guideline to the metric 
 calculated by BCH. Theses guidelines are calculated using the metrics 
@@
 with the guideline. These thresholds are calibrated by BCH using their own
-data/experience --- using open-source and closed software systems. If a project is
+data/experience---using open-source and closed software systems. If a project is
 compliant with a guideline, it means that it is at least $65\%$ better than the
@@
 provided by BCH for a project after finishing its evaluation. The example
-refers to the OpenSSL CVE-$2016$-$6304$ vulnerability refactoring -
+refers to the OpenSSL CVE-$2016$-$6304$ vulnerability refactoring---
 as described by Section~\ref{sec:motivation}. This
@@
 commit(s) interleaved with more programming activities, being the later
-commonly called a floss-refactoring. This study considers mainly $1$-commit 
+commonly called a floss-refactoring. This study considers mainly single-commit 
 patches. Yet, a small percentage of data points had more than one commit 
@@
 In \emph{RQ1}, we report and discuss the impact of patches on open-source
-software maintainability under two different views: guidelines/metrics 
-and formula proposed - the mean of all 
-guidelines/metrics - $M(v)$.
+software maintainability under four different views: guidelines/metrics, formula proposed (i.e., the mean of all 
+guidelines/metrics, $M(v)$), severity and programming languages.
 
@@
 on each guideline individually and the average impact on all guidelines 
-together - $M(v)$. Under each guideline, it is stated the 
+together ($M(v)$). Under each guideline, it is stated the 
 metric used to the calculations. Table~\ref{tab:guidelines} describes 
@@
 we also see that patching vulnerabilities also has a very
-significant negative impact on all guidelines - between $10\%$ and $40\%$. 
+significant negative impact on all guidelines---between $10\%$ and $40\%$. 
 \emph{Write Short Units of 
@@
 severity vulnerabilities. Again, patches have a considerable 
-negative impact on software maintainability - between $20\%$ 
+negative impact on software maintainability---between $20\%$ 
 and $50\%$. Statistical significance was retrieved only for \emph{LOW}
@@
 impact on maintainability. But overall patches in all languages have a 
-considerable negative impact on maintainability - between $35\%$ to $50\%$ - which 
+considerable negative impact on maintainability---between $35\%$ to $50\%$---which 
 confirms the need for better and more secure programming languages. 
@@
 Typically, CWE-79 vulnerabilities do not need extra lines 
-to be fixed, as shown in Listing~\ref{lst:fix} - one simple 
+to be fixed, as shown in Listing~\ref{lst:fix}---one simple 
 \texttt{escape} function patches the issue. On the same type of fix,
@@
 Lifetime (CWE-401)}, or also called \emph{Resource Management Errors (CWE-399)}, can also be the root cause of Denial-of-Service attacks and  
-difficult to patch since it usually requires to add more complexity to the program 
-- as seen in the example presented on Section~\ref{sec:motivation}. 
+difficult to patch since it usually requires to add more complexity to the program---as seen in the example presented on Section~\ref{sec:motivation}. 
 Patches of \emph{SQL Injection
@@
 For both types of changes, there is a negative impact
-on software maintainability $-$ especially, for security changes. 
+on software maintainability---especially, for security changes. 
 The resulting p-value of the Paired Wilcoxon signed-rank test is $0.300$, i.e., 
@@
 	ESLint, Infer, and more) should be leveraged to 
-	assess the risk of patching vulnerabilities - alongside pull
+	assess the risk of patching vulnerabilities---alongside pull
 	requests/code reviews. While static code analysis
@@
 have the same impact on the overall maintainability of both software versions 
-- non-vulnerable and vulnerable - but not necessarily on the same 
+(non-vulnerable and vulnerable) but not necessarily on the same 
 guidelines.
@@
 style and software contribution policies which is not evaluated in this study.
-Furthermore, this evaluation considers that $969$ regular commits - any kind
-of commit - are enough to
+Furthermore, this evaluation considers that $969$ regular commits---any kind
+of commit---are enough to
 alleviate random irregularities in the maintainability differences of the

### 2020-05-27 paper/icsme20.tex
one column extra left

@@
 presents the results and discusses their
-implications; section~\ref{sec:threats} enumerates the threats to the validity of
+implications; section~\ref{sec:implications} elaborates on the implications
+developers should consider in the future; section~\ref{sec:threats} enumerates the threats to the validity of
 this study; section~\ref{sec:rw} describes the different work and existing
@@
 \label{tab:guidelines}
+\vspace{-0.5cm}
 \end{table}
@@
 This study evaluates a final total of $969$ security and baseline commits 
-from $260$ distinct open-source projects. The following section presents and 
-discusses the results and answers to each research question.
-
-\textit{\textbf{RQ1: What is the impact of security patches on the
+from $260$ distinct open-source projects. This section reports/discusses 
+the results for each research question.
+%
+\\\textit{\textbf{RQ1: What is the impact of security patches on the
 maintainability of open-source software?}}
 
-In \emph{RQ1}, we report and discuss the impact of patches on open-source
-software maintainability under four different views: guidelines/metrics, formula proposed (i.e., the mean of all 
-guidelines/metrics, $M(v)$), severity and programming languages.
-
-\textbf{Guidelines/Metrics:} Each patch performs a set of changes
+In \emph{RQ1}, we report/discuss the impact of patches on open-source
+software maintainability under four views: guideline, overall score, 
+severity and programming language.
+
+\textbf{Guideline/Metric:} Each patch performs a set of changes
 on the source code of the software. These changes may have a different
@@
 average impact of all guidelines.
+
+ \begin{figure}[htp]
+ 	\vspace{-0.9cm}
+
+     \begin{adjustwidth}{-1cm}{-1cm}  
+  	\centering
+  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
+
+    \end{adjustwidth}
+  	\caption{Impact of the security patches per guideline.
+For each guideline, $3$ types of impact in the software maintainability are presented: negative, maintainability decreases (red); none, maintainability remains the same (yellow); and, positive, maintainability increases (green).}
+ 	\label{fig:guidelines}	
+	\vspace{-0.5cm}
+ \end{figure}
 
@@
 \begin{figure}[htp]
- 	\vspace{-0.9cm}
-
-     \begin{adjustwidth}{-2.5cm}{-1cm}  
  	\centering 	
-  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
-
-    \end{adjustwidth}
-  	\caption{Impact of the security patches per guideline.
-For each guideline, $3$ types of impact in the software maintainability are presented: negative, maintainability decreases (red); none, maintainability remains the same (yellow); and, positive, maintainability increases (green).}
- 	\label{fig:guidelines}	
+	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
+ 	\caption{Maintainability difference by vulnerability severity.}
+	\label{fig:severity}
 	\vspace{-0.5cm}
@@
 
-\begin{figure}[h]
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
- 	\caption{Maintainability difference by vulnerability severity.}
-	\label{fig:severity}
+\begin{figure}[htp]
+\vspace{-0.3cm}
+  \centering
+  \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
+  \caption{Maintainability difference by programming language.}
+  \label{fig:lang_main}  
 \end{figure}
 
-\textbf{Programming Languages:} The impact on software maintainability per programming 
+\begin{figure*}[htp]
+	\vspace{-0.5cm}
+  \centering
+  \subfigure[Maintainability difference by first-level weaknesses from the 
+  \textit{Research Concepts} list on Common Weakness Enumeration (CWE)]{
+  \includegraphics[scale=0.4]{figures/main_per_cwe.pdf}}\quad\quad
+  \subfigure[Maintainability difference by sub-weaknesses of the 
+  \textit{Improper Neutralization} Weakness (CWE-707)]{
+  \includegraphics[scale=0.4]{figures/main_per_cwe_spec_707.pdf}}\quad\quad
+    \subfigure[Maintainability difference by sub-weaknesses of the 
+	\textit{Improper Control of a Resource Through its Lifetime} Weakness (CWE-664)]{
+	\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
+	\label{fig:pat}
+	\vspace{-0.5cm}
+\end{figure*}
+
+\textbf{Programming Language:} The impact on software maintainability per programming 
 language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
@@
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
-  \caption{Maintainability difference by programming language.}
-  \label{fig:lang_main}  
-\end{figure}
-
 \textbf{Conclusion:} Results show that developers fail to follow the coding
@@
 programming languages are needed.
-
-% \Luis{I think this section is missing more discussion.
-What do you think should be changed in the community to cope with our findings?
-How can practitioners fix security vulnerabilities wihthout hindering maintainability?
-do we need more tools, better programming languages, more strict contribution policies?
-is there a trade-off between secure software and technical debt?
-show us how these findings can change SE}
-% Following the \textit{American Statistical Association}
-% recommendations~\cite{doi:10.1080/00031305.2016.1154108}, we consider the results for
-% Ruby as statistically significant because its p-value is on the borderline.
-%
-
-\textit{\textbf{RQ2: Which weaknesses are more likely to
+%
+\\\textit{\textbf{RQ2: Which weaknesses are more likely to
 affect open-source software maintainability?}}
-
-\begin{figure*}[htp]
-  \centering
-  \subfigure[Maintainability difference by first-level weaknesses from the 
-  \textit{Research Concepts} list on Common Weakness Enumeration (CWE)]{
-  \includegraphics[scale=0.4]{figures/main_per_cwe.pdf}}\quad\quad
-  \subfigure[Maintainability difference by sub-weaknesses of the 
-  \textit{Improper Neutralization} Weakness (CWE-707)]{
-  \includegraphics[scale=0.4]{figures/main_per_cwe_spec_707.pdf}}\quad\quad
-    \subfigure[Maintainability difference by sub-weaknesses of the 
-	\textit{Improper Control of a Resource Through its Lifetime} Weakness (CWE-664)]{
-	\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
-	\label{fig:pat}
-\end{figure*}
 
@@
 research should be perfomed to retrieve conclusions.
-
-\textit{\textbf{RQ3: What is the impact of security changes versus regular 
+%
+\\\textit{\textbf{RQ3: What is the impact of security changes versus regular 
 changes on the maintainability of open-source software?}}
@@
 
+\begin{figure}[h]
+	\vspace{-0.3cm}	
+ 	\centering 	
+	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
+ 	\caption{Maintainability difference for security and baseline patches.}
+	\label{fig:secvsreg}
+	\vspace{-0.2cm}
+\end{figure}
+
 For both types of changes, there is a negative impact
@@
 new vulnerabilities.
-
-\begin{figure}[h]	
- 	\centering 	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
-	\label{fig:secvsreg}
-\end{figure}
 
@@
 %
-The following section presents the potential threats to the validity of this
-study.
-%
-\\\textit{\textbf{Construct:}} The maintainability formula was inferred based 
+This section presents the potential threats to the validity.
+%
+\\\textit{\textbf{Construct:}} The formula was inferred based 
 on the BCH's reports. The high amount of different projects and backgrounds 
@@
 the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
-Maintainability is computed as the mean of all guidelines. A patch can
-have the same impact on the overall maintainability of both software versions 
-(non-vulnerable and vulnerable) but not necessarily on the same 
-guidelines.
+Maintainability is computed as the mean of all guidelines. Different software versions (vulnerable/fixed) of one vulnerability may have the same overal score
+and still be affected by different guidelines.
 %

### 2020-05-28 paper/icsme20.tex
minor fixes
pass in implications section

@@
 \section{Study Implications}\label{sec:implications}
-
-This section presents the implications that we retrieved from this
-study and think are important to consider in the future by developers.
-Our results exhibit evidence that developers may have to
-reduce maintainability for the sake of security. We understand that developers
-should be able to patch and produce secure code without hindering the maintainability of
-their projects. But there are still some concerns that need to be addressed
-and that this study tries to bring awareness for:
+%
+Our results show evidence that developers may have to reduce maintainability for the 
+sake of security. We argue that developers should be able to patch and produce secure 
+code without hindering the maintainability of their projects. But there are still  
+concerns that need to be addressed and that this study brings awareness for:
 %
@@
 	assess the risk of patching vulnerabilities---alongside pull
-	requests/code reviews. While static code analysis
+  requests/code reviews. \Rui{what? (till the ned)-->} While static code analysis
 	may be daunting due to the number of rules and different effects
@@
   reduce the risk of hindering maintainability.
+  \Rui{static analysis vs. source coode?!}
 %
 \\\textit{\textbf{Computer Curricula Needs to be Updated:}}
-	Software Developers lack the knowledge to create secure and
+	We argue that software Developers lack the knowledge to create both secure and
 	maintainable code. Mainly because academic curricula is not yet prepared
 	to properly educate them. Universities should do their part on making
-	this possible. Online services such as BCH have a great potential to help
+	this possible. \Rui{connection to curricula is not clear-->} Online services such as BCH have a great potential to help
 	developers to be more aware of the maintainability issues introduced by their
@@
 \\\textit{\textbf{Poor Coding Practices Documentation:}}
-	Software documentation must provided to developers with the
+  \Rui{too generic?}
+	Software documentation must be provided to developers with the
 	best practices to implement secure and maintainable software. Unmaintainable
@@
 baseline. 
+%
 \\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.
@@
 dataset comprises more commits in Java, i.e., the dataset may not be representative 
-of the population regarding programming languages. Our approach does not consider 
-security patches in any other languages but English.
+of the population regarding programming languages. Our approach does not consider patches in any other languages but English.
 

### 2020-05-28 paper/icsme20.tex
new version

@@
 guidelines using a dataset including 1335 security commits.
-Results show that fixing software vulnerabilities may hinder the maintainability 
-of open-source software and that different levels of attention should be given to 
-different severity levels, programming languages and types of weaknesses. We 
-conclude that changes to code bases to patch 
-vulnerabilities need to be performed with extra care and that security-related 
-code evolution should be subject in computer curricula.\Rui{Needs improvement}
+Results show evidence of a new trade-off between security and maintainability, i.e., 
+when developers may be hindering the software maintainability or even introducing 
+new vulnerabilities while patching vulnerabiliteis.
+We conclude that changes to codebases while patching 
+vulnerabilities need to be performed with extra care (e.g., using tools
+for patch risk assessment), computer science
+curricula needs to be updated and more secure programming languages are necessary.
 \end{abstract}
-
 \begin{IEEEkeywords}
@@
 
+For projects with large codebases, the results retrieved for \emph{Keep Your 
+Codebase Small} were not viable because they retrieved values above the limit 
+set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
+not considered since the tool does not contemplate two
+of the most important techniques to security testing: vulnerability scanning
+and penetration testing. Instead, it only contemplates unit 
+testing. Due to BCH limitations, we detected a few data points
+that retrieved incorrect overall calculations. 
+Those data points were disregarded from the study.
+
 The BCH tool does not compute the final score that our study needs to compare
@@
  	\vspace{-0.9cm}
-
      \begin{adjustwidth}{-1cm}{-1cm}  
@@
   	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
-
     \end{adjustwidth}
-  	\caption{Impact of the security patches per guideline.
-For each guideline, $3$ types of impact in the software maintainability are presented: negative, maintainability decreases (red); none, maintainability remains the same (yellow); and, positive, maintainability increases (green).}
+  	\caption{Impact of the security patches per guideline. For each guideline, $3$ types of impact 
+	in the software maintainability are presented: negative, maintainability decreases (red); none, 
+	maintainability remains the same (yellow); and, positive, maintainability increases (green).}
  	\label{fig:guidelines}	
@@
 we also see that patching vulnerabilities also has a very
-significant negative impact on all guidelines---between $10\%$ and $40\%$. 
-\emph{Write Short Units of 
-Code} ($38.3\%$), \emph{Write Simple Units of Code} ($37.9\%$),  and 
-\emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
-most negativelly affected guidelines. 
-This may imply that developers when patching 
-vulnerabilities fail to design/implement solutions that continue to respect 
-the limit bounds of branch points and function/module 
-sizes that are recommended by coding best practices. Still, on 
+significant number of negative cases for all guidelines---between $10\%$ and $40\%$. 
+\emph{Write Short Units of Code} ($38.3\%$), \emph{Write Simple Units of Code} 
+($37.9\%$), and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
+most negativelly affected guidelines. This may imply that developers when patching 
+vulnerabilities have a hard time to design/implement solutions that continue to 
+respect the limit bounds of branch points and function/module sizes that are 
+recommended by coding best practices. Still, on 
 respecting bound limits, developers also seem to not consider the limit 
@@
 violated when the patch requires to input new information to a function/class 
-and developers fail to use the \emph{Introduce Parameter Object} refactoring 
+and developers struggle to use the \emph{Introduce Parameter Object} refactoring 
 pattern. Results do not provide statistical significance to the \emph{Separate 
@@
 interact with others. However, results may imply that developers
-fail to use techniques such as encapsulation to hide implementation
+forget to use techniques such as encapsulation to hide implementation
 details and make the system more modular.
 
-Software systems typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. 
 The \emph{Write Code Once} guideline results show that duplicated code 
-increased in $171$ ($17.7\%$) of the patches. Previous work showed a 
+increased in $171$ ($17.7\%$) of the patches. Software systems typically have 
+$9\%$-$17\%$ of cloned code~\cite{5773403}. Previous work showed a 
 correlation between code smells and code duplication~\cite{7476787} 
@@
 
-For projects with large codebases, the results retrieved for \emph{Keep Your 
-Codebase Small} were not viable because they retrieved values above the limit 
-set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
-not considered since the tool does not contemplate two
-of the most important techniques to security testing: vulnerability scanning
-and penetration testing. Instead, it only contemplates unit 
-testing. Due to BCH limitations, we detected a few data points
-that retrieved incorrect overall calculations. 
-Those data points were disregarded from the study.
-
 \textbf{Overall Score ($M(v)$):} 
@@
 The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
-on software maintainability. While $188$ ($19.4\%$) have no impact and $375$ ($38.7\%$) cases
-have a negative impact. The larger number of negative cases may be 
-explained by guidelines with higher concentration of negative cases with higher amplitudes, 
-such as, \emph{Write Short Units of 
-Code}, \emph{Write Simple Units of 
-Code} and \emph{Separate Concerns in Modules}.
+on software maintainability. While $188$ ($19.4\%$) cases have no impact, $375$ 
+($38.7\%$) have a negative impact on software maintanibility. The larger number of 
+negative cases may be explained by guidelines with higher concentration of negative 
+cases with higher amplitudes, such as, \emph{Write Short Units of Code}, \emph{Write 
+Simple Units of Code} and \emph{Separate Concerns in Modules}.
 
@@
 \textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
-Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National Vulnerability 
-Database} (NVD) website to collect their severity levels. In total, we 
-retrieved severity scores for $536$ vulnerabilities: $112$ (\emph{HIGH}), 
-$395$ (\emph{MEDIUM}) and $29$ (\emph{LOW}). Figure~\ref{fig:severity} 
+Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National 
+Vulnerability Database} (NVD) website to collect their severity levels. In total, 
+we retrieved severity scores for $536$ vulnerabilities: $112$ \emph{High}, 
+$395$ \emph{Medium} and $29$ \emph{Low}. Figure~\ref{fig:severity} 
 presents the impact of security patches per severity level on the 
 maintainability of open-source software. We observe that patches for 
-\emph{HIGH} ($50.0\%$) and \emph{MEDIUM} ($43.8\%$) severity vulnerabilities 
-hinder more the maintainability of software than \emph{LOW} ($27.6\%$) 
+\emph{High} ($50.0\%$) and \emph{Medium} ($43.8\%$) severity vulnerabilities 
+hinder more the maintainability of software than \emph{Low} ($27.6\%$) 
 severity vulnerabilities. Again, patches have a considerable 
 negative impact on software maintainability---between $20\%$ 
-and $50\%$. Statistical significance was retrieved only for \emph{LOW}
-severity vulnerabilitie, i.e., \emph{LOW} severity vulnerabilities
+and $50\%$. Statistical significance was retrieved only for \emph{Low}
+severity vulnerabilitie, i.e., \emph{Low} severity vulnerabilities
 may have less negative impact on software maintainability thant the 
-other severity levels. However, 
-results should not be disregarded because they seem to confirm the 
+other severity levels. However, results should not be disregarded 
+because they seem to confirm the 
 assumption that higher severity vulnerabilities patches may have a 
@@
 \texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
-\texttt{Python} and \texttt{PHP}. \texttt{Groovy} samples are not presented
-because the number of cases did not fulfill the statistical test requirement. 
-\texttt{C/C++} ($46.5\%$), \texttt{Ruby} ($46.5\%$) and 
-\texttt{PHP} ($38.6\%$) are the programming languages with worse 
-impact on software maintainability. \texttt{Java} ($50.9\%$) and \texttt{Python} 
-($46.5\%$) seem to be the 
-languages less affected by patching, i.e., more than $45\%$ of patches had a positive 
-impact on maintainability. But overall patches in all languages have a 
-considerable negative impact on maintainability---between $35\%$ to $50\%$---which 
+\texttt{Python} and \texttt{PHP}, leaving \texttt{Groovy} out of the analysis.
+\texttt{C/C++}, \texttt{Ruby} and 
+\texttt{PHP} are the programming languages with worse 
+impact on software maintainability, i.e., with the highest number of negative cases (respectively, $46.5\%$, $46.5\%$ and $38.6\%$). \texttt{Java} and \texttt{Python} 
+seem to be the languages less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability (respectively, $50.9\%$ and $46.5\%$). But overall 
+languages have a considerable amount of cases that negatively impact software 
+maintainability---between $35\%$ to $50\%$---which 
 confirms the need for better and more secure programming languages. 
@@
 
-We would expect that the negative impact of programming languages on
-maintainability to be more severe, as arguably poor design of programming
-languages and lack of using the best practices lead to more buggy and vulnerable
+We expected the negative impact for programming languages on
+maintainability to be more severe, as arguably, poor design of programming
+languages for security and the lack of best practices application by developers   leads to more buggy and vulnerable
 code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
@@
 
-\textbf{Conclusion:} Results show that developers fail to follow the coding
-practice guidelines and hinder software maintainability after patching; and, 
+\textbf{Conclusion:} Results show that developers have a hard time followings the coding
+practice guidelines and, consequently, hinder software maintainability while patching; and, 
 that different levels of attention should be paid to each guideline. For instance,
@@
 %
-\\\textit{\textbf{Tools for Patch Risk Detection:}}
+\\\textit{\textbf{Tools for Patch Risk Assessment:}}
 	Design debt of one guideline can lead
@@
 	assess the risk of patching vulnerabilities---alongside pull
-  requests/code reviews. \Rui{what? (till the ned)-->} While static code analysis
+  requests/code reviews. Static code analysis
 	may be daunting due to the number of rules and different effects
-	on maintainability, BCH uses the top-$10$ of the guidelines
+	on maintainability. BCH claims to use the top-$10$ of the guidelines
 	with the highest effect on maintainability. Yet, static analysis 
-	can be used to complement the source code analysis, introduce
-	the capability of vulnerability detection, and, predict and
-  reduce the risk of hindering maintainability.
-  \Rui{static analysis vs. source coode?!}
-%
-\\\textit{\textbf{Computer Curricula Needs to be Updated:}}
-	We argue that software Developers lack the knowledge to create both secure and
+	can be used to complement the BCH analysis by introducing
+	the capability of vulnerability detection and support the prediction/prevention of the risk of hindering maintainability while patching.
+%
+\\\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
+	We argue that developers lack the knowledge to create both secure and
 	maintainable code. Mainly because academic curricula is not yet prepared
-	to properly educate them. Universities should do their part on making
-	this possible. \Rui{connection to curricula is not clear-->} Online services such as BCH have a great potential to help
-	developers to be more aware of the maintainability issues introduced by their
-	changes. Thus, developers can 
+	to properly educate them. Universities should encourage students to use online 	services such as BCH or similar ones. These tools have a great potential to make
+	students aware of maintainability issues and beginner mistakes (e.g.,
+	coding practices violation). Thus, students and future developers can 
 	put more work into improving their code maintainability and avoiding the risk
@@
 	features to implement.
-%
-\\\textit{\textbf{Poor Coding Practices Documentation:}}
-  \Rui{too generic?}
-	Software documentation must be provided to developers with the
-	best practices to implement secure and maintainable software. Unmaintainable
-	code is difficult to test, analyze and re-use. Software companies must
-	do an effort to provide these artifacts to their developers, in order,
-	to avoid the introduction of new vulnerabilities while patching older ones.
+% %
+% \\\textit{\textbf{Poor Coding Practices Documentation:}}
+%   \Rui{too generic?}
+% 	Software documentation must be provided to developers with the
+% 	best practices to implement secure and maintainable software. Unmaintainable
+% 	code is difficult to test, analyze and re-use. Software companies must
+% 	do an effort to provide these artifacts to their developers, in order,
+% 	to avoid the introduction of new vulnerabilities while patching older ones.
 

### 2020-05-28 paper/icsme20.tex
new version

@@
 and $50\%$. Statistical significance was retrieved only for \emph{Low}
-severity vulnerabilitie, i.e., \emph{Low} severity vulnerabilities
+severity vulnerabilities, i.e., \emph{Low} severity vulnerabilities
 may have less negative impact on software maintainability thant the 
@@
 Our results show evidence that developers may have to reduce maintainability for the 
-sake of security. We argue that developers should be able to patch and produce secure 
-code without hindering the maintainability of their projects. But there are still  
-concerns that need to be addressed and that this study brings awareness for:
+sake of security. We argue that developers should be able to patch and produce 
+secure code without hindering the maintainability of their projects. But there are 
+still concerns that need to be addressed and that this study brings awareness for:
 %

### 2020-05-28 paper/icsme20.tex
abstract hot fixes

@@
 that when developers address these vulnerabilities on their codebases, 
-they actually affect the maintainability of software, potentially 
-introducing other vulnerabilities. This paper evaluates the impact of 
-patches to improve security on the maintainability of open-source software. 
+they actually affect the maintainability of software. This paper evaluates the 
+impact of patches to improve security on the maintainability of open-source 
+software. 
 Maintainability is measured based on the Better Code Hub’s model of 10 
-guidelines using a dataset including 1335 security commits.
-Results show evidence of a new trade-off between security and maintainability, i.e., 
-when developers may be hindering the software maintainability or even introducing 
+guidelines using a dataset including 1335 security-related commits.
+Results show evidence of a trade-off between security and maintainability, i.e., 
+developers may be hindering the software maintainability or even introducing 
 new vulnerabilities while patching vulnerabiliteis.
-We conclude that changes to codebases while patching 
+The implications of our study are that changes to codebases while patching 
 vulnerabilities need to be performed with extra care (e.g., using tools

### 2020-05-28 scripts/maintainability/eval_maintainability.py
eval main script update

@@ -13,7 +13,7 @@ import maintainability.better_code_hub as bch
 RETRY_ATTEMPTS = 10
 
 @click.command()
-@click.option('--dataset', prompt=True, default="../dataset/reg.csv",
+@click.option('--dataset', prompt=True, default="../dataset/db_release_security_fixes.csv",
               help="Dataset with commits for comparison.")
 @click.option('--commit', prompt=True, default="security",
               help="type of dataset")

### 2020-05-28 scripts/report.py
scripts update

@@ -24,13 +24,13 @@ def main_calculation(df, cache, dataset):
         sha_p_key = 'sha-p'
     
     for i, r in df.iterrows():
-
+                
         if r[sha_key] is np.nan or r[sha_p_key] is np.nan:
             continue
             
         info_f = cache.get_stored_commit_analysis(r['owner'], r['project'], r[sha_key])
         info_p = cache.get_stored_commit_analysis(r['owner'], r['project'], r[sha_p_key])
-            
+                    
         if info_f is None or info_p is None:
             none+=1
             df.at[i, 'diff'] = np.nan
@@ -104,6 +104,11 @@ def comparison(secdb, regdb, reports):
     df_reg = pd.read_csv(regdb)
     chart.main_comparison_chart(reports, df_sec, df_reg)
     
+def guideline_swarm(secdb, reports):
+    
+    df_sec = pd.read_csv(secdb)
+    chart.main_guideline_swarm_plot(reports, df_sec)
+       
     
 if __name__ == "__main__":
     
@@ -128,7 +133,7 @@ if __name__ == "__main__":
             comparison(secdb=args.secdb, regdb=args.regdb, reports=args.reports)
     elif args.goal == 'guideline':
         if args.secdb != None and args.reports != None:
-            guideline(secdb=args.secdb, reports=args.reports)
+            guideline_swarm(secdb=args.secdb, reports=args.reports)
     elif args.goal == 'language':
         if args.secdb != None and args.reports != None:
             language(secdb=args.secdb, reports=args.reports)

### 2020-05-28 scripts/stats/chart.py

@@ -3,6 +3,7 @@ from matplotlib import rc
 from matplotlib import rcParams
 
 import pandas as pd
+import seaborn as sns
 
 
 from scipy.stats import wilcoxon
@@ -40,13 +41,13 @@ def save_report(path, name):
     plt.tight_layout()
     plt.savefig('{}/{}'.format(path, name))
 
-def main_comparison_chart(reports, df_sec, df_reg):    
-
+def main_comparison_chart(reports, df_sec, df_reg):   
+        
         test = pd.concat((tests.hypothesis_test(df_reg['diff'], 'reg'), 
                             tests.hypothesis_test(df_sec['diff'], 'sec')),
                             ignore_index=True)
                         
-        total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)        
+        total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)   
         db_size = len(df_sec)  
 
         rep = {'neg' : pd.Series([total_reg['neg']/db_size, total_sec['neg']/db_size]),
@@ -202,15 +203,12 @@ def main_per_language_chart(reports, df, wilcoxon = True):
         if lang != None:
             df.at[i, 'Language'] = lang
     
-    langs = data.add_others_group(df,
-                                tests.filter_small_sample_groups(df, 'Language'), 
-                                'Language', 
-                                'Other')
+    langs = tests.filter_small_sample_groups(df, 'Language')
         
     results = {l:data.filter_results_per_field(df, l, 'Language') for l in langs}
     test = pd.concat([tests.hypothesis_test(df[df['Language'] == i]['diff'], i) 
                             for i in langs],
-                            ignore_index=True)
+                            ignore_index=True)    
                                 
     rep = {'neg': pd.Series([results[i][0]/sum(results[i]) for i in langs]),
         'neg_abs': pd.Series([results[i][0] for i in langs]),
@@ -228,11 +226,11 @@ def main_per_language_chart(reports, df, wilcoxon = True):
     plt.subplots_adjust(left=0.25, right=0.85, top=0.9, bottom=0.06)
[…]

### 2020-05-28 scripts/stats/enum.py

@@ -1,18 +1,18 @@
 # data.py 
 
-guidelines = {'Write Short Units of Code':'\\textbf{WShortUC}\nUnit Size',
-                'Write Simple Units of Code':'\\textbf{WSimpleUC}\nMcCabe Complexity',
-                'Write Code Once':'\\textbf{WCO}\nDuplication',
-                'Keep Unit Interfaces Small':'\\textbf{KUIS}\nUnit Interfacing',
-                'Separate Concerns in Modules':'\\textbf{SCM}\nModule Coupling',
-                'Couple Architecture Components Loosely':'\\textbf{CACL}\nComponent Independence',
-                'Keep Architecture Components Balanced':'\\textbf{KACB}\nComponent Balance',
-                'Write Clean Code':'\\textbf{WCC}\nCode Smells'}
+guidelines = {'Write Short Units of Code':'\\textbf{Write Short}\n\\textbf{Units of Code}\nUnit Size',
+                'Write Simple Units of Code':'\\textbf{Write Simple}\n\\textbf{Units of Code}\nMcCabe Complexity',
+                'Write Code Once':'\\textbf{Write Code}\n\\textbf{Once}\nDuplication',
+                'Keep Unit Interfaces Small':'\\textbf{Keep Unit}\n\\textbf{Interfaces Small}\nUnit Interfacing',
+                'Separate Concerns in Modules':'\\textbf{Separate Concerns}\n\\textbf{in Modules}\nModule Coupling',
+                'Couple Architecture Components Loosely':'\\textbf{Couple Architecture}\n\\textbf{Components Loosely}\nComponent Independence',
+                'Keep Architecture Components Balanced':'\\textbf{Keep Architecture}\n\\textbf{Components Balanced}\nComponent Balance',
+                'Write Clean Code':'\\textbf{Write Clean}\n\\textbf{Code}\nCode Smells'}
 
-languages = {'Java': ['java', 'scala'], 
+
+languages = {'Java': ['java', 'scala', 'js'], 
                 'Python': ['py'], 
                 'Groovy': ['groovy'], 
-                'JavaScript': ['js'] , 
                 'PHP': ['ctp', 'php', 'inc', 'tpl'], 
                 'Objective-C/C++': ['m', 'mm'], 
                 'Ruby': ['rb'], 

### 2020-05-28 paper/icsme20.tex
paper update
update

@@
 impact of patches to improve security on the maintainability of open-source 
-software. 
-Maintainability is measured based on the Better Code Hub’s model of 10 
+software. Maintainability is measured based on the Better Code Hub’s model of 10 
 guidelines using a dataset including 1335 security-related commits.
@@
 %
-
 As ISO does not provide any specific guidelines/formulas to calculate 
@@
 \begin{itemize}
-  \item Evidence that supports a new trade-off between security and maintainability:
-  when developers need to patch vulnerabilities, they might be hindering the software
-  maintainability or even introducing new vulnerabilities.
+  \item Evidence that supports the trade-off between security and maintainability:
+  developers may be hindering the software maintainability or even introducing new vulnerabilities, while patching older ones.
 	\item An empirical study on the impact of security changes on software
@@
 	\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
+    \caption{Maintainability difference per weakness.}
 	\label{fig:pat}
@@
 
-In \emph{RQ2}, we report and discuss the impact of security patches on open-source
-software maintainability per weakness. In this study, we use weakness definition 
-mentioned in Section~\ref{sec:motivation} and the CWE enumeration
-system to classify the vulnerabilities.
-Figure~\ref{fig:pat} shows three different charts: \emph{(a)} shows
+In \emph{RQ2}, we report/discuss the impact of security patches on
+software maintainability per weakness. We use the weakness definition
+and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section~\ref{sec:motivation}).
+Figure~\ref{fig:pat} shows three different charts: \emph{(\ref{fig:pat}-a)}, shows
 the impact of all patches grouped using the first-level weaknesses from
-the \emph{Research Concepts} list on CWE; and, the remaining charts
-present the impact on maintainability of the sub-weaknesses from two of
-the weaknesses presented in \emph{(a)}.
-
-In Figure~\ref{fig:pat}-\emph{a}, we observe that patching vulnerabilities
-such as Insufficient Control Flow Management 
-(CWE-691), Protection Mechanism Failure (CWE-693), Improper Access Control 
-(CWE-284), Improper Neutralization (CWE-707) and Improper Control of a Resource 
-Through its Lifetime (CWE-664) may have a negative impact on software 
-maintainability. The CWE-707 ($295$) and CWE-664 ($318$) 
-weaknesses integrate a larger amount of cases compared to the remaining 
-weaknesses. Thus, we present an analysis of their sub-weaknesses on 
+the \emph{Research Concepts} list on CWE; and, \emph{(\ref{fig:pat}-b)} and  
+\emph{(\ref{fig:pat}-c)} present the impact on maintainability per sub-weakness of two 
+weaknesses from \emph{(\ref{fig:pat}-a)}: \emph{Improper Neutralization} (CWE-707) and 
+\emph{Improper Control of a Resource 
+Through its Lifetime} (CWE-664), respectively.
+
+In Figure~\ref{fig:pat}-\emph{a}, there is no clear evidence of the impact on 
+maintainability per weakness. Yet, it is important to note that
+overall there is a very considerable number of cases that hinder
+maintainability---between $30\%$ and $60\%$.
+The CWE-707 ($295$) and CWE-664 ($318$) 
+weaknesses integrate the higher number of cases compared to the remaining 
+ones. Thus, we present an analysis of their sub-weaknesses on 
 Figure~\ref{fig:pat}-\emph{b} and Figure~\ref{fig:pat}-\emph{c}, respectively. 
-Results demonstrate evidence that patching vulnerabilities hinders 
-the maintainability of open-source software for $4$ different weaknesses: 
+Results shows that patching vulnerabilities may hinder 
+the maintainability of open-source software in $4$ different sub-weaknesses: 
 \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
@@
 a Memory Buffer (CWE-119)} patches endure more
-cases with no impact on the open-source software maintainability.
-Typically, CWE-79 vulnerabilities do not need extra lines 
+cases with no impact on the open-source software maintainability. \emph{SQL Injection
+(CWE-89)} patches equally hinder and improve the software maintainability.
+These patches usually follow the same complexity as the CWE-79 patches.
+However, the three weaknesses have a considerable amount of cases that hinder
+the software maintainability---$32.4\%$, $40.7\%$ and, $41.1\%$, respectively---which
+should not be happening. Typically, CWE-79 vulnerabilities do not need extra lines 
 to be fixed, as shown in Listing~\ref{lst:fix}---one simple 
@@
 CVE-$2016$-$0799$\footnote{CVE-$2016$-$0799$ patch details available at 
-\url{https://github.com/openssl/openssl/commit/9cb177301fdab492e4cfef376b28339afe3ef663} (Accessed on \today{})}.
-As CWE-199 
-vulnerabilities, \emph{Missing Release of Memory after Effective 
-Lifetime (CWE-401)}, or also called \emph{Resource Management Errors (CWE-399)}, can also be the root cause of Denial-of-Service attacks and  
-difficult to patch since it usually requires to add more complexity to the program---as seen in the example presented on Section~\ref{sec:motivation}. 
-Patches of \emph{SQL Injection
-(CWE-89)} weaknesses equally hinder and improve the software maintainability. 
-But a very considerable percentage of patches hinder the open-source 
-software maintainability ($41.1\%$).
+\url{https://github.com/openssl/openssl/commit/9cb177301fdab492e4cfef376b28339afe3ef663}
+(Accessed on \today{})}. 
+As CWE-199 vulnerabilities, \emph{Missing Release of Memory after Effective 
+Lifetime (CWE-401)} can also be the root cause of Denial-of-Service attacks and  
+difficult to patch since it usually requires to add more complexity to the program (cf. Section~\ref{sec:motivation}).
 
@@
 Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the most likely 
-weaknesses to affect open-source software maintainability.
-No statistical evidence was retrieved mostly due to the difficult
-of finding a higher numbers of cases of open-source patches per weakness. However, 
-results should not be disregarded because they offer an interesting view on how 
-the maintainability is impacted per weakness which can provide insights to 
-developers/researchers on the weaknesses that need more attention. Further
-research should be perfomed to retrieve conclusions.
+weaknesses to affect software maintainability. No statistical evidence was retrieveds. However, results should not be disregarded because they hint on how 
+the maintainability is impacted per weakness which highlights the weaknesses
+that may need more attention from developers/researchers. 
 %
@@
 
-For both types of changes, there is a negative impact
-on software maintainability---especially, for security changes. 
-The resulting p-value of the Paired Wilcoxon signed-rank test is $0.300$, i.e., 
-conclusions regarding the regular changes are not statistically significant. 
-Yet, it is important to note that \emph{patching security vulnerabilities seem 
-to have a more negative impact on the maintainability of open-source software 
-than performing regular changes}.
-This raises 
-a new trade-off when developers need to patch vulnerabilities on their projects
-because they might be hindering the software maintainability or even introducing
-new vulnerabilities.
+Results show that security-related commits are expected to hinder the
+maintainability of software. In contrast, one cannot say the same about
+regular changes, in which we do not observe significant differences. Thus, we
+urge the importance of adopting maintainability practices while applying
+security patches.
+%
+% For both types of changes, there is a negative impact
+% on software maintainability---especially, for security changes.
+% The resulting p-value of the Paired Wilcoxon signed-rank test is $0.300$, i.e.,
+% conclusions regarding the regular changes are not statistically significant.
+% Yet, it is important to note that \emph{patching security vulnerabilities seem
+% to have a more negative impact on the maintainability of open-source software
+% than performing regular changes}.
+% This raises
+% a new trade-off when developers need to patch vulnerabilities on their projects
+% because they might be hindering the software maintainability or even introducing
+% new vulnerabilities.
 

### 2020-05-28 scripts/data/get_secbench_patterns.py

@@ -1,11 +1,0 @@
-import sys
-import pandas as pd
-
-_SECBENCH = sys.argv[1]
-_ICPC20 = sys.argv[2]
-
-df_sec = pd.read_csv(_SECBENCH)
-df = pf.read_CSV(_ICPC20)
-
-
-

### 2020-05-28 paper/icsme20.tex
t2v

@@
 to fix security flaws. This approach discards security patches that were
-not explicit in commits messages. Pontas et. al\cite{10.1109/MSR.2019.00064} dataset
-has patches that were manually mined in software repositories and, consequently,
-wrong associations may have been performed between vulnerabilities and code snippets. 
-The dataset also included floss-refactorings. Both cases were manually inspected
-and deleted from the dataset. Thus, this study does not consider that a
-developer may perform a patch quickly and later proceed to the refactor.
-Baseline commits are retrieved randomly by selecting one commit with an 
-approximate size and from the same project of a given security patch. 
+not explicit in commits messages. This study assumes that patches were performed 
+using a single-commit or several sequentially. The view that a 
+developer may perform a patch quickly and later proceed to the refactor is not considered.s
+Baseline commits are retrieved randomly from the same project of the security patch.
 This approach softens the differences

### 2020-05-28 paper/icsme20.tex
rw

@@
 of refactorings. Instead, statistical tests were used to find the metrics that
-have the potential to change significantly after refactorings. The differences
-of maintainability between groups of refactored elements and non-refactored were
-analyzed. In addition, the authors concluded that the source code elements
-subjected to refactorings had significantly lower maintainability than the
-elements not affected. Studying the evolution of maintainability issues during
-the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
+have the potential to change significantly after refactorings. Studying the evolution 
+of maintainability issues during the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
 discovered that maintainability decreases over time. Palomba et al.
@@
 should be carefully monitored by programmers since there is a high correlation
-between maintainability aspects and proneness to changes and faults. In 2019, Cruz et 
+between maintainability aspects and proneness to changes/faults. In 2019, Cruz et 
 al.~\cite{cruz2019energyoriented} proposed a formula to calculate maintainability 
 based on the BCH's guidelines and measured the impact of energy-oriented fixes 
-on the maintainability of software.
-The present work uses the same model (BCH) and focuses solely on evaluating 
-the impact of patching vulnerabilities on software maintainability.
+on software maintainability. The present work uses the same model (BCH) and focuses 
+solely on evaluating the impact of security patches on software maintainability.
 
 Researchers investigated the relationship between design patterns and software
-maintainability~\cite{10.1007/978-3-642-35267-6-18}. $300$ revisions of
-\texttt{JHotDraw} were analyzed to conclude that every introduced pattern
-instance improved software maintainability. However, other studies show that the
-use of design patterns may introduce maintainability issues into
+maintainability~\cite{10.1007/978-3-642-35267-6-18}. However, other studies show that 
+the use of design patterns may introduce maintainability issues into
 software~\cite{4493325}. Yskout et. al were not able to detect if the usage of 
@@
 the authors of the second study were not able to reproduce it, not obtaining any
-evidence about the impact of language design. Both have used the number of
-defects as an indicator of software quality. In addition,
+evidence about the impact of language design. In addition,
 Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922,
@@
 previously demonstrated correlation between programming language and software 
-defects. Our work studies how security refactorings affect software quality based 
-on the code maintainability analysis and provides evidence that programming languages 
-have an impact on maintainability.
+defects. Our work studies how security patches affect software quality based 
+on the code maintainability analysis and provides shows that programming languages 
+may have an impact on maintainability.
 

### 2020-05-28 paper/icsme20.tex
10 pages

@@
 	\item An empirical study on the impact of security changes on software
-  maintainability and what weaknesses, severity levels and programming languages need more attention.
+  maintainability (per guideline, severity, weakness and programming language).
 	\item A replication package with all scripts and data created to perform the
@@
 	 	\centering 	
-	\includegraphics[width=0.5\textwidth]{figures/methodology.pdf}
+    \begin{adjustwidth}{0.5cm}{-1cm}  
+	\includegraphics[width=0.49\textwidth]{figures/methodology.pdf}
+ 	\vspace{-0.1cm}
+
  	\caption{Study Methodology}
 	\label{fig:met}
+	 \end{adjustwidth}
+ 	\vspace{-0.1cm}
 \end{figure}
@@
 	\label{fig:bchrep}
+	\vspace{-0.2cm}
 \end{figure}
@@
 
+Aiming to analyze the impact of security patches, we use BCH to compute the
+maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
+
+\begin{itemize}
+	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
+	patch (\emph{sha-p});
+	\item $v_{s}$, the version free of the security flaw, i.e., after the
+	patch (\emph{sha});
+\end{itemize}
+
+Patches can be performed through one commit, several consecutive commits, or 
+commit(s) interleaved with more programming activities (floss-refactorings). This study considers mainly single-commit 
+patches. Yet, a small percentage of data points had more than one commit 
+involved in the patch. We manually inspected these cases and $23$ 
+floss-refactorings were identified and disregarded since it would not
+be fair to measure the maintainability of these cases where other
+programming activities are involved.
+
 \begin{figure}[h]
@@
 \end{figure}
-
-Aiming to analyze the impact of security patches, we use BCH to compute the
-maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
-
-\begin{itemize}
-	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
-	patch (\emph{sha-p});
-	\item $v_{s}$, the version free of the security flaw, i.e., after the
-	patch (\emph{sha});
-\end{itemize}
-
-Patches can be performed through one commit, several consecutive commits, or 
-commit(s) interleaved with more programming activities, being the later
-commonly called a floss-refactoring. This study considers mainly single-commit 
-patches. Yet, a small percentage of data points had more than one commit 
-involved in the patch. We manually inspected these cases and $23$ 
-floss-refactorings were identified and disregarded since it would not
-be fair to measure the maintainability of these cases where other
-programming activities are involved.
 
@@
   	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
+	\vspace{-0.4cm}
+
     \end{adjustwidth}
+
   	\caption{Impact of the security patches per guideline. For each guideline, $3$ types of impact 
@@
 
+ \begin{figure}[htp]
+ 	\vspace{-0.6cm}
+  	\centering 	
+ 	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
+ 	\vspace{-0.3cm}
+  	\caption{Maintainability difference by vulnerability severity.}
+ 	\label{fig:severity}
+ 	\vspace{-0.5cm}
+ \end{figure}
+
 Regarding the impact of security patches for each guideline, we observe
 that overall patches have more cases where the impact is positive. However, 
-we also see that patching vulnerabilities also has a very
+we also see that patching vulnerabilities has a very
 significant number of negative cases for all guidelines---between $10\%$ and $40\%$. 
@@
 respect the limit bounds of branch points and function/module sizes that are 
-recommended by coding best practices. Still, on 
+recommended by coding practices. Still, on 
 respecting bound limits, developers also seem to not consider the limit 
@@
 Simple Units of Code} and \emph{Separate Concerns in Modules}.
-
-\begin{figure}[htp]
- 	\centering 	
-	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
- 	\caption{Maintainability difference by vulnerability severity.}
-	\label{fig:severity}
-	\vspace{-0.5cm}
-\end{figure}
 
@@
 \begin{figure}[htp]
-\vspace{-0.3cm}
+\vspace{-0.4cm}
   \centering
   \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
+  \vspace{-0.3cm}
   \caption{Maintainability difference by programming language.}
   \label{fig:lang_main}  
+  \vspace{-0.2cm}
+
 \end{figure}
@@
 	\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
+	\vspace{-0.3cm}
     \caption{Maintainability difference per weakness.}
 	\label{fig:pat}
-	\vspace{-0.5cm}
+	\vspace{-0.4cm}
 \end{figure*}
@@
 this analysis to programming languages with at least $20$ data points, as this 
-is a requirement for our hypothesis tests. Thus, we compare the results for 
+is a requirement for the hypothesis tests. Thus, we compare the results for 
 \texttt{C/C++}, \texttt{Ruby}, \texttt{Java}, \texttt{Objective C/C++}, 
@@
 \texttt{PHP} are the programming languages with worse 
-impact on software maintainability, i.e., with the highest number of negative cases (respectively, $46.5\%$, $46.5\%$ and $38.6\%$). \texttt{Java} and \texttt{Python} 
-seem to be the languages less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability (respectively, $50.9\%$ and $46.5\%$). But overall 
-languages have a considerable amount of cases that negatively impact software 
+impact on maintainability, i.e., with the highest number of negative cases ($46.5\%$, $46.5\%$ and $38.6\%$, respectively). \texttt{Java} and \texttt{Python} 
+seem to be the less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability ($50.9\%$ and $46.5\%$, respectively). But overall 
+languages have a considerable amount of cases that negatively impact  
 maintainability---between $35\%$ to $50\%$---which 
-confirms the need for better and more secure programming languages. 
+confirms the need for better/more secure programming languages. 
 Statistical significance was only retrieved for the \texttt{C/C++} 
-(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} (p-value = $0.048$) languages.
-The remaining programming languages did not yield statistical significance, 
-thus results should be read with care.
-Yet, data reports very interesting hints on the impact of programming languages 
-on security patches.
+(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} 
+(p-value = $0.048$) languages. Yet, data reports very interesting hints on the impact 
+of programming languages on security patches.
 
@@
 maintainability to be more severe, as arguably, poor design of programming
-languages for security and the lack of best practices application by developers   leads to more buggy and vulnerable
+languages for security and the lack of best practices application by developers lead to more buggy/vulnerable
 code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
 Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
-a significant negative impact approximate to $50\%$ on software 
+a significant negative impact approximate to $50\%$ on 
 maintainability. We suspect that these values are
 the result of projects contributions policies (e.g., coding standards). In our dataset, 
-$9$/$10$ projects with more contributors have strict contribution 
-policies regarding code standards. For
-instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-Standard}\footnote{PSR-2 Standard available at
-\url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
-
-\textbf{Conclusion:} Results show that developers have a hard time followings the coding
-practice guidelines and, consequently, hinder software maintainability while patching; and, 
+$9$/$10$ projects with more contributors follow strict contribution 
+policies for code standards. % For
+% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
+% Standard}\footnote{PSR-2 Standard available at
+% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
+
+\textbf{Conclusion:} Results show that developers may have a hard time following the guidelines and, consequently, hinder software maintainability while patching; and, 
 that different levels of attention should be paid to each guideline. For instance,
@@
 may have a negative impact on the maintainability of open-source software}; and, 
-that tools such as Better Code Hub should be integrated in the CI/CD pipeline
-to help developers evaluate the risk of patches hindering software maintainability
-alongside with Pull Requests/Code Reviews. We also conclude that different severity 
+that tools such as BCH should be integrated in the CI/CD pipelines
+to help developers evaluate the risk of patches hinder software maintainability---alongside Pull Requests/Code Reviews. We also conclude that different severity 
 vulnerabilities may need different levels of attention and better and more secure 
@@
 some buffer overflows may be harder to fix and lead to more complex 
-solutions such as the one performed for 
+solutions (e.g., 
 CVE-$2016$-$0799$\footnote{CVE-$2016$-$0799$ patch details available at 
 \url{https://github.com/openssl/openssl/commit/9cb177301fdab492e4cfef376b28339afe3ef663}
-(Accessed on \today{})}. 
-As CWE-199 vulnerabilities, \emph{Missing Release of Memory after Effective 
-Lifetime (CWE-401)} can also be the root cause of Denial-of-Service attacks and  
-difficult to patch since it usually requires to add more complexity to the program (cf. Section~\ref{sec:motivation}).
+(Accessed on \today{})}). 
+As CWE-199 weaknesses, \emph{Missing Release of Memory after Effective 
+Lifetime (CWE-401)} can also be the cause of Denial-of-Service attacks and  
+difficult to patch since it usually requires to add complexity to the program (cf. Section~\ref{sec:motivation}).
 
@@
 
-The impact of security and baseline changes in the maintainability of
-open-source software is presented in Figure~\ref{fig:secvsreg}.  
+The impact of security and baseline changes on software maintainability is presented in Figure~\ref{fig:secvsreg}.  
 We have seen, previously, a deterioration on software maintainability 
@@
 security patches, the maintainability of regular changes remains the 
-same in $42.5\%$ ($412$) of the cases and performing regular
-changes has a more positive impact than negative on the software 
+same in $42.5\%$ ($412$) of the cases and perform regular
+changes has a more positive impact than negative on  
 maintainability.
+
+\begin{figure}[htp]
+	\vspace{-0.5cm}	
+ 	\centering 	
+	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
+	\vspace{-0.3cm}
+ 	\caption{Maintainability difference for security and baseline patches.}
+	\label{fig:secvsreg}
+	\vspace{-0.4cm}
+\end{figure}
 
@@
 
-\begin{figure}[h]
-	\vspace{-0.3cm}	
- 	\centering 	
-	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
-	\label{fig:secvsreg}
-	\vspace{-0.2cm}
-\end{figure}
-
-Results show that security-related commits are expected to hinder the
-maintainability of software. In contrast, one cannot say the same about
-regular changes, in which we do not observe significant differences. Thus, we
+Results show that security-related commits are expected to harm software maintainability. In contrast, one cannot say the same about
+regular changes, in which no significant differences are observed. Thus, we
 urge the importance of adopting maintainability practices while applying
@@
 	features to implement.
-% %
+% 
 % \\\textit{\textbf{Poor Coding Practices Documentation:}}
@@
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
-commits produced by project developers to classify the changes performed by them
-to fix security flaws. This approach discards security patches that were
-not explicit in commits messages. This study assumes that patches were performed 
+commits produced by project developers to classify the changes performed while patching vulnerabilities. This approach discards patches that were
+not explicit in commits messages. We assume that patches were performed 
 using a single-commit or several sequentially. The view that a 
-developer may perform a patch quickly and later proceed to the refactor is not considered.s
+developer may perform a patch quickly and later proceed to the refactor is not considered.
 Baseline commits are retrieved randomly from the same project of the security patch.
@@
 
-Researchers investigated the relationship between design patterns and software
+Researchers investigated the relationship between design patterns and
 maintainability~\cite{10.1007/978-3-642-35267-6-18}. However, other studies show that 
 the use of design patterns may introduce maintainability issues into
-software~\cite{4493325}. Yskout et. al were not able to detect if the usage of 
-design patterns has a positive impact, but concludes that developers prefer to 
+software~\cite{4493325}. Yskout et. al did not detect if the usage of 
+design patterns has a positive impact, but concluded that developers prefer to 
 work with the support of security patterns~\cite{8077802}. The present work 
-studies how security patterns influence maintainability for open-source software.
+studies how security weaknesses influence maintainability for open-source software.
 
@@
 quality~\cite{Ray:2014:LSS:2635868.2635922,Ray:2017:LSP:3144574.3126905}. The first
-study shows that some programming languages are more buggy-prone than others. However,
-the authors of the second study were not able to reproduce it, not obtaining any
-evidence about the impact of language design. In addition,
+one shows that some programming languages are more buggy-prone than others. However,
+the authors of the second one were not able to reproduce it and did not obtain any
+evidence about the language design impact. 
 Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922,
@@
 patches on the maintainability of $260$ open-source projects. We leveraged
-Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018, cruz2019energyoriented}. Security patches 
-significantly affect the software maintainability of open-source software, i.e., 
-developers hinder maintainability while patching vulnerabilities. We found this evidence in
-$41.9\%$ of the studied cases. When comparing with the impact of regular changes, security patches seem to have a higher negative impact than regular commits.
-Regarding security patches, we also observe that some guidelines are more
-affected than others: \emph{Write Simple Units of Code} ($37.9\%$) and
-\emph{Write Short Units of Code} ($38.3\%$).
-
-The current study also hints that more severe vulnerabilities may have
-a more negative impact on the software maintainability and that patches
-is some programming languages may needed more attention than others (e.g., \texttt{C/C++}, \texttt{Ruby} and \texttt{PHP}. Therefore, 
-we draw attention for the need of better documentation explaining 
-how to use best practices to produce secure and maintainable code; 
-new and better programming languages; new
-academic curricula in the fields of security and software quality; and, 
-tools such as Better Code Hub and efficient static analysis to integrate
-in the CI/CD pipelines to help 
-developers predict the effect of their patches on software maintainability.
+Better Code Hub reports to calculate maintainability based on a model proposed in 
+previous work~\cite{Olivari:2018, cruz2019energyoriented}. Results show evidence of a 
+trade-off between security and maintainability ($41.9\%$ of the cases), i.e., 
+developers may be hindering the software maintainability or even introduces 
+new vulnerabilities while patching older ones. We also observe that some guidelines 
+and programming languages are more likely to be affected than others. This study
+aims to stress that changes to codebases while patching 
+vulnerabilities need to be performed with extra care (e.g., integrating tools
+for patch risk assessment in the CI/CD pipelines), computer science
+curricula needs to be updated and more secure programming languages are necessary.
 

### 2020-05-28 paper/icsme20.tex
adjusting
polishing

@@
 \begin{abstract}
-Security is a requirement of utmost importance to produce software with 
-high quality. However, there is still a considerable amount of 
+Security is a requirement of utmost importance to produce software quality. However, there is still a considerable amount of 
 vulnerabilities being discovered and fixed, almost weekly. We suspect 
@@
 Results show evidence of a trade-off between security and maintainability, i.e., 
-developers may be hindering the software maintainability or even introducing 
-new vulnerabilities while patching vulnerabiliteis.
+developers may be hindering the software maintainability and introducing 
+new vulnerabilities while patching vulnerabilities.
 The implications of our study are that changes to codebases while patching 
@@
 first steps taken by SIG to introduce security on their maintainability model). 
-Nevertheless, our study uses BCH's toolset to provide an assessment of maintainability in 
-software for the following reasons: 1) BCH considers a broad number of code metrics; and, 2)
+Nevertheless, our study uses BCH to provide an assessment of maintainability in 
+software for the following reasons: BCH integrates a broad number of code metrics; and, 
 code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}.
@@
 we intend to highlight the need for tools to predict the impact of patches on 
-maintainability~\cite{4724577}, documentation to help developers adopt security 
-best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
-security-related code evolution in computer curricula.
+maintainability~\cite{4724577}, the importance of integrating
+maintainable security in computer science curricula and better programming languages.
 %
@@
  \begin{figure}[htp]
- 	\vspace{-0.9cm}
+ 	\vspace{-0.6cm}
      \begin{adjustwidth}{-1cm}{-1cm}  
@@
  	\label{fig:guidelines}	
-	\vspace{-0.5cm}
+	\vspace{-0.7cm}
  \end{figure}

### 2020-05-28 paper/icsme20.tex
update

@@
 \ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any
-memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded memory growth - which was not being perfomed before.
+memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded memory growth - which was not being performed before.
 After releasing the unbounded memory, the program performs the same check 
@@
 a large number of lines of code and introduced more complexity to the code
-with new branch points which disrupts two of the guidelines proposed by the Software Improvement Group 
+with new branch points which disrupt two of the guidelines proposed by the Software Improvement Group 
 (SIG) for building maintainable software~\cite{Visser:2016:OREILLY}: \emph{Write Short Units of Code}
@@
 ($37.9\%$), and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
-most negativelly affected guidelines. This may imply that developers when patching 
+most negatively affected guidelines. This may imply that developers when patching 
 vulnerabilities have a hard time to design/implement solutions that continue to 

### 2020-05-29 paper/icsme20.tex
spellcheck conc+rw

@@
 software~\cite{4493325}. Yskout et. al did not detect if the usage of 
-design patterns has a positive impact, but concluded that developers prefer to 
+design patterns has a positive impact but concluded that developers prefer to 
 work with the support of security patterns~\cite{8077802}. The present work 
@@
 Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the 
-previously demonstrated correlation between programming language and software 
+previously demonstrated a correlation between programming language and software 
 defects. Our work studies how security patches affect software quality based 

### 2020-05-29 paper/icsme20.tex
spellcheck t2v

@@
 %
-This section presents the potential threats to the validity.
+This section presents the potential threats to validity.
 %
@@
 the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
-Maintainability is computed as the mean of all guidelines. Different software versions (vulnerable/fixed) of one vulnerability may have the same overal score
+Maintainability is computed as the mean of all guidelines. Different software versions 
+(vulnerable/fixed) of one vulnerability may have the same overall score
 and still be affected by different guidelines.
@@
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
-commits produced by project developers to classify the changes performed while patching vulnerabilities. This approach discards patches that were
+commits produced by project developers to classify the changes performed while 
+patching vulnerabilities. This approach discards patches that were
 not explicit in commits messages. We assume that patches were performed 
@@
 maintainability may still be affected by the developers' experience, coding
-style and software contribution policies which is not evaluated in this study.
+style, and software contribution policies which are not evaluated in this study.
 Furthermore, this evaluation considers that $969$ regular commits---any kind
@@
 dataset comprises more commits in Java, i.e., the dataset may not be representative 
-of the population regarding programming languages. Our approach does not consider patches in any other languages but English.
+of the population regarding programming languages. Our approach does not consider 
+patches in any other language but English.
 

### 2020-05-29 paper/icsme20.tex
small fixes

@@
 \begin{abstract}
-Security is a requirement of utmost importance to produce software quality. However, there is still a considerable amount of 
-vulnerabilities being discovered and fixed, almost weekly. We suspect 
-that when developers address these vulnerabilities on their codebases, 
-they actually affect the maintainability of software. This paper evaluates the 
+Security is a requirement of utmost importance to produce high-quality software. 
+However, there is still a considerable amount of 
+vulnerabilities being discovered and fixed, almost weekly. We suspect that developers 
+affect the maintainability of their codebases when patching vulnerabilities. This paper evaluates the 
 impact of patches to improve security on the maintainability of open-source 
@@
 
-Results show that regular changes typically have less negative impact
+Results show that regular changes typically have a less negative impact
 on the software maintainability of open-source software.
 However, we manually inspected a considerable amount of cases from
-the regular changes distribution where this happens and found that 
+the distribution of the regular changes where this happens and found that 
 identifying regular commits as similar in size as the security-related 
@@
 
-Results show that security-related commits are expected to harm software maintainability. In contrast, one cannot say the same about
+Security-related commits are expected to harm software maintainability. In contrast, one cannot say the same about
 regular changes, in which no significant differences are observed. Thus, we

### 2020-05-29 paper/icsme20.tex
spellcheck

@@
 cases with no impact on the open-source software maintainability. \emph{SQL Injection
-(CWE-89)} patches equally hinder and improve the software maintainability.
+(CWE-89)} patches equally hinder and improve software maintainability.
 These patches usually follow the same complexity as the CWE-79 patches.
@@
 (e.g., replacing the \texttt{strncpy} function by a more secure one 
-\texttt{strlcpy} that checks if the buffer is null terminated). However,
+\texttt{strlcpy} that checks if the buffer is null-terminated). However,
 some buffer overflows may be harder to fix and lead to more complex 
@@
 Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the most likely 
-weaknesses to affect software maintainability. No statistical evidence was retrieveds. However, results should not be disregarded because they hint on how 
+weaknesses to affect software maintainability. No statistical evidence was retrieved. However, results should not be disregarded because they hint on how 
 the maintainability is impacted per weakness which highlights the weaknesses
@@
 negative impact, $38.7\%$ ($375$) remained the same and $19.4\%$ 
-($188$) increased the software maintainability. For regular changes, 
+($188$) increased software maintainability. For regular changes, 
 we observe that the maintainability decreases in $27.0\%$ ($262$) 

### 2020-05-29 paper/icsme20.tex
spellcheck

@@
 impact on maintainability, i.e., with the highest number of negative cases ($46.5\%$, $46.5\%$ and $38.6\%$, respectively). \texttt{Java} and \texttt{Python} 
-seem to be the less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability ($50.9\%$ and $46.5\%$, respectively). But overall 
+seem to be less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability ($50.9\%$ and $46.5\%$, respectively). But overall 
 languages have a considerable amount of cases that negatively impact  
@@
 maintainability. We suspect that these values are
-the result of projects contributions policies (e.g., coding standards). In our dataset, 
+the result of project contributions policies (e.g., coding standards). In our dataset, 
 $9$/$10$ projects with more contributors follow strict contribution 
@@
 may have a negative impact on the maintainability of open-source software}; and, 
-that tools such as BCH should be integrated in the CI/CD pipelines
-to help developers evaluate the risk of patches hinder software maintainability---alongside Pull Requests/Code Reviews. We also conclude that different severity 
+that tools such as BCH should be integrated into the CI/CD pipelines
+to help developers evaluate the risk of patches of hindering software maintainability---alongside Pull Requests/Code Reviews. We also conclude that different severity 
 vulnerabilities may need different levels of attention and better and more secure 

### 2020-05-29 paper/icsme20.tex
spellcheck

@@
 Results show evidence of a trade-off between security and maintainability, i.e., 
-developers may be hindering the software maintainability and introducing 
+developers may be hindering software maintainability and introducing 
 new vulnerabilities while patching vulnerabilities.
@@
 Software quality is important because it is ultimately related to the overall
-cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
+cost of developing and maintaining software applications~\cite{slaughter1998evaluating}. 
 Software quality characteristics include, but are not limited to functional correctness,
-reliability, usability, maintainability, and security. Security is an 
+reliability, usability, maintainability, evolvavility and security. Security is an 
 important non-functional requirement during the development of software systems. 
@@
 Novel tools have been built to 
-detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
+utomatically detect software vulnerabilities (e.g., FindBugs, Infer and more). 
 Despite these efforts, improving software security is not a trivial task and requires 
@@
   \item Evidence that supports the trade-off between security and maintainability:
-  developers may be hindering the software maintainability or even introducing new vulnerabilities, while patching older ones.
+  developers may be hindering software maintainability while patching vulnerabilities.
 	\item An empirical study on the impact of security changes on software
@@
 Thus, while developers 
-tend to ignore refactoring techinques such as the \emph{Replace Method 
+tend to ignore refactoring techniques such as the \emph{Replace Method 
 with Method Object} technique, they may also be introducing new software 
@@
 trade-off between security and maintainability ($41.9\%$ of the cases), i.e., 
-developers may be hindering the software maintainability or even introduces 
+developers may be hindering software maintainability or even introduces 
 new vulnerabilities while patching older ones. We also observe that some guidelines 

### 2020-05-29 paper/icsme20.tex
spellcheck

@@
 maintainability is different from zero~\cite{10.2307/2282543}.
-The Wilcoxon test requires a distribution size of at leat $20$ instances.
+The Wilcoxon test requires a distribution size of at least $20$ instances.
 To understand the effect-size, as advocated by the Common-language effect 
@@
 For each type of guideline, a boxplot 
-and a swarmplot are presented to show the variability and dipersion of
+and a swarmplot are presented to show the variability and dispersion of
 the results alongside the number of absolute
@@
 $26.9\%$ and $11.0\%$, respectively. Component independence and balance
-are important to make easier to find the source code that developers
+are important to make it easier to find the source code that developers
 want to patch/improve and to understand how the high-level components
@@
 on software maintainability. While $188$ ($19.4\%$) cases have no impact, $375$ 
-($38.7\%$) have a negative impact on software maintanibility. The larger number of 
-negative cases may be explained by guidelines with higher concentration of negative 
-cases with higher amplitudes, such as, \emph{Write Short Units of Code}, \emph{Write 
+($38.7\%$) have a negative impact on software maintainability. The larger number of 
+negative cases may be explained by guidelines with a higher concentration of negative 
+cases with higher amplitudes, such as \emph{Write Short Units of Code}, \emph{Write 
 Simple Units of Code} and \emph{Separate Concerns in Modules}.
@@
 severity vulnerabilities, i.e., \emph{Low} severity vulnerabilities
-may have less negative impact on software maintainability thant the 
+may have a less negative impact on software maintainability than the 
 other severity levels. However, results should not be disregarded 

### 2020-05-29 paper/icsme20.tex
spellcheck
fixes

@@
 
-Novel tools have been built to 
-utomatically detect software vulnerabilities (e.g., FindBugs, Infer and more). 
+Novel tools have been built to automatically detect software vulnerabilities (e.g., FindBugs, Infer and more). 
 Despite these efforts, improving software security is not a trivial task and requires 
@@
 impact on the software maintainability and, possibly, even be the cause of the 
-introduction of new vulnerabilities---harming software reliability. Therefore, 
+introduction of new vulnerabilities---harming software reliability and introducing technical debt. Therefore, 
 in this paper, we present an empirical study on the impact of patches of 
@@
 commit ($v_{s-1}$), as illustrated in Figure~\ref{fig:commit}. Thus, we can determine 
-which patches had a positive, negative or null impact on the project maintainability.
+which patches had a positive, negative, or null impact on the project maintainability.
 
@@
 previous work~\cite{Olivari:2018, cruz2019energyoriented}. Results show evidence of a 
-trade-off between security and maintainability ($41.9\%$ of the cases), i.e., 
+trade-off between security and maintainability ($41.9\%$ of security patches yielded a negative impact), i.e., 
 developers may be hindering software maintainability or even introduces 
 new vulnerabilities while patching older ones. We also observe that some guidelines 
-and programming languages are more likely to be affected than others. This study
-aims to stress that changes to codebases while patching 
+and programming languages are more likely to be affected than others. The implications 
+of our study are that changes to codebases while patching 
 vulnerabilities need to be performed with extra care (e.g., integrating tools
-for patch risk assessment in the CI/CD pipelines), computer science
-curricula needs to be updated and more secure programming languages are necessary.
-
-As future work, this study may be extended in several directions: 
+for patch risk assessment in the CI/CD pipelines); computer science
+curricula needs to be updated; and, more secure programming languages are necessary.
+
+As future work, the study extended in several directions: 
 investigate which guidelines affect most the maintainability per

### 2020-05-29 paper/icsme20.tex
fixes

@@
 software for the following reasons: BCH integrates a broad number of code metrics; and, 
-code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}.
+code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented, 8785997}.
 
@@
 based on the BCH's guidelines and measured the impact of energy-oriented fixes 
-on software maintainability. The present work uses the same model (BCH) and focuses 
+on software maintainability. Recent work,
+proposed a new maintainability model to 
+measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
+The present work uses the same model (BCH) and focuses 
 solely on evaluating the impact of security patches on software maintainability.

### 2020-05-29 paper/icsme20.tex
fixes

@@
 previous work~\cite{Olivari:2018, cruz2019energyoriented}. Results show evidence of a 
-trade-off between security and maintainability ($41.9\%$ of security patches yielded a negative impact), i.e., 
-developers may be hindering software maintainability or even introduces 
-new vulnerabilities while patching older ones. We also observe that some guidelines 
+trade-off between security and maintainability, as $41.9\%$ of security patches yielded a negative impact. Hence,
+developers may be hindering software maintainability while patching vulnerabilities. We also observe that some guidelines 
 and programming languages are more likely to be affected than others. The implications 

### 2020-05-29 paper/icsme20.tex
fixes

@@
 software for the following reasons: BCH integrates a broad number of code metrics; and, 
-code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented, 8785997}.
+code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, 8919169, 8785997}.
 
@@
 In this research, we follow a very similar methodology to the one presented 
-in previous work on the maintainability of energy-oriented fixes~\cite{cruz2019energyoriented}. 
+in previous work on the maintainability of energy-oriented fixes~\cite{8919169}. 
 As said before, the web-based source code analysis service \emph{Better Code
@@
 work on measuring the impact of energy-oriented 
-refactorings~\cite{cruz2019energyoriented}. Cruz et al. ($2019$) proposed an 
+refactorings~\cite{8919169}. Cruz et al. ($2019$) proposed an 
 equation to capture the distance between the current state of the project and
 the standard thresholds calculated by the BCH based on the insights provided 
-in~\cite{Olivari:2018}. The equation provided in~\cite{cruz2019energyoriented} 
+in~\cite{Olivari:2018}. The equation provided in~\cite{8919169} 
 considers that the size of project changes does not affect the maintainability 
@@
 is calculated based on previous work, which calculates the number of lines of code that comply and not comply
-with the guideline at a given severity level~\cite{cruz2019energyoriented}. In our analysis, we compute 
+with the guideline at a given severity level~\cite{8919169}. In our analysis, we compute 
 the difference of maintainability between the security commit ($v_{s}$) and its parent 
@@
 between maintainability aspects and proneness to changes/faults. In 2019, Cruz et 
-al.~\cite{cruz2019energyoriented} proposed a formula to calculate maintainability 
+al.~\cite{8919169} proposed a formula to calculate maintainability 
 based on the BCH's guidelines and measured the impact of energy-oriented fixes 
@@
 Better Code Hub reports to calculate maintainability based on a model proposed in 
-previous work~\cite{Olivari:2018, cruz2019energyoriented}. Results show evidence of a 
+previous work~\cite{Olivari:2018, 8919169}. Results show evidence of a 
 trade-off between security and maintainability, as $41.9\%$ of security patches yielded a negative impact. Hence,

### 2020-05-29 paper/icsme20.tex
fixes

@@
 reliability, usability, maintainability, evolvavility and security. Security is an 
-important non-functional requirement during the development of software systems. 
+essential non-functional requirement during the development of software systems. 
 However, there is still a considerable amount of vulnerabilities being
@@
 %
+
 As ISO does not provide any specific guidelines/formulas to calculate 
@@
 %
-This paper is structured as follows: section~\ref{sec:motivation} introduces an
+This paper is structured as follows: Section~\ref{sec:motivation} introduces an
 example of a security refactoring of a known vulnerability found in the
@@
 protocols. Repository available at \url{https://github.com/openssl/openssl}
-(Accessed on \today{})}; section~\ref{sec:methodology} describes the
-methodology used to answer the research questions; section~\ref{sec:results}
+(Accessed on \today{})}; Section~\ref{sec:methodology} describes the
+methodology used to answer the research questions; Section~\ref{sec:results}
 presents the results and discusses their
-implications; section~\ref{sec:implications} elaborates on the implications
-developers should consider in the future; section~\ref{sec:threats} enumerates the threats to the validity of
-this study; section~\ref{sec:rw} describes the different work and existing
-literature in the field of study; and, finally, section~\ref{sec:conclusions}
+implications; Section~\ref{sec:implications} elaborates on the implications
+developers should consider in the future; Section~\ref{sec:threats} enumerates the threats to the validity of
+this study; Section~\ref{sec:rw} describes the different work and existing
+literature in the field of study; and, finally, Section~\ref{sec:conclusions}
 concludes the main findings and elaborates on future work.
@@
 software. The methodology comprises the following steps, as illustrated in
-Figure~\ref{fig:met}.
+Fig.~\ref{fig:met}.
 %
@@
 
-Figure~\ref{fig:bchrep} shows an example of the report
+Fig.~\ref{fig:bchrep} shows an example of the report
 provided by BCH for a project after finishing its evaluation. The example
@@
 constructors in Java). One of the guidelines with which the project does not
-comply is the one presented in the report (cf. Figure~\ref{fig:bchrep}):
+comply is the one presented in the report (cf. Fig.~\ref{fig:bchrep}):
 \emph{Write Simple Units of Code}. BCH analyzes this guideline based on the
@@
 Aiming to analyze the impact of security patches, we use BCH to compute the
-maintainability of two different versions of the project (cf. Figure~\ref{fig:commit}):
+maintainability of two different versions of the project (cf. Fig.~\ref{fig:commit}):
 
@@
 the difference of maintainability between the security commit ($v_{s}$) and its parent 
-commit ($v_{s-1}$), as illustrated in Figure~\ref{fig:commit}. Thus, we can determine 
+commit ($v_{s-1}$), as illustrated in Fig.~\ref{fig:commit}. Thus, we can determine 
 which patches had a positive, negative, or null impact on the project maintainability.
@@
 % security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
-% Figure~\ref{fig:commit}.
+% Fig.~\ref{fig:commit}.
 
@@
 impact on the guidelines/metrics used to measure software maintainability. 
-Figure~\ref{fig:guidelines} shows the impact of security patches 
+Fig.~\ref{fig:guidelines} shows the impact of security patches 
 on each guideline individually and the average impact on all guidelines 
@@
 on software maintainability guidelines, this is not reflected in the average 
-impact of all guidelines ($M(v)$) as we can see in Figure~\ref{fig:guidelines}.
+impact of all guidelines ($M(v)$) as we can see in Fig.~\ref{fig:guidelines}.
 The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
@@
 we retrieved severity scores for $536$ vulnerabilities: $112$ \emph{High}, 
-$395$ \emph{Medium} and $29$ \emph{Low}. Figure~\ref{fig:severity} 
+$395$ \emph{Medium} and $29$ \emph{Low}. Fig.~\ref{fig:severity} 
 presents the impact of security patches per severity level on the 
@@
 \textbf{Programming Language:} The impact on software maintainability per programming 
-language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
+language was also analyzed (Fig.~\ref{fig:lang_main}). We restrict 
 this analysis to programming languages with at least $20$ data points, as this 
@@
 code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
-Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
+Fig.~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
 a significant negative impact approximate to $50\%$ on 
@@
 for \emph{Separate Concerns in Modules}. 
-The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
+The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Fig.~\ref{fig:guidelines}
 is $0.044$. Since the p-value is 
@@
 and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section~\ref{sec:motivation}).
-Figure~\ref{fig:pat} shows three different charts: \emph{(\ref{fig:pat}-a)}, shows
+Fig.~\ref{fig:pat} shows three different charts: \emph{(\ref{fig:pat}-a)}, shows
 the impact of all patches grouped using the first-level weaknesses from
@@
 
-In Figure~\ref{fig:pat}-\emph{a}, there is no clear evidence of the impact on 
+In Fig.~\ref{fig:pat}-\emph{a}, there is no clear evidence of the impact on 
 maintainability per weakness. Yet, it is important to note that
@@
 ones. Thus, we present an analysis of their sub-weaknesses on 
-Figure~\ref{fig:pat}-\emph{b} and Figure~\ref{fig:pat}-\emph{c}, respectively. 
+Fig.~\ref{fig:pat}-\emph{b} and Fig.~\ref{fig:pat}-\emph{c}, respectively. 
 Results shows that patching vulnerabilities may hinder 
@@
 
-The impact of security and baseline changes on software maintainability is presented in Figure~\ref{fig:secvsreg}.  
+The impact of security and baseline changes on software maintainability is presented in Fig.~\ref{fig:secvsreg}.  
 We have seen, previously, a deterioration on software maintainability 
@@
 \\\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
-	We argue that developers lack the knowledge to create both secure and
-	maintainable code. Mainly because academic curricula is not yet prepared
-	to properly educate them. Universities should encourage students to use online 	services such as BCH or similar ones. These tools have a great potential to make
+	Computer Science curricula is not yet prepared to properly educate students for maintainable security. 
+	    Students should be exposed to this problem to gain experience with it. Curricula should focus on producing 
+	    secure and maintainable code and alert to the trade-off between both.
+	Universities should also encourage students to use code quality analysis tools such as BCH or similar ones. These tools have a great potential to make
 	students aware of maintainability issues and beginner mistakes (e.g.,
@@
 previous work~\cite{Olivari:2018, 8919169}. Results show evidence of a 
-trade-off between security and maintainability, as $41.9\%$ of security patches yielded a negative impact. Hence,
-developers may be hindering software maintainability while patching vulnerabilities. We also observe that some guidelines 
+trade-off between security and maintainability, as $41.9\%$ of security patches 
+yielded a negative impact. Hence,
+developers may be hindering software maintainability while patching 
+vulnerabilities. We also observe that some guidelines 
 and programming languages are more likely to be affected than others. The implications 
 of our study are that changes to codebases while patching 
-vulnerabilities need to be performed with extra care (e.g., integrating tools
-for patch risk assessment in the CI/CD pipelines); computer science
+vulnerabilities need to be performed with extra care; tools
+for patch risk assessment should be integrate into the CI/CD pipeline; computer science
 curricula needs to be updated; and, more secure programming languages are necessary.

### 2020-05-29 paper/icsme20.tex
fixes

@@
 Results show evidence of a trade-off between security and maintainability, i.e., 
-developers may be hindering software maintainability and introducing 
-new vulnerabilities while patching vulnerabilities.
-The implications of our study are that changes to codebases while patching 
-vulnerabilities need to be performed with extra care (e.g., using tools
-for patch risk assessment), computer science
-curricula needs to be updated and more secure programming languages are necessary.
+developers may be hindering software maintainability.
+The implications 
+of our study are that changes to codebases while patching 
+vulnerabilities need to be performed with extra care; tools
+for patch risk assessment should be integrate into the CI/CD pipeline; computer science
+curricula needs to be updated; and, more secure programming languages are necessary.
 \end{abstract}

### 2020-05-29 paper/icsme20.tex

@@
 \url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
+
 %These vulnerabilities are gateways for attackers to penetrate systems and 
@@
 %
+
 In this section, we discuss the methodology used to measure
@@
 %
+
 We use a combined dataset of security vulnerabilities which is 
@@
 security patches.
+
 %

### 2020-08-10 paper/emse20.tex
emse template addeed

--- 2020-05-29 paper/icsme20.tex
@@
-\documentclass[10pt,conference]{IEEEtran}
-\IEEEoverridecommandlockouts
-
+%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This is a general template file for the LaTeX package SVJour3
+% for Springer journals.          Springer Heidelberg 2010/09/16
+%
+% Copy it to a new file with a new name and use it as the basis
+% for your article. Delete % signs as needed.
+%
+% This template includes a few options for different layouts and
+% content for various journals. Please consult a previous issue of
+% your journal as needed.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% First comes an example EPS file -- just ignore it and
+% proceed on the \documentclass line
+% your LaTeX will extract the file if required
+%\documentclass{svjour3}                     % onecolumn (standard format)
+%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
+\documentclass[smallextended]{svjour3}       % onecolumn (second format)
+%\documentclass[twocolumn]{svjour3}          % twocolumn
+%
+\smartqed  % flush right qed marks, e.g. at end of proof
+%
 \usepackage[utf8]{inputenc}
@@
 \usepackage{changepage}
+\usepackage{listings}
 
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\usepackage{listings}
 
@@
 
-%
-% Add comments in the text
-%
 \newboolean{showcomments}
@@
 
+%
+% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
+%
+% insert here the call for the packages your document requires
+%\usepackage{latexsym}
+% etc.
+%
+% please place your own definitions here and don't use \def but
+% \newcommand{}{}
+%
+% Insert the name of "your journal" with
+% \journalname{myjournal}
+%
 \begin{document}
 
-\title{Fixing Vulnerabilities Potentially Hinders Maintainability}
-%\title{Coding practices are important to produce secure software: Teach them to your students and developers!}
-%\title{The reason why you should be applying coding practices: safer software.}
-%\title{What is the impact of  corollarysecurity patches in software evolution?}
-%\title{Hey, developers! You should pay more attention to your patches.}
-%\title{Are good coding pratices used when patching software vulnerabilities?}
-
-\author{
-    Anonymou(s) Author(s)
-
-%     \IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-% \textit{name of organization (of Aff.)}\\
-% City, Country \\
-% email address}
-% \and
-% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-% \textit{name of organization (of Aff.)}\\
-% City, Country \\
-% email address}
-% \and
-% \IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-% \textit{name of organization (of Aff.)}\\
-% City, Country \\
-% email address}
+\title{Fixing Vulnerabilities Potentially Hinders Maintainability%\thanks{Grants or other notes
+%about the article that should go on the front page should be
+%placed here. General acknowledgments should be placed at the end of the article.}
 }
+% \subtitle{Do you have a subtitle?\\ If so, write it here}
+
+%\titlerunning{Short form of title}        % if too long for running head
+
+\author{Sofia Reis         \and
+        Rui Abreu           \and
+        Luis Cruz
+}
+
+%\authorrunning{Short form of author list} % if too long for running head
+
+\institute{F. Author \at
+              first address \\
+              Tel.: +123-45-678910\\
+              Fax: +123-45-678910\\
+              \email{fauthor@example.com}           %  \\
+%             \emph{Present address:} of F. Author  %  if needed
+           \and
+           S. Author \at
+              second address
+}
+
+\date{Received: date / Accepted: date}
+% The correct dates will be entered by the editor
 
@@
 developers may be hindering software maintainability.
-The implications 
-of our study are that changes to codebases while patching 
+The implications of our study are that changes to codebases while patching 
 vulnerabilities need to be performed with extra care; tools
@@
 curricula needs to be updated; and, more secure programming languages are necessary.
+\keywords{Security \and Software Maintenance \and Open-Source Software}
+% \PACS{PACS code1 \and PACS code2 \and more}
+% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
-\begin{IEEEkeywords}
-Security, Software Maintenance, Open-Source Software
-\end{IEEEkeywords}
 
@@
 
-%These vulnerabilities are gateways for attackers to penetrate systems and 
-%exploit their resources. As an example, at the beginning of $2020$, academics published a 
-%new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
-%security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
-%address in previous generations and that makes every user using a CPU released before Q$4$ 
-%$2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
-%at the cost of features and/or performance.
-%
-
 In $2011$, the International Organization for Standardization (ISO) issued an
@@
 However, many developers still lack the knowledge about best practices to deliver 
-secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
+secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}.
 ISO describes software maintainability as ``the degree of effectiveness and
@@
 vulnerabilities on software maintenance across open-source software.
-%
-% Software maintainability can also be defined as 
-% the property of software that reveals how easily the software system can be 
-% maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure. 
-% Hedegus et al.
-% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
-% groups of refactored elements and non-refactored and conclude that the source
-% code elements subjected to refactorings had significantly lower maintainability
-% than the elements not modified.
-% There is still little information about the impact of patching vulnerabilities
-% in open-source software maintainability. \textit{Are developers affecting software
-% maintainability when trying to improve software security?}, it is what we want to
-% investigate.
-%
 
@@
 software for the following reasons: BCH integrates a broad number of code metrics; and, 
-code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, 8919169, 8785997}.
+code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
 
@@
 maintainable security in computer science curricula and better programming languages.
-%
-%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
-%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
-%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
-%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
-%
 
@@
 implications; Section~\ref{sec:implications} elaborates on the implications
-developers should consider in the future; Section~\ref{sec:threats} enumerates the threats to the validity of
+developers should consider in the future; Section~\ref{sec:threats} enumerates 
+the threats to the validity of
 this study; Section~\ref{sec:rw} describes the different work and existing
@@
 concludes the main findings and elaborates on future work.
-%
 
@@
 
-%We remove any OCSP_RESPIDs from a previous handshake
-%to prevent unbounded memory growth - CVE-2016-6304
-
 \medskip
@@
 
-% \medskip
-% \setcounter{lstannotation}{0}
-% \begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-% \\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
-%  static SUB_STATE_RETURN read_state_machine(SSL *s){
-%  // [snip]
-% +   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
-% +        && !BUF_MEM_grow_clean(s->init_buf,
-% +                                (int)s->s3->tmp.message_size
-% +                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
-% +         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
-% +         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-% +          return SUB_STATE_ERROR;
-%  // [snip]
-%  }
-% \end{lstlisting}
-%
-% \medskip
-% \setcounter{lstannotation}{0}
-% \begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-% \\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
-%  int tls_get_message_header(SSL *s, int *mt){
-%  // [snip]
-%   l = RECORD_LAYER_get_rrec_length(&s->rlayer) +
-%         SSL3_HM_HEADER_LENGTH;
-% - if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func3}!*/
-% -      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-% -      goto err;
-%    }
-%  // [snip]
-%
-% - if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l +SSL3_HM_HEADER_LENGTH)) {  /*!\annotation{lst:func4}!*/
-% -     SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-% -     goto err;
-% - }
-%
-% \end{lstlisting}
-%
-%\medskip
-%\setcounter{lstannotation}{0}
-%\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
-%\\CVE-2016-6307 vulnerability for file ssl/statem/statem_lib.c },label={lst:vuln}]
-% int tls_get_message_header(SSL *s, int *mt){
-% // [snip]
-%+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
-%+        && !BUF_MEM_grow_clean(s->init_buf, 
-%+                                (int)s->s3->tmp.message_size
-%+                                + SSL3_HM_HEADER_LENGTH)) {
-%+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
-%+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-%+          return SUB_STATE_ERROR;
-% // [snip]
-% }
-%\end{lstlisting}
-
 The code changes performed to patch the vulnerability were:
@@
 \ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any
-memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded memory growth - which was not being performed before.
+memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous 
+handshake to prevent unbounded memory growth - which was not being performed before.
 After releasing the unbounded memory, the program performs the same check 
 in \ref{lst:func2} that used to perform in \ref{lst:func3} which is responsible 
-for handling the application when no OCSP response IDS were being allocated. Now, instead of performing it after packaging processing, it does it previously.
+for handling the application when no OCSP response IDS were being allocated. Now, 
+instead of performing it after packaging processing, it does it previously.
 
@@
 
-% \Sof{The \texttt{read\_state\_machine} function, presented in Listing \ref{lst:vuln}, abstracts a sub-state-machine
-% to handle the messages exchanged between peers. When messages are larger than
-% what is defined by default for TLS, it may trigger a memory corruption issue since the allocation
-% of this resource was being verified very early in the application's flow, using the \texttt{tls\_get\_message\_header} function presented in Listing ~\ref{lst:vuln2} (\ref{lst:func3} and \ref{lst:func4}).
-% These verifications of size allocation were substituted by the one presented in Listing~\ref{lst:vuln}(\ref{lst:func2}).
-% %
-% Although checking the maximum size of a message seems an elementary
-% problem to repair, the example shows that there was a significant amount of
-% changes performed in the code base which produced a negative impact in the
-% maintainability. The developer tries to patch the
-% code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
-% and disrupts
-% one of the guidelines proposed by the Software Improvement Group (SIG) for
-% building maintainable software~\cite{Visser:2016:OREILLY} $-$
-% the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
-% OpenSSL maintainability.}
-
-%%%%%%% 
 Software maintainability is designated as the degree to which an application is understood, 
@@
 approximately $75\%$ of the cost related to a project. To answer the following three 
-research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} 
+research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064} 
 to measure the impact of security patches on the maintainability of open-source software. 
@@
 \emph{Better Code Hub} tool.
-%
 
@@
 	\item Combine the datasets from related work that classifies
-the activities of developers addressing security-oriented patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064}.
+the activities of developers addressing security-oriented patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064}.
 %
@@
 	\includegraphics[width=0.49\textwidth]{figures/methodology.pdf}
- 	\vspace{-0.1cm}
 
@@
 	 \end{adjustwidth}
- 	\vspace{-0.1cm}
 \end{figure}
@@
 	\label{fig:bchrep}
-	\vspace{-0.2cm}
 \end{figure}
@@
 \label{tab:guidelines}
-\vspace{-0.5cm}
 \end{table}
@@
  \begin{figure}[htp]
- 	\vspace{-0.6cm}
      \begin{adjustwidth}{-1cm}{-1cm}  
@@
   	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
-	\vspace{-0.4cm}
-
     \end{adjustwidth}
@@
  	\label{fig:guidelines}	
-	\vspace{-0.7cm}
  \end{figure}
@@
  \begin{figure}[htp]
- 	\vspace{-0.6cm}
   	\centering 	
  	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
- 	\vspace{-0.3cm}
   	\caption{Maintainability difference by vulnerability severity.}
  	\label{fig:severity}
- 	\vspace{-0.5cm}
  \end{figure}
@@
 \begin{figure}[htp]
-\vspace{-0.4cm}
   \centering
   \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
-  \vspace{-0.3cm}
   \caption{Maintainability difference by programming language.}
   \label{fig:lang_main}    
-  \vspace{-0.2cm}
-
 \end{figure}
@@
 \begin{figure*}[htp]
-	\vspace{-0.5cm}
   \centering
@@
 	\includegraphics[scale=0.4]{figures/main_per_cwe_spec_664.pdf}}
-	\vspace{-0.3cm}
     \caption{Maintainability difference per weakness.}
 	\label{fig:pat}
-	\vspace{-0.4cm}
 \end{figure*}
@@
 languages for security and the lack of best practices application by developers lead to more buggy/vulnerable
-code~\cite{Ray:2017:LSP:3144574.3126905, 2019arXiv190110220B}. However,
+code~\cite{Ray:2017:LSP:3144574.3126905,2019arXiv190110220B}. However,
 Fig.~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
@@
 \begin{figure}[htp]
-	\vspace{-0.5cm}	
  	\centering 	
-	\includegraphics[width=0.4\textwidth]{figures/main_comparison.pdf}
-	\vspace{-0.3cm}
+	\includegraphics[width=0.5\textwidth]{figures/main_comparison.pdf}
  	\caption{Maintainability difference for security and baseline patches.}
 	\label{fig:secvsreg}
-	\vspace{-0.4cm}
 \end{figure}
@@
 representative benchmark of closed and open-source software projects to compute 
-the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
+the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,Baggen2012}.
 Maintainability is computed as the mean of all guidelines. Different software versions 
@@
 evidence about the language design impact. 
-Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922,
-Ray:2017:LSP:3144574.3126905} and identified flaws that throw into distrust the 
+Berger et al. ($2019$)~\cite{2019arXiv190110220B} tried to reproduce~\cite{Ray:2014:LSS:2635868.2635922,Ray:2017:LSP:3144574.3126905} 
+and identified flaws that throw into distrust the 
 previously demonstrated a correlation between programming language and software 
@@
 Better Code Hub reports to calculate maintainability based on a model proposed in 
-previous work~\cite{Olivari:2018, 8919169}. Results show evidence of a 
+previous work~\cite{Olivari:2018,8919169}. Results show evidence of a 
 trade-off between security and maintainability, as $41.9\%$ of security patches 
@@
 
-%
-% \section*{Acknowledgment}\label{sec:ack}
-%
-% \noindent
-% We thank the Software Improvement Group (SIG) team for all the support and
-% help in validating the methodology.
-% %ack pedro adao when accepted.
-% \pagebreak
-\balance
-
-{
-  \bibliographystyle{unsrt}
-  \bibliography{icsme20}
-}
+\bibliographystyle{spmpsci}
+\bibliography{emse20}   % name your BibTeX data base
 
 \end{document}
+% end of file template.tex

### 2020-08-13 paper/emse20.tex
esme update

@@
 
-\textbf{Conclusion:} Results show that developers may have a hard time following the guidelines and, consequently, hinder software maintainability while patching; and, 
+\textbf{Conclusion:} Results show that developers may have a hard time following the 
+guidelines and, consequently, hinder software maintainability while patching; and, 
 that different levels of attention should be paid to each guideline. For instance,
@@
 
+\section{Now}
+
+\begin{itemize}
+	\item How this is a problem or has an impact to the software project? What is the message to developers? 
+	\item Implications: Discuss that depending on the project, maybe security is a first class citizen and should be fixed above all. Don't 
+  just state that patch risk assessment should take maintainability into account. Saying that security should be a priority requires maybe a user study
+  as evidence (check if the literature available is not enough).
+  \item Review the secure programming languages implication. Check if the idea of fully secure exists. The idea to 
+\end{itemize}
+
+\section{Later}
+
+\begin{itemize}
+	\item 
+\end{itemize}
+
+\section{Not Sure}
+\begin{itemize}
+	\item I am not sure how the computer science curricula got mentioned from a motivation. I don't think vulnerabilities are been introduced, its 
+ more that they get exposed. In this case, the implication does not make much sense from a practical point of view.
+\end{itemize}
+
 \bibliographystyle{spmpsci}

### 2020-08-21 paper/emse20.tex
update

--- 2020-08-10 paper/emse20.tex
@@
 
+\section{Plan A}
+
+Please, pay more attention to the cases with !.
+
+\begin{itemize}
+	\item How this is a problem or has an impact to the software project? 
+  What is the message to developers? (clarify)
+	\item Implications: Discuss that depending on the project, maybe security 
+  is a first class citizen and should be fixed above all. Don't 
+  just state that patch risk assessment should take maintainability into account. 
+  Saying that security should be a priority requires maybe a user study
+  as evidence (check if the literature available is not enough).
+  \item Review the secure programming languages implication. Check 
+  if the idea of fully secure exists. The idea should be more secure, not
+  fully secure.
+  \item [!] \textit{I am not sure how the computer science curricula got mentioned 
+    from a motivation. I don't think vulnerabilities are been introduced, its 
+    more that they get exposed. In this case, the implication does not make 
+    much sense from a practical point of view.} -> Why not? Shouldn't 
+    maintainable security be introduced in the computer science curricula since
+    we have evidence that developers might not following good practices while
+    patching? If they learn it in college and are aware of it in their jobs
+    won't they be better prepared?
+  \item [!] \textit{It did take me several readings to understand the paper, especially the approach} - 
+  what to do with this @rui?
+  \item [!] \textit{Fig 4, although it looks impressive, I am not sure how to read the figure} - what 
+  should we do @rui?
+  \item [!] \textit{Fig 5, 6 and 7 are not the best ways to portray your findings. For instance, I am 
+  still not sure how the positive, none and negative metrics are calculated and normalized} - what
+  is the best way then? 
+  \item \textit{Authors carefully highlight the key messages of each research question and answer
+  all accordingly} - Improve this.
+  \item [!] \textit{wrong assumption: patching commit should also be responsible to preserve the 
+  maintainability.... The patching commit is to fix the vulnerability and to put the system is 
+  an error-free, executable state so that the fix can be tested} -> I agree with this for vulnerabilities
+  that are the result of errors. For example, a bad expection handling, a memory leak that led to 
+  a denial-of-service attack, NPE, etc. However, there are many other vulnerabilities whose
+  exploits do not have interference on systems execution. So, what to do here? Should we divide
+  the study by vulnerabilities and their implications on systems? Should we divide it by severity levels?
+  \item \textit{authors write "...important to write maintainable code to support the production
+  of more secure software...". Relationship is not clear nor is it backed by some evidence} -> 
+  Improve this.
+  \item Drop the word refactoring. Security patches are not refactorings. Refactorings is perfective
+  or preventive maintenance, whereas patching a vulnerability is corrective maintenance.
+  \item Discuss better the fact of not using other important aspects for the baseline such as the
+  size of the changed method and complexity by metrics. Soften the part where it says we should
+  have gone beyond size.
+  \item \textit{ Figure 4 does not indicate that or support the statement that “patching 
+  vulnerabilities has a very significant number of negative cases for all guidelines”. Besides a 
+  corresponding statistical test that supports this statement, even the swarm and box plotz itself 
+  clearly tell a different story. So, authors may rethink this statement or better explain how they 
+  come to this conclusion.} -> We are overselling here. Improve it.
+  \item \textit{Authors state that they assume for the Pontas dataset that the last commit related 
+  to patching being the actual patch while the parent of the first patch-related commit is assumed 
+  to be the vulnerable version. However, nothing is said whether this assumption holds all the time. 
+  So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.
+  \item \textit{In Section 3-C, for computing and comparing the maintainability, only security commits  
+  (and parent commit) are mentioned. Did you apply the same method also the the regular changes that 
+  you use to compare with security commits? Please make this clear, currently this is ignored in the 
+  particular  (sub) section.} - Improve this.
+  \item \textit{For clarification, the authors explained that they performed a manual investigation 
+  for RQ3. It would be beneficial for more details.} -> Clarify this.
+  \item \textit{In terms of presentation, the results were a bit difficult to follow at times as they 
+  did not contain contextualization. For example, “negative cases” can seem a bit ambiguous when 
+  referencing the relationship between the impact with guidelines.} -> Improve this.
+
+\end{itemize}
+
+\section{Plan B}
+
+Baseado nos reviewers, eu acho que para seguir as orientaçoes deles de refletir e fortalecer 
+a nova assumption e as implicaçoes, teriamos de acrescentar ao Plano A, os proximos pontos
+antes de submeter à ESME.
+
+\begin{itemize}
+\item \textbf{Step 1:} Pegar nos casos em que o impacto na maintainability foi negativa e ver se foram modificados 
+depois e se sim, se houve melhoria da maintainability. -> isto para então resolver a questao da 
+assumption errada.
+\item \textbf{Step 2:} Se não existir nenhuma alteração, entao talvez entrar em contacto com os developers
+e perguntar se quando fizeram o patch tiveram em consideração maintainability do projecto. E se não porquê.
+\item Problemas de baselines e testes acho que nao vale a pena pegar porque vai depender
+do reviewer (eu acho).
+\end{itemize}
+
+Qual é o plano que acha melhor? Como prosseguir?
+
+\section{Later}
+
+\begin{itemize}
+	\item \textbf{Assumptions:}\textit{authors measure the maintainability directly after the patch. inestad it should look 
+  for multiple follow up commits and how maintainability evolves there, as it is likely that certain
+  issues are addresses AFTER the patching commit} -> I think we should see if later commits touch on
+  the patch and measure their maintainability again. Was a less maintainable fix pushed first and then
+  refactored, or does the code stay less maintainable?
+  \item \textbf{To back up implications:} \textit{In terms of the push for changes to curricula, I would agree that it is important 
+  to always stress writing high quality code. However, the support by the study is not strong. Given a 
+  majority of patches did not decrease maintainability, it shows that there is a reasonable focus on maintainability. 
+  Similarly, it is important to understand the circumstances when it decreased. As previously mentioned, organizations 
+  like would prioritize getting the system patched faster (especially in the case of a severe vulnerability) over 
+  its maintainability.} -> Work with practioners to understand if they were taking into consideration maintainability?
+  \item \textbf{Baseline:}\textit{Authors just search for commits of a similar size. There are other important aspects
+  that are entirely neglected such as size of the changed method or how complex it it (by means of 
+  metrics)} -> Improve baseline.
+  \item \textbf{Baseline:}\textit{Similarly, the comparison to “regular” commits really compares security patches to 
+  any other type of change. It would have been more meaningful to compared them to specific types of 
+  commits, like bug fixes (e.g., are security fixes less maintainable than bug fixes?).}
+  \item \textbf{Tests:}\textit{or the chosen statistical test, the hypothesis may be not well chosen as it just checks for 
+  the same distribution of pairs of versions. However, this means if a similar number of pairs have 
+  a negative and a positive impact, the hypothesis holds however, if this is not the case, the statistical 
+  test does not state in which direction the hypothesis is rejected, that is, what eventually is the 
+  alternative hypothesis. But this is necessary to get insights whether maintainability is hampered or not. 
+  Authors may rethink the choice of the hypothesis or even the test itself.} -> What about this?
+  \item \textbf{Thesis idea (PL):} It is important 
+  to understand the correlation between the severity of the vulnerability and the programming language as well 
+  as the maintainability impact. It is not clear how software written a more security-focused language would 
+  inherently be more maintainable (it assumes the flaw is the language and not the person).  
+\end{itemize}
+
 \bibliographystyle{spmpsci}

### 2020-08-21 paper/emse20.tex
images update

@@
 
-\begin{table}[h]
+\begin{table}[H]
+\centering
 \scriptsize
@@
  	\centering 	
-	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
+	\includegraphics[width=0.9\linewidth]{figures/commit.pdf}
  	\caption{Maintainability difference for security commits}
@@
   	\centering
-  	\includegraphics[width=0.51\textwidth]{figures/main_per_guideline.pdf}
+  	\includegraphics[width=0.8\textwidth]{figures/main_per_guideline.pdf}
     \end{adjustwidth}
@@
   	\centering 	
- 	\includegraphics[width=0.4\textwidth]{figures/main_per_severity.pdf}
+ 	\includegraphics[width=0.6\textwidth]{figures/main_per_severity.pdf}
   	\caption{Maintainability difference by vulnerability severity.}
@@
   \centering
-  \includegraphics[width=0.4\textwidth]{figures/main_per_language.pdf}
+  \includegraphics[width=0.6\textwidth]{figures/main_per_language.pdf}
   \caption{Maintainability difference by programming language.}
@@
 \texttt{PHP} are the programming languages with worse 
-impact on maintainability, i.e., with the highest number of negative cases ($46.5\%$, $46.5\%$ and $38.6\%$, respectively). \texttt{Java} and \texttt{Python} 
-seem to be less affected by patching, i.e., integrating a larger amount of cases with positive impact on maintainability ($50.9\%$ and $46.5\%$, respectively). But overall 
+impact on maintainability, i.e., with the highest number of negative cases ($46.5\%$, $46.5\%$ 
+and $38.6\%$, respectively). \texttt{Java} and \texttt{Python} 
+seem to be less affected by patching, i.e., integrating a larger amount of cases with positive 
+impact on maintainability ($50.9\%$ and $46.5\%$, respectively). But overall 
 languages have a considerable amount of cases that negatively impact  
@@
 that tools such as BCH should be integrated into the CI/CD pipelines
-to help developers evaluate the risk of patches of hindering software maintainability---alongside Pull Requests/Code Reviews. We also conclude that different severity 
+to help developers evaluate the risk of patches of hindering software maintainability---alongside 
+Pull Requests/Code Reviews. We also conclude that different severity 
 vulnerabilities may need different levels of attention and better and more secure 
@@
 Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the most likely 
-weaknesses to affect software maintainability. No statistical evidence was retrieved. However, results should not be disregarded because they hint on how 
+weaknesses to affect software maintainability. No statistical evidence was retrieved. 
+However, results should not be disregarded because they hint on how 
 the maintainability is impacted per weakness which highlights the weaknesses
@@
  	\centering 	
-	\includegraphics[width=0.5\textwidth]{figures/main_comparison.pdf}
+	\includegraphics[width=0.7\textwidth]{figures/main_comparison.pdf}
  	\caption{Maintainability difference for security and baseline patches.}

### 2020-09-01 paper/emse20.tex
hot fix
my thoughts
authors names and images and tables resize

@@
 
-\institute{F. Author \at
-              first address \\
-              Tel.: +123-45-678910\\
-              Fax: +123-45-678910\\
-              \email{fauthor@example.com}           %  \\
-%             \emph{Present address:} of F. Author  %  if needed
+\institute{Sofia Reis \at
+			INESC ID and IST, University of Lisbon, Lisbon, Portugal\\
+            \email{sofia.o.reis@tecnico.ulisboa.pt}   
 			\and
-           S. Author \at
-              second address
+			Rui Abreu \at
+			INESC ID and IST, University of Lisbon, Lisbon, Portugal\\
+			\email{rui@computer.org}
+           \and
+			Luis Cruz \at
+			Delft University of Technology, Delft, The Netherlands\\
+			\email{L.Cruz@tudelft.nl}
 }
@@
 curricula needs to be updated; and, more secure programming languages are necessary.
-\keywords{Security \and Software Maintenance \and Open-Source Software}
+\keywords{Software Security \and Software Maintenance \and Open-Source Software}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 \begin{lstlisting}[style={CStyle}, caption={Patch provided by OpenSSL developers to the
-\\CVE-2016-6304 vulnerability on file ssl/t1\_lib.c},label={lst:vuln}]
+CVE-2016-6304 vulnerability on file ssl/t1\_lib.c},label={lst:vuln}]
 static int ssl_scan_clienthello_tlsext(SSL *s, PACKET *pkt, int *al){ 
@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a \\Cross-Site Scripting vulnerability},label={lst:fix}]
+\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a Cross-Site Scripting vulnerability},label={lst:fix}]
    <p class='hint'>
@@
 %
+\begin{figure}[h]
+	 	\centering 	
+    \begin{adjustwidth}{0.5cm}{-1cm}  
+	\includegraphics[width=0.9\textwidth]{figures/methodology.pdf}
+
+ 	\caption{Study Methodology}
+	\label{fig:met}
+	 \end{adjustwidth}
+\end{figure}
+%
 \begin{enumerate}
@@
 %
-\begin{figure}[h]
-	 	\centering 	
-    \begin{adjustwidth}{0.5cm}{-1cm}  
-	\includegraphics[width=0.49\textwidth]{figures/methodology.pdf}
-
- 	\caption{Study Methodology}
-	\label{fig:met}
-	 \end{adjustwidth}
-\end{figure}
+
 %
@@
  	\centering 	
-	\includegraphics[width=0.44\textwidth]{figures/bch_report.png}
+	\includegraphics[width=0.9\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-$2016$-$6304$ vulnerability
@@
 	\caption{Guidelines to produce maintainable code}
-\begin{tabular}{L{1.66cm}L{3.3cm}L{2.7cm}}
+\begin{tabular}{L{2cm}L{4.5cm}L{4cm}}
 
@@
  	\centering 	
-	\includegraphics[width=0.9\linewidth]{figures/commit.pdf}
+	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
  	\caption{Maintainability difference for security commits}
@@
   	\centering
-  	\includegraphics[width=0.8\textwidth]{figures/main_per_guideline.pdf}
+  	\includegraphics[width=0.9\textwidth]{figures/main_per_guideline.pdf}
     \end{adjustwidth}
@@
   	\centering 	
- 	\includegraphics[width=0.6\textwidth]{figures/main_per_severity.pdf}
+ 	\includegraphics[width=0.7\textwidth]{figures/main_per_severity.pdf}
   	\caption{Maintainability difference by vulnerability severity.}
@@
 
-Security-related commits are expected to harm software maintainability. In contrast, one cannot say the same about
+Security-related commits are expected to harm software maintainability. 
+In contrast, one cannot say the same about
 regular changes, in which no significant differences are observed. Thus, we
@@
 %
-This section presents the potential threats to validity.
-%
-\\\textit{\textbf{Construct:}} The formula was inferred based 
+This section presents the potential threats to validity of this study.
+%
+\\\textit{\textbf{Construct Validity:}} The formula was inferred based 
 on the BCH's reports. The high amount of different projects and backgrounds 
@@
 %
-\\\textit{\textbf{Internal:}} The security patches dataset provided by previous
+\\\textit{\textbf{Internal Validity:}} The security patches dataset provided by previous
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
@@
 %
-\\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.
+\\\textit{\textbf{External Validity:}} We use a dataset that only comprises patches of open-source software.
 However, the methodology requires to access data that is not publicly available.
@@
   what to do with this @rui?
+    \begin{itemize}
+      \item Ehmm... strange... we need to improve the approach then. Thoughts?
+    \end{itemize}
   \item [!] \textit{Fig 4, although it looks impressive, I am not sure how to read the figure} - what 
   should we do @rui?
+    \begin{itemize}
+      \item Carefully explain it in the caption?
+    \end{itemize}
   \item [!] \textit{Fig 5, 6 and 7 are not the best ways to portray your findings. For instance, I am 
@@
   is the best way then? 
+    \begin{itemize}
+      \item Strange comment...
+    \end{itemize}
   \item \textit{Authors carefully highlight the key messages of each research question and answer
@@
   the study by vulnerabilities and their implications on systems? Should we divide it by severity levels?
+    \begin{itemize}
+      \item We need to discuss this... I think this related to the next point. 
+    \end{itemize}  
   \item \textit{authors write "...important to write maintainable code to support the production
@@
 Qual é o plano que acha melhor? Como prosseguir?
+
+A discutir, mas acho que por agora era tentar fazer o 1.
 

### 2020-09-03 paper/emse20.tex
changes in sec 1 and sec 2

@@
 
-\title{Fixing Vulnerabilities Potentially Hinders Maintainability%\thanks{Grants or other notes
-%about the article that should go on the front page should be
-%placed here. General acknowledgments should be placed at the end of the article.}
-}
-% \subtitle{Do you have a subtitle?\\ If so, write it here}
-
-%\titlerunning{Short form of title}        % if too long for running head
+\title{Fixing Vulnerabilities Potentially Hinders Maintainability}
 
@@
 }
-
-%\authorrunning{Short form of author list} % if too long for running head
 
@@
 \date{Received: date / Accepted: date}
-% The correct dates will be entered by the editor
 
@@
 \begin{abstract}
-Security is a requirement of utmost importance to produce high-quality software. 
-However, there is still a considerable amount of 
-vulnerabilities being discovered and fixed, almost weekly. We suspect that developers 
-affect the maintainability of their codebases when patching vulnerabilities. This paper evaluates the 
-impact of patches to improve security on the maintainability of open-source 
-software. Maintainability is measured based on the Better Code Hub’s model of 10 
-guidelines using a dataset including 1335 security-related commits.
-Results show evidence of a trade-off between security and maintainability, i.e., 
-developers may be hindering software maintainability.
-The implications of our study are that changes to codebases while patching 
-vulnerabilities need to be performed with extra care; tools
-for patch risk assessment should be integrate into the CI/CD pipeline; computer science
-curricula needs to be updated; and, more secure programming languages are necessary.
+Security is a requirement of utmost importance to produce 
+high-quality software. However, there is still a considerable amount 
+of vulnerabilities being discovered and fixed, almost weekly. We 
+suspect that developers affect the maintainability of their 
+codebases when patching vulnerabilities. This paper evaluates the 
+impact of patches to improve security on the maintainability of 
+open-source software. Maintainability is measured based on the 
+Better Code Hub’s model of 10 guidelines using a dataset including 
+1335 security-related commits. Results show evidence of a trade-off 
+between security and maintainability, i.e., developers may be 
+hindering software maintainability. The implications of our study 
+are that changes to codebases while patching vulnerabilities need to 
+be performed with extra care; tools for patch risk assessment should 
+be integrate into the CI/CD pipeline; computer science curricula 
+needs to be updated; and, more secure programming languages are 
+necessary.
 \keywords{Software Security \and Software Maintenance \and Open-Source Software}
-% \PACS{PACS code1 \and PACS code2 \and more}
-% \subclass{MSC code1 \and MSC code2 \and more}
 \end{abstract}
@@
 %
-Software quality is important because it is ultimately related to the overall
-cost of developing and maintaining software applications~\cite{slaughter1998evaluating}. 
-Software quality characteristics include, but are not limited to functional correctness,
-reliability, usability, maintainability, evolvavility and security. Security is an 
-essential non-functional requirement during the development of software systems. 
-However, there is still a considerable amount of vulnerabilities being
-discovered and fixed, almost weekly, as disclosed by the Zero Day Initiative
+Software quality is important because it is ultimately related to 
+the overall cost of developing and maintaining software 
+applications~\cite{slaughter1998evaluating}. Software quality 
+characteristics include, but are not limited to functional 
+correctness, reliability, usability, maintainability, evolvavility 
+and security. Security is an essential non-functional requirement 
+during the development of software systems. In $2011$, the 
+International Organization for Standardization (ISO) issued an 
+update for software product quality ISO/IEC 25010 considering 
+\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. However, there is still 
+a considerable amount of vulnerabilities being discovered and fixed, 
+almost weekly, as disclosed by the Zero Day Initiative 
 website\footnote{Zero Day Initiative website available at 
-\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
-
-In $2011$, the International Organization for Standardization (ISO) issued an
-update for software product quality ISO/IEC 25010 considering \emph{Security} 
-as one of the main software product quality characteristics~\cite{iso:2011}. 
-However, many developers still lack the knowledge about best practices to deliver 
-secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}.
-ISO describes software maintainability as ``the degree of effectiveness and
-efficiency with which a software product or system can be modified to improve
-it, correct it or adapt it to changes in environment, and in requirements'' on
-software quality ISO/IEC 25010~\cite{iso:2011}. 
-Unmaintainable code is hard to test, analyze and re-use. Therefore, 
-it is important to write maintainable code to support the production 
-of more secure software and prevent 
-the introduction of new vulnerabilities in the code bases.
-
-Novel tools have been built to automatically detect software vulnerabilities (e.g., FindBugs, Infer and more). 
-Despite these efforts, improving software security is not a trivial task and requires 
-implementing patches that might affect software maintainability. 
-Our suspicion is that some of these patches may have a negative 
-impact on the software maintainability and, possibly, even be the cause of the 
-introduction of new vulnerabilities---harming software reliability and introducing technical debt. Therefore, 
-in this paper, we present an empirical study on the impact of patches of 
-vulnerabilities on software maintenance across open-source software.
-
-As ISO does not provide any specific guidelines/formulas to calculate 
-maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
-code analysis service Better Code Hub (BCH) to compute the software compliance 
-with a set of $10$ guidelines/metrics to produce quality software 
-based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. 
-There are other well-known standards and models that have been proposed
-to increase software security: Common Criteria~\cite{common:2009} which received
-negative criticism regarding the costs associated and poor technical evaluation;
-the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
-focused only on web applications, and a model proposed by Xu et al. ($2013$)
-~\cite{6616351} for rating software security (arguably, it was one of the
-first steps taken by SIG to introduce security on their maintainability model). 
-Nevertheless, our study uses BCH to provide an assessment of maintainability in 
-software for the following reasons: BCH integrates a broad number of code metrics; and, 
-code metrics were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
-
-From a methodological 
-perspective, we leveraged a dataset of $1335$ security patches/refactorings collected from open-source 
-software. We calculate software maintainability before and after the patch to measure its 
-impact. This empirical study presents evidence that changes 
-applied in the code bases to patch vulnerabilities affect code maintainability.
-Results also hint that developers should pay different levels of attention 
-to different severity levels and classes of weaknesses when patching vulnerabilities. We also
-show that patches in programming languages such as, \emph{C/C++}, \emph{Ruby} and \emph{PHP}, 
-may have a more negative impact in software maintainability.
-Little information is known about the impact of security patches on the software
-maintainability. With this study, 
-we intend to highlight the need for tools to predict the impact of patches on 
-maintainability~\cite{4724577}, the importance of integrating
-maintainable security in computer science curricula and better programming languages.
-
-The main contributions of the present work are:
+\url{https://www.zerodayinitiative.com/advisories/published/} 
+(Accessed on \today{})}.
+
+Researchers found a correlation between the presence of 
+vulnerabilities on software and code 
+complexity~\cite{shin2010evaluating,10.1145/1774088.1774504}. 
+Security experts claim that complexity hides bugs that may result in 
+security 
+vulnerabilities~\cite{mcgraw2004software,schneier2006beyond}. In 
+practice, an attacker only needs to find one way into the system 
+while a defender needs to find and mitigate all the security issues. 
+Complex code is difficult to understand, maintain and 
+test~\cite{1702388}. Thus, the developer's task gets harder as the 
+codebase grows in size and complexity. But the risk can be minimized 
+by writing clean and maintainable code.
+
+ISO describes \textit{software maintainability} as ``the degree of 
+effectiveness and efficiency with which a software product or system 
+can be modified to improve it, correct it or adapt it to changes in 
+environment, and in requirements'' on software quality ISO/IEC 
+25010~\cite{iso:2011}. Thereby, maintainable security may be 
+defined, briefly, as the degree of effectiveness and efficiency with 
+which software can be changed to mitigate a security 
+vulnerability---corrective maintenance.
+% (hopefully, without introducing new ones). 
+However, many developers still lack knowledge on the best 
+practices to deliver and maintain secure and high-quality 
+software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a world where zero-day vulnerabilities are constantly 
+emerging, mitigation needs to be fast and efficient. Therefore, it 
+is important to write maintainable code to support the production of 
+more secure software---maintainable code is faster 
+to mitigate---and, prevent the introduction of new vulnerabilities 
+in the codebases.
+% Unmaintainable code is hard to test, analyze and re-use. Therefore,
+% it is important to write maintainable code to support the production
+% of more secure software and prevent
+% the introduction of new vulnerabilities in the code bases.
+%
+Novel tools have been built to automatically detect software 
+vulnerabilities (e.g., FindBugs, Infer and more). Despite these 
+efforts, improving software security is not a trivial task and 
+requires implementing patches that might affect software 
+maintainability. Our suspicion is that some of these patches may 
+have a negative impact on the software maintainability and, 
+possibly, even be the cause of the introduction of new 
+vulnerabilities---harming software reliability and introducing 
+technical debt. Therefore, in this paper, we present an empirical 
+study on the impact of patches of vulnerabilities on software 
+maintenance across open-source software.
+
+As ISO does not provide any specific guidelines/formulas to 
+calculate maintainability, we resort to Software Improvement Group 
+(SIG)'s web-based source code analysis service Better Code Hub (BCH) 
+to compute the software compliance with a set of $10$ 
+guidelines/metrics to produce quality software based in ISO/IEC 
+$25010$~\cite{Visser:2016:OREILLY}. There are other well-known 
+standards and models that have been proposed to increase software 
+security: Common Criteria~\cite{common:2009} which received
+negative criticism regarding the costs associated and poor technical 
+evaluation; the OWASP Application Security Verification 
+Criteria~\cite{oswap:2009} which is focused only on web 
+applications, and a model proposed by Xu et al. ($2013$)
+~\cite{6616351} for rating software security (arguably, it was one 
+of the first steps taken by SIG to introduce security on their 
+maintainability model). Nevertheless, our study uses BCH to provide 
+an assessment of maintainability in software for the following 
+reasons: BCH integrates a broad number of code metrics; and, 
+code metrics were empirically validated in previous 
+work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
+
+From a methodological perspective, we leveraged a dataset of $1335$ 
+security patches collected from open-source software. We calculate 
+software maintainability before and after the patch to measure its 
+impact. This empirical study presents evidence that changes applied 
+in the codebases to patch vulnerabilities affect code 
+maintainability. Results also hint that developers 
+should pay different levels of attention to different severity 
+levels and classes of weaknesses when patching vulnerabilities. We 
+also show that patches in programming languages such as, 
+\emph{C/C++}, \emph{Ruby} and \emph{PHP}, may have a more negative 
+impact in software maintainability. Little information is known 
+about the impact of security patches on software 
+maintainability. Developers need to be aware of the impact of their 
+changes on software maintainability while patching security 
+vulnerabilities. The harm of maintainability can increase the time 
+of response of future mitigations or even other maintainability 
+tasks. Thus, it is of utmost importance to find means to assist 
+mitigation and reduce its risks. With this study, we intend to 
+highlight the need for tools to assess the impact of patches on 
+software maintainability~\cite{4724577}; the importance of 
+integrating maintainable security in computer science curricula; 
+and, the demand for better programming languages, designed by 
+leveraging security principles~\cite{kurilova2014wyvern,10.1145/2489828.2489830}.
+
+This research performs the following main contributions:
 %
 \begin{itemize}
-  \item Evidence that supports the trade-off between security and maintainability:
-  developers may be hindering software maintainability while patching vulnerabilities.
-	\item An empirical study on the impact of security changes on software
-  maintainability (per guideline, severity, weakness and programming language).
-	\item A replication package with all scripts and data created to perform the
-	empirical evaluation, for reproducibility. Available online:
-  \url{https://figshare.com/s/4861207064900dfb3372}.
+  \item Evidence that supports the trade-off between security and 
+  maintainability: developers may be hindering software 
+  maintainability while patching vulnerabilities.
+	\item An empirical study on the impact of security patches on 
+	software maintainability (per guideline, severity, weakness and 
+	programming language).
+	\item A replication package with all scripts and data created to 
+	perform the empirical evaluation, for reproducibility. Available 
+	online: \url{https://figshare.com/s/4861207064900dfb3372}.
 \end{itemize}
 %
-This paper is structured as follows: Section~\ref{sec:motivation} introduces an
-example of a security refactoring of a known vulnerability found in the
-protocol implementation of OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
+This paper is structured as follows: Section~\ref{sec:motivation} 
+introduces an example of a security patch of a known vulnerability 
+found in the protocol implementation of 
+OpenSSL\footnote{\label{openssl}OpenSSL is a toolkit that
 contains open-source implementations of the SSL and TLS cryptographic
-protocols. Repository available at \url{https://github.com/openssl/openssl}
-(Accessed on \today{})}; Section~\ref{sec:methodology} describes the
-methodology used to answer the research questions; Section~\ref{sec:results}
-presents the results and discusses their
-implications; Section~\ref{sec:implications} elaborates on the implications
-developers should consider in the future; Section~\ref{sec:threats} enumerates 
-the threats to the validity of
-this study; Section~\ref{sec:rw} describes the different work and existing
-literature in the field of study; and, finally, Section~\ref{sec:conclusions}
-concludes the main findings and elaborates on future work.
+protocols. Repository available at 
+\url{https://github.com/openssl/openssl} (Accessed on \today{})}; 
+Section~\ref{sec:methodology} describes the methodology used to 
+answer the research questions; Section~\ref{sec:results} presents 
+the results and discusses their implications; 
+Section~\ref{sec:implications} elaborates on the implications
+developers should consider in the future; Section~\ref{sec:threats} 
+enumerates the threats to the validity of this study; 
+Section~\ref{sec:rw} describes the different work and existing
+literature in the field of study; and, finally, 
+Section~\ref{sec:conclusions} concludes the main findings and 
+elaborates on future work.
 
@@
 %
-Due to time-to-market pressure and the lack of security expertise~\cite{8077802}, code-related
-security flaws are generally detected \textit{after the fact}, i.e., when
-hackers exploit them. Patching these issues is usually done by modifying the code base. 
-However, these patches may have a
-negative impact on software maintenance (e.g., developers might follow the
-quickest solution to fix the problem but not necessarily the most elegant and
-performant one). Thus, we want to understand how security patches impact
-the maintainability of software.
-
-As an example, consider the patch of the TLS state machine protocol implementation
-in OpenSSL\footnoteref{openssl} to address a memory leak flaw found in the way 
-OpenSSL handled TLS status request extension data during session 
-renegotiation, and where a malicious client could cause a Denial-of-Service
-(DoS) attack via large OCSP Status Request extensions when OCSP stapling support
-was enabled. Online Certificate Status Procotol (OCSP) stapling, formally known as the TLS Certificate 
-Status Request extension, is a standard for checking the revocation status of
-certificates.
-
-This vulnerability is listed at the Common
-Vulnerabilities and Exposures dictionary as CVE-$2016$-$6304$~\footnote{CVE-$2016$-$6304$
-details available at \url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6304}
-(Accessed on \today{})}. It is amongst the vulnerabilities studied in our
-research. The snippet, in Listing~\ref{lst:vuln}, presents the changes performed on the
-\emph{ssl/t1$\_$lib.c} file\footnote{CVE-$2016$-$6304$ fix available  at
+Due to time-to-market pressure and the lack of security 
+expertise~\cite{8077802}, code-related security flaws are generally 
+detected \textit{after the fact}, i.e., when hackers exploit them. 
+Patching these issues is usually done by modifying the codebase. 
+However, these patches may have a negative impact on software 
+maintenance (e.g., developers might follow the quickest solution to 
+fix the problem but not necessarily the most elegant and performant 
+one). Thus, we want to understand how security patches impact
+software maintainability.
+
+As an example, consider the patch of the TLS state machine protocol 
+implementation in OpenSSL\footnoteref{openssl} to address a memory 
+leak flaw found in the way how OpenSSL handled TLS status request 
+extension data during session renegotiation, and where a malicious 
+client could cause a Denial-of-Service (DoS) attack via large OCSP 
+Status Request extensions when OCSP stapling support was enabled. 
+Online Certificate Status Procotol (OCSP) stapling, formally known 
+as the TLS Certificate Status Request extension, is a standard for 
+checking the revocation status of certificates.
+
+This vulnerability is listed at the Common Vulnerabilities and 
+Exposures dictionary as CVE-$2016$-$6304$~\footnote{CVE-$2016$-$6304$
+details available at 
+\url{http://cve.mitre.org/cgi-bin/cvename.cgi?name=CVE-2016-6304}
+(Accessed on \today{})}. It is amongst the vulnerabilities studied 
+in our research. The snippet, in Listing~\ref{lst:vuln}, presents 
+the changes performed on the
+\emph{ssl/t1$\_$lib.c} file\footnote{CVE-$2016$-$6304$ fix available 
+ at
 \url{https://github.com/openssl/openssl/commit/e408c09bbf7c3057bda4b8d20bec1b3a7771c15b}
-(Accessed on \today{})} by the OpenSSL developers to patch the vulnerability. 
-Every SSL/TLS connection begins with a handshake 
+(Accessed on \today{})} by the OpenSSL developers to patch the 
+vulnerability. Every SSL/TLS connection begins with a handshake 
 which is responsible for the negotiation between the two parties. 
-The OSCP Status Request extension allows the client to verify
-the server certificate and enables a TLS server to include
-its response in the handshake. The problem in CVE-$2016$-$6304$
-is a flaw in the logic of OpenSSL that does not handle memory efficiently
-when large OCSP Status Request extensions are sent each time a client 
-requests renegotiation. This was possible because the OCSP responses IDs 
-were not released between handshakes. Instead, they would be allocated 
-again and again. Thus, if a malicious client does it several times it may lead to 
-an unbounded memory growth on the server and, eventually, lead to a 
-DoS attack through memory exhaustion.
+The OSCP Status Request extension allows the client to verify the 
+server certificate and enables a TLS server to include its response 
+in the handshake. The problem in CVE-$2016$-$6304$ is a flaw in the 
+logic of OpenSSL that does not handle memory efficiently when large 
+OCSP Status Request extensions are sent each time a client requests 
+renegotiation. This was possible because the OCSP responses IDs 
+were not released between handshakes. Instead, they would be 
+allocated again and again. Thus, if a malicious client does it 
+several times it may lead to an unbounded memory growth on the 
+server and, eventually, lead to a DoS attack through memory 
+exhaustion.
+
+The code changes performed to patch the vulnerability were:
+
+\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function 
+removes any memory allocated to the OCSP response IDs 
+(OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded 
+memory growth - which was not being performed before.
+After releasing the unbounded memory, the logic condition in 
+\ref{lst:func3} was shifted to \ref{lst:func2} which is responsible 
+for handling the application when no OCSP response IDS are 
+allocated. Now, instead of performing it after the packages 
+processing, it does it previously.
+
+Patching this vulnerability seems a rudimentary task. Yet,
+a considerable amount of changes were performed in the codebase 
+which yielded to a negative impact on software maintainability.
+While patching, the developer introduced $6$ new lines in a method 
+already with a large number of lines of code and introduced more 
+complexity to the code with new branch points which disrupt two of 
+the guidelines proposed by the Software Improvement Group 
+(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}: 
+\emph{Write Short Units of Code} and \emph{Write Simple Units of 
+Code}.
 
@@
 
-The code changes performed to patch the vulnerability were:
-
-\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function removes any
-memory allocated to the OCSP response IDs (OCSP$\_$RESPIDs) from a previous 
-handshake to prevent unbounded memory growth - which was not being performed before.
-After releasing the unbounded memory, the program performs the same check 
-in \ref{lst:func2} that used to perform in \ref{lst:func3} which is responsible 
-for handling the application when no OCSP response IDS were being allocated. Now, 
-instead of performing it after packaging processing, it does it previously.
-
-Patching this vulnerability seems a rudimentary task. Yet,
-a considerable amount of changes were performed
-in the codebase which yielded to a negative impact on software maintainability.
-While patching, the developer introduced $6$ new lines in a method already with 
-a large number of lines of code and introduced more complexity to the code
-with new branch points which disrupt two of the guidelines proposed by the Software Improvement Group 
-(SIG) for building maintainable software~\cite{Visser:2016:OREILLY}: \emph{Write Short Units of Code}
-and \emph{Write Simple Units of Code}.
-
-Software maintainability is designated as the degree to which an application is understood, 
-repaired or enhanced. In this paper, our concern is to study whether, while improving software
-security, developers are also negatively impacting the maintainability of their
-software applications. This is important because software maintainability is 
-approximately $75\%$ of the cost related to a project. To answer the following three 
-research questions, we use two datasets of security patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064} 
-to measure the impact of security patches on the maintainability of open-source software. 
+Software maintainability is designated as the degree to which an 
+application is understood, repaired or enhanced. In this paper, our 
+concern is to study whether, while improving software security, 
+developers are also negatively impacting the maintainability of their
+software applications. This is important because software 
+maintainability is approximately $75\%$ of the cost related to a 
+project. To answer the following three research questions, we use 
+two datasets of security 
+patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064} to measure the 
+impact of security patches on the maintainability of open-source 
+software. 
 %
@@
 \textit{\textbf{RQ1: What is the impact of security patches on the
-maintainability of open-source software?}} Often security flaws require patching 
-code to make software more secure.
-However, \textbf{there is no evidence yet of how security patches impact the
-maintainability of open-source software}. Our suspicion is that developers tend
-to introduce technical debt in their software when patching the code base to
-address a security flaw because they tend to choose the easiest path to solve
-it. To address it, we compute the maintainability of $1335$ commits using the
-\emph{Better Code Hub} tool.
+maintainability of open-source software?}} Often security flaws 
+require patching code to make software more secure. However, 
+\textbf{there is no evidence yet of how security patches impact the
+maintainability of open-source software}. Our suspicion is that 
+developers tend to introduce technical debt in their software when 
+patching the codebases to address a security flaw because they tend 
+to choose the easiest path to solve it. To address it, we compute 
+the maintainability of $1335$ commits using the \emph{Better Code Hub} tool.
 
 \textit{\textbf{RQ2: Which weaknesses are more likely to
-affect open-source software maintainability?}}
-There are security flaws that are more difficult to patch than others. For
-instance, implementing secure authentication is not as easy as patching a
-cross-site scripting vulnerability since usually the last one can be fixed
-without adding new lines of code/complexity to the code. A typical fix for 
-the cross-site scripting vulnerability is presented in Listing~\ref{lst:fix}. 
-The developer added the function \texttt{htmlentities} to escape the data given by the variable
-\texttt{\$\_['file']}. \textbf{Knowing which weaknesses are more likely to increase
-maintainability issues is one step forward to bring awareness to security
-engineers of what weaknesses need more attention}. The taxonomy of security
-patterns used to answer this question is the one provided by the Common Weakness Enumeration
-(CWE). \emph{Weakness}, according to the Common Weakness Enumeration (CWE) glossary, 
-is a type of code-flaw that could contribute to the 
-introduction of vulnerabilities within that product. Maintainability is measured
-separately for each weakness.
+affect open-source software maintainability?}} There are security 
+flaws that are more difficult to patch than others. For instance, 
+implementing secure authentication is not as easy as patching a
+cross-site scripting vulnerability since the latter can be fixed
+without adding new lines of code/complexity to the code. A typical 
+fix for the cross-site scripting vulnerability is presented in 
+Listing~\ref{lst:fix}. The developer added the function 
+\texttt{htmlentities} to escape the data given by the variable
+\texttt{\$\_['file']}. \textbf{Knowing which weaknesses are more 
+likely to increase maintainability issues is one step forward to 
+bring awareness to security engineers of what weaknesses need more 
+attention}. The taxonomy of security patterns used to answer this 
+question is the one provided by the Common Weakness Enumeration
+(CWE). \emph{Weakness}, according to the Common Weakness Enumeration 
+(CWE) glossary, is a type of code-flaw that could contribute to the 
+introduction of vulnerabilities within that product. Maintainability 
+is measured separately for each weakness.
 %
@@
 %
-\textit{\textbf{RQ3: What is the impact of security patches versus regular changes on the
-maintainability of open-source software?}}
-%
-We also compute the maintainability of random regular commits using the
-\emph{Better Code Hub} tool---baseline.  We use them to understand \textbf{how
-maintainability evolves when security patches are performed versus when
-they are not}.
+\textit{\textbf{RQ3: What is the impact of security patches versus 
+regular changes on the maintainability of open-source software?}}
+%
+We also compute the maintainability of random regular commits using 
+the \emph{Better Code Hub} tool---baseline. We use them to 
+understand \textbf{how maintainability evolves when security patches 
+are performed versus when they are not}.
 %
@@
 the impact of security patches on the maintainbility of open-source
-software. The methodology comprises the following steps, as illustrated in
-Fig.~\ref{fig:met}.
+software. The methodology comprises the following steps, as 
+illustrated in Fig.~\ref{fig:met}.
 %
@@
 	code bases from the combined dataset containing $1335$ security
-	refactorings collected from open-source software available on
+	patches collected from open-source software available on
 	GitHub.
@@
 The $1335$ patches in the dataset were run against the BCH toolset to
-calculate their maintainability reports. However, there are refactorings
+calculate their maintainability reports. However, there are patches
 discarded from our study due to limitations of BCH: in particular, lack of
@@
 % This pattern comprises
-%  		several other security refactorings that do not have a CWE assigned and patterns that
-%  		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
+%  		several other security patches that do not have a CWE assigned and patterns that
+%  		do not satisfy the Wilcoxon test size requirement of more than $20$ patches.\\
 % \bottomrule
@@
 % 	\item \textbf{Miscellaneous (673 cases)} This pattern comprises
-% 		several other security refactorings that do not have a CWE assigned and patterns that
-% 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.
+% 		several other security patches that do not have a CWE assigned and patterns that
+% 		do not satisfy the Wilcoxon test size requirement of more than $20$ patches.
 % \end{itemize}
@@
 open-source software maintainability~\cite{HEGEDUS2018313}. However, there is no
-previous work focused on comparing the impact of security refactorings with
+previous work focused on comparing the impact of security patches with
 regular refactorings on maintainability.
@@
 As said before, the web-based source code analysis service \emph{Better Code
-Hub} (BCH) is used to collect the maintainability reports of the refactorings of
+Hub} (BCH) is used to collect the maintainability reports of the patches of
 each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
@@
  	\caption{Maintainability report of OpenSSL's CVE-$2016$-$6304$ vulnerability
-refactoring for the guideline \emph{Write Simple Units of Code} provided by
+patch for the guideline \emph{Write Simple Units of Code} provided by
 \emph{Better Code Hub}. This version of OpenSSL does not comply with the
@@
 provided by BCH for a project after finishing its evaluation. The example
-refers to the OpenSSL CVE-$2016$-$6304$ vulnerability refactoring---
+refers to the OpenSSL CVE-$2016$-$6304$ vulnerability patch---
 as described by Section~\ref{sec:motivation}. This
@@
 work on measuring the impact of energy-oriented 
-refactorings~\cite{8919169}. Cruz et al. ($2019$) proposed an 
+patches~\cite{8919169}. Cruz et al. ($2019$) proposed an 
 equation to capture the distance between the current state of the project and
@@
 violated when the patch requires to input new information to a function/class 
-and developers struggle to use the \emph{Introduce Parameter Object} refactoring 
+and developers struggle to use the \emph{Introduce Parameter Object} patch 
 pattern. Results do not provide statistical significance to the \emph{Separate 
@@
 Thus, while developers 
-tend to ignore refactoring techniques such as the \emph{Replace Method 
+tend to ignore patch techniques such as the \emph{Replace Method 
 with Method Object} technique, they may also be introducing new software 
@@
 commit is limiting the type of regular commits (e.g, input 
-refactorings, variables or functions, type conversion, etc).  
+patches, variables or functions, type conversion, etc).  
 We strongly believe that this phenomenon lead
@@
 
-Many studies have investigated the relationship between refactorings and
+Many studies have investigated the relationship between patches and
 software quality. Previous work focused on object-oriented metrics has evaluated the
-impact of refactorings and exhibited proof that quantifying effectively the
-impact of refactorings on maintainability may help to choose the appropriate
-refactoring type~\cite{1167822}. In contrast to this work, Hegedus et
+impact of patches and exhibited proof that quantifying effectively the
+impact of patches on maintainability may help to choose the appropriate
+patch type~\cite{1167822}. In contrast to this work, Hegedus et
 al.~\cite{HEGEDUS2018313} did not select particular metrics to assess the effect
-of refactorings. Instead, statistical tests were used to find the metrics that
-have the potential to change significantly after refactorings. Studying the evolution 
+of patches. Instead, statistical tests were used to find the metrics that
+have the potential to change significantly after patches. Studying the evolution 
 of maintainability issues during the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
@@
 investigate which guidelines affect most the maintainability per
-weakness; check if vulnerability patches are followed by refactoring
+weakness; check if vulnerability patches are followed by patch
 commits and how much time does it take to do it; 
@@
   Improve this.
-  \item Drop the word refactoring. Security patches are not refactorings. Refactorings is perfective
+  \item Drop the word patch. Security patches are not patches. Refactorings is perfective
   or preventive maintenance, whereas patching a vulnerability is corrective maintenance.

### 2020-09-03 paper/emse20.tex
hot fixes

@@
 
-\newcommand\Sof[1]{\nb{Sofia}{red}{#1}}
+\newcommand\Sofia[1]{\nb{Sofia}{red}{#1}}
 \newcommand\Luis[1]{\nb{Luis}{mygreen}{#1}}
@@
 practices to deliver and maintain secure and high-quality 
-software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a world where zero-day vulnerabilities are constantly 
-emerging, mitigation needs to be fast and efficient. Therefore, it 
+software~\cite{Pothamsetty:2005:SEL:1107622.1107635,8077802}. In a 
+world where zero-day vulnerabilities are constantly emerging, 
+mitigation needs to be fast and efficient. Therefore, it 
 is important to write maintainable code to support the production of 
-more secure software---maintainable code is faster 
-to mitigate---and, prevent the introduction of new vulnerabilities 
-in the codebases.
+more secure software---maintainable code is less complex and, 
+consequently, less prone to
+vulnerabilities~\cite{shin2010evaluating,10.1145/1774088.1774504})---
+and, prevent the introduction of new vulnerabilities.
 % Unmaintainable code is hard to test, analyze and re-use. Therefore,
@@
 \begin{itemize}
+	\vspace{-0.3em}
   \item Evidence that supports the trade-off between security and 
@@
 %
+\begin{enumerate}
+	\item Combine the datasets from related work that classifies
+	the activities of developers addressing security-oriented 
+	patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064}.
+%
+	\item
+	Extract relevant data (e.g., owner and name of
+	the repository, sha key of the vulnerable version, sha key of 	
+	the fixed version and codebases) from the combined dataset 	
+	containing $1335$ security patches collected from open-source 	
+	software available on GitHub.
+%
+	\item A baseline of regular commits was randomly collected from 
+	the list of projects of the final dataset to evaluate the impact 
+	of regular commits on the maintainability of open-source 
+	software.
+%
+  \item Use the Software Improvement Group's\footnote{SIG's website: 
+  \url{https://www.sig.eu/} (Accessed on \today{})} web-based source 
+  code analysis service \emph{Better Code Hub} (BCH)\footnote{BCH's 
+  website: \url{https://bettercodehub.com/} (Accessed on \today{})} 
+  to quantify maintainability for both security and regular commits. 
+  The codebases that resemble to the commits of our datasets are 
+  collected before the analysis performed by BCH.
+\end{enumerate}
+%
 \begin{figure}[h]
@@
 	\includegraphics[width=0.9\textwidth]{figures/methodology.pdf}
-
  	\caption{Study Methodology}
@@
 %
-\begin{enumerate}
-	\item Combine the datasets from related work that classifies
-the activities of developers addressing security-oriented patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064}.
-%
-	\item
-	Extract relevant data (e.g., owner and name of
-	the repository, sha key of the vulnerable version, sha key of the fixed version) to the experiments and 
-	code bases from the combined dataset containing $1335$ security
-	patches collected from open-source software available on
-	GitHub.
-%
-	\item A baseline of regular commits was randomly collected from the list of
-	projects of the final dataset to evaluate the impact of regular commits on the
-	maintainability of open-source software.
-%
-  \item Use the Software Improvement Group's\footnote{SIG's website: \url{https://www.sig.eu/} 
-  (Accessed on \today{})} web-based source code analysis
-  service \emph{Better Code Hub} (BCH)\footnote{BCH's website:
-  \url{https://bettercodehub.com/} (Accessed on \today{})} to quantify maintainability
-  for both security and regular commits. The code bases that resemble to the commits
-  of our datasets are collected before the analysis performed by BCH.
-\end{enumerate}
-%
-
-%
 \subsection{Datasets}
@@
 We use a combined dataset of security vulnerabilities which is 
-the outcome of mining and manually inspect a total of $265$ GitHub projects for both
-datasets. Reis and Abreu 
-($2017$) mined open-source
-software aiming at the extraction of real examples---created by real
-developers---to test and assess the performance of static analysis tools~\cite{Reis:2017:IJSSE} since
-using hand-seeded test cases or mutations could lead to misleading assessments
-of the capabilities of the tools~\cite{just2014mutants}. The study yielded a
-dataset of $711$ test cases for $16$ different security vulnerability types. Some of the vulnerability types
-are based on the OWASP Top $10$ of $2013$~\cite{oswap:2013} and OWASP Top $10$ of
-$2017$~\cite{oswap:2017}. Each test case of the
-dataset is a triplet: the commit before the patching, the commit responsible
-for the patching, and the snippets of code that differ from one version to
-another (typically, called \textit{diffs})---where one can easily review the
-code used to fix the security flaw. 
-%
-
-Ponta et al. ($2019$) tracked the \url{pivotal.io} website for vulnerabilities 
-from $2014$ to $2019$. For each new vulnerability, the authors manually searched 
-for references to commits involved in the patch on the National Vulnerability 
-Database (NVD) website. However, $70\%$
-of the vulnerabilities did not have any references to commits. Thus, the authors
-used their expertise to locate the commits in the repositories. This technique 
-yielded a database of $624$ patches~\cite{10.1109/MSR.2019.00064}. The original 
-dataset contains $1282$ commits. One patch can have multiple commits assigned.
-To fit the dataset in our methodology, we search for the first and last commits
-used to patch the vulnerability. We assumed the last commit of the patch as 
-the fix (\emph{sha}) and the parent of the first commit as the vulnerable version 
+the outcome of mining and manually inspecting a total of $265$ 
+GitHub projects for both datasets. Reis and Abreu 
+($2017$) mined open-source software aiming at the extraction of 
+real---created by developers---samples of security vulnerabilities 
+to test and assess the performance of static analysis 
+tools~\cite{Reis:2017:IJSSE} since using hand-seeded test cases or 
+mutations can lead to misleading assessments of the capabilities of 
+the tools~\cite{just2014mutants}. The study yielded a dataset of 
+$711$ test cases for $16$ different security vulnerability types. 
+Some of the vulnerability types are based on the OWASP Top $10$ of 
+$2013$~\cite{oswap:2013} and OWASP Top $10$ of 
+$2017$~\cite{oswap:2017}. Each test case of the dataset is a 
+triplet: the commit before the patching, the commit responsible
+for the patching, and the snippets of code that differ from one 
+version to another (typically, called \textit{diffs})---where one 
+can easily review the code used to fix the security flaw. 
+%
+
+Ponta et al. ($2019$) used a different approach and tracked the 
+\url{pivotal.io} website for vulnerabilities from $2014$ to $2019$. 
+For each new vulnerability, the authors manually searched for 
+references to commits involved in the patch on the National 
+Vulnerability Database (NVD) website. However, $70\%$ of the 
+vulnerabilities did not have any references to commits. Thus, the 
+authors used their expertise to locate the commits in the 
+repositories. This technique yielded a dataset of $624$ 
+patches~\cite{10.1109/MSR.2019.00064}. The original dataset contains 
+$1282$ commits. One patch can have multiple commits assigned.
+To fit the dataset in our methodology, we located the first and 
+last commits used to patch the vulnerability. For these cases, we 
+used the GitHub API to retrieve automatically the dates of the 
+commits. Then, for each patch, the group of commits was ordered from 
+the oldest commit to the newest one. We assumed the last commit 
+(newest one) as the fix (\emph{sha}) and the parent of the 
+first commit (oldest commit) as the vulnerable version 
 (\emph{sha-p}). 
@@
 
-In this study, we focus on computing the
-maintainability of the commits before and after the security patching to
-evaluate if its impact was positive, negative or none.
-The $1335$ patches in the dataset were run against the BCH toolset to
-calculate their maintainability reports. However, there are patches
-discarded from our study due to limitations of BCH: in particular, lack of
-language support and project size. The final dataset used in this paper comprises
-$969$ security patches. $867$ of these patches are identified with Common Weakness
-Enumeration IDs (CWE) where $62.1\%$ were scraped from the National Vulnerability 
-Dataset (NVD) and the remaining $37.9\%$ were manually classified following the 
+In this study, we focus on computing the maintainability of the 
+commits before and after the security patching to evaluate if its 
+impact was positive, negative or none. The $1335$ patches in the 
+dataset were run against the BCH toolset to calculate their 
+maintainability reports. However, there are patches discarded from 
+our study due to limitations of BCH: in particular, lack of
+language support and project size. The final dataset used in this 
+paper comprises $969$ security patches. $867$ of these patches are 
+identified with Common Weakness Enumeration IDs (CWE) where $62.1\%$ 
+were scraped from the National Vulnerability Dataset (NVD) and the 
+remaining $37.9\%$ were manually classified following the 
 \emph{Research Concepts} CWE's list.
@@
 %
-Previous studies attempted to measure the impact of regular refactorings on
-open-source software maintainability~\cite{HEGEDUS2018313}. However, there is no
-previous work focused on comparing the impact of security patches with
-regular refactorings on maintainability.
-We analyze the maintainability of regular commits (i.e., commits not related
-to security fixes) and use them as a baseline.
+Previous studies attempted to measure the impact of regular commits 
+on open-source software maintainability~\cite{HEGEDUS2018313}. 
+However, there is no previous work focused on comparing the impact 
+of security patches with regular commits on maintainability.
+We analyze the maintainability of regular commits (i.e., commits not 
+related to security patches) and use them as a baseline.
 The baseline dataset is generated from the security commits dataset.
-For each
-security commit, a commit with an approximate size is fetched. Due to the complexity 
-of some patches, it was not possible to find patches with the exact number of added and 
-deleted lines. Our search uses the patch \emph{diff} to look for a new similar \emph{diff}. 
-The algorithm integrates the following steps:
+For both vulnerable/sha-p and patch/sha commits, a commit 
+with an approximate size is fetched. Due to the complexity of some 
+patches, it was not possible to find patches with the exact number 
+of added and deleted lines. Our search uses the patch \emph{diff} to 
+look for a \emph{diff} of the same size. The algorithm integrates 
+the following steps:
 
@@
 \begin{itemize}
-	\item How this is a problem or has an impact to the software project? 
-  What is the message to developers? (clarify)
+	% \item How this is a problem or has an impact to the software project?
+	%   What is the message to developers? (clarify)
 	\item Implications: Discuss that depending on the project, maybe security 
@@
   as evidence (check if the literature available is not enough).
-  \item Review the secure programming languages implication. Check 
+  \item Implications: Review the secure programming languages implication. Check 
   if the idea of fully secure exists. The idea should be more secure, not
@@
     won't they be better prepared?
-  \item [!] \textit{It did take me several readings to understand the paper, especially the approach} - 
-  what to do with this @rui?
-    \begin{itemize}
-      \item Ehmm... strange... we need to improve the approach then. Thoughts?
-    \end{itemize}
+  % \item [!] \textit{It did take me several readings to understand the paper, especially the approach} -
+ %  what to do with this @rui?
+ %    \begin{itemize}
+ %      \item Ehmm... strange... we need to improve the approach then. Thoughts?
+ %    \end{itemize}
   \item [!] \textit{Fig 4, although it looks impressive, I am not sure how to read the figure} - what 
@@
     \end{itemize}  
-  \item \textit{authors write "...important to write maintainable code to support the production
-  of more secure software...". Relationship is not clear nor is it backed by some evidence} -> 
-  Improve this.
-  \item Drop the word patch. Security patches are not patches. Refactorings is perfective
-  or preventive maintenance, whereas patching a vulnerability is corrective maintenance.
+  % \item \textit{authors write "...important to write maintainable code to support the production
+  % of more secure software...". Relationship is not clear nor is it backed by some evidence} ->
+  % Improve this.
+  % \item Drop the word patch. Security patches are not patches. Refactorings is perfective
+  % or preventive maintenance, whereas patching a vulnerability is corrective maintenance.
   \item Discuss better the fact of not using other important aspects for the baseline such as the
@@
   come to this conclusion.} -> We are overselling here. Improve it.
-  \item \textit{Authors state that they assume for the Pontas dataset that the last commit related 
-  to patching being the actual patch while the parent of the first patch-related commit is assumed 
-  to be the vulnerable version. However, nothing is said whether this assumption holds all the time. 
-  So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.
+  % \item \textit{Authors state that they assume for the Pontas dataset that the last commit related
+  % to patching being the actual patch while the parent of the first patch-related commit is assumed
+  % to be the vulnerable version. However, nothing is said whether this assumption holds all the time.
+  % So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.
   \item \textit{In Section 3-C, for computing and comparing the maintainability, only security commits  

### 2020-09-03 paper/emse20.tex
meth

@@
 
-Ponta et al. ($2019$) used a different approach and tracked the 
-\url{pivotal.io} website for vulnerabilities from $2014$ to $2019$. 
-For each new vulnerability, the authors manually searched for 
-references to commits involved in the patch on the National 
-Vulnerability Database (NVD) website. However, $70\%$ of the 
-vulnerabilities did not have any references to commits. Thus, the 
-authors used their expertise to locate the commits in the 
-repositories. This technique yielded a dataset of $624$ 
-patches~\cite{10.1109/MSR.2019.00064}. The original dataset contains 
-$1282$ commits. One patch can have multiple commits assigned.
-To fit the dataset in our methodology, we located the first and 
-last commits used to patch the vulnerability. For these cases, we 
-used the GitHub API to retrieve automatically the dates of the 
-commits. Then, for each patch, the group of commits was ordered from 
-the oldest commit to the newest one. We assumed the last commit 
+Ponta et al. ($2019$) tracked the \url{pivotal.io} website for 
+vulnerabilities from $2014$ to $2019$. For each new vulnerability, 
+the authors manually searched for references to commits involved in 
+the patch on the National Vulnerability Database (NVD) website. 
+However, $70\%$ of the vulnerabilities did not have any references 
+to commits. Thus, the authors used their expertise to locate the 
+commits in the repositories. This technique yielded a dataset of 
+$624$ patches~\cite{10.1109/MSR.2019.00064}. The original dataset 
+contains $1282$ commits. One patch can have multiple commits 
+assigned. To fit the dataset in our methodology, we located the 
+first and last commits used to patch the vulnerability. For these 
+cases, we used the GitHub API to retrieve automatically the dates of 
+the commits. Then, for each patch, the group of commits was ordered 
+from the oldest commit to the newest one. We assumed the last commit 
 (newest one) as the fix (\emph{sha}) and the parent of the 
@@
 The baseline dataset is generated from the security commits dataset. 
-For both vulnerable/sha-p and patch/sha commits, a commit 
-with an approximate size is fetched. Due to the complexity of some 
-patches, it was not possible to find patches with the exact number 
-of added and deleted lines. Our search uses the patch \emph{diff} to 
-look for a \emph{diff} of the same size. The algorithm integrates 
-the following steps:
+As for security patches, for each regular fix or refatoring, we need 
+a the commit performing the regular change (sha-reg) and version of 
+the software before the change (sha-reg-p). For each security patch, 
+a \emph{diff} with an approximate size was fetched to a regular 
+change. Due to the complexity of some patches, it was not possible 
+to find patches with the exact number of added and deleted lines. 
+Our search uses the security patch \emph{diff} to look for a 
+\emph{diff} of the same size. The algorithm integrates the following 
+steps:
 
 \begin{enumerate}
-\item Calculates the \emph{diff} between the security patch commit (\emph{sha}) and its parent (\emph{sha-p}).
+\item Calculates the \emph{diff} between the security patch commit 
+(\emph{sha}) and its parent (\emph{sha-p}).
 \item Selects a random commit from the project (\emph{sha-reg}).
 \item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
-\item Checks if the size from the regular change has the same/approximated size to 
-the security patch.
+\item Checks if the size from the regular change has the 
+same/approximated size to the security patch.
 \end{enumerate}
 
-The pair of the random commit and its parent is accepted if the \emph{diff} 
-size fits in the range size. This range widens every $10$ attempts to
-search for a change with the same size. We originate the regular commits from the 
-security commits to ensure that differences in maintainability are not a 
-consequence of characteristics of different projects.
+The pair of the random commit and its parent is accepted if the 
+\emph{diff} size fits in the range size. This range widens every 
+$10$ attempts to search for a change with the same size. We 
+originate the regular commits from the security commits to ensure 
+that differences in maintainability are not a consequence of 
+characteristics of different projects.
 
@@
 
-In this research, we follow a very similar methodology to the one presented 
-in previous work on the maintainability of energy-oriented fixes~\cite{8919169}. 
-As said before, the web-based source code analysis service \emph{Better Code
-Hub} (BCH) is used to collect the maintainability reports of the patches of
-each project. Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
+In this research, we follow a very similar methodology to the one 
+presented in previous work on the maintainability of energy-oriented 
+fixes~\cite{8919169}. As said before, the web-based source code 
+analysis service \emph{Better Code Hub} (BCH) is used to collect the 
+maintainability reports of the patches of each project. 
+Table~\ref{tab:guidelines} presents the $10$ guidelines proposed
 by BCH's authors for delivering software that is not difficult to
-maintain~\cite{Visser:2016:OREILLY} and, maps each guideline to the metric 
-calculated by BCH. Theses guidelines are calculated using the metrics 
-presented in~\cite{criteria:2017} and also briefly explained on Table~\ref{tab:guidelines}. 
-During each guideline evaluation, the tool determines the compliance towards 
-one guideline by establishing limits for the percentage of code allowed to be 
-in each of the $4$ risk severity levels
-(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
-risk}). If the project does not violate the thresholds, then it is compliant
-with the guideline. These thresholds are calibrated by BCH using their own
-data/experience---using open-source and closed software systems. If a project is
-compliant with a guideline, it means that it is at least $65\%$ better than the
-software used by BCH to calculate the thresholds\footnote{Check the answer to
-\emph{How can I adjust the threshold for passing/not passing a guideline?} at
+maintain~\cite{Visser:2016:OREILLY} and, maps each guideline to the 
+metric calculated by BCH. Theses guidelines are calculated using the 
+metrics presented in~\cite{criteria:2017} and are also briefly 
+explained on Table~\ref{tab:guidelines}. 
+During each guideline evaluation, the tool determines the compliance 
+towards one guideline by establishing limits for the percentage of 
+code allowed to be in each of the $4$ risk severity levels
+(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and 
+\emph{very high risk}). If the project does not violate those 
+thresholds, then the BCH considers that the code is compliant with 
+a guideline. These thresholds are determined by BCH using their own
+data/experience---using open-source and closed software systems. If 
+a project is compliant with a guideline, it means that it is at 
+least $65\%$ better than the software used by BCH to calculate the 
+thresholds\footnote{Check the answer to \emph{How can I adjust the 
+threshold for passing/not passing a guideline?} at
 \url{https://bettercodehub.com/docs/faq} (Accessed on \today{})}.
@@
 	\includegraphics[width=0.9\textwidth]{figures/bch_report.png}
- 	\caption{Maintainability report of OpenSSL's CVE-$2016$-$6304$ vulnerability
-patch for the guideline \emph{Write Simple Units of Code} provided by
-\emph{Better Code Hub}. This version of OpenSSL does not comply with the
-guideline in the example since the bars do not reach the threshold points. This
-example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
+ 	\caption{Maintainability report of OpenSSL's CVE-$2016$-$6304$ 
+	vulnerability patch for the guideline \emph{Write Simple Units 
+	of Code} provided by \emph{Better Code Hub}. This version of 
+	OpenSSL does not comply with the guideline in the example since 
+	the bars do not reach the threshold points. This example only 
+	complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean 
+	Code}).}
 	\label{fig:bchrep}
@@
 \scriptsize
-	\caption{Guidelines to produce maintainable code}
+	\caption{Guidelines to produce maintainable code.}
 \begin{tabular}{L{2cm}L{4.5cm}L{4cm}}
@@
 
-Fig.~\ref{fig:bchrep} shows an example of the report
-provided by BCH for a project after finishing its evaluation. The example
+Fig.~\ref{fig:bchrep} shows an example of the report provided by BCH 
+for a project after finishing its evaluation. The example
 refers to the OpenSSL CVE-$2016$-$6304$ vulnerability patch---
-as described by Section~\ref{sec:motivation}. This
-version of OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write
+as described by Section~\ref{sec:motivation}. This version of 
+OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write 
 Clean Code}.
 
-SIG defines \emph{Units} as the smallest groups of code that can be maintained
-and executed independently~\cite{Visser:2016:OREILLY} (e.g., methods and
-constructors in Java). One of the guidelines with which the project does not
-comply is the one presented in the report (cf. Fig.~\ref{fig:bchrep}):
-\emph{Write Simple Units of Code}. BCH analyzes this guideline based on the
-McCabe Complexity~\cite{1702388} to calculate the number of branch points of a
-method. The bar at the bottom of the figure represents the top $30$ units that
-violate the guideline, sorted by severity. The violation severities are
-indicated using colors, and there is a legend to help to interpret them. The
-green bar represents the number of compliant branch points per unit (\emph{at
-most $5$}), i.e., the number of units are compliant with ISO
-$25010$~\cite{iso:2011}. Yellow, orange and red bars represent units that do not
-comply with medium (\emph{above $5$}), high (\emph{above $10$}) and very high
-(\emph{above $25$}) severity levels. In the bar, there are marks that pinpoint
-the compliance thresholds for each severity level. If the green mark is
-somewhere in the green bar it means it is compliant with a low level of
+SIG defines \emph{Units} as the smallest groups of code that can be 
+maintained and executed independently~\cite{Visser:2016:OREILLY} 
+(e.g., methods and constructors in Java). One of the guidelines with 
+which the project does not comply is the one presented in the report 
+(cf. Fig.~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. BCH 
+analyzes this guideline based on the McCabe 
+Complexity~\cite{1702388} to calculate the number of branch points 
+of a method. The bar at the bottom of the figure represents the top 
+$30$ units that violate the guideline, sorted by severity. The 
+different severities of violating the guideline are indicated using 
+colors, and there is a legend to help to interpret them. The green 
+bar represents the number of compliant branch points per unit 
+(\emph{at most $5$}), i.e., the number of units are compliant with 
+ISO $25010$~\cite{iso:2011}. Yellow, orange and red bars represent 
+units that do not comply with medium (\emph{above $5$}), high 
+(\emph{above $10$}) and very high (\emph{above $25$}) severity 
+levels. In the bar, there are marks that pinpoint the compliance 
+thresholds for each severity level. If the green mark is somewhere 
+in the green bar it means it is compliant with a low level of
 severity.
 
-Aiming to analyze the impact of security patches, we use BCH to compute the
-maintainability of two different versions of the project (cf. Fig.~\ref{fig:commit}):
+Aiming to analyze the impact of security patches, we use BCH to compute the maintainability of two different 
+versions of the project (cf. Fig.~\ref{fig:commit}):
 
 \begin{itemize}
-	\item $v_{s-1}$, the version containing the security flaw, i.e., before the
-	patch (\emph{sha-p});
-	\item $v_{s}$, the version free of the security flaw, i.e., after the
-	patch (\emph{sha});
+	\item $v_{s-1}$, the version containing the security flaw, i.e., 
+	before the patch (\emph{sha-p});
+	\item $v_{s}$, the version free of the security flaw, i.e., 
+	after the patch (\emph{sha});
 \end{itemize}
 
-Patches can be performed through one commit, several consecutive commits, or 
-commit(s) interleaved with more programming activities (floss-refactorings). This study considers mainly single-commit 
-patches. Yet, a small percentage of data points had more than one commit 
-involved in the patch. We manually inspected these cases and $23$ 
-floss-refactorings were identified and disregarded since it would not
-be fair to measure the maintainability of these cases where other
-programming activities are involved.
+The same approach is used analyze the impact of regular 
+changes. Security patches can be performed through one commit, 
+several consecutive commits, or commit(s) interleaved with more 
+programming activities (floss-refactorings). This study considers 
+mainly single-commit patches. Yet, a small percentage of data points 
+had more than one commit involved in the patch. We manually 
+inspected these cases and $23$ floss-refactorings were identified 
+and disregarded since it would not be fair to measure the 
+maintainability of these cases where other programming activities 
+are involved.
 
@@
 	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
- 	\caption{Maintainability difference for security commits}
+ 	\caption{Maintainability difference for security commits.}
 	\label{fig:commit}
@@
 
-For projects with large codebases, the results retrieved for \emph{Keep Your 
-Codebase Small} were not viable because they retrieved values above the limit 
-set by BCH ($20$ Person-years). The \emph{Automated Tests} guideline was also 
-not considered since the tool does not contemplate two
-of the most important techniques to security testing: vulnerability scanning
-and penetration testing. Instead, it only contemplates unit 
-testing. Due to BCH limitations, we detected a few data points
-that retrieved incorrect overall calculations. 
-Those data points were disregarded from the study.
-
-The BCH tool does not compute the final score that our study needs to compare
-maintainability amongst different project versions. For this, we follow previous 
-work on measuring the impact of energy-oriented 
-patches~\cite{8919169}. Cruz et al. ($2019$) proposed an 
-equation to capture the distance between the current state of the project and
-the standard thresholds calculated by the BCH based on the insights provided 
-in~\cite{Olivari:2018}. The equation provided in~\cite{8919169} 
-considers that the size of project changes does not affect the maintainability 
-and that the distance to lower severity levels is less penalized than to the
+For projects with large codebases, the results retrieved for 
+\emph{Keep Your Codebase Small} were not viable because they 
+retrieved values above the limit set by BCH ($20$ Person-years). The 
+\emph{Automated Tests} guideline was also not considered since the 
+tool does not contemplate two of the most important techniques to 
+security testing: vulnerability scanning and penetration testing. 
+Instead, it only contemplates unit testing. Due to BCH limitations, 
+we detected a few data points that retrieved incorrect overall 
+calculations. Those data points were disregarded from the study.
+
+The BCH tool does not compute the final score that our study needs 
+to compare maintainability amongst different project versions. For 
+this, we follow previous work on measuring the impact of 
+energy-oriented patches~\cite{8919169}. Cruz et al. ($2019$) 
+proposed an equation to capture the distance between the current 
+state of the project and the standard thresholds calculated by the 
+BCH based on the insights provided in~\cite{Olivari:2018}. The 
+equation provided in~\cite{8919169} considers that the size of 
+project changes does not affect the maintainability and that the 
+distance to lower severity levels is less penalized than to the
 thresholds in high severity levels.
@@
 
-Given the violations for the BCH guidelines, the maintainability score is computed
-$M(v)$ as follows:
+Given the violations for the BCH guidelines, the maintainability 
+score is computed $M(v)$ as follows:
 
@@
 where $G$ is the group of maintainability guidelines from BCH
-(Table~\ref{tab:guidelines}) and $v$ is the version of the software under
-evaluation. $M(v) < 0$ indicates that version $v$ is violating (some of) the
-guidelines, while $M(v) > 0$ indicates that version $v$ is following
-the BCH guidelines. The maintenance for the guideline $g$, $M_g$ for a given version
-of a project is computed as the summation of the compliance with the 
-maintainability guideline for the given severity level (medium, high, and very high).
-\Luis{The compliance for a severity level
-is calculated based on previous work, which calculates the number of lines of code that comply and not comply
-with the guideline at a given severity level~\cite{cruz2019energyoriented}.}
-The compliance for a severity level
-is calculated based on previous work, which calculates the number of lines of code that comply and not comply
-with the guideline at a given severity level~\cite{8919169}. In our analysis, we compute 
-the difference of maintainability between the security commit ($v_{s}$) and its parent 
-commit ($v_{s-1}$), as illustrated in Fig.~\ref{fig:commit}. Thus, we can determine 
-which patches had a positive, negative, or null impact on the project maintainability.
+(Table~\ref{tab:guidelines}) and $v$ is the version of the software 
+under evaluation. $M(v) < 0$ indicates that version $v$ is violating 
+(some of) the guidelines, while $M(v) > 0$ indicates that version 
+$v$ is following the BCH guidelines. The maintenance for the 
+guideline $g$, $M_g$ for a given version of a project is computed as 
+the summation of the compliance with the maintainability guideline 
+for the given severity level (medium, high, and very high).
+The compliance for a severity level is calculated based on previous 
+work, which calculates the number of lines of code that comply and 
+not comply with the guideline at a given severity 
+level~\cite{8919169}. In our analysis, we compute the difference of 
+maintainability between the security commit ($v_{s}$) and its parent 
+commit ($v_{s-1}$), as illustrated in Fig.~\ref{fig:commit}. Thus, 
+we can determine which patches had a positive, negative, or null 
+impact on the project maintainability.
 
@@
 %
-To validate the maintainability differences in different groups of commits
-(e.g., baseline and security commits), we use the Paired Wilcoxon signed-rank
-test with the significance level $\alpha = 0.05$~\cite{10.2307/3001968}. In
-other words, we test the null hypothesis that the maintainability difference
-between pairs of versions $v_{s-1}$, $v_s$ (i.e., before and after a
-security-commit) come from the same distribution. Nevertheless, this test 
+To validate the maintainability differences in different groups of 
+commits (e.g., baseline and security commits), we use the Paired 
+Wilcoxon signed-rank test with the significance level $\alpha = 
+0.05$~\cite{10.2307/3001968}. In other words, we test the null 
+hypothesis that the maintainability difference between pairs of 
+versions $v_{s-1}$, $v_s$ (i.e., before and after a security-commit) 
+come from the same distribution. Nevertheless, this test 
 has a limitation: it does not consider the groups of commits with a 
-zero-difference maintainability. In $1959$, Pratt provided an improvement to 
-the test to solve this issue making the test more robust. Thus, we use a 
-version of the Wilcoxon test that incorporates the cases where 
-maintainability is different from zero~\cite{10.2307/2282543}.
-The Wilcoxon test requires a distribution size of at least $20$ instances.
-To understand the effect-size, as advocated by the Common-language effect 
-sizes\cite{graw:1992}, we compute the mean difference, the median of the 
-difference, and the percentage of cases that reduce maintainability.
+zero-difference maintainability. In $1959$, Pratt provided an 
+improvement to the test to solve this issue making the test more 
+robust. Thus, we use a version of the Wilcoxon test that 
+incorporates the cases where maintainability is different from 
+zero~\cite{10.2307/2282543}. The Wilcoxon test requires a 
+distribution size of at least $20$ instances. To understand the 
+effect-size, as advocated by the Common-language effect 
+sizes\cite{graw:1992}, we compute the mean difference, the median of 
+the difference, and the percentage of cases that reduce maintainability.
 %
@@
   fully secure.
-  \item [!] \textit{I am not sure how the computer science curricula got mentioned 
+  \item [!] Implications: \textit{I am not sure how the computer science curricula got mentioned 
     from a motivation. I don't think vulnerabilities are been introduced, its 
@@
   all accordingly} - Improve this.
-  \item [!] \textit{wrong assumption: patching commit should also be responsible to preserve the 
+  \item [!] T2v: \textit{wrong assumption: patching commit should also be responsible to preserve the 
   maintainability.... The patching commit is to fix the vulnerability and to put the system is 
@@
   % or preventive maintenance, whereas patching a vulnerability is corrective maintenance.
-  \item Discuss better the fact of not using other important aspects for the baseline such as the
-  size of the changed method and complexity by metrics. Soften the part where it says we should
-  have gone beyond size.
+  % \item Discuss better the fact of not using other important aspects for the baseline such as the
+  % size of the changed method and complexity by metrics. Soften the part where it says we should
+  % have gone beyond size.
   \item \textit{ Figure 4 does not indicate that or support the statement that “patching 
@@
   % So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.
-  \item \textit{In Section 3-C, for computing and comparing the maintainability, only security commits  
-  (and parent commit) are mentioned. Did you apply the same method also the the regular changes that 
-  you use to compare with security commits? Please make this clear, currently this is ignored in the 
-  particular  (sub) section.} - Improve this.
+  % \item \textit{In Section 3-C, for computing and comparing the maintainability, only security commits
+  % (and parent commit) are mentioned. Did you apply the same method also the the regular changes that
+  % you use to compare with security commits? Please make this clear, currently this is ignored in the
+  % particular  (sub) section.} - Improve this.
   \item \textit{For clarification, the authors explained that they performed a manual investigation 

### 2020-09-04 paper/emse20.tex
results

@@
 software. The methodology comprises the following steps, as 
-illustrated in Fig.~\ref{fig:met}.
+illustrated in Figure~\ref{fig:met}.
 %
@@
 
-Fig.~\ref{fig:bchrep} shows an example of the report provided by BCH 
+Figure~\ref{fig:bchrep} shows an example of the report provided by BCH 
 for a project after finishing its evaluation. The example
@@
 which the project does not comply is the one presented in the report 
-(cf. Fig.~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. BCH 
+(cf. Figure~\ref{fig:bchrep}): \emph{Write Simple Units of Code}. BCH 
 analyzes this guideline based on the McCabe 
@@
 Aiming to analyze the impact of security patches, we use BCH to compute the maintainability of two different 
-versions of the project (cf. Fig.~\ref{fig:commit}):
+versions of the project (cf. Figure~\ref{fig:commit}):
 
@@
 maintainability between the security commit ($v_{s}$) and its parent 
-commit ($v_{s-1}$), as illustrated in Fig.~\ref{fig:commit}. Thus, 
+commit ($v_{s-1}$), as illustrated in Figure~\ref{fig:commit}. Thus, 
 we can determine which patches had a positive, negative, or null 
@@
 % security commit ($v_{s}$) and its parent commit ($v_{s-1}$), as illustrated in
-% Fig.~\ref{fig:commit}.
+% Figure~\ref{fig:commit}.
 
@@
 
-This study evaluates a final total of $969$ security and baseline commits 
-from $260$ distinct open-source projects. This section reports/discusses 
-the results for each research question.
-%
-\\\textit{\textbf{RQ1: What is the impact of security patches on the
-maintainability of open-source software?}}
-
-In \emph{RQ1}, we report/discuss the impact of patches on open-source
-software maintainability under four views: guideline, overall score, 
+This study evaluates a final total of $969$ security and baseline 
+commits from $260$ distinct open-source projects. This section 
+reports/discusses the results for each research question.
+%
+
+\textit{\textbf{RQ1: What is the impact of security patches on the
+maintainability of open-source software?}} In \emph{RQ1}, we 
+report and discuss the impact of patches on open-source software 
+maintainability under four groups: guideline, overall score, 
 severity and programming language.
@@
 \textbf{Guideline/Metric:} Each patch performs a set of changes
-on the source code of the software. These changes may have a different
-impact on the guidelines/metrics used to measure software maintainability. 
-Fig.~\ref{fig:guidelines} shows the impact of security patches 
-on each guideline individually and the average impact on all guidelines 
-together ($M(v)$). Under each guideline, it is stated the 
-metric used to the calculations. Table~\ref{tab:guidelines} describes 
-in more detail the metrics behind the guidelines. 
-For each type of guideline, a boxplot 
-and a swarmplot are presented to show the variability and dispersion of
-the results alongside the number of absolute
-and relative cases of each impact. Next to each
-type of guideline, it is presented the mean ($\overline{x}$) and median (M)
-of the maintainability difference and the p-value resulting from the
-Paired Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
-average impact of all guidelines.
+on the source code of the software. These changes may have a 
+different impact on the guidelines/metrics used to measure software 
+maintainability. Figure~\ref{fig:guidelines} shows the impact of 
+security patches on each guideline individually and the average 
+impact on all guidelines together ($M(v)$). Under each guideline, it 
+is stated the metric used to the calculations. For instance, for the 
+\emph{Write Short Units of Code} guideline, the metric used is 
+\emph{Unit Size}. Table~\ref{tab:guidelines} describes in more 
+detail the metrics behind the guidelines. For each type of 
+guideline, a swarmplot is presented to show the variability and 
+dispersion of the results alongside the number of absolute and 
+relative cases of each impact. Next to each type of guideline, it is 
+presented the mean ($\overline{x}$) and median (M) of the 
+maintainability difference and the p-value resulting from the Paired 
+Wilcoxon signed-rank test. $M(v)$ is not a guideline but rather the 
+average impact of all guidelines. Each point of the plot represents 
+the impact of a security patch on software maintainbility. Red, 
+means the impact was negative, i.e., the patch harmed 
+maintainability. Yellow, means the patch did not have any kind of 
+impact on maintainability. Green, means the impact was positive, i.e., the patch improver software maintainability.
 
@@
 
-  	\caption{Impact of the security patches per guideline. For each guideline, $3$ types of impact 
-	in the software maintainability are presented: negative, maintainability decreases (red); none, 
-	maintainability remains the same (yellow); and, positive, maintainability increases (green).}
+  	\caption{Impact of the security patches per guideline. For each 
+	guideline, $3$ types of impact in the software maintainability 
+	are presented: negative, maintainability decreases (red); none, 
+	maintainability remains the same (yellow); and, positive, 
+	maintainability increases (green). For instance, in the 
+	\emph{Write Short Units of Code} guideline, $38.29\%$ of 
+	security patches harmed software maitainabilitty; $23.53\%$ of 
+	security patches had no impact on maintainability; and, 
+	$38.18\%$ of security patches actually improved software 
+	maintainability.}
  	\label{fig:guidelines}	
@@
  \begin{figure}[htp]
-  	\centering 	
- 	\includegraphics[width=0.7\textwidth]{figures/main_per_severity.pdf}
+  	\centering 	 	\includegraphics[width=0.7\textwidth]{figures/main_per_severity.pdf}
   	\caption{Maintainability difference by vulnerability severity.}
@@
 
-Regarding the impact of security patches for each guideline, we observe
-that overall patches have more cases where the impact is positive. However, 
-we also see that patching vulnerabilities has a very
-significant number of negative cases for all guidelines---between $10\%$ and $40\%$. 
-\emph{Write Short Units of Code} ($38.3\%$), \emph{Write Simple Units of Code} 
-($37.9\%$), and \emph{Separate Concerns in Modules} ($33.8\%$) seem to be the 
-most negatively affected guidelines. This may imply that developers when patching 
-vulnerabilities have a hard time to design/implement solutions that continue to 
-respect the limit bounds of branch points and function/module sizes that are 
-recommended by coding practices. Still, on 
-respecting bound limits, developers also seem to not consider the limit 
-of $4$ parameters per function for the \emph{Keep Unit Interfaces Small} 
-guideline required by BCH, in $24.8\%$ of the cases. This guideline is usually
-violated when the patch requires to input new information to a function/class 
-and developers struggle to use the \emph{Introduce Parameter Object} patch 
-pattern. Results do not provide statistical significance to the \emph{Separate 
-Concerns in Modules} guideline, i.e., results should be read carefully.
-
-Software architecture is also affected while patching vulnerabilities.
-Both \emph{Couple Architecture Component Loosely} and \emph{Keep
-Architecture Components Balanced} guidelines suffer a negative impact of 
-$26.9\%$ and $11.0\%$, respectively. Component independence and balance
-are important to make it easier to find the source code that developers
-want to patch/improve and to understand how the high-level components
-interact with others. However, results may imply that developers
-forget to use techniques such as encapsulation to hide implementation
-details and make the system more modular.
-
-The \emph{Write Code Once} guideline results show that duplicated code 
-increased in $171$ ($17.7\%$) of the patches. Software systems typically have 
-$9\%$-$17\%$ of cloned code~\cite{5773403}. Previous work showed a 
-correlation between code smells and code duplication~\cite{7476787} 
-which may also be reflected in the \emph{Write Clean Code} guideline results. 
-BCH reported new code smells for $145$ ($15.0\%$) patches which according 
-to previous work may harm software security since some of these new code
-smells may be new vulnerabilities~\cite{8819456}.
-Thus, while developers 
+Regarding the impact of security patches per guideline, we 
+observe that overall patches have more cases where the impact is 
+positive---where maintainability increases. However, we also 
+see that patching vulnerabilities has a very significant number of 
+negative cases per guideline---between $10\%$ and $40\%$. 
+\emph{Write Short Units of Code} ($38.3\%$), \emph{Write Simple 
+Units of Code} ($37.9\%$), and \emph{Separate Concerns in Modules} 
+($33.8\%$) seem to be the most negatively affected guidelines. This 
+may imply that developers when patching vulnerabilities have a hard 
+time to design/implement solutions that continue to respect the 
+limit bounds of branch points and function/module sizes that are 
+recommended by coding practices. Still on respecting bound limits, 
+developers also seem to not consider the limit of $4$ parameters per 
+function for the \emph{Keep Unit Interfaces Small} guideline 
+required by BCH, in $24.8\%$ of the cases. This guideline is usually
+violated when the patch requires to input new information to a 
+function/class and developers struggle to use the \emph{Introduce 
+Parameter Object} patch pattern. Results do not provide statistical 
+significance to the \emph{Separate Concerns in Modules} guideline, 
+i.e., results should be read carefully.
+
+Software architecture is also affected while patching 
+vulnerabilities. Both \emph{Couple Architecture Component Loosely} 
+and \emph{Keep Architecture Components Balanced} guidelines suffer a 
+negative impact of $26.9\%$ and $11.0\%$, respectively. Component 
+independence and balance are important to make it easier to find the 
+source code that developers want to patch/improve and to understand 
+how the high-level components interact with others. However, results 
+may imply that developers forget to use techniques such as 
+encapsulation to hide implementation details and make the system 
+more modular.
+
+The \emph{Write Code Once} guideline results show that duplicated 
+code increased in $171$ ($17.7\%$) of the patches. Software systems 
+typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. Previous 
+work showed a correlation between code smells and code
+duplication~\cite{7476787} which may also be reflected in the 
+\emph{Write Clean Code} guideline results. BCH reported new code 
+smells for $145$ ($15.0\%$) patches which according to previous work 
+may harm software security since some of these new code smells may 
+be new vulnerabilities~\cite{8819456}. Thus, while developers 
 tend to ignore patch techniques such as the \emph{Replace Method 
-with Method Object} technique, they may also be introducing new software 
-vulnerabilities and, consequently, harm the market value and economy of companies
-~\cite{4267025}.
+with Method Object} technique, they may also be introducing new 
+software vulnerabilities and, consequently, harm the market value 
+and economy of companies~\cite{4267025}.
 
@@
 Although overall patching vulnerabilities has a less negative impact
-on software maintainability guidelines, this is not reflected in the average 
-impact of all guidelines ($M(v)$) as we can see in Fig.~\ref{fig:guidelines}.
-The $M(v)$ plots show that $406$ ($41.9\%$) cases have negative impact
-on software maintainability. While $188$ ($19.4\%$) cases have no impact, $375$ 
-($38.7\%$) have a negative impact on software maintainability. The larger number of 
-negative cases may be explained by guidelines with a higher concentration of negative 
-cases with higher amplitudes, such as \emph{Write Short Units of Code}, \emph{Write 
-Simple Units of Code} and \emph{Separate Concerns in Modules}.
-
-\textbf{Severity:} Some of the vulnerabilities are identified with \emph{Common 
-Vulnerabilities and Exposure} (CVE) entries. We leveraged the \emph{National 
-Vulnerability Database} (NVD) website to collect their severity levels. In total, 
-we retrieved severity scores for $536$ vulnerabilities: $112$ \emph{High}, 
-$395$ \emph{Medium} and $29$ \emph{Low}. Fig.~\ref{fig:severity} 
+on software maintainability guidelines, this is not reflected in the 
+average impact of all guidelines ($M(v)$) as we can see in 
+Figure~\ref{fig:guidelines}. Remember that each point of the plot 
+represents the impact of a security patch on software 
+maintainbility. Red, means the impact was negative, i.e., the patch 
+harmed maintainability. Yellow, means the patch did not have any 
+kind of impact on maintainability. Green, means the impact was 
+positive, i.e., the patch improver software maintainability.
+The $M(v)$ plot show that $406$ ($41.9\%$) cases have a negative 
+impact on software maintainability. While $188$ ($19.4\%$) cases 
+have no impact at all and $375$ ($38.7\%$) have a positive impact on 
+software maintainability. The larger number of negative cases may be 
+explained by guidelines with higher concentrations of negative 
+cases with higher amplitudes, such as \emph{Write Short Units of 
+Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns 
+in Modules}---more red points on the left, being $0$ the middle.
+
+\textbf{Severity:} Some of the vulnerabilities are identified with 
+\emph{Common Vulnerabilities and Exposure} (CVE) entries. We 
+leveraged the \emph{National Vulnerability Database} (NVD) website 
+to collect their severity levels. In total, we retrieved severity 
+scores for $536$ vulnerabilities: $112$ \emph{High}, $395$ 
+\emph{Medium} and $29$ \emph{Low}. Figure~\ref{fig:severity} 
 presents the impact of security patches per severity level on the 
 maintainability of open-source software. We observe that patches for 
-\emph{High} ($50.0\%$) and \emph{Medium} ($43.8\%$) severity vulnerabilities 
-hinder more the maintainability of software than \emph{Low} ($27.6\%$) 
-severity vulnerabilities. Again, patches have a considerable 
-negative impact on software maintainability---between $20\%$ 
-and $50\%$. Statistical significance was retrieved only for \emph{Low}
-severity vulnerabilities, i.e., \emph{Low} severity vulnerabilities
-may have a less negative impact on software maintainability than the 
+\emph{High} ($50.0\%$) and \emph{Medium} ($43.8\%$) severity 
+vulnerabilities hinder more the maintainability of software than 
+\emph{Low} ($27.6\%$) severity vulnerabilities. Again, patches have 
+a considerable negative impact on software maintainability---between 
+$20\%$ and $50\%$. Statistical significance was retrieved only for 
+\emph{Low} severity vulnerabilities, i.e., \emph{Low} severity 
+vulnerabilities may have a less negative impact on software maintainability than the 
 other severity levels. However, results should not be disregarded 
@@
 \textbf{Programming Language:} The impact on software maintainability per programming 
-language was also analyzed (Fig.~\ref{fig:lang_main}). We restrict 
+language was also analyzed (Figure~\ref{fig:lang_main}). We restrict 
 this analysis to programming languages with at least $20$ data points, as this 
@@
 impact on maintainability ($50.9\%$ and $46.5\%$, respectively). But overall 
-languages have a considerable amount of cases that negatively impact  
-maintainability---between $35\%$ to $50\%$---which 
+languages have a considerable amount of cases that negatively impact maintainability---between $35\%$ to $50\%$---which 
 confirms the need for better/more secure programming languages. 
@@
 code~\cite{Ray:2017:LSP:3144574.3126905,2019arXiv190110220B}. However,
-Fig.~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
+Figure~\ref{fig:lang_main} shows that only \emph{C/C++} and \emph{Ruby} have 
 a significant negative impact approximate to $50\%$ on 
@@
 
-\textbf{Conclusion:} Results show that developers may have a hard time following the 
-guidelines and, consequently, hinder software maintainability while patching; and, 
-that different levels of attention should be paid to each guideline. For instance,
+\textbf{Summary:} Results show that developers may have a 
+hard time following the guidelines and, consequently, hinder 
+software maintainability while patching vulnerabilities; and, that different levels of 
+attention should be paid to each guideline. For instance,
 \emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
@@
 for \emph{Separate Concerns in Modules}. 
-The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Fig.~\ref{fig:guidelines}
+The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
 is $0.044$. Since the p-value is 
-below the significance level of $0.05$, we argue that \emph{security patches 
+below the significance level of $0.05$, we argue that \textbf{security patches 
 may have a negative impact on the maintainability of open-source software}; and, 
@@
 %
-\\\textit{\textbf{RQ2: Which weaknesses are more likely to
+
+\textit{\textbf{RQ2: Which weaknesses are more likely to
 affect open-source software maintainability?}}
@@
 and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section~\ref{sec:motivation}).
-Fig.~\ref{fig:pat} shows three different charts: \emph{(\ref{fig:pat}-a)}, shows
+Figure~\ref{fig:pat} shows three different charts: \emph{(\ref{fig:pat}-a)}, shows
 the impact of all patches grouped using the first-level weaknesses from
@@
 
-In Fig.~\ref{fig:pat}-\emph{a}, there is no clear evidence of the impact on 
+In Figure~\ref{fig:pat}-\emph{a}, there is no clear evidence of the impact on 
 maintainability per weakness. Yet, it is important to note that
@@
 ones. Thus, we present an analysis of their sub-weaknesses on 
-Fig.~\ref{fig:pat}-\emph{b} and Fig.~\ref{fig:pat}-\emph{c}, respectively. 
+Figure~\ref{fig:pat}-\emph{b} and Figure~\ref{fig:pat}-\emph{c}, respectively. 
 Results shows that patching vulnerabilities may hinder 
@@
 
-\emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
+\textbf{Summary:} \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
 (CWE-200)}, \emph{Missing Release of Memory after Effective 
-Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the most likely 
-weaknesses to affect software maintainability. No statistical evidence was retrieved. 
+Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the \textbf{most likely 
+weaknesses to affect software maintainability}. No statistical evidence was retrieved. 
 However, results should not be disregarded because they hint on how 
@@
 
-The impact of security and baseline changes on software maintainability is presented in Fig.~\ref{fig:secvsreg}.  
+The impact of security and baseline changes on software maintainability is presented in Figure~\ref{fig:secvsreg}.  
 We have seen, previously, a deterioration on software maintainability 
@@
 However, we manually inspected a considerable amount of cases from
-the distribution of the regular changes where this happens and found that 
-identifying regular commits as similar in size as the security-related 
-commit is limiting the type of regular commits (e.g, input 
+the distribution of the regular changes where maintainability 
+increases and found that identifying regular commits as similar in 
+size as the security-related commit is limiting the type of regular 
+commits being randomly chosen (e.g, input 
 patches, variables or functions, type conversion, etc).  
-We strongly believe that this phenomenon lead
+We presume that this phenomenon lead
 to the significant number of cases where there is no impact
@@
 
-Security-related commits are expected to harm software maintainability. 
+\textbf{Summary:} Security-related commits are expected to harm software maintainability. 
 In contrast, one cannot say the same about
-regular changes, in which no significant differences are observed. Thus, we
+regular changes, in which \textbf{no significant differences are observed}. Thus, we
 urge the importance of adopting maintainability practices while applying
@@
 	Design debt of one guideline can lead
-	to severe impacts on the software quality~\cite{10.1145/1985362.1985366}. 
-	Thus, tools like Better Code Hub and static analysis (e.g., SonarQube, Codacy,
-	ESLint, Infer, and more) should be leveraged to 
-	assess the risk of patching vulnerabilities---alongside pull
-  requests/code reviews. Static code analysis
-	may be daunting due to the number of rules and different effects
+	to severe impacts on the software 
+	quality~\cite{10.1145/1985362.1985366}. 
+	Thus, tools like Better Code Hub and static analysis (e.g., 
+	SonarQube, Codacy, ESLint, Infer, and more) should be leveraged 
+	to assess the risk of patching vulnerabilities---alongside pull
+	requests/code reviews. Static code analysis may be daunting due 
+	to the number of rules and different effects
 	on maintainability. BCH claims to use the top-$10$ of the guidelines
@@
 not explicit in commits messages. We assume that patches were performed 
-using a single-commit or several sequentially. The view that a 
-developer may perform a patch quickly and later proceed to the refactor is not considered.
-Baseline commits are retrieved randomly from the same project of the security patch.
+using a single-commit or several sequentially. The 
+perspective that a 
+developer may perform a patch quickly and later 
+proceed to the refactor is not considered. We assume 
+that all patches were only performed once. Depending 
+on the impact of the vulnerability in the system, some 
+vulnerabilities may have more urgency to be fixed than 
+others. Baseline commits are retrieved randomly from 
+the same project of the security patch.
 This approach softens the differences
@@
  %    \end{itemize}
-  \item [!] \textit{Fig 4, although it looks impressive, I am not sure how to read the figure} - what 
-  should we do @rui?
-    \begin{itemize}
-      \item Carefully explain it in the caption?
-    \end{itemize}
-  \item [!] \textit{Fig 5, 6 and 7 are not the best ways to portray your findings. For instance, I am 
-  still not sure how the positive, none and negative metrics are calculated and normalized} - what
-  is the best way then? 
-    \begin{itemize}
-      \item Strange comment...
-    \end{itemize}
-  \item \textit{Authors carefully highlight the key messages of each research question and answer
-  all accordingly} - Improve this.
-  \item [!] T2v: \textit{wrong assumption: patching commit should also be responsible to preserve the 
-  maintainability.... The patching commit is to fix the vulnerability and to put the system is 
-  an error-free, executable state so that the fix can be tested} -> I agree with this for vulnerabilities
-  that are the result of errors. For example, a bad expection handling, a memory leak that led to 
-  a denial-of-service attack, NPE, etc. However, there are many other vulnerabilities whose
-  exploits do not have interference on systems execution. So, what to do here? Should we divide
-  the study by vulnerabilities and their implications on systems? Should we divide it by severity levels?
-    \begin{itemize}
-      \item We need to discuss this... I think this related to the next point. 
-    \end{itemize}  
+  % \item [!] \textit{Fig 4, although it looks impressive, I am not sure how to read the figure} - what
+  % should we do @rui?
+  %   \begin{itemize}
+  %     \item Carefully explain it in the caption?
+  %   \end{itemize}
+  % \item [!] \textit{Fig 5, 6 and 7 are not the best ways to portray your findings. For instance, I am
+  % still not sure how the positive, none and negative metrics are calculated and normalized} - what
+  % is the best way then?
+  %   \begin{itemize}
+  %     \item Strange comment...
+  %   \end{itemize}
+  % \item \textit{Authors carefully highlight the key messages of each research question and answer
+  % all accordingly} - Improve this.
+  % \item [!] T2v: \textit{wrong assumption: patching commit should also be responsible to preserve the
+  % maintainability.... The patching commit is to fix the vulnerability and to put the system is
+  % an error-free, executable state so that the fix can be tested} -> I agree with this for vulnerabilities
+  % that are the result of errors. For example, a bad expection handling, a memory leak that led to
+  % a denial-of-service attack, NPE, etc. However, there are many other vulnerabilities whose
+  % exploits do not have interference on systems execution. So, what to do here? Should we divide
+  % the study by vulnerabilities and their implications on systems? Should we divide it by severity levels?
+  %   \begin{itemize}
+  %     \item We need to discuss this... I think this related to the next point.
+  %   \end{itemize}
   % \item \textit{authors write "...important to write maintainable code to support the production
@@
   % have gone beyond size.
-  \item \textit{ Figure 4 does not indicate that or support the statement that “patching 
-  vulnerabilities has a very significant number of negative cases for all guidelines”. Besides a 
-  corresponding statistical test that supports this statement, even the swarm and box plotz itself 
-  clearly tell a different story. So, authors may rethink this statement or better explain how they 
-  come to this conclusion.} -> We are overselling here. Improve it.
+  % \item \textit{ Figure 4 does not indicate that or support the statement that “patching
+  % vulnerabilities has a very significant number of negative cases for all guidelines”. Besides a
+  % corresponding statistical test that supports this statement, even the swarm and box plotz itself
+  % clearly tell a different story. So, authors may rethink this statement or better explain how they
+  % come to this conclusion.} -> We are overselling here. Improve it.
   % \item \textit{Authors state that they assume for the Pontas dataset that the last commit related
@@
   % particular  (sub) section.} - Improve this.
-  \item \textit{For clarification, the authors explained that they performed a manual investigation 
-  for RQ3. It would be beneficial for more details.} -> Clarify this.
-  \item \textit{In terms of presentation, the results were a bit difficult to follow at times as they 
-  did not contain contextualization. For example, “negative cases” can seem a bit ambiguous when 
-  referencing the relationship between the impact with guidelines.} -> Improve this.
+  % \item \textit{For clarification, the authors explained that they performed a manual investigation
+  % for RQ3. It would be beneficial for more details.} -> Clarify this.
+  % \item \textit{In terms of presentation, the results were a bit difficult to follow at times as they
+  % did not contain contextualization. For example, “negative cases” can seem a bit ambiguous when
+  % referencing the relationship between the impact with guidelines.} -> Improve this.
 
@@
 
-\section{Plan B}
-
-Baseado nos reviewers, eu acho que para seguir as orientaçoes deles de refletir e fortalecer 
-a nova assumption e as implicaçoes, teriamos de acrescentar ao Plano A, os proximos pontos
-antes de submeter à ESME.
-
-\begin{itemize}
-\item \textbf{Step 1:} Pegar nos casos em que o impacto na maintainability foi negativa e ver se foram modificados 
-depois e se sim, se houve melhoria da maintainability. -> isto para então resolver a questao da 
-assumption errada.
-\item \textbf{Step 2:} Se não existir nenhuma alteração, entao talvez entrar em contacto com os developers
-e perguntar se quando fizeram o patch tiveram em consideração maintainability do projecto. E se não porquê.
-\item Problemas de baselines e testes acho que nao vale a pena pegar porque vai depender
-do reviewer (eu acho).
-\end{itemize}
-
-Qual é o plano que acha melhor? Como prosseguir?
-
-A discutir, mas acho que por agora era tentar fazer o 1.
-
-\section{Later}
-
-\begin{itemize}
-	\item \textbf{Assumptions:}\textit{authors measure the maintainability directly after the patch. inestad it should look 
-  for multiple follow up commits and how maintainability evolves there, as it is likely that certain
-  issues are addresses AFTER the patching commit} -> I think we should see if later commits touch on
-  the patch and measure their maintainability again. Was a less maintainable fix pushed first and then
-  refactored, or does the code stay less maintainable?
-  \item \textbf{To back up implications:} \textit{In terms of the push for changes to curricula, I would agree that it is important 
-  to always stress writing high quality code. However, the support by the study is not strong. Given a 
-  majority of patches did not decrease maintainability, it shows that there is a reasonable focus on maintainability. 
-  Similarly, it is important to understand the circumstances when it decreased. As previously mentioned, organizations 
-  like would prioritize getting the system patched faster (especially in the case of a severe vulnerability) over 
-  its maintainability.} -> Work with practioners to understand if they were taking into consideration maintainability?
-  \item \textbf{Baseline:}\textit{Authors just search for commits of a similar size. There are other important aspects
-  that are entirely neglected such as size of the changed method or how complex it it (by means of 
-  metrics)} -> Improve baseline.
-  \item \textbf{Baseline:}\textit{Similarly, the comparison to “regular” commits really compares security patches to 
-  any other type of change. It would have been more meaningful to compared them to specific types of 
-  commits, like bug fixes (e.g., are security fixes less maintainable than bug fixes?).}
-  \item \textbf{Tests:}\textit{or the chosen statistical test, the hypothesis may be not well chosen as it just checks for 
-  the same distribution of pairs of versions. However, this means if a similar number of pairs have 
-  a negative and a positive impact, the hypothesis holds however, if this is not the case, the statistical 
-  test does not state in which direction the hypothesis is rejected, that is, what eventually is the 
-  alternative hypothesis. But this is necessary to get insights whether maintainability is hampered or not. 
-  Authors may rethink the choice of the hypothesis or even the test itself.} -> What about this?
-  \item \textbf{Thesis idea (PL):} It is important 
-  to understand the correlation between the severity of the vulnerability and the programming language as well 
-  as the maintainability impact. It is not clear how software written a more security-focused language would 
-  inherently be more maintainable (it assumes the flaw is the language and not the person).  
-\end{itemize}
+% \section{Plan B}
+%
+% Baseado nos reviewers, eu acho que para seguir as orientaçoes deles de refletir e fortalecer
+% a nova assumption e as implicaçoes, teriamos de acrescentar ao Plano A, os proximos pontos
+% antes de submeter à ESME.
+%
+% \begin{itemize}
+% \item \textbf{Step 1:} Pegar nos casos em que o impacto na maintainability foi negativa e ver se foram modificados
+% depois e se sim, se houve melhoria da maintainability. -> isto para então resolver a questao da
+% assumption errada.
+% \item \textbf{Step 2:} Se não existir nenhuma alteração, entao talvez entrar em contacto com os developers
+% e perguntar se quando fizeram o patch tiveram em consideração maintainability do projecto. E se não porquê.
+% \item Problemas de baselines e testes acho que nao vale a pena pegar porque vai depender
+% do reviewer (eu acho).
+% \end{itemize}
+%
+% Qual é o plano que acha melhor? Como prosseguir?
+%
+% A discutir, mas acho que por agora era tentar fazer o 1.
+%
+% \section{Later}
+%
+% \begin{itemize}
+% 	\item \textbf{Assumptions:}\textit{authors measure the maintainability directly after the patch. inestad it should look
+%   for multiple follow up commits and how maintainability evolves there, as it is likely that certain
+%   issues are addresses AFTER the patching commit} -> I think we should see if later commits touch on
+%   the patch and measure their maintainability again. Was a less maintainable fix pushed first and then
+%   refactored, or does the code stay less maintainable?
+%   \item \textbf{To back up implications:} \textit{In terms of the push for changes to curricula, I would agree that it is important
+%   to always stress writing high quality code. However, the support by the study is not strong. Given a
+%   majority of patches did not decrease maintainability, it shows that there is a reasonable focus on maintainability.
+%   Similarly, it is important to understand the circumstances when it decreased. As previously mentioned, organizations
+%   like would prioritize getting the system patched faster (especially in the case of a severe vulnerability) over
+%   its maintainability.} -> Work with practioners to understand if they were taking into consideration maintainability?
+%   \item \textbf{Baseline:}\textit{Authors just search for commits of a similar size. There are other important aspects
+%   that are entirely neglected such as size of the changed method or how complex it it (by means of
+%   metrics)} -> Improve baseline.
+%   \item \textbf{Baseline:}\textit{Similarly, the comparison to “regular” commits really compares security patches to
+%   any other type of change. It would have been more meaningful to compared them to specific types of
+%   commits, like bug fixes (e.g., are security fixes less maintainable than bug fixes?).}
+%   \item \textbf{Tests:}\textit{or the chosen statistical test, the hypothesis may be not well chosen as it just checks for
+%   the same distribution of pairs of versions. However, this means if a similar number of pairs have
+%   a negative and a positive impact, the hypothesis holds however, if this is not the case, the statistical
+%   test does not state in which direction the hypothesis is rejected, that is, what eventually is the
+%   alternative hypothesis. But this is necessary to get insights whether maintainability is hampered or not.
+%   Authors may rethink the choice of the hypothesis or even the test itself.} -> What about this?
+%   \item \textbf{Thesis idea (PL):} It is important
+%   to understand the correlation between the severity of the vulnerability and the programming language as well
+%   as the maintainability impact. It is not clear how software written a more security-focused language would
+%   inherently be more maintainable (it assumes the flaw is the language and not the person).
+% \end{itemize}
 

### 2020-09-04 paper/emse20.tex
i think it's ready

@@
 \\\textit{\textbf{Tools for Patch Risk Assessment:}}
-	Design debt of one guideline can lead
-	to severe impacts on the software 
-	quality~\cite{10.1145/1985362.1985366}. 
-	Thus, tools like Better Code Hub and static analysis (e.g., 
-	SonarQube, Codacy, ESLint, Infer, and more) should be leveraged 
+	Design debt of one guideline can lead to severe impacts on the 
+	software quality~\cite{10.1145/1985362.1985366}. Some software 
+	producers consider security as a first class citizen while others 
+	do not.
+As mentioned in previous work, security it is very important and should bee considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}.
+However, the lack of experts or awareness to the ease of producing vulnerable software leads companies to ship low quality software.
+Providing automated tools to developers assess the risk of their patches it is important to help companies shipping software of higuer quality.
+Tools like Better Code Hub and static analysis (e.g.,
+	SonarQube, Codacy, ESLint, Infer, and more) can be leveraged 
 	to assess the risk of patching vulnerabilities---alongside pull
@@
 	on maintainability. BCH claims to use the top-$10$ of the guidelines
-	with the highest effect on maintainability. Yet, static analysis 
+	with the highest effects on maintainability. Yet, static analysis 
 	can be used to complement the BCH analysis by introducing
-	the capability of vulnerability detection and support the prediction/prevention of the risk of hindering maintainability while patching.
+	the capability of vulnerability detection and support the prediction/prevention of the risk of a patch of hindering maintainability.
 %
@@
 	Computer Science curricula is not yet prepared to properly educate students for maintainable security. 
-	    Students should be exposed to this problem to gain experience with it. Curricula should focus on producing 
-	    secure and maintainable code and alert to the trade-off between both.
+	    Students should be exposed to this problem to gain experience with it. Curricula should focus on the production of 
+	    secure and maintainable code and alert to the trade-off between both. These matters should be discussed and presented to students in software engineering courses.
 	Universities should also encourage students to use code quality analysis tools such as BCH or similar ones. These tools have a great potential to make
 	students aware of maintainability issues and beginner mistakes (e.g.,
-	coding practices violation). Thus, students and future developers can 
-	put more work into improving their code maintainability and avoiding the risk
-	of introducing new vulnerabilities while patching older ones.
-%	
-\\\textit{\textbf{Secure Programming Languages Wanted:}} Programming Languages 
-	should provide new design patterns to easily implement security weaknesses 
+	coding practices violation).
+%	
+\\\textit{\textbf{More Secure Programming Languages Wanted:}} Programming Languages 
+	should provide new design patterns to easily patch security weaknesses 
 	without endangering maintainability.
-	Ultimately, new programming languages, both secure and maintainable by design,
-	may be designed to help developers be highly productive when writing and
-	maintaining secure applications. One example is the need for designing new
+	Ultimately, new programming languages, both secure and maintainable by design, such as Wyvern~\cite{10.1145/2489828.2489830,kurilova2014wyvern}, 
+	should be designed to help developers be less vulnerability prone when writing and maintaining secure applications. One example is the need for designing new
 	authorization mechanisms since these are one of the most complex security
@@
 
-\section{Plan A}
-
-Please, pay more attention to the cases with !.
-
-\begin{itemize}
+% \section{Plan A}
+%
+% Please, pay more attention to the cases with !.
+%
+% \begin{itemize}
 	% \item How this is a problem or has an impact to the software project?
 	%   What is the message to developers? (clarify)
-	\item Implications: Discuss that depending on the project, maybe security 
-  is a first class citizen and should be fixed above all. Don't 
-  just state that patch risk assessment should take maintainability into account. 
-  Saying that security should be a priority requires maybe a user study
-  as evidence (check if the literature available is not enough).
-  \item Implications: Review the secure programming languages implication. Check 
-  if the idea of fully secure exists. The idea should be more secure, not
-  fully secure.
-  \item [!] Implications: \textit{I am not sure how the computer science curricula got mentioned 
-    from a motivation. I don't think vulnerabilities are been introduced, its 
-    more that they get exposed. In this case, the implication does not make 
-    much sense from a practical point of view.} -> Why not? Shouldn't 
-    maintainable security be introduced in the computer science curricula since
-    we have evidence that developers might not following good practices while
-    patching? If they learn it in college and are aware of it in their jobs
-    won't they be better prepared?
+	% \item Implications: Discuss that depending on the project, maybe security
+	%   is a first class citizen and should be fixed above all. Don't
+	%   just state that patch risk assessment should take maintainability into account.
+	%   Saying that security should be a priority requires maybe a user study
+	%   as evidence (check if the literature available is not enough).
+	%   \item Implications: Review the secure programming languages implication. Check
+	%   if the idea of fully secure exists. The idea should be more secure, not
+	%   fully secure.
+  % \item [!] Implications: \textit{I am not sure how the computer science curricula got mentioned
+  %   from a motivation. I don't think vulnerabilities are been introduced, its
+  %   more that they get exposed. In this case, the implication does not make
+  %   much sense from a practical point of view.} -> Why not? Shouldn't
+  %   maintainable security be introduced in the computer science curricula since
+  %   we have evidence that developers might not following good practices while
+  %   patching? If they learn it in college and are aware of it in their jobs
+  %   won't they be better prepared?
   % \item [!] \textit{It did take me several readings to understand the paper, especially the approach} -
@@
 
-\end{itemize}
+% \end{itemize}
 

### 2020-09-04 paper/emse20.tex
caption fig  4, improved

@@
 In this section, we discuss the methodology used to measure
-the impact of security patches on the maintainbility of open-source
+the impact of security patches on the maintainability of open-source
 software. The methodology comprises the following steps, as 
@@
 average impact of all guidelines. Each point of the plot represents 
-the impact of a security patch on software maintainbility. Red, 
+the impact of a security patch on software maintainability. Red, 
 means the impact was negative, i.e., the patch harmed 
@@
 
-  	\caption{Impact of the security patches per guideline. For each 
-	guideline, $3$ types of impact in the software maintainability 
-	are presented: negative, maintainability decreases (red); none, 
-	maintainability remains the same (yellow); and, positive, 
-	maintainability increases (green). For instance, in the 
+  	\caption{Impact of the security patches per guideline and overall mean, $M(v)$. 
+	Each point of the plot represents the impact of a security patch on software 
+	maintainability. Red, means the impact was negative, i.e., the patch 
+	harmed maintainability. Yellow, means the patch did not have any 
+	kind of impact on maintainability. Green, means the impact was 
+	positive, i.e., the patch improver software maintainability. For instance, in the 
 	\emph{Write Short Units of Code} guideline, $38.29\%$ of 
@@
 represents the impact of a security patch on software 
-maintainbility. Red, means the impact was negative, i.e., the patch 
+maintainability. Red, means the impact was negative, i.e., the patch 
 harmed maintainability. Yellow, means the patch did not have any 
@@
 
-The impact of security and baseline changes on software maintainability is presented in Figure~\ref{fig:secvsreg}.  
+The impact of security and baseline changes on software maintainability 
+is presented in Figure~\ref{fig:secvsreg}.  
 We have seen, previously, a deterioration on software maintainability 
@@
 
-\textbf{Summary:} Security-related commits are expected to harm software maintainability. 
-In contrast, one cannot say the same about
-regular changes, in which \textbf{no significant differences are observed}. Thus, we
-urge the importance of adopting maintainability practices while applying
-security patches.
+\textbf{Summary:} Security-related commits are expected to harm software 
+maintainability. In contrast, one cannot say the same about
+regular changes, in which \textbf{no significant differences are observed}. 
+Thus, we urge the importance of adopting maintainability practices while 
+applying security patches.
 
@@
 	producers consider security as a first class citizen while others 
-	do not.
-As mentioned in previous work, security it is very important and should bee considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}.
-However, the lack of experts or awareness to the ease of producing vulnerable software leads companies to ship low quality software.
-Providing automated tools to developers assess the risk of their patches it is important to help companies shipping software of higuer quality.
+	do not. As mentioned in previous work, security it is very important 
+	and should bee considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}. 
+	However, the lack of experts or awareness to the ease of producing 
+	vulnerable software leads companies to ship low quality software. 
+	Providing automated tools to developers assess the risk of their 
+	patches it is important to help companies shipping software of higuer quality. 
 	Tools like Better Code Hub and static analysis (e.g., 
@@
 	can be used to complement the BCH analysis by introducing
-	the capability of vulnerability detection and support the prediction/prevention of the risk of a patch of hindering maintainability.
+	the capability of vulnerability detection and support the prediction 
+	and prevention of the risk of a patch of hindering maintainability.
 %
 \\\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
-	Computer Science curricula is not yet prepared to properly educate students for maintainable security. 
-	    Students should be exposed to this problem to gain experience with it. Curricula should focus on the production of 
-	    secure and maintainable code and alert to the trade-off between both. These matters should be discussed and presented to students in software engineering courses.
-	Universities should also encourage students to use code quality analysis tools such as BCH or similar ones. These tools have a great potential to make
-	students aware of maintainability issues and beginner mistakes (e.g.,
-	coding practices violation).
-%	
-\\\textit{\textbf{More Secure Programming Languages Wanted:}} Programming Languages 
-	should provide new design patterns to easily patch security weaknesses 
-	without endangering maintainability.
-	Ultimately, new programming languages, both secure and maintainable by design, such as Wyvern~\cite{10.1145/2489828.2489830,kurilova2014wyvern}, 
-	should be designed to help developers be less vulnerability prone when writing and maintaining secure applications. One example is the need for designing new
-	authorization mechanisms since these are one of the most complex security
-	features to implement.
+	Computer Science curricula is not yet prepared to properly educate 
+	students for maintainable security. Students should be exposed to 
+	this problem to gain experience with it. Curricula should focus on 
+	the production of secure and maintainable code and alert to the 
+	trade-off between both. These matters should be discussed and 
+	presented to students in software engineering courses.
+	Universities should also encourage students to use code quality 
+	analysis tools such as BCH or similar ones. These tools have a 
+	great potential to make students aware of maintainability issues 
+	and beginner mistakes (e.g., coding practices violation).
+%	
+\\\textit{\textbf{More Secure Programming Languages Wanted:}} Programming 
+	Languages should provide new design patterns to easily patch security weaknesses 
+	without endangering maintainability. Ultimately, new programming languages, 
+	both secure and maintainable by design, such as Wyvern~\cite{10.1145/2489828.2489830,kurilova2014wyvern}, 
+	should be designed to help developers be less vulnerability prone when 
+	writing and maintaining secure applications. One example is the need for 
+	designing new authorization mechanisms since these are one of the most 
+	complex security features to implement.
 % 

### 2020-09-04 paper/emse20.tex
small fix

@@
 maintainability. Yellow, means the patch did not have any kind of 
-impact on maintainability. Green, means the impact was positive, i.e., the patch improver software maintainability.
+impact on maintainability. Green, means the impact was positive, i.e., 
+the patch improver software maintainability.
 
@@
 Code}, \emph{Write Simple Units of Code} and \emph{Separate Concerns 
-in Modules}---more red points on the left, being $0$ the middle.
+in Modules}---more red points on the left, being $0$ the reference point.
 
@@
 
-\textbf{Summary:} Security-related commits are expected to harm software 
+\textbf{Summary:} Security-related commits are observed to harm software 
 maintainability. In contrast, one cannot say the same about

### 2020-09-04 paper/emse20.tex
rui affiliation changed

@@
 			Rui Abreu \at
-			INESC ID and IST, University of Lisbon, Lisbon, Portugal\\
+			INESC ID and FEUP, University of Porto, Porto, Portugal\\
 			\email{rui@computer.org}
@@
 %
-\\\textit{\textbf{Tools for Patch Risk Assessment:}}
+
+\textit{\textbf{Tools for Patch Risk Assessment:}}
 	Design debt of one guideline can lead to severe impacts on the 
@@
 %
-\\\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
+
+\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
 	Computer Science curricula is not yet prepared to properly educate 
@@
 %
-\\\textit{\textbf{More Secure Programming Languages Wanted:}} Programming 
+
+\textit{\textbf{More Secure Programming Languages Wanted:}} Programming 
 	Languages should provide new design patterns to easily patch security weaknesses 
@@
 %
-\\\textit{\textbf{Construct Validity:}} The formula was inferred based 
+
+\textit{\textbf{Construct Validity:}} The formula was inferred based 
 on the BCH's reports. The high amount of different projects and backgrounds 
@@
 %
-\\\textit{\textbf{Internal Validity:}} The security patches dataset provided by previous
+
+\textit{\textbf{Internal Validity:}} The security patches dataset provided by previous
 work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
@@
 %
-\\\textit{\textbf{External Validity:}} We use a dataset that only comprises patches of open-source software.
+
+\textit{\textbf{External Validity:}} We use a dataset that only comprises patches of open-source software.
 However, the methodology requires to access data that is not publicly available.
@@
 % \end{itemize}
+\pagebreak
 

### 2020-09-10 paper/emse20.tex
ready for submission

@@
 Better Code Hub’s model of 10 guidelines using a dataset including 
-1335 security-related commits. Results show evidence of a trade-off 
+1300 security-related commits. Results show evidence of a trade-off 
 between security and maintainability, i.e., developers may be 
@@
 
-From a methodological perspective, we leveraged a dataset of $1335$ 
+From a methodological perspective, we leveraged a dataset of $1300$ 
 security patches collected from open-source software. We calculate 
@@
 
-The code changes performed to patch the vulnerability were:
-
-\ref{lst:func1} The \texttt{sk\_OCSP\_RESPID\_pop\_free} function 
-removes any memory allocated to the OCSP response IDs 
-(OCSP$\_$RESPIDs) from a previous handshake to prevent unbounded 
-memory growth - which was not being performed before.
+The code changes performed to patch the CVE-$2016$-$6304$ vulnerability are 
+presented in Listing~\ref{lst:vuln}. The \texttt{sk\_OCSP\_RESPID\_pop\_free} function 
+(\ref{lst:func1}) removes any memory allocated to the OCSP response IDs 
+(\texttt{OCSP\_RESPID}s) from a previous handshake to prevent unbounded 
+memory growth---which was not being performed before.
 After releasing the unbounded memory, the logic condition in 
@@
 for handling the application when no OCSP response IDS are 
-allocated. Now, instead of performing it after the packages 
-processing, it does it previously.
+allocated. After the patch, in the new version of the software, the condition is checked before the package processing 
+instead of after. Thereby, the system avoids the increase of unbounded memory (and, a potential DoS attack).
 
@@
 to choose the easiest path to solve it. To address it, we compute 
-the maintainability of $1335$ commits using the \emph{Better Code Hub} tool.
+the maintainability of $1300$ commits using the \emph{Better Code Hub} tool.
 
@@
 	the fixed version and codebases) from the combined dataset 	
-	containing $1335$ security patches collected from open-source 	
+	containing $1300$ security patches collected from open-source 	
 	software available on GitHub.
@@
 
-We use a combined dataset of security vulnerabilities which is 
-the outcome of mining and manually inspecting a total of $265$ 
-GitHub projects for both datasets. Reis and Abreu 
+We use a combined dataset of $1330$ security patches which is 
+the outcome of mining and manually inspecting a total of $312$ 
+GitHub projects. The combined dataset integrates two different
+works: SecBench~\cite{Reis:2017:IJSSE} and Ponta et al.~\cite{10.1109/MSR.2019.00064}.
+
+Reis and Abreu 
 ($2017$) mined open-source software aiming at the extraction of 
-real---created by developers---samples of security vulnerabilities 
+real---created by developers---patches of security vulnerabilities 
 to test and assess the performance of static analysis 
@@
 the tools~\cite{just2014mutants}. The study yielded a dataset of 
-$711$ test cases for $16$ different security vulnerability types. 
-Some of the vulnerability types are based on the OWASP Top $10$ of 
+$676$ patches for $16$ different security vulnerability types, dubbed as SecBench. 
+The vulnerability types are based on the OWASP Top $10$ of 
 $2013$~\cite{oswap:2013} and OWASP Top $10$ of 
 $2017$~\cite{oswap:2017}. Each test case of the dataset is a 
-triplet: the commit before the patching, the commit responsible
-for the patching, and the snippets of code that differ from one 
-version to another (typically, called \textit{diffs})---where one 
-can easily review the code used to fix the security flaw. 
+triplet: the commit before the patching (\emph{sha-p}), the commit responsible
+for the patching (\emph{sha}), and the snippets of code that differ from one 
+version to another (typically, called \emph{diffs})---where one 
+can easily review the code used to fix the vulnerability. 
 %
@@
 commits before and after the security patching to evaluate if its 
-impact was positive, negative or none. The $1335$ patches in the 
+impact was positive, negative or none. The $1300$ patches in the 
 dataset were run against the BCH toolset to calculate their 
-maintainability reports. However, there are patches discarded from 
-our study due to limitations of BCH: in particular, lack of
-language support and project size. The final dataset used in this 
-paper comprises $969$ security patches. $867$ of these patches are 
-identified with Common Weakness Enumeration IDs (CWE) where $62.1\%$ 
-were scraped from the National Vulnerability Dataset (NVD) and the 
-remaining $37.9\%$ were manually classified following the 
-\emph{Research Concepts} CWE's list.
+maintainability reports. Due to the limitations of BCH, in particular, 
+lack of language support and project size, we had
+to clean the dataset and select the most relevant and compatible projects
+with BCH. The final dataset used in this 
+paper comprises $969$ security patches from $260$ projects. We used the 
+Common Weakness Enumeration (CWE) taxonomy to classify each vulnerability.
+For instance,  the \texttt{Fix CVE-2014-1608: mc\_issue\_attachment\_get SQL injection} (\texttt{00b4c17}\footnote{CVE-$2014$-$1608$ details available at \url{https://github.com/mantisbt/mantisbt/commit/00b4c17088fa56594d85fe46b6c6057bb3421102} (Accessed on \today)}) is a \emph{CWE-89: Improper Neutralization of Special Elements used in an SQL Command ('SQL Injection')}\footnote{CWE-$89$ details available at \url{https://cwe.mitre.org/data/definitions/89.html} (Accessed on \today)} according to the CWE taxonomy.
+We were able to classify a total of $866$ patches using the CWE taxonomy
+where the CWEs for $536$ patches were automatically scraped from the National Vulnerability Dataset (NVD); while for the other $330$ patches, the CWEs were manually chosen by the authors following the \emph{Research Concepts} CWE's list.
 
@@
 The baseline dataset is generated from the security commits dataset. 
-As for security patches, for each regular fix or refatoring, we need 
-a the commit performing the regular change (sha-reg) and version of 
-the software before the change (sha-reg-p). For each security patch, 
+As for the security patches, for each regular fix or refatoring, we also need the commit performing the regular change (\emph{sha-reg}) and version of 
+the software before the change (\emph{sha-reg-p}). For each security patch, 
 a \emph{diff} with an approximate size was fetched to a regular 
@@
 
+\begin{table}[h]
+\centering
+\scriptsize
+	\caption{Guidelines to produce maintainable code.}
+\begin{tabular}{L{2cm}L{4.5cm}L{4cm}}
+
+\toprule
+\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
+\midrule
+
+\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
+ units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of 
+ LOCs within each unit~\cite{criteria:2017} \\\midrule
+
+\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
+it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision 
+points~\cite{1702388,criteria:2017}\\\midrule
+
+\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
+multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant 
+LOCs~\cite{criteria:2017}\\\midrule
+
+\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
+$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} 
+\# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
+
+\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
+loosely coupled databases are easier to oversee and execute & \textbf{Module Coupling:} \# of
+incoming dependencies~\cite{criteria:2017} \\\midrule
+
+\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
+within modules that are exposed to modules in other components & \textbf{Component Independence:} 
+\% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
+
+\textbf{Keep Architecture Components Balanced} & Balancing the number of
+components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} 
+Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
+
+\textbf{Keep your code base Small} & Reduce and avoid the system size because
+small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted 
+to man-month/man-year~\cite{criteria:2017} \\\midrule
+
+\textbf{Automate Tests} & Test your code base because it makes development
+predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit 
+complexity, component independence and volume~\cite{Visser:2016:OREILLY}
+ \\\midrule
+
+\textbf{Write Clean Code} & Avoid producing software with code smells because
+it is more likely to be maintainable in the future & \textbf{Code Smells:} 
+\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
+identifier names) \\
+\bottomrule
+\end{tabular}
+\label{tab:guidelines}
+\end{table}
+
+Figure~\ref{fig:bchrep} shows an example of the report provided by BCH 
+for a project after finishing its evaluation. The example
+refers to the OpenSSL CVE-$2016$-$6304$ vulnerability patch---
+as described by Section~\ref{sec:motivation}. This version of 
+OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write 
+Clean Code}.
+
 \begin{figure}[h]
  	\centering 	
-	\includegraphics[width=0.9\textwidth]{figures/bch_report.png}
+	\includegraphics[width=0.8\textwidth]{figures/bch_report.png}
  	\caption{Maintainability report of OpenSSL's CVE-$2016$-$6304$ 
@@
 \end{figure}
-
-\begin{table}[H]
-\centering
-\scriptsize
-	\caption{Guidelines to produce maintainable code.}
-\begin{tabular}{L{2cm}L{4.5cm}L{4cm}}
-
-\toprule
-\textbf{10 Guidelines} & \textbf{Description} & \textbf{Metric}\\
-\midrule
-
-\textbf{Write Short Units of Code} & Limit code units to $15$ LOCs because smaller
- units are easier to understand, reuse and test them & \textbf{Unit Size:} \% of 
- LOCs within each unit~\cite{criteria:2017} \\\midrule
-
-\textbf{Write Simple Units of Code} & Limit branch points to $4$ per unit because
-it makes units easier to test and modify & \textbf{McCabe Complexity:} \# of decision 
-points~\cite{1702388,criteria:2017}\\\midrule
-
-\textbf{Write Code Once} & Do not copy code because bugs tend to replicate at
-multiple places (inefficient and error-prone) & \textbf{Duplication:} \% of redundant 
-LOCs~\cite{criteria:2017}\\\midrule
-
-\textbf{Keep Unit Interfaces Small} & Limit the number of parameters to at most
-$4$ because it makes units easier to understand and reuse & \textbf{Unit Interfacing:} 
-\# of parameters defined in a signature of a unit~\cite{criteria:2017} \\\midrule
-
-\textbf{Separate Concerns in Modules} & Avoid large modules because changes in
-loosely coupled databases are easier to oversee and execute & \textbf{Module Coupling:} \# of
-incoming dependencies~\cite{criteria:2017} \\\midrule
-
-\textbf{Couple Architecture Components Loosely} & Minimize the amount of code
-within modules that are exposed to modules in other components & \textbf{Component Independence:} 
-\% of code in modules classified as hidden~\cite{criteria:2017}\\\midrule
-
-\textbf{Keep Architecture Components Balanced} & Balancing the number of
-components ease locating code and allow for isolated maintenance & \textbf{Component Balance:} 
-Gini coefficient to measure the inequality of distribution between components~\cite{criteria:2017} \\\midrule
-
-\textbf{Keep your code base Small} & Reduce and avoid the system size because
-small products are easier to manage and maintain & \textbf{Volume:} \# of LOCs converted 
-to man-month/man-year~\cite{criteria:2017} \\\midrule
-
-\textbf{Automate Tests} & Test your code base because it makes development
-predictable and less risky & \textbf{Testability:} Ratings aggregation $-$ unit 
-complexity, component independence and volume~\cite{Visser:2016:OREILLY}
- \\\midrule
-
-\textbf{Write Clean Code} & Avoid producing software with code smells because
-it is more likely to be maintainable in the future & \textbf{Code Smells:} 
-\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
-identifier names) \\
-\bottomrule
-\end{tabular}
-\label{tab:guidelines}
-\end{table}
-
-Figure~\ref{fig:bchrep} shows an example of the report provided by BCH 
-for a project after finishing its evaluation. The example
-refers to the OpenSSL CVE-$2016$-$6304$ vulnerability patch---
-as described by Section~\ref{sec:motivation}. This version of 
-OpenSSL only complies with $1$ out of $10$ guidelines: \emph{Write 
-Clean Code}.
 
@@
 
-The same approach is used analyze the impact of regular 
+\begin{figure}[h]
+ 	\centering 	
+	\includegraphics[width=0.5\linewidth]{figures/commit.pdf}
+ 	\caption{Maintainability difference for security commits.}
+	\label{fig:commit}
+\end{figure}
+
+The same approach is used to analyze the impact of regular 
 changes. Security patches can be performed through one commit, 
@@
 are involved.
-
-\begin{figure}[h]
- 	\centering 	
-	\includegraphics[width=0.6\linewidth]{figures/commit.pdf}
- 	\caption{Maintainability difference for security commits.}
-	\label{fig:commit}
-\end{figure}
 
@@
 effect-size, as advocated by the Common-language effect 
-sizes\cite{graw:1992}, we compute the mean difference, the median of 
-the difference, and the percentage of cases that reduce maintainability.
+sizes, we compute the mean difference, the median of 
+the difference, and the percentage of cases that reduce maintainability~\cite{graw:1992}.
 %
@@
 commits from $260$ distinct open-source projects. This section 
-reports/discusses the results for each research question.
+reports and discusses the results for each research question.
 %
@@
 impact on all guidelines together ($M(v)$). Under each guideline, it 
-is stated the metric used to the calculations. For instance, for the 
+is stated the metric used for the calculations. For instance, for the 
 \emph{Write Short Units of Code} guideline, the metric used is 
@@
  \begin{figure}[htp]
-  	\centering 	 	\includegraphics[width=0.7\textwidth]{figures/main_per_severity.pdf}
+  	\centering 	 	\includegraphics[width=0.65\textwidth]{figures/main_per_severity.pdf}
   	\caption{Maintainability difference by vulnerability severity.}
@@
 in Modules}---more red points on the left, being $0$ the reference point.
+The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ 
+is $0.044$ (cf. Figure~\ref{fig:guidelines}). Since the p-value is 
+below the significance level of $0.05$, we argue that security patches 
+may have a negative impact on the maintainability of open-source software.
 
@@
 \emph{Low} severity vulnerabilities, i.e., \emph{Low} severity 
-vulnerabilities may have a less negative impact on software maintainability than the 
+vulnerabilities may have more cases where software maintainability was improved than the 
 other severity levels. However, results should not be disregarded 
@@
 Statistical significance was only retrieved for the \texttt{C/C++} 
-(p-value = $2.24$x$10^{-05}$), \texttt{Ruby} (p-value = $0.041$) and \texttt{PHP} 
-(p-value = $0.048$) languages. Yet, data reports very interesting hints on the impact 
+($p = 2.24$x$10^{-05}$), \texttt{Ruby} ($p = 0.041$) and \texttt{PHP} 
+($p = 0.048$) languages. Yet, data reports very interesting hints on the impact 
 of programming languages on security patches.
@@
 \emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
-are the most affected ones. No statistical significance was retrieved
+are the most affected ones. No statistical significance was observed
 for \emph{Separate Concerns in Modules}.
-The resulting p-value of the Paired Wilcoxon signed-rank test for $M(v)$ in Figure~\ref{fig:guidelines}
-is $0.044$. Since the p-value is 
-below the significance level of $0.05$, we argue that \textbf{security patches 
-may have a negative impact on the maintainability of open-source software}; and, 
-that tools such as BCH should be integrated into the CI/CD pipelines
+As shown in Figure~\ref{fig:guidelines}, there is statistical significance ($p=0.044 < 0.05$) to support our findings: \textbf{security patches
+may have a negative impact on the maintainability of open-source software}. Therefore, tools such as BCH should be integrated into the CI/CD pipelines
 to help developers evaluate the risk of patches of hindering software maintainability---alongside
-Pull Requests/Code Reviews. We also conclude that different severity 
-vulnerabilities may need different levels of attention and better and more secure 
-programming languages are needed.
+Pull Requests/Code Reviews. Different severity
+vulnerabilities may need different levels of attention---high/medium vulnerabilities need more attention (cf.
+Figure~\ref{fig:severity}).
+However, statistical significance was only observed for low severity vulnerabilities.
+Better and more secure programming languages are needed.
+We observed statistical significance for \emph{C/C++}, \emph{Ruby} and \emph{PHP} that support that security patches in those languages may hinder software maintainability (cf.
+Figure~\ref{fig:lang_main}).
 %
@@
 affect open-source software maintainability?}}
-
 In \emph{RQ2}, we report/discuss the impact of security patches on
@@
 
-\textbf{Summary:} \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure 
+\textbf{Summary:}
+Although results did not yield statistical significance, we show preliminary evidence that researchers and developers ought to pay more attention to maintainability when fixing the following types of weaknesses: \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure
 (CWE-200)}, \emph{Missing Release of Memory after Effective
-Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)} seem to be the \textbf{most likely 
-weaknesses to affect software maintainability}. No statistical evidence was retrieved. 
-However, results should not be disregarded because they hint on how 
-the maintainability is impacted per weakness which highlights the weaknesses
-that may need more attention from developers/researchers. 
-%
-\\\textit{\textbf{RQ3: What is the impact of security changes versus regular 
+Lifetime (CWE-401)} and \emph{Path Traversal (CWE-22)}.
+
+% No statistical evidence was observed.
+% However, results should not be disregarded because they hint on how
+% the maintainability is impacted per weakness which highlights the weaknesses
+% that may need more attention from developers/researchers.
+%
+
+\textit{\textbf{RQ3: What is the impact of security changes versus regular 
 changes on the maintainability of open-source software?}}
-
 The impact of security and baseline changes on software maintainability 
@@
 
-Results show that regular changes typically have a less negative impact
-on the software maintainability of open-source software.   
-However, we manually inspected a considerable amount of cases from
+Results show that regular changes are less prone to hinder the software maintainability of open-source software.   
+We manually inspected a considerable amount of cases from
 the distribution of the regular changes where maintainability 
@@
 
-\textit{\textbf{Construct Validity:}} The formula was inferred based 
+\textit{\textbf{Construct Validity:}} The formula to calculate the maintainability value ($M(v)$) was inferred based 
 on the BCH's reports. The high amount of different projects and backgrounds 
@@
 (vulnerable/fixed) of one vulnerability may have the same overall score
-and still be affected by different guidelines.
-%
+and still be affected by different guidelines. Therefore, we provide
+an analysis per guideline and our results are all available on figshare for future reproductions and deeper analysis.
 
@@
 perspective that a 
-developer may perform a patch quickly and later 
+developer may quickly perform a patch and later 
 proceed to the refactor is not considered. We assume 
@@
 on the impact of the vulnerability in the system, some 
-vulnerabilities may have more urgency to be fixed than 
-others. Baseline commits are retrieved randomly from 
+vulnerabilities may have more urgency to be patched than 
+others. For instance, a vulnerability performing a Denial-of-Service attack
+which usually brings entire systems down may be more urgent to patch than a 
+cross-site scripting vulnerability which usually does not have impact
+on the systems execution but rather on the data accessibility.
+
+Baseline commits are retrieved randomly from 
 the same project of the security patch.
@@
 
-\textit{\textbf{External Validity:}} We use a dataset that only comprises patches of open-source software.
-However, the methodology requires to access data that is not publicly available.
-Thus, our findings may not extend to non-open source software. Different programming 
+\textit{\textbf{External Validity:}} The BCH tool uses private and open-source data to determine the thresholds for each guideline. We only analyze patches of open-source software.
+Thus, our findings may not extend to private/non-open source software. Different programming 
 languages may require different coding practices to address software safety. The 
 dataset comprises more commits in Java, i.e., the dataset may not be representative 
-of the population regarding programming languages. Our approach does not consider 
+of the population regarding programming languages.
+For both datasets, we a manual validation of the message of the commits was performed.
+Only commits in English were considered.
+Thus, our approach does not consider
 patches in any other language but English.
@@
 % \end{itemize}
-\pagebreak
 

### 2020-09-10 paper/emse20.tex
spellcheck

@@
 several consecutive commits, or commit(s) interleaved with more 
-programming activities (floss-refactorings). This study considers 
+programming activities (floss-refactoring). This study considers 
 mainly single-commit patches. Yet, a small percentage of data points 
@@
 impact on maintainability. Green, means the impact was positive, i.e., 
-the patch improver software maintainability.
+the patch improved software maintainability.
 
@@
 	kind of impact on maintainability. Green, means the impact was 
-	positive, i.e., the patch improver software maintainability. For instance, in the 
+	positive, i.e., the patch improve software maintainability. For instance, in the 
 	\emph{Write Short Units of Code} guideline, $38.29\%$ of 
-	security patches harmed software maitainabilitty; $23.53\%$ of 
+	security patches harmed software maintainability; $23.53\%$ of 
 	security patches had no impact on maintainability; and, 
@@
 kind of impact on maintainability. Green, means the impact was 
-positive, i.e., the patch improver software maintainability.
+positive, i.e., the patch improved software maintainability.
 The $M(v)$ plot show that $406$ ($41.9\%$) cases have a negative 
@@
 	Providing automated tools to developers assess the risk of their 
-	patches it is important to help companies shipping software of higuer quality. 
+	patches it is important to help companies shipping software of higher quality. 
 	Tools like Better Code Hub and static analysis (e.g., 

### 2021-01-20 paper/emse20.tex
first changes to the paper

@@
 of vulnerabilities being discovered and fixed, almost weekly. We 
-suspect that developers affect the maintainability of their 
+hypothesize that developers affect the maintainability of their 
 codebases when patching vulnerabilities. This paper evaluates the 
@@
 open-source software. Maintainability is measured based on the 
-Better Code Hub’s model of 10 guidelines using a dataset including 
+Better Code Hub’s model of 10 guidelines on a dataset including 
 1300 security-related commits. Results show evidence of a trade-off 
-between security and maintainability, i.e., developers may be 
-hindering software maintainability. The implications of our study 
+between security and maintainability in $41.9\%$ of the cases, i.e., developers 
+may hinder software maintainability. More than $35\%$ of the security 
+patches studied increased software complexity and the percentage 
+of LOCs per unit. The implications of our study 
 are that changes to codebases while patching vulnerabilities need to 

### 2021-01-20 response/response.tex
adding reviewer 1 response

\documentclass[11pt,fleqn]{article}

\usepackage{vmargin}
\setpapersize{A4}
\setmarginsrb{2.5cm}{2.5cm}{2.5cm}{2.5cm}%
{\baselineskip}{2\baselineskip}{baselineskip}{2\baselineskip}
\setlength{\parindent}{0pt}
\setlength{\parskip}{5pt}

\usepackage{caption}
\usepackage{subcaption}

\usepackage{epsfig}
\usepackage{latexsym}
\usepackage{url}
\usepackage{tgheros}
\renewcommand*\familydefault{\sfdefault}

\newcommand{\eline}{\vspace*{.75\baselineskip}}
\newcommand{\Referee}[1]{\eline \noindent {\bf Reviewer comment #1:} \\}
\newcommand{\Us}{\eline \noindent {\bf Response:}\\}
\newcommand{\TBD}{{\bf To Be Done}}
\newcommand{\newreviewer}[1]{\section*{Reviewer #1}\vspace*{-1.05\baselineskip}}
\newcommand{\editor}[1]{\section*{Editor #1}\vspace*{-1.05\baselineskip}}

%\usepackage{tikz}
\usepackage[skins,breakable]{tcolorbox}
\tcbset{textmarker/.style={
%
skin=enhancedmiddle jigsaw,breakable,parbox=false,before skip=-1mm, after skip=0mm,
boxrule=0mm,leftrule=2mm,rightrule=2mm,boxsep=0mm,arc=0mm,outer arc=0mm,
left=2mm,right=2mm,top=2mm,bottom=2mm,toptitle=1mm,bottomtitle=1mm,oversize}}

\newtcolorbox{rcomment}{textmarker,colback=gray!15!white,colframe=gray!80!white}

\newenvironment{revcomment}[1][]
{\Referee{#1}\begin{rcomment}}
{\end{rcomment}}

\newenvironment{reveditor}
{\begin{rcomment}}
{\end{rcomment}}

\usepackage{todo}
\usepackage{hyperref}

%customized
\usepackage{xfrac}

% more lenient line breaking (avoids text protruding into margins)
\sloppy

\title{\vspace*{-2cm}{\bf Authors' Response to the Review of
 EMSE-D-20-00300:\\
 ``Fixing Vulnerabilities Potentially Hinders Maintainability''}}

\author{Sofia Reis, Rui Abreu, Luis Cruz}
\date{}

\begin{document}

\maketitle

\editor{}

\begin{reveditor}
    Overall, I would like to give you a chance to respond to 
    these issues in a revision of your manuscript but I have 
    to make it clear that it does not guarantee a path towards 
    acceptance. Your rebuttal to \#1 will be of critical importance 
    for me judge the validity of the use of the BCH tool and in 
    making a final decision. Methodology is, of course, very 
    important for ESE journal and \#2 \& \#3s comments on methodological 
    details need to be carefully addressed, too. 
\end{reveditor}

\Us We thank the editor and reviewers for their valuable feedback. In the 
following points, we address the concerns raised by the reviewers.

\newreviewer{1}

\begin{revcomment}[1.1]

    This paper investigates the impact of patches to improve 
    security on the maintainability of open-source software. 
    The paper is well written and easy to read. I really 
    appreciated the replication package: complete, and well 
    described. Really a good work!

    However, I found a crucial issue in this paper. The authors 
    adopted as static analysis tool Better Code Hub. Better Code 
    Hub' model includes 10 guidelines that can help the developers to 
    write better code. Unfortunately, these guidelines are not related 
    to maintainability. Avoiding introducing the issues associated to 
    these guidelines do not imply increasing the maintainability of the 
    code, since the tools has no possibility to measure "maintainability".

\end{revcomment}

\Us Thank you for your positive feedback on the paper presentation and replication
package. BCH is not a static analysis tool instead it checks codebases for 
compliance with the ten guidelines. The set of guidelines presented in our paper
is the same published by Software Improvement Group (SIG) in their ebook called
"Building Maintainable Software: Ten Guidelines for Future-Proof Code"
\footnote{Available here: https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
According to the authors, all the guidelines were infered from analyzing
hundres of real-world systems and measure maintainability.

We clarified the difference between BCH and static analysis tools in
the Section 1 (Introduction) and Section 3 (Methodology).

\begin{revcomment}[1.2]

    As far as I know, the company that developed Better Code Hub 
    developed also the Delta Maintainability Model [diBiase2019] 
    that aims at measuring the maintainability of a code change 
    and a score and compare change-based maintainability measurements.

    [diBiase2019] M. di Biase and A. Rastogi and M. Bruntink and A. van 
    Deursen. The Delta Maintainability Model: Measuring Maintainability 
    of Fine-Grained Code Changes. IEEE/ACM International Conference 
    on Technical Debt (TechDebt) in 2019.

\end{revcomment}

\Us Thank you for suggesting Marco's paper. Our paper uses a model 
for measuring maintainability that was published previously by one of the co-authors
of this paper [Cruz2019]. The entire methodology and calculations 
were actually validated by Marco.

The model created by Marco focus on measuring maintainability of fine-grained 
changes using 5 of the same guidelines we use. The SIG-MM model in 
Marco's paper is the model behind Better Code Hub.

We added Marco's work to our related work.

[Cruz2019] Luis Cruz, Rui Abreu, John Grundy, Li Li, Xin Xia. 
Do Energy-oriented Changes Hinder Maintainability?. International 
Conference on Software Maintenance and Evolution (ICSME) in 2019.

\begin{revcomment}[1.3]

    Moreover, I am not sure that Better Code Hub is 
    the best static analysis tools to detect security 
    issues. One of the most adopted in security domain 
    is Coverity Scan for examples, but also other tools 
    might be a better choice (e.g. Kiuwan).

\end{revcomment}

\Us Thank your for your comment. In this paper, we do not
try to use static analysis to detect security issues.
We have already a dataset of security patches (vulnerable versions
and safe versions) and our goal is to understand what 
is the impact of those patches on the maintainability
guidelines/metrics. We arguee that these guidelines/metrics
can in the future complement static analysis tools 
by assiting developers with more information on the 
risks associated with their patches.

SonarQube uses 4 of the set of metrics we are using and 
also rates maintainability. In Section X, 
we compare the results between BCH and SonarQube. However, 
we arguee that BCH performs a more complete analysis.

\begin{revcomment}[1.4]

    The authors could have adopted both approaches, Better 
    Code Hub guidelines or another one first and then measure 
    the code maintainability with the Delta Maintainability Model 
    since the goal of the paper is to improve security on the 
    maintainability.

\end{revcomment}

\Us Thank you for your comment. As we explained before, while 
addressing comment 1.2, both models assess maintainability.
Delta Maintainability Model is an alternative to our approach 
that uses only 5 of the set of guidelines our study considers.

\begin{revcomment}[1.5]

    In my opinion, the idea behind this work is very interesting, 
    but the research questions cannot be answered with the tools 
    and metrics selected.

    My recommendation is to select a specific static analysis tool 
    for security issues and to include a maintainability model to 
    evaluate the maintenance.

\end{revcomment}

\Us Thank you for the positive interest on the work. As we 
explained before Better Code Hub calculates maintainability metrics.
Our goal was to measure the impact of security patches on software 
maintainability. For future research, it would be interesting to 
understand if the results of these metrics can actually help 
security engineers assessing the risk of their patches and guide 
them on performing better patches. However, in this work, 
we only focus on assessing software maintainability.

\newreviewer{2}

\begin{revcomment}[2.1]

    Question:  Is your dataset available?

\end{revcomment}

\Us Thank you for your question. The dataset is fully available at 
\texttt{dataset/db\_release\_security\_fixes.csv} in the figshare package
provided in the contributions of the paper: 
\url{https://figshare.com/s/4861207064900dfb3372}. If the paper is 
accepted, the package will be available on GitHub.

\begin{revcomment}[2.2]

    Abstract:
    Throughout the paper, "hypothesize" is probably a better word than 
    "suspect" for sounding more scientific.

    Your results should be more specific than "show evidence of trace-off". 
    Briefly tell us about what metrics you used and what the numerical results 
    indicate.

\end{revcomment}

\Us We thank the reviewer for the suggestions. We addressed
both of the issues in the abstract. For the later issue, we reported 
the overall result for maintainability and results
for the two of the guidelines with more negative impact in software 
maintainability: software complexity and unit size.

\end{document}

### 2021-01-20 response/response.tex
more changes and responses for reviewer 2

@@
 
+\begin{revcomment}[2.3]
+
+    Intro:
+    \begin{itemize}
+        \item First sentence - quality is not ONLY related to cost but also to security 
+        and safety.
+        \item Provide a URL to Software Improvement Group and Better Code Hub.
+        \item Page 3, Line 1 - Application Security Verification Standard (ASVS).
+        \item Page 3, Line 7 - instead of a "broad number of code metrics" - tell 
+        us exactly the number of metrics.
+        \item Page 3, Line 13 - "suggest" sounds more scientific than "hint at"
+        \item Page 3, Line 24 - "we intend to highlight the need …". Is that the 
+        broad goal of your paper?  The goal should be explicitly stated in the 
+        abstract and intro.
+    \end{itemize}
+\end{revcomment}
+
+\Us Thank you for reporting these minor issues. All of them were addressed in 
+the new version. In the last point, where the reviewer ask ``Is that the 
+broad goal of your paper?'', we want to clarify that the goal of our study
+is to show the impact of security patches on sofwtare maintainability and
+highlight the need for solutions which are described in the introduction
+and end of the abstract.
+
+\begin{revcomment}[2.4]
+
+    Intro:\\
+    You should check out this paper: 
+    Li, Frank and Paxson, Vern. A large-scale empirical study of security patches.
+    ACM SIGSAC Conference on Computer and Communications Security in 2017.
+
+\end{revcomment}
+
+\Us Thank you for bringing our attention to this paper. We integrated
+this paper in the Related Work section and used it to support some 
+of our claims and findings.
+
 \end{document}

### 2021-01-20 paper/emse20.tex
small fix to abstract
more changes and responses for reviewer 2

@@
 between security and maintainability in $41.9\%$ of the cases, i.e., developers 
-may hinder software maintainability. More than $35\%$ of the security 
-patches studied increased software complexity and the percentage 
+may hinder software maintainability. More than $35\%$ of security 
+patches increased software complexity and the percentage 
 of LOCs per unit. The implications of our study 
@@
 update for software product quality ISO/IEC 25010 considering 
-\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. However, there is still 
+\emph{Security} as one of the main software product quality 
+characteristics~\cite{iso:2011}. However, there is still 
 a considerable amount of vulnerabilities being discovered and fixed, 
@@
 Researchers found a correlation between the presence of 
-vulnerabilities on software and code 
-complexity~\cite{shin2010evaluating,10.1145/1774088.1774504}. 
+vulnerabilities on software and code complexity~\cite{shin2010evaluating,10.1145/1774088.1774504}. 
 Security experts claim that complexity hides bugs that may result in 
-security 
-vulnerabilities~\cite{mcgraw2004software,schneier2006beyond}. In 
+security vulnerabilities~\cite{mcgraw2004software,schneier2006beyond}. In 
 practice, an attacker only needs to find one way into the system 
@@
 Complex code is difficult to understand, maintain and 
-test~\cite{1702388}. Thus, the developer's task gets harder as the 
+test~\cite{1702388}. Thus, the task of a developer gets harder as the 
 codebase grows in size and complexity. But the risk can be minimized 
@@
 %
-Novel tools have been built to automatically detect software 
-vulnerabilities (e.g., FindBugs, Infer and more). Despite these 
-efforts, improving software security is not a trivial task and 
+Static analysis tools (SATs) have been built to automatically detect software 
+vulnerabilities (e.g., FindBugs, Infer and more). Developers
+use those tools to locate the issues in the code. However,
+while performing the patches to those issues, SATs are not 
+capable of providing information on the quality of the patch.
+Improving software security is not a trivial task and 
 requires implementing patches that might affect software 
-maintainability. Our suspicion is that some of these patches may 
+maintainability. Our hypothesis is that some of these patches may 
 have a negative impact on the software maintainability and, 
@@
 vulnerabilities---harming software reliability and introducing 
-technical debt. Therefore, in this paper, we present an empirical 
-study on the impact of patches of vulnerabilities on software 
-maintenance across open-source software.
+technical debt. Research found that $34\%$ of the security patches 
+performed introduce new problems and $52\%$ are incomplete and do not 
+fully secure systems~\cite{10.1145/3133956.3134072}. Therefore, in this paper, 
+we present an empirical study on the impact of patches of 
+vulnerabilities on software maintenance across open-source software.
+We arguee that tools that assess these type of metrics may
+complement SATs with valuable information to understand and
+address better the issue.
 
@@
 calculate maintainability, we resort to Software Improvement Group 
-(SIG)'s web-based source code analysis service Better Code Hub (BCH) 
+(SIG\footnote{SIG's website: 
+\url{https://www.sig.eu/} (Accessed on \today{})})'s web-based source 
+code analysis service Better Code Hub (BCH)\footnote{BCH's 
+website: \url{https://bettercodehub.com/} (Accessed on \today{})}  
 to compute the software compliance with a set of $10$ 
 guidelines/metrics to produce quality software based in ISO/IEC 
-$25010$~\cite{Visser:2016:OREILLY}. There are other well-known 
+$25010$~\cite{Visser:2016:OREILLY}. SIG 
+has been helping business and technology leaders drive their organizational 
+objectives by fundamentally improving the health and security of 
+their software applications for more than 20 years. Their 
+models are scientically proven~\cite{5609747,6113040,baggen2012} and certified.
+
+There are other well-known 
 standards and models that have been proposed to increase software 
@@
 evaluation; the OWASP Application Security Verification 
-Criteria~\cite{oswap:2009} which is focused only on web 
+Standard (ASVS)~\cite{oswap:2009} which is focused only on web 
 applications, and a model proposed by Xu et al. ($2013$)
@@
 an assessment of maintainability in software for the following 
-reasons: BCH integrates a broad number of code metrics; and, 
+reasons: BCH integrates a total of $10$ different code metrics; and, 
 code metrics were empirically validated in previous 
@@
 in the codebases to patch vulnerabilities affect code 
-maintainability. Results also hint that developers 
+maintainability. Results also suggest that developers 
 should pay different levels of attention to different severity 
@@
 	programming language).
-	\item A replication package with all scripts and data created to 
+	\item A replication package with the scripts and data created to 
 	perform the empirical evaluation, for reproducibility. Available 
@@
 %
-Due to time-to-market pressure and the lack of security 
-expertise~\cite{8077802}, code-related security flaws are generally 
-detected \textit{after the fact}, i.e., when hackers exploit them. 
-Patching these issues is usually done by modifying the codebase. 
-However, these patches may have a negative impact on software 
-maintenance (e.g., developers might follow the quickest solution to 
-fix the problem but not necessarily the most elegant and performant 
-one). Thus, we want to understand how security patches impact
-software maintainability.
+% Due to time-to-market pressure and the lack of security 
+% expertise~\cite{8077802}, code-related security flaws are generally 
+% detected \textit{after the fact}, i.e., when hackers exploit them. 
+% Patching these issues is usually done by modifying the codebase. 
+% However, these patches may have a negative impact on software 
+% maintenance (e.g., developers might follow the quickest solution to 
+% fix the problem but not necessarily the most elegant and performant 
+% one). Thus, we want to understand how security patches impact
+% software maintainability.
 
@@
 representative benchmark of closed and open-source software projects to compute 
-the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,Baggen2012}.
+the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,baggen2012}.
 Maintainability is computed as the mean of all guidelines. Different software versions 

### 2021-01-20 paper/emse20.tex
small fix

@@
 the overall cost of developing and maintaining software 
-applications~\cite{slaughter1998evaluating}. Software quality 
+applications, security and safety~\cite{slaughter1998evaluating}. Software quality 
 characteristics include, but are not limited to functional 

### 2021-01-26 response/response.tex
more changes

@@
 
+\begin{quote}
+    Introduction\\
+    ...\\
+    Static analysis tools (SATs) have been built to automatically detect software 
+    vulnerabilities (e.g., FindBugs, Infer and more). Developers
+    use those tools to locate the issues in the code. However,
+    while performing the patches to those issues, SATs are not 
+    capable of providing information on the quality of the patch.
+    Improving software security is not a trivial task and 
+    requires implementing patches that might affect software 
+    maintainability. Our hypothesis is that some of these patches may 
+    have a negative impact on the software maintainability and, 
+    possibly, even be the cause of the introduction of new 
+    vulnerabilities---harming software reliability and introducing 
+    technical debt. Research found that $34\%$ of the security patches 
+    performed introduce new problems and $52\%$ are incomplete and do not 
+    fully secure systems[Li2017]. Therefore, in this paper, 
+    we present an empirical study on the impact of patches of 
+    vulnerabilities on software maintenance across open-source software.
+    We arguee that tools that assess these type of metrics may
+    complement SATs with valuable information to help the developer
+    understand better the risk of its patch.\\
+    ...
+\end{quote}
+
+[Li2017] F. Li and V. Paxson. A Large-Scale Empirical Study of Security Patches. 
+A Large-Scale Empirical Study of Security Patches in 2017.
+
 \begin{revcomment}[1.2]
@@
 The model created by Marco focus on measuring maintainability of fine-grained 
-changes using 5 of the same guidelines we use. The SIG-MM model in 
+changes using 5 of the same guidelines we already use. The SIG-MM model in 
 Marco's paper is the model behind Better Code Hub.
 
-We added Marco's work to our related work.
+We added Marco's work to the related work.
 
@@
 
-SonarQube uses 4 of the set of metrics we are using and 
-also rates maintainability. In Section X, 
+SonarQube uses 4 metrics of the set of metrics we are using and 
+also rates maintainability. In Section 3 (\textit{Why Bettter Code Hub?}), 
 we compare the results between BCH and SonarQube. However, 
@@
 Delta Maintainability Model is an alternative to our approach 
-that uses only 5 of the set of guidelines our study considers.
+and only uses 5 guidelines of the set of guidelines our study considers.
 
@@
 understand if the results of these metrics can actually help 
-security engineers assessing the risk of their patches and guide 
+security engineers assess the risk of their patches and guide 
 them on performing better patches. However, in this work, 
@@
 the overall result for maintainability and results
-for the two of the guidelines with more negative impact in software 
+for the two of the guidelines with more negative impact on software 
 maintainability: software complexity and unit size.
@@
 
-\Us Thank you for reporting these minor issues. All of them were addressed in 
-the new version. In the last point, where the reviewer ask ``Is that the 
+\Us Thank you for reporting these issues. All of them were addressed in 
+the new version of the paper. In the last point, where the reviewer asks ``Is that the 
 broad goal of your paper?'', we want to clarify that the goal of our study
-is to show the impact of security patches on sofwtare maintainability and
-highlight the need for solutions which are described in the introduction
-and end of the abstract.
+is to show the impact of security patches on software maintainability and
+highlight the need for solutions which are described in the end of 
+the abstract, Section 1 (\textit{Introduction}) and Section 5 (\textit{Study Implications}).
 
@@
 
+\begin{revcomment}[2.5]
+
+    Section 2\\
+    \begin{itemize}
+        \item The first paragraph is redundant with Section 1.  
+        \item Page 4, Line 15 - You define the OCSP acronym late in the paragraph.
+        \item Page 5, Line 6 - how many new branch points?
+        \item Page 6, Line 19 - I think you want to say "In this study, maintainability is …."  
+        Since it currently implies this information is available in the CWE.
+        \item Page 6, line 47 - "regular commits" non-security commits to be more clear; maybe a 
+        few more words to explain "randomly collected."  You also call them "baseline commits" in Figure 
+        1 which gives a different phrase for the same concept.  Pick one and use it everywhere.
+    \end{itemize}
+\end{revcomment}
+
+\Us Thank you for pointing out all of these issues. We addressed all of them. We decided
+to use regular changes/regular commits instead of baseline and non-security commits.
+
+\begin{revcomment}[2.6]
+    Section 3\\
+    \begin{itemize}
+        \item Sometime you say 1330 patches (e.g. page 7, line 16) and sometimes 1300 (e.g. page 8, line 11, 
+        and the abstract).  In science it's better to say the exact number and use it always and not round.  
+        \item You say the dataset had 1330 (or 1300) patches and 1282 commits though you also say one patch 
+        can have multiple commits assigned.  So how did you end up with less than one commit/patch?
+    \end{itemize}
+\end{revcomment}
+
+\Us Thank you for raising this concern. The right number of security patches is $1300$ ($624$ from Ponta et al. and 
+$676$ from Secbench). The $1330$ value is a typo, we fixed the issue. Regarding the second point, $1282$ commits is
+the equivalent to the $624$ patches that integrate the Pontas et al. dataset where one patch
+can have multiple commits. The $1282$ commits are referring to the Ponta et al. dataset and 
+not to the combined dataset.
+
+\begin{revcomment}[2.7]
+    Section 3\\
+    \begin{itemize}
+        \item I would really like a better picture of what SIG is to establish credibility.  I was at first 
+        thinking it had something to do with ACM SIG's but now after going to the website, it seems like a 
+        commercial/consulting organization? I also feel irritated that I don't understand what you really mean 
+        by "running against the BCH toolset" (page 8, line 13) since you have repeatedly mentioned BCH like it was 
+        an industry standard - but you have not told me if it's a static analysis tool or what it is.  The explanation 
+        and link to introduce SIG and BCH should be on Page 2 lines 41-42 rather than Page 7 - though an explanation 
+        of the BCH toolset is best a paragraph in the methodology.  
+    \end{itemize}
+\end{revcomment}
+
+\Us Thank you for raising the concern regarding SIG's credibility. SIG is a company that 
+has more than $20$ years of experience and research in software quality production. 
+Their models are scientifically proven and certified. We added this information and 
+the links the reviewer refer in the comment to the Introduction section.
+
+\begin{quote}
+    As ISO does not provide any specific guidelines/formulas to 
+calculate maintainability, we resort to Software Improvement Group 
+(SIG\footnote{SIG's website: 
+\url{https://www.sig.eu/} (Accessed on \today{})})'s web-based source 
+code analysis service Better Code Hub (BCH)\footnote{BCH's 
+website: \url{https://bettercodehub.com/} (Accessed on \today{})}  
+to compute the software compliance with a set of $10$ 
+guidelines/metrics to produce quality software based in ISO/IEC 
+$25010$ [Visser2016]. SIG 
+has been helping business and technology leaders drive their organizational 
+objectives by fundamentally improving the health and security of 
+their software applications for more than 20 years. Their 
+models are scientifically proven and certified [Alves2010, Alves2011, Baggen2012].
+\end{quote}
+
+We rephrased "running against the BCH toolset" to "analyzed using the BCH toolset". We mean
+that we used BCH to calculate the metrics presented on Table 1.
+
+[Visser2016] J. Visser. Building Maintainable Software, Java Edition: Ten Guidelines 
+for Future-Proof Code. O'Reilly Media, Inc. in 2016.
+
+[Alves2010] T. L. Alves and C. Ypma and J. Visser. Deriving metric thresholds from 
+benchmark data. IEEE International Conference on Software Maintenance in 2010.
+
+[Alves2011] T. L. Alves and J. P. Correia and J. Visser. Benchmark-Based Aggregation 
+of Metrics to Ratings. Joint Conference of the 21st International Workshop on Software 
+Measurement and the 6th International Conference on Software Process and Product Measurement
+in 2011.
+
+[Baggen2012] R. Baggen, J. P. Correia, K. Schill, J. Visser. Standardized code quality 
+benchmarking for improving software maintainability. Software Quality Journal in 2012.
+
+\begin{revcomment}[2.8]
+    Section 3\\
+    \begin{itemize}
+        \item Page 7, line 4 - I don't think "resemble" is the word you are looking for - but I don't 
+        know what you are trying to say so I can't make a suggestion.
+        \item Please explain your repeatable methodology to "clean the dataset and select the most relevant and 
+        compatible projects" (page 8, line 14) that brings the patches from 1282 to 969.  From 969 to 866 patches - 
+        you just could not classify?  Please explain.
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.9]
+    Section 3\\
+    \begin{itemize}
+        \item Page 8, line 33 "regular commit" - pick baseline or regular (or non-security) commits and use it always.  
+        I don't understand the sentence "The baseline dataset is generated from the security commits dataset."   
+        Do you mean you extract these commits from the projects in the security commits dataset?
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.10]
+    Section 3\\
+    \begin{itemize}
+        \item Section 3.2 is confusing enough that I'm not sure of the analysis - 
+        I will need to review the revision that you would submit to gain confidence.  
+        Are you saying you want your non-security commits (regular commits) to be about 
+        the same size as a given security commit so you are using this criteria to choose 
+        your regular commits?  Justify this reasoning and make the methodology clear.  
+        I would have expected that you randomly collected 1300 non-security commits to compare 
+        with the 1300 security commits. Why constrain to make them be of "similar" size?  
+        Why not allow the characteristics of regular versus security commits to be what 
+        they are without such curating?
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.11]
+    Section 3\\
+    \begin{itemize}
+        \item Section 3.2 is confusing enough that I'm not sure of the analysis - 
+        I will need to review the revision that you would submit to gain confidence.  
+        Are you saying you want your non-security commits (regular commits) to be about 
+        the same size as a given security commit so you are using this criteria to choose 
+        your regular commits?  Justify this reasoning and make the methodology clear.  
+        I would have expected that you randomly collected 1300 non-security commits to compare 
+        with the 1300 security commits. Why constrain to make them be of "similar" size?  
+        Why not allow the characteristics of regular versus security commits to be what 
+        they are without such curating?
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.12]
+    Section 3\\
+    \begin{itemize}
+        \item Since SIG seems to be "just another consulting organization" - I'd like 
+        to see about the acceptance of the 10 guidelines in the software engineering 
+        discipline.  I would guess you could find peer-reviewed references to support 
+        these guidelines, or even books by Martin Fowler.   That has more credibility 
+        than the reference by the SIG consulting group. For example on Page 9, line 42 
+        you cite McCabe Complexity --  that has more credibility than a consulting 
+        organizations guidelines that they use to make money on their tool.  Tell us more 
+        about the BCH data experience since that is your comparison point for compliance.
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.13]
+    Section 3\\
+    \begin{itemize}
+        \item Since SIG seems to be "just another consulting organization" - I'd like 
+        to see about the acceptance of the 10 guidelines in the software engineering 
+        discipline.  I would guess you could find peer-reviewed references to support 
+        these guidelines, or even books by Martin Fowler.   That has more credibility 
+        than the reference by the SIG consulting group. For example on Page 9, line 42 
+        you cite McCabe Complexity --  that has more credibility than a consulting 
+        organizations guidelines that they use to make money on their tool.  Tell us more 
+        about the BCH data experience since that is your comparison point for compliance.
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.13]
+    Section 3\\
+    \begin{itemize}
+        \item Page 10, line 49 "mainly single-commit patches" … "small percentage 
+        of data points" - give us the exact numbers for each of these.  Page 11, 
+        line 38 - define "floss refactoring" and justify how these were objectively 
+        and repeatably identified in your set.   You say "these cases" which makes 
+        it seems that you were including the set of patches that had more than one 
+        commit as your (only) floss candidates.  
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.14]
+    Section 3\\
+    \begin{itemize}
+        \item Page 11, Line 42 - I think you are saying you removed these two guidelines 
+        from the analysis of all projects (as shown in Figure 4) but this paragraph seems 
+        to say this was only for projects with large code bases.
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.15]
+    Section 3\\
+    \begin{itemize}
+        \item Page 11  line 48.  How many were disregarded?
+    \end{itemize}
+\end{revcomment}
+
+\Us until here
+
+\begin{revcomment}[2.16]
+    Section 4\\
+    \begin{itemize}
+        \item Page 13, Line 3 - 969 security commits and 969 
+        baseline commits?  It's hard to know how to parse that.
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.17]
+    Section 4\\
+    \begin{itemize}
+        \item Page 13, Line 31 "overall patches …" rather than having us 
+        rely on looking at the Figure, tell us the number or percentage of 
+        when the impact is positive like you do with the negative in the next 
+        sentence. [You do provide the specifics starting on page 15 line 49.  
+        I wanted it here.]
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.18]
+    Section 4\\
+    \begin{itemize}
+        \item Page 19, Line 7 - for the unindoctrinated - I'd 
+        explain the concept of Research Concepts.  Also explain 
+        your choice of 7a and 7b and how representative they are 
+        from the several hundred choices. Similar for your explanations 
+        in the second two paragraphs on this page.  How far down the 
+        concept/vulnerability hierarchy tree are these CWE choices or 
+        are they lower level vulnerabilities chosen from the 700+ CWE types?
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.18]
+    Section 4\\
+    \begin{itemize}
+        \item Page 20, Line 42 - how many is "a considerable number of cases"?  
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.18]
+    Section 4\\
+    \begin{itemize}
+        \item Page 20, Line 45 - it seems choosing commits of similar size 
+        to the security commits ended up causing a limitation.  I'm still 
+        wondering why you chose that methodology anyway.  
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.19]
+    Section 4\\
+    \begin{itemize}
+        \item RQ3 is answered to succinctly and with too little 
+        detail to be convincing to me that the differences between 
+        security and regular changes is different.  
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.20]
+    Section 5\\
+    \begin{itemize}
+        \item Line 32 - since RQ3 was not convincing, 
+        I don't feel a separate effort for maintainable security 
+        is justified as separate from general focus on maintainability. 
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[2.20]
+    Section 6\\
+    \begin{itemize}
+        \item It might not be a threat - but I feel the curation 
+        of the non-security commits is a limitation to the generalizability 
+        of the characteristics of non-security commits
+    \end{itemize}
+\end{revcomment}
+
+\Us
+
+\newreviewer{3}
+
+\begin{revcomment}[3.1]
+
+    The paper addresses a relevant and up to date topic. 
+    The authors have analyzed a large set of project wrt to 
+    quality assessment focused on vulnerabilities in line with 
+    ISO25000 standard. There are some points worth considering: 
+    - authors should better motivate the choice for BCH as web 
+    based source code analysis service with respect to other 
+    options that are available in literature. Tools such as Kiuwan, 
+    sonar cloud, codify etc are also able to provide specific metrics 
+    on maintainability and security, vulnerability features of code. 
+    Furthermore Kiuwan is also an example designed based on the ISO25000 
+    standard. Authors discuss the need for tools for risk assessment in 
+    the study implications (section5) but fail to assess why their choice 
+    falls on BCH. Furthermore all tools provide different conceptualization 
+    and evaluation criteria for calculating quality characteristics such as Vulnerabilities 
+    and Security. Authors should also explicit how these quality characteristics are measured in BCH.
+
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[3.2]
+
+    In section 3a better state/formulate the hypothesis authors 
+    analyze wrt to the Research questions in previous sections. 
+    In general the empirical study is not structured according to any 
+    of the guidelines or explicit states any of them.
+
+\end{revcomment}
+
+\Us
+
+\begin{revcomment}[3.3]
+
+    The findings of the research should better emphasize what the 
+    lessons learned and what better practices should be put in place 
+    by developers to assure better quality of software. The take away 
+    message remains hindered.
+
+\end{revcomment}
+
+\Us
+
 \end{document}

### 2021-01-26 paper/emse20.tex
more changes

@@
 consequently, less prone to
-vulnerabilities~\cite{shin2010evaluating,10.1145/1774088.1774504})---
+vulnerabilities~\cite{shin2010evaluating,10.1145/1774088.1774504}---
 and, prevent the introduction of new vulnerabilities.
+
 % Unmaintainable code is hard to test, analyze and re-use. Therefore,
@@
 We arguee that tools that assess these type of metrics may
-complement SATs with valuable information to understand and
-address better the issue.
+complement SATs with valuable information to help the developer
+understand better the risk of its patch.
 
@@
 their software applications for more than 20 years. Their 
-models are scientically proven~\cite{5609747,6113040,baggen2012} and certified.
+models are scientifically proven and certified~\cite{5609747,6113040,baggen2012}.
 
@@
 \section{Motivation and Research Questions}\label{sec:motivation}
-%
+% Rev 2 says it is redundant 
 % Due to time-to-market pressure and the lack of security 
@@
 extension data during session renegotiation, and where a malicious 
-client could cause a Denial-of-Service (DoS) attack via large OCSP 
-Status Request extensions when OCSP stapling support was enabled. 
-Online Certificate Status Procotol (OCSP) stapling, formally known 
+client could cause a Denial-of-Service (DoS) attack via large Online 
+Certificate Status Procotol (OCSP) Status Request extensions when 
+OCSP stapling support was enabled. OCSP stapling, formally known 
 as the TLS Certificate Status Request extension, is a standard for 
@@
 for handling the application when no OCSP response IDS are 
-allocated. After the patch, in the new version of the software, the condition is checked before the package processing 
-instead of after. Thereby, the system avoids the increase of unbounded memory (and, a potential DoS attack).
+allocated. After the patch, in the new version of the software, the 
+condition is checked before the package processing 
+instead of after. Thereby, the system avoids the increase of 
+unbounded memory (and, a potential DoS attack).
 
@@
 already with a large number of lines of code and introduced more 
-complexity to the code with new branch points which disrupt two of 
+complexity to the code with $2$ new branch points which disrupt two of 
 the guidelines proposed by the Software Improvement Group 
@@
 (CWE) glossary, is a type of code-flaw that could contribute to the 
-introduction of vulnerabilities within that product. Maintainability 
+introduction of vulnerabilities within that product. In this study, maintainability 
 is measured separately for each weakness.
@@
 %
+\item After combining both datasets, we removed the duplicates
+and filtered out every floss-refactoring.
+%
 	\item
@@
 %
-	\item A baseline of regular commits was randomly collected from 
+  \item A baseline of regular commits was randomly collected (i.e., for 
+  each security commit a random commit from the same project was selected) from 
 	the list of projects of the final dataset to evaluate the impact 
@@
   to quantify maintainability for both security and regular commits. 
-  The codebases that resemble to the commits of our datasets are 
-  collected before the analysis performed by BCH.
+  BCH evaluates the codebase available in the default branch of a GitHub project. 
+  We created a tool that pulls the codebase of each commit of our dataset 
+  to a new branch; it sets the new branch as the default branch; it runs 
+  the BCH analysis in the codebase; and, finally saves the BCH metrics
+  results.
 \end{enumerate}
@@
 	\centering 	
-    \begin{adjustwidth}{0.5cm}{-1cm}  
-	\includegraphics[width=0.9\textwidth]{figures/methodology.pdf}
+    \begin{adjustwidth}{-0.5cm}{-0.5cm}  
+      \vspace{-2.5cm} 
+	\includegraphics[width=1.1\textwidth]{figures/methodology.pdf}
+  \vspace{-3cm} 
   \caption{Study Methodology}
@@
 
-We use a combined dataset of $1330$ security patches which is 
+We use a combined dataset of $1300$ security patches which is 
 the outcome of mining and manually inspecting a total of $312$ 
@@
 commits in the repositories. This technique yielded a dataset of 
-$624$ patches~\cite{10.1109/MSR.2019.00064}. The original dataset 
-contains $1282$ commits. One patch can have multiple commits 
+$624$ patches~\cite{10.1109/MSR.2019.00064} and $1282$ commits. 
+One patch can have multiple commits 
 assigned. To fit the dataset in our methodology, we located the 
@@
 impact was positive, negative or none. The $1300$ patches in the 
-dataset were run against the BCH toolset to calculate their 
+dataset were analyzed using the BCH toolset to calculate their 
 maintainability reports. Due to the limitations of BCH, in particular, 
@@
 Common Weakness Enumeration (CWE) taxonomy to classify each vulnerability. 
-For instance,  the \texttt{Fix CVE-2014-1608: mc\_issue\_attachment\_get SQL injection} (\texttt{00b4c17}\footnote{CVE-$2014$-$1608$ details available at \url{https://github.com/mantisbt/mantisbt/commit/00b4c17088fa56594d85fe46b6c6057bb3421102} (Accessed on \today)}) is a \emph{CWE-89: Improper Neutralization of Special Elements used in an SQL Command ('SQL Injection')}\footnote{CWE-$89$ details available at \url{https://cwe.mitre.org/data/definitions/89.html} (Accessed on \today)} according to the CWE taxonomy.
+For instance,  the \texttt{Fix CVE-2014-1608: mc\_issue\_attachment\_get SQL injection} 
+(\texttt{00b4c17}\footnote{CVE-$2014$-$1608$ details available at 
+\url{https://github.com/mantisbt/mantisbt/commit/00b4c17088fa56594d85fe46b6c6057bb3421102} 
+(Accessed on \today)}) is a \emph{CWE-89: Improper Neutralization of Special Elements used 
+in an SQL Command ('SQL Injection')}\footnote{CWE-$89$ details available at \url{https://cwe.mitre.org/data/definitions/89.html} 
+(Accessed on \today)} according to the CWE taxonomy.
 We were able to classify a total of $866$ patches using the CWE taxonomy
-where the CWEs for $536$ patches were automatically scraped from the National Vulnerability Dataset (NVD); while for the other $330$ patches, the CWEs were manually chosen by the authors following the \emph{Research Concepts} CWE's list.
+where the CWEs for $536$ patches were automatically scraped from the National 
+Vulnerability Dataset (NVD); while for the other $330$ patches, the CWEs were manually 
+chosen by the authors following the \emph{Research Concepts} CWE's list.
 
@@
 characteristics of different projects.
+
+\subsection{Why Bettter Code Hub?}
 

### 2021-01-26 response/response.tex
more fixes

@@
 
-We added Marco's work to the related work.
+We mention Marco's work in our related work:
+\begin{quote}
+    Recent work,
+proposed a new maintainability model to 
+measure fine-grained code changes by adapting/extending the BCH model[diBiase2019].
+The present work uses the same model (SIG-MM), considers more guidelines and focuses 
+solely on evaluating the impact of security patches on software maintainability.
+\end{quote}
 
@@
 Conference on Software Maintenance and Evolution (ICSME) in 2019.
+
+[diBiase2019] M. di Biase and A. Rastogi and M. Bruntink and A. van 
+Deursen. The Delta Maintainability Model: Measuring Maintainability 
+of Fine-Grained Code Changes. IEEE/ACM International Conference 
+on Technical Debt (TechDebt) in 2019.
 
@@
 risks associated with their patches.
-
-SonarQube uses 4 metrics of the set of metrics we are using and 
-also rates maintainability. In Section 3 (\textit{Why Bettter Code Hub?}), 
-we compare the results between BCH and SonarQube. However, 
-we arguee that BCH performs a more complete analysis.
 
@@
         know what you are trying to say so I can't make a suggestion.
+    \end{itemize}
+\end{revcomment}
+
+\Us Thank you for raising this issue. In the sentence ``The codebases that resemble to the commits of our datasets 
+are collected before the analysis performed by BCH.'', we mean that the codebases of each commit in 
+our dataset were analyzed by BCH. We improved the text for:
+
+\begin{quote}
+    BCH evaluates the codebase available in the default branch of a GitHub project. 
+  We created a tool that pulls the codebase of each commit of our dataset 
+  to a new branch; it sets the new branch as the default branch; it runs 
+  the BCH analysis in the codebase; and, finally saves the BCH metrics
+  results.
+\end{quote}
+
+\begin{revcomment}[2.9]
+    Section 3\\
+    \begin{itemize}
         \item Please explain your repeatable methodology to "clean the dataset and select the most relevant and 
@@
 
-\Us
+\Us As explained in comment $2.6$, the $1282$ commits refer only to the Ponta et al. dataset (a total of $624$ 
+patches). In total, we analyzed $1300$ patches. However, BCH as some limitations such as
+analyzing projects with very large codebases and was uncapable to analyze some patches. In addition,
+we also detected floss-refactorings which we decided not to consider. All of this is explained
+in more detail later in Section 3.4.
 
-\begin{revcomment}[2.9]
+\begin{quote}
+    Security patches can be performed through one commit, 
+several consecutive commits, or commit(s) interleaved with more 
+programming activities (floss-refactoring). This study considers 
+mainly single-commit patches. Yet, a small percentage of data points 
+had more than one commit involved in the patch. We manually 
+inspected these cases and $23$ floss-refactorings were identified 
+and disregarded since it would not be fair to measure the 
+maintainability of these cases where other programming activities 
+are involved.
+
+For projects with large codebases, the results retrieved for 
+\emph{Keep Your Codebase Small} were not viable because they 
+retrieved values above the limit set by BCH ($20$ Person-years). The 
+\emph{Automated Tests} guideline was also not considered since the 
+tool does not contemplate two of the most important techniques to 
+security testing: vulnerability scanning and penetration testing. 
+Instead, it only contemplates unit testing. Due to BCH limitations, 
+we detected a few data points that retrieved incorrect overall 
+calculations. Those data points were disregarded from the study.
+\end{quote}
+
+Those are the cases that were tossed from the initial $1300$ patches.
+
+Regarding the second question, thank you for pointing out ... (still working on this)
+
+\begin{revcomment}[2.10]
     Section 3\\
@@
 
-\Us
-
-\begin{revcomment}[2.10]
-    Section 3\\
-    \begin{itemize}
-        \item Section 3.2 is confusing enough that I'm not sure of the analysis - 
-        I will need to review the revision that you would submit to gain confidence.  
-        Are you saying you want your non-security commits (regular commits) to be about 
-        the same size as a given security commit so you are using this criteria to choose 
-        your regular commits?  Justify this reasoning and make the methodology clear.  
-        I would have expected that you randomly collected 1300 non-security commits to compare 
-        with the 1300 security commits. Why constrain to make them be of "similar" size?  
-        Why not allow the characteristics of regular versus security commits to be what 
-        they are without such curating?
-    \end{itemize}
-\end{revcomment}
-
-\Us
+\Us Yes, for each security commit we collected a random regular change from the same project.
 
@@
 
-\Us
+\Us Yes, we are looking for regular changes with the same size as security commits. We 
+have collected a baseline of random changes only (no size restrictions) in the past and 
+performed the same evaluation. We now present the results for both baselines: \textit{size-baseline}, where we consider
+size appproximation; and, \textit{random-baseline}, where we consider all the regular
+changes characteristics.
 
@@
 
-\Us
-
-\begin{revcomment}[2.13]
-    Section 3\\
-    \begin{itemize}
-        \item Since SIG seems to be "just another consulting organization" - I'd like 
-        to see about the acceptance of the 10 guidelines in the software engineering 
-        discipline.  I would guess you could find peer-reviewed references to support 
-        these guidelines, or even books by Martin Fowler.   That has more credibility 
-        than the reference by the SIG consulting group. For example on Page 9, line 42 
-        you cite McCabe Complexity --  that has more credibility than a consulting 
-        organizations guidelines that they use to make money on their tool.  Tell us more 
-        about the BCH data experience since that is your comparison point for compliance.
-    \end{itemize}
-\end{revcomment}
-
-\Us
+\Us (I'll do this in the end.)
 
@@
 
-\Us
+\Us working on this
 
@@
 
-\Us until here
+\Us Due to BCH limitations, $308$ data points were disregarded.
 
@@
 
-\begin{revcomment}[2.18]
+\begin{revcomment}[2.19]
     Section 4\\
@@
 
-\begin{revcomment}[2.18]
+\begin{revcomment}[2.20]
     Section 4\\
@@
 
-\begin{revcomment}[2.19]
+\begin{revcomment}[2.21]
     Section 4\\
@@
 
-\begin{revcomment}[2.20]
+\begin{revcomment}[2.22]
     Section 5\\
@@
 
-\begin{revcomment}[2.20]
+\begin{revcomment}[2.23]
     Section 6\\
@@
 
-\Us
+\Us We kindly thank you the reviewer for the comments. We added a sub Section to 
+the paper called \textit{Why Better Code Hub?}. We compare BCH with other tools
+that perform code analysis and motivate why we find BCH to be the best fit for us.
 

### 2021-01-26 paper/emse20.tex
more fixes

@@
 their software applications for more than 20 years. Their 
-models are scientifically proven and certified~\cite{5609747,6113040,baggen2012}.
+models are scientifically proven and certified~\cite{4335232,5609747,6113040,baggen2012}.
 
@@
 \begin{enumerate}
-	\item Combine the datasets from related work that classifies
+	\item Combine the datasets from related work that classify
 	the activities of developers addressing security-oriented 
-	patches~\cite{Reis:2017:IJSSE,10.1109/MSR.2019.00064}.
-%
-\item After combining both datasets, we removed the duplicates
-and filtered out every floss-refactoring.
+  patches~\cite{reis2017secbench,10.1109/MSR.2019.00064}. The
+  duplicated patches were tossed (only $2$ were found).
 %
@@
 	the repository, sha key of the vulnerable version, sha key of 	
-	the fixed version and codebases) from the combined dataset 	
+	the fixed version) from the combined dataset 	
 	containing $1300$ security patches collected from open-source 	
@@
 %
-  \item A baseline of regular commits was randomly collected (i.e., for 
-  each security commit a random commit from the same project was selected) from 
-	the list of projects of the final dataset to evaluate the impact 
-	of regular commits on the maintainability of open-source 
+  \item Two baselines of regular changes were collected:
+  \textit{random-baseline} (for each security commit, a random change 
+  was collected from the same project) and \textit{size-baseline} (for 
+  each security commit, a random change with the same size
+  was collected from the same project). Our goal is to evaluate the impact 
+	of regular changes on the maintainability of open-source 
 	software.
 %
-  \item Use the Software Improvement Group's\footnote{SIG's website: 
-  \url{https://www.sig.eu/} (Accessed on \today{})} web-based source 
-  code analysis service \emph{Better Code Hub} (BCH)\footnote{BCH's 
-  website: \url{https://bettercodehub.com/} (Accessed on \today{})} 
+  \item Use the Software Improvement Group (SIG)'s web-based source 
+  code analysis service \emph{Better Code Hub} (BCH)
   to quantify maintainability for both security and regular commits. 
@@
 GitHub projects. The combined dataset integrates two different
-works: SecBench~\cite{Reis:2017:IJSSE} and Ponta et al.~\cite{10.1109/MSR.2019.00064}.
+works: SECBENCH~\cite{reis2017secbench,Reis:2017:IJSSE} and Ponta et al.~\cite{10.1109/MSR.2019.00064}.
 
@@
 to test and assess the performance of static analysis 
-tools~\cite{Reis:2017:IJSSE} since using hand-seeded test cases or 
+tools~\cite{reis2017secbench,Reis:2017:IJSSE} since using hand-seeded test cases or 
 mutations can lead to misleading assessments of the capabilities of 
 the tools~\cite{just2014mutants}. The study yielded a dataset of 
-$676$ patches for $16$ different security vulnerability types, dubbed as SecBench. 
+$676$ patches for $16$ different security vulnerability types, dubbed as SECBENCH. 
 The vulnerability types are based on the OWASP Top $10$ of 
@@
 %
-
 Ponta et al. ($2019$) tracked the \url{pivotal.io} website for 
@@
 commits in the repositories. This technique yielded a dataset of 
-$624$ patches~\cite{10.1109/MSR.2019.00064} and $1282$ commits. 
-One patch can have multiple commits 
-assigned. To fit the dataset in our methodology, we located the 
+$624$ patches~\cite{10.1109/MSR.2019.00064} and $1282$ commits---one 
+patch can have multiple commits assigned. To fit the dataset in our methodology, we located the 
 first and last commits used to patch the vulnerability. For these 
@@
 dataset were analyzed using the BCH toolset to calculate their 
-maintainability reports. Due to the limitations of BCH, in particular, 
-lack of language support and project size, we had
-to clean the dataset and select the most relevant and compatible projects
-with BCH. The final dataset used in this 
+maintainability reports. Due to the limitations of BCH (in particular, 
+lack of language support and project size) and the presence of
+floss-refactorings, $331$ patches were tossed---explained in 
+more detail in Section~\ref{sec:main_analysis}.
+The final dataset used in this 
 paper comprises $969$ security patches from $260$ projects. We used the 
@@
 (Accessed on \today)} according to the CWE taxonomy.
-We were able to classify a total of $866$ patches using the CWE taxonomy
+We were able to classify a total of $866$ patches \Sofia{not sure why i did not classified all of them}using the CWE taxonomy
 where the CWEs for $536$ patches were automatically scraped from the National 
@@
 %
-\subsection{Security vs. Baseline Commits}
-%
-Previous studies attempted to measure the impact of regular commits 
+\subsection{Security Patches vs. Regular Changes}
+%
+Previous studies attempted to measure the impact of regular changes 
 on open-source software maintainability~\cite{HEGEDUS2018313}. 
 However, there is no previous work focused on comparing the impact 
-of security patches with regular commits on maintainability.
-We analyze the maintainability of regular commits (i.e., commits not 
-related to security patches) and use them as a baseline.
-The baseline dataset is generated from the security commits dataset. 
-As for the security patches, for each regular fix or refatoring, we also need the commit performing the regular change (\emph{sha-reg}) and version of 
-the software before the change (\emph{sha-reg-p}). For each security patch, 
-a \emph{diff} with an approximate size was fetched to a regular 
-change. Due to the complexity of some patches, it was not possible 
+of security patches with regular changes on maintainability---only
+with bug-fixes~\cite{10.1145/3133956.3134072}.
+We analyze the maintainability of regular changes---changes not 
+related to security patches---and, use them as a baseline.
+The baseline dataset is generated from the security commits dataset, i.e.,
+for each security commit in the dataset we collect a random regular
+change from the same project. We created two different baselines: 
+\textit{random-baseline},
+considering random changes and all their characteristics; and,
+\textit{size-baseline}, considering also random changes
+but with the same size as security patches.
+
+\subsubsection{Size-Baseline}
+
+As for the security patches, for each regular change or refactoring, we also 
+need the commit performing the regular change (\emph{sha-reg}) and version of 
+the software before the change (\emph{sha-reg-p}). For each security patch, a 
+\emph{diff} with an approximate 
+size was fetched to a regular change. Due to the complexity of some patches, it was not possible 
 to find patches with the exact number of added and deleted lines. 
@@
 \item Selects a random commit from the project (\emph{sha-reg}).
-\item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) and its parent (\emph{sha-reg-p}).
+\item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) 
+and its parent (\emph{sha-reg-p}).
 \item Checks if the size from the regular change has the 
@@
 $10$ attempts to search for a change with the same size. We 
-originate the regular commits from the security commits to ensure 
+originate the regular changes from the security commits to ensure 
 that differences in maintainability are not a consequence of 
@@
 
+\subsubsection{Random-Baseline}
+
+As for the security patches, for each regular change or refactoring, we also 
+need the commit performing the regular change (\emph{sha-reg}) and version of 
+the software before the change (\emph{sha-reg-p}). For each security patch, 
+a random commit is selected from the the project (\emph{sha-reg}).
+
 \subsection{Why Bettter Code Hub?}
 
-\subsection{Maintainability Analysis}
+SIG---the company behind BCH--has been helping business and technology 
+leaders drive their organizational 
+objectives by fundamentally improving the health and security of 
+their software applications for more than $20$ years. The
+inner-working of their SIG-MM model---which is the one behind BCH---was 
+scientifically proven~\cite{4335232,5609747,6113040,baggen2012}.
+BCH not only considers security on their maintainability model~\cite{6616351}
+but also integrates a total of $10$ different code metrics; and, 
+code metrics were empirically validated in previous 
+work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
+
+As BCH there are other tools that do the same kind of code analysis.
+Two examples are Kiuwan and SonarCloud. Kiuwan is not free, their 
+inner-workings do not seem to be published anywhere and there is
+no clear list of metrics available---we inspected their datasheet.
+SonarCloud is free but uses only $4$ of the guidelines (duplication,
+size, complexity and code smells) that BCH considers. It also scores 
+maintainability. In the following table, we
+present a brief comparison between BCH and SonarCloud:
+
+\Sofia{Currently building a script to retrieve the SonarCloud results}
+
+\subsection{Maintainability Analysis}\label{sec:main_analysis}
 
@@
 presented in previous work on the maintainability of energy-oriented 
-fixes~\cite{8919169}. As said before, the web-based source code 
+fixes~\cite{8919169}. The inner-workings of BCH were proposed
+originally in 2007~\cite{4335232} and suffer refinements later~\cite{5609747,6113040,baggen2012}. 
+As said before, the web-based source code 
 analysis service \emph{Better Code Hub} (BCH) is used to collect the 
@@
 \scriptsize
-	\caption{Guidelines to produce maintainable code.}
+	\caption{Guidelines to produce maintainable code. \Sofia{Still need to add more references}}
 \begin{tabular}{L{2cm}L{4.5cm}L{4cm}}
@@
 
-Aiming to analyze the impact of security patches, we use BCH to compute the maintainability of two different 
-versions of the project (cf. Figure~\ref{fig:commit}):
+Aiming to analyze the impact of security patches, we use BCH to compute 
+the maintainability of two different versions of the project 
+(cf. Figure~\ref{fig:commit}):
 
@@
 Instead, it only contemplates unit testing. Due to BCH limitations, 
-we detected a few data points that retrieved incorrect overall 
+we detected $308$ data points that retrieved incorrect overall 
 calculations. Those data points were disregarded from the study.
@@
 of patches. Instead, statistical tests were used to find the metrics that
-have the potential to change significantly after patches. Studying the evolution 
+have the potential to change significantly after patches.
+
+Researchers
+performed a large-scale empirical study to understand the characteristics of security patches
+and their differences against security bug fixes~\cite{10.1145/3133956.3134072}.
+The main findings were that security patches are smaller and less complex compared
+to bug fixes and usually performed at function-level.
+
+Studying the evolution 
 of maintainability issues during the development of Android apps, Malavolta et al. ($2018$)~\cite{8530041}
@@
 measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
-The present work uses the same model (BCH) and focuses 
+The present work uses the same model (SIG-MM), considers more guidelines and focuses 
 solely on evaluating the impact of security patches on software maintainability.

### 2021-01-31 response/response.tex

@@
     these guidelines do not imply increasing the maintainability of the 
-    code, since the tools has no possibility to measure "maintainability".
+    code, since the tools has no possibility to measure ``maintainability''.
 
@@
 \Us Thank you for your positive feedback on the paper presentation and replication
-package. BCH is not a static analysis tool instead it checks codebases for 
-compliance with the ten guidelines. The set of guidelines presented in our paper
-is the same published by Software Improvement Group (SIG) in their ebook called
-"Building Maintainable Software: Ten Guidelines for Future-Proof Code"
+package.
+
+We would like to rebut your concern about Better Code Hub (BCH)'s guidelines not 
+being related to maintainability. First, as mentioend in the paper, BCH checks GitHub 
+codebases against 10 engineering guidelines as devised by Software Improvement Group 
+(SIG). SIG has devised these guidelies after many years of experience: analyzing more 
+than 15 million lines of code every week, SIG maintains the industry’s largest 
+benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
+is the only lab in the world certified by TÜViT to issue ISO 25010 certificates.
+
+Second, each of the guidelines in isolation is pretty simple to understand, but, when 
+combined with the others, and, when evaluated as a whole, it becomes easy to see why 
+they would comprise the definition of good software. Furthermore, the relation of the 
+guidelines to maintainability is well detailed in the SIG's book ``Building Maintainable 
+Software: Ten Guidelines for Future-Proof Code''
 \footnote{Available here: https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
-According to the authors, all the guidelines were infered from analyzing
-hundres of real-world systems and measure maintainability.
+In fact, the underlying maintainability model for BCH is documented with several
+(non-academic) reports linked at this page:
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/} as 
+well as validated by the academic community in the following paper:
+
+\begin{verbatim}
+Heitlager, Ilja, Tobias Kuipers, and Joost Visser. 
+``A practical model for measuring maintainability.'' 
+6th international conference on the quality of information and 
+        communications technology (QUATIC 2007). 
+IEEE, 2007.
+\end{verbatim}
+
+BCH's compliance criterion is derived from the requirements for 4 star level maintainability (cf. ISO 25010). 
+The concrete values of the thresholds used in BCH are documented also in our book Building Maintainability 
+Software https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/.
+
+Further, the methodology for arriving at the thresholds is documented in
+the papers you already point out. SIG performs the threshold calibration
+yearly on a proprietary data set to satisfy the requirements of TUViT to
+be a certified measurement model. However, please note that this data set 
+cannot be shared due to non-disclosure agreements between SIG and its 
+clients.
+
+All in all, we argue that BCH's guidelines are a good proxy for building maintainable software. This is 
+backed up by the observations of SIG in the field, as it was told us in personal communications:
+
+\begin{quote}
+We saw the value of a tool like BCH in helping in writing clean and maintainable code and most importantly, 
+it offers us so much more:
+
+Over time, the guidelines in BCH will become part of your coding standards, and the overall quality of your 
+work will improve, slowly, but surely.
+\end{quote}
+
+\textbf{PAREI AQUI.}
 

### 2021-01-31 response/response.tex
completing 1.1

@@
 \usepackage{subcaption}
-
+\usepackage{hyperref}
 \usepackage{epsfig}
@@
 We would like to rebut your concern about Better Code Hub (BCH)'s guidelines not 
-being related to maintainability. First, as mentioend in the paper, BCH checks GitHub 
+being related to maintainability. First, as mentioned in the paper, BCH checks GitHub 
 codebases against 10 engineering guidelines as devised by Software Improvement Group 
-(SIG). SIG has devised these guidelies after many years of experience: analyzing more 
+(SIG). SIG has devised these guidelines after many years of experience: analyzing more 
 than 15 million lines of code every week, SIG maintains the industry’s largest 
 benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is the only lab in the world certified by TÜViT to issue ISO 25010 certificates.
+is the only lab in the world certified by TÜViT to issue ISO 25010:2011 
+certificates\footnote{Information available here: 
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
 
@@
 Software: Ten Guidelines for Future-Proof Code''
-\footnote{Available here: https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
+\footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
 In fact, the underlying maintainability model for BCH is documented with several
@@
 
-BCH's compliance criterion is derived from the requirements for 4 star level maintainability (cf. ISO 25010). 
-The concrete values of the thresholds used in BCH are documented also in our book Building Maintainability 
-Software https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/.
+BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
+The concrete values of the thresholds used in BCH are documented also in their book Building Maintainability 
+Software \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
 
-Further, the methodology for arriving at the thresholds is documented in
-the papers you already point out. SIG performs the threshold calibration
+Further, the methodology for arriving at the thresholds is documented in the following 
+literature:
+
+\begin{verbatim}
+    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
+    ``Deriving metric thresholds from benchmark data.'' 
+    IEEE International Conference on Software Maintenance (ICSME 2010). 
+    IEEE, 2010.
+
+    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
+    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
+    2011 Joint Conference of the 21st International Workshop on Software 
+    Measurement and the 6th International Conference on Software Process 
+    and Product Measurement. 
+    IEEE, 2011.
+
+    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
+    ``Standardized code quality benchmarking for improving software maintainability.'' 
+    Software Quality Journal.
+    Springer, 2011.
+
+\end{verbatim}
+
+SIG performs the threshold calibration
 yearly on a proprietary data set to satisfy the requirements of TUViT to
@@
 
-We clarified the difference between BCH and static analysis tools in
-the Section 1 (Introduction) and Section 3 (Methodology).
+% We clarified the difference between BCH and static analysis tools in
+% the Section 1 (Introduction) and Section 3 (Methodology).
 
-\begin{quote}
-    Introduction\\
-    ...\\
-    Static analysis tools (SATs) have been built to automatically detect software 
-    vulnerabilities (e.g., FindBugs, Infer and more). Developers
-    use those tools to locate the issues in the code. However,
-    while performing the patches to those issues, SATs are not 
-    capable of providing information on the quality of the patch.
-    Improving software security is not a trivial task and 
-    requires implementing patches that might affect software 
-    maintainability. Our hypothesis is that some of these patches may 
-    have a negative impact on the software maintainability and, 
-    possibly, even be the cause of the introduction of new 
-    vulnerabilities---harming software reliability and introducing 
-    technical debt. Research found that $34\%$ of the security patches 
-    performed introduce new problems and $52\%$ are incomplete and do not 
-    fully secure systems[Li2017]. Therefore, in this paper, 
-    we present an empirical study on the impact of patches of 
-    vulnerabilities on software maintenance across open-source software.
-    We arguee that tools that assess these type of metrics may
-    complement SATs with valuable information to help the developer
-    understand better the risk of its patch.\\
-    ...
-\end{quote}
+% \begin{quote}
+%     Introduction\\
+%     ...\\
+%     Static analysis tools (SATs) have been built to automatically detect software 
+%     vulnerabilities (e.g., FindBugs, Infer and more). Developers
+%     use those tools to locate the issues in the code. However,
+%     while performing the patches to those issues, SATs are not 
+%     capable of providing information on the quality of the patch.
+%     Improving software security is not a trivial task and 
+%     requires implementing patches that might affect software 
+%     maintainability. Our hypothesis is that some of these patches may 
+%     have a negative impact on the software maintainability and, 
+%     possibly, even be the cause of the introduction of new 
+%     vulnerabilities---harming software reliability and introducing 
+%     technical debt. Research found that $34\%$ of the security patches 
+%     performed introduce new problems and $52\%$ are incomplete and do not 
+%     fully secure systems[Li2017]. Therefore, in this paper, 
+%     we present an empirical study on the impact of patches of 
+%     vulnerabilities on software maintenance across open-source software.
+%     We arguee that tools that assess these type of metrics may
+%     complement SATs with valuable information to help the developer
+%     understand better the risk of its patch.\\
+%     ...
+% \end{quote}
 
-[Li2017] F. Li and V. Paxson. A Large-Scale Empirical Study of Security Patches. 
-A Large-Scale Empirical Study of Security Patches in 2017.
+% [Li2017] F. Li and V. Paxson. A Large-Scale Empirical Study of Security Patches. 
+% A Large-Scale Empirical Study of Security Patches in 2017.
 
@@
 
-\Us Thank you for suggesting Marco's paper. Our paper uses a model 
+\Us Thank you for suggesting DiBiase's paper. Our paper uses a model 
 for measuring maintainability that was published previously by one of the co-authors
 of this paper [Cruz2019]. The entire methodology and calculations 
-were actually validated by Marco.
+were actually validated by DiBiase.
 
@@
 changes using 5 of the same guidelines we already use. The SIG-MM model in 
-Marco's paper is the model behind Better Code Hub.
+DiBiase's paper is the model behind Better Code Hub.
 

### 2021-01-31 response/response.tex
improving 1.2

@@
 
-\Us Thank you for suggesting DiBiase's paper. Our paper uses a model 
-for measuring maintainability that was published previously by one of the co-authors
-of this paper [Cruz2019]. The entire methodology and calculations 
-were actually validated by DiBiase.
+\Us Thank you for suggesting the DiBiase's paper. 
+We evaluate the impact of security patches in software 
+maintainability using an alternative maintainability model [Cruz2019].
 
-The model created by Marco focus on measuring maintainability of fine-grained 
-changes using 5 of the same guidelines we already use. The SIG-MM model in 
-DiBiase's paper is the model behind Better Code Hub.
+DiBiase's work was developed in paralell with [Cruz2019]. As
+the outcome of a master thesis, it only considers $5$ of the same 
+guidelines we use. Both models build a new maintainability model
+over the SIG-MM model---the model behind BCH.
 
-We mention Marco's work in our related work:
+Our model was validated by DiBiase in several discussions between 
+the authors of both papers.
+
+We also mention DiBiase's work in our \textit{Related Work} section:
 \begin{quote}

### 2021-01-31 response/response.tex
improving 1.5

@@
 We have already a dataset of security patches (vulnerable versions
-and safe versions) and our goal is to understand what 
+and non-vulnerable versions) and our goal is to understand what 
 is the impact of those patches on the maintainability
 guidelines/metrics. We arguee that these guidelines/metrics
-can in the future complement static analysis tools 
+may in the future complement static analysis tools 
 by assiting developers with more information on the 
@@
 explained before Better Code Hub calculates maintainability metrics.
-Our goal was to measure the impact of security patches on software 
-maintainability. For future research, it would be interesting to 
-understand if the results of these metrics can actually help 
+Our goal is to measure the impact of security patches on software 
+maintainability and not finding vulnerabilities---vulnerabilities
+are already located. Thus, for now, 
+we do not need to leverage static analysis tools.
+
+For future research, it would be interesting to 
+understand how BCH can be fused with a static analysis tool 
+and how the BCH results can help 
 security engineers assess the risk of their patches and guide 
-them on performing better patches. However, in this work, 
-we only focus on assessing software maintainability.
+them on performing better patches.
 

### 2021-01-31 response/response.tex
improving 2.15
improving 2.23

@@
 
-\Us Due to BCH limitations, $308$ data points were disregarded.
+\Us Due to BCH limitations, in particular, lack of language support 
+and project size by BCH, $308$ data points were disregarded.
 
@@
 
-\Us
+\Us We agree with the reviewer. We built this baseline following the concerns of
+previous reviewers with our random baseline. Reviewers were afraid that 
+comparing regular changes with security changes with very different sizes---a scenario 
+that could happen---was unfair. It made sense to us, thus we created this baseline.
+Nevertheless, we also feel that considering size may be limitating the generalizability
+of the characteristics of regular changes. Thus, we added to RQ3 our initial random 
+baseline and completed the work by presenting now two baselines (size-baseline and random-baseline).
 

### 2021-02-01 response/response.tex
response to 2.17, 2.16 and 1st item of 2.13

@@
         \item Page 10, line 49 "mainly single-commit patches" … "small percentage 
-        of data points" - give us the exact numbers for each of these.  Page 11, 
-        line 38 - define "floss refactoring" and justify how these were objectively 
+        of data points" - give us the exact numbers for each of these.  
+        \item Page 11, line 38 - define "floss refactoring" and justify how these were objectively 
         and repeatably identified in your set. You say "these cases" which makes 
@@
 
-\Us
+\Us Our dataset integrates $89.3\%$ of single-commit patches and 
+$10.7\%$ of patches involving more than one commit. We replaced 
+line 49 by the following one:
+
+\begin{quote}
+    Only $10.7\%$ of the data points 
+of our dataset involve more than one commits, the other
+$89.3\%$ of the cases are single-commit patches.
+\end{quote}
+
+(working on 2nd item)
 
@@
 
-\Us
+\Us We mean that our study is an empirical evaluation of $969$ security patches
+and $969$ regular changes. We improved the sentence.
+
+\begin{quote}
+    This study evaluates a final total of $969$ security patches from $260$ 
+distinct open-source projects. 
+\end{quote}
 
@@
 
-\Us
+\Us We replaced overall by the number of patches where maintainability increases ($38.7\%$).
+
+\begin{quote}
+    Regarding the impact of security patches per guideline, we 
+observe that $38.7\%$ of patches have more cases where the impact is 
+positive---where maintainability increases.
+\end{quote}
 

### 2021-02-01 paper/emse20.tex
more fixes
response to 2.17, 2.16 and 1st item of 2.13

@@
 several consecutive commits, or commit(s) interleaved with more 
-programming activities (floss-refactoring). This study considers 
-mainly single-commit patches. Yet, a small percentage of data points 
-had more than one commit involved in the patch. We manually 
+programming activities (floss-refactoring). Only $10.7\%$ of the data points 
+of our dataset involve more than one commits, the other
+$89.3\%$ of the cases are single-commit patches. We manually 
 inspected these cases and $23$ floss-refactorings were identified 
@@
 sizes, we compute the mean difference, the median of 
-the difference, and the percentage of cases that reduce maintainability~\cite{graw:1992}.
+the difference, and the percentage of cases that reduce 
+maintainability~\cite{graw:1992}.
 %
@@
 
-This study evaluates a final total of $969$ security and baseline 
-commits from $260$ distinct open-source projects. This section 
+This study evaluates a final total of $969$ security patches from $260$ 
+distinct open-source projects. This section 
 reports and discusses the results for each research question.
@@
 Regarding the impact of security patches per guideline, we 
-observe that overall patches have more cases where the impact is 
+observe that $38.7\%$ of patches have more cases where the impact is 
 positive---where maintainability increases. However, we also 

### 2021-02-01 response/response.tex
removing textbf
polishing

@@
 
-\textbf{PAREI AQUI.}
-
 % We clarified the difference between BCH and static analysis tools in
@@
 
-DiBiase's work was developed in paralell with [Cruz2019]. As
-the outcome of a master thesis, it only considers $5$ of the same 
-guidelines we use. Both models build a new maintainability model
-over the SIG-MM model---the model behind BCH.
+DiBiase's work was developed in paralell with the one by Cruz et al [Cruz2019]. 
+Both models attempt to measure maintainability, but DiBiase's is not as complete 
+since it only considers $5$ of the same set of guidelines. Both models build a 
+new maintainability model over the SIG-MM model---the model behind BCH.
 
@@
 proposed a new maintainability model to 
-measure fine-grained code changes by adapting/extending the BCH model[diBiase2019].
+measure fine-grained code changes by adapting/extending the BCH model [diBiase2019].
 The present work uses the same model (SIG-MM), considers more guidelines and focuses 
@@
 is the impact of those patches on the maintainability
-guidelines/metrics. We arguee that these guidelines/metrics
+guidelines/metrics. We argue that these guidelines/metrics
 may in the future complement static analysis tools 

### 2021-02-01 response/response.tex
a few more polishing

@@
 
-\Us Thank your for your comment. In this paper, we do not
-try to use static analysis to detect security issues.
-We have already a dataset of security patches (vulnerable versions
-and non-vulnerable versions) and our goal is to understand what 
-is the impact of those patches on the maintainability
-guidelines/metrics. We argue that these guidelines/metrics
+\Us Thank your for your comment. Note, however, that the goal of our work it not 
+to use static analysis to detect security issues. We have already a dataset of 
+security patches (pairs of vulnerable versions and non-vulnerable versions) and 
+our objective is to understand what 
+the impact of those patches on the maintainability
+guidelines/metrics is. We argue that these guidelines/metrics
 may in the future complement static analysis tools 
@@
 
-\Us Thank you for your comment. As we explained before, while 
-addressing comment 1.2, both models assess maintainability.
-Delta Maintainability Model is an alternative to our approach 
-and only uses 5 guidelines of the set of guidelines our study considers.
+\Us Thank you for your comment. As we explained before, namely while addressing 
+comment 1.2, both models assess maintainability. However, the Delta Maintainability 
+Model only uses 5 guidelines of the set of guidelines our study considers. Therefore, 
+we consider it to be an alternative to our model, but a weaker one. This has been 
+discussed in the related work section.
 
@@
 
-\Us Thank you for the positive interest on the work. As we 
+\Us Thank you for the positive feedback on the work. As we 
 explained before Better Code Hub calculates maintainability metrics.

### 2021-02-01 response/response.tex
fix small things

--- 2021-02-01 response/response.tex
@@
 \begin{quote}
-    Recent work,
+    Recent work
 proposed a new maintainability model to 
@@
 
-\Us Thank you for the positive interest on the work. As we 
+\Us Thank you for the positive interest on the work.
+
+We would like to rebut the reviewer concern. As we 
 explained before Better Code Hub calculates maintainability metrics.
@@
 
-\Us We thank the reviewer for the suggestions. We addressed
-both of the issues in the abstract. For the later issue, we reported 
-the overall result for maintainability and results
-for the two of the guidelines with more negative impact on software 
+\Us We thank the reviewer for the suggestions. We replaced ``suspect'' by
+``hypothesize''. For the later issue, we added to the abstract 
+the results for maintainability and results
+for the $2$ guidelines with more negative impact on software 
 maintainability: software complexity and unit size.
+
+\begin{quote}
+    Results show evidence of a trade-off 
+between security and maintainability for $41.90\%$ of the cases, i.e., developers 
+may hinder software maintainability. Our analysis shows that $38.29\%$ of patches increased
+software complexity and $37.87\%$ of patches increased the percentage 
+of LOCs per unit. 
+\end{quote}
 
@@
 
-\Us We agree with the reviewer. We built this baseline following the concerns of
-previous reviewers with our random baseline. Reviewers were afraid that 
+\Us We agree with the reviewer. Previous reviewers were afraid that 
 comparing regular changes with security changes with very different sizes---a scenario 

### 2021-02-01 paper/emse20.tex
fix small things

@@
 1300 security-related commits. Results show evidence of a trade-off 
-between security and maintainability in $41.9\%$ of the cases, i.e., developers 
-may hinder software maintainability. More than $35\%$ of security 
-patches increased software complexity and the percentage 
+between security and maintainability for $41.90\%$ of the cases, i.e., developers 
+may hinder software maintainability. Our analysis shows that $38.29\%$ of patches increased
+software complexity and $37.87\%$ of patches increased the percentage 
 of LOCs per unit. The implications of our study 
@@
 based on the BCH's guidelines and measured the impact of energy-oriented fixes 
-on software maintainability. Recent work,
+on software maintainability. Recent work
 proposed a new maintainability model to 

### 2021-02-01 response/response.tex
improving 1.5

@@
 
-% We clarified the difference between BCH and static analysis tools in
-% the Section 1 (Introduction) and Section 3 (Methodology).
-
-% \begin{quote}
-%     Introduction\\
-%     ...\\
-%     Static analysis tools (SATs) have been built to automatically detect software 
-%     vulnerabilities (e.g., FindBugs, Infer and more). Developers
-%     use those tools to locate the issues in the code. However,
-%     while performing the patches to those issues, SATs are not 
-%     capable of providing information on the quality of the patch.
-%     Improving software security is not a trivial task and 
-%     requires implementing patches that might affect software 
-%     maintainability. Our hypothesis is that some of these patches may 
-%     have a negative impact on the software maintainability and, 
-%     possibly, even be the cause of the introduction of new 
-%     vulnerabilities---harming software reliability and introducing 
-%     technical debt. Research found that $34\%$ of the security patches 
-%     performed introduce new problems and $52\%$ are incomplete and do not 
-%     fully secure systems[Li2017]. Therefore, in this paper, 
-%     we present an empirical study on the impact of patches of 
-%     vulnerabilities on software maintenance across open-source software.
-%     We arguee that tools that assess these type of metrics may
-%     complement SATs with valuable information to help the developer
-%     understand better the risk of its patch.\\
-%     ...
-% \end{quote}
-
-% [Li2017] F. Li and V. Paxson. A Large-Scale Empirical Study of Security Patches. 
-% A Large-Scale Empirical Study of Security Patches in 2017.
-
 \begin{revcomment}[1.2]
@@
 
-\Us Thank you for the positive feedback on the work. As we 
-explained before Better Code Hub calculates maintainability metrics.
-Our goal is to measure the impact of security patches on software 
+\Us Thank you for the positive feedback on the work.
+
+We would like to rebut your concern on the tools and 
+metrics we are using. As we explained previously, in comment 1.1, 
+BCH checks GitHub codebases against 10 maintainability guidelines 
+as devised by Software Improvement Group (SIG). SIG has devised 
+these guidelines after many years of experience: analyzing more 
+than 15 million lines of code every week, SIG maintains the industry’s largest 
+benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
+is the *only* lab in the world certified by TÜViT to issue ISO 25010:2011 
+certificates\footnote{Information available here: 
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}
+The guidelines and their relation to maintainability is well 
+detailed in the SIG's book ``Building Maintainable 
+Software: Ten Guidelines for Future-Proof Code''
+\footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
+
+SIG also performs the threshold calibration
+yearly on a proprietary data set to satisfy the requirements of TUViT to
+be a certified measurement model.
+
+Note, that our goal is to measure the impact of security patches on software 
 maintainability and not finding vulnerabilities---vulnerabilities
-are already located. Thus, for now, 
+are already identified and located. Thus, for now, 
 we do not need to leverage static analysis tools.
 
-For future research, it would be interesting to 
+For future research, it is indeed interesting to 
 understand how BCH can be fused with a static analysis tool 
@@
 security engineers assess the risk of their patches and guide 
-them on performing better patches.
+them on performing better patches. But for now we only try
+to understand the impact of security patches on those metrics.
+
+All in all, we arguee that BCH is the right tool and enough
+to perform the current analysis.
 
@@
         \item Please explain your repeatable methodology to "clean the dataset and select the most relevant and 
-        compatible projects" (page 8, line 14) that brings the patches from 1282 to 969.  From 969 to 866 patches - 
+        compatible projects" (page 8, line 14) that brings the patches from 1282 to 969.  
+        \item From 969 to 866 patches - 
         you just could not classify?  Please explain.
@@
 
-\Us We replaced overall by the number of patches where maintainability increases ($38.7\%$).
+\Us We replaced overall by the number of patches where maintainability 
+increases ($38.7\%$).
 

### 2021-02-01 response/response.tex
improving 1.5
improving 2.23

@@
 \footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
-
-SIG also performs the threshold calibration
-yearly on a proprietary data set to satisfy the requirements of TUViT to
-be a certified measurement model.
 
@@
 
-\Us We agree with the reviewer. Previous reviewers were afraid that 
+\Us We agree with the reviewer. Our concern by presenting a total random baseline 
+was that 
 comparing regular changes with security changes with very different sizes---a scenario 
-that could happen---was unfair. It made sense to us, thus we created this baseline.
+that could happen---was unfair. 
 Nevertheless, we also feel that considering size may be limitating the generalizability

### 2021-02-01 response/response.tex
finished 2.9

@@
 
-Regarding the second question, thank you for pointing out ... (still working on this)
+Regarding the second question, thank you for pointing out the issue. 
+While manually inspecting the patches, we were not able to map
+the issues to any CWE ID due to the lack of quality information on the 
+problem and solution. Yet, we manually validated the list again and after discussion 
+we classified more $40$ cases. We also clarified this on the paper:
+
+\begin{quote}
+    We were able to classify a total of $906$ patches using the CWE taxonomy:
+the CWE's for $536$ patches were automatically scraped from the National 
+Vulnerability Dataset (NVD); while the other $370$ patches were manually classified by 
+the authors following the \emph{Research Concepts} CWE's list. A total 
+of $63$ patches were not classified because we were not able to map
+the issues to any CWE ID due to the lack of quality information on the 
+problem and solution.
+\end{quote}
 

### 2021-02-01 paper/emse20.tex
finished 2.9

@@
 (Accessed on \today)} according to the CWE taxonomy.
-We were able to classify a total of $866$ patches \Sofia{not sure why i did not classified all of them}using the CWE taxonomy
-where the CWEs for $536$ patches were automatically scraped from the National 
-Vulnerability Dataset (NVD); while for the other $330$ patches, the CWEs were manually 
-chosen by the authors following the \emph{Research Concepts} CWE's list.
+We were able to classify a total of $906$ patches using the CWE taxonomy:
+the CWE's for $536$ patches were automatically scraped from the National 
+Vulnerability Dataset (NVD); while the other $370$ patches were manually classified by 
+the authors following the \emph{Research Concepts} CWE's list. A total 
+of $63$ patches were not classified because we were not able to map
+the issues to any CWE ID due to the lack of quality information on the 
+problem and solution.
 

### 2021-02-01 response/response.tex
finished 2.13

@@
 
-(working on 2nd item)
+We define floss-refactoring before in the paragraph
+as you can see below. Regarding the methodology to detect floss-refactorings, 
+we manually inspected all the cases to ensure that 
+floss-refactorings were not being considered.
+
+\begin{quote}
+    The same approach is used to analyze the impact of regular 
+    changes. Security patches can be performed through one commit (single-commit); 
+    several consecutive commits (multi-commits); or, commit(s) interleaved with more 
+    programming activities (floss-refactoring). Only $10.7\%$ of the data points 
+    of our dataset involve more than one commit, the other
+    $89.3\%$ of the cases are single-commit patches. We manually 
+    inspected all the cases and identified $23$ floss-refactorings. 
+    These were disregarded since it would not be fair to measure the 
+    maintainability of these cases where other programming activities 
+    are involved.
+\end{quote}
 

### 2021-02-01 paper/emse20.tex
finished 2.13

@@
 The same approach is used to analyze the impact of regular 
-changes. Security patches can be performed through one commit, 
-several consecutive commits, or commit(s) interleaved with more 
+changes. Security patches can be performed through one commit (single-commit); 
+several consecutive commits (multi-commits); or, commit(s) interleaved with more 
 programming activities (floss-refactoring). Only $10.7\%$ of the data points 
-of our dataset involve more than one commits, the other
+of our dataset involve more than one commit, the other
 $89.3\%$ of the cases are single-commit patches. We manually 
-inspected these cases and $23$ floss-refactorings were identified 
-and disregarded since it would not be fair to measure the 
+inspected all the cases and identified $23$ floss-refactorings. 
+These were disregarded since it would not be fair to measure the 
 maintainability of these cases where other programming activities 

### 2021-02-02 response/response.tex
2.18

@@
 
-\Us working on this
+\Us Yes, we started with all the guidelines but after analyzing the BCH results 
+we found that some of the results of the \emph{Keep Your Codebase Small} guidelines were not viable
+because they retrieved values above the limit set by BCH ($20$ Person-years). We did not want 
+to lose more data points. Thus, we decided to not consider the guideline for all projects.
+
+We also did not consider the \emph{Automated Tests} guideline because
+this guideline does not contemplate vulnerability scanning and penetration testing
+which are the two most important testing techniques in security.
+
+We improved the paper in the following text:
+
+\begin{quote}
+    For projects with large codebases, the results retrieved for 
+\emph{Keep Your Codebase Small} were not viable because they 
+retrieved values above the limit set by BCH ($20$ Person-years). Thus, 
+we decided to not consider this guideline in our research. The 
+\emph{Automated Tests} guideline was also not considered since the 
+tool does not contemplate two of the most important techniques to 
+security testing: vulnerability scanning and penetration testing. 
+\end{quote}
 

### 2021-02-02 response/response.tex
changes

@@
 
-All in all, we arguee that BCH is the right tool and enough
+All in all, we argue that BCH is appropriate 
 to perform the current analysis.
@@
 
-\Us Thank you for bringing our attention to this paper. We integrated
+\Us Thank you for bringing our attention to this paper. We now discuss
 this paper in the Related Work section and used it to support some 
@@
 
-\Us Thank you for raising this concern. The right number of security patches is $1300$ ($624$ from Ponta et al. and 
+\Us Thank you for raising this issue. The right number of security patches is $1300$ ($624$ from Ponta et al. and 
 $676$ from Secbench). The $1330$ value is a typo, we fixed the issue. Regarding the second point, $1282$ commits is
@@
 
-Regarding the second question, thank you for pointing out the issue. 
-While manually inspecting the patches, we were not able to map
-the issues to any CWE ID due to the lack of quality information on the 
+Regarding the second question:
+while manually inspecting the patches, we were not able to map
+the issues to any CWE ID due to the lack of information on the 
 problem and solution. Yet, we manually validated the list again and after discussion 
@@
 
-\Us Yes, for each security commit we collected a random regular change from the same project.
+\Us Yes, for each security commit we collected a random regular change from the same project. We have clarified it 
+in the paper.
 
@@
 
-\Us Yes, we started with all the guidelines but after analyzing the BCH results 
+\Us We started with all the guidelines but after analyzing the BCH results 
 we found that some of the results of the \emph{Keep Your Codebase Small} guidelines were not viable

### 2021-02-02 response/response.tex
some fixes

@@
 while manually inspecting the patches, we were not able to map
-the issues to any CWE ID due to the lack of information on the 
-problem and solution. Yet, we manually validated the list again and after discussion 
-we classified more $40$ cases. We also clarified this on the paper:
-
-\begin{quote}
-    We were able to classify a total of $906$ patches using the CWE taxonomy:
-the CWE's for $536$ patches were automatically scraped from the National 
-Vulnerability Dataset (NVD); while the other $370$ patches were manually classified by 
-the authors following the \emph{Research Concepts} CWE's list. A total 
-of $63$ patches were not classified because we were not able to map
-the issues to any CWE ID due to the lack of quality information on the 
-problem and solution.
-\end{quote}
+the issue to any CWE with confidence due to the lack of quality
+information on the vulnerability/patch. We clarified this in the paper.
 

### 2021-02-02 paper/emse20.tex
2.18
some fixes

@@
 of $63$ patches were not classified because we were not able to map
-the issues to any CWE ID due to the lack of quality information on the 
-problem and solution.
+the issue to any CWE with confidence due to the lack of quality
+information on the vulnerability/patch.
 
@@
 fixes~\cite{8919169}. The inner-workings of BCH were proposed
-originally in 2007~\cite{4335232} and suffer refinements later~\cite{5609747,6113040,baggen2012}. 
+originally in 2007~\cite{4335232} and suffered refinements later~\cite{5609747,6113040,baggen2012}. 
 As said before, the web-based source code 
@@
 \emph{Keep Your Codebase Small} were not viable because they 
-retrieved values above the limit set by BCH ($20$ Person-years). The 
+retrieved values above the limit set by BCH ($20$ Person-years). Thus, 
+we decided to not consider this guideline in our research. The 
 \emph{Automated Tests} guideline was also not considered since the 

### 2021-02-02 response/response.tex
starting 2.18

@@
 
-\Us
+\Us We added an explanation of the Research Concepts list to the paper
+in a footnote:
+
+\begin{quote}
+    Research Concepts is a tree-view provided by the Common Weakness 
+    Enumeration (CWE) website that intends to facilitate research into 
+    weaknesses. It is organized 
+    according to 
+    abstractions of behaviors instead of how they can be detected, 
+    their usual location in code, and when they are introduced in the 
+    development life cycle. Available here: 
+    \url{https: //cwe.mitre.org/data/definitions/1000.html}
+\end{quote}
 

### 2021-02-02 paper/emse20.tex
starting 2.18

@@
 GitHub projects. The combined dataset integrates two different
-works: SECBENCH~\cite{reis2017secbench,Reis:2017:IJSSE} and Ponta et al.~\cite{10.1109/MSR.2019.00064}.
-
-Reis and Abreu 
-($2017$) mined open-source software aiming at the extraction of 
+works: Secbench~\cite{reis2017secbench,Reis:2017:IJSSE} and Ponta et al.~\cite{10.1109/MSR.2019.00064}.
+
+\textbf{Reis and Abreu 
+($2017$)} mined open-source software aiming at the extraction of 
 real---created by developers---patches of security vulnerabilities 
@@
 the tools~\cite{just2014mutants}. The study yielded a dataset of 
-$676$ patches for $16$ different security vulnerability types, dubbed as SECBENCH. 
+$676$ patches for $16$ different security vulnerability types, dubbed as Secbench. 
 The vulnerability types are based on the OWASP Top $10$ of 
@@
 can easily review the code used to fix the vulnerability.
-%
-Ponta et al. ($2019$) tracked the \url{pivotal.io} website for 
+
+\textbf{Ponta et al. ($2019$)} tracked the \url{pivotal.io} website for 
 vulnerabilities from $2014$ to $2019$. For each new vulnerability, 
@@
 Vulnerability Dataset (NVD); while the other $370$ patches were manually classified by 
-the authors following the \emph{Research Concepts} CWE's list. A total 
+the authors following the \emph{Research Concepts} CWE's list\footnote{Research Concepts list available at \url{https://cwe.mitre.org/data/definitions/1000.html}}. A total 
 of $63$ patches were not classified because we were not able to map
@@
 In \emph{RQ2}, we report/discuss the impact of security patches on
-software maintainability per weakness. We use the weakness definition
+software maintainability per weakness (CWE). We use the weakness definition
 and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section~\ref{sec:motivation}).
@@
 the impact of all patches grouped using the first-level weaknesses from
-the \emph{Research Concepts} list on CWE; and, \emph{(\ref{fig:pat}-b)} and  
+the \emph{Research Concepts}~\footnote{Research Concepts 
+ is a tree-view provided by the Common Weakness Enumeration (CWE) website 
+ that intends to facilitate research into weaknesses. It is organized 
+ according to 
+ abstractions of behaviors instead of how they can be detected, 
+ their usual location in code, and when they are introduced in the 
+ development life cycle. Available here: \url{https://cwe.mitre.org/data/definitions/1000.html}
+} list on CWE; and, \emph{(\ref{fig:pat}-b)} and  
 \emph{(\ref{fig:pat}-c)} present the impact on maintainability per sub-weakness of two 

### 2021-02-02 response/response.tex
fix to 2.9 and 2.18 finished

@@
 the issue to any CWE with confidence due to the lack of quality
-information on the vulnerability/patch. We clarified this in the paper.
+information on the vulnerability/patch for $103$ patches. 
+We clarified this in the paper.
+
+\begin{quote}
+    A total 
+of $103$ patches were not classified because we were not able to map
+the issue to any CWE with confidence due to the lack of quality
+information on the vulnerability/patch.
+\end{quote}
 
@@
 
+In Figure 7a, we present the entire distribution of the $969$ vulnerabilities per
+each node of the 1st-level of the Research Concepts tree.
+Looking at those results, we detected that the percentage of cases for CWE-707 ($30.4\%$) 
+and CWE-664 ($32.8\%$) were considerably higher in the dataset than the rest of CWEs.
+Thus, we decided to look at lower levels of behavior for both CWEs: CWE-707
+results are presented in Figure-7b and CWE-664 results are presented in Figure-7c.
+In Figure-7b, all the CWEs presented are sub levels of the CWE-707. The same
+happens for Figure-7c, but the CWEs are sub levels if the CWE-664.
+
+We added the previous numbers to the paper and clarified that we are considering lower levels
+of vulnerabilities.
+
+\begin{quote}
+    In \emph{RQ2}, we report/discuss the impact of security patches on
+    software maintainability per weakness (CWE). We use the weakness definition
+    and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section 2).
+    Figure 7 shows three different charts. Figure \emph{7-a}, presents
+    the impact of the $969$ patches grouped by the first level weaknesses from
+    the \emph{Research Concepts} list. While the Figures \emph{7-b} and
+    \emph{7-c} present the impact on maintainability for lower levels of 
+    weaknesses for the most prevalent weaknesses in Figure \emph{7-a}:
+    \emph{Improper Neutralization} (CWE-707) 
+    and \emph{Improper Control of a Resource 
+    Through its Lifetime} (CWE-664), respectively.
+
+    In Figure~7-\emph{a}, there is no clear evidence of the impact on 
+    maintainability per weakness. Yet, it is important to note that
+    overall there is a very considerable number of cases that hinder
+    maintainability---between $30\%$ and $60\%$. The CWE-707 and CWE-664 
+    weaknesses integrate the higher number of cases compared to the remaining 
+    ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
+    Thus, we present an analysis of their sub-weaknesses on 
+    Figure~7-\emph{b} and Figure~7-\emph{c}, respectively. 
+\end{quote}
+
 \begin{revcomment}[2.19]

### 2021-02-02 paper/emse20.tex
fix to 2.9 and 2.18 finished

@@
 (Accessed on \today)} according to the CWE taxonomy.
-We were able to classify a total of $906$ patches using the CWE taxonomy:
+We were able to classify a total of $866$ patches using the CWE taxonomy:
 the CWE's for $536$ patches were automatically scraped from the National 
@@
 the authors following the \emph{Research Concepts} CWE's list\footnote{Research Concepts list available at \url{https://cwe.mitre.org/data/definitions/1000.html}}. A total 
-of $63$ patches were not classified because we were not able to map
+of $103$ patches were not classified because we were not able to map
 the issue to any CWE with confidence due to the lack of quality
@@
 and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section~\ref{sec:motivation}).
-Figure~\ref{fig:pat} shows three different charts: \emph{(\ref{fig:pat}-a)}, shows
-the impact of all patches grouped using the first-level weaknesses from
-the \emph{Research Concepts}~\footnote{Research Concepts 
+Figure~\ref{fig:pat} shows three different charts. Figure \emph{\ref{fig:pat}-a}, presents
+the impact of the $969$ patches grouped by the first level weaknesses from
+the \emph{Research Concepts}\footnote{Research Concepts 
  is a tree-view provided by the Common Weakness Enumeration (CWE) website 
@@
  their usual location in code, and when they are introduced in the 
- development life cycle. Available here: \url{https://cwe.mitre.org/data/definitions/1000.html}
-} list on CWE; and, \emph{(\ref{fig:pat}-b)} and  
-\emph{(\ref{fig:pat}-c)} present the impact on maintainability per sub-weakness of two 
-weaknesses from \emph{(\ref{fig:pat}-a)}: \emph{Improper Neutralization} (CWE-707) and 
-\emph{Improper Control of a Resource 
+ development life cycle. The list is available here: \url{https://cwe.mitre.org/data/definitions/1000.html}
+} list. While the Figures \emph{\ref{fig:pat}-b} and
+\emph{\ref{fig:pat}-c} present the impact on maintainability for lower levels of 
+weaknesses for the most prevalent weaknesses in Figure \emph{\ref{fig:pat}-a}:
+\emph{Improper Neutralization} (CWE-707) 
+and \emph{Improper Control of a Resource 
 Through its Lifetime} (CWE-664), respectively.
@@
 overall there is a very considerable number of cases that hinder
-maintainability---between $30\%$ and $60\%$.
-The CWE-707 ($295$) and CWE-664 ($318$) 
+maintainability---between $30\%$ and $60\%$. The CWE-707 and CWE-664 
 weaknesses integrate the higher number of cases compared to the remaining 
-ones. Thus, we present an analysis of their sub-weaknesses on 
+ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
+Thus, we present an analysis of their sub-weaknesses on 
 Figure~\ref{fig:pat}-\emph{b} and Figure~\ref{fig:pat}-\emph{c}, respectively.
+
 Results shows that patching vulnerabilities may hinder 

### 2021-02-02 response/response.tex
improving 2.11

@@
 
-\Us Yes, we are looking for regular changes with the same size as security commits. We 
+\Us Yes, we are looking for regular changes with the same size as security commits
+from the same project. However, 
+for some cases, it was difficult to find the exact same size. Thus, we  
+searched for an approximation. Every 10 attempts of failling to get a commit with 
+the same size, we spanned the range size. We clear the methodology in the paper
+in Section 3.2 \textit{(Security Patches vs. Regular Changes)}.
+
+``I would have expected that you randomly collected 1300 non-security commits to compare 
+with the 1300 security commits.'': Although we started with 1300 commits, 
+$331$ cases were tossed due to BCH limitations. In addition, BCH
+takes substantial time to process this amount of cases. 
+Since the comparison is between 
+the final dataset of $969$ patches and the baselines. We produced d
+baselines with the final number of viable results obtained for the security 
+dataset.
+
+We 
 have collected a baseline of random changes only (no size restrictions) in the past and 
@@
 
-\Us
+\Us Yes, cases with low
 

### 2021-02-02 paper/emse20.tex
improving 2.11

@@
 
-\subsubsection{Size-Baseline}
-
-As for the security patches, for each regular change or refactoring, we also 
+\subsubsection{Random-Baseline}
+
+As for the security patches, for each regular change, we also 
 need the commit performing the regular change (\emph{sha-reg}) and version of 
 the software before the change (\emph{sha-reg-p}). For each security patch, a 
-\emph{diff} with an approximate 
-size was fetched to a regular change. Due to the complexity of some patches, it was not possible 
+random commit from the same project is selected (\emph{sha-reg}).
+\emph{sha-reg-p} is the commit performed previous to the 
+regular change, \emph{sha-reg}.
+
+\subsubsection{Size-Baseline}
+
+For the size-baseline, we also need to obtain the regular change (\emph{sha-reg})
+and the version of the software before the change (\emph{sha-reg-p}). 
+First, the \emph{diff}
+between the security patch and its parent is calculated.
+Then, a random commit/regular change from 
+the same project is selected (\emph{sha-reg}). The 
+\emph{diff} between \emph{sha-reg} and its parent (\emph{sha-reg-p})
+is calculated and compared to the \emph{diff} from the security patch.
+Due to the complexity of some patches, it was not possible 
 to find patches with the exact number of added and deleted lines. 
-Our search uses the security patch \emph{diff} to look for a 
-\emph{diff} of the same size. The algorithm integrates the following 
-steps:
-
-\begin{enumerate}
-\item Calculates the \emph{diff} between the security patch commit 
-(\emph{sha}) and its parent (\emph{sha-p}).
-\item Selects a random commit from the project (\emph{sha-reg}).
-\item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) 
-and its parent (\emph{sha-reg-p}).
-\item Checks if the size from the regular change has the 
-same/approximated size to the security patch.
-\end{enumerate}
-
-The pair of the random commit and its parent is accepted if the 
+Thus, we looked for an approximation.
+% The algorithm integrates the following steps:
+
+% \begin{enumerate}
+% \item Calculates the \emph{diff} between the security patch commit 
+% (\emph{sha}) and its parent (\emph{sha-p}).
+% \item Selects a random commit from the project (\emph{sha-reg}).
+% \item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) 
+% and its parent (\emph{sha-reg-p}).
+% \item Checks if the size from the regular change has the 
+% same/approximated size to the security patch.
+% \end{enumerate}
+
+The pair of the regular change (\emph{sha-reg}) and its parent (\emph{sha-reg-p}) is accepted if the 
 \emph{diff} size fits in the range size. This range widens every 
@@
 characteristics of different projects.
-
-\subsubsection{Random-Baseline}
-
-As for the security patches, for each regular change or refactoring, we also 
-need the commit performing the regular change (\emph{sha-reg}) and version of 
-the software before the change (\emph{sha-reg-p}). For each security patch, 
-a random commit is selected from the the project (\emph{sha-reg}).
 

### 2021-02-03 scripts/maintainability/cache.py
scripts used for merging cache

@@ -56,6 +56,9 @@ class Cache():
     
     def set_data(self, cache):
         self._data = cache
+    
+    def set_storage_path(self, path):
+        self.storage_path = path
 
     def get_value(self, key):
         """Return value for given key."""

### 2021-02-03 scripts/maintainability/merge_cache.py

@@ -1,0 +1,44 @@
+import argparse
+from os import listdir
+from os.path import isfile, join
+import json
+from pathlib import Path
+from zipfile import ZipFile, ZIP_DEFLATED
+
+from maintainability.cache import Cache
+
+def change_extension(file_path, extension):
+    return Path(file_path).stem + extension
+
+
+def merge_cache(cache, output):
+    cache_files = [f for f in listdir(cache) if isfile(join(cache, f)) and 'cache' in f]
+    caches = [Cache(join(cache, f)) for f in cache_files]    
+    merged_cache = caches[0]
+    for cache in caches[1::]:
+        merged_cache = merged_cache.merge_cache(cache)
+
+    merged_cache.set_storage_path(output)
+    print(merged_cache.storage_path)
+    merged_cache.save_data()
+    # json_filename = change_extension(output, '.json')
+    # print(json_filename)
+    # datas = json.dumps(merged_cache)
+    # with ZipFile(output, 'w', compression=ZIP_DEFLATED) as myzip:
+    #     myzip.writestr(json_filename, datas)
+
+
+
+if __name__ == "__main__":
+    
+    parser = argparse.ArgumentParser(description='Report results')    
+                        
+    parser.add_argument('-cache', type=str, metavar='cache folder', help='cache folder path')  
+    parser.add_argument('-output', type=str, metavar='output path', help='output folder path')      
+
+    args = parser.parse_args()
+
+    if args.cache != None and args.output != None:
+         merge_cache(cache=args.cache, output=args.output)
+    else:
+        print('Something is wrong. Verify your parameters')

### 2021-02-03 scripts/maintainability/create_reg_dataset.py
old script used to create random baseline

@@ -3,91 +3,44 @@
 
 import csv
 import random
-import requests
 
 from contexttimer import timer
-import click
-import time
 import pandas as pd
+import click
 
 from maintainability import ghutils, gitutils, log
 
 CLONAGE_DIR = "./tmp"
 
 @click.command()
-@click.option('--security-dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
+@click.option('--security-dataset', prompt=True, default="../baselines/db_regular_changes_random.csv",
               help="Input dataset of security commits.")
-@click.option('--regular-dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
+@click.option('--regular-dataset', prompt=True, default="../baselines/db_regular_changes_random.csv",
               help="Output dataset with regular commits.")
 def tool(security_dataset, regular_dataset):
     """CLI to collect random commits and analyze maintainability."""
     df = pd.read_csv(security_dataset)
-    rows = add_random_regular_commits(df, regular_dataset)
+    add_random_regular_commits(df, regular_dataset)
 
 @timer()
 def add_random_regular_commits(rows, regular_dataset):
     """Add refactoring commits in the dataset."""
     log.info(f"Will analyze {len(rows)} commits.")
-
-    for i, row in rows.iterrows():        
-        if 'random_commit' not in row.keys() and row['reg'] != 'FOUND':
+    for i, row in rows[~pd.notnull(rows['sha-reg'])][::-1].iterrows():        
+        if 'random_commit' not in row.keys():
             user = row['owner']
             project = row['project']
-            sha = row['sha']
-            sha_p = row['sha-p']
-
-            repo = ghutils.get_repo(user, project)
-            print(user, project, sha, sha_p)
-            diff_files = repo.compare(sha_p, sha).files
-            size = len(diff_files)
-            adds = sum([i.additions for i in diff_files])
-            dels = sum([i.deletions for i in diff_files])
-            print(user, project, sha, sha_p, adds, dels)
-
             with ghutils.GithubCloneRepoPersistent(user, project, CLONAGE_DIR):
-
-                commits = gitutils.get_all_commits('.')
-                
-                count = 0; step = 1; limit = 301
-                while count < limit:
-
+                try:
+                    commits = gitutils.get_all_commits('.')
                     regular_commit = random.choice(commits)
-                    regular_commit_parent = gitutils.resolve_commit('.', f"{regular_commit}^1")
-
[…]

### 2021-02-03 scripts/maintainability/create_size_reg_dataset.py
updating name of script to create size baseline

@@ -1,0 +1,93 @@
+"""Module with all the instrumentation used for SZZ.
+"""
+
+import csv
+import random
+import requests
+
+from contexttimer import timer
+import click
+import time
+import pandas as pd
+
+from maintainability import ghutils, gitutils, log
+
+CLONAGE_DIR = "./tmp"
+
+@click.command()
+@click.option('--security-dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
+              help="Input dataset of security commits.")
+@click.option('--regular-dataset', prompt=True, default="../dataset/db_release_regular_fixes.csv",
+              help="Output dataset with regular commits.")
+def tool(security_dataset, regular_dataset):
+    """CLI to collect random commits and analyze maintainability."""
+    df = pd.read_csv(security_dataset)
+    rows = add_random_regular_commits(df, regular_dataset)
+
+@timer()
+def add_random_regular_commits(rows, regular_dataset):
+    """Add refactoring commits in the dataset."""
+    log.info(f"Will analyze {len(rows)} commits.")
+
+    for i, row in rows.iterrows():        
+        if 'random_commit' not in row.keys() and row['reg'] != 'FOUND':
+            user = row['owner']
+            project = row['project']
+            sha = row['sha']
+            sha_p = row['sha-p']
+
+            repo = ghutils.get_repo(user, project)
+            print(user, project, sha, sha_p)
+            diff_files = repo.compare(sha_p, sha).files
+            size = len(diff_files)
+            adds = sum([i.additions for i in diff_files])
+            dels = sum([i.deletions for i in diff_files])
+            print(user, project, sha, sha_p, adds, dels)
+
+            with ghutils.GithubCloneRepoPersistent(user, project, CLONAGE_DIR):
+
+                commits = gitutils.get_all_commits('.')
+                
+                count = 0; step = 1; limit = 301
+                while count < limit:
+
+                    regular_commit = random.choice(commits)
+                    regular_commit_parent = gitutils.resolve_commit('.', f"{regular_commit}^1")
+
+                    if regular_commit_parent == None:
+                        continue
+
+                    diff_files_reg = repo.compare(regular_commit_parent, regular_commit).files
+
+                    size_reg = len(diff_files_reg)
+                    adds_reg = sum([i.additions for i in diff_files_reg])
+                    dels_reg = sum([i.deletions for i in diff_files_reg])
+
[…]

### 2021-02-03 scripts/stats/chart.py
updating scripts to consider both baselines

@@ -41,40 +41,46 @@ def save_report(path, name):
     plt.tight_layout()
     plt.savefig('{}/{}'.format(path, name))
 
-def main_comparison_chart(reports, df_sec, df_reg):   
+def main_comparison_chart(reports, dfs):   
         
-        test = pd.concat((tests.hypothesis_test(df_reg['diff'], 'reg'), 
-                            tests.hypothesis_test(df_sec['diff'], 'sec')),
+        df_sec, df_rand_reg, df_size_reg = dfs['security'], dfs['random'], dfs['size']
+
+        test = pd.concat((tests.hypothesis_test(df_rand_reg['diff'], 'random-reg'), 
+                            tests.hypothesis_test(df_size_reg['diff'], 'size-reg'),
+                            tests.hypothesis_test(df_sec['diff'], 'security')),
                             ignore_index=True)
                         
-        total_sec, total_reg = data.filter_results(df_sec), data.filter_results(df_reg)   
+        total_sec, total_rand_reg, total_size_reg = data.filter_results(df_sec), \
+                                                    data.filter_results(df_rand_reg[pd.notnull(df_rand_reg['diff'])]),  \
+                                                    data.filter_results(df_size_reg)    
         db_size = len(df_sec)  
+        db_rand_size = len(df_rand_reg[pd.notnull(df_rand_reg['diff'])])
 
-        rep = {'neg' : pd.Series([total_reg['neg']/db_size, total_sec['neg']/db_size]),
-                'neg_abs' : pd.Series([total_reg['neg'], total_sec['neg']]),
-                'pos' : pd.Series([total_reg['pos']/db_size, total_sec['pos']/db_size]),
-                'pos_abs' : pd.Series([total_reg['pos'], total_sec['pos']]),
-                'nul' : pd.Series([total_reg['nul']/db_size, total_sec['nul']/db_size]),
-                'nul_abs' : pd.Series([total_reg['nul'], total_sec['nul']]),
-                'type' : pd.Series(['Regular Change','Security Change'])}
+        rep = {'neg' : pd.Series([total_rand_reg['neg']/db_rand_size, total_size_reg['neg']/db_size , total_sec['neg']/db_size]),
+                'neg_abs' : pd.Series([total_rand_reg['neg'],  total_size_reg['neg'], total_sec['neg']]),
[…]

### 2021-02-03 scripts/maintainability/merge_cache.py
update merge

@@ -19,14 +19,7 @@ def merge_cache(cache, output):
         merged_cache = merged_cache.merge_cache(cache)
 
     merged_cache.set_storage_path(output)
-    print(merged_cache.storage_path)
     merged_cache.save_data()
-    # json_filename = change_extension(output, '.json')
-    # print(json_filename)
-    # datas = json.dumps(merged_cache)
-    # with ZipFile(output, 'w', compression=ZIP_DEFLATED) as myzip:
-    #     myzip.writestr(json_filename, datas)
-
 
 
 if __name__ == "__main__":

### 2021-02-03 response/response.tex
adding response to 2.19 and 2.20

@@
 
-\Us
+\Us In total, we inspected around $25$ cases whose maintainability difference/
+impact was zero to understand what was leading to a higher number of cases with no impact.
 
@@
 
-\Us Yes, cases with low
+\Us As we described in comment $2.19$, we manually inspected $25$ cases.
+Many of those cases were changes of small size (< $10$ LOCs) where
+refactorings were usually improving names of variables, functions, cases of type
+conversion, etc. Changes that have no impact on the metrics we are 
+evaluating.
+
+Our concern by presenting a total random baseline 
+was that comparing regular changes with security changes with very different sizes---a scenario 
+that could happen---was unfair. Thus, we chose to create this baseline.
+However, we also curated a total random baseline in the past and we now 
+present both baselines in the paper.
 
@@
 
-\Us We kindly thank you the reviewer for the comments. We added a sub Section to 
+\Us We kindly thank the reviewer for the comments. We added a sub Section to 
 the paper called \textit{Why Better Code Hub?}. We compare BCH with other tools

### 2021-02-03 paper/emse20.tex
updating baseline chart
RQ3 wip

@@
 \usepackage{algorithmic}
-\usepackage{graphicx}
+\usepackage{graphicx,subfigure}
 \usepackage{textcomp}
@@
 $9$/$10$ projects with more contributors follow strict contribution 
-policies for code standards. % For
-% instance, \texttt{laravel/framework} requires the \texttt{PSR-2
-% Standard}\footnote{PSR-2 Standard available at
-% \url{https://www.php-fig.org/psr/psr-2/} (Accessed on \today{})} that is a coding style guide for \texttt{PHP}.
+policies for code standards.
 
@@
 \textbf{Summary:}
-Although results did not yield statistical significance, we show preliminary evidence that researchers and developers ought to pay more attention to maintainability when fixing the following types of weaknesses: \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure
+Although results did not yield statistical significance, we show preliminary evidence that 
+researchers and developers ought to pay more attention to maintainability when fixing the 
+following types of weaknesses: \emph{Improper Input Validation (CWE-20)}, \emph{Information Exposure
 (CWE-200)}, \emph{Missing Release of Memory after Effective
@@
 
-% No statistical evidence was observed.
-% However, results should not be disregarded because they hint on how
-% the maintainability is impacted per weakness which highlights the weaknesses
-% that may need more attention from developers/researchers.
-%
-
 \textit{\textbf{RQ3: What is the impact of security changes versus regular 
 changes on the maintainability of open-source software?}}
-The impact of security and baseline changes on software maintainability 
-is presented in Figure~\ref{fig:secvsreg}.  
+The impact of security and regular changes on software maintainability 
+is presented in Figure~\ref{fig:secvsreg}. In this section, 
+we present a comparison of security patches with 
+two different baselines of regular changes: \emph{random-baseline},
+a dataset of random regular changes;
+and \emph{size-baseline}, a dataset of  
+regular changes with the same size as security 
+patches. Our hypothesis in this research question is
+that \emph{Security patches hinder more maintainability and 
+need more attention than regular changes on software maintainability}.
+
 We have seen, previously, a deterioration on software maintainability 
@@
   \centering
-	\includegraphics[width=0.7\textwidth]{figures/main_comparison.pdf}
- 	\caption{Maintainability difference for security and baseline patches.}
+  \includegraphics[width=0.8\textwidth]{figures/baseline.pdf}
+  \caption{Maintainability difference of security patches versus regular changes.}
   \label{fig:secvsreg}    
@@
 applying security patches.
-
-%
-% For both types of changes, there is a negative impact
-% on software maintainability---especially, for security changes.
-% The resulting p-value of the Paired Wilcoxon signed-rank test is $0.300$, i.e.,
-% conclusions regarding the regular changes are not statistically significant.
-% Yet, it is important to note that \emph{patching security vulnerabilities seem
-% to have a more negative impact on the maintainability of open-source software
-% than performing regular changes}.
-% This raises
-% a new trade-off when developers need to patch vulnerabilities on their projects
-% because they might be hindering the software maintainability or even introducing
-% new vulnerabilities.
 
@@
 	complex security features to implement.
-% 
-% \\\textit{\textbf{Poor Coding Practices Documentation:}}
-%   \Rui{too generic?}
-% 	Software documentation must be provided to developers with the
-% 	best practices to implement secure and maintainable software. Unmaintainable
-% 	code is difficult to test, analyze and re-use. Software companies must
-% 	do an effort to provide these artifacts to their developers, in order,
-% 	to avoid the introduction of new vulnerabilities while patching older ones.
 
@@
 
-\textit{\textbf{External Validity:}} The BCH tool uses private and open-source data to determine the thresholds for each guideline. We only analyze patches of open-source software.
+\textit{\textbf{External Validity:}} The BCH tool uses private and open-source data to 
+determine the thresholds for each guideline. We only analyze patches of open-source software.
 Thus, our findings may not extend to private/non-open source software. Different programming 
@@
 dataset comprises more commits in Java, i.e., the dataset may not be representative 
-of the population regarding programming languages.
-For both datasets, we a manual validation of the message of the commits was performed.
-Only commits in English were considered.
-Thus, our approach does not consider
+of the population regarding programming languages. For both datasets, we a manual validation 
+of the message of the commits was performed. Only commits in English were considered. Thus, our approach does not consider 
 patches in any other language but English.
@@
 
-% \section{Plan A}
-%
-% Please, pay more attention to the cases with !.
-%
-% \begin{itemize}
-	% \item How this is a problem or has an impact to the software project?
-	%   What is the message to developers? (clarify)
-	% \item Implications: Discuss that depending on the project, maybe security
-	%   is a first class citizen and should be fixed above all. Don't
-	%   just state that patch risk assessment should take maintainability into account.
-	%   Saying that security should be a priority requires maybe a user study
-	%   as evidence (check if the literature available is not enough).
-	%   \item Implications: Review the secure programming languages implication. Check
-	%   if the idea of fully secure exists. The idea should be more secure, not
-	%   fully secure.
-  % \item [!] Implications: \textit{I am not sure how the computer science curricula got mentioned
-  %   from a motivation. I don't think vulnerabilities are been introduced, its
-  %   more that they get exposed. In this case, the implication does not make
-  %   much sense from a practical point of view.} -> Why not? Shouldn't
-  %   maintainable security be introduced in the computer science curricula since
-  %   we have evidence that developers might not following good practices while
-  %   patching? If they learn it in college and are aware of it in their jobs
-  %   won't they be better prepared?
-  % \item [!] \textit{It did take me several readings to understand the paper, especially the approach} -
- %  what to do with this @rui?
- %    \begin{itemize}
- %      \item Ehmm... strange... we need to improve the approach then. Thoughts?
- %    \end{itemize}
-  % \item [!] \textit{Fig 4, although it looks impressive, I am not sure how to read the figure} - what
-  % should we do @rui?
-  %   \begin{itemize}
-  %     \item Carefully explain it in the caption?
-  %   \end{itemize}
-  % \item [!] \textit{Fig 5, 6 and 7 are not the best ways to portray your findings. For instance, I am
-  % still not sure how the positive, none and negative metrics are calculated and normalized} - what
-  % is the best way then?
-  %   \begin{itemize}
-  %     \item Strange comment...
-  %   \end{itemize}
-  % \item \textit{Authors carefully highlight the key messages of each research question and answer
-  % all accordingly} - Improve this.
-  % \item [!] T2v: \textit{wrong assumption: patching commit should also be responsible to preserve the
-  % maintainability.... The patching commit is to fix the vulnerability and to put the system is
-  % an error-free, executable state so that the fix can be tested} -> I agree with this for vulnerabilities
-  % that are the result of errors. For example, a bad expection handling, a memory leak that led to
-  % a denial-of-service attack, NPE, etc. However, there are many other vulnerabilities whose
-  % exploits do not have interference on systems execution. So, what to do here? Should we divide
-  % the study by vulnerabilities and their implications on systems? Should we divide it by severity levels?
-  %   \begin{itemize}
-  %     \item We need to discuss this... I think this related to the next point.
-  %   \end{itemize}
-  % \item \textit{authors write "...important to write maintainable code to support the production
-  % of more secure software...". Relationship is not clear nor is it backed by some evidence} ->
-  % Improve this.
-  % \item Drop the word patch. Security patches are not patches. Refactorings is perfective
-  % or preventive maintenance, whereas patching a vulnerability is corrective maintenance.
-  % \item Discuss better the fact of not using other important aspects for the baseline such as the
-  % size of the changed method and complexity by metrics. Soften the part where it says we should
-  % have gone beyond size.
-  % \item \textit{ Figure 4 does not indicate that or support the statement that “patching
-  % vulnerabilities has a very significant number of negative cases for all guidelines”. Besides a
-  % corresponding statistical test that supports this statement, even the swarm and box plotz itself
-  % clearly tell a different story. So, authors may rethink this statement or better explain how they
-  % come to this conclusion.} -> We are overselling here. Improve it.
-  % \item \textit{Authors state that they assume for the Pontas dataset that the last commit related
-  % to patching being the actual patch while the parent of the first patch-related commit is assumed
-  % to be the vulnerable version. However, nothing is said whether this assumption holds all the time.
-  % So, how did authors validate their assumption? This is missing, but should be added.} - Improve this.
-  % \item \textit{In Section 3-C, for computing and comparing the maintainability, only security commits
-  % (and parent commit) are mentioned. Did you apply the same method also the the regular changes that
-  % you use to compare with security commits? Please make this clear, currently this is ignored in the
-  % particular  (sub) section.} - Improve this.
-  % \item \textit{For clarification, the authors explained that they performed a manual investigation
-  % for RQ3. It would be beneficial for more details.} -> Clarify this.
-  % \item \textit{In terms of presentation, the results were a bit difficult to follow at times as they
-  % did not contain contextualization. For example, “negative cases” can seem a bit ambiguous when
-  % referencing the relationship between the impact with guidelines.} -> Improve this.
-
-% \end{itemize}
-
-% \section{Plan B}
-%
-% Baseado nos reviewers, eu acho que para seguir as orientaçoes deles de refletir e fortalecer
-% a nova assumption e as implicaçoes, teriamos de acrescentar ao Plano A, os proximos pontos
-% antes de submeter à ESME.
-%
-% \begin{itemize}
-% \item \textbf{Step 1:} Pegar nos casos em que o impacto na maintainability foi negativa e ver se foram modificados
-% depois e se sim, se houve melhoria da maintainability. -> isto para então resolver a questao da
-% assumption errada.
-% \item \textbf{Step 2:} Se não existir nenhuma alteração, entao talvez entrar em contacto com os developers
-% e perguntar se quando fizeram o patch tiveram em consideração maintainability do projecto. E se não porquê.
-% \item Problemas de baselines e testes acho que nao vale a pena pegar porque vai depender
-% do reviewer (eu acho).
-% \end{itemize}
-%
-% Qual é o plano que acha melhor? Como prosseguir?
-%
-% A discutir, mas acho que por agora era tentar fazer o 1.
-%
-% \section{Later}
-%
-% \begin{itemize}
-% 	\item \textbf{Assumptions:}\textit{authors measure the maintainability directly after the patch. inestad it should look
-%   for multiple follow up commits and how maintainability evolves there, as it is likely that certain
-%   issues are addresses AFTER the patching commit} -> I think we should see if later commits touch on
-%   the patch and measure their maintainability again. Was a less maintainable fix pushed first and then
-%   refactored, or does the code stay less maintainable?
-%   \item \textbf{To back up implications:} \textit{In terms of the push for changes to curricula, I would agree that it is important
-%   to always stress writing high quality code. However, the support by the study is not strong. Given a
-%   majority of patches did not decrease maintainability, it shows that there is a reasonable focus on maintainability.
-%   Similarly, it is important to understand the circumstances when it decreased. As previously mentioned, organizations
-%   like would prioritize getting the system patched faster (especially in the case of a severe vulnerability) over
-%   its maintainability.} -> Work with practioners to understand if they were taking into consideration maintainability?
-%   \item \textbf{Baseline:}\textit{Authors just search for commits of a similar size. There are other important aspects
-%   that are entirely neglected such as size of the changed method or how complex it it (by means of
-%   metrics)} -> Improve baseline.
-%   \item \textbf{Baseline:}\textit{Similarly, the comparison to “regular” commits really compares security patches to
-%   any other type of change. It would have been more meaningful to compared them to specific types of
-%   commits, like bug fixes (e.g., are security fixes less maintainable than bug fixes?).}
-%   \item \textbf{Tests:}\textit{or the chosen statistical test, the hypothesis may be not well chosen as it just checks for
-%   the same distribution of pairs of versions. However, this means if a similar number of pairs have
-%   a negative and a positive impact, the hypothesis holds however, if this is not the case, the statistical
-%   test does not state in which direction the hypothesis is rejected, that is, what eventually is the
-%   alternative hypothesis. But this is necessary to get insights whether maintainability is hampered or not.
-%   Authors may rethink the choice of the hypothesis or even the test itself.} -> What about this?
-%   \item \textbf{Thesis idea (PL):} It is important
-%   to understand the correlation between the severity of the vulnerability and the programming language as well
-%   as the maintainability impact. It is not clear how software written a more security-focused language would
-%   inherently be more maintainable (it assumes the flaw is the language and not the person).
-% \end{itemize}
-
 \bibliographystyle{spmpsci}

### 2021-02-03 paper/emse20.tex
small fix

@@
 patches. Our hypothesis in this research question is
-that \emph{Security patches hinder more maintainability and 
-need more attention than regular changes on software maintainability}.
+that \emph{Security patches hinder more software maintainability and 
+need more attention than regular changes}.
 

### 2021-02-05 response/response.tex
answer to comment 3.1

@@
 to perform the current analysis.
+
+\pagebreak
 
@@
 \Us As we described in comment $2.19$, we manually inspected $25$ cases.
-Many of those cases were changes of small size (< $10$ LOCs) where
+Many of those cases were changes of small size ($< 10$ LOCs) where
 refactorings were usually improving names of variables, functions, cases of type
@@
 
+\pagebreak
+
 \newreviewer{3}
@@
 
-\Us We kindly thank the reviewer for the comments. We added a sub Section to 
-the paper called \textit{Why Better Code Hub?}. We compare BCH with other tools
-that perform code analysis and motivate why we find BCH to be the best fit for us.
+\Us We kindly thank the reviewer for the comments.
+
+As mentioned in the paper, BCH checks GitHub codebases against $10$ engineering 
+guidelines as devised by Software Improvement Group (SIG). SIG has devised 
+these guidelines after many years of experience: analyzing more than $15$ million 
+lines of code every week, SIG maintains the industry’s largest 
+benchmark, containing more than $10$ billion lines of code across $200$+ 
+technologies; SIG is the only lab in the world certified by TÜViT to issue ISO 25010:2011
+certificates for Trusted Product Maintainability\footnote{Information available here: 
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
+
+Each of the guidelines in isolation is pretty simple to understand, but, when 
+combined with the others, and, when evaluated as a whole, it becomes easy to see why 
+they would comprise the definition of good software. Furthermore, the relation of the entire set of
+guidelines to maintainability is well detailed in the SIG's book ``Building Maintainable 
+Software: Ten Guidelines for Future-Proof Code''
+\footnote{Available here: 
+\url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
+
+In this study, to clarify, we only focus on measuring software maintainability, we do not focus 
+on vulnerability detection or measuring security. Instead, we want to understand how software 
+maintainability is impacted by security patches. However, as we explained before, BCH is 
+based on the ISO 25010 which considers \emph{Security} as one of the main software product quality 
+characteristics since $2011$.
+
+There are other tools that measure the same metrics such as as Kiuwan; or, even 
+provide a model to calculate maintainability such as SonarCloud. However, we argue
+that BCH is a better tool for the following reasons:
+
+\begin{itemize}
+    \item Kiuwan measures 4 guidelines from the total set of $10$
+    guidelines that BCH measures for maintainability: Complexity, Duplicated 
+    Code, Size and Coupling\footnote{List available here: \url{https://www.kiuwan.com/docs/display/K5/Metrics}}. 
+    We did not find the full description of those metrics anywhere. In constrant, 
+    SIG provides a full description of the guidelines/metrics behind better code
+    hub plus their relation with maintainability on the book and several scientific papers. 
+    \item SonarCloud measures 5 guidelines from the total set of $10$
+    guidelines that BCH measures for maintainability: Complexity, Duplication, 
+    Issues, Size and Tests\footnote{List available here: \url{https://sonarcloud.io/documentation/user-guide/metric-definitions/}}. 
+    The tool rates maintainability with a letter based on the technical 
+    debt ratio which is calculated as the ratio between the cost to develop 
+    the software and the cost to fix it. But no information is provided 
+    on how the remediation cost and developmet cost are calculated. 
+    Better Code Hub contemplates a larger set of maintainability metrics 
+    and our formula for maintainability is based on metrics that are 
+    fully described.  
+\end{itemize}
+
+We do not know codify and did find the tool anywhere in static analysis tools 
+collections or even through organic search at Google.
+
+We added a Section to 
+the paper called \textit{Why Better Code Hub?} where we motivate the Better Code Hub tool and 
+compare it with Kiuwan and SonarCloud.
 
@@
 \Us
+
+In fact, the underlying maintainability model for BCH is documented with several
+(non-academic) reports linked at this page:
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/} as 
+well as validated by the academic community in the following paper:
+
+\begin{verbatim}
+    Heitlager, Ilja, Tobias Kuipers, and Joost Visser. 
+    ``A practical model for measuring maintainability.'' 
+    6th international conference on the quality of information and 
+    communications technology (QUATIC 2007). 
+    IEEE, 2007.
+\end{verbatim}
+
+BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
+The concrete values of the thresholds used in BCH are documented also in their book Building Maintainability 
+Software: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
+
+Further, the methodology for arriving at the thresholds is documented in the following 
+literature:
+
+\begin{verbatim}
+    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
+    ``Deriving metric thresholds from benchmark data.'' 
+    IEEE International Conference on Software Maintenance (ICSME 2010). 
+    IEEE, 2010.
+
+    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
+    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
+    2011 Joint Conference of the 21st International Workshop on Software 
+    Measurement and the 6th International Conference on Software Process 
+    and Product Measurement. 
+    IEEE, 2011.
+
+    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
+    ``Standardized code quality benchmarking for improving software maintainability.'' 
+    Software Quality Journal.
+    Springer, 2011.
+
+\end{verbatim}
+
+SIG performs the threshold calibration
+yearly on a proprietary data set to satisfy the requirements of TUViT to
+be a certified measurement model. However, please note that this data set 
+cannot be shared due to non-disclosure agreements between SIG and its 
+clients.
 

### 2021-02-05 response/response.tex
response to 2.2 and 2.3, still missing the lessons learned section on the paper.. wip

@@
 
-\Us
-
-In fact, the underlying maintainability model for BCH is documented with several
-(non-academic) reports linked at this page:
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/} as 
-well as validated by the academic community in the following paper:
+\Us In our work, we follow the same methodology as the previous work 
+mentioned bellow but instead of measuring
+the impact of energy-oriented changes on software maintainability, 
+we measure the impact of security patches.
 
 \begin{verbatim}
-    Heitlager, Ilja, Tobias Kuipers, and Joost Visser. 
-    ``A practical model for measuring maintainability.'' 
-    6th international conference on the quality of information and 
-    communications technology (QUATIC 2007). 
-    IEEE, 2007.
+    Luis Cruz, Rui Abreu, John Grundy, Li Li and Xin Xia. 
+    ``Do Energy-oriented Changes Hinder Maintainability?.'' 
+    IEEE International Conference on Software Maintenance and 
+    Evolution (ICSME). 
+    IEEE, 2019.
 \end{verbatim}
 
-BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
-The concrete values of the thresholds used in BCH are documented also in their book Building Maintainability 
-Software: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
-
-Further, the methodology for arriving at the thresholds is documented in the following 
-literature:
-
-\begin{verbatim}
-    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
-    ``Deriving metric thresholds from benchmark data.'' 
-    IEEE International Conference on Software Maintenance (ICSME 2010). 
-    IEEE, 2010.
-
-    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
-    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
-    2011 Joint Conference of the 21st International Workshop on Software 
-    Measurement and the 6th International Conference on Software Process 
-    and Product Measurement. 
-    IEEE, 2011.
-
-    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
-    ``Standardized code quality benchmarking for improving software maintainability.'' 
-    Software Quality Journal.
-    Springer, 2011.
-
-\end{verbatim}
-
-SIG performs the threshold calibration
-yearly on a proprietary data set to satisfy the requirements of TUViT to
-be a certified measurement model. However, please note that this data set 
-cannot be shared due to non-disclosure agreements between SIG and its 
-clients.
+Software quality is important because it is ultimately related to 
+the overall cost of developing and maintaining secure and safe software 
+applications. Static analysis tools (SATs) have been built to automatically 
+detect software vulnerabilities. However, these tools only locate 
+the issues in the code and do not provide any assistance or guidance 
+on the impact of the problem (vulnerability) or on the quality of the solution 
+(patch). Improving software security is not a trivial task and 
+requires implementing patches that might affect software 
+maintainability. Many times developers even introduce 
+new vulnerabilies while patching software vulnerabilities
+---harming software reliability and introducing 
+technical debt.
+Research found that $34\%$ of the security patches 
+performed introduce new problems and $52\%$ are incomplete and do not 
+fully secure systems.
+
+Our hypothesis is that some of these patches may 
+have a negative impact on the software maintainability. With this 
+study, we want to understand which maintainability metrics can guide 
+security engineers on producing patches with better quality---which 
+we evaluate with \textit{\textbf{RQ1: What is the impact of security patches on the
+maintainability of open-source software?}}. In this research 
+question, we present the results by metric, overall score, severity and programming language
+to have a better understanding of the impact of each of these parameters in 
+software maintainability.
+
+Different weaknesses may require patches with different characteristics. 
+Our hypothesis
+is that security patches for different weaknesses can have different 
+impacts on software maintainability. Thus, we evaluate this in \textit{\textbf{RQ2: Which weaknesses are more likely to
+affect open-source software maintainability?}}. The results can help
+security engineers on prioritizing weaknesses and bring awareness to the 
+ones that need more attention.
+
+Performing a regular change/refactoring, for instance, to improve the name of 
+a variable or function, is different than performing a security patch.
+Therefore, we created \textit{\textbf{RQ3: What is the impact of security patches versus 
+regular changes on the maintainability of open-source software?}}
+to explore if there is a need to pay more attention to 
+security patches than regular changes during software maintenance tasks.
 
@@
 
-\Us
+\Us We agree with the reviewer that we should be clear on the lessons 
+learned from this study. We now provide a Section titled as \textit{Lessons Learned} where 
+we describe te best practices we think developers should follow
+when patching vulnerabilities.
 

### 2021-02-06 paper/emse20.tex
answer to comment 3.1
lessons learned wip

@@
 \textbf{there is no evidence yet of how security patches impact the
-maintainability of open-source software}. Our suspicion is that 
+maintainability of open-source software}. Our hypothesis is that 
 developers tend to introduce technical debt in their software when 
@@
 
+\section{Lessons Learned}\label{sec:lessons}
+
+\Sofia{WIP} In this section, we summarize the lessons learned and 
+best practices developers should follow to perform 
+quality patches and not avoid hindering software maintainability.
+
+\textit{\textbf{Best Practices:}} There is a set of 
+guidelines which have a higher negative impact in software maintainability.
+While patching security vulnerabilities, developers should 
+take special attention to the size and complexity of their patches
+by designing and implementing solutions that continue to respect the 
+limit bounds of branch points and function sizes that are
+recommended by coding practices.
+
+avoid adding new parameters to functions; avoid large 
+modules and the amount of code within modules 
+that are explosed to other modules.
+
+\textit{\textbf{Severity:}} Vulnerabilities of high and
+medium severity should be prioritized in software maintainability.
+
+\textit{\textbf{Programming Language:}} Developers should 
+pay more attention on patching vulnerabilities in C/C++, Ruby and
+PHP code.
+
+\textit{\textbf{Types of vulnerabilities:}}
+
 \section{Study Implications}\label{sec:implications}

### 2021-02-07 paper/emse20.tex
adding lessons learned

--- 2021-02-03 paper/emse20.tex
@@
 The \emph{Write Code Once} guideline results show that duplicated 
-code increased in $171$ ($17.7\%$) of the patches. Software systems 
+code increased in $17.7\%$ ($171$/$969$) of the patches. Software systems 
 typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. Previous 
@@
 \emph{Write Clean Code} guideline results. BCH reported new code 
-smells for $145$ ($15.0\%$) patches which according to previous work 
-may harm software security since some of these new code smells may 
-be new vulnerabilities~\cite{8819456}. Thus, while developers 
-tend to ignore patch techniques such as the \emph{Replace Method 
-with Method Object} technique, they may also be introducing new 
-software vulnerabilities and, consequently, harm the market value 
-and economy of companies~\cite{4267025}.
+smells for $15.0\%$ ($145$/$969$) of the patches which according to previous work 
+may be the source of new software vulnerabilities~\cite{8819456,10.1145/3133956.3134072}
+capable of harming the market value and economy of companies~\cite{4267025}.
+Developers should never reuse code by copying and pasting 
+existing code fragments. Instead, they should create a method and call 
+it every time needed. The \emph{Extract Method} refactoring 
+technique solves many duplication problems. This makes spoting and 
+solving the issue faster because you only need to fix the method used
+instead of locating and fixing the issue multiple times.
+Clone detection tools can also help on locating the issues.
 
@@
 
+\section{Lessons Learned}\label{sec:lessons}
+
+In this section, we summarize the lessons learned and 
+best practices developers should follow to perform 
+quality security patches and avoid hindering 
+software maintainability.
+
+\textit{\textbf{Best Practices:}} Developers are not 
+paying attention to some quality aspects of their solutions/patches, 
+as seen in Figure~\ref{fig:guidelines}, ending up harming 
+software maintainability. We argue that developers
+should design and implement solutions that respect the 
+limit bounds of branch points and function/module sizes that are 
+recommended by best practices to avoid increasing the 
+size, complexity and dependencies of their patches.
+Still on respecting bound limits, developers 
+should also consider the limit of $4$ parameters per function---to 
+keep unit interfaces small and easy to use and understand.
+Patterns such as \emph{Introduce Parameter Object} are useful 
+to send information to a new function/class through an object
+and keep the number of parameters small and the information well-organized.
+
+Security patches also have a negative impact on the maintenance of 
+software architecture. Maintain the components independence and balance 
+is important to make it easier to find the source code that 
+developers want to patch/improve and to better understand
+how the high-level componenet interact with others. Applying 
+\emph{encapsulation} to hide implementation details and make the system 
+more modular is a step forward to not hinder software architecture maintainability.
+
+According to previous research, there is a correlation between 
+code duplication and code smells~\cite{7476787}---duplicates are a source 
+of regression bugs. BCH reports new code smells
+for $15\%$ of the patches under study which supports previous research---$34\%$ of security patches 
+introduce new problems~\cite{8819456,10.1145/3133956.3134072}. 
+Developers should never reuse code by copying and pasting 
+existing code fragments. Instead, they should create a method and call 
+it every time needed. The \emph{Extract Method} refactoring 
+technique solves many duplication problems. This makes spoting and 
+solving the issue faster because you only need to fix one method instead
+of multiple vulnerabilities. Clone detection tools such as CPD can help 
+on locating duplicates.
+
+\textit{\textbf{Severity:}} Research in the past exhibits proof that developers prioritize
+higher impact vulnerabilities~\cite{10.1145/3133956.3134072}.
+Our study results show that vulnerabilities of high and
+medium severity should be prioritized in software maintainability
+by developers.
+
+\textit{\textbf{Programming Language:}} In general, our study 
+shows that there is over $35\%$ of patches with negative impact 
+on software maintainability per programming language. Developers implementing code in   
+\texttt{C/C++}, \texttt{Ruby} and \texttt{PHP} should pay extra care 
+to their solutions/patches.
+
+\textit{\textbf{Types of vulnerabilities:}} We tried to understand 
+the impact of different types of vulnerabilities on software maintainability.
+Overall, all the CWEs studied in more depth present a negative impact
+over $30\%$ on software maintainability. 
+\emph{Cross-Site Scripting} (CWE-79) and \emph{Improper 
+Restriction of Operations within the Bounds of a Memory Buffer} (CWE-119) 
+are less prone to have impact on open-source software maintainability.
+Developers should pay special attention to \emph{Improper Input Validation} (CWE-20), 
+\emph{Information Exposure} (CWE-200), \emph{Missing Release of Memory after Effective Lifetime} 
+(CWE-401) and \emph{Path Traversal} (CWE-22). However, 
+more research should be performed to better understand 
+the impact of each guideline on each CWE.
+
 \section{Study Implications}\label{sec:implications}

### 2021-02-07 response/response.tex
iterating response with rui's feedback on rev 3

@@
 \begin{itemize}
-    \item Kiuwan measures 4 guidelines from the total set of $10$
-    guidelines that BCH measures for maintainability: Complexity, Duplicated 
+    \item Kiuwan measures 4 guidelines that by name appear to be similar to 
+    some of the guidelines measured by BCH: Complexity, Duplicated 
     Code, Size and Coupling\footnote{List available here: \url{https://www.kiuwan.com/docs/display/K5/Metrics}}. 
-    We did not find the full description of those metrics anywhere. In constrant, 
-    SIG provides a full description of the guidelines/metrics behind better code
+    However, we did not find the full description of kiuwan's metrics anywhere. Thus, 
+    we are ensure that the metrics are the same but we are able to conclude that
+    BCH integrates a larger set of metrics.  
+    In contrast, 
+    SIG provides a full description of the $10$ guidelines/metrics behind better code
     hub plus their relation with maintainability on the book and several scientific papers. 
-    \item SonarCloud measures 5 guidelines from the total set of $10$
-    guidelines that BCH measures for maintainability: Complexity, Duplication, 
+    \item SonarCloud measures 5 similar metrics to the ones measured by BCH: Complexity, Duplication, 
     Issues, Size and Tests\footnote{List available here: \url{https://sonarcloud.io/documentation/user-guide/metric-definitions/}}. 
@@
     Better Code Hub contemplates a larger set of maintainability metrics 
-    and our formula for maintainability is based on metrics that are 
-    fully described.  
+    and our formula for maintainability is not only fully explained in our paper but also 
+    based on metrics that are fully and publicly described.  
 \end{itemize}
 
-We do not know codify and did find the tool anywhere in static analysis tools 
+We do not know codify and did not find the tool anywhere in static analysis tools 
 collections or even through organic search at Google.
@@
 
-Software quality is important because it is ultimately related to 
-the overall cost of developing and maintaining secure and safe software 
-applications. Static analysis tools (SATs) have been built to automatically 
-detect software vulnerabilities. However, these tools only locate 
-the issues in the code and do not provide any assistance or guidance 
-on the impact of the problem (vulnerability) or on the quality of the solution 
-(patch). Improving software security is not a trivial task and 
+Improving software security is not a trivial task and 
 requires implementing patches that might affect software 
-maintainability. Many times developers even introduce 
-new vulnerabilies while patching software vulnerabilities
----harming software reliability and introducing 
-technical debt.
-Research found that $34\%$ of the security patches 
-performed introduce new problems and $52\%$ are incomplete and do not 
-fully secure systems.
-
-Our hypothesis is that some of these patches may 
+maintainability. Our hypothesis is that some of these patches may 
 have a negative impact on the software maintainability. With this 
@@
 learned from this study. We now provide a Section titled as \textit{Lessons Learned} where 
-we describe te best practices we think developers should follow
+we describe the best practices we argue developers should follow
 when patching vulnerabilities.

### 2021-02-07 paper/emse20.tex
adding rqs improvement

@@
 concern is to study whether, while improving software security, 
-developers are also negatively impacting the maintainability of their
-software applications. This is important because software 
+developers are also hindering software maintainability. This is 
+important because software 
 maintainability is approximately $75\%$ of the cost related to a 
@@
 developers tend to introduce technical debt in their software when 
-patching the codebases to address a security flaw because they tend 
-to choose the easiest path to solve it. To address it, we compute 
-the maintainability of $1300$ commits using the \emph{Better Code Hub} tool.
+patching software vulnerabilities because they tend to not pay enough 
+attention to the quality of those patches. To address it, 
+we follow the same methodology as previous research~\cite{8919169} and compute 
+the maintainability of $1300$ patches using the \emph{Better Code Hub} tool. 
+We present the maintainability impact by guideline/metric, overall score, 
+severity and programming language.
 
@@
 \texttt{htmlentities} to escape the data given by the variable
-\texttt{\$\_['file']}. \textbf{Knowing which weaknesses are more 
+\texttt{\$\_['file']}. Our hypothesis
+is that security patches for different weaknesses can have different 
+impacts on software maintainability. \textbf{Understanding which weaknesses are more 
 likely to increase maintainability issues is one step forward to 
@@
 %
-We also compute the maintainability of random regular commits using 
+Performing a regular change/refactoring, for instance, to improve the name of 
+a variable or function, is different than performing a security patch.
+Therefore, we also computed the maintainability of random regular commits using 
 the \emph{Better Code Hub} tool---baseline. We use them to 

### 2021-02-07 response/response.tex
improving response to rev 3.2
Better Code Hub section

@@
 We added a Section to 
-the paper called \textit{Why Better Code Hub?} where we motivate the Better Code Hub tool and 
+the paper called \textit{Better Code Hub} where we motivate the Better Code Hub tool and 
 compare it with Kiuwan and SonarCloud.
@@
 
+The previous information is already included in the paper. However, 
+we clarified these points in Section 2, where we present the research questions 
+and our hypothesis.
+
 \begin{revcomment}[3.3]

### 2021-02-07 paper/emse20.tex
Better Code Hub section

@@
 
-\subsection{Why Bettter Code Hub?}
-
-SIG---the company behind BCH--has been helping business and technology 
-leaders drive their organizational 
-objectives by fundamentally improving the health and security of 
-their software applications for more than $20$ years. The
-inner-working of their SIG-MM model---which is the one behind BCH---was 
-scientifically proven~\cite{4335232,5609747,6113040,baggen2012}.
-BCH not only considers security on their maintainability model~\cite{6616351}
-but also integrates a total of $10$ different code metrics; and, 
-code metrics were empirically validated in previous 
-work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
-
-As BCH there are other tools that do the same kind of code analysis.
-Two examples are Kiuwan and SonarCloud. Kiuwan is not free, their 
-inner-workings do not seem to be published anywhere and there is
-no clear list of metrics available---we inspected their datasheet.
-SonarCloud is free but uses only $4$ of the guidelines (duplication,
-size, complexity and code smells) that BCH considers. It also scores 
-maintainability. In the following table, we
-present a brief comparison between BCH and SonarCloud:
-
-\Sofia{Currently building a script to retrieve the SonarCloud results}
+\subsection{Bettter Code Hub}
+
+SIG---the company behind BCH---has been helping business and technology 
+leaders drive their organizational objectives by fundamentally improving 
+the health and security of their software applications for more than $20$ years. 
+The inner-workings of their SIG-MM model---the one behind BCH---are 
+scientifically proven and certified~\cite{4335232,5609747,6113040,baggen2012}.
+
+BCH checks GitHub codebases against 10 maintainability 
+guidelines~\cite{Visser:2016:OREILLY} that were empirically 
+validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
+SIG has devised these guidelines after many years of experience: analyzing more 
+than 15 million lines of code every week, SIG maintains the industry’s largest 
+benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
+is the only lab in the world certified by TÜViT to issue ISO 25010 certificates\footnote{Information available here: 
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
+BCH's compliance criterion is derived from the requirements for 4-star 
+level maintainability (cf. ISO 25010)~\cite{5609747,6113040,baggen2012,Visser:2016:OREILLY}.
+SIG performs the threshold calibration
+yearly on a proprietary data set to satisfy the requirements of TUViT to
+be a certified measurement model.
+
+As BCH there are other tools that also perform code analysis for similar 
+metrics. Two examples are Kiuwan and SonarCloud. However, Kiuwan does
+not provide the full description of the metrics it measures; and, SonarCloud
+although it provides a way of rating software maintainability, the variables 
+description of their formula are not available. Both analyze less maintainability
+guidelines than BCH and do not have their inner-workings fully and publicly described.
 

### 2021-02-07 response/response.tex
response to 2.12

@@
 
-\Us (I'll do this in the end.)
+\Us All the guidelines are fully described on their book ``Building Maintainable 
+Software: Ten Guidelines for Future-Proof Code''
+\footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
+
+We found other papers using some of the metrics measured by BCH. But we did not find
+extra theoretical papers explaining metrics in detail as we found for McCabe Complexity.
+However, metrics like Unit Size, Duplication, Unit Interfacing, Volume, Testability
+and Code Smells are well-known metrics in the Software Engineering field.
+
+We would like to address your concern regarding the credibility of BCH. BCH checks GitHub 
+codebases against 10 engineering guidelines as devised by Software Improvement Group 
+(SIG). SIG has devised these guidelines after many years of experience: analyzing more 
+than 15 million lines of code every week, SIG maintains the industry’s largest 
+benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
+is the only lab in the world certified by TÜViT to issue ISO 25010:2011 
+certificates\footnote{Information available here: 
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
+
+The underlying maintainability model for BCH is documented with several
+(non-academic) reports linked at this page:
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/} as 
+well as validated by the academic community in the following paper:
+
+\begin{verbatim}
+Heitlager, Ilja, Tobias Kuipers, and Joost Visser. 
+``A practical model for measuring maintainability.'' 
+6th international conference on the quality of information and 
+        communications technology (QUATIC 2007). 
+IEEE, 2007.
+\end{verbatim}
+
+BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
+The concrete values of the thresholds used in BCH are documented also in their book Building Maintainability 
+Software \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
+
+Further, the methodology for arriving at the thresholds is documented in the following 
+literature:
+
+\begin{verbatim}
+    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
+    ``Deriving metric thresholds from benchmark data.'' 
+    IEEE International Conference on Software Maintenance (ICSME 2010). 
+    IEEE, 2010.
+
+    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
+    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
+    2011 Joint Conference of the 21st International Workshop on Software 
+    Measurement and the 6th International Conference on Software Process 
+    and Product Measurement. 
+    IEEE, 2011.
+
+    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
+    ``Standardized code quality benchmarking for improving software maintainability.'' 
+    Software Quality Journal.
+    Springer, 2011.
+
+\end{verbatim}
+
+SIG performs the threshold calibration
+yearly on a proprietary data set to satisfy the requirements of TUViT to
+be a certified measurement model. However, please note that this data set 
+cannot be shared due to non-disclosure agreements between SIG and its 
+clients. BCH has a free plan available that allows the analysis of public
+projects with a size limit of 100.000 lines of code.
 

### 2021-02-07 paper/emse20.tex
response to 2.12
improving study implications

@@
 \scriptsize
-	\caption{Guidelines to produce maintainable code. \Sofia{Still need to add more references}}
+	\caption{Guidelines to produce maintainable code.}
 \begin{tabular}{L{2cm}L{4.5cm}L{4cm}}
@@
 	do not. As mentioned in previous work, security it is very important 
-	and should bee considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}. 
-	However, the lack of experts or awareness to the ease of producing 
+	and should be considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}. 
+	However, the lack of experts and awareness to the ease of producing 
 	vulnerable software leads companies to ship low quality software. 
@@
   patches it is important to help companies shipping software of higher quality. 
-	Tools like Better Code Hub and static analysis (e.g., 
-	SonarQube, Codacy, ESLint, Infer, and more) can be leveraged 
-	to assess the risk of patching vulnerabilities---alongside pull
-	requests/code reviews. Static code analysis may be daunting due 
+  Bryan O’Sullivan, VP of Engineering at Facebook, advocated for new computer science
+  risk models to detect vulnerabilities in scale and predict the level of security of the software under production in
+  his talk “‘Challenges in Making Software Work at Scale”’ at FaceTAV’20.
+
+	Tools like Better Code Hub can complement static analysis (e.g., 
+  SonarQube, Codacy, ESLint, Infer, and more) to provide more information 
+  to security engineers on the vulnerabilities per se and the risk of their patches
+  ---alongside pull requests/code reviews.
+  Static code analysis may be daunting due 
 	to the number of rules and different effects

### 2021-02-08 response/response.tex
adding RQ3

@@
 
-\Us
+\Us We improved the discussion of the comparison between 
+the results for security patches and regular changes. We now
+consider two baselines: \emph{size-baseline}, where 
+we collected random changes with the same size as security 
+patches; and, \emph{random-baseline}, where we collected
+random changes without any restrictions. The overall results
+show that regular changes are less prone to hinder software 
+maintainability than security patches. Security
+patches results have more cases with negative 
+impact on software maintainability than any baseline 
+of regular changes. By providing results 
+for both baselines, we now clearly see that the limitation 
+found for the \emph{size-baseline} does not change 
+the need to pay more attention to security patches 
+in software maintenance tasks.
 
@@
 
-\Us
+\Us As explained in the previous response, security patches 
+hinder more software maintainability than regular changes with size 
+restriction or no size restriction. Thus, we argue that
+special attention should be given to maintainable security
+when teaching software maintainability.
 

### 2021-02-08 paper/emse20.tex
adding RQ3

@@
 \begin{itemize}
-	\vspace{-0.3em}
   \item Evidence that supports the trade-off between security and 
@@
 answer the research questions; Section~\ref{sec:results} presents 
-the results and discusses their implications; 
+the results and discusses their implications; Section~\ref{sec:lessons}
+present the lessons learned of this study;
 Section~\ref{sec:implications} elaborates on the implications
@@
 we present a comparison of security patches with 
-two different baselines of regular changes: \emph{random-baseline},
-a dataset of random regular changes;
-and \emph{size-baseline}, a dataset of  
-regular changes with the same size as security 
-patches. Our hypothesis in this research question is
+two different baselines of regular changes: 
+\emph{size-baseline}, a dataset of  
+random regular changes with the same size as security 
+patches becausecomparing changes with considerable different sizes may be unfair; and, \emph{random-baseline},
+a dataset of random regular changes. Our hypothesis in this research question is
 that \emph{Security patches hinder more software maintainability and 
@@
 We have seen, previously, a deterioration on software maintainability 
-when patching vulnerabilities: $41.9\%$ ($406$) suffered a 
-negative impact, $38.7\%$ ($375$) remained the same and $19.4\%$ 
-($188$) increased software maintainability. For regular changes, 
+when patching vulnerabilities: $41.9\%$ ($406$) of patches suffered a 
+negative impact, $38.7\%$ ($375$) of patches remained the same and $19.4\%$ 
+($188$) of patches increased software maintainability. For regular changes,
+when considering the size of the changes (\emph{size-basline}),
 we observe that the maintainability decreases in $27.0\%$ ($262$) 
@@
 changes has a more positive impact than negative on  
-maintainability.
+maintainability. Fully regular changes (\emph{random-baseline}) 
+are less prone to hinder software maintainability than 
+security changes. About $34.4\%$ ($333$) of the regular changes
+have a negative impact on software maintainability compared with 
+the $41.9\%$ of security patches.
 
@@
 
-Results show that regular changes are less prone to hinder the software maintainability of open-source software.   
+Overall, the results for both baselines, show that regular 
+changes are less prone to hinder the software 
+maintainability of open-source software.   
 We manually inspected a considerable amount of cases from
-the distribution of the regular changes where maintainability 
-increases and found that identifying regular commits as similar in 
+the distribution of regular changes that considers size (\emph{size-baseline}), 
+where the impact on maintainability 
+increases, and, found that identifying regular changes with the same 
 size as the security-related commit is limiting the type of regular 
-commits being randomly chosen (e.g, input 
-patches, variables or functions, type conversion, etc).  
+commits being randomly chosen: input 
+patches, variables or functions, type conversion, i.e., changes
+with no impact on the software metrics BCh evaluates.
 We presume that this phenomenon lead
 to the significant number of cases where there is no impact
-on the software maintainability.
+on the software maintainability. On the other hand, 
+identifying regular changes without any restrictions (\emph{random-baseline})
+shows that regular changes have less negative impact 
+on software maintainability compared with security patches
+with statistical significance ($p = 5.34$x$10^{-8}$) and that 
+special attention should be given to security patches.
 
 \textbf{Summary:} Security-related commits are observed to harm software 
-maintainability. In contrast, one cannot say the same about
-regular changes, in which \textbf{no significant differences are observed}. 
+maintainability. While regular changes with no characteristics restrictions 
+show that regular changes are less prone to harm software maintainability.
 Thus, we urge the importance of adopting maintainability practices while 

### 2021-02-08 scripts/stats/chart.py
random baseline

@@ -62,7 +62,7 @@ def main_comparison_chart(reports, dfs):
                 'pos_abs' : pd.Series([total_rand_reg['pos'],  total_size_reg['pos'], total_sec['pos']]),
                 'nul' : pd.Series([total_rand_reg['nul']/db_rand_size, total_size_reg['nul']/db_size , total_sec['nul']/db_size]),
                 'nul_abs' : pd.Series([total_rand_reg['nul'],  total_size_reg['nul'], total_sec['nul']]),
-                'type' : pd.Series(['Regular Change\n(Random Baseline)', 'Regular Change\n(Size Baseline)', 'Security Change'])}
+                'type' : pd.Series(['Regular Changes\n(random-baseline)', 'Regular Changes\n(size-baseline)', 'Security Patches'])}
         
         df, idx, fig, ax = config_report(rep, len(rep['type']), 7, 5, 111)
         

### 2021-02-08 response/response.tex
few changes

@@
 \Us We thank the editor and reviewers for their valuable feedback. In the 
-following points, we address the concerns raised by the reviewers.
+following, we address the concerns raised by the reviewers, and consequently 
+to your concerns too.
 
@@
 codebases against 10 engineering guidelines as devised by Software Improvement Group 
-(SIG). SIG has devised these guidelines after many years of experience: analyzing more 
+(SIG). These guidelines are the result of many years of experience: analyzing more 
 than 15 million lines of code every week, SIG maintains the industry’s largest 
@@
 
-\Us Thank your for your comment. Note, however, that the goal of our work it not 
+\Us We do not think that the concern of the reviewer is warranted. 
+Note that the goal of our work it not 
 to use static analysis to detect security issues. We have already a dataset of 
@@
 
-\Us Thank you for your comment. As we explained before, namely while addressing 
-comment 1.2, both models assess maintainability. However, the Delta Maintainability 
+\Us As we explained before, namely while addressing comment 1.2, both models assess 
+maintainability. However, the Delta Maintainability 
 Model only uses 5 guidelines of the set of guidelines our study considers. Therefore, 

### 2021-02-08 response/response.tex
improving rev 1 points with feedback from Rui and Luis

@@
 benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is the only lab in the world certified by TÜViT to issue ISO 25010:2011 
+is, according to their website, the only lab in the world certified by TÜViT to issue ISO 25010:2011 
 certificates\footnote{Information available here: 
 \url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
+TÜViT is an IT Security company. Being certified by TÜViT means 
+that the company checked if BCH respects the ISO 25010 guidelines succesfully.
 
 Second, each of the guidelines in isolation is pretty simple to understand, but, when 
-combined with the others, and, when evaluated as a whole, it becomes easy to see why 
+combined and evaluated as a whole, it becomes easy to see why 
 they would comprise the definition of good software. Furthermore, the relation of the 
@@
 
-Further, the methodology for arriving at the thresholds is documented in the following 
+Further, the methodology used to derive the thresholds is documented in the following 
 literature:
@@
 \end{quote}
+
+We changed the flow of the text and clarified some of these points in the \emph{Introduction} Section. We also 
+added a section motivating Better Code Hub called \emph{Better Code Hub} in Section $3$.
 
@@
 may in the future complement static analysis tools 
-by assiting developers with more information on the 
+by assisting developers with more information on the 
 risks associated with their patches.
@@
 Model only uses 5 guidelines of the set of guidelines our study considers. Therefore, 
-we consider it to be an alternative to our model, but a weaker one. This has been 
+we consider it to be an alternative to our model. However, it considers
+a smaller set of guidelines than ours. This has been 
 discussed in the related work section.
@@
 $10.7\%$ of patches involving more than one commit. We replaced 
-line 49 by the following one:
+line 49 by the following sentence:
 
@@
 
-We define floss-refactoring before in the paragraph
-as you can see below. Regarding the methodology to detect floss-refactorings, 
-we manually inspected all the cases to ensure that 
+We define floss-refactoring with the paragraph below. 
+Regarding the methodology to detect floss-refactorings, 
+we manually inspected all the cases to ensure 
 floss-refactorings were not being considered.
+
+\textbf{!!! question: it is a common trend to say that we do manual inspections. Reviewers are getting tired of it because some times is not true. Can we add anything here to make this trustworthy?}
 
@@
 \begin{quote}
-    This study evaluates a final total of $969$ security patches from $260$ 
-distinct open-source projects. 
+    This study evaluates a total of 969 security patches from 
+    260 distinct open-source projects.
 \end{quote}

### 2021-02-08 paper/emse20.tex
improving rev 1 points with feedback from Rui and Luis

@@
 
-% Unmaintainable code is hard to test, analyze and re-use. Therefore,
-% it is important to write maintainable code to support the production
-% of more secure software and prevent
-% the introduction of new vulnerabilities in the code bases.
-%
+As ISO does not provide any specific guidelines/formulas to 
+calculate maintainability, we resort to Software Improvement Group 
+(SIG\footnote{SIG's website: 
+\url{https://www.sig.eu/} (Accessed on \today{})})'s web-based source 
+code analysis service Better Code Hub (BCH)\footnote{BCH's 
+website: \url{https://bettercodehub.com/} (Accessed on \today{})}  
+to compute the software compliance with a set of $10$ 
+guidelines/metrics to produce quality software based in ISO/IEC 
+$25010$~\cite{Visser:2016:OREILLY}. SIG 
+has been helping business and technology leaders drive their organizational 
+objectives by fundamentally improving the health and security of 
+their software applications for more than 20 years. Their 
+models are scientifically proven and certified~\cite{4335232,5609747,6113040,baggen2012}.
+
+There are other well-known 
+standards and models that have been proposed to increase software 
+security: Common Criteria~\cite{common:2009} which received
+negative criticism regarding the costs associated and poor technical 
+evaluation; the OWASP Application Security Verification 
+Standard (ASVS)~\cite{oswap:2009} which is focused only on web 
+applications, and a model proposed by Xu et al. ($2013$)
+~\cite{6616351} for rating software security (arguably, it was one 
+of the first steps taken by SIG to introduce security on their 
+maintainability model). Nevertheless, our study uses BCH to provide 
+an assessment of maintainability in software for the following 
+reasons: BCH integrates a total of $10$ different code metrics; and, 
+code metrics were empirically validated in previous 
+work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
+
 Static analysis tools (SATs) have been built to automatically detect software 
@@
 vulnerabilities on software maintenance across open-source software.
-We arguee that tools that assess these type of metrics may
+We arguee that tools that assess these type of code metrics may
 complement SATs with valuable information to help the developer
-understand better the risk of its patch.
-
-As ISO does not provide any specific guidelines/formulas to 
-calculate maintainability, we resort to Software Improvement Group 
-(SIG\footnote{SIG's website: 
-\url{https://www.sig.eu/} (Accessed on \today{})})'s web-based source 
-code analysis service Better Code Hub (BCH)\footnote{BCH's 
-website: \url{https://bettercodehub.com/} (Accessed on \today{})}  
-to compute the software compliance with a set of $10$ 
-guidelines/metrics to produce quality software based in ISO/IEC 
-$25010$~\cite{Visser:2016:OREILLY}. SIG 
-has been helping business and technology leaders drive their organizational 
-objectives by fundamentally improving the health and security of 
-their software applications for more than 20 years. Their 
-models are scientifically proven and certified~\cite{4335232,5609747,6113040,baggen2012}.
-
-There are other well-known 
-standards and models that have been proposed to increase software 
-security: Common Criteria~\cite{common:2009} which received
-negative criticism regarding the costs associated and poor technical 
-evaluation; the OWASP Application Security Verification 
-Standard (ASVS)~\cite{oswap:2009} which is focused only on web 
-applications, and a model proposed by Xu et al. ($2013$)
-~\cite{6616351} for rating software security (arguably, it was one 
-of the first steps taken by SIG to introduce security on their 
-maintainability model). Nevertheless, our study uses BCH to provide 
-an assessment of maintainability in software for the following 
-reasons: BCH integrates a total of $10$ different code metrics; and, 
-code metrics were empirically validated in previous 
-work~\cite{Bijlsma:2012:FIR:2317098.2317124,8530041,8919169,8785997}.
+understand the risk of its patch.
 
@@
 
-This study evaluates a final total of $969$ security patches from $260$ 
-distinct open-source projects. This section 
+This study evaluates a total of $969$ security patches 
+from $260$ distinct open-source projects. This section 
 reports and discusses the results for each research question.

### 2021-02-08 response/response.tex
small fixes do 1.1

@@
 TÜViT is an IT Security company. Being certified by TÜViT means 
-that the company checked if BCH respects the ISO 25010 guidelines succesfully.
+that the company checked if BCH actually respects and follows the ISO 25010 guidelines.
 
@@
 
-We changed the flow of the text and clarified some of these points in the \emph{Introduction} Section. We also 
+We changed the flow of the text and clarified these points in the \emph{Introduction} Section. We also 
 added a section motivating Better Code Hub called \emph{Better Code Hub} in Section $3$.

### 2021-02-08 response/response.tex
small fixes do 1.2

@@
 measure fine-grained code changes by adapting/extending the BCH model [diBiase2019].
-The present work uses the same model (SIG-MM), considers more guidelines and focuses 
+The present work uses the same base model (SIG-MM), considers more guidelines and focuses 
 solely on evaluating the impact of security patches on software maintainability.

### 2021-02-08 paper/emse20.tex
small fixes do 1.2

@@
 measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
-The present work uses the same model (SIG-MM), considers more guidelines and focuses 
+The present work uses the same base model (SIG-MM), considers more guidelines and focuses 
 solely on evaluating the impact of security patches on software maintainability.

### 2021-02-08 response/response.tex
improving flow of response 1.4

@@
 \Us As we explained before, namely while addressing comment 1.2, both models assess 
-maintainability. However, the Delta Maintainability 
-Model only uses 5 guidelines of the set of guidelines our study considers. Therefore, 
-we consider it to be an alternative to our model. However, it considers
-a smaller set of guidelines than ours. This has been 
+maintainability. We consider the Delta Maintainability 
+Model to be an alternative to our model. However, it only uses 5 guidelines of the 
+set of guidelines our study considers. This has been 
 discussed in the related work section.

### 2021-02-08 response/response.tex
rev 1 final touches

@@
 certificates\footnote{Information available here: 
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}
+\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
+
 The guidelines and their relation to maintainability is well 
@@
 security engineers assess the risk of their patches and guide 
-them on performing better patches. But for now we only try
+them on performing better patches. But, for now, we only try
 to understand the impact of security patches on those metrics.

### 2021-02-08 response/response.tex
reviewer 3

@@
 Each of the guidelines in isolation is pretty simple to understand, but, when 
-combined with the others, and, when evaluated as a whole, it becomes easy to see why 
+combined and evaluated as a whole, it becomes easy to see why 
 they would comprise the definition of good software. Furthermore, the relation of the entire set of
@@
 There are other tools that measure the same metrics such as as Kiuwan; or, even 
-provide a model to calculate maintainability such as SonarCloud. However, we argue
+provide a model to calculate maintainability such as SonarCloud. In this study, it is not
+our goal to study other tools. However, we argue
 that BCH is a better tool for the following reasons:

### 2021-02-08 response/response.tex
merging lessons learned and study implications

@@
 is, according to their website, the only lab in the world certified by TÜViT to issue ISO 25010:2011 
-certificates\footnote{Information available here: 
+certificates for Trusted Product Maintainability\footnote{Information available here: 
 \url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.

### 2021-02-08 paper/emse20.tex
merging lessons learned and study implications

@@
 answer the research questions; Section~\ref{sec:results} presents 
-the results and discusses their implications; Section~\ref{sec:lessons}
-present the lessons learned of this study;
+the results and discusses their implications; 
 Section~\ref{sec:implications} elaborates on the implications
@@
 However, there is no previous work focused on comparing the impact 
-of security patches with regular changes on maintainability---only
+of security patches with regular changes on maintainability, only
 with bug-fixes~\cite{10.1145/3133956.3134072}.
@@
 The baseline dataset is generated from the security commits dataset, i.e.,
-for each security commit in the dataset we collect a random regular
+for each security commit in the dataset, we collect a random regular
 change from the same project. We created two different baselines: 
@@
 \textit{size-baseline}, considering also random changes
-but with the same size as security patches.
+but with an approximate size as security patches---we argue 
+that comparing changes with considerable different sizes is unfair.
 
@@
 
-As for the security patches, for each regular change, we also 
+As for the security patches, for each regular change, we  
 need the commit performing the regular change (\emph{sha-reg}) and version of 
 the software before the change (\emph{sha-reg-p}). For each security patch, a 
-random commit from the same project is selected (\emph{sha-reg}).
-\emph{sha-reg-p} is the commit performed previous to the 
-regular change, \emph{sha-reg}.
+random commit from the same project is selected, \emph{sha-reg}.
+The parent of \emph{sha-reg} or the commit performed before \emph{sha-reg} is \emph{sha-reg-p}.
 
@@
 random regular changes with the same size as security 
-patches becausecomparing changes with considerable different sizes may be unfair; and, \emph{random-baseline},
+patches---we argue that comparing changes with considerable different 
+sizes may be unfair; and, \emph{random-baseline},
 a dataset of random regular changes. Our hypothesis in this research question is
@@
 
-\section{Lessons Learned}\label{sec:lessons}
-
-In this section, we summarize the lessons learned and 
-best practices developers should follow to perform 
-quality security patches and avoid hindering 
-software maintainability.
-
-\textit{\textbf{Best Practices:}} Developers are not 
+\section{Study Implications}\label{sec:implications}
+%
+Our results show evidence that developers may have to reduce maintainability for the 
+sake of security. We argue that developers should be able to patch and produce 
+secure code without hindering the maintainability of their projects. But there are 
+still concerns that need to be addressed and that this study brings awareness for:
+%
+
+\textit{\textbf{Best Practices to Follow:}} Developers are not 
 paying attention to some quality aspects of their solutions/patches, 
@@
 size, complexity and dependencies of their patches.
-Still on respecting bound limits, developers 
-should also consider the limit of $4$ parameters per function---to 
+Developers should also consider the threshold limit 
+for parameters per function---to 
 keep unit interfaces small and easy to use and understand.
@@
 Security patches also have a negative impact on the maintenance of 
-software architecture. Maintain the components independence and balance 
+software architecture. Maintaining the components independence and balance 
 is important to make it easier to find the source code that 
@@
 \emph{encapsulation} to hide implementation details and make the system 
-more modular is a step forward to not hinder software architecture maintainability.
+more modular is a step forward not to hinder software architecture maintainability.
 
@@
 
-\textit{\textbf{Severity:}} Research in the past exhibits proof that developers prioritize
+\textit{\textbf{Prioritize High and Medium Severity:}} Previous research 
+exhibits proof that developers prioritize
 higher impact vulnerabilities~\cite{10.1145/3133956.3134072}.
-Our study results show that vulnerabilities of high and
-medium severity should be prioritized in software maintainability
+Our study shows that vulnerabilities of high and
+medium severity should be prioritized in software maintainability tasks
 by developers.
 
-\textit{\textbf{Programming Language:}} In general, our study 
-shows that there is over $35\%$ of patches with negative impact 
-on software maintainability per programming language. Developers implementing code in   
-\texttt{C/C++}, \texttt{Ruby} and \texttt{PHP} should pay extra care 
-to their solutions/patches.
-
-\textit{\textbf{Types of vulnerabilities:}} We tried to understand 
-the impact of different types of vulnerabilities on software maintainability.
+\textit{\textbf{Some Types of Vulnerabilities Need More Attention:}} 
+Our study attempted at shedding light at the impact 
+of different types of vulnerabilities on software maintainability.
 Overall, all the CWEs studied in more depth present a negative impact
@@
 
-\section{Study Implications}\label{sec:implications}
-%
-Our results show evidence that developers may have to reduce maintainability for the 
-sake of security. We argue that developers should be able to patch and produce 
-secure code without hindering the maintainability of their projects. But there are 
-still concerns that need to be addressed and that this study brings awareness for:
-%
-
-\textit{\textbf{Tools for Patch Risk Assessment:}}
+\textit{\textbf{Tools for Patch Risk Assessment Wanted:}}
 	Design debt of one guideline can lead to severe impacts on the 
@@
 
-\textit{\textbf{More Secure Programming Languages Wanted:}} Programming 
-	Languages should provide new design patterns to easily patch security weaknesses 
+\textit{\textbf{Secure Programming Languages by Design:}} 
+In general, our study 
+shows that there is over $35\%$ of patches with negative impact 
+on software maintainability per programming language. Developers implementing code in   
+\texttt{C/C++}, \texttt{Ruby} and \texttt{PHP} should pay extra care 
+to their solutions/patches. In addition, programming 
+	languages should provide new design patterns to easily patch security weaknesses 
 	without endangering maintainability. Ultimately, new programming languages, 

### 2021-02-08 paper/emse20.tex
small fix

@@
 
-\textit{\textbf{Best Practices to Follow:}} Developers are not 
+\textit{\textbf{Follow Best Practices:}} Developers are not 
 paying attention to some quality aspects of their solutions/patches, 

### 2021-02-08 paper/emse20.tex
final touches to study implications and lessons learned

@@
 Our study shows that vulnerabilities of high and
-medium severity should be prioritized in software maintainability tasks
-by developers.
+medium severity should be prioritized in software maintainability tasks.
 
@@
 of different types of vulnerabilities on software maintainability.
-Overall, all the CWEs studied in more depth present a negative impact
+Overall, all the CWEs under study present a negative impact
 over $30\%$ on software maintainability. 

### 2021-02-08 paper/emse20.tex
pass on RQ3

@@
 but with an approximate size as security patches---we argue 
-that comparing changes with considerable different sizes is unfair.
+that comparing changes with considerable different sizes may be unfair.
 
@@
 the software before the change (\emph{sha-reg-p}). For each security patch, a 
-random commit from the same project is selected, \emph{sha-reg}.
-The parent of \emph{sha-reg} or the commit performed before \emph{sha-reg} is \emph{sha-reg-p}.
+random commit from the same project is selected, \emph{sha-reg}. The parent commit
+of \emph{sha-reg} is the \emph{sha-reg-p}.
 
@@
 and the version of the software before the change (\emph{sha-reg-p}). 
-First, the \emph{diff}
-between the security patch and its parent is calculated.
-Then, a random commit/regular change from 
-the same project is selected (\emph{sha-reg}). The 
+First, our tool calculates the \emph{diff}
+between the security patch and its parent.
+Second, a random commit/regular change from 
+the same project is selected, \emph{sha-reg}. The 
 \emph{diff} between \emph{sha-reg} and its parent (\emph{sha-reg-p})
-is calculated and compared to the \emph{diff} from the security patch.
+is calculated. Then, the regular change \emph{diff} is compared 
+to the security patch \emph{diff}.
 Due to the complexity of some patches, it was not possible 
-to find patches with the exact number of added and deleted lines. 
+to find patches with the exact same number of added and deleted lines. 
 Thus, we looked for an approximation.
-% The algorithm integrates the following steps:
-
-% \begin{enumerate}
-% \item Calculates the \emph{diff} between the security patch commit 
-% (\emph{sha}) and its parent (\emph{sha-p}).
-% \item Selects a random commit from the project (\emph{sha-reg}).
-% \item Calculates the \emph{diff} between the random commit (\emph{sha-reg}) 
-% and its parent (\emph{sha-reg-p}).
-% \item Checks if the size from the regular change has the 
-% same/approximated size to the security patch.
-% \end{enumerate}
 
@@
 \emph{diff} size fits in the range size. This range widens every 
-$10$ attempts to search for a change with the same size. We 
+$10$ attempts to search for a change with an approximate size. We 
 originate the regular changes from the security commits to ensure 
@@
 robust. Thus, we use a version of the Wilcoxon test that 
-incorporates the cases where maintainability is different from 
+incorporates the cases where maintainability is equal to 
 zero~\cite{10.2307/2282543}. The Wilcoxon test requires a 
@@
 
-\textit{\textbf{RQ3: What is the impact of security changes versus regular 
+\textit{\textbf{RQ3: What is the impact of security patches versus regular 
 changes on the maintainability of open-source software?}}
@@
 sizes may be unfair; and, \emph{random-baseline},
-a dataset of random regular changes. Our hypothesis in this research question is
-that \emph{Security patches hinder more software maintainability and 
-need more attention than regular changes}.
-
+a dataset of random regular changes.
+
+Our hypothesis is
+that \emph{security patches hinder more software maintainability 
+than regular changes}. 
 We have seen, previously, a deterioration on software maintainability 
@@
 changes has a more positive impact than negative on  
-maintainability. Fully regular changes (\emph{random-baseline}) 
-are less prone to hinder software maintainability than 
+maintainability. However, no statistical significance was 
+obtained. Regular changes (\emph{random-baseline}), with no size 
+restrictions, are less prone to hinder software maintainability than 
 security changes. About $34.4\%$ ($333$) of the regular changes 
-have a negative impact on software maintainability compared with 
-the $41.9\%$ of security patches.
+have negative impact on software maintainability---less than 
+in the security patches. For the \emph{random-baseline}, 
+statistical significance was retrieved ($p = 5.34$x$10^{-8}$).
 
@@
 patches, variables or functions, type conversion, i.e., changes
-with no impact on the software metrics BCh evaluates.
+with no impact on the software metrics analyzed by BCH.
 We presume that this phenomenon lead
@@
 on software maintainability compared with security patches
-with statistical significance ($p = 5.34$x$10^{-8}$) and that 
-special attention should be given to security patches.
+and that special attention should be given to security patches.
 
 \textbf{Summary:} Security-related commits are observed to harm software 
-maintainability. While regular changes with no characteristics restrictions 
-show that regular changes are less prone to harm software maintainability.
+maintainability, while regular changes are less prone to harm software maintainability.
 Thus, we urge the importance of adopting maintainability practices while 

### 2021-02-08 response/response.tex
2.16 fix

@@
 
-For projects with large codebases, the results retrieved for 
+For projects with large codebases, the results obtained for 
 \emph{Keep Your Codebase Small} were not viable because they 
@@
 floss-refactorings were not being considered.
-
-\textbf{!!! question: it is a common trend to say that we do manual inspections. Reviewers are getting tired of it because some times is not true. Can we add anything here to make this trustworthy?}
 
@@
 \begin{quote}
-    This study evaluates a total of 969 security patches from 
-    260 distinct open-source projects.
+    This study evaluates a total of $969$ security patches and 
+    $969$ regular changes 
+    from $260$ distinct open-source projects. 
 \end{quote}

### 2021-02-09 response/response.tex
luis feedback on 2.8

@@
   We created a tool that pulls the codebase of each commit of our dataset 
-  to a new branch; it sets the new branch as the default branch; it runs 
-  the BCH analysis in the codebase; and, finally saves the BCH metrics
-  results.
+  to a new branch; it sets the new branch as the default branch; and, runs 
+  the BCH analysis on the codebase; after the analysis is finished, the tool saves 
+  the BCH metrics results to a cache file.
 \end{quote}

### 2021-02-09 paper/emse20.tex
2.16 fix
luis feedback on 2.8

@@
   We created a tool that pulls the codebase of each commit of our dataset 
-  to a new branch; it sets the new branch as the default branch; it runs 
-  the BCH analysis in the codebase; and, finally saves the BCH metrics
-  results.
+  to a new branch; it sets the new branch as the default branch; and, runs 
+  the BCH analysis on the codebase; after the analysis is finished, the tool saves 
+  the BCH metrics results to a cache file.
 \end{enumerate}
@@
 
-This study evaluates a total of $969$ security patches 
-from $260$ distinct open-source projects. This section 
+This study evaluates a total of $969$ security patches and 
+$969$ regular changes 
+from $260$ distinct open-source projects. 
+This section 
 reports and discusses the results for each research question.

### 2021-02-09 response/response.tex
more fixes:

@@
 
-Second, each of the guidelines in isolation is pretty simple to understand, but, when 
+Second, each of the guidelines in isolation is easy to understand, but, when 
 combined and evaluated as a whole, it becomes easy to see why 
@@
 
-\Us As we explained before, namely while addressing comment 1.2, both models assess 
+\Us As we explained above, namely while addressing comment 1.2, both models assess 
 maintainability. We consider the Delta Maintainability 
@@
 \Us Thank you for your question. The dataset is fully available at 
-\texttt{dataset/db\_release\_security\_fixes.csv} in the figshare package
+\texttt{dataset/db\_security\_changes.csv} in the figshare package
 provided in the contributions of the paper: 
@@
 accepted, the package will be available on GitHub.
+
+This information is described in Section 1 where we present 
+the contributions of our work:
+
+\begin{quote}
+    This research performs the following main contributions:
+
+\begin{itemize}
+  \item Evidence that supports the trade-off between security and 
+  maintainability: developers may be hindering software 
+  maintainability while patching vulnerabilities.
+	\item An empirical study on the impact of security patches on 
+	software maintainability (per guideline, severity, weakness and 
+	programming language).
+	\item A replication package with the scripts and data created to 
+	perform the empirical evaluation, for reproducibility. Available 
+	online: \url{https://figshare.com/s/4861207064900dfb3372}.
+\end{itemize}
+
+\end{quote}
 
@@
 \Us As explained in comment $2.6$, the $1282$ commits refer only to the Ponta et al. dataset (a total of $624$ 
-patches). In total, we analyzed $1300$ patches. However, BCH as some limitations such as
-analyzing projects with very large codebases and was uncapable to analyze some patches. In addition,
-we also detected floss-refactorings which we decided not to consider. All of this is explained
+patches). In total, we analyzed $1300$ patches ($624$ patches from Ponta et al. and $676$ from Secbench). 
+However, our study only considers $969$ patches. We detected a total of $23$ patches with floss-refactorings. 
+In addition, BCH has some limitations, in particular, the lack of language support and project size,
+and was uncapable of analyzing the rest of the $308$ data points. All of this is explained
 in more detail later in Section 3.4.
@@
 in the paper.
+
+\begin{quote}
+    The baseline dataset is generated from the security commits dataset, i.e.,
+for each security commit in the dataset, we collect a random regular
+change from the same project. 
+\end{quote}
 
@@
 and project size by BCH, $308$ data points were disregarded.
+We clarified it in the paper (Section 3.4):
+
+\begin{quotation}
+    Due to BCH limitations, 
+we detected $308$ data points that retrieved incorrect overall 
+calculations. Those data points were disregarded from the study.
+\end{quotation}
 
@@
 
-\Us In total, we inspected around $25$ cases whose maintainability difference/
-impact was zero to understand what was leading to a higher number of cases with no impact.
+\Us In total, we inspected around $25$ regular changes with no impact on software maintainability  
+ to understand what was leading to the higher number of cases. This point
+ was clarified in Section 4, with the following paragraph:
+
+ \begin{quote}
+    Overall, the results for both baselines, show that regular 
+    changes are less prone to hinder the software 
+    maintainability of open-source software.   
+    However, the \emph{size-baseline} integrates a larger 
+    number of cases with no impact on software maintainability. 
+    We manually inspected a total of $25$ cases from
+    that distribution of regular changes with no impact on maintainability, 
+    and, found that identifying regular changes with the same 
+    size as the security-related commit is limiting the type of regular 
+    commits being randomly chosen: input 
+    patches, variables or functions, type conversion (i.e., changes
+    with no impact on the software metrics analyzed by BCH).
+ \end{quote}
 
@@
 hinder more software maintainability than regular changes with size 
-restriction or no size restriction. Thus, we argue that
+restriction (\emph{size-baseline}) or no size restriction (\emph{random-baseline}). Thus, we argue that
 special attention should be given to maintainable security
-when teaching software maintainability.
+when teaching software maintainability. We improved the analysis 
+provided in Section 4 for the research question 3 (RQ3). We added 
+an extra baseline for comparison with no size restrictions and 
+improved the discussion over our results.
 
@@
 
-Each of the guidelines in isolation is pretty simple to understand, but, when 
+Each of the guidelines in isolation is easy to understand, but, when 
 combined and evaluated as a whole, it becomes easy to see why 
@@
 We added a Section to 
-the paper called \textit{Better Code Hub} where we motivate the Better Code Hub tool and 
+the paper called \textit{3.3 Better Code Hub} where we motivate the Better Code Hub tool and 
 compare it with Kiuwan and SonarCloud.
@@
 
-\Us We agree with the reviewer that we should be clear on the lessons 
-learned from this study. We now provide a Section titled as \textit{Lessons Learned} where 
-we describe the best practices we argue developers should follow
-when patching vulnerabilities.
+\Us We agree with the reviewer that we should better emphasize the lessons 
+learned and what best practices can be followed by developers to assure better 
+vulnerability patches. We improved these points in Section 5.
 

### 2021-02-09 paper/emse20.tex
more fixes:

@@
   patches~\cite{reis2017secbench,10.1109/MSR.2019.00064}. The
-  duplicated patches were tossed (only $2$ were found).
+  duplicated patches were tossed.
 %
@@
 maintainability of open-source software.   
-We manually inspected a considerable amount of cases from
-the distribution of regular changes that considers size (\emph{size-baseline}), 
-where the impact on maintainability 
-increases, and, found that identifying regular changes with the same 
+However, the \emph{size-baseline} integrates a larger 
+number of cases with no impact on software maintainability. 
+We manually inspected a total of $25$ cases from
+that distribution of regular changes with no impact on maintainability, 
+and, found that identifying regular changes with the same 
 size as the security-related commit is limiting the type of regular 
 commits being randomly chosen: input 
-patches, variables or functions, type conversion, i.e., changes
-with no impact on the software metrics analyzed by BCH.
+patches, variables or functions, type conversion (i.e., changes
+with no impact on the software metrics analyzed by BCH).
 We presume that this phenomenon lead

### 2021-02-09 response/response.tex
section 5 fixes, 2.4 in progress

@@
 
-\Us Thank you for suggesting the DiBiase's paper. 
+\Us Thank you for suggesting the Di Biase et al.'s paper. 
 We evaluate the impact of security patches in software 
@@
 
-DiBiase's work was developed in paralell with the one by Cruz et al [Cruz2019]. 
-Both models attempt to measure maintainability, but DiBiase's is not as complete 
+Di Biase et al.'s work was developed in paralell with the one by Cruz et al [Cruz2019]. 
+Both models attempt to measure maintainability, but Di Biase et al.'s is not as complete 
 since it only considers $5$ of the same set of guidelines. Both models build a 
@@
 
-We also mention DiBiase's work in our \textit{Related Work} section:
+We also mention Di Biase et al.'s work in our \textit{Related Work} section:
 \begin{quote}
@@
 measure fine-grained code changes by adapting/extending the BCH model [diBiase2019].
-The present work uses the same base model (SIG-MM), considers more guidelines and focuses 
-solely on evaluating the impact of security patches on software maintainability.
+Our work uses the same base model (SIG-MM), but considers a broader set of guidelines. 
+Moreover, we solely focus on evaluating the impact of security patches on software maintainability.
 \end{quote}
@@
     You should check out this paper: 
-    Li, Frank and Paxson, Vern. A large-scale empirical study of security patches.
+    [Li2017] Li, Frank and Paxson, Vern. A large-scale empirical study of security patches.
     ACM SIGSAC Conference on Computer and Communications Security in 2017.
@@
 of our claims and findings.
+
+We added the following paragraph to Section 7 (\emph{Related Work}):
+
+\begin{quote}
+    Researchers
+performed a large-scale empirical study to understand the characteristics of security patches
+and their differences against bug fixes [Li2017].
+The main findings were that security patches are smaller and less complex compared
+to bug fixes and usually performed at function-level. Our study compares
+the impact of security patches on software maintainability with the impact 
+of regular changes.
+\end{quote}
+
+We added the following information to Section 1 (\emph{Introduction}):
+
+\begin{quote}
+    Our hypothesis is that some of these patches may 
+have a negative impact on the software maintainability and, 
+possibly, even be the cause of the introduction of new 
+vulnerabilities---harming software reliability and introducing 
+technical debt. Research found that $34\%$ of the security patches 
+performed introduce new problems and $52\%$ are incomplete and do not 
+fully secure systems [Li2017].
+\end{quote}
 

### 2021-02-09 paper/emse20.tex
section 5 fixes, 2.4 in progress

@@
 size, complexity and dependencies of their patches.
-Developers should also consider the threshold limit 
-for parameters per function---to 
-keep unit interfaces small and easy to use and understand.
+Developers should also keep function parameters below 
+the recommended limit. It helps keep unit interfaces small 
+and easy to use and understand.
 Patterns such as \emph{Introduce Parameter Object} are useful 
@@
 performed a large-scale empirical study to understand the characteristics of security patches
-and their differences against security bug fixes~\cite{10.1145/3133956.3134072}.
+and their differences against bug fixes~\cite{10.1145/3133956.3134072}.
 The main findings were that security patches are smaller and less complex compared
-to bug fixes and usually performed at function-level.
+to bug fixes and usually performed at function-level. Our study compares
+the impact of security patches on software maintainability with the impact 
+of regular changes.
 
@@
 measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
-The present work uses the same base model (SIG-MM), considers more guidelines and focuses 
-solely on evaluating the impact of security patches on software maintainability.
+Our work uses the same base model (SIG-MM), but considers a broader set of guidelines. 
+Moreover, we solely focus on evaluating the impact of security patches on software maintainability.
 

### 2021-02-09 response/response.tex
improving response

@@
 
-\Us As we explained above, namely while addressing comment 1.2, both models assess 
+\Us As we explain above, namely while addressing comment 1.2, both models assess 
 maintainability. We consider the Delta Maintainability 
@@
 
-We added the following information to Section 1 (\emph{Introduction}):
-
+We also used the paper to suport our research in different Sections of our paper:
+
+Section 1 (\emph{Introduction})
 \begin{quote}
@@
 fully secure systems [Li2017].
+\end{quote}
+
+Section 3.2 (\emph{Security Patches vs. Regular Changes})
+\begin{quote}
+    Previous studies attempted to measure the impact of regular changes 
+on open-source software maintainability. 
+However, there is no previous work focused on comparing the impact 
+of security patches with regular changes on maintainability, only
+with bug-fixes [Li2017].
+\end{quote}
+
+Section 5 (\emph{Study Implications})
+\begin{quote}
+    \textit{\textbf{Prioritize High and Medium Severity:}} Previous research 
+exhibits proof that developers prioritize
+higher impact vulnerabilities [Li2017].
+Our study shows that vulnerabilities of high and
+medium severity should be prioritized in software maintainability tasks.
 \end{quote}
@@
 not to the combined dataset.
+
+\begin{quote}
+    We use a combined dataset of $1300$ security patches which is 
+the outcome of mining and manually inspecting a total of $312$ 
+GitHub projects.
+\end{quote}
 
@@
 learned and what best practices can be followed by developers to assure better 
-vulnerability patches. We improved these points in Section 5.
+vulnerability patches. We improved the discussion of these points in Section 5.
 

### 2021-02-09 response/response.tex
improving 2.13

@@
 \begin{quote}
-    Security patches can be performed through one commit, 
-several consecutive commits, or commit(s) interleaved with more 
-programming activities (floss-refactoring). This study considers 
-mainly single-commit patches. Yet, a small percentage of data points 
-had more than one commit involved in the patch. We manually 
-inspected these cases and $23$ floss-refactorings were identified 
-and disregarded since it would not be fair to measure the 
-maintainability of these cases where other programming activities 
-are involved.
-
-For projects with large codebases, the results obtained for 
+    Security patches can be performed through one commit (single-commit); 
+    several consecutive commits (multi-commits); or, commit(s) interleaved with more 
+    programming activities (floss-refactoring). Only $10.7\%$ of the data points 
+    of our dataset involve more than one commit, the other
+    $89.3\%$ of the cases are single-commit patches. We manually 
+    inspected $25.1\%$ ($244$/$969$) of the security patches---$122$ from each dataset---and, 
+    identified a total of $23$ floss-refactorings. Many of these 
+    are patches with many changes where it is hard to understand which parts 
+    involve the security patch. These were disregarded to minimize the impact 
+    of measuring other programming activities rather than solely security patches
+    can have in our results.
+
+    For projects with large codebases, the results retrieved for 
     \emph{Keep Your Codebase Small} were not viable because they 
-retrieved values above the limit set by BCH ($20$ Person-years). The 
+    retrieved values above the limit set by BCH ($20$ Person-years). Thus, 
+    we decided to not consider this guideline in our research. The 
     \emph{Automated Tests} guideline was also not considered since the 
@@
     Instead, it only contemplates unit testing. Due to BCH limitations, 
-we detected a few data points that retrieved incorrect overall 
+    we detected $308$ data points that retrieved incorrect overall 
     calculations. Those data points were disregarded from the study.
@@
 Regarding the methodology to detect floss-refactorings, 
-we manually inspected all the cases to ensure 
-floss-refactorings were not being considered.
-
-\begin{quote}
-    The same approach is used to analyze the impact of regular 
-    changes. Security patches can be performed through one commit (single-commit); 
+we manually inspected $25.1\%$ ($244$/$969$) of the security 
+patches---$122$ from each dataset---and, 
+identified a total of $23$ floss-refactorings.
+
+\begin{quote}
+    Security patches can be performed through one commit (single-commit); 
     several consecutive commits (multi-commits); or, commit(s) interleaved with more 
@@
     $89.3\%$ of the cases are single-commit patches. We manually 
-    inspected all the cases and identified $23$ floss-refactorings. 
-    These were disregarded since it would not be fair to measure the 
-    maintainability of these cases where other programming activities 
-    are involved.
+    inspected $25.1\%$ ($244$/$969$) of the security patches---$122$ from each dataset---and, 
+    identified a total of $23$ floss-refactorings. Many of these 
+    are patches with many changes where it is hard to understand which parts 
+    involve the security patch. These were disregarded to minimize the impact 
+    of measuring other programming activities rather than solely security patches
+    can have in our results. 
 \end{quote}
@@
 provided in Section 4 for the research question 3 (RQ3). We added 
-an extra baseline for comparison with no size restrictions and 
+an extra baseline for comparison with no size restrictions (\emph{random-baseline}) and 
 improved the discussion over our results.

### 2021-02-09 paper/emse20.tex
improving 2.13

@@
 
-The same approach is used to analyze the impact of regular 
-changes. Security patches can be performed through one commit (single-commit); 
+Security patches can be performed through one commit (single-commit); 
 several consecutive commits (multi-commits); or, commit(s) interleaved with more 
@@
 $89.3\%$ of the cases are single-commit patches. We manually 
-inspected all the cases and identified $23$ floss-refactorings. 
-These were disregarded since it would not be fair to measure the 
-maintainability of these cases where other programming activities 
-are involved.
+inspected $25.1\%$ ($244$/$969$) of the security patches---$122$ from each dataset---and, 
+identified a total of $23$ floss-refactorings. Many of these 
+are patches with many changes where it is hard to understand which parts 
+involve the security patch. These were disregarded to minimize the impact 
+of measuring other programming activities rather than solely security patches
+can have in our results.
 
@@
 on the systems execution but rather on the data accessibility.
+We manually inspected $25.1\%$ of the security patches looking 
+for floss-refactorings---$122$ from each dataset. We did 
+find $23$ cases we argue to be floss-refactorings and toss them
+in order to minimize the impact of this threat.
 
@@
 dataset comprises more commits in Java, i.e., the dataset may not be representative 
-of the population regarding programming languages. For both datasets, we a manual validation 
+of the population regarding programming languages. For both datasets, a manual validation 
 of the message of the commits was performed. Only commits in English were considered. Thus, our approach does not consider 

### 2021-02-09 response/response.tex
small fix
polishing

--- 2021-02-09 response/response.tex
@@
 
-Second, each of the guidelines in isolation is easy to understand, but, when 
-combined and evaluated as a whole, it becomes easy to see why 
-they would comprise the definition of good software. Furthermore, the relation of the 
-guidelines to maintainability is well detailed in the SIG's book ``Building Maintainable 
-Software: Ten Guidelines for Future-Proof Code''
-\footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
+Second, the relation of the entire set of
+guidelines to maintainability is well detailed in the O'Reilly's book ``Building Maintainable 
+Software: Ten Guidelines for Future-Proof Code'' by Professor Joost Visser
+\footnote{Available here: 
+\url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
 In fact, the underlying maintainability model for BCH is documented with several
@@
 
-Each of the guidelines in isolation is easy to understand, but, when 
-combined and evaluated as a whole, it becomes easy to see why 
-they would comprise the definition of good software. Furthermore, the relation of the entire set of
-guidelines to maintainability is well detailed in the SIG's book ``Building Maintainable 
-Software: Ten Guidelines for Future-Proof Code''
+% Each of the guidelines in isolation is easy to understand, but, when 
+% combined and evaluated as a whole, it becomes easy to see why 
+%they would comprise the definition of good software. 
+Furthermore, the relation of the entire set of
+guidelines to maintainability is well detailed in the O'Reilly's book ``Building Maintainable 
+Software: Ten Guidelines for Future-Proof Code'' by Professor Joost Visser
 \footnote{Available here: 

### 2021-02-09 response/response.tex
improving response and paper to the guidelines that were not considered

--- 2021-02-09 response/response.tex
@@
 
-    For projects with large codebases, the results retrieved for 
-    \emph{Keep Your Codebase Small} were not viable because they 
-    retrieved values above the limit set by BCH ($20$ Person-years). Thus, 
-    we decided to not consider this guideline in our research. The 
+    For projects with large codebases, the results calculated for 
+    the \emph{Keep Your Codebase Small} guideline were way above the 
+    limit set by BCH ($20$ Person-years). We suspect this threshold 
+    may not be well calibrated. Thus, we decided not to consider 
+    this guideline in our research. The 
     \emph{Automated Tests} guideline was also not considered since the 
@@
     security testing: vulnerability scanning and penetration testing. 
-    Instead, it only contemplates unit testing. Due to BCH limitations, 
-    we detected $308$ data points that retrieved incorrect overall 
-    calculations. Those data points were disregarded from the study.
+    Instead, it only contemplates unit testing. In total, 
+    we detected $308$ data points suffering from these two limitations. 
+    Those data points were disregarded from the study.
 \end{quote}
@@
 patches---$122$ from each dataset---and, 
-identified a total of $23$ floss-refactorings.
+identified a total of $23$ floss-refactorings. We clarified 
+this in the paper:
 
@@
 
-\Us We started with all the guidelines but after analyzing the BCH results 
-we found that some of the results of the \emph{Keep Your Codebase Small} guidelines were not viable
-because they retrieved values above the limit set by BCH ($20$ Person-years). We did not want 
+\Us We started with all the guidelines but after analyzing the BCH results, 
+we found that some of the results for the \emph{Keep Your Codebase Small} guideline 
+were way above the limit set by BCH ($20$ Person-years). We suspect this threshold 
+may not be well calibrated. Thus, we decided not to consider 
+this guideline in our research. We did not want 
 to lose more data points. Thus, we decided to not consider the guideline for all projects.
@@
 
-We improved the paper in the following text:
-
-\begin{quote}
-    For projects with large codebases, the results retrieved for 
-\emph{Keep Your Codebase Small} were not viable because they 
-retrieved values above the limit set by BCH ($20$ Person-years). Thus, 
-we decided to not consider this guideline in our research. The 
+We improved the paper in the following paragraph:
+
+\begin{quote}
+    Due to BCH limitations, in particular, lack of language support 
+    and project size by BCH, $308$ data points were not analyzed and automatically 
+    disregarded from our study. After performing the BCH analysis and 
+    performing the maintainability calculations, we found the following 
+    limitations regarding two BCH's guidelines:
+
+    1) For projects with large codebases, the results calculated for 
+    the \emph{Keep Your Codebase Small} guideline were way above the 
+    limit set by BCH ($20$ Person-years). We suspect this threshold 
+    may not be well calibrated. Thus, we decided not to consider 
+    this guideline in our research.
+
+    2) The 
     \emph{Automated Tests} guideline was also not considered since the 
@@
     security testing: vulnerability scanning and penetration testing. 
+    Instead, it only contemplates unit testing. 
 \end{quote}
@@
 \begin{quotation}
-    Due to BCH limitations, 
-we detected $308$ data points that retrieved incorrect overall 
-calculations. Those data points were disregarded from the study.
+    Due to BCH limitations, in particular, lack of language support 
+    and project size by BCH, $308$ data points were not analyzed and automatically 
+    disregarded from our study. 
 \end{quotation}

### 2021-02-09 paper/emse20.tex
improving response and paper to the guidelines that were not considered

@@
 
-For projects with large codebases, the results retrieved for 
-\emph{Keep Your Codebase Small} were not viable because they 
-retrieved values above the limit set by BCH ($20$ Person-years). Thus, 
-we decided to not consider this guideline in our research. The 
+Due to BCH limitations, in particular, lack of language support 
+and project size by BCH, $308$ data points were not analyzed and automatically 
+disregarded from our study. After performing the BCH analysis and 
+performing the maintainability calculations, we found the following 
+limitations regarding two BCH's guidelines:
+
+1) For projects with large codebases, the results calculated for 
+the \emph{Keep Your Codebase Small} guideline were way above the 
+limit set by BCH ($20$ Person-years). We suspect this threshold 
+may not be well calibrated. Thus, we decided not to consider 
+this guideline in our research.
+
+2) The 
 \emph{Automated Tests} guideline was also not considered since the 
@@
 security testing: vulnerability scanning and penetration testing. 
-Instead, it only contemplates unit testing. Due to BCH limitations, 
-we detected $308$ data points that retrieved incorrect overall 
-calculations. Those data points were disregarded from the study.
+Instead, it only contemplates unit testing.
 

### 2021-02-09 response/response.tex
small fix
more fixes

@@
 We define floss-refactoring with the paragraph below. 
-Regarding the methodology to detect floss-refactorings, 
-we manually inspected $25.1\%$ ($244$/$969$) of the security 
-patches---$122$ from each dataset---and, 
-identified a total of $23$ floss-refactorings. We clarified 
+To mitigate the impact of 
+floss-refactorings, we extracted and manually inspected a random sample with 
+    $25\%$ of security patches from each dataset. From this sample, we identified 
+    $23$ floss-refactorings. Most floss-refactoring patches include many changes 
+    making it difficult to understand which parts involve the security patch. 
+    Although we suspect that more floss-refactorings may occur, we argue that 
+    they occur in a small portion of the data.We clarified 
 this in the paper:
@@
     of our dataset involve more than one commit, the other
-    $89.3\%$ of the cases are single-commit patches. We manually 
-    inspected $25.1\%$ ($244$/$969$) of the security patches---$122$ from each dataset---and, 
-    identified a total of $23$ floss-refactorings. Many of these 
-    are patches with many changes where it is hard to understand which parts 
-    involve the security patch. These were disregarded to minimize the impact 
-    of measuring other programming activities rather than solely security patches
-    can have in our results. 
+    $89.3\%$ of the cases are single-commit patches. To mitigate the impact of 
+    floss-refactorings, we extracted and manually inspected a random sample with 
+    $25\%$ of security patches from each dataset. From this sample, we identified 
+    $23$ floss-refactorings. Most floss-refactoring patches include many changes 
+    making it difficult to understand which parts involve the security patch. 
+    Although we suspect that more floss-refactorings may occur, we argue that 
+    they occur in a small portion of the data.
 \end{quote}
@@
     disregarded from our study. After performing the BCH analysis and 
-    performing the maintainability calculations, we found the following 
+    the maintainability calculations, we found the following 
     limitations regarding two BCH's guidelines:

### 2021-02-09 paper/emse20.tex
small fix
more fixes

@@
 of our dataset involve more than one commit, the other
-$89.3\%$ of the cases are single-commit patches. We manually 
-inspected $25.1\%$ ($244$/$969$) of the security patches---$122$ from each dataset---and, 
-identified a total of $23$ floss-refactorings. Many of these 
-are patches with many changes where it is hard to understand which parts 
-involve the security patch. These were disregarded to minimize the impact 
-of measuring other programming activities rather than solely security patches
-can have in our results.
+$89.3\%$ of the cases are single-commit patches. To mitigate the impact of 
+floss-refactorings, we extracted and manually inspected a random sample with 
+$25\%$ of security patches from each dataset. From this sample, we identified 
+$23$ floss-refactorings. Most floss-refactoring patches include many changes 
+making it difficult to understand which parts involve the security patch. 
+Although we suspect that more floss-refactorings may occur, we argue that 
+they occur in a small portion of the data.
 
@@
 and project size by BCH, $308$ data points were not analyzed and automatically 
-disregarded from our study. After performing the BCH analysis and 
-performing the maintainability calculations, we found the following 
+disregarded from our study. After performing the BCH analysis and the maintainability calculations, we found the following 
 limitations regarding two BCH's guidelines:

### 2021-02-10 response/response.tex
new pass

@@
 \Us We do not think that the concern of the reviewer is warranted. 
-Note that the goal of our work it not 
+Note that the goal of our work is not 
 to use static analysis to detect security issues. We have already a dataset of 
@@
 
-This information is described in Section 1 where we present 
+This information is also available in Section 1 where we present 
 the contributions of our work:
@@
 Since the comparison is between 
-the final dataset of $969$ patches and the baselines. We produced d
+the final dataset of $969$ patches and the baselines, we produced 
 baselines with the final number of viable results obtained for the security 
@@
     Although we suspect that more floss-refactorings may occur, we argue that 
-    they occur in a small portion of the data.We clarified 
+    they occur in a small portion of the data. We clarified 
 this in the paper:
@@
     the maintainability calculations, we found the following 
-    limitations regarding two BCH's guidelines:
+    limitations regarding two of BCH's guidelines:
 
@@
     Regarding the impact of security patches per guideline, we 
-observe that $38.7\%$ of patches have more cases where the impact is 
-positive---where maintainability increases.
+observe that $38.7\%$ of the security patches have positive impact
+on software maintainability.
 \end{quote}
@@
 Looking at those results, we detected that the percentage of cases for CWE-707 ($30.4\%$) 
-and CWE-664 ($32.8\%$) were considerably higher in the dataset than the rest of CWEs.
+and CWE-664 ($32.8\%$) were considerably higher in the dataset than the rest of the CWEs.
 Thus, we decided to look at lower levels of behavior for both CWEs: CWE-707
@@
 In Figure-7b, all the CWEs presented are sub levels of the CWE-707. The same
-happens for Figure-7c, but the CWEs are sub levels if the CWE-664.
+happens for Figure-7c, but the CWEs are sub levels of the CWE-664.
 

### 2021-02-10 paper/emse20.tex
new pass

@@
 disregarded from our study. After performing the BCH analysis and the maintainability calculations, we found the following 
-limitations regarding two BCH's guidelines:
+limitations regarding two of BCH's guidelines:
 
@@
 Regarding the impact of security patches per guideline, we 
-observe that $38.7\%$ of patches have more cases where the impact is 
-positive---where maintainability increases. However, we also 
+observe that $38.7\%$ of the security patches have positive impact
+on software maintainability. However, we also 
 see that patching vulnerabilities has a very significant number of 

### 2021-02-10 response/response.tex
more fixes

@@
 \usepackage{subcaption}
-\usepackage{hyperref}
+\PassOptionsToPackage{hyphens}{url}\usepackage{hyperref}
 \usepackage{epsfig}
@@
 BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
-The concrete values of the thresholds used in BCH are documented also in their book Building Maintainability 
+The concrete values of the thresholds used in BCH are also documented in their book Building Maintainability 
 Software \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
@@
     Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
-    ``Standardized code quality benchmarking for improving software maintainability.'' 
+    ``Standardized code quality benchmarking for improving software 
+    maintainability.'' 
     Software Quality Journal.
@@
 We do not know codify and did not find the tool anywhere in static analysis tools 
-collections or even through organic search at Google.
+collections or even through an organic search at Google.
 
@@
 \Us In our work, we follow the same methodology as the previous work 
-mentioned bellow but instead of measuring
+mentioned below but instead of measuring
 the impact of energy-oriented changes on software maintainability, 
@@
 affect open-source software maintainability?}}. The results can help
-security engineers on prioritizing weaknesses and bring awareness to the 
+security engineers prioritizing weaknesses and bring awareness to the 
 ones that need more attention.

### 2021-02-10 response/response.tex
fix

--- 2021-02-10 response/response.tex
@@
 We saw the value of a tool like BCH in helping in writing clean and maintainable code and most importantly, 
-it offers us so much more:
-
-Over time, the guidelines in BCH will become part of your coding standards, and the overall quality of your 
-work will improve, slowly, but surely.
+it offers us so much more: over time, the guidelines in BCH will become part of your coding standards, and 
+the overall quality of your work will improve, slowly, but surely.
 \end{quote}

### 2021-02-10 response/response.tex
fix

@@
 
-We changed the flow of the text and clarified these points in the \emph{Introduction} Section. We also 
+We changed the flow of the text and clarified these points in the \emph{Introduction}. We also 
 added a section motivating Better Code Hub called \emph{Better Code Hub} in Section $3$.
@@
 Both models attempt to measure maintainability, but Di Biase et al.'s is not as complete 
-since it only considers $5$ of the same set of guidelines. Both models build a 
-new maintainability model over the SIG-MM model---the model behind BCH.
+since it only considers $5$ of the same set of guidelines. Both models build their 
+maintainability model over the SIG-MM model---the model behind BCH.
 

### 2021-02-10 response/response.tex
more fixes

--- 2021-02-10 response/response.tex
@@
 We saw the value of a tool like BCH in helping in writing clean and maintainable code and most importantly, 
-it offers us so much more: over time, the guidelines in BCH will become part of your coding standards, and 
-the overall quality of your work will improve, slowly, but surely.
+it offers us so much more:
+
+Over time, the guidelines in BCH will become part of your coding standards, and the overall quality of your 
+work will improve, slowly, but surely.
 \end{quote}
@@
 
-\Us Thank you for suggesting the Di Biase et al.'s paper. 
+\Us Thank you for suggesting Di Biase et al.'s paper. 
 We evaluate the impact of security patches in software 
@@
 
-Di Biase et al.'s work was developed in paralell with the one by Cruz et al [Cruz2019]. 
+Di Biase et al.'s work was developed in parallel with the one by Cruz et al [Cruz2019]. 
 Both models attempt to measure maintainability, but Di Biase et al.'s is not as complete 
@@
 
-We would like to rebut your concern on the tools and 
+We would like to rebut your concern about the tools and 
 metrics we are using. As we explained previously, in comment 1.1, 
@@
 
-Note, that our goal is to measure the impact of security patches on software 
+Note that our goal is to measure the impact of security patches on software 
 maintainability and not finding vulnerabilities---vulnerabilities
@@
 \Us We thank the reviewer for the suggestions. We replaced ``suspect'' by
-``hypothesize''. For the later issue, we added to the abstract 
+``hypothesize''. For the latter issue, we added to the abstract 
 the results for maintainability and results
-for the $2$ guidelines with more negative impact on software 
+for the $2$ guidelines with a more negative impact on software 
 maintainability: software complexity and unit size.
@@
 is to show the impact of security patches on software maintainability and
-highlight the need for solutions which are described in the end of 
+highlight the need for solutions, which are described at the end of 
 the abstract, Section 1 (\textit{Introduction}) and Section 5 (\textit{Study Implications}).

### 2021-02-10 response/response.tex

@@
 We saw the value of a tool like BCH in helping in writing clean and maintainable code and most importantly, 
-it offers us so much more:
-
-Over time, the guidelines in BCH will become part of your coding standards, and the overall quality of your 
-work will improve, slowly, but surely.
-\end{quote}
-
-We changed the flow of the text and clarified these points in the \emph{Introduction} Section. We also 
+it offers us so much more: over time, the guidelines in BCH will become part of your coding standards, and 
+the overall quality of your work will improve, slowly, but surely.
+\end{quote}
+
+We changed the flow of the text and clarified these points in the \emph{Introduction}. We also 
 added a section motivating Better Code Hub called \emph{Better Code Hub} in Section $3$.

### 2021-02-10 response/response.tex
more fixes

@@
 
-[Cruz2019] Luis Cruz, Rui Abreu, John Grundy, Li Li, Xin Xia. 
-Do Energy-oriented Changes Hinder Maintainability?. International 
-Conference on Software Maintenance and Evolution (ICSME) in 2019.
-
-[diBiase2019] M. di Biase and A. Rastogi and M. Bruntink and A. van 
-Deursen. The Delta Maintainability Model: Measuring Maintainability 
-of Fine-Grained Code Changes. IEEE/ACM International Conference 
-on Technical Debt (TechDebt) in 2019.
+The literature that supports this response is the following:
+
+\begin{verbatim}
+    Luis Cruz, Rui Abreu, John Grundy, Li Li, Xin Xia. 
+    ``Do Energy-oriented Changes Hinder Maintainability?'' 
+    International Conference on Software Maintenance and 
+    Evolution (ICSME 2019). 
+    IEEE, 2019.
+
+    M. di Biase and A. Rastogi and M. Bruntink and A. van 
+    Deursen. 
+    ``The Delta Maintainability Model: Measuring Maintainability 
+    of Fine-Grained Code Changes.'' 
+    IEEE/ACM International Conference on Technical Debt (TechDebt). 
+    IEEE, 2019.
+
+\end{verbatim}
 
@@
 benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is the *only* lab in the world certified by TÜViT to issue ISO 25010:2011 
+is the only lab in the world certified by TÜViT to issue ISO 25010:2011 
 certificates\footnote{Information available here: 
@@
 to perform the current analysis.
-
-\pagebreak
 
@@
 for the $2$ guidelines with a more negative impact on software 
-maintainability: software complexity and unit size.
+maintainability (software complexity and unit size), in particular:
 
@@
 baseline and completed the work by presenting now two baselines (size-baseline and random-baseline).
-
-\pagebreak
 

### 2021-02-10 response/response.tex
final changes

@@
 
-[Visser2016] J. Visser. Building Maintainable Software, Java Edition: Ten Guidelines 
-for Future-Proof Code. O'Reilly Media, Inc. in 2016.
-
-[Alves2010] T. L. Alves and C. Ypma and J. Visser. Deriving metric thresholds from 
-benchmark data. IEEE International Conference on Software Maintenance in 2010.
-
-[Alves2011] T. L. Alves and J. P. Correia and J. Visser. Benchmark-Based Aggregation 
-of Metrics to Ratings. Joint Conference of the 21st International Workshop on Software 
-Measurement and the 6th International Conference on Software Process and Product Measurement
-in 2011.
-
-[Baggen2012] R. Baggen, J. P. Correia, K. Schill, J. Visser. Standardized code quality 
-benchmarking for improving software maintainability. Software Quality Journal in 2012.
+The literature that supports this response is the following:
+
+\begin{verbatim}
+    J. Visser. 
+    `` Building Maintainable Software, Java Edition: Ten Guidelines 
+    for Future-Proof Code'' 
+    O'Reilly Media, Inc. 2016.
+
+    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
+    ``Deriving metric thresholds from benchmark data.'' 
+    IEEE International Conference on Software Maintenance (ICSME 2010). 
+    IEEE, 2010.
+
+    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
+    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
+    2011 Joint Conference of the 21st International Workshop on Software 
+    Measurement and the 6th International Conference on Software Process 
+    and Product Measurement. 
+    IEEE, 2011.
+
+    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
+    ``Standardized code quality benchmarking for improving software 
+    maintainability.'' 
+    Software Quality Journal.
+    Springer, 2011.
+
+\end{verbatim}
 

### 2021-02-10 paper/emse20.tex

@@
 limit set by BCH ($20$ Person-years). We suspect this threshold 
-may not be well calibrated. Thus, we decided not to consider 
+may not be well calibrated, and hence biasing our results. Thus, we decided not to consider 
 this guideline in our research.
@@
 
-As future work, the study extended in several directions: 
+As future work, the study can be extended in several directions: 
 investigate which guidelines affect most the maintainability per

### 2021-04-06 paper/emse20.tex
small fixes

@@
 
-%
-% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
-%
-% insert here the call for the packages your document requires
-%\usepackage{latexsym}
-% etc.
-%
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
-%
-% Insert the name of "your journal" with
-% \journalname{myjournal}
-%
 \begin{document}
@@
 high-quality software. However, there is still a considerable amount 
-of vulnerabilities being discovered and fixed, almost weekly. We 
+of vulnerabilities being discovered and fixed almost weekly. We 
 hypothesize that developers affect the maintainability of their 
@@
 	perform the empirical evaluation, for reproducibility. Available 
-	online: \url{https://figshare.com/s/4861207064900dfb3372}.
+	online: \url{https://github.com/TQRG/maintainable-security}.
 \end{itemize}
@@
 \section{Motivation and Research Questions}\label{sec:motivation}
-% Rev 2 says it is redundant 
-% Due to time-to-market pressure and the lack of security 
-% expertise~\cite{8077802}, code-related security flaws are generally 
-% detected \textit{after the fact}, i.e., when hackers exploit them. 
-% Patching these issues is usually done by modifying the codebase. 
-% However, these patches may have a negative impact on software 
-% maintenance (e.g., developers might follow the quickest solution to 
-% fix the problem but not necessarily the most elegant and performant 
-% one). Thus, we want to understand how security patches impact
-% software maintainability.
 

### 2021-04-30 response/response_2/response.tex
adding final review draft

--- 2021-02-10 response/response.tex
@@
 \begin{reveditor}
-    Overall, I would like to give you a chance to respond to 
-    these issues in a revision of your manuscript but I have 
-    to make it clear that it does not guarantee a path towards 
-    acceptance. Your rebuttal to \#1 will be of critical importance 
-    for me judge the validity of the use of the BCH tool and in 
-    making a final decision. Methodology is, of course, very 
-    important for ESE journal and \#2 \& \#3s comments on methodological 
-    details need to be carefully addressed, too. 
+    Many thanks for submitting your revision to Empirical Software Engineering journal. 
+    All reviewers and I thank you for the effort you have put in preparing the detailed 
+    revision which successfully addresses all points raised earlier. Accordingly, I am happy 
+    to recommend the acceptance of your manuscript to the EiCs. Please note that reviewers 
+    highlight some editing issues, so please clarify Reviewer \#2's points and conduct an 
+    overall proof read in preparing your final proofs.
+
 \end{reveditor}
 
-\Us We thank the editor and reviewers for their valuable feedback. In the 
-following, we address the concerns raised by the reviewers, and consequently 
-to your concerns too.
-
-\newreviewer{1}
-
-\begin{revcomment}[1.1]
-
-    This paper investigates the impact of patches to improve 
-    security on the maintainability of open-source software. 
-    The paper is well written and easy to read. I really 
-    appreciated the replication package: complete, and well 
-    described. Really a good work!
-
-    However, I found a crucial issue in this paper. The authors 
-    adopted as static analysis tool Better Code Hub. Better Code 
-    Hub' model includes 10 guidelines that can help the developers to 
-    write better code. Unfortunately, these guidelines are not related 
-    to maintainability. Avoiding introducing the issues associated to 
-    these guidelines do not imply increasing the maintainability of the 
-    code, since the tools has no possibility to measure ``maintainability''.
-
-\end{revcomment}
-
-\Us Thank you for your positive feedback on the paper presentation and replication
-package.
-
-We would like to rebut your concern about Better Code Hub (BCH)'s guidelines not 
-being related to maintainability. First, as mentioned in the paper, BCH checks GitHub 
-codebases against 10 engineering guidelines as devised by Software Improvement Group 
-(SIG). These guidelines are the result of many years of experience: analyzing more 
-than 15 million lines of code every week, SIG maintains the industry’s largest 
-benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is, according to their website, the only lab in the world certified by TÜViT to issue ISO 25010:2011 
-certificates for Trusted Product Maintainability\footnote{Information available here: 
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
-TÜViT is an IT Security company. Being certified by TÜViT means 
-that the company checked if BCH actually respects and follows the ISO 25010 guidelines.
-
-Second, the relation of the entire set of
-guidelines to maintainability is well detailed in the O'Reilly's book ``Building Maintainable 
-Software: Ten Guidelines for Future-Proof Code'' by Professor Joost Visser
-\footnote{Available here: 
-\url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
-In fact, the underlying maintainability model for BCH is documented with several
-(non-academic) reports linked at this page:
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/} as 
-well as validated by the academic community in the following paper:
-
-\begin{verbatim}
-Heitlager, Ilja, Tobias Kuipers, and Joost Visser. 
-``A practical model for measuring maintainability.'' 
-6th international conference on the quality of information and 
-        communications technology (QUATIC 2007). 
-IEEE, 2007.
-\end{verbatim}
-
-BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
-The concrete values of the thresholds used in BCH are also documented in their book Building Maintainability 
-Software \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
-
-Further, the methodology used to derive the thresholds is documented in the following 
-literature:
-
-\begin{verbatim}
-    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
-    ``Deriving metric thresholds from benchmark data.'' 
-    IEEE International Conference on Software Maintenance (ICSME 2010). 
-    IEEE, 2010.
-
-    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
-    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
-    2011 Joint Conference of the 21st International Workshop on Software 
-    Measurement and the 6th International Conference on Software Process 
-    and Product Measurement. 
-    IEEE, 2011.
-
-    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
-    ``Standardized code quality benchmarking for improving software 
-    maintainability.'' 
-    Software Quality Journal.
-    Springer, 2011.
-
-\end{verbatim}
-
-SIG performs the threshold calibration
-yearly on a proprietary data set to satisfy the requirements of TUViT to
-be a certified measurement model. However, please note that this data set 
-cannot be shared due to non-disclosure agreements between SIG and its 
-clients.
-
-All in all, we argue that BCH's guidelines are a good proxy for building maintainable software. This is 
-backed up by the observations of SIG in the field, as it was told us in personal communications:
-
-\begin{quote}
-We saw the value of a tool like BCH in helping in writing clean and maintainable code and most importantly, 
-it offers us so much more: over time, the guidelines in BCH will become part of your coding standards, and 
-the overall quality of your work will improve, slowly, but surely.
-\end{quote}
-
-We changed the flow of the text and clarified these points in the \emph{Introduction}. We also 
-added a section motivating Better Code Hub called \emph{Better Code Hub} in Section $3$.
-
-\begin{revcomment}[1.2]
-
-    As far as I know, the company that developed Better Code Hub 
-    developed also the Delta Maintainability Model [diBiase2019] 
-    that aims at measuring the maintainability of a code change 
-    and a score and compare change-based maintainability measurements.
-
-    [diBiase2019] M. di Biase and A. Rastogi and M. Bruntink and A. van 
-    Deursen. The Delta Maintainability Model: Measuring Maintainability 
-    of Fine-Grained Code Changes. IEEE/ACM International Conference 
-    on Technical Debt (TechDebt) in 2019.
-
-\end{revcomment}
-
-\Us Thank you for suggesting Di Biase et al.'s paper. 
-We evaluate the impact of security patches in software 
-maintainability using an alternative maintainability model [Cruz2019].
-
-Di Biase et al.'s work was developed in parallel with the one by Cruz et al [Cruz2019]. 
-Both models attempt to measure maintainability, but Di Biase et al.'s is not as complete 
-since it only considers $5$ of the same set of guidelines. Both models build their 
-maintainability model over the SIG-MM model---the model behind BCH.
-
-Our model was validated by DiBiase in several discussions between 
-the authors of both papers.
-
-We also mention Di Biase et al.'s work in our \textit{Related Work} section:
-\begin{quote}
-    Recent work
-proposed a new maintainability model to 
-measure fine-grained code changes by adapting/extending the BCH model [diBiase2019].
-Our work uses the same base model (SIG-MM), but considers a broader set of guidelines. 
-Moreover, we solely focus on evaluating the impact of security patches on software maintainability.
-\end{quote}
-
-The literature that supports this response is the following:
-
-\begin{verbatim}
-    Luis Cruz, Rui Abreu, John Grundy, Li Li, Xin Xia. 
-    ``Do Energy-oriented Changes Hinder Maintainability?'' 
-    International Conference on Software Maintenance and 
-    Evolution (ICSME 2019). 
-    IEEE, 2019.
-
-    M. di Biase and A. Rastogi and M. Bruntink and A. van 
-    Deursen. 
-    ``The Delta Maintainability Model: Measuring Maintainability 
-    of Fine-Grained Code Changes.'' 
-    IEEE/ACM International Conference on Technical Debt (TechDebt). 
-    IEEE, 2019.
-
-\end{verbatim}
-
-\begin{revcomment}[1.3]
-
-    Moreover, I am not sure that Better Code Hub is 
-    the best static analysis tools to detect security 
-    issues. One of the most adopted in security domain 
-    is Coverity Scan for examples, but also other tools 
-    might be a better choice (e.g. Kiuwan).
-
-\end{revcomment}
-
-\Us We do not think that the concern of the reviewer is warranted. 
-Note that the goal of our work is not 
-to use static analysis to detect security issues. We have already a dataset of 
-security patches (pairs of vulnerable versions and non-vulnerable versions) and 
-our objective is to understand what 
-the impact of those patches on the maintainability
-guidelines/metrics is. We argue that these guidelines/metrics
-may in the future complement static analysis tools 
-by assisting developers with more information on the 
-risks associated with their patches.
-
-\begin{revcomment}[1.4]
-
-    The authors could have adopted both approaches, Better 
-    Code Hub guidelines or another one first and then measure 
-    the code maintainability with the Delta Maintainability Model 
-    since the goal of the paper is to improve security on the 
-    maintainability.
-
-\end{revcomment}
-
-\Us As we explain above, namely while addressing comment 1.2, both models assess 
-maintainability. We consider the Delta Maintainability 
-Model to be an alternative to our model. However, it only uses 5 guidelines of the 
-set of guidelines our study considers. This has been 
-discussed in the related work section.
-
-\begin{revcomment}[1.5]
-
-    In my opinion, the idea behind this work is very interesting, 
-    but the research questions cannot be answered with the tools 
-    and metrics selected.
-
-    My recommendation is to select a specific static analysis tool 
-    for security issues and to include a maintainability model to 
-    evaluate the maintenance.
-
-\end{revcomment}
-
-\Us Thank you for the positive feedback on the work.
-
-We would like to rebut your concern about the tools and 
-metrics we are using. As we explained previously, in comment 1.1, 
-BCH checks GitHub codebases against 10 maintainability guidelines 
-as devised by Software Improvement Group (SIG). SIG has devised 
-these guidelines after many years of experience: analyzing more 
-than 15 million lines of code every week, SIG maintains the industry’s largest 
-benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is the only lab in the world certified by TÜViT to issue ISO 25010:2011 
-certificates\footnote{Information available here: 
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
-
-The guidelines and their relation to maintainability is well 
-detailed in the SIG's book ``Building Maintainable 
-Software: Ten Guidelines for Future-Proof Code''
-\footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
-
-Note that our goal is to measure the impact of security patches on software 
-maintainability and not finding vulnerabilities---vulnerabilities
-are already identified and located. Thus, for now, 
-we do not need to leverage static analysis tools.
-
-For future research, it is indeed interesting to 
-understand how BCH can be fused with a static analysis tool 
-and how the BCH results can help 
-security engineers assess the risk of their patches and guide 
-them on performing better patches. But, for now, we only try
-to understand the impact of security patches on those metrics.
-
-All in all, we argue that BCH is appropriate 
-to perform the current analysis.
+\Us We thank the editor and reviewers for their valuable feedback and acceptance. In the 
+following, we answer the questions raised by the reviewer 2.
 
@@
 \begin{revcomment}[2.1]
+    What is meant by "20 person years"?  How would one obtain this measure?
+\end{revcomment}
 
-    Question:  Is your dataset available?
+\Us A person-year is the equivalent to 12 person-months. A person-month is a standard measure of 
+source code volume where the total volume in a codebase is the volume in lines of code
+converted to person-months. BCH uses person-years instead of volume 
+because it facilitates the comparisons of source code volume between technologies
+(different programming languages have different levels of of verbosity). 
+We do not consider this metric on our study. However, according to the "Building
+Maintainable Software" book, 20 person-years of rebuild value 
+is the equivalent to at most 1750000 lines of code in Java.
+
+\begin{revcomment}[2.2]
+
+    The bullet for automated tests.  Can you think of another work 
+    instead of "contemplate"?  I'm not sure of what you are trying to say.
 
@@
 
-\Us Thank you for your question. The dataset is fully available at 
-\texttt{dataset/db\_security\_changes.csv} in the figshare package
-provided in the contributions of the paper: 
-\url{https://figshare.com/s/4861207064900dfb3372}. If the paper is 
-accepted, the package will be available on GitHub.
-
-This information is also available in Section 1 where we present 
-the contributions of our work:
-
-\begin{quote}
-    This research performs the following main contributions:
-
-\begin{itemize}
-  \item Evidence that supports the trade-off between security and 
-  maintainability: developers may be hindering software 
-  maintainability while patching vulnerabilities.
-	\item An empirical study on the impact of security patches on 
-	software maintainability (per guideline, severity, weakness and 
-	programming language).
-	\item A replication package with the scripts and data created to 
-	perform the empirical evaluation, for reproducibility. Available 
-	online: \url{https://figshare.com/s/4861207064900dfb3372}.
-\end{itemize}
-
-\end{quote}
-
-\begin{revcomment}[2.2]
-
-    Abstract:
-    Throughout the paper, "hypothesize" is probably a better word than 
-    "suspect" for sounding more scientific.
-
-    Your results should be more specific than "show evidence of trace-off". 
-    Briefly tell us about what metrics you used and what the numerical results 
-    indicate.
-
-\end{revcomment}
-
-\Us We thank the reviewer for the suggestions. We replaced ``suspect'' by
-``hypothesize''. For the latter issue, we added to the abstract 
-the results for maintainability and results
-for the $2$ guidelines with a more negative impact on software 
-maintainability (software complexity and unit size), in particular:
-
-\begin{quote}
-    Results show evidence of a trade-off 
-between security and maintainability for $41.90\%$ of the cases, i.e., developers 
-may hinder software maintainability. Our analysis shows that $38.29\%$ of patches increased
-software complexity and $37.87\%$ of patches increased the percentage 
-of LOCs per unit. 
-\end{quote}
-
-\begin{revcomment}[2.3]
-
-    Intro:
-    \begin{itemize}
-        \item First sentence - quality is not ONLY related to cost but also to security 
-        and safety.
-        \item Provide a URL to Software Improvement Group and Better Code Hub.
-        \item Page 3, Line 1 - Application Security Verification Standard (ASVS).
-        \item Page 3, Line 7 - instead of a "broad number of code metrics" - tell 
-        us exactly the number of metrics.
-        \item Page 3, Line 13 - "suggest" sounds more scientific than "hint at"
-        \item Page 3, Line 24 - "we intend to highlight the need …". Is that the 
-        broad goal of your paper?  The goal should be explicitly stated in the 
-        abstract and intro.
-    \end{itemize}
-\end{revcomment}
-
-\Us Thank you for reporting these issues. All of them were addressed in 
-the new version of the paper. In the last point, where the reviewer asks ``Is that the 
-broad goal of your paper?'', we want to clarify that the goal of our study
-is to show the impact of security patches on software maintainability and
-highlight the need for solutions, which are described at the end of 
-the abstract, Section 1 (\textit{Introduction}) and Section 5 (\textit{Study Implications}).
-
-\begin{revcomment}[2.4]
-
-    Intro:\\
-    You should check out this paper: 
-    [Li2017] Li, Frank and Paxson, Vern. A large-scale empirical study of security patches.
-    ACM SIGSAC Conference on Computer and Communications Security in 2017.
-
-\end{revcomment}
-
-\Us Thank you for bringing our attention to this paper. We now discuss
-this paper in the Related Work section and used it to support some 
-of our claims and findings.
-
-We added the following paragraph to Section 7 (\emph{Related Work}):
-
-\begin{quote}
-    Researchers
-performed a large-scale empirical study to understand the characteristics of security patches
-and their differences against bug fixes [Li2017].
-The main findings were that security patches are smaller and less complex compared
-to bug fixes and usually performed at function-level. Our study compares
-the impact of security patches on software maintainability with the impact 
-of regular changes.
-\end{quote}
-
-We also used the paper to suport our research in different Sections of our paper:
-
-Section 1 (\emph{Introduction})
-\begin{quote}
-    Our hypothesis is that some of these patches may 
-have a negative impact on the software maintainability and, 
-possibly, even be the cause of the introduction of new 
-vulnerabilities---harming software reliability and introducing 
-technical debt. Research found that $34\%$ of the security patches 
-performed introduce new problems and $52\%$ are incomplete and do not 
-fully secure systems [Li2017].
-\end{quote}
-
-Section 3.2 (\emph{Security Patches vs. Regular Changes})
-\begin{quote}
-    Previous studies attempted to measure the impact of regular changes 
-on open-source software maintainability. 
-However, there is no previous work focused on comparing the impact 
-of security patches with regular changes on maintainability, only
-with bug-fixes [Li2017].
-\end{quote}
-
-Section 5 (\emph{Study Implications})
-\begin{quote}
-    \textit{\textbf{Prioritize High and Medium Severity:}} Previous research 
-exhibits proof that developers prioritize
-higher impact vulnerabilities [Li2017].
-Our study shows that vulnerabilities of high and
-medium severity should be prioritized in software maintainability tasks.
-\end{quote}
-
-\begin{revcomment}[2.5]
-
-    Section 2\\
-    \begin{itemize}
-        \item The first paragraph is redundant with Section 1.  
-        \item Page 4, Line 15 - You define the OCSP acronym late in the paragraph.
-        \item Page 5, Line 6 - how many new branch points?
-        \item Page 6, Line 19 - I think you want to say "In this study, maintainability is …."  
-        Since it currently implies this information is available in the CWE.
-        \item Page 6, line 47 - "regular commits" non-security commits to be more clear; maybe a 
-        few more words to explain "randomly collected."  You also call them "baseline commits" in Figure 
-        1 which gives a different phrase for the same concept.  Pick one and use it everywhere.
-    \end{itemize}
-\end{revcomment}
-
-\Us Thank you for pointing out all of these issues. We addressed all of them. We decided
-to use regular changes/regular commits instead of baseline and non-security commits.
-
-\begin{revcomment}[2.6]
-    Section 3\\
-    \begin{itemize}
-        \item Sometime you say 1330 patches (e.g. page 7, line 16) and sometimes 1300 (e.g. page 8, line 11, 
-        and the abstract).  In science it's better to say the exact number and use it always and not round.  
-        \item You say the dataset had 1330 (or 1300) patches and 1282 commits though you also say one patch 
-        can have multiple commits assigned.  So how did you end up with less than one commit/patch?
-    \end{itemize}
-\end{revcomment}
-
-\Us Thank you for raising this issue. The right number of security patches is $1300$ ($624$ from Ponta et al. and 
-$676$ from Secbench). The $1330$ value is a typo, we fixed the issue. Regarding the second point, $1282$ commits is
-the equivalent to the $624$ patches that integrate the Pontas et al. dataset where one patch
-can have multiple commits. The $1282$ commits are referring to the Ponta et al. dataset and 
-not to the combined dataset.
-
-\begin{quote}
-    We use a combined dataset of $1300$ security patches which is 
-the outcome of mining and manually inspecting a total of $312$ 
-GitHub projects.
-\end{quote}
-
-\begin{revcomment}[2.7]
-    Section 3\\
-    \begin{itemize}
-        \item I would really like a better picture of what SIG is to establish credibility.  I was at first 
-        thinking it had something to do with ACM SIG's but now after going to the website, it seems like a 
-        commercial/consulting organization? I also feel irritated that I don't understand what you really mean 
-        by "running against the BCH toolset" (page 8, line 13) since you have repeatedly mentioned BCH like it was 
-        an industry standard - but you have not told me if it's a static analysis tool or what it is.  The explanation 
-        and link to introduce SIG and BCH should be on Page 2 lines 41-42 rather than Page 7 - though an explanation 
-        of the BCH toolset is best a paragraph in the methodology.  
-    \end{itemize}
-\end{revcomment}
-
-\Us Thank you for raising the concern regarding SIG's credibility. SIG is a company that 
-has more than $20$ years of experience and research in software quality production. 
-Their models are scientifically proven and certified. We added this information and 
-the links the reviewer refer in the comment to the Introduction section.
-
-\begin{quote}
-    As ISO does not provide any specific guidelines/formulas to 
-calculate maintainability, we resort to Software Improvement Group 
-(SIG\footnote{SIG's website: 
-\url{https://www.sig.eu/} (Accessed on \today{})})'s web-based source 
-code analysis service Better Code Hub (BCH)\footnote{BCH's 
-website: \url{https://bettercodehub.com/} (Accessed on \today{})}  
-to compute the software compliance with a set of $10$ 
-guidelines/metrics to produce quality software based in ISO/IEC 
-$25010$ [Visser2016]. SIG 
-has been helping business and technology leaders drive their organizational 
-objectives by fundamentally improving the health and security of 
-their software applications for more than 20 years. Their 
-models are scientifically proven and certified [Alves2010, Alves2011, Baggen2012].
-\end{quote}
-
-We rephrased "running against the BCH toolset" to "analyzed using the BCH toolset". We mean
-that we used BCH to calculate the metrics presented on Table 1.
-
-The literature that supports this response is the following:
-
-\begin{verbatim}
-    J. Visser. 
-    `` Building Maintainable Software, Java Edition: Ten Guidelines 
-    for Future-Proof Code'' 
-    O'Reilly Media, Inc. 2016.
-
-    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
-    ``Deriving metric thresholds from benchmark data.'' 
-    IEEE International Conference on Software Maintenance (ICSME 2010). 
-    IEEE, 2010.
-
-    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
-    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
-    2011 Joint Conference of the 21st International Workshop on Software 
-    Measurement and the 6th International Conference on Software Process 
-    and Product Measurement. 
-    IEEE, 2011.
-
-    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
-    ``Standardized code quality benchmarking for improving software 
-    maintainability.'' 
-    Software Quality Journal.
-    Springer, 2011.
-
-\end{verbatim}
-
-\begin{revcomment}[2.8]
-    Section 3\\
-    \begin{itemize}
-        \item Page 7, line 4 - I don't think "resemble" is the word you are looking for - but I don't 
-        know what you are trying to say so I can't make a suggestion.
-    \end{itemize}
-\end{revcomment}
-
-\Us Thank you for raising this issue. In the sentence ``The codebases that resemble to the commits of our datasets 
-are collected before the analysis performed by BCH.'', we mean that the codebases of each commit in 
-our dataset were analyzed by BCH. We improved the text for:
-
-\begin{quote}
-BCH evaluates the codebase available in the default branch of a GitHub project. 
-  We created a tool that pulls the codebase of each commit of our dataset 
-  to a new branch; it sets the new branch as the default branch; and, runs 
-  the BCH analysis on the codebase; after the analysis is finished, the tool saves 
-  the BCH metrics results to a cache file.
-\end{quote}
-
-\begin{revcomment}[2.9]
-    Section 3\\
-    \begin{itemize}
-        \item Please explain your repeatable methodology to "clean the dataset and select the most relevant and 
-        compatible projects" (page 8, line 14) that brings the patches from 1282 to 969.  
-        \item From 969 to 866 patches - 
-        you just could not classify?  Please explain.
-    \end{itemize}
-\end{revcomment}
-
-\Us As explained in comment $2.6$, the $1282$ commits refer only to the Ponta et al. dataset (a total of $624$ 
-patches). In total, we analyzed $1300$ patches ($624$ patches from Ponta et al. and $676$ from Secbench). 
-However, our study only considers $969$ patches. We detected a total of $23$ patches with floss-refactorings. 
-In addition, BCH has some limitations, in particular, the lack of language support and project size,
-and was uncapable of analyzing the rest of the $308$ data points. All of this is explained
-in more detail later in Section 3.4.
-
-\begin{quote}
-    Security patches can be performed through one commit (single-commit); 
-    several consecutive commits (multi-commits); or, commit(s) interleaved with more 
-    programming activities (floss-refactoring). Only $10.7\%$ of the data points 
-    of our dataset involve more than one commit, the other
-    $89.3\%$ of the cases are single-commit patches. We manually 
-    inspected $25.1\%$ ($244$/$969$) of the security patches---$122$ from each dataset---and, 
-    identified a total of $23$ floss-refactorings. Many of these 
-    are patches with many changes where it is hard to understand which parts 
-    involve the security patch. These were disregarded to minimize the impact 
-    of measuring other programming activities rather than solely security patches
-    can have in our results.
-
-    For projects with large codebases, the results calculated for 
-    the \emph{Keep Your Codebase Small} guideline were way above the 
-    limit set by BCH ($20$ Person-years). We suspect this threshold 
-    may not be well calibrated. Thus, we decided not to consider 
-    this guideline in our research. The 
-    \emph{Automated Tests} guideline was also not considered since the 
-    tool does not contemplate two of the most important techniques to 
-    security testing: vulnerability scanning and penetration testing. 
-    Instead, it only contemplates unit testing. In total, 
-    we detected $308$ data points suffering from these two limitations. 
-    Those data points were disregarded from the study.
-\end{quote}
-
-Those are the cases that were tossed from the initial $1300$ patches.
-
-Regarding the second question:
-while manually inspecting the patches, we were not able to map
-the issue to any CWE with confidence due to the lack of quality
-information on the vulnerability/patch for $103$ patches. 
-We clarified this in the paper.
-
-\begin{quote}
-    A total 
-of $103$ patches were not classified because we were not able to map
-the issue to any CWE with confidence due to the lack of quality
-information on the vulnerability/patch.
-\end{quote}
-
-\begin{revcomment}[2.10]
-    Section 3\\
-    \begin{itemize}
-        \item Page 8, line 33 "regular commit" - pick baseline or regular (or non-security) commits and use it always.  
-        I don't understand the sentence "The baseline dataset is generated from the security commits dataset."   
-        Do you mean you extract these commits from the projects in the security commits dataset?
-    \end{itemize}
-\end{revcomment}
-
-\Us Yes, for each security commit we collected a random regular change from the same project. We have clarified it 
-in the paper.
-
-\begin{quote}
-    The baseline dataset is generated from the security commits dataset, i.e.,
-for each security commit in the dataset, we collect a random regular
-change from the same project. 
-\end{quote}
-
-\begin{revcomment}[2.11]
-    Section 3\\
-    \begin{itemize}
-        \item Section 3.2 is confusing enough that I'm not sure of the analysis - 
-        I will need to review the revision that you would submit to gain confidence.  
-        Are you saying you want your non-security commits (regular commits) to be about 
-        the same size as a given security commit so you are using this criteria to choose 
-        your regular commits?  Justify this reasoning and make the methodology clear.  
-        I would have expected that you randomly collected 1300 non-security commits to compare 
-        with the 1300 security commits. Why constrain to make them be of "similar" size?  
-        Why not allow the characteristics of regular versus security commits to be what 
-        they are without such curating?
-    \end{itemize}
-\end{revcomment}
-
-\Us Yes, we are looking for regular changes with the same size as security commits
-from the same project. However, 
-for some cases, it was difficult to find the exact same size. Thus, we  
-searched for an approximation. Every 10 attempts of failling to get a commit with 
-the same size, we spanned the range size. We clear the methodology in the paper
-in Section 3.2 \textit{(Security Patches vs. Regular Changes)}.
-
-``I would have expected that you randomly collected 1300 non-security commits to compare 
-with the 1300 security commits.'': Although we started with 1300 commits, 
-$331$ cases were tossed due to BCH limitations. In addition, BCH
-takes substantial time to process this amount of cases. 
-Since the comparison is between 
-the final dataset of $969$ patches and the baselines, we produced 
-baselines with the final number of viable results obtained for the security 
-dataset.
-
-We 
-have collected a baseline of random changes only (no size restrictions) in the past and 
-performed the same evaluation. We now present the results for both baselines: \textit{size-baseline}, where we consider
-size appproximation; and, \textit{random-baseline}, where we consider all the regular
-changes characteristics.
-
-\begin{revcomment}[2.12]
-    Section 3\\
-    \begin{itemize}
-        \item Since SIG seems to be "just another consulting organization" - I'd like 
-        to see about the acceptance of the 10 guidelines in the software engineering 
-        discipline.  I would guess you could find peer-reviewed references to support 
-        these guidelines, or even books by Martin Fowler.   That has more credibility 
-        than the reference by the SIG consulting group. For example on Page 9, line 42 
-        you cite McCabe Complexity --  that has more credibility than a consulting 
-        organizations guidelines that they use to make money on their tool.  Tell us more 
-        about the BCH data experience since that is your comparison point for compliance.
-    \end{itemize}
-\end{revcomment}
-
-\Us All the guidelines are fully described on their book ``Building Maintainable 
-Software: Ten Guidelines for Future-Proof Code''
-\footnote{Available here: \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
-
-We found other papers using some of the metrics measured by BCH. But we did not find
-extra theoretical papers explaining metrics in detail as we found for McCabe Complexity.
-However, metrics like Unit Size, Duplication, Unit Interfacing, Volume, Testability
-and Code Smells are well-known metrics in the Software Engineering field.
-
-We would like to address your concern regarding the credibility of BCH. BCH checks GitHub 
-codebases against 10 engineering guidelines as devised by Software Improvement Group 
-(SIG). SIG has devised these guidelines after many years of experience: analyzing more 
-than 15 million lines of code every week, SIG maintains the industry’s largest 
-benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is the only lab in the world certified by TÜViT to issue ISO 25010:2011 
-certificates\footnote{Information available here: 
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
-
-The underlying maintainability model for BCH is documented with several
-(non-academic) reports linked at this page:
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/} as 
-well as validated by the academic community in the following paper:
-
-\begin{verbatim}
-Heitlager, Ilja, Tobias Kuipers, and Joost Visser. 
-``A practical model for measuring maintainability.'' 
-6th international conference on the quality of information and 
-        communications technology (QUATIC 2007). 
-IEEE, 2007.
-\end{verbatim}
-
-BCH's compliance criterion is derived from the requirements for 4-star level maintainability (cf. ISO 25010). 
-The concrete values of the thresholds used in BCH are documented also in their book Building Maintainability 
-Software \url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}.
-
-Further, the methodology for arriving at the thresholds is documented in the following 
-literature:
-
-\begin{verbatim}
-    Tiago L. Alves, Christiaan Ypma and Joost Visser. 
-    ``Deriving metric thresholds from benchmark data.'' 
-    IEEE International Conference on Software Maintenance (ICSME 2010). 
-    IEEE, 2010.
-
-    Tiago L. Alves, Jose Pedro Correia and Joost Visser. 
-    ``Benchmark-Based Aggregation of Metrics to Ratings.'' 
-    2011 Joint Conference of the 21st International Workshop on Software 
-    Measurement and the 6th International Conference on Software Process 
-    and Product Measurement. 
-    IEEE, 2011.
-
-    Robert Baggen, Jose Pedro Correia, Katrin Schill and Joost Visser. 
-    ``Standardized code quality benchmarking for improving software maintainability.'' 
-    Software Quality Journal.
-    Springer, 2011.
-
-\end{verbatim}
-
-SIG performs the threshold calibration
-yearly on a proprietary data set to satisfy the requirements of TUViT to
-be a certified measurement model. However, please note that this data set 
-cannot be shared due to non-disclosure agreements between SIG and its 
-clients. BCH has a free plan available that allows the analysis of public
-projects with a size limit of 100.000 lines of code.
-
-\begin{revcomment}[2.13]
-    Section 3\\
-    \begin{itemize}
-        \item Page 10, line 49 "mainly single-commit patches" … "small percentage 
-        of data points" - give us the exact numbers for each of these.  
-        \item Page 11, line 38 - define "floss refactoring" and justify how these were objectively 
-        and repeatably identified in your set. You say "these cases" which makes 
-        it seems that you were including the set of patches that had more than one 
-        commit as your (only) floss candidates.  
-    \end{itemize}
-\end{revcomment}
-
-\Us Our dataset integrates $89.3\%$ of single-commit patches and 
-$10.7\%$ of patches involving more than one commit. We replaced 
-line 49 by the following sentence:
-
-\begin{quote}
-    Only $10.7\%$ of the data points 
-of our dataset involve more than one commits, the other
-$89.3\%$ of the cases are single-commit patches.
-\end{quote}
-
-We define floss-refactoring with the paragraph below. 
-To mitigate the impact of 
-floss-refactorings, we extracted and manually inspected a random sample with 
-    $25\%$ of security patches from each dataset. From this sample, we identified 
-    $23$ floss-refactorings. Most floss-refactoring patches include many changes 
-    making it difficult to understand which parts involve the security patch. 
-    Although we suspect that more floss-refactorings may occur, we argue that 
-    they occur in a small portion of the data. We clarified 
-this in the paper:
-
-\begin{quote}
-    Security patches can be performed through one commit (single-commit); 
-    several consecutive commits (multi-commits); or, commit(s) interleaved with more 
-    programming activities (floss-refactoring). Only $10.7\%$ of the data points 
-    of our dataset involve more than one commit, the other
-    $89.3\%$ of the cases are single-commit patches. To mitigate the impact of 
-    floss-refactorings, we extracted and manually inspected a random sample with 
-    $25\%$ of security patches from each dataset. From this sample, we identified 
-    $23$ floss-refactorings. Most floss-refactoring patches include many changes 
-    making it difficult to understand which parts involve the security patch. 
-    Although we suspect that more floss-refactorings may occur, we argue that 
-    they occur in a small portion of the data.
-\end{quote}
-
-\begin{revcomment}[2.14]
-    Section 3\\
-    \begin{itemize}
-        \item Page 11, Line 42 - I think you are saying you removed these two guidelines 
-        from the analysis of all projects (as shown in Figure 4) but this paragraph seems 
-        to say this was only for projects with large code bases.
-    \end{itemize}
-\end{revcomment}
-
-\Us We started with all the guidelines but after analyzing the BCH results, 
-we found that some of the results for the \emph{Keep Your Codebase Small} guideline 
-were way above the limit set by BCH ($20$ Person-years). We suspect this threshold 
-may not be well calibrated. Thus, we decided not to consider 
-this guideline in our research. We did not want 
-to lose more data points. Thus, we decided to not consider the guideline for all projects.
-
-We also did not consider the \emph{Automated Tests} guideline because
-this guideline does not contemplate vulnerability scanning and penetration testing
-which are the two most important testing techniques in security.
-
-We improved the paper in the following paragraph:
-
-\begin{quote}
-    Due to BCH limitations, in particular, lack of language support 
-    and project size by BCH, $308$ data points were not analyzed and automatically 
-    disregarded from our study. After performing the BCH analysis and 
-    the maintainability calculations, we found the following 
-    limitations regarding two of BCH's guidelines:
-
-    1) For projects with large codebases, the results calculated for 
-    the \emph{Keep Your Codebase Small} guideline were way above the 
-    limit set by BCH ($20$ Person-years). We suspect this threshold 
-    may not be well calibrated. Thus, we decided not to consider 
-    this guideline in our research.
-
-    2) The 
-    \emph{Automated Tests} guideline was also not considered since the 
-    tool does not contemplate two of the most important techniques to 
-    security testing: vulnerability scanning and penetration testing. 
-    Instead, it only contemplates unit testing. 
-\end{quote}
-
-\begin{revcomment}[2.15]
-    Section 3\\
-    \begin{itemize}
-        \item Page 11  line 48.  How many were disregarded?
-    \end{itemize}
-\end{revcomment}
-
-\Us Due to BCH limitations, in particular, lack of language support 
-and project size by BCH, $308$ data points were disregarded.
-We clarified it in the paper (Section 3.4):
-
-\begin{quotation}
-    Due to BCH limitations, in particular, lack of language support 
-    and project size by BCH, $308$ data points were not analyzed and automatically 
-    disregarded from our study. 
-\end{quotation}
-
-\begin{revcomment}[2.16]
-    Section 4\\
-    \begin{itemize}
-        \item Page 13, Line 3 - 969 security commits and 969 
-        baseline commits?  It's hard to know how to parse that.
-    \end{itemize}
-\end{revcomment}
-
-\Us We mean that our study is an empirical evaluation of $969$ security patches
-and $969$ regular changes. We improved the sentence.
-
-\begin{quote}
-    This study evaluates a total of $969$ security patches and 
-    $969$ regular changes 
-    from $260$ distinct open-source projects. 
-\end{quote}
-
-\begin{revcomment}[2.17]
-    Section 4\\
-    \begin{itemize}
-        \item Page 13, Line 31 "overall patches …" rather than having us 
-        rely on looking at the Figure, tell us the number or percentage of 
-        when the impact is positive like you do with the negative in the next 
-        sentence. [You do provide the specifics starting on page 15 line 49.  
-        I wanted it here.]
-    \end{itemize}
-\end{revcomment}
-
-\Us We replaced overall by the number of patches where maintainability 
-increases ($38.7\%$).
-
-\begin{quote}
-    Regarding the impact of security patches per guideline, we 
-observe that $38.7\%$ of the security patches have positive impact
-on software maintainability.
-\end{quote}
-
-\begin{revcomment}[2.18]
-    Section 4\\
-    \begin{itemize}
-        \item Page 19, Line 7 - for the unindoctrinated - I'd 
-        explain the concept of Research Concepts. Also explain 
-        your choice of 7a and 7b and how representative they are 
-        from the several hundred choices. Similar for your explanations 
-        in the second two paragraphs on this page.  How far down the 
-        concept/vulnerability hierarchy tree are these CWE choices or 
-        are they lower level vulnerabilities chosen from the 700+ CWE types?
-    \end{itemize}
-\end{revcomment}
-
-\Us We added an explanation of the Research Concepts list to the paper
-in a footnote:
-
-\begin{quote}
-    Research Concepts is a tree-view provided by the Common Weakness 
-    Enumeration (CWE) website that intends to facilitate research into 
-    weaknesses. It is organized 
-    according to 
-    abstractions of behaviors instead of how they can be detected, 
-    their usual location in code, and when they are introduced in the 
-    development life cycle. Available here: 
-    \url{https: //cwe.mitre.org/data/definitions/1000.html}
-\end{quote}
-
-In Figure 7a, we present the entire distribution of the $969$ vulnerabilities per
-each node of the 1st-level of the Research Concepts tree.
-Looking at those results, we detected that the percentage of cases for CWE-707 ($30.4\%$) 
-and CWE-664 ($32.8\%$) were considerably higher in the dataset than the rest of the CWEs.
-Thus, we decided to look at lower levels of behavior for both CWEs: CWE-707
-results are presented in Figure-7b and CWE-664 results are presented in Figure-7c.
-In Figure-7b, all the CWEs presented are sub levels of the CWE-707. The same
-happens for Figure-7c, but the CWEs are sub levels of the CWE-664.
-
-We added the previous numbers to the paper and clarified that we are considering lower levels
-of vulnerabilities.
-
-\begin{quote}
-    In \emph{RQ2}, we report/discuss the impact of security patches on
-    software maintainability per weakness (CWE). We use the weakness definition
-    and taxonomy proposed by the \emph{Common Weakness Enumeration} (cf. Section 2).
-    Figure 7 shows three different charts. Figure \emph{7-a}, presents
-    the impact of the $969$ patches grouped by the first level weaknesses from
-    the \emph{Research Concepts} list. While the Figures \emph{7-b} and
-    \emph{7-c} present the impact on maintainability for lower levels of 
-    weaknesses for the most prevalent weaknesses in Figure \emph{7-a}:
-    \emph{Improper Neutralization} (CWE-707) 
-    and \emph{Improper Control of a Resource 
-    Through its Lifetime} (CWE-664), respectively.
-
-    In Figure~7-\emph{a}, there is no clear evidence of the impact on 
-    maintainability per weakness. Yet, it is important to note that
-    overall there is a very considerable number of cases that hinder
-    maintainability---between $30\%$ and $60\%$. The CWE-707 and CWE-664 
-    weaknesses integrate the higher number of cases compared to the remaining 
-    ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
-    Thus, we present an analysis of their sub-weaknesses on 
-    Figure~7-\emph{b} and Figure~7-\emph{c}, respectively. 
-\end{quote}
-
-\begin{revcomment}[2.19]
-    Section 4\\
-    \begin{itemize}
-        \item Page 20, Line 42 - how many is "a considerable number of cases"?  
-    \end{itemize}
-\end{revcomment}
-
-\Us In total, we inspected around $25$ regular changes with no impact on software maintainability  
- to understand what was leading to the higher number of cases. This point
- was clarified in Section 4, with the following paragraph:
-
- \begin{quote}
-    Overall, the results for both baselines, show that regular 
-    changes are less prone to hinder the software 
-    maintainability of open-source software.   
-    However, the \emph{size-baseline} integrates a larger 
-    number of cases with no impact on software maintainability. 
-    We manually inspected a total of $25$ cases from
-    that distribution of regular changes with no impact on maintainability, 
-    and, found that identifying regular changes with the same 
-    size as the security-related commit is limiting the type of regular 
-    commits being randomly chosen: input 
-    patches, variables or functions, type conversion (i.e., changes
-    with no impact on the software metrics analyzed by BCH).
- \end{quote}
-
-\begin{revcomment}[2.20]
-    Section 4\\
-    \begin{itemize}
-        \item Page 20, Line 45 - it seems choosing commits of similar size 
-        to the security commits ended up causing a limitation.  I'm still 
-        wondering why you chose that methodology anyway.  
-    \end{itemize}
-\end{revcomment}
-
-\Us As we described in comment $2.19$, we manually inspected $25$ cases.
-Many of those cases were changes of small size ($< 10$ LOCs) where
-refactorings were usually improving names of variables, functions, cases of type
-conversion, etc. Changes that have no impact on the metrics we are 
-evaluating.
-
-Our concern by presenting a total random baseline 
-was that comparing regular changes with security changes with very different sizes---a scenario 
-that could happen---was unfair. Thus, we chose to create this baseline.
-However, we also curated a total random baseline in the past and we now 
-present both baselines in the paper.
-
-\begin{revcomment}[2.21]
-    Section 4\\
-    \begin{itemize}
-        \item RQ3 is answered to succinctly and with too little 
-        detail to be convincing to me that the differences between 
-        security and regular changes is different.  
-    \end{itemize}
-\end{revcomment}
-
-\Us We improved the discussion of the comparison between 
-the results for security patches and regular changes. We now
-consider two baselines: \emph{size-baseline}, where 
-we collected random changes with the same size as security 
-patches; and, \emph{random-baseline}, where we collected
-random changes without any restrictions. The overall results
-show that regular changes are less prone to hinder software 
-maintainability than security patches. Security
-patches results have more cases with negative 
-impact on software maintainability than any baseline 
-of regular changes. By providing results 
-for both baselines, we now clearly see that the limitation 
-found for the \emph{size-baseline} does not change 
-the need to pay more attention to security patches 
-in software maintenance tasks.
-
-\begin{revcomment}[2.22]
-    Section 5\\
-    \begin{itemize}
-        \item Line 32 - since RQ3 was not convincing, 
-        I don't feel a separate effort for maintainable security 
-        is justified as separate from general focus on maintainability. 
-    \end{itemize}
-\end{revcomment}
-
-\Us As explained in the previous response, security patches 
-hinder more software maintainability than regular changes with size 
-restriction (\emph{size-baseline}) or no size restriction (\emph{random-baseline}). Thus, we argue that
-special attention should be given to maintainable security
-when teaching software maintainability. We improved the analysis 
-provided in Section 4 for the research question 3 (RQ3). We added 
-an extra baseline for comparison with no size restrictions (\emph{random-baseline}) and 
-improved the discussion over our results.
-
-\begin{revcomment}[2.23]
-    Section 6\\
-    \begin{itemize}
-        \item It might not be a threat - but I feel the curation 
-        of the non-security commits is a limitation to the generalizability 
-        of the characteristics of non-security commits
-    \end{itemize}
-\end{revcomment}
-
-\Us We agree with the reviewer. Our concern by presenting a total random baseline 
-was that 
-comparing regular changes with security changes with very different sizes---a scenario 
-that could happen---was unfair. 
-Nevertheless, we also feel that considering size may be limitating the generalizability
-of the characteristics of regular changes. Thus, we added to RQ3 our initial random 
-baseline and completed the work by presenting now two baselines (size-baseline and random-baseline).
-
-\newreviewer{3}
-
-\begin{revcomment}[3.1]
-
-    The paper addresses a relevant and up to date topic. 
-    The authors have analyzed a large set of project wrt to 
-    quality assessment focused on vulnerabilities in line with 
-    ISO25000 standard. There are some points worth considering: 
-    - authors should better motivate the choice for BCH as web 
-    based source code analysis service with respect to other 
-    options that are available in literature. Tools such as Kiuwan, 
-    sonar cloud, codify etc are also able to provide specific metrics 
-    on maintainability and security, vulnerability features of code. 
-    Furthermore Kiuwan is also an example designed based on the ISO25000 
-    standard. Authors discuss the need for tools for risk assessment in 
-    the study implications (section5) but fail to assess why their choice 
-    falls on BCH. Furthermore all tools provide different conceptualization 
-    and evaluation criteria for calculating quality characteristics such as Vulnerabilities 
-    and Security. Authors should also explicit how these quality characteristics are measured in BCH.
-
-\end{revcomment}
-
-\Us We kindly thank the reviewer for the comments.
-
-As mentioned in the paper, BCH checks GitHub codebases against $10$ engineering 
-guidelines as devised by Software Improvement Group (SIG). SIG has devised 
-these guidelines after many years of experience: analyzing more than $15$ million 
-lines of code every week, SIG maintains the industry’s largest 
-benchmark, containing more than $10$ billion lines of code across $200$+ 
-technologies; SIG is the only lab in the world certified by TÜViT to issue ISO 25010:2011
-certificates for Trusted Product Maintainability\footnote{Information available here: 
-\url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
-
-% Each of the guidelines in isolation is easy to understand, but, when 
-% combined and evaluated as a whole, it becomes easy to see why 
-%they would comprise the definition of good software. 
-Furthermore, the relation of the entire set of
-guidelines to maintainability is well detailed in the O'Reilly's book ``Building Maintainable 
-Software: Ten Guidelines for Future-Proof Code'' by Professor Joost Visser
-\footnote{Available here: 
-\url{https://www.softwareimprovementgroup.com/resources/ebook-building-maintainable-software/}}.
-
-In this study, to clarify, we only focus on measuring software maintainability, we do not focus 
-on vulnerability detection or measuring security. Instead, we want to understand how software 
-maintainability is impacted by security patches. However, as we explained before, BCH is 
-based on the ISO 25010 which considers \emph{Security} as one of the main software product quality 
-characteristics since $2011$.
-
-There are other tools that measure the same metrics such as as Kiuwan; or, even 
-provide a model to calculate maintainability such as SonarCloud. In this study, it is not
-our goal to study other tools. However, we argue
-that BCH is a better tool for the following reasons:
-
-\begin{itemize}
-    \item Kiuwan measures 4 guidelines that by name appear to be similar to 
-    some of the guidelines measured by BCH: Complexity, Duplicated 
-    Code, Size and Coupling\footnote{List available here: \url{https://www.kiuwan.com/docs/display/K5/Metrics}}. 
-    However, we did not find the full description of kiuwan's metrics anywhere. Thus, 
-    we are ensure that the metrics are the same but we are able to conclude that
-    BCH integrates a larger set of metrics.  
-    In contrast, 
-    SIG provides a full description of the $10$ guidelines/metrics behind better code
-    hub plus their relation with maintainability on the book and several scientific papers. 
-    \item SonarCloud measures 5 similar metrics to the ones measured by BCH: Complexity, Duplication, 
-    Issues, Size and Tests\footnote{List available here: \url{https://sonarcloud.io/documentation/user-guide/metric-definitions/}}. 
-    The tool rates maintainability with a letter based on the technical 
-    debt ratio which is calculated as the ratio between the cost to develop 
-    the software and the cost to fix it. But no information is provided 
-    on how the remediation cost and developmet cost are calculated. 
-    Better Code Hub contemplates a larger set of maintainability metrics 
-    and our formula for maintainability is not only fully explained in our paper but also 
-    based on metrics that are fully and publicly described.  
-\end{itemize}
-
-We do not know codify and did not find the tool anywhere in static analysis tools 
-collections or even through an organic search at Google.
-
-We added a Section to 
-the paper called \textit{3.3 Better Code Hub} where we motivate the Better Code Hub tool and 
-compare it with Kiuwan and SonarCloud.
-
-\begin{revcomment}[3.2]
-
-    In section 3a better state/formulate the hypothesis authors 
-    analyze wrt to the Research questions in previous sections. 
-    In general the empirical study is not structured according to any 
-    of the guidelines or explicit states any of them.
-
-\end{revcomment}
-
-\Us In our work, we follow the same methodology as the previous work 
-mentioned below but instead of measuring
-the impact of energy-oriented changes on software maintainability, 
-we measure the impact of security patches.
-
-\begin{verbatim}
-    Luis Cruz, Rui Abreu, John Grundy, Li Li and Xin Xia. 
-    ``Do Energy-oriented Changes Hinder Maintainability?.'' 
-    IEEE International Conference on Software Maintenance and 
-    Evolution (ICSME). 
-    IEEE, 2019.
-\end{verbatim}
-
-Improving software security is not a trivial task and 
-requires implementing patches that might affect software 
-maintainability. Our hypothesis is that some of these patches may 
-have a negative impact on the software maintainability. With this 
-study, we want to understand which maintainability metrics can guide 
-security engineers on producing patches with better quality---which 
-we evaluate with \textit{\textbf{RQ1: What is the impact of security patches on the
-maintainability of open-source software?}}. In this research 
-question, we present the results by metric, overall score, severity and programming language
-to have a better understanding of the impact of each of these parameters in 
-software maintainability.
-
-Different weaknesses may require patches with different characteristics. 
-Our hypothesis
-is that security patches for different weaknesses can have different 
-impacts on software maintainability. Thus, we evaluate this in \textit{\textbf{RQ2: Which weaknesses are more likely to
-affect open-source software maintainability?}}. The results can help
-security engineers prioritizing weaknesses and bring awareness to the 
-ones that need more attention.
-
-Performing a regular change/refactoring, for instance, to improve the name of 
-a variable or function, is different than performing a security patch.
-Therefore, we created \textit{\textbf{RQ3: What is the impact of security patches versus 
-regular changes on the maintainability of open-source software?}}
-to explore if there is a need to pay more attention to 
-security patches than regular changes during software maintenance tasks.
-
-The previous information is already included in the paper. However, 
-we clarified these points in Section 2, where we present the research questions 
-and our hypothesis.
-
-\begin{revcomment}[3.3]
-
-    The findings of the research should better emphasize what the 
-    lessons learned and what better practices should be put in place 
-    by developers to assure better quality of software. The take away 
-    message remains hindered.
-
-\end{revcomment}
-
-\Us We agree with the reviewer that we should better emphasize the lessons 
-learned and what best practices can be followed by developers to assure better 
-vulnerability patches. We improved the discussion of these points in Section 5.
+\Us Thank you for pointing this out. We changed the work for ``consider''.
+This metric only considers unit testing, which is usually not applied for 
+security testing but rather to functionality testing.
 

### 2021-04-30 paper/emse20.tex
adding final review draft

@@
 \emph{Automated Tests} guideline was also not considered since the 
-tool does not contemplate two of the most important techniques to 
+tool does not consider two of the most important techniques to 
 security testing: vulnerability scanning and penetration testing. 
-Instead, it only contemplates unit testing.
+Instead, it only considers unit testing.
 

### 2021-04-30 paper/emse20.tex
grammarly

@@
 open-source software. Maintainability is measured based on the 
-Better Code Hub’s model of 10 guidelines on a dataset including 
+Better Code Hub’s model of 10 guidelines on a dataset, including 
 1300 security-related commits. Results show evidence of a trade-off 
@@
 characteristics include, but are not limited to functional 
-correctness, reliability, usability, maintainability, evolvavility 
+correctness, reliability, usability, maintainability, evolvability 
 and security. Security is an essential non-functional requirement 
@@
 Complex code is difficult to understand, maintain and 
-test~\cite{1702388}. Thus, the task of a developer gets harder as the 
+test~\cite{1702388}. Thus, the task of a developer gets more challenging as the 
 codebase grows in size and complexity. But the risk can be minimized 
@@
 to compute the software compliance with a set of $10$ 
-guidelines/metrics to produce quality software based in ISO/IEC 
+guidelines/metrics to produce quality software based on ISO/IEC 
 $25010$~\cite{Visser:2016:OREILLY}. SIG 
@@
 
-Static analysis tools (SATs) have been built to automatically detect software 
-vulnerabilities (e.g., FindBugs, Infer and more). Developers
+Static analysis tools (SATs) have been built to detect software 
+vulnerabilities automatically (e.g., FindBugs, Infer, and more). Developers
 use those tools to locate the issues in the code. However,
-while performing the patches to those issues, SATs are not 
-capable of providing information on the quality of the patch. 
-Improving software security is not a trivial task and 
-requires implementing patches that might affect software 
-maintainability. Our hypothesis is that some of these patches may 
+while performing the patches to those issues, SATs cannot provide 
+information on the quality of the patch. 
+Improving software security is not a trivial task and requires 
+implementing patches that might affect software maintainability. 
+We hypothesize that some of these patches may 
 have a negative impact on the software maintainability and, 
@@
 vulnerabilities on software maintenance across open-source software.
-We arguee that tools that assess these type of code metrics may
+We argue that tools that assess these type of code metrics may
 complement SATs with valuable information to help the developer
@@
 \emph{C/C++}, \emph{Ruby} and \emph{PHP}, may have a more negative 
-impact in software maintainability. Little information is known 
+impact on software maintainability. Little information is known 
 about the impact of security patches on software 
@@
 vulnerabilities. The harm of maintainability can increase the time 
-of response of future mitigations or even other maintainability 
+of response of future mitigations or even of other maintainability 
 tasks. Thus, it is of utmost importance to find means to assist 
@@
 	\item A replication package with the scripts and data created to 
-	perform the empirical evaluation, for reproducibility. Available 
+	perform the empirical evaluation for reproducibility. Available 
 	online: \url{https://github.com/TQRG/maintainable-security}.
@@
 client could cause a Denial-of-Service (DoS) attack via large Online 
-Certificate Status Procotol (OCSP) Status Request extensions when 
+Certificate Status Protocol (OCSP) Status Request extensions when 
 OCSP stapling support was enabled. OCSP stapling, formally known 
@@
 instead of after. Thereby, the system avoids the increase of 
-unbounded memory (and, a potential DoS attack).
+unbounded memory (and a potential DoS attack).
 
@@
 a considerable amount of changes were performed in the codebase 
-which yielded to a negative impact on software maintainability.
+which yielded a negative impact on software maintainability.
 While patching, the developer introduced $6$ new lines in a method 
 already with a large number of lines of code and introduced more 
-complexity to the code with $2$ new branch points which disrupt two of 
+complexity to the code with $2$ new branch points, which disrupt two of 
 the guidelines proposed by the Software Improvement Group 
@@
 Software maintainability is designated as the degree to which an 
-application is understood, repaired or enhanced. In this paper, our 
-concern is to study whether, while improving software security, 
+application is understood, repaired, or enhanced. In this paper, our 
+concern is to study whether while improving software security, 
 developers are also hindering software maintainability. This is 
@@
 \textit{\textbf{RQ1: What is the impact of security patches on the
-maintainability of open-source software?}} Often security flaws 
+maintainability of open-source software?}} Often, security flaws 
 require patching code to make software more secure. However, 
 \textbf{there is no evidence yet of how security patches impact the
-maintainability of open-source software}. Our hypothesis is that 
+maintainability of open-source software}. We hypothesize that 
 developers tend to introduce technical debt in their software when 
-patching software vulnerabilities because they tend to not pay enough 
+patching software vulnerabilities because they tend not to pay enough 
 attention to the quality of those patches. To address it, 
@@
 We present the maintainability impact by guideline/metric, overall score, 
-severity and programming language.
+severity, and programming language.
 
@@
 \texttt{htmlentities} to escape the data given by the variable
-\texttt{\$\_['file']}. Our hypothesis
-is that security patches for different weaknesses can have different 
-impacts on software maintainability. \textbf{Understanding which weaknesses are more 
-likely to increase maintainability issues is one step forward to 
-bring awareness to security engineers of what weaknesses need more 
-attention}. The taxonomy of security patterns used to answer this 
+\texttt{\$\_['file']}. We hypothesize that security patches for 
+different weaknesses can have different impacts on software maintainability. 
+\textbf{Understanding which weaknesses are more likely to increase maintainability 
+issues is one step toward bringing awareness to security engineers of what 
+weaknesses need more attention.} The taxonomy of security patterns used to answer this 
 question is the one provided by the Common Weakness Enumeration
@@
 Performing a regular change/refactoring, for instance, to improve the name of 
-a variable or function, is different than performing a security patch.
+a variable or function is different than performing a security patch.
 Therefore, we also computed the maintainability of random regular commits using 
@@
 first and last commits used to patch the vulnerability. For these 
-cases, we used the GitHub API to retrieve automatically the dates of 
-the commits. Then, for each patch, the group of commits was ordered 
+cases, we used the GitHub API to retrieve the dates of 
+the commits automatically. Then, for each patch, the group of commits was ordered 
 from the oldest commit to the newest one. We assumed the last commit 
@@
 commits before and after the security patching to evaluate if its 
-impact was positive, negative or none. The $1300$ patches in the 
+impact was positive, negative, or none. The $1300$ patches in the 
 dataset were analyzed using the BCH toolset to calculate their 
@@
 but with an approximate size as security patches---we argue 
-that comparing changes with considerable different sizes may be unfair.
+that comparing changes with considerably different sizes may be unfair.
 
@@
 As for the security patches, for each regular change, we  
-need the commit performing the regular change (\emph{sha-reg}) and version of 
-the software before the change (\emph{sha-reg-p}). For each security patch, a 
-random commit from the same project is selected, \emph{sha-reg}. The parent commit
+need the commit performing the regular change (\emph{sha-reg}) and version of the software before the change (\emph{sha-reg-p}). A 
+random commit from the same project is selected for each security patch, \emph{sha-reg}. The parent commit
 of \emph{sha-reg} is the \emph{sha-reg-p}.
@@
 
-BCH checks GitHub codebases against 10 maintainability 
+BCH checks GitHub codebases against $10$ maintainability 
 guidelines~\cite{Visser:2016:OREILLY} that were empirically 
@@
 SIG has devised these guidelines after many years of experience: analyzing more 
-than 15 million lines of code every week, SIG maintains the industry’s largest 
-benchmark, containing more than 10 billion lines of code across 200+ technologies; SIG 
-is the only lab in the world certified by TÜViT to issue ISO 25010 certificates\footnote{Information available here: 
+than $15$ million lines of code every week, SIG maintains the industry’s largest 
+benchmark, containing more than $10$ billion lines of code across $200$+ technologies; SIG 
+is the only lab in the world certified by TÜViT to issue ISO $25010$ certificates\footnote{Information available here: 
 \url{https://www.softwareimprovementgroup.com/methodologies/iso-iec-25010-2011-standard/}}.
 BCH's compliance criterion is derived from the requirements for 4-star 
-level maintainability (cf. ISO 25010)~\cite{5609747,6113040,baggen2012,Visser:2016:OREILLY}.
+level maintainability (cf. ISO $25010$)~\cite{5609747,6113040,baggen2012,Visser:2016:OREILLY}.
 SIG performs the threshold calibration
@@
 
-As BCH there are other tools that also perform code analysis for similar 
+As BCH, other tools also perform code analysis for similar 
 metrics. Two examples are Kiuwan and SonarCloud. However, Kiuwan does
@@
 description of their formula are not available. Both analyze less maintainability
-guidelines than BCH and do not have their inner-workings fully and publicly described.
+guidelines than BCH and do not have their inner workings fully and publicly described.
 
@@
 presented in previous work on the maintainability of energy-oriented 
-fixes~\cite{8919169}. The inner-workings of BCH were proposed
+fixes~\cite{8919169}. The inner workings of BCH were proposed
 originally in 2007~\cite{4335232} and suffered refinements later~\cite{5609747,6113040,baggen2012}. 
@@
 maintain~\cite{Visser:2016:OREILLY} and, maps each guideline to the 
-metric calculated by BCH. Theses guidelines are calculated using the 
-metrics presented in~\cite{criteria:2017} and are also briefly 
-explained on Table~\ref{tab:guidelines}. 
+metric calculated by BCH. These guidelines are calculated using the 
+metrics presented in~\cite{criteria:2017} and are also briefly explained in Table~\ref{tab:guidelines}. 
 During each guideline evaluation, the tool determines the compliance 
@@
 different severities of violating the guideline are indicated using 
-colors, and there is a legend to help to interpret them. The green 
+colors, and there is a legend to help interpret them. The green 
 bar represents the number of compliant branch points per unit 
 (\emph{at most $5$}), i.e., the number of units are compliant with 
-ISO $25010$~\cite{iso:2011}. Yellow, orange and red bars represent 
+ISO $25010$~\cite{iso:2011}. Yellow, orange, and red bars represent 
 units that do not comply with medium (\emph{above $5$}), high 
@@
 thresholds for each severity level. If the green mark is somewhere 
-in the green bar it means it is compliant with a low level of
-severity.
+in the green bar, it is compliant with a low severity level.
 
@@
 the \emph{Keep Your Codebase Small} guideline were way above the 
-limit set by BCH ($20$ Person-years). We suspect this threshold 
-may not be well calibrated, and hence biasing our results. Thus, we decided not to consider 
-this guideline in our research.
+limit set by BCH ($20$ person-years). We suspect this threshold 
+may not be well-calibrated, and hence biasing our results. Thus, we decided not to consider this guideline in our research.
 
@@
 The BCH tool does not compute the final score that our study needs 
-to compare maintainability amongst different project versions. For 
-this, we follow previous work on measuring the impact of 
+to compare maintainability amongst different project versions. We 
+follow previous work on measuring the impact of 
 energy-oriented patches~\cite{8919169}. Cruz et al. ($2019$) 
@@
 equation provided in~\cite{8919169} considers that the size of 
-project changes does not affect the maintainability and that the 
+project changes do not affect the maintainability, and that the 
 distance to lower severity levels is less penalized than to the
@@
 hypothesis that the maintainability difference between pairs of 
-versions $v_{s-1}$, $v_s$ (i.e., before and after a security-commit) 
-come from the same distribution. Nevertheless, this test 
-has a limitation: it does not consider the groups of commits with a 
-zero-difference maintainability. In $1959$, Pratt provided an 
-improvement to the test to solve this issue making the test more 
-robust. Thus, we use a version of the Wilcoxon test that 
+versions $v_{s-1}$, $v_s$ (i.e., before and after a security commit) 
+come from the same distribution. Nevertheless, this test has a 
+limitation: it does not consider the groups of commits with 
+zero-difference maintainability. In $1959$, Pratt improved the 
+test to solve this issue, making the test more robust. Thus, 
+we use a version of the Wilcoxon test that 
 incorporates the cases where maintainability is equal to 
@@
 \textbf{Guideline/Metric:} Each patch performs a set of changes
-on the source code of the software. These changes may have a 
+on the software's source code. These changes may have a 
 different impact on the guidelines/metrics used to measure software 
@@
 detail the metrics behind the guidelines. For each type of 
-guideline, a swarmplot is presented to show the variability and 
-dispersion of the results alongside the number of absolute and 
+guideline, a swarm plot is presented to show the variability/dispersion 
+of the results alongside the number of absolute and 
 relative cases of each impact. Next to each type of guideline, it is 
@@
 average impact of all guidelines. Each point of the plot represents 
-the impact of a security patch on software maintainability. Red, 
+the impact of a security patch on software maintainability. Red 
 means the impact was negative, i.e., the patch harmed 
-maintainability. Yellow, means the patch did not have any kind of 
-impact on maintainability. Green, means the impact was positive, i.e., 
+maintainability. Yellow means the patch did not have any kind of 
+impact on maintainability. Green means the impact was positive, i.e., 
 the patch improved software maintainability.
@@
     Each point of the plot represents the impact of a security patch on software 
-	maintainability. Red, means the impact was negative, i.e., the patch 
-	harmed maintainability. Yellow, means the patch did not have any 
-	kind of impact on maintainability. Green, means the impact was 
+    maintainability. Red means the impact was negative, i.e., the patch 
+    harmed maintainability. Yellow means the patch did not have any 
+    kind of impact on maintainability. Green means the impact was 
     positive, i.e., the patch improve software maintainability. For instance, in the 
@@
     security patches had no impact on maintainability; and, 
-	$38.18\%$ of security patches actually improved software 
-	maintainability.}
+    $38.18\%$ of security patches improved software maintainability.}
  	\label{fig:guidelines}	
@@
  on software maintainability. However, we also 
-see that patching vulnerabilities has a very significant number of 
+ see that patching vulnerabilities have a very significant number of 
  negative cases per guideline---between $10\%$ and $40\%$. 
@@
  ($33.8\%$) seem to be the most negatively affected guidelines. This 
-may imply that developers when patching vulnerabilities have a hard 
-time to design/implement solutions that continue to respect the 
+ may imply that developers, when patching vulnerabilities, have a hard 
+ time designing/implementing solutions that continue to respect the 
  limit bounds of branch points and function/module sizes that are 
-recommended by coding practices. Still on respecting bound limits, 
-developers also seem to not consider the limit of $4$ parameters per 
+ recommended by coding practices. Still, on respecting bound limits, 
+ developers also seem not to consider the limit of $4$ parameters per 
  function for the \emph{Keep Unit Interfaces Small} guideline 
@@
  violated when the patch requires to input new information to a 
-function/class and developers struggle to use the \emph{Introduce 
+ function/class, and developers struggle to use the \emph{Introduce 
  Parameter Object} patch pattern. Results do not provide statistical 
@@
 The \emph{Write Code Once} guideline results show that duplicated 
-code increased in $17.7\%$ ($171$/$969$) of the patches. Software systems 
-typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. Previous 
-work showed a correlation between code smells and code
+code increased in $17.7\%$ ($171$/$969$) of the patches. Software systems typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. Previous work showed a correlation between code smells and code
 duplication~\cite{7476787} which may also be reflected in the 
 \emph{Write Clean Code} guideline results. BCH reported new code 
-smells for $15.0\%$ ($145$/$969$) of the patches which according to previous work 
+smells for $15.0\%$ ($145$/$969$) of the patches, which according to previous work, 
 may be the source of new software vulnerabilities~\cite{8819456,10.1145/3133956.3134072}
@@
 it every time needed. The \emph{Extract Method} refactoring 
-technique solves many duplication problems. This makes spoting and 
-solving the issue faster because you only need to fix the method used
+technique solves many duplication problems. This makes spotting and solving the issue faster because you only need to fix the method used
 instead of locating and fixing the issue multiple times.
-Clone detection tools can also help on locating the issues.
+Clone detection tools can also help in locating the issues.
 
@@
 represents the impact of a security patch on software 
-maintainability. Red, means the impact was negative, i.e., the patch 
-harmed maintainability. Yellow, means the patch did not have any 
-kind of impact on maintainability. Green, means the impact was 
-positive, i.e., the patch improved software maintainability.
-The $M(v)$ plot show that $406$ ($41.9\%$) cases have a negative 
-impact on software maintainability. While $188$ ($19.4\%$) cases 
-have no impact at all and $375$ ($38.7\%$) have a positive impact on 
+maintainability. Red means the impact was negative, i.e., the patch 
+harmed maintainability. Yellow means the patch did not have any kind of impact on maintainability. Green means the impact was positive, i.e., the patch improved software maintainability.
+The $M(v)$ plot shows that $406$ ($41.9\%$) cases have a negative impact on software maintainability. While $188$ ($19.4\%$) cases 
+have no impact at all, and $375$ ($38.7\%$) have a positive impact on 
 software maintainability. The larger number of negative cases may be 
@@
 other severity levels. However, results should not be disregarded 
-because they seem to confirm the 
+because they somehow confirm the 
 assumption that higher severity vulnerabilities patches may have a 
-more negative impact on maintainability and, consequently, that high/medium 
+more negative impact on maintainability, i.e., high/medium 
 severity vulnerabilities may need more attention than low severity while 
-patching vulnerabilities.
+patching.
 
@@
 hard time following the guidelines and, consequently, hinder
-software maintainability while patching vulnerabilities; and, that different levels of
+software maintainability while patching vulnerabilities; and that different levels of
 attention should be paid to each guideline. For instance,
 \emph{Write Simple Units of Code} and \emph{Write Short Units of Code}
-are the most affected ones. No statistical significance was observed
+guidelines are the most affected ones. No statistical significance was observed
 for \emph{Separate Concerns in Modules}.
@@
 maintainability---between $30\%$ and $60\%$. The CWE-707 and CWE-664 
-weaknesses integrate the higher number of cases compared to the remaining 
-ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
+weaknesses integrate the higher number of cases compared to the remaining ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
 Thus, we present an analysis of their sub-weaknesses on 
@@
 The impact of a patch depends on its complexity, i.e., if the patch
-adds complexity to the code base it is probably affecting the software
+adds complexity to the code base, it is probably affecting the software
 maintainability. \emph{Cross-Site Scripting (CWE-79)} and 
@@
 the software maintainability---$32.4\%$, $40.7\%$ and, $41.1\%$, respectively---which
-should not be happening. Typically, CWE-79 vulnerabilities do not need extra lines 
-to be fixed, as shown in Listing~\ref{lst:fix}---one simple 
+should not be happening. Typically, CWE-79 vulnerabilities do not need extra lines to be fixed, as shown in Listing~\ref{lst:fix}---one simple 
 \texttt{escape} function patches the issue. On the same type of fix,
 CWE-199 vulnerabilities may also be fixed without adding new source code
-(e.g., replacing the \texttt{strncpy} function by a more secure one 
+(e.g., replacing the \texttt{strncpy} function with a more secure one 
 \texttt{strlcpy} that checks if the buffer is null-terminated). However,
@@
 As CWE-199 weaknesses, \emph{Missing Release of Memory after Effective 
-Lifetime (CWE-401)} can also be the cause of Denial-of-Service attacks and  
-difficult to patch since it usually requires to add complexity to the program (cf. Section~\ref{sec:motivation}).
+Lifetime (CWE-401)} can also be the cause of Denial-of-Service attacks and difficult to 
+patch since it usually requires adding complexity to the program (cf. Section~\ref{sec:motivation}).
 

### 2021-04-30 response/response_2/response.tex
small fix
small changes to response 2 and paper

@@
 \Us A person-year is the equivalent to 12 person-months. A person-month is a standard measure of 
-source code volume where the total volume in a codebase is the volume in lines of code
-converted to person-months. BCH uses person-years instead of volume 
+source code volume. It is the amount
+of source code that one developer with average productivity could write in one
+month. According to the "Building
+Maintainable Software" book, 20 person-years of rebuild value 
+is the equivalent to at most 175000 lines of code in Java. BCH uses person-years instead of volume 
 because it facilitates the comparisons of source code volume between technologies
 (different programming languages have different levels of of verbosity).
-We do not consider this metric on our study. However, according to the "Building
-Maintainable Software" book, 20 person-years of rebuild value 
-is the equivalent to at most 1750000 lines of code in Java.
 
 \begin{revcomment}[2.2]
-
     The bullet for automated tests.  Can you think of another work 
     instead of "contemplate"?  I'm not sure of what you are trying to say.
-
 \end{revcomment}
 
-\Us Thank you for pointing this out. We changed the work for ``consider''.
+\Us Thank you for pointing this out. We changed the work for ``include''.
 This metric only considers unit testing, which is usually not applied for 

### 2021-04-30 paper/emse20.tex
small changes to response 2 and paper

@@
 and project size by BCH, $308$ data points were not analyzed and automatically 
-disregarded from our study. After performing the BCH analysis and the maintainability calculations, we found the following 
+disregarded from our study. After performing the BCH analysis and the 
+maintainability calculations, we found the following 
 limitations regarding two of BCH's guidelines:
@@
 limit set by BCH ($20$ person-years). We suspect this threshold 
-may not be well-calibrated, and hence biasing our results. Thus, we decided not to consider this guideline in our research.
+may not be well-calibrated, and hence biasing our results. Thus, 
+we decided not to consider this guideline in our research.
 
@@
 \emph{Automated Tests} guideline was also not considered since the 
-tool does not consider two of the most important techniques to 
+tool does not include two of the most important techniques to 
 security testing: vulnerability scanning and penetration testing. 
-Instead, it only considers unit testing.
+Instead, it only integrates unit testing.
 
@@
 than regular changes}. 
-We have seen, previously, a deterioration on software maintainability 
+We have seen, previously, a deterioration in software maintainability 
 when patching vulnerabilities: $41.9\%$ ($406$) of patches suffered a 
-negative impact, $38.7\%$ ($375$) of patches remained the same and $19.4\%$ 
+negative impact, $38.7\%$ ($375$) of patches remained the same, and $19.4\%$ 
 ($188$) of patches increased software maintainability. For regular changes,
@@
 security patches, the maintainability of regular changes remains the 
-same in $42.5\%$ ($412$) of the cases and perform regular
+same in $42.5\%$ ($412$) of the cases, i.e., performing regular
 changes has a more positive impact than negative on  
@@
 restrictions, are less prone to hinder software maintainability than 
-security changes. About $34.4\%$ ($333$) of the regular changes 
-have negative impact on software maintainability---less than 
-in the security patches. For the \emph{random-baseline}, 
+security changes. About $34.4\%$ ($333$) of the regular changes hinder 
+software maintainability---less than in the security patches. For the \emph{random-baseline}, 
 statistical significance was retrieved ($p = 5.34$x$10^{-8}$).
@@
 
-Overall, the results for both baselines, show that regular 
+Overall, the results for both baselines show that regular 
 changes are less prone to hinder the software 
@@
 that distribution of regular changes with no impact on maintainability, 
-and, found that identifying regular changes with the same 
+and found that identifying regular changes with the same 
 size as the security-related commit is limiting the type of regular 
@@
 identifying regular changes without any restrictions (\emph{random-baseline})
-shows that regular changes have less negative impact 
-on software maintainability compared with security patches
+shows that regular changes have a less negative impact 
+on software maintainability when compared to security patches
 and that special attention should be given to security patches.
@@
 %
-
 \textit{\textbf{Follow Best Practices:}} Developers are not 
@@
 recommended by best practices to avoid increasing the 
-size, complexity and dependencies of their patches.
+size, complexity, and dependencies of their patches.
 Developers should also keep function parameters below 
@@
 
-Security patches also have a negative impact on the maintenance of 
-software architecture. Maintaining the components independence and balance 
+Security patches also harm the maintenance of software architecture. Maintaining the components independence and balance 
 is important to make it easier to find the source code that 
 developers want to patch/improve and to better understand
-how the high-level componenet interact with others. Applying 
+how the high-level components interact with others. Applying 
 \emph{encapsulation} to hide implementation details and make the system 
@@
 it every time needed. The \emph{Extract Method} refactoring 
-technique solves many duplication problems. This makes spoting and 
+technique solves many duplication problems. This makes spotting and 
 solving the issue faster because you only need to fix one method instead
@@
 \textit{\textbf{Some Types of Vulnerabilities Need More Attention:}} 
-Our study attempted at shedding light at the impact 
-of different types of vulnerabilities on software maintainability.
+Our study attempted to shed light on the impact of different types of vulnerabilities on software maintainability.
 Overall, all the CWEs under study present a negative impact
@@
 Restriction of Operations within the Bounds of a Memory Buffer} (CWE-119) 
-are less prone to have impact on open-source software maintainability.
+are less prone to have an impact on open-source software maintainability.
 Developers should pay special attention to \emph{Improper Input Validation} (CWE-20), 
@@
 (CWE-401) and \emph{Path Traversal} (CWE-22). However, 
-more research should be performed to better understand 
-the impact of each guideline on each CWE.
+more research should be performed to better understand the impact of each guideline on each CWE.
 
@@
   software quality~\cite{10.1145/1985362.1985366}. Some software 
-	producers consider security as a first class citizen while others 
-	do not. As mentioned in previous work, security it is very important 
-	and should be considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}. 
-	However, the lack of experts and awareness to the ease of producing 
-	vulnerable software leads companies to ship low quality software. 
-	Providing automated tools to developers assess the risk of their 
-  patches it is important to help companies shipping software of higher quality. 
-  Bryan O’Sullivan, VP of Engineering at Facebook, advocated for new computer science
-  risk models to detect vulnerabilities in scale and predict the level of security of the software under production in
-  his talk “‘Challenges in Making Software Work at Scale”’ at FaceTAV’20.
+  producers consider security as a first-class citizen while others do not.
+As mentioned in previous work, security is critical and should be considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}.
+  However, the lack of experts and awareness of developers for security while producing/patching software leads companies to ship low-quality software. 
+  Providing automated tools to developers to assess the risk of their patches is essential to help companies shipping software of higher quality. 
+  Bryan O'Sullivan, VP of Engineering at Facebook, advocated for new computer science
+  risk models to detect vulnerabilities in scale and predict the level of security of the software under production in his talk ```Challenges in Making Software Work at Scale''' at FaceTAV’20.
 
   Tools like Better Code Hub can complement static analysis (e.g., 
-  SonarQube, Codacy, ESLint, Infer, and more) to provide more information 
-  to security engineers on the vulnerabilities per se and the risk of their patches
-  ---alongside pull requests/code reviews.
-  Static code analysis may be daunting due 
-	to the number of rules and different effects
-	on maintainability. BCH claims to use the top-$10$ of the guidelines
-	with the highest effects on maintainability. Yet, static analysis 
-	can be used to complement the BCH analysis by introducing
-	the capability of vulnerability detection and support the prediction 
-	and prevention of the risk of a patch of hindering maintainability.
-%
-
-\textit{\textbf{Computer Science Curricula Needs to be Updated:}}
-	Computer Science curricula is not yet prepared to properly educate 
-	students for maintainable security. Students should be exposed to 
-	this problem to gain experience with it. Curricula should focus on 
-	the production of secure and maintainable code and alert to the 
-	trade-off between both. These matters should be discussed and 
-	presented to students in software engineering courses.
-	Universities should also encourage students to use code quality 
-	analysis tools such as BCH or similar ones. These tools have a 
-	great potential to make students aware of maintainability issues 
-	and beginner mistakes (e.g., coding practices violation).
-%
-
-\textit{\textbf{Secure Programming Languages by Design:}} 
-In general, our study 
-shows that there is over $35\%$ of patches with negative impact 
-on software maintainability per programming language. Developers implementing code in   
-\texttt{C/C++}, \texttt{Ruby} and \texttt{PHP} should pay extra care 
-to their solutions/patches. In addition, programming 
-	languages should provide new design patterns to easily patch security weaknesses 
-	without endangering maintainability. Ultimately, new programming languages, 
-	both secure and maintainable by design, such as Wyvern~\cite{10.1145/2489828.2489830,kurilova2014wyvern}, 
-	should be designed to help developers be less vulnerability prone when 
-	writing and maintaining secure applications. One example is the need for 
-	designing new authorization mechanisms since these are one of the most 
-	complex security features to implement.
+  SonarQube, Codacy, ESLint, Infer, and more) to provide more informatio
 
@@
 %
-This section presents the potential threats to validity of this study.
+This section presents the potential threats to the validity of this study.
 %
@@
 \textit{\textbf{Construct Validity:}} The formula to calculate the maintainability value ($M(v)$) was inferred based 
-on the BCH's reports. The high amount of different projects and backgrounds 
-may require other maintainability standards. However, BCH does use a 
-representative benchmark of closed and open-source software projects to compute 
-the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,baggen2012}.
+on the BCH's reports. The high amount of different projects and backgrounds may require other maintainability standards. However, BCH does use a 
+representative benchmark of closed and open-source software projects to compute the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,baggen2012}.
 Maintainability is computed as the mean of all guidelines. Different software versions 
@@
 and still be affected by different guidelines. Therefore, we provide
-an analysis per guideline and our results are all available on figshare for future reproductions and deeper analysis.
+an analysis per guideline, and our results are all available on GitHub for future reproductions and deeper analysis.
 
@@
 patching vulnerabilities. This approach discards patches that were
-not explicit in commits messages. We assume that patches were performed 
-using a single-commit or several sequentially. The 
-perspective that a 
-developer may quickly perform a patch and later 
-proceed to the refactor is not considered. We assume 
-that all patches were only performed once. Depending 
-on the impact of the vulnerability in the system, some 
-vulnerabilities may have more urgency to be patched than 
-others. For instance, a vulnerability performing a Denial-of-Service attack
-which usually brings entire systems down may be more urgent to patch than a 
-cross-site scripting vulnerability which usually does not have impact
-on the systems execution but rather on the data accessibility.
-We manually inspected $25.1\%$ of the security patches looking 
-for floss-refactorings---$122$ from each dataset. We did 
+not explicit in commits messages. We assume that patches were performed using a single commit or several sequentially. The perspective that a 
+developer may quickly perform a patch and later proceed to the refactor is not considered.
+We assume that all patches were only performed once.
+Depending on the impact of the vulnerability in the system, some vulnerabilities may have more urgency to be patched than others.
+For instance, a vulnerability performing a Denial-of-Service attack
+that usually brings entire systems down may be more urgent to patch than a 
+cross-site scripting vulnerability which generally does not have an impact
+on the execution of the system but rather on the data accessibility.
+We manually inspected $25.1\%$ of the security patches looking for floss-refactorings---$122$ from each dataset. We did 
 find $23$ cases we argue to be floss-refactorings and toss them
-in order to minimize the impact of this threat.
+to minimize the impact of this threat.
 
 Baseline commits are retrieved randomly from 
-the same project of the security patch.
+the same project as the security patch.
 This approach softens the differences
@@
 
-\textit{\textbf{External Validity:}} The BCH tool uses private and open-source data to 
-determine the thresholds for each guideline. We only analyze patches of open-source software.
-Thus, our findings may not extend to private/non-open source software. Different programming 
-languages may require different coding practices to address software safety. The 
-dataset comprises more commits in Java, i.e., the dataset may not be representative 
-of the population regarding programming languages. For both datasets, a manual validation 
-of the message of the commits was performed. Only commits in English were considered. Thus, our approach does not consider 
+\textit{\textbf{External Validity:}} The BCH tool uses private and open-source data to determine the thresholds for each guideline. We only analyze patches of open-source software.
+Thus, our findings may not extend to private/non-open source software.
+Different programming languages may require different coding practices to address software safety.
+The dataset comprises more commits in Java, i.e., the dataset may not be representative of the population regarding programming languages.
+For both datasets, manual validation of the message of the commits was performed.
+Only commits in English were considered.
+Thus, our approach does not consider
 patches in any other language but English.
@@
 software quality. Previous work focused on object-oriented metrics has evaluated the
-impact of patches and exhibited proof that quantifying effectively the
+impact of patches and exhibited proof that quantifying the
 impact of patches on maintainability may help to choose the appropriate
@@
 and their differences against bug fixes~\cite{10.1145/3133956.3134072}.
-The main findings were that security patches are smaller and less complex compared
-to bug fixes and usually performed at function-level. Our study compares
+The main findings were that security patches are smaller and less complex than bug fixes and are usually performed at the function level. Our study compares
 the impact of security patches on software maintainability with the impact 
@@
 between maintainability aspects and proneness to changes/faults. In 2019, Cruz et 
-al.~\cite{8919169} proposed a formula to calculate maintainability 
-based on the BCH's guidelines and measured the impact of energy-oriented fixes 
+al.~\cite{8919169} proposed a formula to calculate maintainability based on the BCH's guidelines and measured the impact of energy-oriented fixes 
 on software maintainability. Recent work
-proposed a new maintainability model to 
-measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
-Our work uses the same base model (SIG-MM), but considers a broader set of guidelines. 
+proposed a new maintainability model to measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
+Our work uses the same base model (SIG-MM) but considers a broader set of guidelines. 
 Moreover, we solely focus on evaluating the impact of security patches on software maintainability.
@@
 Researchers investigated the relationship between design patterns and
-maintainability~\cite{10.1007/978-3-642-35267-6-18}. However, other studies show that 
-the use of design patterns may introduce maintainability issues into
-software~\cite{4493325}. Yskout et. al did not detect if the usage of 
-design patterns has a positive impact but concluded that developers prefer to 
-work with the support of security patterns~\cite{8077802}. The present work 
-studies how security weaknesses influence maintainability for open-source software.
+maintainability~\cite{10.1007/978-3-642-35267-6-18}. However, other studies show that the use of design patterns may introduce maintainability issues into
+software~\cite{4493325}.
+Yskout et.
+al did not detect if the usage of design patterns has a positive impact but concluded that developers prefer to work with the support of security patterns~\cite{8077802}.
+The present work studies how security weaknesses influence maintainability for open-source software.
 
@@
 one shows that some programming languages are more buggy-prone than others. However,
-the authors of the second one were not able to reproduce it and did not obtain any
+the authors of the second one could not reproduce it and did not obtain any
 evidence about the language design impact. 
@@
 patches on the maintainability of $260$ open-source projects. We leveraged
-Better Code Hub reports to calculate maintainability based on a model proposed in 
-previous work~\cite{Olivari:2018,8919169}. Results show evidence of a 
-trade-off between security and maintainability, as $41.9\%$ of security patches 
-yielded a negative impact. Hence,
-developers may be hindering software maintainability while patching 
-vulnerabilities. We also observe that some guidelines 
-and programming languages are more likely to be affected than others. The implications 
-of our study are that changes to codebases while patching 
-vulnerabilities need to be performed with extra care; tools
-for patch risk assessment should be integrate into the CI/CD pipeline; computer science
-curricula needs to be updated; and, more secure programming languages are necessary.
+Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018,8919169}. Results show evidence of a 
+trade-off between security and maintainability, as $41.9\%$ of security patches yielded a negative impact. Hence,
+developers may be hindering software maintainability while patching vulnerabilities.
+We also observe that some guidelines and programming languages are more likely to be affected than others.
+The implications of our study are that changes to codebases while patching vulnerabilities need to be performed with extra care; tools
+for patch risk assessment should be integrated into the CI/CD pipeline; computer science
+curricula need to be updated; and more secure programming languages are necessary.
 
@@
 investigate which guidelines affect most the maintainability per
-weakness; check if vulnerability patches are followed by patch
+weakness; check if vulnerability patches are followed by new
 commits and how much time does it take to do it; 

### 2021-07-14 paper/emse20.tex
adding acks

@@
 \setcounter{lstannotation}{0}
-\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} developers to a Cross-Site Scripting vulnerability},label={lst:fix}]
+\begin{lstlisting}[style={PHPStyle}, caption={Fix provided by \texttt{nextcloud/server} 
+  developers to a Cross-Site Scripting vulnerability},label={lst:fix}]
    <p class='hint'>
@@
 As for the security patches, for each regular change, we  
-need the commit performing the regular change (\emph{sha-reg}) and version of the software before the change (\emph{sha-reg-p}). A 
+need the commit performing the regular change (\emph{sha-reg}) and 
+version of the software before the change (\emph{sha-reg-p}). A 
 random commit from the same project is selected for each security patch, \emph{sha-reg}. The parent commit
@@
 The \emph{Write Code Once} guideline results show that duplicated 
-code increased in $17.7\%$ ($171$/$969$) of the patches. Software systems typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. Previous work showed a correlation between code smells and code
+code increased in $17.7\%$ ($171$/$969$) of the patches. Software systems 
+typically have $9\%$-$17\%$ of cloned code~\cite{5773403}. Previous work showed a correlation between code smells and code
 duplication~\cite{7476787} which may also be reflected in the 
@@
 it every time needed. The \emph{Extract Method} refactoring 
-technique solves many duplication problems. This makes spotting and solving the issue faster because you only need to fix the method used
+technique solves many duplication problems. This makes spotting and solving 
+the issue faster because you only need to fix the method used
 instead of locating and fixing the issue multiple times.
@@
 maintainability. Red means the impact was negative, i.e., the patch 
-harmed maintainability. Yellow means the patch did not have any kind of impact on maintainability. Green means the impact was positive, i.e., the patch improved software maintainability.
-The $M(v)$ plot shows that $406$ ($41.9\%$) cases have a negative impact on software maintainability. While $188$ ($19.4\%$) cases 
+harmed maintainability. Yellow means the patch did not have any kind of 
+impact on maintainability. Green means the impact was positive, i.e., the patch improved software maintainability.
+The $M(v)$ plot shows that $406$ ($41.9\%$) cases have a negative impact 
+on software maintainability. While $188$ ($19.4\%$) cases 
 have no impact at all, and $375$ ($38.7\%$) have a positive impact on 
@@
 for \emph{Separate Concerns in Modules}.
-As shown in Figure~\ref{fig:guidelines}, there is statistical significance ($p=0.044 < 0.05$) to support our findings: \textbf{security patches
-may have a negative impact on the maintainability of open-source software}. Therefore, tools such as BCH should be integrated into the CI/CD pipelines
+As shown in Figure~\ref{fig:guidelines}, there is statistical significance ($p=0.044 < 0.05$) 
+to support our findings: \textbf{security patches
+may have a negative impact on the maintainability of open-source software}. Therefore, tools 
+such as BCH should be integrated into the CI/CD pipelines
 to help developers evaluate the risk of patches of hindering software maintainability---alongside
 Pull Requests/Code Reviews. Different severity
-vulnerabilities may need different levels of attention---high/medium vulnerabilities need more attention (cf.
-Figure~\ref{fig:severity}).
-However, statistical significance was only observed for low severity vulnerabilities.
-Better and more secure programming languages are needed.
-We observed statistical significance for \emph{C/C++}, \emph{Ruby} and \emph{PHP} that support that security patches in those languages may hinder software maintainability (cf.
-Figure~\ref{fig:lang_main}).
+vulnerabilities may need different levels of attention---high/medium 
+vulnerabilities need more attention (cf. Figure~\ref{fig:severity}). However, 
+statistical significance was only observed for low severity vulnerabilities. Better 
+and more secure programming languages are needed. We observed statistical 
+significance for \emph{C/C++}, \emph{Ruby} and \emph{PHP} that support that 
+security patches in those languages may hinder software maintainability (cf. Figure~\ref{fig:lang_main}).
 %
@@
 maintainability---between $30\%$ and $60\%$. The CWE-707 and CWE-664 
-weaknesses integrate the higher number of cases compared to the remaining ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
+weaknesses integrate the higher number of cases compared to the remaining 
+ones: $295$ ($30.4\%$) data points and $318$ ($32.8\%$) data points, respectively. 
 Thus, we present an analysis of their sub-weaknesses on 
@@
 the software maintainability---$32.4\%$, $40.7\%$ and, $41.1\%$, respectively---which
-should not be happening. Typically, CWE-79 vulnerabilities do not need extra lines to be fixed, as shown in Listing~\ref{lst:fix}---one simple 
+should not be happening. Typically, CWE-79 vulnerabilities do not need extra lines 
+to be fixed, as shown in Listing~\ref{lst:fix}---one simple 
 \texttt{escape} function patches the issue. On the same type of fix,
@@
 %
+
 \textit{\textbf{Follow Best Practices:}} Developers are not 
@@
   software quality~\cite{10.1145/1985362.1985366}. Some software 
-  producers consider security as a first-class citizen while others do not.
-As mentioned in previous work, security is critical and should be considered as a default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}.
-  However, the lack of experts and awareness of developers for security while producing/patching software leads companies to ship low-quality software. 
-  Providing automated tools to developers to assess the risk of their patches is essential to help companies shipping software of higher quality. 
+  producers consider security as a first-class citizen while others do not. As 
+  mentioned in previous work, security is critical and should be considered as a 
+  default feature~\cite{10.1145/2489828.2489830,kurilova2014wyvern,mcgraw2004software}. 
+  However, the lack of experts and awareness of developers for security while 
+  producing/patching software leads companies to ship low-quality software. 
+  Providing automated tools to developers to assess the risk of their patches is 
+  essential to help companies shipping software of higher quality. 
   Bryan O'Sullivan, VP of Engineering at Facebook, advocated for new computer science
-  risk models to detect vulnerabilities in scale and predict the level of security of the software under production in his talk ```Challenges in Making Software Work at Scale''' at FaceTAV’20.
+  risk models to detect vulnerabilities in scale and predict the level of security 
+  of the software under production in his talk ```Challenges in Making Software Work at Scale''' at FaceTAV’20.
 
@@
 
-\textit{\textbf{Construct Validity:}} The formula to calculate the maintainability value ($M(v)$) was inferred based 
-on the BCH's reports. The high amount of different projects and backgrounds may require other maintainability standards. However, BCH does use a 
-representative benchmark of closed and open-source software projects to compute the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,baggen2012}.
+\textit{\textbf{Construct Validity:}} The formula to calculate the 
+maintainability value ($M(v)$) was inferred based 
+on the BCH's reports. The high amount of different projects and backgrounds 
+may require other maintainability standards. However, BCH does use a 
+representative benchmark of closed and open-source software projects to 
+compute the thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY,baggen2012}.
 Maintainability is computed as the mean of all guidelines. Different software versions 
@@
 patching vulnerabilities. This approach discards patches that were
-not explicit in commits messages. We assume that patches were performed using a single commit or several sequentially. The perspective that a 
-developer may quickly perform a patch and later proceed to the refactor is not considered.
-We assume that all patches were only performed once.
-Depending on the impact of the vulnerability in the system, some vulnerabilities may have more urgency to be patched than others.
-For instance, a vulnerability performing a Denial-of-Service attack
+not explicit in commits messages. We assume that patches were performed using 
+a single commit or several sequentially. The perspective that a 
+developer may quickly perform a patch and later proceed to the refactor is 
+not considered. We assume that all patches were only performed once. Depending 
+on the impact of the vulnerability in the system, some vulnerabilities may have 
+more urgency to be patched than others. For instance, a vulnerability performing a Denial-of-Service attack
 that usually brings entire systems down may be more urgent to patch than a 
@@
 
-\textit{\textbf{External Validity:}} The BCH tool uses private and open-source data to determine the thresholds for each guideline. We only analyze patches of open-source software.
+\textit{\textbf{External Validity:}} The BCH tool uses private and open-source 
+data to determine the thresholds for each guideline. We only analyze patches of open-source software.
 Thus, our findings may not extend to private/non-open source software. 
-Different programming languages may require different coding practices to address software safety.
-The dataset comprises more commits in Java, i.e., the dataset may not be representative of the population regarding programming languages.
+Different programming languages may require different coding practices to address 
+software safety. The dataset comprises more commits in Java, i.e., the dataset 
+may not be representative of the population regarding programming languages. 
 For both datasets, manual validation of the message of the commits was performed. 
-Only commits in English were considered.
-Thus, our approach does not consider
+Only commits in English were considered. Thus, our approach does not consider 
 patches in any other language but English.
@@
 and their differences against bug fixes~\cite{10.1145/3133956.3134072}.
-The main findings were that security patches are smaller and less complex than bug fixes and are usually performed at the function level. Our study compares
+The main findings were that security patches are smaller and less complex than bug 
+fixes and are usually performed at the function level. Our study compares
 the impact of security patches on software maintainability with the impact 
@@
 between maintainability aspects and proneness to changes/faults. In 2019, Cruz et 
-al.~\cite{8919169} proposed a formula to calculate maintainability based on the BCH's guidelines and measured the impact of energy-oriented fixes 
+al.~\cite{8919169} proposed a formula to calculate maintainability based on the 
+BCH's guidelines and measured the impact of energy-oriented fixes 
 on software maintainability. Recent work
-proposed a new maintainability model to measure fine-grained code changes by adapting/extending the BCH model~\cite{8785997}.
+proposed a new maintainability model to measure fine-grained code changes by 
+adapting/extending the BCH model~\cite{8785997}.
 Our work uses the same base model (SIG-MM) but considers a broader set of guidelines. 
@@
 Researchers investigated the relationship between design patterns and
-maintainability~\cite{10.1007/978-3-642-35267-6-18}. However, other studies show that the use of design patterns may introduce maintainability issues into
-software~\cite{4493325}.
-Yskout et.
-al did not detect if the usage of design patterns has a positive impact but concluded that developers prefer to work with the support of security patterns~\cite{8077802}.
-The present work studies how security weaknesses influence maintainability for open-source software.
+maintainability~\cite{10.1007/978-3-642-35267-6-18}. However, other studies 
+show that the use of design patterns may introduce maintainability issues into
+software~\cite{4493325}. Yskout et. al did not detect if the usage of design 
+patterns has a positive impact but concluded that developers prefer to work with 
+the support of security patterns~\cite{8077802}. The present work studies how 
+security weaknesses influence maintainability for open-source software.
 
@@
 patches on the maintainability of $260$ open-source projects. We leveraged
-Better Code Hub reports to calculate maintainability based on a model proposed in previous work~\cite{Olivari:2018,8919169}. Results show evidence of a 
-trade-off between security and maintainability, as $41.9\%$ of security patches yielded a negative impact. Hence,
+Better Code Hub reports to calculate maintainability based on a model proposed 
+in previous work~\cite{Olivari:2018,8919169}. Results show evidence of a 
+trade-off between security and maintainability, as $41.9\%$ of security patches 
+yielded a negative impact. Hence,
 developers may be hindering software maintainability while patching vulnerabilities. 
-We also observe that some guidelines and programming languages are more likely to be affected than others.
-The implications of our study are that changes to codebases while patching vulnerabilities need to be performed with extra care; tools
+We also observe that some guidelines and programming languages are more likely to 
+be affected than others. The implications of our study are that changes to codebases 
+while patching vulnerabilities need to be performed with extra care; tools
 for patch risk assessment should be integrated into the CI/CD pipeline; computer science
@@
 
+\section*{Acknowledgements}
+
+We thank SIG’s Better Code Hub team for all the support as
+well as help in validating our methodology and results; and, Pedro Adão 
+for the invaluable feedback in the early stages of the project.
+
+This work is financed by National Funds through the
+Portuguese funding agency, FCT - Fundação para a Ciência
+e a Tecnologia with reference UIDB/50021/2020, a PhD scholarship 
+(ref. SFRH/BD/143319/2019), the
+SecurityAware Project (ref. CMU/TIC/0064/2019)---also funded
+by the Carnegie Mellon Program---, and
+the FaultLocker Project (ref. PTDC/CCI-COM/29300/2017).
+
 \bibliographystyle{spmpsci}
